### 2016-02-13 docs/manuscript.Rmd
done some stuff it's good

---
title: "Determining the direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

Gibran Hemani* and George Davey Smith

MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK

\* Correspondence to: g.hemani@bristol.ac.uk

```{r setup, echo=FALSE, cache=FALSE}

suppressWarnings(suppressPackageStartupMessages(library(knitr)))
suppressWarnings(suppressPackageStartupMessages(library(pander)))
suppressWarnings(suppressPackageStartupMessages(library(captioner)))
panderOptions("table.caption.prefix", "")

read_chunk("~/repo/cit_measurement_error/scripts/mr_directionality_analysis.R")
opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE)

fig_nums <- captioner()
tab_nums <- captioner(prefix = "Table")
stab_nums <- captioner(prefix = "Supplementary Table")
sfig_nums <- captioner(prefix = "Supplementary Figure")

```

```{r tablelist, cache=FALSE}

ctab <- function(x) tab_nums(x, display="c")

tab_nums(
	"", 
	caption="", 
	display=FALSE)

```

```{r figurelist, cache=FALSE}

cfig <- function(x) fig_nums(x, display="c")

fig_nums(
	"", 
	caption="",
	display=FALSE
)

```

```{r stablelist, cache=FALSE}

cstab <- function(x) stab_nums(x, display="c")

stab_nums(
	"",
	caption="",
	display=FALSE
)

```

```{r sfigurelist, cache=FALSE}

csfig <- function(x) sfig_nums(x, display="c")

sfig_nums("", 
	caption="", 
	display=FALSE
)

```

### Abstract

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple method for inferring if an exposure is on the causal pathway to an outcome is through mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the alarming effect of increasing certainty in the wrong answer. We argue that because measurement error is ubiquitous in phenotypic data, that mediation-based causal inference should be treated with caution. Finally, we demonstrate that Mendelian randomization is a method for causal inference that is robust to measurement error.

## Introduction

With the explosion in data that measures the human phenome, of great interest is understanding the causal nature of the emerging putitive correlations between measures of interest. Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling. The principle behind mediation analysis can be explained as follows. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the causal inference test (CIT) [@Millstein2009], which have been employed by a number of recent publications make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results. One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013]. Measurement (or observational) error is the difference between a measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect sensitivity, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Sensitivity is also an issue, for example if the unit of measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error can indeed arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and obesity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically have both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

An alternative statistical method that gained prominence at least in part as a solution to the problem of measurement error [@Ashenfelter1994] is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014]. Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to proxy the exposure in an association test with the outcome. If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. This approach also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.

Here, however, arises a potential issue. Often the biological effect of a SNP is not known, and therefore it can be difficult to determine for which of the two variables in a putative exposure-outcome association is the SNP a valid instrument. By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are by design hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

Mediation based methods, for example CIT and other related approaches typically attempt to use the data to resolve this issue using the following logic: if the SNP-exposure association is stronger than the SNP-outcome association then the causal effect must be in the direction of exposure to outcome. The same logic can indeed be applied to MR, but often when applied the sensitivity of these approaches to measurement error is not considered.

When being objective one must assume that measurement error is ubiquitous, and any measured variable is only an imperfect proxy of the biological quantity that the researcher intended to obtain. Here we show using theory and simulations how measurement error can lead to unreliable causal inference in the mediation-based CIT method, and we propose an extension to MR that, although does not eliminate the problem of measurement error, provides a framework in which to ascertain the causal direction when the biology of the instrument is not fully understood.

## Methods

### Simulations

### CIT direction

### MR causal test

## Results

### The influence of measurement error on CIT

Here we consider a simple causal model where an exposure, $x$, has an effect on an outcome of interest, $y$, and that there is a known SNP, $g$, which has a direct effect on the exposure. Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$. For example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where $\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$, and $\epsilon_m$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

$$
x = \alpha_g + \beta_g + \epsilon_g
$$

where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the residual value of $x$; and the outcome is partially explained by the exposure

$$
y = \alpha_x + \beta_x + \epsilon_x
$$

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $g$ is associated with $x|y$; and 4) $g$ is independent of $y|x$. The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error is introduced we can show through simple algebra that

$$
\begin{aligned}
cov(g, y - \hat{y}) &= cov(g, y) - cov(g, \hat{y})  \\
                    &= \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
\end{aligned}
$$

where

$$
D = \frac{\beta^2_m var(x)} {\beta^2_m var(x) + var(\epsilon_m)}
$$

Thus an observational study will find $cov(g, y - \hat{y}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_m) \neq 0$) then there will remain an association between $g$ and $y|x$, which is in violation of the the 4th condition of the CIT. Note that measurement bias alone is insufficient to lead to a violation of the test statistic assumptions.

An alternative method to infer causality that is in principle robust to measurement error in the exposure variable, is Mendelian randomization (MR). The origins and applications of MR have been explained in detail, but the principle in which this is achieved is straightforward. Whereas mediation makes causal inference by contrasting the association of the instrument on the outcome before and after adjusting for the exposure, MR uses the instrument as a proxy for the exposure that is assumed to be devoid of measurement error and independent of potential confounders. In this setting, the problem is reformulated: instead of the researcher being agnostic about whether the instrument, $g$, has a direct influence on $x$ or $y$, the researcher must select an instrument which is known to have a direct effect on $x$. Thus if $g$ serves as a proxy for $x$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

$$
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
$$

which reduces to

$$
\lim_{n \to \infty} \hat{\beta}_{MR} = \frac{\beta_x cov(x + \epsilon_x)}{cov(x + \epsilon_x)} = \beta_x
$$

thus we obtain an estimate of the effect of $x_O$ on $y$ that is unbiased by measurement error in $x$.

## References

### 2016-02-13 scripts/mr_directionality_analysis.R

@@ -1,7 +1,11 @@
+
+## ---- setup ----
+
 library(dplyr)
 library(ggplot2)
 library(tidyr)
 library(psych)
+library(lattice)
 
 get_cor_from_pval <- function(p, n)
 {
@@ -10,67 +14,191 @@ get_cor_from_pval <- function(p, n)
 	return(R2)
 }
 
-load("~/repo/cit_measurement_error/results/p_comp_20160210.RData")
+
+## ---- read_parameters ----
+
+load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
 
 parameters$p_az <- -log10(parameters$p_az)
 parameters$p_bz <- -log10(parameters$p_bz)
-parameters$sig_mr <- parameters$p_bz > -log10(0.05)
-parameters$correct_direction <- parameters$p_az > parameters$p_bz
-parameters$bigna <- parameters$noisea > parameters$noiseb
-
-parameters$p_test <- parameters$p_bz
-parameters$p_test[!parameters$correct_direction] <- parameters$p_az[!parameters$correct_direction]
 parameters$r_ab <- parameters$r_ab^2
+parameters$r_ab <- round(parameters$r_ab, 1)
 parameters$r_za <- parameters$r_za^2
 parameters$noisea <- parameters$noisea^2
 parameters$noiseb <- parameters$noiseb^2
-
-parameters$rhs <- sqrt(parameters$r_ab) * parameters$cor_bbp
+parameters$rhs <- parameters$cor_ab * parameters$cor_bbp
 parameters$lhs <- parameters$cor_aap
 parameters$rhs_lhs_diff <- parameters$lhs - parameters$rhs
 
+parameters$correct_direction <- parameters$p_az > parameters$p_bz
+
+parameters$p_test <- parameters$p_bz
+parameters$p_test[!parameters$correct_direction] <- parameters$p_az[!parameters$correct_direction]
+
+# Calculate correlation test
 a <- r.test(n=parameters$n, r12=parameters$cor_zap, r13=parameters$cor_zbp, r23=parameters$cor_abp)
 parameters$cortest_t <- a$t
 parameters$cortest_p <- -log10(a$p)
[…]

### 2016-02-13 docs/manuscript.Rmd
updated

@@
 tab_nums <- captioner(prefix = "Table")
+sfig_nums <- captioner(prefix = "Supplementary Figure")
 stab_nums <- captioner(prefix = "Supplementary Table")
-sfig_nums <- captioner(prefix = "Supplementary Figure")
 
@@
 fig_nums(
-	"", 
-	caption="",
+	"cit_measurement_error_figure", 
+	caption="The CIT was performed on simulated variables where the exposure influenced the outcome and the exposure was instrumented by a SNP.
+The test statistic from CIT when testing if the exposure caused the outcome (the true model) is in red, and the test for the outcome causing the exposure (false model) is in green.
+Rows of plots represent the sample sizes used for the simulations.
+As measurement error increases (decreasing values on x-axis) the test statistic for the incorrect model gets stronger and the test statistic for the correct model gets weaker.",
+	display=FALSE
+)
+
+fig_nums(
+	"cit_mr_comparison_figure", 
+	caption="Outcomes were simulated to be causally influenced by exposures, with varying degrees of measurement error applied to both.
+CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
+The value of $d = \\rho_{x, x_O} & - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Williams' test to be more likely to be wrong about the direction of causality.
+Rows of graphs represent the sample size used in the simulations.",
 	display=FALSE
@@
 )
+
+```
+
+```{r load_up }
+
+```
+```{r read_data }
 
@@
 
-An alternative method to infer causality that is in principle robust to measurement error in the exposure variable, is Mendelian randomization (MR).
-The origins and applications of MR have been explained in detail, but the principle in which this is achieved is straightforward.
-Whereas mediation makes causal inference by contrasting the association of the instrument on the outcome before and after adjusting for the exposure, MR uses the instrument as a proxy for the exposure that is assumed to be devoid of measurement error and independent of potential confounders.
-In this setting, the problem is reformulated: instead of the researcher being agnostic about whether the instrument, $g$, has a direct influence on $x$ or $y$, the researcher must select an instrument which is known to have a direct effect on $x$.
-Thus if $g$ serves as a proxy for $x$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
+We performed simulations to verify that this problem does arise using the CIT method.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
+Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.
+
+### Inferring direction of causality using MR
+
+If one is to select an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
 Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
@@
 
+However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
+A simple approach, similar in principle to that used in the CIT, is to assume that $g$ has a direct influence on the variable $x$ if its association with $x$ is larger than its association with $y$ (or vice versa).
+This can be formally tested using the Williams' test for a difference between $\rho_{g, x_O}$ and $\rho_{g, y_O}$.
+If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Williams' test will obtain the correct result if the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0.
+
+It can be easily shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 1).
+Therefore, using the MR approach there are two potential advantages over the CIT.
+First, we can conduct a formal test for the direction of causality.
+Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome.
+
+We performed simulations to explore the performance of this approach in comparison to CIT.
+`r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ is violated the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely with increasing sample size.
+However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT.
+When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+
+## Discussion
+
+## Appendix 1
+
+$$
+\begin{aligned}
+\rho_{g, x_O} & > \rho_{g, y_O} \\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
+\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
+\end{aligned}
+$$
+
+\newpage
+
+## Tables
+
+\newpage
+
+## Figures
+
+```{r cit_measurement_error_figure }
+
+```
+
+`r fig_nums("cit_measurement_error_figure")`
+
+\newpage
+
+## Supplementary tables
+
+\newpage
+
+## Supplementary figures
+
 ## References

### 2016-02-13 scripts/mr_directionality_analysis.R

@@ -1,11 +1,10 @@
+## ---- load_up ----
 
-## ---- setup ----
-
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(psych)
-library(lattice)
+suppressPackageStartupMessages(library(dplyr))
+suppressPackageStartupMessages(library(ggplot2))
+suppressPackageStartupMessages(library(tidyr))
+suppressPackageStartupMessages(library(psych))
+suppressPackageStartupMessages(library(lattice))
 
 get_cor_from_pval <- function(p, n)
 {
@@ -15,7 +14,7 @@ get_cor_from_pval <- function(p, n)
 }
 
 
-## ---- read_parameters ----
+## ---- read_data ----
 
 load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
 
@@ -85,6 +84,40 @@ psum1 <- pl %>%
 		prop_nonsig=sum(direction_p_value <= -log10(0.05), na.rm=T)/n()
 	)
 psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=10)
+
+
+## ---- cit_measurement_error_figure ----
+
+load("~/repo/cit_measurement_error/results/20141114.RData")
+dat <- gather(dat, eval, pval, GT, TG, factor_key=TRUE)
+levels(dat$eval) <- c("Correct inference", "Incorrect inference")
+
+ggplot(dat, aes(x=sqrt(rsq),y=-log10(pval))) +
+geom_point(aes(colour=eval)) +
+labs(y=expression(-log[10]*p), x=expression(cor(x, x[O])), colour="") +
+facet_grid(n ~ ., scale="free_y")
+
+
+
+## ---- cit_mr_comparison_figure ----
+
+psum2 <- gather(psum1, eval, value, prop_sig_correct, prop_sig_incorrect, prop_nonsig, factor_key=TRUE) %>%
+	group_by(n, rhs_lhs_diff_bin, r_za, eval, test) %>%
+	summarise(value=mean(value, na.rm=TRUE))	
+
+temp <- do.call(rbind, strsplit(as.character(psum2$rhs_lhs_diff_bin), split=","))
+psum2$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp[,1]))
+psum2$rhs_lhs_diff_bin_lab <- as.factor(paste0("d = ", psum2$rhs_lhs_diff_bin_numeric))
[…]

### 2016-02-14 docs/manuscript.Rmd
updates

@@
 fig_nums(
+	"d_relationship_figure", 
+	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.",
+	display=FALSE
+)
+
+fig_nums(
 	"cit_mr_comparison_figure", 
@@
 CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
-The value of $d = \\rho_{x, x_O} & - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Williams' test to be more likely to be wrong about the direction of causality.
+The value of $d = \\rho_{x, x_O} - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality.
 Rows of graphs represent the sample size used in the simulations.",
@@
 
+Supposing there are two correlated variables, $x$ and $y$, with a variant $g$ associated with both.
+Assuming no horizontal pleiotropy (i.e.
+$g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on.
+This can be achieved by testing for difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population.
+It is calculated as
+
+$$
+Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
+$$
+
+where Fisher's z-transformation is used to obtain $Z_{g*} = \frac{1}{2} \ln  \left (  \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )$,
+
+$$
+h = \frac{1 - (frm^2)} {1 - rm^2}
+$$
+
+where
+
+$$
+f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
+$$
+
+and
+
+$$
+rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
+$$
+
 ## Results
@@
 $$
-x = \alpha_g + \beta_g + \epsilon_g
+x = \alpha_g + \beta_g g + \epsilon_g
 $$
@@
 $$
-y = \alpha_x + \beta_x + \epsilon_x
+y = \alpha_x + \beta_x x + \epsilon_x
 $$
@@
 The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$.
-When measurement error is introduced we can show through simple algebra that
+When measurement error is introduced we can show using basic covariance properties (Appendix 1) that
 
@@
 \begin{aligned}
-cov(g, y - \hat{y}) &= cov(g, y) - cov(g, \hat{y})  \\
-                    &= \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
+cov(g, y - \hat{y}) & = cov(g, y) - cov(g, \hat{y})  \\
+                    & = \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
 \end{aligned}
@@
 However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
-A simple approach, similar in principle to that used in the CIT, is to assume that $g$ has a direct influence on the variable $x$ if its association with $x$ is larger than its association with $y$ (or vice versa).
-This can be formally tested using the Williams' test for a difference between $\rho_{g, x_O}$ and $\rho_{g, y_O}$.
-If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Williams' test will obtain the correct result if the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0.
-
-It can be easily shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 1).
+If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result if the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0.
+It can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2).
 Therefore, using the MR approach there are two potential advantages over the CIT.
@@
 Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome.
+`r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very high, the condition of $d > 0$ is satisfied.
 
 We performed simulations to explore the performance of this approach in comparison to CIT.
-`r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ is violated the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely with increasing sample size.
+`r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
 However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT.
@@
 ## Appendix 1
+
+Given the following model
+
+$$
+x = \alpha_g + \beta_g g + \epsilon_g \\
+x_O = \alpha_m + \beta_m x + \epsilon_m \\
+y = \alpha_x + \beta_x x + \epsilon_x
+$$
+
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, and $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
+Under the CIT, this is expected to be $cov(g, y - \hat{y}) = 0$ when $x$ causes $y$, where $\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}$ is the predicted value of $y$ using the measured value of $x_O$.
+
+We can split $cov(g, y - \hat{y})$ into two parts, $cov(g, y)$ and $cov(g, \hat{y})$.
+
+**Part 1**
+
+$$
+\begin{aligned}
+cov(g, y) & = cov(g, \beta_x x) \\
+          & = cov(g, \beta_x \beta_g g) \\
+          & = \beta_x\beta_g var(g)
+\end{aligned}
+$$
+
+**Part 2**
+
+$$
+\begin{aligned}
+cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
+                & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
+                & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
+                & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
+\end{aligned}
+$$
+
+Simpifying further
+
+$$
+\begin{aligned}
+\hat{\beta}_{x_O} & = \frac{cov(y, x_O)} {var(x_O)} \\
+                  & = \frac{\beta_m cov(y, x)} {var(x_O)} \\
+                  & = \frac{\beta_m \beta_x var(x)} {var(x_O)} \\
+                  & = \frac{\beta_m \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
+\end{aligned}
+$$
+
+which can be substituted back to give
+
+$$
+\begin{aligned}
+cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                & = D\beta_x\beta_g var(g)
+\end{aligned}
+$$
+
+where
+
+$$
+D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
+$$
+
+## Appendix 2
+
+Steiger test is used to infer if the variant $g$ has a direct influence on $x$ or $y$ when it is known that it associates with both, but the direction of causality between $x$ and $y$ is unknown.
+Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:
+
+$$
+x = \alpha_1 + \beta_1 g + e_1
+$$
+
+for the first stage and
+
+$$
+y = \alpha_2 + \beta_2 \hat{x} + e_2
+$$
+
+where $\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g$.
+Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
+In the presence of measurement error in $x$ and $y$, however, Steiger test will instead be assessing the inequality $\rho_{g, x_O} > \rho_{g, y_O}$, which can be simplified:
 
@@
 
-```{r cit_measurement_error_figure }
+```{r cit_measurement_error_figure, fig.width=10, fig.height=10 }
 
@@
 `r fig_nums("cit_measurement_error_figure")`
+
+\newpage
+
+```{r d_relationship_figure, fig.width=10, fig.height=6 }
+
+```
+
+`r fig_nums("d_relationship_figure")`
+
+\newpage
+
+```{r cit_mr_comparison_figure, fig.width=10, fig.height=10 }
+
+```
+
+`r fig_nums("cit_mr_comparison_figure")`
 

### 2016-02-14 docs/manuscript.tex
updates

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Determining the direction of causality in the face of measurement error},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Determining the direction of causality in the face of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{14 February 2016}

\begin{document}

\maketitle

Gibran Hemani* and George Davey Smith

MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
School of Social and Community Medicine, Bristol, UK

* Correspondence to:
\href{mailto:g.hemani@bristol.ac.uk}{\nolinkurl{g.hemani@bristol.ac.uk}}

\subsubsection{Abstract}\label{abstract}

With our ability to characterize the human phenome ever improving it is
becoming increasingly common to use statistical tests to infer the
causal relationships between correlated variables. A simple method for
inferring if an exposure is on the causal pathway to an outcome is
through mediation analysis, where the effect of an instrument on the
outcome is tested before and after adjusting for the exposure. We show
that in the face of measurement error this can lead to erroneous
results, and that increasing sample size, rather than reducing bias, can
often have the alarming effect of increasing certainty in the wrong
answer. We argue that because measurement error is ubiquitous in
phenotypic data, that mediation-based causal inference should be treated
with caution. Finally, we demonstrate that Mendelian randomization is a
method for causal inference that is robust to measurement error.

\subsection{Introduction}\label{introduction}

With the explosion in data that measures the human phenome, of great
interest is understanding the causal nature of the emerging putitive
correlations between measures of interest. Mediation is a commonly used
statistical method for making causal inference in observational
data\textsuperscript{1--4}, and it exists in many forms, from simple
regression based systems to structural equation modeling. The principle
behind mediation analysis can be explained as follows. Supposing an
exposure has a causal effect on an outcome, then an `instrument' that
causes the exposure (a single nucleotide polymorophism (SNP), for
example) should also influence the outcome. Therefore the influence of
the instrument on the outcome conditional on the exposure should be
zero. This forms the basis of a number of methods, such as the causal
inference test (CIT)\textsuperscript{1}, which have been employed by a
number of recent publications make causal inferences, often in large
scale `omics datasets\textsuperscript{5--7}.

While mediation based approaches are simple and convenient to implement,
it is important to note where they may lead to unreliable results. One
such mechanism is measurement error\textsuperscript{8--11}. Measurement
(or observational) error is the difference between a measured value of a
quantity and its true value. Such variability can arise through a whole
plethora of mechanisms, which are often unique to the study design and
difficult to avoid\textsuperscript{12,13}. Array technology is now
commonly used to obtain high throughput phenotyping at low cost, but
they come with the problem of having imperfect sensitivity, so for
example methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{14,15}. Sensitivity is also an issue, for example
if the unit of measurement of biological interest is the methylation
level in a T cell, then measurement error of this value can be
introduced by using methylation levels from whole blood samples because
the measured value will be an assay of many cell
types\textsuperscript{16}.

Measurement error can indeed arise in more low-tech data too, for
example when measuring body mass index (BMI) one is typically interested
in using this as a proxy for obesity, but it is clear that the
correlation between BMI and obesity is not perfect\textsuperscript{17}.
A similar problem of biological misspecification is unavoidable in
disease diagnosis. Measurement error can also be introduced after the
data has been collected, for example the transformation of non-normal
data for the purpose of statistical analysis will lead to a new variable
that will typically have both bias and imprecision compared to the
original variable. The sources of measurement error are not limited to
this list\textsuperscript{13}, and its impact has been explored in the
context of mediation analysis in the epidemiological literature
extensively\textsuperscript{8,11}.

An alternative statistical method that gained prominence at least in
part as a solution to the problem of measurement
error\textsuperscript{18} is Mendelian randomisation
(MR)\textsuperscript{19,20}. Here an instrumental variable (typically a
SNP) that has a robost, causal association with the exposure is used to
proxy the exposure in an association test with the outcome. If there is
no measurement error in the SNP then the causal association between the
exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure. This approach also has the useful property that it guards
against unmeasured confounders driving the association between exposure
and outcome; and if the biological effect of the SNP is understood then
it can also aid against reverse causality driving the association.

Here, however, arises a potential issue. Often the biological effect of
a SNP is not known, and therefore it can be difficult to determine for
which of the two variables in a putative exposure-outcome association is
the SNP a valid instrument. By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are by
design hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{21}. Another situation is where the
causal direction between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{7}?

Mediation based methods, for example CIT and other related approaches
typically attempt to use the data to resolve this issue using the
following logic: if the SNP-exposure association is stronger than the
SNP-outcome association then the causal effect must be in the direction
of exposure to outcome. The same logic can indeed be applied to MR, but
often when applied the sensitivity of these approaches to measurement
error is not considered.

When being objective one must assume that measurement error is
ubiquitous, and any measured variable is only an imperfect proxy of the
biological quantity that the researcher intended to obtain. Here we show
using theory and simulations how measurement error can lead to
unreliable causal inference in the mediation-based CIT method, and we
propose an extension to MR that, although does not eliminate the problem
of measurement error, provides a framework in which to ascertain the
causal direction when the biology of the instrument is not fully
understood.

\subsection{Methods}\label{methods}

\subsubsection{Simulations}\label{simulations}

\subsubsection{CIT direction}\label{cit-direction}

\subsubsection{MR causal test}\label{mr-causal-test}

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population. It is calculated as

\[
Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
\]

where Fisher's z-transformation is used to obtain
\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),

\[
h = \frac{1 - (frm^2)} {1 - rm^2}
\]

where

\[
f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
\]

and

\[
rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
\]

\subsection{Results}\label{results}

\subsubsection{The influence of measurement error on
CIT}\label{the-influence-of-measurement-error-on-cit}

Here we consider a simple causal model where an exposure, \(x\), has an
effect on an outcome of interest, \(y\), and that there is a known SNP,
\(g\), which has a direct effect on the exposure. Measurement error of
an exposure can be modeled as some transformation of the true value that
leads to the observed value, \(x_O=f(x)\). For example, we can define
\(f(x) = \alpha_m + \beta_m x + \epsilon_m\), where \(\alpha_m\) and
\(\beta_m\) influence the bias in the measurement of \(x\), and
\(\epsilon_m\) represents the imprecision in the measurement of \(x\).
Here the true value of the exposure is partially explained by the
genetic instrument, \(g\), such that

\[
x = \alpha_g + \beta_g g + \epsilon_g
\]

where \(\beta_g\) is the effect of the SNP on the exposure, and
\(\epsilon_g\) is the residual value of \(x\); and the outcome is
partially explained by the exposure

\[
y = \alpha_x + \beta_x x + \epsilon_x
\]

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(g\) is associated with \(x|y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition is necessary for causal
inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement error
is introduced we can show using basic covariance properties (Appendix 1)
that

\[
\begin{aligned}
cov(g, y - \hat{y}) & = cov(g, y) - cov(g, \hat{y})  \\
                    & = \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
\end{aligned}
\]

where

\[
D = \frac{\beta^2_m var(x)} {\beta^2_m var(x) + var(\epsilon_m)}
\]

Thus an observational study will find \(cov(g, y - \hat{y}) = 0\) when
the true model is causal only when \(D = 1\). Therefore, if there is any
measurement error that incurs imprecision (i.e.
\(var(\epsilon_m) \neq 0\)) then there will remain an association
between \(g\) and \(y|x\), which is in violation of the the 4th
condition of the CIT. Note that measurement bias alone is insufficient
to lead to a violation of the test statistic assumptions.

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure (\(\rho_{x, x_O}=1\)) the CIT is reliable in identifying
the correct causal direction. However, as measurement error increases,
eventually the CIT is more likely to infer a robust causal association
in the wrong direction. Also of concern here is that increasing sample
size does not solve the issue, indeed it only strengthens the evidence
of the incorrect inference.

\subsubsection{Inferring direction of causality using
MR}\label{inferring-direction-of-causality-using-mr}

If one is to select an instrument \(g\) that has a direct influence on
\(x\) and \(x\) is causally related to \(y\), one can reason that the
influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as

\[
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
\]

which reduces to

\[
\lim_{n \to \infty} \hat{\beta}_{MR} = \frac{\beta_x cov(x + \epsilon_x)}{cov(x + \epsilon_x)} = \beta_x
\]

thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
unbiased by measurement error in \(x\).

However, if we do not know whether the SNP \(g\) has a direct influence
on \(x\) or \(y\) then further efforts are required. If \(x\) causes
\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
Steiger test will obtain the correct result if the value
\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0. It can be shown
that in the presence of measurement error,
\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2). Therefore,
using the MR approach there are two potential advantages over the CIT.
First, we can conduct a formal test for the direction of causality.
Second the correct direction of causality can be inferred when
measurement error in the exposure variable is present, as long as it is
lower than the product of the measurement error in the outcome and the
causal correlation between the exposure and the outcome. Figure 2 shows
that in most cases, especially when the causal effect between \(x\) and
\(y\) is not very high, the condition of \(d > 0\) is satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
MR analysis is liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, in all cases the MR analysis infers the wrong
direction of causality at the same or a lower rate than the CIT. When
\(d > 0\) is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality.

\subsection{Discussion}\label{discussion}

\subsection{Appendix 1}\label{appendix-1}

Given the following model

\[
x = \alpha_g + \beta_g g + \epsilon_g \\
x_O = \alpha_m + \beta_m x + \epsilon_m \\
y = \alpha_x + \beta_x x + \epsilon_x
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), and \(x_O\) is the measured quantity
of \(x\), where measurement error is incurred from bias in \(\alpha_m\)
and \(\beta_m\) and imprecision from \(\epsilon_m\), our objective is to
estimate the expected magnitude of association between \(g\) and \(y\)
after conditioning on \(x\). Under the CIT, this is expected to be
\(cov(g, y - \hat{y}) = 0\) when \(x\) causes \(y\), where
\(\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}\) is the
predicted value of \(y\) using the measured value of \(x_O\).

We can split \(cov(g, y - \hat{y})\) into two parts, \(cov(g, y)\) and
\(cov(g, \hat{y})\).

\textbf{Part 1}

\[
\begin{aligned}
cov(g, y) & = cov(g, \beta_x x) \\
          & = cov(g, \beta_x \beta_g g) \\
          & = \beta_x\beta_g var(g)
\end{aligned}
\]

\textbf{Part 2}

\[
\begin{aligned}
cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
\end{aligned}
\]

Simpifying further

\[
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y, x_O)} {var(x_O)} \\
                  & = \frac{\beta_m cov(y, x)} {var(x_O)} \\
                  & = \frac{\beta_m \beta_x var(x)} {var(x_O)} \\
                  & = \frac{\beta_m \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\end{aligned}
\]

which can be substituted back to give

\[
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = D\beta_x\beta_g var(g)
\end{aligned}
\]

where

\[
D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\]

\subsection{Appendix 2}\label{appendix-2}

Steiger test is used to infer if the variant \(g\) has a direct
influence on \(x\) or \(y\) when it is known that it associates with
both, but the direction of causality between \(x\) and \(y\) is unknown.
Assuming the causal direction is \(x \to y\), two stage MR is formulated
using the following regression models:

\[
x = \alpha_1 + \beta_1 g + e_1
\]

for the first stage and

\[
y = \alpha_2 + \beta_2 \hat{x} + e_2
\]

where \(\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1). In the presence of measurement error in \(x\) and \(y\),
however, Steiger test will instead be assessing the inequality
\(\rho_{g, x_O} > \rho_{g, y_O}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
\]

\newpage

\subsection{Tables}\label{tables}

\newpage

\subsection{Figures}\label{figures}

\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}\\
Figure 1: The CIT was performed on simulated variables where the
exposure influenced the outcome and the exposure was instrumented by a
SNP. The test statistic from CIT when testing if the exposure caused the
outcome (the true model) is in red, and the test for the outcome causing
the exposure (false model) is in green. Rows of plots represent the
sample sizes used for the simulations. As measurement error increases
(decreasing values on x-axis) the test statistic for the incorrect model
gets stronger and the test statistic for the correct model gets weaker.

\newpage

\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
Figure 2: Plots depicting the function
\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).

\newpage

\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}\\
Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations.

\newpage

\subsection{Supplementary tables}\label{supplementary-tables}

\newpage

\subsection{Supplementary figures}\label{supplementary-figures}

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
molecular relationships with a causal inference test. \emph{BMC
genetics} \textbf{10,} 23 (2009).

2.Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
causal associations between gene expression and disease. \emph{Nature
Genetics} \textbf{37,} 710--717 (2005).

3.Gayan, J., Gonzalez-Perez, A. \& Ruiz, A. `Does replication groups
scoring reduce false positive rate in SNP interaction discovery?
Response'. \emph{BMC Genomics} \textbf{11,} 403 (2010).

4.Tavtigian, S. V. \emph{et al.} Rare, Evolutionarily Unlikely Missense
Substitutions in ATM Confer Increased Risk of Breast Cancer. \emph{The
American Journal of Human Genetics} \textbf{85,} 427--446 (2009).

5.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
epigenetic marks that mediate genetic risk for epithelial ovarian
cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).

6.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
methylation as an intermediary of genetic risk in rheumatoid arthritis.
\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).

7.Waszak, S. M. \emph{et al.} Variation and genetic control of chromatin
architecture in humans. \emph{Cell} \textbf{162,} 1039--1050 (2015).

8.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
Vandenbroucke, J. P. Quantification of bias in direct effects estimates
due to different types of measurement error in the mediator.
\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).

9.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
inference. \emph{PloS one} \textbf{8,} e80735 (2013).

10.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
covariate adjustment for estimating causal effects. \emph{Proceedings of
the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
(UAI-10)} 527--536 (2010).

11.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
mediator matters when estimating indirect effects. \emph{Journal of
epidemiology and community health} \textbf{67,} 458--66 (2013).

12.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
3--34 (2011).

13.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
measurement bias. \emph{American journal of epidemiology} \textbf{170,}
959--62; discussion 963--4 (2009).

14.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
pathway analysis: two potential perils in cancer studies involving DNA
methylation array analysis. \emph{Cancer epidemiology, biomarkers \&
prevention : a publication of the American Association for Cancer
Research, cosponsored by the American Society of Preventive Oncology}
\textbf{22,} 1052--60 (2013).

15.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
polymorphic CpGs in the Illumina Infinium HumanMethylation450
microarray. \emph{Epigenetics : official journal of the DNA Methylation
Society} \textbf{8,} 203--9 (2013).

16.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
measures of cell mixture distribution. \emph{BMC bioinformatics}
\textbf{13,} 86 (2012).

17.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
\textbf{341,} 856--8 (2013).

18.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
Schooling from a New Sample of Twins. \emph{The American Economic
Review} \textbf{84,} 1157--1173 (1994).

19.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
potentials, and limitations. \emph{International journal of
epidemiology} \textbf{33,} 30--42 (2004).

20.{Davey Smith}, G. \& Hemani, G. Mendelian randomization: genetic
anchors for causal inference in epidemiological studies. \emph{Human
molecular genetics} \textbf{23,} R89-----R98 (2014).

21.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
Adipocyte Browning in Humans. \emph{The New England journal of medicine}
\textbf{373,} 895--907 (2015).

\end{document}

### 2016-02-14 scripts/mr_directionality_analysis.R

@@ -5,6 +5,7 @@ suppressPackageStartupMessages(library(ggplot2))
 suppressPackageStartupMessages(library(tidyr))
 suppressPackageStartupMessages(library(psych))
 suppressPackageStartupMessages(library(lattice))
+suppressPackageStartupMessages(library(latex2exp))
 
 get_cor_from_pval <- function(p, n)
 {
@@ -92,11 +93,29 @@ load("~/repo/cit_measurement_error/results/20141114.RData")
 dat <- gather(dat, eval, pval, GT, TG, factor_key=TRUE)
 levels(dat$eval) <- c("Correct inference", "Incorrect inference")
 
-ggplot(dat, aes(x=sqrt(rsq),y=-log10(pval))) +
+ggplot(subset(dat, n %in% c(100, 1000, 10000)), aes(x=sqrt(rsq),y=-log10(pval))) +
 geom_point(aes(colour=eval)) +
 labs(y=expression(-log[10]*p), x=expression(cor(x, x[O])), colour="") +
 facet_grid(n ~ ., scale="free_y")
 
+
+## ---- d_relationship_figure ----
+
+ineq <- expand.grid(
+	cora = seq(0, 1, by=0.02),
+	corb = seq(0, 1, by=0.1),
+	ab = c(0.1, 0.5, 0.9)
+)
+
+ineq$lhs <- ineq$cora
+ineq$rhs <- ineq$corb * ineq$ab
+ineq$d <- ineq$lhs - ineq$rhs
+ineq$ablab <- paste0("cor(x,y) = ", ineq$ab)
+
+ggplot(ineq, aes(x=cora, y=d)) +
+geom_line(aes(colour=corb, group=corb)) +
+facet_wrap(~ ablab) +
+labs(x=TeX("$cor(x, x_O)$"), y="d", colour=TeX("$cor(y, y_O)$"))
 
 
 ## ---- cit_mr_comparison_figure ----
@@ -110,13 +129,14 @@ psum2$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp[,1]))
 psum2$rhs_lhs_diff_bin_lab <- as.factor(paste0("d = ", psum2$rhs_lhs_diff_bin_numeric))
 psum2$rhs_lhs_diff_bin_lab <- factor(psum2$rhs_lhs_diff_bin_lab, levels=levels(psum2$rhs_lhs_diff_bin_lab)[order(as.numeric(as.character(levels(psum2$rhs_lhs_diff_bin_numeric))))])
 
[…]

### 2016-02-14 scripts/theory.R

@@ -1,112 +1,48 @@
-n <- 10000
-gx <- rbinom(n, 2, 0.3)
-gz <- rbinom(n, 2, 0.3)
-u1 <- rnorm(n)
-u2 <- rnorm(n)
-u3 <- rnorm(n)
-ex <- rnorm(n)
-ez <- rnorm(n)
-ey <- rnorm(n)
-x <- 0.3 * gx + u1 + u2 + ex
-z <- 0.5 * gz + x + u1 + u3 + ez
-y <- x + z + u2 + u3 + ey
+nsim <- 100
 
+c_gy <- array(0, nsim)
+c_gyhat <- array(0, nsim)
+c_gy_exp <- array(0, nsim)
+c_gyhat_exp <- array(0, nsim)
 
+n <- 100000
 
-
-standardise <- function(x) 
+for(i in 1:nsim)
 {
-	a <- mean(x)
-	return((x - mean(x)) / sd(x) + a)
+	g <- rbinom(n, 2, 0.5)
+
+	a_g <- runif(1)
+	b_g <- runif(1)
+	e_g <- runif(1)
+
+	a_m <- runif(1)
+	b_m <- runif(1)
+	e_m <- runif(1)
+
+	a_x <- runif(1)
+	b_x <- runif(1)
+	e_x <- runif(1)
+
+	x <- a_g + b_g * g + rnorm(n, sd=sqrt(e_g))
+	x0 <- a_m + b_m * x + rnorm(n, sd=sqrt(e_m))
+	y <- a_x + b_x * x + rnorm(n, sd=sqrt(e_x))
+	yhat <- fitted.values(lm(y ~ x0))
+
+	c_gy[i] <- cov(g, y)
+	c_gy_exp[i] <- b_x * b_g * var(g)
+
+	c_gyhat[i] <- cov(g, yhat)
+	c_gyhat_exp[i] <- b_x * b_g * var(g) * b_m^2 * var(x) / (b_m^2 * var(x) + e_m)
 }
 
 
-n <- 100000
-g <- rbinom(n, 2, 0.3)
-e1 <- rnorm(n)
-e2 <- rnorm(n)
-e3 <- rnorm(n)
-x1 <- standardise(0.4 * g + e1)
-x2 <- standardise(1 * x1 + e2)
-y <- 0.7 * x1 + e3
+dat <- rbind(
[…]

### 2016-02-14 docs/manuscript.Rmd
updated

@@
 We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the alarming effect of increasing certainty in the wrong answer.
-We argue that because measurement error is ubiquitous in phenotypic data, that mediation-based causal inference should be treated with caution.
-Finally, we demonstrate that Mendelian randomization is a method for causal inference that is robust to measurement error.
+Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible.
+We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
+We argue that because measurement error is ubiquitous in phenotypic data, mediation-based causal inference should be treated with caution.
 
@@
 
-### Simulations
+The analysis proceeds by simulating variables that have known causal effects and valid instruments, and then applying measurement error in different configurations.
+We then assess the performance of the CIT test and MR analysis for a) inference of a causal relationship and b) inferring the correct direction of causality.
+Included in the MR analysis is a simple extension that is used to make inference of the causal direction when it is unknown which of the correlated variables the instrument is directly acting upon.
+All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]().
 
-### CIT direction
+### CIT test
+
+The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
+The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$.
+To infer the direction of causality using the CIT method, the omnibus p-value was estimated for $x$ causing $y$, and for $y$ causing $x$, and the model that returned the lowest p-value was taken as the one denoting the correct causal direction.
 
@@
 $g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on.
-This can be achieved by testing for difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population.
+This can be achieved by testing for a difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
 It is calculated as
@@
 
+The $Z$ value is interpreted such that
+
+$$
+Z \left\{
+\begin{array}{ll}
+> 0, & x \to y\\
+< 0, & y \to x\\
+= 0, & x \perp\!\!\!\perp y 
+\end{array} \right.
+$$
+
+and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$.
+The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test.
+For simplicity, the test is deemed to have obtained a robust association if the Z score test has a p-value $<  0.05$ and the two sample MR analysis has a p-value $< 0.05$.
+Note that the same approach can be applied to a two-sample MR setting, where
+
+### Simulations
+
+Simulations were conducting by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_O$, the outcome $y$, the measured values of the outcome $y_O$ and the instrument $g$.
+In all models $x$ causes $y$ and $g$ is an instrument for $x$.
+Each variable was simulated such that:
+
+$$
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \beta_g g + \epsilon_g \\
+x_O & = \beta_{mx} x + \epsilon_{mx} \\
+y & = \beta_x x + \epsilon_x \\
+y_O & = \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+$$
+
+All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated such that
+
+$$
+\begin{aligned}
+cor(g, x)^2 & = 0.1 \\
+cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
+var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+n & = \{100, 1000, 10000\}
+\end{aligned}
+$$
+
+giving a total of 432 combinations of parameters.
+Each of these sets of variables was performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simualted variables.
+Similar patterns of results were obtained for different values of $cor(g, x)$.
+
 ## Results
@@
 
-Here we consider a simple causal model where an exposure, $x$, has an effect on an outcome of interest, $y$, and that there is a known SNP, $g$, which has a direct effect on the exposure.
 Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$.
@@
 We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction.
-However, as measurement error increases, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
 Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.
@@
 
-If one is to select an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
+When selecting an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
 Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
@@
 $$
-\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
+\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
 $$
@@
 However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
-If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result if the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0.
-It can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2).
-Therefore, using the MR approach there are two potential advantages over the CIT.
+If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$.
+But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0.
+However, using the MR approach there are two potential advantages over the CIT.
 First, we can conduct a formal test for the direction of causality.
 Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome.
-`r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very high, the condition of $d > 0$ is satisfied.
+`r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
 
@@
 However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT.
-When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+When $d > 0$ is satisfied we observe that in many cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
 
 ## Discussion
+
+Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and unfortunately there are no solutions that work perfectly in all scenarios.
+In the face of measurement error it is evident that causal inference drawn through mediation analysis will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error for most experimental designs.
+
+In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis.
+Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality.
+We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it increases the significance of the incorrect inference.
+
+Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods.
+Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces.
+This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest.
+But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
+Here we have described a simple extension to MR which can be used as an alternative to mediation based methods.
+We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages.
+First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality.
+Second, after testing in a comprehensive range of scenarios the MR based approach is much less likely to infer the wrong direction of causality compared to CIT.
+
+There are a number of limitations to the analysis that we performed.
+We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
+
+The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly liable to measurement error.
+
+Limitations
+
+Pleiotropy
 

### 2016-02-14 docs/manuscript.tex
updated

@@
 often have the alarming effect of increasing certainty in the wrong
-answer. We argue that because measurement error is ubiquitous in
-phenotypic data, that mediation-based causal inference should be treated
-with caution. Finally, we demonstrate that Mendelian randomization is a
-method for causal inference that is robust to measurement error.
+answer. Commonly used tools that are predicated on this methodology such
+as the causal inference test (CIT) are shown to be susceptible. We
+demonstrate that Mendelian randomization (MR) is a method for causal
+inference that is robust to measurement error, and that a simple
+extension to MR can provide a method to estimate the direction of
+causality between correlated variables even when the biology of the
+instrument is unknown. We argue that because measurement error is
+ubiquitous in phenotypic data, mediation-based causal inference should
+be treated with caution.
 
@@
 
-\subsubsection{Simulations}\label{simulations}
+\subsubsection{CIT test}\label{cit-test}
 
-\subsubsection{CIT direction}\label{cit-direction}
+The CIT method\textsuperscript{1} is implemented in the R package
+\emph{R/cit}\textsuperscript{22}. The \emph{cit.cp} function was used to
+obtain an omnibus p-value of causality between an exposure \(x\) and an
+outcome \(y\) using a mediator \(g\). To infer the direction of
+causality using the CIT method, the omnibus p-value was estimated for
+\(x\) causing \(y\), and for \(y\) causing \(x\), and the model that
+returned the lowest p-value was taken as the one denoting the correct
+causal direction.
 
@@
 variable) it is desirable to know which of the variables it has a direct
-influence on. This can be achieved by testing for difference in the
+influence on. This can be achieved by testing for a difference in the
 correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a population. It is calculated as
+for correlated correlations within a
+population\textsuperscript{\textbf{???}}. It is calculated as
 
@@
 
+The \(Z\) value is interpreted such that
+
+\[
+Z \left\{
+\begin{array}{ll}
+> 0, & x \to y\\
+< 0, & y \to x\\
+= 0, & x \perp\!\!\!\perp y 
+\end{array} \right.
+\]
+
+and a p-value is generated from the \(Z\) value to indicate confidence
+of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
+for a causal association is then obtained by performing two sample MR in
+the direction that is predicted to be the correct model based on the Z
+score test. For simplicity, the test is deemed to have obtained a robust
+association if the Z score test has a p-value \(< 0.05\) and the two
+sample MR analysis has a p-value \(< 0.05\). Note that the same approach
+can be applied to a two-sample MR setting, where
+
+\subsubsection{Simulations}\label{simulations}
+
+Simulations were conducting by creating variables of sample size \(n\)
+for the exposure \(x\), the measured values of the exposure \(x_O\), the
+outcome \(y\), the measured values of the outcome \(y_O\) and the
+instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
+instrument for \(x\). Each variable was simulated such that:
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \beta_g g + \epsilon_g \\
+x_O & = \beta_{mx} x + \epsilon_{mx} \\
+y & = \beta_x x + \epsilon_x \\
+y_O & = \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+\]
+
+All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
+generated such that
+
+\[
+\begin{aligned}
+cor(g, x) & = \{0.01, 0.05, 0.1\} \\
+cor(x, y) & = \{0.2, 0.4, 0.6, 0.8\} \\
+var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+n & = \{100, 1000, 10000\}
+\end{aligned}
+\]
+
+giving a total of 432 combinations of parameters. Each of these sets of
+variables was performed 100 times, and the CIT and MR methods were
+applied to each in order to evaluate the causal association of the
+simualted variables.
+
 \subsection{Results}\label{results}
@@
 
-Here we consider a simple causal model where an exposure, \(x\), has an
-effect on an outcome of interest, \(y\), and that there is a known SNP,
-\(g\), which has a direct effect on the exposure. Measurement error of
-an exposure can be modeled as some transformation of the true value that
-leads to the observed value, \(x_O=f(x)\). For example, we can define
-\(f(x) = \alpha_m + \beta_m x + \epsilon_m\), where \(\alpha_m\) and
-\(\beta_m\) influence the bias in the measurement of \(x\), and
-\(\epsilon_m\) represents the imprecision in the measurement of \(x\).
-Here the true value of the exposure is partially explained by the
-genetic instrument, \(g\), such that
+Measurement error of an exposure can be modeled as some transformation
+of the true value that leads to the observed value, \(x_O=f(x)\). For
+example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
+where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
+of \(x\), and \(\epsilon_m\) represents the imprecision in the
+measurement of \(x\). Here the true value of the exposure is partially
+explained by the genetic instrument, \(g\), such that
 
@@
 the CIT method. Figure 1 shows that when there is no measurement error
-in the exposure (\(\rho_{x, x_O}=1\)) the CIT is reliable in identifying
-the correct causal direction. However, as measurement error increases,
-eventually the CIT is more likely to infer a robust causal association
-in the wrong direction. Also of concern here is that increasing sample
-size does not solve the issue, indeed it only strengthens the evidence
-of the incorrect inference.
+in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
+reliable in identifying the correct causal direction. However, as
+measurement error increases in the exposure variable, eventually the CIT
+is more likely to infer a robust causal association in the wrong
+direction. Also of concern here is that increasing sample size does not
+solve the issue, indeed it only strengthens the evidence of the
+incorrect inference.
 
@@
 
-If one is to select an instrument \(g\) that has a direct influence on
-\(x\) and \(x\) is causally related to \(y\), one can reason that the
+When selecting an instrument \(g\) that has a direct influence on \(x\)
+and \(x\) is causally related to \(y\), one can reason that the
 influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
@@
 \[
-\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O,g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
+\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
 \]
@@
 \(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
-Steiger test will obtain the correct result if the value
-\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0. It can be shown
-that in the presence of measurement error,
-\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2). Therefore,
-using the MR approach there are two potential advantages over the CIT.
-First, we can conduct a formal test for the direction of causality.
-Second the correct direction of causality can be inferred when
-measurement error in the exposure variable is present, as long as it is
-lower than the product of the measurement error in the outcome and the
-causal correlation between the exposure and the outcome. Figure 2 shows
-that in most cases, especially when the causal effect between \(x\) and
-\(y\) is not very high, the condition of \(d > 0\) is satisfied.
+Steiger test will obtain the correct result because the value
+\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0 when the causal
+association between \(x\) and \(y\) is \(\rho_{x, y} < 1\). But it can
+be shown that in the presence of measurement error,
+\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), thus it is
+important to note that under certain conditions when measurement error
+is present the Steiger test could make erroneous inference about causal
+direction because the value of \(d\) is not restricted to being greater
+than 0. However, using the MR approach there are two potential
+advantages over the CIT. First, we can conduct a formal test for the
+direction of causality. Second the correct direction of causality can be
+inferred when measurement error in the exposure variable is present, as
+long as it is lower than the product of the measurement error in the
+outcome and the causal correlation between the exposure and the outcome.
+Figure 2 shows that in most cases, especially when the causal effect
+between \(x\) and \(y\) is not very large, the condition of \(d > 0\) is
+satisfied.
 
@@
 direction of causality at the same or a lower rate than the CIT. When
-\(d > 0\) is satisfied we observe that in most cases the MR method has
+\(d > 0\) is satisfied we observe that in many cases the MR method has
 greater power to obtain evidence for causality than CIT, and always
@@
 \subsection{Discussion}\label{discussion}
+
+Researchers are often confronted with the problem of making causal
+inferences using a statistical framework on observational data, and
+unfortunately there are no solutions that work perfectly in all
+scenarios. In the face of measurement error it is evident that causal
+inference drawn through mediation analysis will be difficult to
+interpret, and this is a serious problem considering that it is often
+impossible to estimate the extent of measurement error for most
+experimental designs.
+
+In the epidemiological literature this is relatively well understood and
+our initial analysis simply confirms that for widely used methods such
+as CIT that it is indeed liable to the same issues as standard mediation
+based analysis. Under many circumstances a practical solution to this
+problem is to use Mendelian randomisation instead of mediation based
+methods. Mendelian randomisation is robust in the face of measurement
+error and, given that the researcher has knowledge about the biology of
+the instrument being used in the analysis, can offer a direct solution
+to the issues that CIT faces. This assumption is often reasonable, for
+example exposures such as CRP
+
+Pleiotropy
 
@@
 
+22.Millstein, J. \emph{cit: Causal Inference Test}. (2016). at
+\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
+
 \end{document}

### 2016-02-14 scripts/mr_directionality_analysis.R

@@ -19,6 +19,8 @@ get_cor_from_pval <- function(p, n)
 
 load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
 
+parameters <- subset(parameters, r_ab != 1)
+
 parameters$p_az <- -log10(parameters$p_az)
 parameters$p_bz <- -log10(parameters$p_bz)
 parameters$r_ab <- parameters$r_ab^2
@@ -30,26 +32,31 @@ parameters$rhs <- parameters$cor_ab * parameters$cor_bbp
 parameters$lhs <- parameters$cor_aap
 parameters$rhs_lhs_diff <- parameters$lhs - parameters$rhs
 
-parameters$correct_direction <- parameters$p_az > parameters$p_bz
-
-parameters$p_test <- parameters$p_bz
-parameters$p_test[!parameters$correct_direction] <- parameters$p_az[!parameters$correct_direction]
-
 # Calculate correlation test
 a <- r.test(n=parameters$n, r12=parameters$cor_zap, r13=parameters$cor_zbp, r23=parameters$cor_abp)
 parameters$cortest_t <- a$t
 parameters$cortest_p <- -log10(a$p)
 parameters$cortest_correct_direction <- sign(parameters$cortest_t) == 1
+parameters$cortest_correct_direction[is.na(parameters$cortest_correct_direction)] <- FALSE
 
 
+# Get the p-value for the MR analysis
+# Use the cortest_t to determine which direction MR association to use
+parameters$p_test <- parameters$p_bz
+parameters$p_test[!parameters$cortest_correct_direction] <- parameters$p_az[!parameters$cortest_correct_direction]
+
+
+# Get the CIT direction
 parameters$cit_correct_direction <- parameters$cit_AB < parameters$cit_BA
 parameters$cit_correct_direction[is.na(parameters$cit_correct_direction)] <- FALSE
 
+# Get the p-value for the CIT direction
 parameters$cit_p <- parameters$cit_AB
 parameters$cit_p[!parameters$cit_correct_direction] <- parameters$cit_BA[!parameters$cit_correct_direction]
 parameters$cit_p <- -log10(parameters$cit_p)
[…]

### 2016-02-14 docs/manuscript.tex
updated

@@
 
+The analysis proceeds by simulating variables that have known causal
+effects and valid instruments, and then applying measurement error in
+different configurations. We then assess the performance of the CIT test
+and MR analysis for a) inference of a causal relationship and b)
+inferring the correct direction of causality. Included in the MR
+analysis is a simple extension that is used to make inference of the
+causal direction when it is unknown which of the correlated variables
+the instrument is directly acting upon. All analysis was performed using
+the R programming language\textsuperscript{22} and code is made
+available at \href{}{github url to go here}.
+
 \subsubsection{CIT test}\label{cit-test}
@@
 The CIT method\textsuperscript{1} is implemented in the R package
-\emph{R/cit}\textsuperscript{22}. The \emph{cit.cp} function was used to
+\emph{R/cit}\textsuperscript{23}. The \emph{cit.cp} function was used to
 obtain an omnibus p-value of causality between an exposure \(x\) and an
@@
 correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a
-population\textsuperscript{\textbf{???}}. It is calculated as
+for correlated correlations within a population\textsuperscript{24}. It
+is calculated as
 
@@
 \begin{aligned}
-cor(g, x) & = \{0.01, 0.05, 0.1\} \\
-cor(x, y) & = \{0.2, 0.4, 0.6, 0.8\} \\
+cor(g, x)^2 & = 0.1 \\
+cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
 var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
@@
 applied to each in order to evaluate the causal association of the
-simualted variables.
+simualted variables. Similar patterns of results were obtained for
+different values of \(cor(g, x)\).
 
@@
 our initial analysis simply confirms that for widely used methods such
-as CIT that it is indeed liable to the same issues as standard mediation
-based analysis. Under many circumstances a practical solution to this
-problem is to use Mendelian randomisation instead of mediation based
-methods. Mendelian randomisation is robust in the face of measurement
-error and, given that the researcher has knowledge about the biology of
-the instrument being used in the analysis, can offer a direct solution
-to the issues that CIT faces. This assumption is often reasonable, for
-example exposures such as CRP
+as CIT they are indeed liable to the same issues as standard mediation
+based analysis. Specifically, we show that as measurement error in the
+exposure variable increases, CIT is likely to infer the wrong direction
+of causality. We also demonstrate that, though seemingly unintuitive,
+increasing sample size does not resolve the issue, rather it increases
+the significance of the incorrect inference.
 
-Pleiotropy
+Under many circumstances a practical solution to this problem is to use
+Mendelian randomisation instead of mediation based methods. Mendelian
+randomisation is robust in the face of measurement error and, given that
+the researcher has knowledge about the biology of the instrument being
+used in the analysis, can offer a direct solution to the issues that CIT
+faces. This assumption is often reasonable, for example SNPs are
+commonly used as instruments when they are found in genes with known
+biological relevance to the trait of interest. But on many occasions
+this is not the case, and it may be tempting to resort to using methods
+based on mediation in order to be able to both ascertain if there is a
+causal association and to infer the direction of causality. Here we have
+described a simple extension to MR which can be used as an alternative
+to mediation based methods. We show that this method is still liable to
+measurement error, but because it has different properties to the CIT it
+offers two main advantages. First, it uses a formal statistical
+framework to test for the robostness of the assumed direction of
+causality. Second, after testing in a comprehensive range of scenarios
+the MR based approach is much less likely to infer the wrong direction
+of causality compared to CIT.
+
+There are a number of limitations to the analysis that we performed. We
+focused on the simple case of a single instrument in a single sample
+setting, and assumed that pleiotropy (the influence of the instrument on
+the outcome through a mechanism other than the exposure) was not
+present.
+
+The overarching result from our simulations is that, regardless of the
+method used, inferring causal direction using an instrument of unknown
+biology is very susceptible to measurement error. With the presence of
+measurement error near ubiquitous in most cross sectional studies, and
+our ability to measure it limited, we argue that it should be given
+serious consideration in designing statistical analyses that attempt to
+make causal inference, and that any putitive results should be
+accompanied with appropriate sensitivity analysis that assesses their
+robustness under varying levels of measurement error.
+
+\newpage
+
+\subsection{Figures}\label{figures}
+
+\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}\\
+Figure 1: The CIT was performed on simulated variables where the
+exposure influenced the outcome and the exposure was instrumented by a
+SNP. The test statistic from CIT when testing if the exposure caused the
+outcome (the true model) is in red, and the test for the outcome causing
+the exposure (false model) is in green. Rows of plots represent the
+sample sizes used for the simulations. As measurement error increases
+(decreasing values on x-axis) the test statistic for the incorrect model
+gets stronger and the test statistic for the correct model gets weaker.
+
+\newpage
+
+\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
+Figure 2: Plots depicting the function
+\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).
+
+\newpage
+
+Figure 3: Outcomes were simulated to be causally influenced by
+exposures, with varying degrees of measurement error applied to both.
+CIT and MR were used to infer evidence for causality between the
+exposure and outcome, and to infer the direction of causality. The value
+of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
+is negative we expect the Steiger test to be more likely to be wrong
+about the direction of causality. Rows of graphs represent the sample
+size used in the simulations.
+
+\newpage
 
@@
 \[
-x = \alpha_g + \beta_g g + \epsilon_g \\
-x_O = \alpha_m + \beta_m x + \epsilon_m \\
-y = \alpha_x + \beta_x x + \epsilon_x
+\begin{aligned}
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_O & = \alpha_m + \beta_m x + \epsilon_m \\
+y & = \alpha_x + \beta_x x + \epsilon_x
+\end{aligned}
 \]
@@
 cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                & = D\beta_x\beta_g var(g)
+                & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
 \end{aligned}
@@
 
-where
-
-\[
-D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
-\]
+\newpage
 
@@
 \newpage
-
-\subsection{Tables}\label{tables}
-
-\newpage
-
-\subsection{Figures}\label{figures}
-
-\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}\\
-Figure 1: The CIT was performed on simulated variables where the
-exposure influenced the outcome and the exposure was instrumented by a
-SNP. The test statistic from CIT when testing if the exposure caused the
-outcome (the true model) is in red, and the test for the outcome causing
-the exposure (false model) is in green. Rows of plots represent the
-sample sizes used for the simulations. As measurement error increases
-(decreasing values on x-axis) the test statistic for the incorrect model
-gets stronger and the test statistic for the correct model gets weaker.
-
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
-Figure 2: Plots depicting the function
-\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).
-
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}\\
-Figure 3: Outcomes were simulated to be causally influenced by
-exposures, with varying degrees of measurement error applied to both.
-CIT and MR were used to infer evidence for causality between the
-exposure and outcome, and to infer the direction of causality. The value
-of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
-is negative we expect the Steiger test to be more likely to be wrong
-about the direction of causality. Rows of graphs represent the sample
-size used in the simulations.
-
-\newpage
-
-\subsection{Supplementary tables}\label{supplementary-tables}
-
-\newpage
-
-\subsection{Supplementary figures}\label{supplementary-figures}
 
@@
 
-22.Millstein, J. \emph{cit: Causal Inference Test}. (2016). at
+22.R Core Team. \emph{R: A Language and Environment for Statistical
+Computing}. (R Foundation for Statistical Computing, 2015). at
+\textless{}\url{https://www.r-project.org/}\textgreater{}
+
+23.Millstein, J. \emph{cit: Causal Inference Test. R package version
+1.9}. (2016). at
 \textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
 
+24.Steiger, J. H. Tests for comparing elements of a correlation matrix.
+\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
+
 \end{document}

### 2016-02-16 docs/manuscript.Rmd
fixed missing pic

@@
 output:
-  pdf_document:
-    keep_tex: true
+  word_document
 bibliography: manuscript.bib
@@
 
-The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly liable to measurement error.
+The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is very susceptible to measurement error.
+With the presence of measurement error near ubiquitous in most cross sectional studies, and our ability to measure it limited, we argue that it should be given serious consideration in designing statistical analyses that attempt to make causal inference, and that any putitive results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
 
-Limitations
+\newpage
 
-Pleiotropy
+## Figures
+
+```{r cit_measurement_error_figure, fig.width=10, fig.height=10 }
+
+```
+
+`r fig_nums("cit_measurement_error_figure")`
+
+\newpage
+
+```{r d_relationship_figure, fig.width=10, fig.height=6 }
+
+```
+
+`r fig_nums("d_relationship_figure")`
+
+\newpage
+
+```{r cit_mr_comparison_figure, fig.width=10, fig.height=10 }
+
+```
+
+`r fig_nums("cit_mr_comparison_figure")`
+
+\newpage
 
@@
 $$
-x = \alpha_g + \beta_g g + \epsilon_g \\
-x_O = \alpha_m + \beta_m x + \epsilon_m \\
-y = \alpha_x + \beta_x x + \epsilon_x
+\begin{aligned}
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_O & = \alpha_m + \beta_m x + \epsilon_m \\
+y & = \alpha_x + \beta_x x + \epsilon_x
+\end{aligned}
 $$
@@
 
-where
-
-$$
-D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
-$$
+\newpage
 
@@
 
-## Tables
-
-\newpage
-
-## Figures
-
-```{r cit_measurement_error_figure, fig.width=10, fig.height=10 }
-
-```
-
-`r fig_nums("cit_measurement_error_figure")`
-
-\newpage
-
-```{r d_relationship_figure, fig.width=10, fig.height=6 }
-
-```
-
-`r fig_nums("d_relationship_figure")`
-
-\newpage
-
-```{r cit_mr_comparison_figure, fig.width=10, fig.height=10 }
-
-```
-
-`r fig_nums("cit_mr_comparison_figure")`
-
-\newpage
-
-## Supplementary tables
-
-\newpage
-
-## Supplementary figures
-
 ## References

### 2016-02-16 docs/manuscript.tex
fixed missing pic

@@
   \postdate{\par}
-  \date{14 February 2016}
+  \date{16 February 2016}
 
@@
 
+\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}\\
 Figure 3: Outcomes were simulated to be causally influenced by

### 2016-02-16 scripts/mr_directionality_analysis.R

@@ -139,7 +139,6 @@ psum2$rhs_lhs_diff_bin_lab <- factor(psum2$rhs_lhs_diff_bin_lab, levels=levels(p
 
 levels(psum2$eval) <- c("Evidence for causality\nwith correct direction", "Evidence for causality\nwith incorrect direction", "No evidence for causality")
 
-dev.new()
 ggplot(subset(psum2, round(r_za,2)==0.1), aes(x=test, y=value)) +
 geom_bar(stat="identity", aes(fill=eval)) +
 facet_grid(n ~ rhs_lhs_diff_bin_lab) +

### 2016-02-16 docs/manuscript.Rmd
word doc works well

@@
 output:
-  word_document
+  pdf_document:
+    keep_tex: true
 bibliography: manuscript.bib

### 2016-02-25 docs/manuscript.Rmd
updated analysis

@@
 The value of $d = \\rho_{x, x_O} - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality.
-Rows of graphs represent the sample size used in the simulations.",
+Rows of graphs represent the sample size used in the simulations.
+For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.",
 	display=FALSE
@@
 
-With the explosion in data that measures the human phenome, of great interest is understanding the causal nature of the emerging putitive correlations between measures of interest.
-Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling.
+With the explosion in data that assess the human phenome, we are presented with the challenge of understanding the causal nature of the emerging putitive correlations between measures of interest.
+Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008].
 The principle behind mediation analysis can be explained as follows.
@@
 Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
-This forms the basis of a number of methods, such as the causal inference test (CIT) [@Millstein2009], which have been employed by a number of recent publications make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
+This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
 
 While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results.
-One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013].
+One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013], which has been noted to be an explicit issue with CIT style approaches [@Wang2015; @Lagani2015].
 Measurement (or observational) error is the difference between a measured value of a quantity and its true value.
@@
 
-Measurement error can indeed arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and obesity is not perfect [@Ahima2013].
+Measurement error will of course arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
 A similar problem of biological misspecification is unavoidable in disease diagnosis.
@@
 
-An alternative statistical method that gained prominence at least in part as a solution to the problem of measurement error [@Ashenfelter1994] is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014].
-Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to proxy the exposure in an association test with the outcome.
+An alternative statistical method that gained prominence as a solution to the problem of measurement error is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014].
+Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to stand for the exposure in an association test with the outcome.
 If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
-This approach also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.
+Indeed this approach was in part pioneered (in the form of instrumental variable (IV) analysis) as a correction for measurement error [@Ashenfelter1994], but it also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.
 
-Here, however, arises a potential issue.
-Often the biological effect of a SNP is not known, and therefore it can be difficult to determine for which of the two variables in a putative exposure-outcome association is the SNP a valid instrument.
+Here, however, arises a potential issue: often the biological effect of a SNP is not known.
+If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?).
 By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
 Such a situation can arise in many scenarios.
-For example genome wide association studies (GWASs) that identify genetic associations for complex traits are by design hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
+For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
 Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?
@@
 The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
-The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$.
-To infer the direction of causality using the CIT method, the omnibus p-value was estimated for $x$ causing $y$, and for $y$ causing $x$, and the model that returned the lowest p-value was taken as the one denoting the correct causal direction.
+The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$ by making inferences based on a number of conditions that we examine in the results.
+To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$.
+For some significance threshold $\alpha$, causal direction was inferred based on the following scenarios:
+
+1. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
+2. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
+3. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then the an alternative causal model is accepted
+4. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made
+
+For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value.
 
@@
 The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test.
-For simplicity, the test is deemed to have obtained a robust association if the Z score test has a p-value $<  0.05$ and the two sample MR analysis has a p-value $< 0.05$.
-Note that the same approach can be applied to a two-sample MR setting, where
+For the purposes of compiling simulation results, the test is deemed to have obtained a robust association if the Z score test has a p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$, though in practice we do not advocate using an arbitrary $p$ value threshold for denoting significance in making causal inference.
+
+Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common.
+An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different.
 
@@
 `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
-However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT.
-When $d > 0$ is satisfied we observe that in many cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+However, in a large number of cases the MR analysis infers the wrong direction of causality at a much lower rate than the CIT.
+When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
 

### 2016-02-25 docs/manuscript.tex
updated analysis

--- 2016-02-14 docs/manuscript.tex
@@
   \postdate{\par}
-  \date{14 February 2016}
+  \date{25 February 2016}
 
@@
 
-With the explosion in data that measures the human phenome, of great
-interest is understanding the causal nature of the emerging putitive
-correlations between measures of interest. Mediation is a commonly used
-statistical method for making causal inference in observational
-data\textsuperscript{1--4}, and it exists in many forms, from simple
-regression based systems to structural equation modeling. The principle
-behind mediation analysis can be explained as follows. Supposing an
-exposure has a causal effect on an outcome, then an `instrument' that
-causes the exposure (a single nucleotide polymorophism (SNP), for
-example) should also influence the outcome. Therefore the influence of
-the instrument on the outcome conditional on the exposure should be
-zero. This forms the basis of a number of methods, such as the causal
-inference test (CIT)\textsuperscript{1}, which have been employed by a
-number of recent publications make causal inferences, often in large
-scale `omics datasets\textsuperscript{5--7}.
+With the explosion in data that assess the human phenome, we are
+presented with the challenge of understanding the causal nature of the
+emerging putitive correlations between measures of interest. Mediation
+is a commonly used statistical method for making causal inference in
+observational data\textsuperscript{1--4}, and it exists in many forms,
+from simple regression based systems to structural equation modeling
+(SEM)\textsuperscript{5}. The principle behind mediation analysis can be
+explained as follows. Supposing an exposure has a causal effect on an
+outcome, then an `instrument' that causes the exposure (a single
+nucleotide polymorophism (SNP), for example) should also influence the
+outcome. Therefore the influence of the instrument on the outcome
+conditional on the exposure should be zero. This forms the basis of a
+number of methods, such as the regression-based causal inference test
+(CIT)\textsuperscript{1}, an SEM implementation in the NEO
+software\textsuperscript{5}, and various Bayesian methods, which have
+been employed by a number of recent publications that make causal
+inferences, often in large scale `omics datasets\textsuperscript{6--8}.
 
@@
 it is important to note where they may lead to unreliable results. One
-such mechanism is measurement error\textsuperscript{8--11}. Measurement
-(or observational) error is the difference between a measured value of a
-quantity and its true value. Such variability can arise through a whole
-plethora of mechanisms, which are often unique to the study design and
-difficult to avoid\textsuperscript{12,13}. Array technology is now
-commonly used to obtain high throughput phenotyping at low cost, but
-they come with the problem of having imperfect sensitivity, so for
-example methylation levels as measured by the Illumina450k chip are
-prone to have some amount of noise around the true
-value\textsuperscript{14,15}. Sensitivity is also an issue, for example
-if the unit of measurement of biological interest is the methylation
-level in a T cell, then measurement error of this value can be
-introduced by using methylation levels from whole blood samples because
-the measured value will be an assay of many cell
-types\textsuperscript{16}.
+such mechanism is measurement error\textsuperscript{9--12}, which has
+been noted to be an explicit issue with CIT style
+approaches\textsuperscript{13,14}. Measurement (or observational) error
+is the difference between a measured value of a quantity and its true
+value. Such variability can arise through a whole plethora of
+mechanisms, which are often unique to the study design and difficult to
+avoid\textsuperscript{15,16}. Array technology is now commonly used to
+obtain high throughput phenotyping at low cost, but they come with the
+problem of having imperfect sensitivity, so for example methylation
+levels as measured by the Illumina450k chip are prone to have some
+amount of noise around the true value\textsuperscript{17,18}.
+Sensitivity is also an issue, for example if the unit of measurement of
+biological interest is the methylation level in a T cell, then
+measurement error of this value can be introduced by using methylation
+levels from whole blood samples because the measured value will be an
+assay of many cell types\textsuperscript{19}.
 
-Measurement error can indeed arise in more low-tech data too, for
+Measurement error will of course arise in more low-tech data too, for
 example when measuring body mass index (BMI) one is typically interested
 in using this as a proxy for obesity, but it is clear that the
-correlation between BMI and obesity is not perfect\textsuperscript{17}.
-A similar problem of biological misspecification is unavoidable in
-disease diagnosis. Measurement error can also be introduced after the
-data has been collected, for example the transformation of non-normal
-data for the purpose of statistical analysis will lead to a new variable
-that will typically have both bias and imprecision compared to the
-original variable. The sources of measurement error are not limited to
-this list\textsuperscript{13}, and its impact has been explored in the
-context of mediation analysis in the epidemiological literature
-extensively\textsuperscript{8,11}.
+correlation between BMI and underlying adiposity is not
+perfect\textsuperscript{20}. A similar problem of biological
+misspecification is unavoidable in disease diagnosis. Measurement error
+can also be introduced after the data has been collected, for example
+the transformation of non-normal data for the purpose of statistical
+analysis will lead to a new variable that will typically have both bias
+and imprecision compared to the original variable. The sources of
+measurement error are not limited to this list\textsuperscript{16}, and
+its impact has been explored in the context of mediation analysis in the
+epidemiological literature extensively\textsuperscript{9,12}.
 
-An alternative statistical method that gained prominence at least in
-part as a solution to the problem of measurement
-error\textsuperscript{18} is Mendelian randomisation
-(MR)\textsuperscript{19,20}. Here an instrumental variable (typically a
+An alternative statistical method that gained prominence as a solution
+to the problem of measurement error is Mendelian randomisation
+(MR)\textsuperscript{21,22}. Here an instrumental variable (typically a
 SNP) that has a robost, causal association with the exposure is used to
-proxy the exposure in an association test with the outcome. If there is
-no measurement error in the SNP then the causal association between the
-exposure and the outcome can be estimated by scaling the association
+stand for the exposure in an association test with the outcome. If there
+is no measurement error in the SNP then the causal association between
+the exposure and the outcome can be estimated by scaling the association
 between the SNP and the outcome by the association between the SNP and
-the exposure. This approach also has the useful property that it guards
-against unmeasured confounders driving the association between exposure
-and outcome; and if the biological effect of the SNP is understood then
-it can also aid against reverse causality driving the association.
+the exposure. Indeed this approach was in part pioneered (in the form of
+instrumental variable (IV) analysis) as a correction for measurement
+error\textsuperscript{23}, but it also has the useful property that it
+guards against unmeasured confounders driving the association between
+exposure and outcome; and if the biological effect of the SNP is
+understood then it can also aid against reverse causality driving the
+association.
 
-Here, however, arises a potential issue. Often the biological effect of
-a SNP is not known, and therefore it can be difficult to determine for
-which of the two variables in a putative exposure-outcome association is
-the SNP a valid instrument. By definition, we expect that if the
+Here, however, arises a potential issue: often the biological effect of
+a SNP is not known. If there exists a putative association between two
+variables, with a SNP being robustly associated with each, it can be
+difficult to determine which of the two variables is subject to the
+direct effect of the SNP (i.e., for which of the two variables is the
+SNP a valid instrument?). By definition, we expect that if the
 association is causal then a SNP for the exposure will be associated
@@
 arise in many scenarios. For example genome wide association studies
-(GWASs) that identify genetic associations for complex traits are by
-design hypothesis free and agnostic of genomic function, and it often
+(GWASs) that identify genetic associations for complex traits are, by
+design, hypothesis free and agnostic of genomic function, and it often
 takes years of follow up studies to understand the biological nature of
-a putative GWAS hit\textsuperscript{21}. Another situation is where the
+a putative GWAS hit\textsuperscript{24}. Another situation is where the
 causal direction between 'omic measures need to be determined, for
@@
 adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level\textsuperscript{7}?
+methylation level, or the gene expression level\textsuperscript{8}?
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{22} and code is made
+the R programming language\textsuperscript{25} and code is made
 available at \href{}{github url to go here}.
@@
 The CIT method\textsuperscript{1} is implemented in the R package
-\emph{R/cit}\textsuperscript{23}. The \emph{cit.cp} function was used to
+\emph{R/cit}\textsuperscript{26}. The \emph{cit.cp} function was used to
 obtain an omnibus p-value of causality between an exposure \(x\) and an
-outcome \(y\) using a mediator \(g\). To infer the direction of
-causality using the CIT method, the omnibus p-value was estimated for
-\(x\) causing \(y\), and for \(y\) causing \(x\), and the model that
-returned the lowest p-value was taken as the one denoting the correct
-causal direction.
+outcome \(y\) using a mediator \(g\) by making inferences based on a
+number of conditions that we examine in the results. To infer the
+direction of causality using the CIT method, an omnibus p-value
+generated by CIT, \(p_{CIT}\), was estimated for the correct model of
+\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
+some significance threshold \(\alpha\), causal direction was inferred
+based on the following scenarios:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
+  then the correct model is accepted
+\item
+  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
+  then the incorrect model is accepted
+\item
+  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
+  then the an alternative causal model is accepted
+\item
+  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
+  then no inference is made
+\end{enumerate}
+
+For the purposes of compiling simulation results we use an arbitrary
+\(\alpha = 0.05\) value.
 
@@
 correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a population\textsuperscript{24}. It
+for correlated correlations within a population\textsuperscript{27}. It
 is calculated as
@@
 the direction that is predicted to be the correct model based on the Z
-score test. For simplicity, the test is deemed to have obtained a robust
-association if the Z score test has a p-value \(< 0.05\) and the two
-sample MR analysis has a p-value \(< 0.05\). Note that the same approach
-can be applied to a two-sample MR setting, where
+score test. For the purposes of compiling simulation results, the test
+is deemed to have obtained a robust association if the Z score test has
+a p-value \(< 0.05\) and the two sample MR analysis has a p-value
+\(< 0.05\), though in practice we do not advocate using an arbitrary
+\(p\) value threshold for denoting significance in making causal
+inference.
+
+Note that the same correlation test approach can be applied to a
+two-sample MR\textsuperscript{28} setting, where the Steiger test of two
+independent correlations is applied instead of the case detailed here
+where correlations share one variable in common. An advantage of using
+the Steiger test in the two sample context is that it can compare
+correlations in independent samples where sample sizes are different.
 
@@
 that this erroneous result is more likely to occur with increasing
-sample size. However, in all cases the MR analysis infers the wrong
-direction of causality at the same or a lower rate than the CIT. When
-\(d > 0\) is satisfied we observe that in many cases the MR method has
+sample size. However, in a large number of cases the MR analysis infers
+the wrong direction of causality at a much lower rate than the CIT. When
+\(d > 0\) is satisfied we observe that in most cases the MR method has
 greater power to obtain evidence for causality than CIT, and always
@@
 
-\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}\\
+\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}
+
 Figure 1: The CIT was performed on simulated variables where the
@@
 
+\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}
+
 Figure 3: Outcomes were simulated to be causally influenced by
@@
 about the direction of causality. Rows of graphs represent the sample
-size used in the simulations.
+size used in the simulations. For the CIT method, outcome 1 denoted
+evidence for causality with correct model, outcomes 2 or 3 denoted
+evidence for causality with incorrect model, and outcome 4 denoted no
+evidence for causality.
 
@@
 
-5.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
+5.Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
+markers to orient the edges in quantitative trait networks: the NEO
+software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
+
+6.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
 epigenetic marks that mediate genetic risk for epithelial ovarian
@@
 
-6.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+7.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
 methylation as an intermediary of genetic risk in rheumatoid arthritis.
@@
 
-7.Waszak, S. M. \emph{et al.} Variation and genetic control of chromatin
+8.Waszak, S. M. \emph{et al.} Variation and genetic control of chromatin
 architecture in humans. \emph{Cell} \textbf{162,} 1039--1050 (2015).
 
-8.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+9.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
 Vandenbroucke, J. P. Quantification of bias in direct effects estimates
@@
 
-9.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+10.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
 
-10.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+11.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 
-11.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+12.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
 mediator matters when estimating indirect effects. \emph{Journal of
@@
 
-12.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
+13.Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
+
+14.Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+I. in \emph{Uncertainty in biology: A computational modeling approach}
+47 (Springer, 2015). at
+\textless{}\url{https://books.google.com/books?id=8SLUCgAAQBAJ\{/\&\}pgis=1}\textgreater{}
+
+15.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
 meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
@@
 
-13.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
+16.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
 measurement bias. \emph{American journal of epidemiology} \textbf{170,}
@@
 
-14.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
+17.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
 pathway analysis: two potential perils in cancer studies involving DNA
@@
 
-15.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
+18.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
 polymorphic CpGs in the Illumina Infinium HumanMethylation450
@@
 
-16.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
+19.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
 measures of cell mixture distribution. \emph{BMC bioinformatics}
@@
 
-17.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
+20.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
 obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
@@
 
-18.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
+21.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
+potentials, and limitations. \emph{International journal of
+epidemiology} \textbf{33,} 30--42 (2004).
+
+22.{Davey Smith}, G. \& Hemani, G. Mendelian randomization: genetic
+anchors for causal inference in epidemiological studies. \emph{Human
+molecular genetics} \textbf{23,} R89-----R98 (2014).
+
+23.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
 Schooling from a New Sample of Twins. \emph{The American Economic
@@
 
-19.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
-potentials, and limitations. \emph{International journal of
-epidemiology} \textbf{33,} 30--42 (2004).
-
-20.{Davey Smith}, G. \& Hemani, G. Mendelian randomization: genetic
-anchors for causal inference in epidemiological studies. \emph{Human
-molecular genetics} \textbf{23,} R89-----R98 (2014).
-
-21.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+24.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
 Adipocyte Browning in Humans. \emph{The New England journal of medicine}
@@
 
-22.R Core Team. \emph{R: A Language and Environment for Statistical
+25.R Core Team. \emph{R: A Language and Environment for Statistical
 Computing}. (R Foundation for Statistical Computing, 2015). at
@@
 
-23.Millstein, J. \emph{cit: Causal Inference Test. R package version
+26.Millstein, J. \emph{cit: Causal Inference Test. R package version
 1.9}. (2016). at
@@
 
-24.Steiger, J. H. Tests for comparing elements of a correlation matrix.
+27.Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
 
+28.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+randomization studies: subsample and 2-sample instrumental variable
+estimators. \emph{American journal of epidemiology} \textbf{178,}
+1177--84 (2013).
+
 \end{document}

### 2016-02-25 scripts/mr_directionality_analysis.R

@@ -47,12 +47,35 @@ parameters$p_test[!parameters$cortest_correct_direction] <- parameters$p_az[!par
 
 
 # Get the CIT direction
-parameters$cit_correct_direction <- parameters$cit_AB < parameters$cit_BA
-parameters$cit_correct_direction[is.na(parameters$cit_correct_direction)] <- FALSE
+
+get_cit_direction <- function(p1, p2, thresh)
+{
+	res <- array(0, length(p1))
+	res[p1 <= thresh & p2 > thresh] <- 1
+	res[p1 > thresh & p2 <= thresh] <- 2
+	res[p1 <= thresh & p2 <= thresh] <- 3
+	res[p1 > thresh & p2 > thresh] <- 4
+	return(res)
+}
+
+parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.05)
+parameters$cit_correct_direction <- parameters$cit_res == 1
 
 # Get the p-value for the CIT direction
+# ab < 0.05 & ba > 0.05 = correct
+# ab > 0.05 & ba < 0.05 = incorrect
+# ab < 0.05 & ba < 0.05 = no call
+# ab > 0.05 & ba > 0.05 = incorrect
+
+
 parameters$cit_p <- parameters$cit_AB
-parameters$cit_p[!parameters$cit_correct_direction] <- parameters$cit_BA[!parameters$cit_correct_direction]
+parameters$cit_p[parameters$cit_res == 2] <- parameters$cit_BA[parameters$cit_res == 2]
+parameters$cit_p[parameters$cit_res == 3] <- 0.5
+parameters$cit_p[parameters$cit_res == 4] <- 0.01
+
+# Do we classify 4 as being significant wrong causal direction, or no causal inference?
+
+
 parameters$cit_p <- -log10(parameters$cit_p)
 parameters$cit_p[is.na(parameters$cit_AB) | is.na(parameters$cit_BA)] <- NA
 
@@ -99,7 +122,7 @@ psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=10)
 
 load("~/repo/cit_measurement_error/results/20141114.RData")
 dat <- gather(dat, eval, pval, GT, TG, factor_key=TRUE)
-levels(dat$eval) <- c("Correct inference", "Incorrect inference")
[…]

### 2016-02-25 docs/manuscript.Rmd
updated

@@
 ---
-title: "Determining the direction of causality in the face of measurement error"
+title: "Determining the existance and direction of causality in the face of measurement error"
 date: "`r format(Sys.time(), '%d %B %Y')`"
 output:
+  <!-- word_document -->
   pdf_document:
@@
 For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
+To illustrate a trivial example - the same FTO variant that is famously associated with BMI is also associated with C-reactive protein (CRP).
+It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would infer that CRP causes BMI.
 Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?
@@
 First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality.
-Second, after testing in a comprehensive range of scenarios the MR based approach is much less likely to infer the wrong direction of causality compared to CIT.
+Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.
+
+Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
+But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
+Recent advances in MR methodology, for example network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust solutions alternatives for these more complicated problems.
 
@@
 We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
+To give an example in which this might arise consider the case that C-reactive protein (CRP) is being tested for causal association with some outcome, and the FTO variant is used as an instrument.
+Though FTO is robustly associated with CRP, it is through the upstream influence of BMI [@Timpson2011].
+Thus, if BMI has a direct influence on the outcome, the conditions of MR will be violated and the Steiger test proposed here will be unreliable.
 

### 2016-02-25 docs/manuscript.tex
updated

@@
             pdfauthor={},
-            pdftitle={Determining the direction of causality in the face of measurement error},
+            pdftitle={Determining the existance and direction of causality in the face of measurement error},
             colorlinks=true,
@@
 \setlength{\droptitle}{-2em}
-  \title{Determining the direction of causality in the face of measurement error}
+  \title{Determining the existance and direction of causality in the face of
+measurement error}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
 takes years of follow up studies to understand the biological nature of
-a putative GWAS hit\textsuperscript{24}. Another situation is where the
-causal direction between 'omic measures need to be determined, for
-example if a DNA methylation probe is associated with expression of an
-adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level\textsuperscript{8}?
+a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
+- the same FTO variant that is famously associated with BMI is also
+associated with C-reactive protein (CRP). It is known that this is
+because BMI causes CRP\textsuperscript{25}, but supposing we only had
+\emph{a priori} knowledge that FTO associated with CRP, and we used it
+to instrument CRP in an MR analysis of CRP against BMI, we would infer
+that CRP causes BMI. Another situation is where the causal direction
+between 'omic measures need to be determined, for example if a DNA
+methylation probe is associated with expression of an adjacent gene,
+then is a cis-acting SNP an instrument for the DNA methylation level, or
+the gene expression level\textsuperscript{8}?
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{25} and code is made
+the R programming language\textsuperscript{26} and code is made
 available at \href{}{github url to go here}.
@@
 The CIT method\textsuperscript{1} is implemented in the R package
-\emph{R/cit}\textsuperscript{26}. The \emph{cit.cp} function was used to
+\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
 obtain an omnibus p-value of causality between an exposure \(x\) and an
@@
 correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a population\textsuperscript{27}. It
+for correlated correlations within a population\textsuperscript{28}. It
 is calculated as
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{28} setting, where the Steiger test of two
+two-sample MR\textsuperscript{29} setting, where the Steiger test of two
 independent correlations is applied instead of the case detailed here
@@
 causality. Second, after testing in a comprehensive range of scenarios
-the MR based approach is much less likely to infer the wrong direction
-of causality compared to CIT.
+the MR based approach is less likely to infer the wrong direction of
+causality compared to CIT, while substantially improving power over CIT
+in the cases where \(d > 0\).
+
+Mediation based network approaches are very well
+established\textsuperscript{14} and have a number of extensions that
+make them valuable tools, including for example network construction.
+But because they are predicated on the basic underlying principles of
+mediation they are liable to suffer from the same issues of measurement
+error. Recent advances in MR methodology, for example network
+MR\textsuperscript{30} and mediation through MR\textsuperscript{31} may
+offer more robust solutions alternatives for these more complicated
+problems.
 
@@
 the outcome through a mechanism other than the exposure) was not
-present.
+present. To give an example in which this might arise consider the case
+that C-reactive protein (CRP) is being tested for causal association
+with some outcome, and the FTO variant is used as an instrument. Though
+FTO is robustly associated with CRP, it is through the upstream
+influence of BMI\textsuperscript{25}. Thus, if BMI has a direct
+influence on the outcome, the conditions of MR will be violated and the
+Steiger test proposed here will be unreliable.
 
@@
 
-25.R Core Team. \emph{R: A Language and Environment for Statistical
+25.Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+index: elucidating direction of causation through reciprocal Mendelian
+randomization. \emph{International journal of obesity (2005)}
+\textbf{35,} 300--8 (2011).
+
+26.R Core Team. \emph{R: A Language and Environment for Statistical
 Computing}. (R Foundation for Statistical Computing, 2015). at
@@
 
-26.Millstein, J. \emph{cit: Causal Inference Test. R package version
+27.Millstein, J. \emph{cit: Causal Inference Test. R package version
 1.9}. (2016). at
@@
 
-27.Steiger, J. H. Tests for comparing elements of a correlation matrix.
+28.Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
 
-28.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+29.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
 randomization studies: subsample and 2-sample instrumental variable
@@
 
+30.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+Network Mendelian randomization: using genetic variants as instrumental
+variables to investigate mediation in causal pathways.
+\emph{International journal of epidemiology} \textbf{44,} 484--95
+(2015).
+
+31.Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+cholesterol, and blood pressure as mediators from obesity to ischemic
+heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).
+
 \end{document}

### 2016-05-04 scripts/theory.R
updated theory to allow measurement error in outcome as well as exposure

@@ -5,10 +5,17 @@ c_gyhat <- array(0, nsim)
 c_gy_exp <- array(0, nsim)
 c_gyhat_exp <- array(0, nsim)
 
+c_gy0 <- array(0, nsim)
+c_gy0hat <- array(0, nsim)
+c_gy0_exp <- array(0, nsim)
+c_gy0hat_exp <- array(0, nsim)
+
+
 n <- 100000
 
 for(i in 1:nsim)
 {
+	message(i)
 	g <- rbinom(n, 2, 0.5)
 
 	a_g <- runif(1)
@@ -23,9 +30,14 @@ for(i in 1:nsim)
 	b_x <- runif(1)
 	e_x <- runif(1)
 
+	a_n <- runif(1)
+	b_n <- runif(1)
+	e_n <- runif(1)
+
 	x <- a_g + b_g * g + rnorm(n, sd=sqrt(e_g))
 	x0 <- a_m + b_m * x + rnorm(n, sd=sqrt(e_m))
 	y <- a_x + b_x * x + rnorm(n, sd=sqrt(e_x))
+	y0 <- a_n + b_n * y + rnorm(n, sd=sqrt(e_n))
 	yhat <- fitted.values(lm(y ~ x0))
 
 	c_gy[i] <- cov(g, y)
@@ -33,6 +45,16 @@ for(i in 1:nsim)
 
 	c_gyhat[i] <- cov(g, yhat)
 	c_gyhat_exp[i] <- b_x * b_g * var(g) * b_m^2 * var(x) / (b_m^2 * var(x) + e_m)
+
+
+	y0hat <- fitted.values(lm(y0 ~ x0))
+
+	c_gy0[i] <- cov(g, y0)
+	c_gy0_exp[i] <- b_n * b_x * b_g * var(g)
+
+	c_gy0hat[i] <- cov(g, y0hat)
+	c_gy0hat_exp[i] <- b_n * b_x * b_g * var(g) * b_m^2 * var(x) / (b_m^2 * var(x) + e_m)
+
 }
 
 
@@ -41,8 +63,18 @@ dat <- rbind(
 	data.frame(obs=c_gyhat, exp=c_gyhat_exp, what="cov_gyhat")
 )
 
+dat0 <- rbind(
+	data.frame(obs=c_gy0, exp=c_gy0_exp, what="cov_gy"),
+	data.frame(obs=c_gy0hat, exp=c_gy0hat_exp, what="cov_gyhat")
+)
+
[…]

### 2016-05-04 scripts/proxy.R
doing proxy again - with x and y measurement error

@@ -164,7 +164,7 @@ testProxy <- function(x, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1))
 	return(dat)	
 }
 
-runSim <- function(N, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1))
+runSim <- function(N, NoiseT=seq(0.25, 1, by=0.25), BiasT=seq(0.25, 1, by=0.25), NoiseG=seq(0.25, 1, by=0.25), BiasG=seq(0.25, 1, by=0.25))
 {
 	require(plyr)
 	d <- list()
@@ -175,44 +175,17 @@ runSim <- function(N, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1))
 		L2 <- rbinom(n, 2, 0.5)
 		G <- L1 + rnorm(n)
 		T <- G + rnorm(n) + L2
-		dat <- expand.grid(noise=Noise, bias=Bias, GT = NA, TG = NA, rsq = NA, bdmr = NA, bdmr_p = NA)
+		dat <- expand.grid(noiseT=NoiseT, biasT=BiasT, noiseG=NoiseG, biasG=BiasG, GT = NA, TG = NA, rsqG = NA, rsqT = NA, bdmr = NA)
 		for(i in 1:nrow(dat))
 		{
-			cat(i, "\n")
-			G1 <- makeProxy(G, dat$noise[i], dat$bias[i])
-			dat$GT[i] <- cit(L1, G1, T, maxit=1000)$p_cit
-			dat$TG[i] <- cit(L1, T, G1, maxit=1000)$p_cit
-			dat$bdmr[i] <- inferCausality(twoStageLS(G, T, L1, L2, FALSE))
-			dat$bdmr_p[i] <- inferCausality(twoStageLS(G1, T, L1, L2, FALSE))
-			dat$rsq[i] <- cor(G1, G)^2
-		}
-		dat$n <- n
-		d[[j]] <- dat
-	}
-	dat <- rbind.fill(d)
-	return(dat)
-}
-
-runSimMR <- function(N, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1), Nsim=1:100)
-{
-	require(plyr)
-	d <- list()
-	for(j in 1:length(N))
-	{
-		n <- N[j]
[…]

### 2016-05-04 scripts/proxy.R
new results generated

@@ -196,4 +196,4 @@ runSim <- function(N, NoiseT=seq(0.25, 1, by=0.25), BiasT=seq(0.25, 1, by=0.25),
 
 
 dat <- runSim(c(100, 500, 1000, 5000, 10000))
-save(dat, file="~/repo/MethylationIV/cit/results/20160504.RData")
+save(dat, file="~/repo/cit_measurement_error/results/20160504.RData")

### 2016-05-05 scripts/aggregate_proxy.R
added proxy res

@@ -1,0 +1,10 @@
+n <- c(100, 500, 1000, 5000, 10000)
+l <- list()
+for(i in 1:length(n))
+{
+	load(paste0("~/repo/cit_measurement_error/results/20160504_", n[i], ".RData"))
+	l[[i]] <- dat
+}
+
+dat <- do.call(rbind, l)
+save(dat, file="~/repo/cit_measurement_error/results/20160504.RData")

### 2016-05-05 scripts/proxy.R

@@ -164,7 +164,7 @@ testProxy <- function(x, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1))
 	return(dat)	
 }
 
-runSim <- function(N, NoiseT=seq(0.25, 1, by=0.25), BiasT=seq(0.25, 1, by=0.25), NoiseG=seq(0.25, 1, by=0.25), BiasG=seq(0.25, 1, by=0.25))
+runSim <- function(N, NoiseT=seq(0.1, 1, by=0.1), BiasT=seq(0.1, 1, by=0.1), NoiseG=seq(0.1, 1, by=0.1), BiasG=seq(0.1, 1, by=0.1))
 {
 	require(plyr)
 	d <- list()
@@ -175,7 +175,8 @@ runSim <- function(N, NoiseT=seq(0.25, 1, by=0.25), BiasT=seq(0.25, 1, by=0.25),
 		L2 <- rbinom(n, 2, 0.5)
 		G <- L1 + rnorm(n)
 		T <- G + rnorm(n) + L2
-		dat <- expand.grid(noiseT=NoiseT, biasT=BiasT, noiseG=NoiseG, biasG=BiasG, GT = NA, TG = NA, rsqG = NA, rsqT = NA, bdmr = NA)
+		dat <- expand.grid(noiseT=NoiseT, biasT=BiasT, noiseG=NoiseG, biasG=BiasG, GT = NA, TG = NA, rsqG = NA, rsqT = NA, bdmr = NA, bdmr_p = NA)
+		message(nrow(dat))
 		for(i in 1:nrow(dat))
 		{
 			message(i)
@@ -183,7 +184,8 @@ runSim <- function(N, NoiseT=seq(0.25, 1, by=0.25), BiasT=seq(0.25, 1, by=0.25),
 			T1 <- makeProxy(T, dat$noiseT[i], dat$biasT[i])
 			dat$GT[i] <- cit.cp(L1, G1, T1)[1]
 			dat$TG[i] <- cit.cp(L1, T1, G1)[1]
-			dat$bdmr[i] <- inferCausality(twoStageLS(G1, T1, L1, L2, FALSE))
+			dat$bdmr[i] <- inferCausality(twoStageLS(G, T, L1, L2, FALSE))
+			dat$bdmr_p[i] <- inferCausality(twoStageLS(G1, T1, L1, L2, FALSE))
[…]

### 2016-05-05 scripts/run_proxy.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+
+module add libraries/gnu_builds/gsl-1.16
+
+Rscript proxy.R 100 &
+Rscript proxy.R 500 &
+Rscript proxy.R 1000 &
+Rscript proxy.R 5000 &
+Rscript proxy.R 10000 &
+wait
+
+Rscript aggregate_proxy.R
+

### 2016-05-05 docs/manuscript.Rmd
updated

@@
 output:
-  <!-- word_document -->
   pdf_document:
@@
 x_O & = \alpha_m + \beta_m x + \epsilon_m \\
-y & = \alpha_x + \beta_x x + \epsilon_x
+y   & = \alpha_x + \beta_x x + \epsilon_x \\
+y_O & = \alpha_n + \beta_n y + \epsilon_n
 \end{aligned}
@@
 
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, and $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y - \hat{y}) = 0$ when $x$ causes $y$, where $\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}$ is the predicted value of $y$ using the measured value of $x_O$.
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
+Under the CIT, this is expected to be $cov(g, y_O - \hat{y_O}) = 0$ when $x$ causes $y$, where $\hat{y}_O = a_n + \beta_n(\hat{a}_{x_O} + \hat{\beta}_{x_O} x_O)$ is the predicted value of $y_O$ using the measured value of $x_O$.
 
-We can split $cov(g, y - \hat{y})$ into two parts, $cov(g, y)$ and $cov(g, \hat{y})$.
+We can split $cov(g, y_O - \hat{y}_O)$ into two parts, $cov(g, y_O)$ and $cov(g, \hat{y}_O)$.
 
@@
 \begin{aligned}
-cov(g, y) & = cov(g, \beta_x x) \\
-          & = cov(g, \beta_x \beta_g g) \\
-          & = \beta_x\beta_g var(g)
+cov(g, y_O) & = cov(g, \beta_n y) \\
+            & = cov(g, \beta_n \beta_x x) \\
+            & = cov(g, \beta_n \beta_x \beta_g g) \\
+            & = \beta_n \beta_x \beta_g var(g)
 \end{aligned}
@@
 \begin{aligned}
-cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
-                & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
-                & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
-                & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
+cov(g, \hat{y}_O) & = cov(g, \alpha_n + \beta_n \alpha_x + \beta_n \hat{\beta}_{x_O} x_O) \\
+                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m x) \\
+                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m \beta_g g) \\
+                  & = \hat{\beta}_{x_O} \beta_n \beta_m \beta_g var(g)
 \end{aligned}
@@
 \begin{aligned}
-\hat{\beta}_{x_O} & = \frac{cov(y, x_O)} {var(x_O)} \\
-                  & = \frac{\beta_m cov(y, x)} {var(x_O)} \\
-                  & = \frac{\beta_m \beta_x var(x)} {var(x_O)} \\
-                  & = \frac{\beta_m \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
+\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
+                  & = \frac{cov(\beta_n y, \beta_m x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m \beta_n cov(y, x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
 \end{aligned}
@@
 \begin{aligned}
-cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                   & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)

### 2016-05-05 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{25 February 2016}
+  \date{05 May 2016}
 
@@
 * Correspondence to:
-\href{mailto:g.hemani@bristol.ac.uk}{\nolinkurl{g.hemani@bristol.ac.uk}}
+\href{mailto:g.hemani@bristol.ac.uk}{g.hemani@bristol.ac.uk}
 
@@
 is a commonly used statistical method for making causal inference in
-observational data\textsuperscript{1--4}, and it exists in many forms,
+observational data \textsuperscript{1--4}, and it exists in many forms,
 from simple regression based systems to structural equation modeling
-(SEM)\textsuperscript{5}. The principle behind mediation analysis can be
-explained as follows. Supposing an exposure has a causal effect on an
+(SEM) \textsuperscript{5}. The principle behind mediation analysis can
+be explained as follows. Supposing an exposure has a causal effect on an
 outcome, then an `instrument' that causes the exposure (a single
@@
 number of methods, such as the regression-based causal inference test
-(CIT)\textsuperscript{1}, an SEM implementation in the NEO
-software\textsuperscript{5}, and various Bayesian methods, which have
-been employed by a number of recent publications that make causal
-inferences, often in large scale `omics datasets\textsuperscript{6--8}.
+(CIT) \textsuperscript{1}, an SEM implementation in the NEO software
+\textsuperscript{5}, and various Bayesian methods, which have been
+employed by a number of recent publications that make causal inferences,
+often in large scale `omics datasets \textsuperscript{6--8}.
 
@@
 it is important to note where they may lead to unreliable results. One
-such mechanism is measurement error\textsuperscript{9--12}, which has
-been noted to be an explicit issue with CIT style
-approaches\textsuperscript{13,14}. Measurement (or observational) error
-is the difference between a measured value of a quantity and its true
-value. Such variability can arise through a whole plethora of
-mechanisms, which are often unique to the study design and difficult to
-avoid\textsuperscript{15,16}. Array technology is now commonly used to
-obtain high throughput phenotyping at low cost, but they come with the
-problem of having imperfect sensitivity, so for example methylation
-levels as measured by the Illumina450k chip are prone to have some
-amount of noise around the true value\textsuperscript{17,18}.
-Sensitivity is also an issue, for example if the unit of measurement of
-biological interest is the methylation level in a T cell, then
-measurement error of this value can be introduced by using methylation
-levels from whole blood samples because the measured value will be an
-assay of many cell types\textsuperscript{19}.
+such mechanism is measurement error \textsuperscript{9--12}, which has
+been noted to be an explicit issue with CIT style approaches
+\textsuperscript{13,14}. Measurement (or observational) error is the
+difference between a measured value of a quantity and its true value.
+Such variability can arise through a whole plethora of mechanisms, which
+are often unique to the study design and difficult to avoid
+\textsuperscript{15,16}. Array technology is now commonly used to obtain
+high throughput phenotyping at low cost, but they come with the problem
+of having imperfect sensitivity, so for example methylation levels as
+measured by the Illumina450k chip are prone to have some amount of noise
+around the true value \textsuperscript{17,18}. Sensitivity is also an
+issue, for example if the unit of measurement of biological interest is
+the methylation level in a T cell, then measurement error of this value
+can be introduced by using methylation levels from whole blood samples
+because the measured value will be an assay of many cell types
+\textsuperscript{19}.
 
@@
 in using this as a proxy for obesity, but it is clear that the
-correlation between BMI and underlying adiposity is not
-perfect\textsuperscript{20}. A similar problem of biological
-misspecification is unavoidable in disease diagnosis. Measurement error
-can also be introduced after the data has been collected, for example
-the transformation of non-normal data for the purpose of statistical
+correlation between BMI and underlying adiposity is not perfect
+\textsuperscript{20}. A similar problem of biological misspecification
+is unavoidable in disease diagnosis. Measurement error can also be
+introduced after the data has been collected, for example the
+transformation of non-normal data for the purpose of statistical
 analysis will lead to a new variable that will typically have both bias
 and imprecision compared to the original variable. The sources of
-measurement error are not limited to this list\textsuperscript{16}, and
+measurement error are not limited to this list \textsuperscript{16}, and
 its impact has been explored in the context of mediation analysis in the
-epidemiological literature extensively\textsuperscript{9,12}.
+epidemiological literature extensively \textsuperscript{9,12}.
 
 An alternative statistical method that gained prominence as a solution
-to the problem of measurement error is Mendelian randomisation
-(MR)\textsuperscript{21,22}. Here an instrumental variable (typically a
-SNP) that has a robost, causal association with the exposure is used to
-stand for the exposure in an association test with the outcome. If there
-is no measurement error in the SNP then the causal association between
-the exposure and the outcome can be estimated by scaling the association
+to the problem of measurement error is Mendelian randomisation (MR)
+\textsuperscript{21,22}. Here an instrumental variable (typically a SNP)
+that has a robost, causal association with the exposure is used to stand
+for the exposure in an association test with the outcome. If there is no
+measurement error in the SNP then the causal association between the
+exposure and the outcome can be estimated by scaling the association
 between the SNP and the outcome by the association between the SNP and
@@
 instrumental variable (IV) analysis) as a correction for measurement
-error\textsuperscript{23}, but it also has the useful property that it
+error \textsuperscript{23}, but it also has the useful property that it
 guards against unmeasured confounders driving the association between
@@
 takes years of follow up studies to understand the biological nature of
-a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
-- the same FTO variant that is famously associated with BMI is also
-associated with C-reactive protein (CRP). It is known that this is
-because BMI causes CRP\textsuperscript{25}, but supposing we only had
+a putative GWAS hit \textsuperscript{24}. To illustrate a trivial
+example - the same FTO variant that is famously associated with BMI is
+also associated with C-reactive protein (CRP). It is known that this is
+because BMI causes CRP \textsuperscript{25}, but supposing we only had
 \emph{a priori} knowledge that FTO associated with CRP, and we used it
@@
 then is a cis-acting SNP an instrument for the DNA methylation level, or
-the gene expression level\textsuperscript{8}?
+the gene expression level \textsuperscript{8}?
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{26} and code is made
+the R programming language \textsuperscript{26} and code is made
 available at \href{}{github url to go here}.
@@
 
-The CIT method\textsuperscript{1} is implemented in the R package
-\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
-obtain an omnibus p-value of causality between an exposure \(x\) and an
-outcome \(y\) using a mediator \(g\) by making inferences based on a
-number of conditions that we examine in the results. To infer the
-direction of causality using the CIT method, an omnibus p-value
-generated by CIT, \(p_{CIT}\), was estimated for the correct model of
-\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
-some significance threshold \(\alpha\), causal direction was inferred
-based on the following scenarios:
+The CIT method \textsuperscript{1} is implemented in the R package
+\emph{R/cit} \textsuperscript{27}. The \emph{cit.cp} function was used
+to obtain an omnibus p-value of causality between an exposure $x$ and an
+outcome $y$ using a mediator $g$ by making inferences based on a number
+of conditions that we examine in the results. To infer the direction of
+causality using the CIT method, an omnibus p-value generated by CIT,
+$p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and
+for incorrect model of $y$ causing $x$. For some significance threshold
+$\alpha$, causal direction was inferred based on the following
+scenarios:
 
@@
 \item
-  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
-  then the correct model is accepted
+  If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then
+  the correct model is accepted
 \item
-  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
-  then the incorrect model is accepted
+  If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then
+  the incorrect model is accepted
 \item
-  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
-  then the an alternative causal model is accepted
+  If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then
+  the an alternative causal model is accepted
 \item
-  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
-  then no inference is made
+  If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no
+  inference is made
 \end{enumerate}
@@
 For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value.
+$\alpha = 0.05$ value.
 
@@
 
-Supposing there are two correlated variables, \(x\) and \(y\), with a
-variant \(g\) associated with both. Assuming no horizontal pleiotropy
-(i.e. \(g\) only has an effect on one of the variables through the other
+Supposing there are two correlated variables, $x$ and $y$, with a
+variant $g$ associated with both. Assuming no horizontal pleiotropy
+(i.e. $g$ only has an effect on one of the variables through the other
 variable) it is desirable to know which of the variables it has a direct
 influence on. This can be achieved by testing for a difference in the
-correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a population\textsuperscript{28}. It
-is calculated as
+correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for
+correlated correlations within a population \textsuperscript{28}. It is
+calculated as
 
@@
 where Fisher's z-transformation is used to obtain
-\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
+$Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )$,
 
@@
 
-The \(Z\) value is interpreted such that
+The $Z$ value is interpreted such that
 
@@
 
-and a p-value is generated from the \(Z\) value to indicate confidence
-of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
-for a causal association is then obtained by performing two sample MR in
-the direction that is predicted to be the correct model based on the Z
-score test. For the purposes of compiling simulation results, the test
-is deemed to have obtained a robust association if the Z score test has
-a p-value \(< 0.05\) and the two sample MR analysis has a p-value
-\(< 0.05\), though in practice we do not advocate using an arbitrary
-\(p\) value threshold for denoting significance in making causal
-inference.
+and a p-value is generated from the $Z$ value to indicate confidence of
+difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a
+causal association is then obtained by performing two sample MR in the
+direction that is predicted to be the correct model based on the Z score
+test. For the purposes of compiling simulation results, the test is
+deemed to have obtained a robust association if the Z score test has a
+p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$,
+though in practice we do not advocate using an arbitrary $p$ value
+threshold for denoting significance in making causal inference.
 
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{29} setting, where the Steiger test of two
-independent correlations is applied instead of the case detailed here
-where correlations share one variable in common. An advantage of using
-the Steiger test in the two sample context is that it can compare
+two-sample MR \textsuperscript{29} setting, where the Steiger test of
+two independent correlations is applied instead of the case detailed
+here where correlations share one variable in common. An advantage of
+using the Steiger test in the two sample context is that it can compare
 correlations in independent samples where sample sizes are different.
@@
 
-Simulations were conducting by creating variables of sample size \(n\)
-for the exposure \(x\), the measured values of the exposure \(x_O\), the
-outcome \(y\), the measured values of the outcome \(y_O\) and the
-instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
-instrument for \(x\). Each variable was simulated such that:
+Simulations were conducting by creating variables of sample size $n$ for
+the exposure $x$, the measured values of the exposure $x_O$, the outcome
+$y$, the measured values of the outcome $y_O$ and the instrument $g$. In
+all models $x$ causes $y$ and $g$ is an instrument for $x$. Each
+variable was simulated such that:
 
@@
 
-All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
-generated such that
+All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated
+such that
 
@@
 simualted variables. Similar patterns of results were obtained for
-different values of \(cor(g, x)\).
+different values of $cor(g, x)$.
 
@@
 Measurement error of an exposure can be modeled as some transformation
-of the true value that leads to the observed value, \(x_O=f(x)\). For
-example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
-where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
-of \(x\), and \(\epsilon_m\) represents the imprecision in the
-measurement of \(x\). Here the true value of the exposure is partially
-explained by the genetic instrument, \(g\), such that
+of the true value that leads to the observed value, $x_O=f(x)$. For
+example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where
+$\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$,
+and $\epsilon_m$ represents the imprecision in the measurement of $x$.
+Here the true value of the exposure is partially explained by the
+genetic instrument, $g$, such that
 
@@
 
-where \(\beta_g\) is the effect of the SNP on the exposure, and
-\(\epsilon_g\) is the residual value of \(x\); and the outcome is
-partially explained by the exposure
+where $\beta_g$ is the effect of the SNP on the exposure, and
+$\epsilon_g$ is the residual value of $x$; and the outcome is partially
+explained by the exposure
 
@@
 
-where \(\beta_x\) is the true effect of the exposure on the outcome. In
+where $\beta_x$ is the true effect of the exposure on the outcome. In
 the causal inference test (CIT), an omnibus p-value is generated from
-four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
-associated with \(y\); 3) \(g\) is associated with \(x|y\); and 4) \(g\)
-is independent of \(y|x\). The 4th condition is necessary for causal
-inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\), where
-\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement error
+four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is
+associated with $y$; 3) $g$ is associated with $x|y$; and 4) $g$ is
+independent of $y|x$. The 4th condition is necessary for causal
+inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where
+$\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error
 is introduced we can show using basic covariance properties (Appendix 1)
@@
 
-Thus an observational study will find \(cov(g, y - \hat{y}) = 0\) when
-the true model is causal only when \(D = 1\). Therefore, if there is any
+Thus an observational study will find $cov(g, y - \hat{y}) = 0$ when the
+true model is causal only when $D = 1$. Therefore, if there is any
 measurement error that incurs imprecision (i.e.
-\(var(\epsilon_m) \neq 0\)) then there will remain an association
-between \(g\) and \(y|x\), which is in violation of the the 4th
-condition of the CIT. Note that measurement bias alone is insufficient
-to lead to a violation of the test statistic assumptions.
+$var(\epsilon_m) \neq 0$) then there will remain an association between
+$g$ and $y|x$, which is in violation of the the 4th condition of the
+CIT. Note that measurement bias alone is insufficient to lead to a
+violation of the test statistic assumptions.
 
@@
 the CIT method. Figure 1 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
+in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is
 reliable in identifying the correct causal direction. However, as
@@
 
-When selecting an instrument \(g\) that has a direct influence on \(x\)
-and \(x\) is causally related to \(y\), one can reason that the
-influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
-\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
-effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as
+When selecting an instrument $g$ that has a direct influence on $x$ and
+$x$ is causally related to $y$, one can reason that the influence of $g$
+on $y$ is the proportion of the effect of $x$ on $y$ that is explained
+by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of
+$x$ on $y$ is estimated as
 
@@
 
-thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
-unbiased by measurement error in \(x\).
+thus we obtain an estimate of the effect of $x_O$ on $y$ that is
+unbiased by measurement error in $x$.
 
-However, if we do not know whether the SNP \(g\) has a direct influence
-on \(x\) or \(y\) then further efforts are required. If \(x\) causes
-\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
-Steiger test will obtain the correct result because the value
-\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0 when the causal
-association between \(x\) and \(y\) is \(\rho_{x, y} < 1\). But it can
-be shown that in the presence of measurement error,
-\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), thus it is
+However, if we do not know whether the SNP $g$ has a direct influence on
+$x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$
+is a valid instrument for $x$ then we expect that the Steiger test will
+obtain the correct result because the value
+$d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal
+association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be
+shown that in the presence of measurement error,
+$d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is
 important to note that under certain conditions when measurement error
 is present the Steiger test could make erroneous inference about causal
-direction because the value of \(d\) is not restricted to being greater
+direction because the value of $d$ is not restricted to being greater
 than 0. However, using the MR approach there are two potential
@@
 Figure 2 shows that in most cases, especially when the causal effect
-between \(x\) and \(y\) is not very large, the condition of \(d > 0\) is
+between $x$ and $y$ is not very large, the condition of $d > 0$ is
 satisfied.
@@
 We performed simulations to explore the performance of this approach in
-comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
+comparison to CIT. Figure 3 shows that, as predicted, when $d < 0$ the
 MR analysis is liable to infer the wrong direction of causality, and
@@
 the wrong direction of causality at a much lower rate than the CIT. When
-\(d > 0\) is satisfied we observe that in most cases the MR method has
+$d > 0$ is satisfied we observe that in most cases the MR method has
 greater power to obtain evidence for causality than CIT, and always
@@
 causality compared to CIT, while substantially improving power over CIT
-in the cases where \(d > 0\).
+in the cases where $d > 0$.
 
-Mediation based network approaches are very well
-established\textsuperscript{14} and have a number of extensions that
-make them valuable tools, including for example network construction.
-But because they are predicated on the basic underlying principles of
-mediation they are liable to suffer from the same issues of measurement
-error. Recent advances in MR methodology, for example network
-MR\textsuperscript{30} and mediation through MR\textsuperscript{31} may
-offer more robust solutions alternatives for these more complicated
-problems.
+Mediation based network approaches are very well established
+\textsuperscript{14} and have a number of extensions that make them
+valuable tools, including for example network construction. But because
+they are predicated on the basic underlying principles of mediation they
+are liable to suffer from the same issues of measurement error. Recent
+advances in MR methodology, for example network MR \textsuperscript{30}
+and mediation through MR \textsuperscript{31} may offer more robust
+solutions alternatives for these more complicated problems.
 
@@
 FTO is robustly associated with CRP, it is through the upstream
-influence of BMI\textsuperscript{25}. Thus, if BMI has a direct
+influence of BMI \textsuperscript{25}. Thus, if BMI has a direct
 influence on the outcome, the conditions of MR will be violated and the
@@
 
-\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
+\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}
+
 Figure 2: Plots depicting the function
-\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).
+$d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.
 
@@
 exposure and outcome, and to infer the direction of causality. The value
-of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
-is negative we expect the Steiger test to be more likely to be wrong
-about the direction of causality. Rows of graphs represent the sample
-size used in the simulations. For the CIT method, outcome 1 denoted
-evidence for causality with correct model, outcomes 2 or 3 denoted
-evidence for causality with incorrect model, and outcome 4 denoted no
-evidence for causality.
+of $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$, such that when $d$ is
+negative we expect the Steiger test to be more likely to be wrong about
+the direction of causality. Rows of graphs represent the sample size
+used in the simulations. For the CIT method, outcome 1 denoted evidence
+for causality with correct model, outcomes 2 or 3 denoted evidence for
+causality with incorrect model, and outcome 4 denoted no evidence for
+causality.
 
@@
 
-where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), and \(x_O\) is the measured quantity
-of \(x\), where measurement error is incurred from bias in \(\alpha_m\)
-and \(\beta_m\) and imprecision from \(\epsilon_m\), our objective is to
-estimate the expected magnitude of association between \(g\) and \(y\)
-after conditioning on \(x\). Under the CIT, this is expected to be
-\(cov(g, y - \hat{y}) = 0\) when \(x\) causes \(y\), where
-\(\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}\) is the
-predicted value of \(y\) using the measured value of \(x_O\).
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that
+has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where
+measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and
+imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of
+$y$, where measurement error is incurred from bias in $\alpha_n$ and
+$\beta_n$ and imprecision from $\epsilon_n$, our objective is to
+estimate the expected magnitude of association between $g$ and $y$ after
+conditioning on $x$. Under the CIT, this is expected to be
+$cov(g, y_O - \hat{y_O}) = 0$ when $x$ causes $y$, where
+$\hat{y}_O = a_n + \beta_n(\hat{a}_{x_O} + \hat{\beta}_{x_O} x_O)$ is
+the predicted value of $y_O$ using the measured value of $x_O$.
 
-We can split \(cov(g, y - \hat{y})\) into two parts, \(cov(g, y)\) and
-\(cov(g, \hat{y})\).
+We can split $cov(g, y_O - \hat{y}_O)$ into two parts, $cov(g, y_O)$ and
+$cov(g, \hat{y}_O)$.
 
@@
 \begin{aligned}
-\hat{\beta}_{x_O} & = \frac{cov(y, x_O)} {var(x_O)} \\
-                  & = \frac{\beta_m cov(y, x)} {var(x_O)} \\
-                  & = \frac{\beta_m \beta_x var(x)} {var(x_O)} \\
-                  & = \frac{\beta_m \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
+\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
+                  & = \frac{cov(\beta_n y, \beta_m x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m \beta_n cov(y, x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
 \end{aligned}
@@
 \begin{aligned}
-cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                 & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
@@
 
-Steiger test is used to infer if the variant \(g\) has a direct
-influence on \(x\) or \(y\) when it is known that it associates with
-both, but the direction of causality between \(x\) and \(y\) is unknown.
-Assuming the causal direction is \(x \to y\), two stage MR is formulated
-using the following regression models:
+Steiger test is used to infer if the variant $g$ has a direct influence
+on $x$ or $y$ when it is known that it associates with both, but the
+direction of causality between $x$ and $y$ is unknown. Assuming the
+causal direction is $x \to y$, two stage MR is formulated using the
+following regression models:
 
@@
 
-where \(\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g\). Writing in scale
-free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
-the exposure variable \(x\), and it is expected that
-\(\rho_{g, x} > \rho_{g, y}\) because
-\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
-causal association between \(x\) and \(y\) (which is likely to be less
-than 1). In the presence of measurement error in \(x\) and \(y\),
-however, Steiger test will instead be assessing the inequality
-\(\rho_{g, x_O} > \rho_{g, y_O}\), which can be simplified:
+where $\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g$. Writing in scale free
+terms, $\rho_{g, x}$ denotes the correlation between $g$ and the
+exposure variable $x$, and it is expected that
+$\rho_{g, x} > \rho_{g, y}$ because
+$\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the
+causal association between $x$ and $y$ (which is likely to be less than
+1). In the presence of measurement error in $x$ and $y$, however,
+Steiger test will instead be assessing the inequality
+$\rho_{g, x_O} > \rho_{g, y_O}$, which can be simplified:
 
@@
 
-1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
-molecular relationships with a causal inference test. \emph{BMC
-genetics} \textbf{10,} 23 (2009).
+1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E.\emph{et al.}
+Disentangling molecular relationships with a causal inference test.
+\emph{BMC genetics} \textbf{10,} 23 (2009).
 
@@
 
-3.Gayan, J., Gonzalez-Perez, A. \& Ruiz, A. `Does replication groups
-scoring reduce false positive rate in SNP interaction discovery?
-Response'. \emph{BMC Genomics} \textbf{11,} 403 (2010).
+3.Gayan, J., Gonzalez-Perez, A. \& Ruiz, A.\emph{et al.} `Does
+replication groups scoring reduce false positive rate in SNP interaction
+discovery? Response'. \emph{BMC Genomics} \textbf{11,} 403 (2010).
 
@@
 
-5.Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
-markers to orient the edges in quantitative trait networks: the NEO
-software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
+5.Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S.\emph{et al.}
+Using genetic markers to orient the edges in quantitative trait
+networks: the NEO software. \emph{BMC systems biology} \textbf{2,} 34
+(2008).
 
@@
 9.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
-Vandenbroucke, J. P. Quantification of bias in direct effects estimates
-due to different types of measurement error in the mediator.
-\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
+Vandenbroucke, J. P.\emph{et al.} Quantification of bias in direct
+effects estimates due to different types of measurement error in the
+mediator. \emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60
+(2012).
 
-10.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
-inference. \emph{PloS one} \textbf{8,} e80735 (2013).
+10.Nagarajan, R. \& Scutari, M. \emph{et al.} Impact of noise on
+molecular network inference. \emph{PloS one} \textbf{8,} e80735 (2013).
 
-11.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
-covariate adjustment for estimating causal effects. \emph{Proceedings of
-the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
-(UAI-10)} 527--536 (2010).
+11.Shpitser, I., VanderWeele, T. \& Robins, J.\emph{et al.} On the
+validity of covariate adjustment for estimating causal effects.
+\emph{Proceedings of the Twenty Sixth Conference on Uncertainty in
+Artificial Intelligence (UAI-10)} 527--536 (2010).
 
-12.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
-mediator matters when estimating indirect effects. \emph{Journal of
-epidemiology and community health} \textbf{67,} 458--66 (2013).
+12.Blakely, T., McKenzie, S. \& Carter, K.\emph{et al.}
+Misclassification of the mediator matters when estimating indirect
+effects. \emph{Journal of epidemiology and community health}
+\textbf{67,} 458--66 (2013).
 
-13.Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
-gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
+13.Wang, L. \& Michoel, T. \emph{et al.} Detection of regulator genes
+and eQTLs in gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
 
 14.Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
-I. in \emph{Uncertainty in biology: A computational modeling approach}
-47 (Springer, 2015). at
+I.\emph{et al.} in \emph{Uncertainty in biology: A computational
+modeling approach} 47 (Springer, 2015). at
 \textless{}\url{https://books.google.com/books?id=8SLUCgAAQBAJ\{/\&\}pgis=1}\textgreater{}
 
-15.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
-meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
-3--34 (2011).
+15.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T.\emph{et al.}
+Measurement and meaning in biology. \emph{The Quarterly Review of
+Biology} \textbf{86,} 3--34 (2011).
 
-16.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
-measurement bias. \emph{American journal of epidemiology} \textbf{170,}
-959--62; discussion 963--4 (2009).
+16.Hern{á}n, M. a \& Cole, S. R. \emph{et al.} Invited Commentary:
+Causal diagrams and measurement bias. \emph{American journal of
+epidemiology} \textbf{170,} 959--62; discussion 963--4 (2009).
 
-17.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
-pathway analysis: two potential perils in cancer studies involving DNA
-methylation array analysis. \emph{Cancer epidemiology, biomarkers \&
-prevention : a publication of the American Association for Cancer
-Research, cosponsored by the American Society of Preventive Oncology}
-\textbf{22,} 1052--60 (2013).
+17.Harper, K. N., Peters, B. a \& Gamble, M. V.\emph{et al.} Batch
+effects and pathway analysis: two potential perils in cancer studies
+involving DNA methylation array analysis. \emph{Cancer epidemiology,
+biomarkers \& prevention : a publication of the American Association for
+Cancer Research, cosponsored by the American Society of Preventive
+Oncology} \textbf{22,} 1052--60 (2013).
 
@@
 
-20.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
-obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
-\textbf{341,} 856--8 (2013).
+20.Ahima, R. S. \& Lazar, M. A. \emph{et al.} Physiology. The health
+risk of obesity--better metrics imperative. \emph{Science (New York,
+N.Y.)} \textbf{341,} 856--8 (2013).
 
-21.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
-potentials, and limitations. \emph{International journal of
-epidemiology} \textbf{33,} 30--42 (2004).
+21.{Davey Smith}, G. \& Ebrahim, S. \emph{et al.} Mendelian
+randomization: prospects, potentials, and limitations.
+\emph{International journal of epidemiology} \textbf{33,} 30--42 (2004).
 
-22.{Davey Smith}, G. \& Hemani, G. Mendelian randomization: genetic
-anchors for causal inference in epidemiological studies. \emph{Human
-molecular genetics} \textbf{23,} R89-----R98 (2014).
+22.{Davey Smith}, G. \& Hemani, G. \emph{et al.} Mendelian
+randomization: genetic anchors for causal inference in epidemiological
+studies. \emph{Human molecular genetics} \textbf{23,} R89-----R98
+(2014).
 
-23.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
-Schooling from a New Sample of Twins. \emph{The American Economic
-Review} \textbf{84,} 1157--1173 (1994).
+23.Ashenfelter, O. \& Krueger, A. B. \emph{et al.} Estimates of the
+Economic Return to Schooling from a New Sample of Twins. \emph{The
+American Economic Review} \textbf{84,} 1157--1173 (1994).
 
@@
 
-26.R Core Team. \emph{R: A Language and Environment for Statistical
-Computing}. (R Foundation for Statistical Computing, 2015). at
-\textless{}\url{https://www.r-project.org/}\textgreater{}
+26.R Core Team \emph{et al.} \emph{R: A Language and Environment for
+Statistical Computing}. (R Foundation for Statistical Computing, 2015).
+at \textless{}\url{https://www.r-project.org/}\textgreater{}
 
-27.Millstein, J. \emph{cit: Causal Inference Test. R package version
-1.9}. (2016). at
+27.Millstein, J. \emph{et al.} \emph{cit: Causal Inference Test. R
+package version 1.9}. (2016). at
 \textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
 
-28.Steiger, J. H. Tests for comparing elements of a correlation matrix.
-\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
+28.Steiger, J. H. \emph{et al.} Tests for comparing elements of a
+correlation matrix. \emph{Psychological Bulletin} \textbf{87,} 245--251
+(1980).
 
-29.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
-randomization studies: subsample and 2-sample instrumental variable
-estimators. \emph{American journal of epidemiology} \textbf{178,}
-1177--84 (2013).
+29.Pierce, B. L. \& Burgess, S. \emph{et al.} Efficient design for
+Mendelian randomization studies: subsample and 2-sample instrumental
+variable estimators. \emph{American journal of epidemiology}
+\textbf{178,} 1177--84 (2013).
 
-30.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
-Network Mendelian randomization: using genetic variants as instrumental
-variables to investigate mediation in causal pathways.
+30.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S.
+G.\emph{et al.} Network Mendelian randomization: using genetic variants
+as instrumental variables to investigate mediation in causal pathways.
 \emph{International journal of epidemiology} \textbf{44,} 484--95

### 2016-05-05 scripts/analysis.R

@@ -3,10 +3,10 @@ library(reshape2)
 library(ggplot2)
 library(plyr)
 
-load("~/repo/MethylationIV/cit/results/20141114.RData")
+load("~/repo/cit_measurement_error/results/20160504.RData")
 
 
-datmrp <- ddply(datmr, .(noise, bias, n), function(x){
+datmrp <- ddply(dat, .(noiseT, biasT, noiseG, biasG, n), function(x){
 	x <- mutate(x)
 	a <- sum(x$bdmr == "AtoB") / nrow(x)
 	ap <- sum(x$bdmr_p == "AtoB") / nrow(x)
@@ -16,9 +16,11 @@ datmrp <- ddply(datmr, .(noise, bias, n), function(x){
 	conp <- sum(x$bdmr_p == "Confounder") / nrow(x)
 	u <- sum(x$bdmr == "Unknown") / nrow(x)
 	up <- sum(x$bdmr_p == "Unknown") / nrow(x)
-	rsq <- mean(x$rsq)
+	rsqG <- mean(x$rsqG)
+	rsqT <- mean(x$rsqT)
 	y <- expand.grid(cause = c("Correct inference", "Incorrect inference", "Unknown"), variable = c("True", "Proxy"))
-	y$rsq <- rsq
+	y$rsqG <- rsqG
+	y$rsqT <- rsqT
 	y$prob <- c(a,b,u,ap,bp,up)
 	return(y)
 })
@@ -33,29 +35,36 @@ dat2$tim <- dat2$time
 dat3 <- dat2
 dat3$GT[dat3$GT < 1e-100] <- 1e-100
 
-ggplot(subset(dat3, sqrt(rsq) > 0.75 & tim=="Incorrect inference"), aes(x=sqrt(rsq),y=-log10(GT))) +
+ggplot(subset(dat3, sqrt(rsqG) > 0.75 & tim=="Incorrect inference" & biasT == 1 & noiseT==0.1), aes(x=sqrt(rsqG),y=-log10(GT))) +
 stat_smooth(aes(colour=factor(n)), se=FALSE) +
 facet_grid(tim ~ .) +
 labs(y=expression(-log[10]*p), x=expression(cor(X, X[p])), colour="")
 ggsave(file="~/repo/MethylationIV/cit/images/samplesize_zoomed.pdf")
 
 
-ggplot(subset(dat2), aes(x=sqrt(rsq),y=-log10(GT))) +
[…]

### 2016-05-05 scripts/theory.R

@@ -78,3 +78,26 @@ ggplot(dat0, aes(x=exp, y=obs)) +
 geom_point() +
 facet_grid(. ~ what)
 
+
+
+##
+
+n <- 10000
+x <- rnorm(n)
+y <- 3 + 2 * x + rnorm(n)
+y0 <- 10 + 0.1 * y + rnorm(n)/10
+
+py <- fitted.values(lm(y ~ x))
+py0 <- fitted.values(lm(y0 ~ x))
+
+plot(py ~ py0)
+
+lm(py0 ~ py)
+
+lm(y ~ x)
+lm(y0 ~ x)
+
+pred <- 10 + 0.1 * (3 + 2 * x)
+
+plot(pred ~ py0)
+

### 2016-05-05 docs/manuscript.Rmd
updated

@@
 
-Gibran Hemani* and George Davey Smith
+Gibran Hemani*, Kate Tilling and George Davey Smith
 
@@
 With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables.
-A simple method for inferring if an exposure is on the causal pathway to an outcome is through mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure.
-We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the alarming effect of increasing certainty in the wrong answer.
+A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure.
+We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
 Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible.
@@
 
-With the explosion in data that assess the human phenome, we are presented with the challenge of understanding the causal nature of the emerging putitive correlations between measures of interest.
+With rapid acceleration in technological advances that enable us to measure the human phenome, we are presented with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
+For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations.
+Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.
+
+An established tool that can resolve this problem is randomisation.
+Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the association by randomly splitting the sample into two groups, applying the exposure to one group and leaving the other as a control.
+The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship.
+In practical terms randomisation is often impossible to implement in such a controlled manner, but the principles can be employed using existing cross sectional data through the use of genetic instruments.
+
 Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008].
-The principle behind mediation analysis can be explained as follows.
+A Mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
+If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the independent variable through the mediator.
+
+This simplistic scenario, however, is seldom borne out in reality.
+Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator.
+An important tool that has shown promise in resolving this problem is randomisation,
+
+ The principle behind mediation analysis can be explained as follows. Supposing
+
  Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
@@
 
-Given the following model
+We assume the following model
 
@@
 
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y_O - \hat{y_O}) = 0$ when $x$ causes $y$, where $\hat{y}_O = a_n + \beta_n(\hat{a}_{x_O} + \hat{\beta}_{x_O} x_O)$ is the predicted value of $y_O$ using the measured value of $x_O$.
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n.
+Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
+Under the CIT, this is expected to be $cov(g, y_O - \hat{y}_O) = 0$ when $x$ causes $y$, where $\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O$ is the predicted value of $y_O$ using the measured value of $x_O$.
 
@@
 \begin{aligned}
-cov(g, \hat{y}_O) & = cov(g, \alpha_n + \beta_n \alpha_x + \beta_n \hat{\beta}_{x_O} x_O) \\
-                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m x) \\
-                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m \beta_g g) \\
-                  & = \hat{\beta}_{x_O} \beta_n \beta_m \beta_g var(g)
+cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
+                  & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
+                  & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
+                  & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
 \end{aligned}
@@
 cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
+                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
 \end{aligned}
 $$
+
+Finally
+
+$$
+\begin{aligned}
+cov(g, y_O - \hat{y}_O) & = \beta_n \beta_x \beta_g var(g) - \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
+\end{aligned}
+$$
+
+thus $cov(g, y_O - \hat{y}_O) = 0$ if the measurement imprecision in $x_O$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_O - \hat{y}_O) = 0$ will not hold.
 
@@
 
-Steiger test is used to infer if the variant $g$ has a direct influence on $x$ or $y$ when it is known that it associates with both, but the direction of causality between $x$ and $y$ is unknown.
+The Steiger test is used to infer if the variant $g$ has a direct influence on $x$ or $y$ when it is known that it associates with both, but the direction of causality between $x$ and $y$ is unknown.
 Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:

### 2016-05-05 docs/manuscript.tex
updated

--- 2016-02-25 docs/manuscript.tex
@@
 where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), and \(x_O\) is the measured quantity
-of \(x\), where measurement error is incurred from bias in \(\alpha_m\)
-and \(\beta_m\) and imprecision from \(\epsilon_m\), our objective is to
-estimate the expected magnitude of association between \(g\) and \(y\)
-after conditioning on \(x\). Under the CIT, this is expected to be
-\(cov(g, y - \hat{y}) = 0\) when \(x\) causes \(y\), where
-\(\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}\) is the
-predicted value of \(y\) using the measured value of \(x_O\).
+that has a direct effect on \(x\), \(x_O\) is the measured quantity of
+\(x\), where measurement error is incurred from bias in \(\alpha_m\) and
+\(\beta_m\) and imprecision from \(\epsilon_m\), and \(y_O\) is the
+measured quantity of \(y\), where measurement error is incurred from
+bias in \(\alpha_n\) and \(\beta_n\) and imprecision from
+\(\epsilon_n\), our objective is to estimate the expected magnitude of
+association between \(g\) and \(y\) after conditioning on \(x\). Under
+the CIT, this is expected to be \(cov(g, y_O - \hat{y}_O) = 0\) when
+\(x\) causes \(y\), where
+\(\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O\) is the predicted
+value of \(y_O\) using the measured value of \(x_O\).
 
-We can split \(cov(g, y - \hat{y})\) into two parts, \(cov(g, y)\) and
-\(cov(g, \hat{y})\).
+We can split \(cov(g, y_O - \hat{y}_O)\) into two parts, \(cov(g, y_O)\)
+and \(cov(g, \hat{y}_O)\).
 
@@
 \begin{aligned}
-cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
+cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                   & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
@@
 \begin{aligned}
-cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
+cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
+                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
 \end{aligned}
 \]
+
+Finally
+
+\[
+\begin{aligned}
+cov(g, y_O - \hat{y}_O) & = \beta_n \beta_x \beta_g var(g) - \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
+\end{aligned}
+\]
+
+thus \(cov(g, y_O - \hat{y}_O) = 0\) if the measurement imprecision in
+\(x_O\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
+then the condition \(cov(g, y_O - \hat{y}_O) = 0\) will not hold.
 

### 2016-05-05 scripts/theory.R

@@ -24,7 +24,7 @@ for(i in 1:nsim)
 
 	a_m <- runif(1)
 	b_m <- runif(1)
-	e_m <- runif(1)
+	e_m <- runif(1)*10
 
 	a_x <- runif(1)
 	b_x <- runif(1)
@@ -78,26 +78,70 @@ ggplot(dat0, aes(x=exp, y=obs)) +
 geom_point() +
 facet_grid(. ~ what)
 
+lm(obs ~ exp, dat)
+lm(obs ~ exp, dat0)
 
 
 ##
 
 n <- 10000
 x <- rnorm(n)
-y <- 3 + 2 * x + rnorm(n)
-y0 <- 10 + 0.1 * y + rnorm(n)/10
+
+bm <- 0.5
+am <- 5
+em <- rnorm(n) / 5
+em <- rep(0, n)
+
+bx <- 2
+ax <- 3
+ex <- rnorm(n)
+
+bn <- 50
+an <- 10
+en <- rnorm(n)/10
+
+x0 <- am + bm * x + em
+y <- ax + bx * x + ex
+y0 <- an + bn * y + en
 
 py <- fitted.values(lm(y ~ x))
 py0 <- fitted.values(lm(y0 ~ x))
+py00 <- fitted.values(lm(y0 ~ x0))
 
 plot(py ~ py0)
+plot(py00 ~ py0)
 
-lm(py0 ~ py)
+summary(lm(py0 ~ py00))
+summary(lm(py0 ~ py))
+
+cov(py0, py00)
+bx^2 * bn^2 * bm * var(x)
 
 lm(y ~ x)
 lm(y0 ~ x)
 
+var(py0)
+var(py00)
+
 pred <- 10 + 0.1 * (3 + 2 * x)
 
 plot(pred ~ py0)
+
+x <- rnorm(n)
+e <- rnorm(n)/2
+b <- 10
+x0 <- 3 + b * x + e
+
+cov(x, x0)
+b * var(x)
+
+
+
+n <- 1000
+u <- rnorm(n)
+a <- u + rnorm(n)
+b <- u + rnorm(n)
+summary(lm(a ~ b + u))
+
+
 

### 2016-05-06 docs/manuscript.Rmd
updating manuscript

@@
 
-With rapid acceleration in technological advances that enable us to measure the human phenome, we are presented with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
+Rapid acceleration in technological advances have occurred over the last decade that enable us to measure the human phenome in increasing detail.
+This presents us with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
 For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations.
-Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.
+Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.
 
@@
 The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship.
-In practical terms randomisation is often impossible to implement in such a controlled manner, but the principles can be employed using existing cross sectional data through the use of genetic instruments.
+This is the theoretical basis behind randomised control trials, but in practical terms randomisation is often impossible to implement in such a controlled manner.
+The principle, however, can be employed using existing cross sectional data through the use of genetic variants (instruments), which have the potentially useful property of proxying for random allocation of exposure levels by virtue of whether a causal allele was inherited.
+There have been two broad statistical approaches to exploiting the properties of genetic instruments: mediation and Mendelian randomisation.
+Comparisons are often made between the two, but a crucial difference in how they behave is in their robustness in the presence of measurement error.
+Here we will introduce the two basic approaches and summarise why measurement error is a serious issue that needs to be considered in method choice for causal analysis.
 
@@
 Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator.
+A proposed solution to this is to use a genetic instrument as the independant variable.
+Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
+Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
+This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods.
+They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
+
+This simplistic scenario, however, is seldom borne out in reality.
+Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator.
 An important tool that has shown promise in resolving this problem is randomisation,
@@
  The principle behind mediation analysis can be explained as follows. Supposing
-
- Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
-Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
-This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
 

### 2016-05-07 docs/manuscript.Rmd
updated manuscript

--- 2016-05-05 docs/manuscript.Rmd
@@
 
-With rapid acceleration in technological advances that enable us to measure the human phenome, we are presented with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
-For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations.
-Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.
+The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits.
+This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
+For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene expression all depend on understanding the causal nature of any putative observational associations.
+Despite the ubiquity of the question, commonly used cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding.
 
-An established tool that can resolve this problem is randomisation.
-Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the association by randomly splitting the sample into two groups, applying the exposure to one group and leaving the other as a control.
-The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship.
-In practical terms randomisation is often impossible to implement in such a controlled manner, but the principles can be employed using existing cross sectional data through the use of genetic instruments.
+An established statistical design that can resolve this problem is randomisation.
+Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control.
+The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship.
+This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
+The principle, however, can be employed in extant cross sectional data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels.
+Two statistical approaches to exploiting the properties of genetic instruments are commonly used: mediation and Mendelian randomisation (MR)  [@DaveySmith2004; @DaveySmithHemani2014].
 
 Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008].
-A Mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
-If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the independent variable through the mediator.
+A mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
+If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.
 
@@
 Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator.
-An important tool that has shown promise in resolving this problem is randomisation,
-
- The principle behind mediation analysis can be explained as follows. Supposing
-
+A proposed solution to this is to use a genetic instrument as the independent variable.
 Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
 Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
-This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
+This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods.
+They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
 
-While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results.
-One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013], which has been noted to be an explicit issue with CIT style approaches [@Wang2015; @Lagani2015].
-Measurement (or observational) error is the difference between a measured value of a quantity and its true value.
+MR employs the same data under a completely different model.
+Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome.
+Here the problems of confounding or reverse causation are avoided by testing for association between the outcome and the SNP.
+The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
+
+By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages.
+The main advantage of mediation is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
+In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables.
+
+Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.
+This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?).
+By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
+Such a situation can arise in many scenarios.
+Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
+For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP).
+It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI.
+Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?
+
+MR does however have some important advantages over mediation based approaches.
+First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size.
+Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR relax this assumption, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
+This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes.
+Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure variable.
+Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
+We argue that this is a serious limitation.
+
+Measurement (or observational) error is the difference between the measured value of a quantity and its true value.
 Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
-Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect sensitivity, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
-Sensitivity is also an issue, for example if the unit of measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
+Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
+Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
 
-Measurement error will of course arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
-A similar problem of biological misspecification is unavoidable in disease diagnosis.
-Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically have both bias and imprecision compared to the original variable.
+Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
+Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both bias and imprecision compared to the original variable.
 The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].
 
-An alternative statistical method that gained prominence as a solution to the problem of measurement error is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014].
-Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to stand for the exposure in an association test with the outcome.
-If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
-Indeed this approach was in part pioneered (in the form of instrumental variable (IV) analysis) as a correction for measurement error [@Ashenfelter1994], but it also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.
-
-Here, however, arises a potential issue: often the biological effect of a SNP is not known.
-If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?).
-By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
-Such a situation can arise in many scenarios.
-For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
-To illustrate a trivial example - the same FTO variant that is famously associated with BMI is also associated with C-reactive protein (CRP).
-It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would infer that CRP causes BMI.
-Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?
-
-Mediation based methods, for example CIT and other related approaches typically attempt to use the data to resolve this issue using the following logic: if the SNP-exposure association is stronger than the SNP-outcome association then the causal effect must be in the direction of exposure to outcome.
-The same logic can indeed be applied to MR, but often when applied the sensitivity of these approaches to measurement error is not considered.
-
-When being objective one must assume that measurement error is ubiquitous, and any measured variable is only an imperfect proxy of the biological quantity that the researcher intended to obtain.
-Here we show using theory and simulations how measurement error can lead to unreliable causal inference in the mediation-based CIT method, and we propose an extension to MR that, although does not eliminate the problem of measurement error, provides a framework in which to ascertain the causal direction when the biology of the instrument is not fully understood.
+Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference.
+Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
+We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood.
+This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
 
@@
 The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
-The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$ by making inferences based on a number of conditions that we examine in the results.
+The methodology of the CIT is as follows.
+Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ is causally related to an outcome $y$.
+Thus,
+
+$$
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+\end{aligned}
+$$
+
+The following tests are then performed:
+
+1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
+2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
+3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
+4. $H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*
+
+where $y - \hat{y} = y - \hat{alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
+If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$.
+The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.
+
+In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
 To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$.
-For some significance threshold $\alpha$, causal direction was inferred based on the following scenarios:
+For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:
 
-1. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
-2. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
-3. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then the an alternative causal model is accepted
-4. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made
+- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
+- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
+- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then an alternative causal model is accepted
+- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made
 
-For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value.
+For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to use p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds.
 
@@
 
-Supposing there are two correlated variables, $x$ and $y$, with a variant $g$ associated with both.
+Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. Assuming the following model
+
+$$
+y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
+$$
+
+where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected.
+A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$.
+
+The direction of the causal association was inferred as follows.
 Assuming no horizontal pleiotropy (i.e.
-$g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on.
-This can be achieved by testing for a difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
+$g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables, $x$ or $y$, it has a direct influence on.
+This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2).
+This test can be formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
 It is calculated as
@@
 
-and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$.
-The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test.
-For the purposes of compiling simulation results, the test is deemed to have obtained a robust association if the Z score test has a p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$, though in practice we do not advocate using an arbitrary $p$ value threshold for denoting significance in making causal inference.
+and a p-value is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.
+The existence of causality and its direction is inferred based on the following scenarios:
+
+- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
+- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
+- Otherwise no inference is made
+
+For the purposes of compiling simulation results, we use an arbitrary $\alpha = 0.05$ value.
 
@@
 An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different.
+The method and R function are provided in Appendix 3.
 
@@
 g & \sim Binom(2, 0.5) \\
-x & = \beta_g g + \epsilon_g \\
-x_O & = \beta_{mx} x + \epsilon_{mx} \\
-y & = \beta_x x + \epsilon_x \\
-y_O & = \beta_{my} y + \epsilon_{my} \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
 \end{aligned}
@@
 
-All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated such that
+where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and $*_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $*_{my}$ are parameters for non-differential measurement error in the outcome $y$.
+
+All $\alpha$ values were set to 0, and $\beta$ values set to 1. Values of $\epsilon_*$ were generated such that
 
@@
 cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
-var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
 n & = \{100, 1000, 10000\}
@@
 giving a total of 432 combinations of parameters.
-Each of these sets of variables was performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simualted variables.
+Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables.
 Similar patterns of results were obtained for different values of $cor(g, x)$.
@@
 
-Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$.
-For example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where $\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$, and $\epsilon_m$ represents the imprecision in the measurement of $x$.
+Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O = f(x)$.
+For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$, and $\epsilon_{mx}$ represents the imprecision in the measurement of $x$.
 Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that
@@
 
-where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the residual value of $x$; and the outcome is partially explained by the exposure
+where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure
 
@@
 where $\beta_x$ is the true effect of the exposure on the outcome.
-In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $g$ is associated with $x|y$; and 4) $g$ is independent of $y|x$.
+In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
 The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$.
@@
 \begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y) - cov(g, \hat{y})  \\
-                    & = \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
+cov(g, y - \hat{y}) & = cov(g, y_O) - cov(g, \hat{y}_O)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
 \end{aligned}
@@
 $$
-D = \frac{\beta^2_m var(x)} {\beta^2_m var(x) + var(\epsilon_m)}
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
 $$
 
-Thus an observational study will find $cov(g, y - \hat{y}) = 0$ when the true model is causal only when $D = 1$.
+Thus an observational study will find $cov(g, y_O - \hat{y_O}) = 0$ when the true model is causal only when $D = 1$.
 Therefore, if there is any measurement error that incurs imprecision (i.e.
-$var(\epsilon_m) \neq 0$) then there will remain an association between $g$ and $y|x$, which is in violation of the the 4th condition of the CIT.
-Note that measurement bias alone is insufficient to lead to a violation of the test statistic assumptions.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_O | x$, which is in violation of the the 4th condition of the CIT.
+Note that linear transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions.
 
@@
 However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
-Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.
+Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
 
@@
 $$
-\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
+\begin{aligned}
+\beta_{MR} = \frac{\beta_{y_O}}{\beta_g} & = \frac{cov(y_O, g)}{cov(x_O, g)} \\
+                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
+                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
+\end{aligned}
 $$
 
-which reduces to
-
-$$
-\lim_{n \to \infty} \hat{\beta}_{MR} = \frac{\beta_x cov(x + \epsilon_x)}{cov(x + \epsilon_x)} = \beta_x
-$$
-
-thus we obtain an estimate of the effect of $x_O$ on $y$ that is unbiased by measurement error in $x$.
+thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
+Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate, but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
 
 However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
-If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$.
+If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result.
+This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$.
 But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0.

### 2016-05-07 scripts/mr_directionality_analysis.R

@@ -376,3 +376,51 @@ facet_grid(noisea ~ noiseb)
 
 
 
+
+
+
+
+## ---- causality_exists ----
+
+
+temp <- 
+
+
+
+
+# Gather
+pl <- gather(parameters, test, direction_p_value, cit_p, cortest_p)
+pl$correct_direction <- pl$cortest_correct_direction
+pl$correct_direction[pl$test == "cit_p"] <- pl$cit_correct_direction[pl$test == "cit_p"]
+pl$test_p_value <- pl$p_test
+pl$test_p_value[pl$test == "cit_p"] <- pl$direction_p_value[pl$test == "cit_p"]
+pl <- subset(pl, select=-c(cortest_correct_direction, cit_correct_direction, p_test))
+pl$test[pl$test == "cit_p"] <- "CIT"
+pl$test[pl$test == "cortest_p"] <- "MR"
+
+# Fix NAs
+pl$test_p_value[is.infinite(pl$test_p_value)] <- max(is.finite(pl$test_p_value))
+pl$direction_p_value[is.infinite(pl$direction_p_value)] <- max(is.finite(pl$direction_p_value))
+
+psum2 <- pl %>%
+	group_by(n, p, r_ab, r_za, noisea, noiseb, test) %>%
+	summarise(
+		prop_correct_direction=sum(correct_direction, na.rm=TRUE)/n(),
+		prop_correct_direction_sig=sum(correct_direction & direction_p_value > -log10(0.05) & test_p_value > -log10(0.05), na.rm=TRUE)/n(),
+		cor_aap=mean(cor_aap, na.rm=TRUE),
+		cor_bbp=mean(cor_bbp, na.rm=TRUE),
+		cor_abp=mean(cor_abp, na.rm=TRUE),
+		cor_ab=mean(cor_ab, na.rm=TRUE),
+		cor_zap=mean(cor_zap, na.rm=TRUE),
+		cor_zbp=mean(cor_zbp, na.rm=TRUE),
+		nsim=n(),
+		test_sig = sum(test_p_value > -log10(0.05), na.rm=TRUE)/n(),
+		direction_sig = sum(direction_p_value > -log10(0.05), na.rm=TRUE)/n(),
+		rhs=mean(rhs),
+		lhs=mean(cor_aap),
+		rhs_lhs_diff=lhs-rhs,
[…]

### 2016-05-08 docs/manuscript.Rmd
updated

@@
 	"d_relationship_figure", 
-	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.",
+	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.
+$d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis.
+Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$.
+The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
 	display=FALSE
+)
+
+fig_nums(
+	"causality_exists_tpr",
+	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
+True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes)."
 )
@@
 
-### The influence of measurement error on CIT
+### The influence of measurement error on mediation
 
@@
 
-### Inferring direction of causality using MR
+### Inferring the existence of causality
 
@@
 
-thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
-Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate, but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
+Thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
+Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
 
-However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
+We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
+We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
+The performance of MR in detecting an association is unrelated to measurement error in.
+Measurement error in the outcome does not induce a substantive difference in performance between CIT and MR.
+
+### Inferring the direction of causality
+
+If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
 If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result.

### 2016-05-08 scripts/mr_directionality_analysis.R

@@ -115,7 +115,7 @@ psum1 <- pl %>%
 		prop_sig_correct=sum(correct_direction & direction_p_value > -log10(0.05) & test_p_value > -log10(0.05), na.rm=T)/n(),
 		prop_nonsig=sum(direction_p_value <= -log10(0.05) | test_p_value <= -log10(0.05), na.rm=T)/n()
 	)
-psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=10)
+psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=seq(-0.2, 1.1, by=0.1))
 
 
 ## ---- cit_measurement_error_figure ----
@@ -146,6 +146,7 @@ ineq$ablab <- paste0("cor(x,y) = ", ineq$ab)
 ggplot(ineq, aes(x=cora, y=d)) +
 geom_line(aes(colour=corb, group=corb)) +
 facet_wrap(~ ablab) +
+geom_hline(yintercept=0, linetype=2) +
 labs(x=TeX("$cor(x, x_O)$"), y="d", colour=TeX("$cor(y, y_O)$"))
 
 
@@ -153,7 +154,7 @@ labs(x=TeX("$cor(x, x_O)$"), y="d", colour=TeX("$cor(y, y_O)$"))
 
 psum2 <- gather(psum1, eval, value, prop_sig_correct, prop_sig_incorrect, prop_nonsig, factor_key=TRUE) %>%
 	group_by(n, rhs_lhs_diff_bin, r_za, eval, test) %>%
-	summarise(value=mean(value, na.rm=TRUE))
+	summarise(value=mean(value, na.rm=TRUE), nsim=n())
 
 temp2 <- do.call(rbind, strsplit(as.character(psum2$rhs_lhs_diff_bin), split=","))
 psum2$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp2[,1]))
@@ -375,52 +376,74 @@ facet_grid(noisea ~ noiseb)
 
 
 
+## ---- causality_exists_tpr ----
+
+pl$causality_exists <- TRUE
+pl$causality_exists[pl$test == "MR"] <- pl$p_bz[pl$test == "MR"] > -log10(0.05)
+pl$causality_exists[pl$test == "CIT"] <- pl$cit_res[pl$test == "CIT"] <= 3
+pl$rhs_lhs_diff_bin <- cut(pl$rhs_lhs_diff, breaks=10)
[…]

### 2016-05-08 docs/manuscript.Rmd
udpated

@@
 	"d_relationship_figure", 
-	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.
+	caption="Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative.
+When $d$ is negative the Steiger test is liable to infer the wrong direction of causality.
 $d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis.
 Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$.
+Shaded regions show the parameter space where $d$ is negative.
 The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
@@
 In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
-The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$.
-When measurement error is introduced we can show using basic covariance properties (Appendix 1) that
+The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$.
+When measurement error is introduced it can be shown using basic covariance properties (Appendix 1) that
 
@@
 
-### Inferring the existence of causality
+### Using MR to infer the existence of causality
 
-When selecting an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
+When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
 Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
@@
 
-### Inferring the direction of causality
+### Using MR to infer the direction of causality
 
 If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
-If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result.
-This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$.
-But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0.
-However, using the MR approach there are two potential advantages over the CIT.
-First, we can conduct a formal test for the direction of causality.
-Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome.
-`r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
+If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that $\rho_{g,x}$ to be greater than $\rho_{g,y}$ because the association of $g$ with $y$ is the proportion of the true causal association between $x$ and $y$ scaled by the association between $g$ and $x$.
+We can use the Steiger test of correlated correlations to formally test the difference in magnitude of $\rho_{g,x}$ and $\rho_{g,y}$, where the variable with the largest correlation on $g$ is deemed to be the variable which is imposing the causal effect.
 
-We performed simulations to explore the performance of this approach in comparison to CIT.
-`r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
-However, in a large number of cases the MR analysis infers the wrong direction of causality at a much lower rate than the CIT.
+This is a simple method to orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.
+But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
+Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0.
+However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $y$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $x$.
+It also shows that in most cases where $y$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
+
+We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect.
+`r cfig("cit_mr_comparison_figure")` shows that when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
+However, the CIT is in general much more fallable to reporting a robust causal association for the wrong direction of causality.
 When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
@@
 
-Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and unfortunately there are no solutions that work perfectly in all scenarios.
-In the face of measurement error it is evident that causal inference drawn through mediation analysis will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error for most experimental designs.
+Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and using genetic instruments as surrogates for randomisation is a useful approach.
+MR and the CIT employ genetic instruments in different ways, and unfortunately neither offer solutions that work perfectly in all scenarios.
+In the face of measurement error it is evident that causal inference drawn through mediation based methods such as the CIT analysis typically fare worse than MR and will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error present in observational data.
 
@@
 Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality.
-We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it increases the significance of the incorrect inference.
+We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
 
-Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods.
-Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces.
-This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest.
+Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation.
+Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces.
+This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest.
 But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
@@
 But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
-Recent advances in MR methodology, for example network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust solutions alternatives for these more complicated problems.
+Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012] network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.
 

### 2016-05-08 scripts/mr_directionality_analysis.R

@@ -28,8 +28,8 @@ parameters$r_ab <- round(parameters$r_ab, 1)
 parameters$r_za <- parameters$r_za^2
 parameters$noisea <- parameters$noisea^2
 parameters$noiseb <- parameters$noiseb^2
-parameters$rhs <- parameters$cor_ab * parameters$cor_bbp
-parameters$lhs <- parameters$cor_aap
+parameters$rhs <- abs(parameters$cor_ab * parameters$cor_bbp)
+parameters$lhs <- abs(parameters$cor_aap)
 parameters$rhs_lhs_diff <- parameters$lhs - parameters$rhs
 
 # Calculate correlation test
@@ -134,7 +134,7 @@ facet_grid(n ~ ., scale="free_y")
 
 ineq <- expand.grid(
 	cora = seq(0, 1, by=0.02),
-	corb = seq(0, 1, by=0.1),
+	corb = seq(0, 1, by=0.2),
 	ab = c(0.1, 0.5, 0.9)
 )
 
@@ -143,10 +143,30 @@ ineq$rhs <- ineq$corb * ineq$ab
 ineq$d <- ineq$lhs - ineq$rhs
 ineq$ablab <- paste0("cor(x,y) = ", ineq$ab)
 
+
+area <- group_by(ineq, ablab) %>%
+	do({
+
+		temp1 <- subset(., corb == max(corb) & d <= 0)
+
+		y1 <- min(temp1$d)
+		x1 <- min(temp1$cora)
+
+		y2 <- min(abs(temp1$d), 0)
+		x2 <- max(temp1$cora)
+
+		temp2 <- subset(., corb == min(corb))
+		y3 <- min(abs(temp2$d), 0)
+		x3 <- min(temp2$cora)
+
+		data.frame(cora=c(x1, x2, x3), d=c(y1, y2, y3), f=1)
+	})
+
 ggplot(ineq, aes(x=cora, y=d)) +
+geom_polygon(data=area, fill="grey", aes(x=cora, y=d)) +
 geom_line(aes(colour=corb, group=corb)) +
-facet_wrap(~ ablab) +
 geom_hline(yintercept=0, linetype=2) +
+facet_wrap(~ ablab) +
 labs(x=TeX("$cor(x, x_O)$"), y="d", colour=TeX("$cor(y, y_O)$"))
 
 
[…]

### 2016-05-09 docs/manuscript.Rmd
finished new draft

@@
 tab_nums(
-	"", 
-	caption="", 
+	"shakhtab", 
+	caption="Summary of MR analysis on gene expression and DNA methylation associations.
+'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the Steiger test.
+P(+ve effect) denotes the proportion of associations that were positive.
+$cor(\\rho_{MR}, \\rho_{P})$ is the correlation of the MR estimate obtained in this analysis and the Pearson correlation between reported in Shakhbazov et al 2016.",
 	display=FALSE)
@@
 	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes)."
+True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+	display=FALSE
 )
@@
 
+```{r shakhbazov }
+
+```
+
 ### Abstract
@@
 We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
-We argue that because measurement error is ubiquitous in phenotypic data, mediation-based causal inference should be treated with caution.
+Using this method we demonstrate that for correlations between DNA methylation and gene expression levels, DNA methylation is likely to be the causal factor.
+We show that using MR to infer causal directions is straightforward, but because measurement error is ubiquitous in observational data, mediation-based causal inference in particular should be treated with caution.
 
@@
 This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
-For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene expression all depend on understanding the causal nature of any putative observational associations.
-Despite the ubiquity of the question, commonly used cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding.
+For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.
+Despite the ubiquity of the question, commonly used observational study data are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004].
 
@@
 This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
-The principle, however, can be employed in extant cross sectional data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels.
-Two statistical approaches to exploiting the properties of genetic instruments are commonly used: mediation and Mendelian randomisation (MR)  [@DaveySmith2004; @DaveySmithHemani2014].
+The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004].
+Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation and Mendelian randomisation (MR).
 
-Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008].
-A mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
+Mediation is a commonly used statistical method for making causal inference in observational data, and it exists in many forms, from regression based systems  [@Baron1986] to structural equation modeling (SEM) [@Aten2008].
+In its simplest form mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
 If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.
 
-This simplistic scenario, however, is seldom borne out in reality.
-Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator.
-A proposed solution to this is to use a genetic instrument as the independent variable.
+This simple form, however, is prone to being unreliable.
+Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator [@Fewell2007].
+Introducing pseudo-randomisation by using a genetic instrument as the independent variable can be used to improve on these issues.
 Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
 Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
-This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods.
-They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].
+This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015].
+They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].
 
-MR employs the same data under a completely different model.
-Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome.
-Here the problems of confounding or reverse causation are avoided by testing for association between the outcome and the SNP.
+MR employs the same data under but uses a different approach in how the instrumental variable is employed.
+Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome, avoiding the problems of confounding or reverse causation by testing for association between the outcome and the SNP.
 The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
@@
 By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages.
-The main advantage of mediation is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
-In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables.
+The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
+In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.
 
@@
 It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI.
-Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?
+Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016]?
 
@@
 First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size.
-Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR relax this assumption, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
+Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes.
-Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure variable.
+Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure and outcome variables.
 Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
@@
 Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
-Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
+Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
 Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
@@
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
-Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both bias and imprecision compared to the original variable.
-The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].
+Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
+The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
 
@@
 We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood.
-This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
+This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
 
@@
 
-where $y - \hat{y} = y - \hat{alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
+where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
 If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$.
@@
 where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected.
-A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$.
+A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].
 
@@
 
-Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common.
+Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting.
+Two-sample MR refers to the case where the SNP-exposure association and SNP-outcome association are calculated in different samples (or from publicly available summary statistics).
+Here the Steiger test of two independent correlations can be applied where.
+
+$$
+Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
+$$
+
 An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different.
-The method and R function are provided in Appendix 3.
+Steiger test statistics were calculated using the *r.test* function in the R package *R/psych* [@Revelle2015].
 
@@
 
-where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and $*_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $*_{my}$ are parameters for non-differential measurement error in the outcome $y$.
+where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.
 
-All $\alpha$ values were set to 0, and $\beta$ values set to 1. Values of $\epsilon_*$ were generated such that
+All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distribted values of $\epsilon_*$ were generated such that
 
@@
 Similar patterns of results were obtained for different values of $cor(g, x)$.
+
+## Two sample MR
+
+Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels.
+Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations.
+The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.
 
@@
 
+### MR analysis of gene expression and DNA methylation
+
+We used the Steiger test to test if DNA methylation levels are more likely to causally influence gene expression levels or vice versa.
+To do this we obtained a list of `r nrow(qtl)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
+These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
+Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association.
+We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).
+
+We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test.
+We observed that the sign of the MR estimate was always in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], and that there was a moderate correlation between the magnitude of the causal correlation and the observational Pearson correlation (`r ctab("shakhtab")`).
+
+We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`)), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
+
 ## Discussion
@@
 
-In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis.
-Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality.
+In the epidemiological literature issues of measurement error in mediation is relatively well explored [@Fewell2007].
+Our analysis extends this to related  methods such as CIT that are widely used in predominantly 'omic data.
+These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely have reduced statistical power, and liable to infer the wrong direction of causality.
 We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
@@
 But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
-Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012] network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.
+Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a], network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.
 
-There are a number of limitations to the analysis that we performed.
-We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
-To give an example in which this might arise consider the case that C-reactive protein (CRP) is being tested for causal association with some outcome, and the FTO variant is used as an instrument.
-Though FTO is robustly associated with CRP, it is through the upstream influence of BMI [@Timpson2011].
-Thus, if BMI has a direct influence on the outcome, the conditions of MR will be violated and the Steiger test proposed here will be unreliable.
+In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
+Recent method developments in MR [@Bowden2015] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
+An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.
+However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood.
+We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies.
+We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
 
-The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is very susceptible to measurement error.
-With the presence of measurement error near ubiquitous in most cross sectional studies, and our ability to measure it limited, we argue that it should be given serious consideration in designing statistical analyses that attempt to make causal inference, and that any putitive results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
+Our analysis of DNA methylation and gene expression summary data demonstates the simplicity of inferring causal existence, direction and magnitude in the two sample MR framework.
+We note that verifying these results with regards to the extent to which measurement error has influenced them empirically is a challenging task and beyond the scope of this work.
+
+The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly sensitive to measurement error.
+With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
+
+\newpage
+
+## Tables
+
+```{r shakhtab }
+
+```
+
+`r tab_nums("shakhtab")`
 
@@
 
+```{r causality_exists_tpr, fig.width=10, fig.height=10 }
+
+```
+
+`r fig_nums("causality_exists_tpr")`
+
+\newpage
+
 ```{r cit_mr_comparison_figure, fig.width=10, fig.height=10 }
@@
 x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_O & = \alpha_m + \beta_m x + \epsilon_m \\
+x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
 y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_O & = \alpha_n + \beta_n y + \epsilon_n
+y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
 \end{aligned}
@@
 
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n.
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
 Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
@@
 \begin{aligned}
-cov(g, y_O) & = cov(g, \beta_n y) \\
-            & = cov(g, \beta_n \beta_x x) \\
-            & = cov(g, \beta_n \beta_x \beta_g g) \\
-            & = \beta_n \beta_x \beta_g var(g)
+cov(g, y_O) & = cov(g, \beta_{my} y) \\
+            & = cov(g, \beta_{my} \beta_x x) \\
+            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
+            & = \beta_{my} \beta_x \beta_g var(g)
 \end{aligned}
@@
 cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
-                  & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
-                  & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
-                  & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
+                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} x) \\
+                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} \beta_g g) \\
+                  & = \hat{\beta}_{x_O} \beta_{mx} \beta_g var(g)
 \end{aligned}
@@
 \hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
-                  & = \frac{cov(\beta_n y, \beta_m x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                  & = \frac{\beta_m \beta_n cov(y, x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
+                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
+                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
+                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
 \end{aligned}
@@
 \begin{aligned}
-cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
-                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
+cov(g, \hat{y}_O) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
+                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
 \end{aligned}
@@
 \begin{aligned}
-cov(g, y_O - \hat{y}_O) & = \beta_n \beta_x \beta_g var(g) - \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
+cov(g, y_O - \hat{y}_O) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
 \end{aligned}
@@
 
-where $\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g$.
+where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$.
 Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).

### 2016-05-09 docs/manuscript.tex
finished new draft

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs}
 \usepackage{graphicx}
@@
   \postdate{\par}
-  \date{05 May 2016}
+  \date{09 May 2016}
 
@@
 becoming increasingly common to use statistical tests to infer the
-causal relationships between correlated variables. A simple method for
-inferring if an exposure is on the causal pathway to an outcome is
-through mediation analysis, where the effect of an instrument on the
-outcome is tested before and after adjusting for the exposure. We show
-that in the face of measurement error this can lead to erroneous
-results, and that increasing sample size, rather than reducing bias, can
-often have the alarming effect of increasing certainty in the wrong
-answer. Commonly used tools that are predicated on this methodology such
-as the causal inference test (CIT) are shown to be susceptible. We
-demonstrate that Mendelian randomization (MR) is a method for causal
-inference that is robust to measurement error, and that a simple
-extension to MR can provide a method to estimate the direction of
-causality between correlated variables even when the biology of the
-instrument is unknown. We argue that because measurement error is
-ubiquitous in phenotypic data, mediation-based causal inference should
-be treated with caution.
+causal relationships between correlated variables. A simple and commonly
+used method for inferring if an exposure is on the causal pathway to an
+outcome is through using genetic instruments in mediation analysis,
+where the effect of an instrument on the outcome is tested before and
+after adjusting for the exposure. We show that in the face of
+measurement error this can lead to erroneous results, and that
+increasing sample size, rather than reducing bias, can often have the
+effect of increasing certainty in the wrong answer. Commonly used tools
+that are predicated on this methodology such as the causal inference
+test (CIT) are shown to be susceptible. We demonstrate that Mendelian
+randomization (MR) is a method for causal inference that is robust to
+measurement error, and that a simple extension to MR can provide a
+method to estimate the direction of causality between correlated
+variables even when the biology of the instrument is unknown. Using this
+method we demonstrate that for correlations between DNA methylation and
+gene expression levels, DNA methylation is likely to be the causal
+factor. We show that using MR to infer causal directions is
+straightforward, but because measurement error is ubiquitous in
+observational data, mediation-based causal inference in particular
+should be treated with caution.
 
@@
 
-With the explosion in data that assess the human phenome, we are
-presented with the challenge of understanding the causal nature of the
-emerging putitive correlations between measures of interest. Mediation
-is a commonly used statistical method for making causal inference in
-observational data\textsuperscript{1--4}, and it exists in many forms,
-from simple regression based systems to structural equation modeling
-(SEM)\textsuperscript{5}. The principle behind mediation analysis can be
-explained as follows. Supposing an exposure has a causal effect on an
-outcome, then an `instrument' that causes the exposure (a single
-nucleotide polymorophism (SNP), for example) should also influence the
-outcome. Therefore the influence of the instrument on the outcome
-conditional on the exposure should be zero. This forms the basis of a
-number of methods, such as the regression-based causal inference test
-(CIT)\textsuperscript{1}, an SEM implementation in the NEO
-software\textsuperscript{5}, and various Bayesian methods, which have
-been employed by a number of recent publications that make causal
-inferences, often in large scale `omics datasets\textsuperscript{6--8}.
-
-While mediation based approaches are simple and convenient to implement,
-it is important to note where they may lead to unreliable results. One
-such mechanism is measurement error\textsuperscript{9--12}, which has
-been noted to be an explicit issue with CIT style
-approaches\textsuperscript{13,14}. Measurement (or observational) error
-is the difference between a measured value of a quantity and its true
-value. Such variability can arise through a whole plethora of
-mechanisms, which are often unique to the study design and difficult to
-avoid\textsuperscript{15,16}. Array technology is now commonly used to
-obtain high throughput phenotyping at low cost, but they come with the
-problem of having imperfect sensitivity, so for example methylation
-levels as measured by the Illumina450k chip are prone to have some
-amount of noise around the true value\textsuperscript{17,18}.
-Sensitivity is also an issue, for example if the unit of measurement of
+The last decade has seem a rapid acceleration in technological advances
+that enable us to measure the human phenome in ever increasing detail,
+furnishing biologists with the tantalising opportunity to map the
+molecular basis of complex traits. This does, however, require
+overcoming the statistical challenge of correctly defining the causal
+architectures that underlie these high dimensional data. For example,
+identifying metabolic perturbations that lead to
+disease\textsuperscript{1}, or understanding the influence of epigenomic
+influences on gene expression\textsuperscript{2} all depend on
+understanding the causal nature of any putative observational
+associations. Despite the ubiquity of the question, commonly used
+observational study data are notoriously susceptible to pitfalls in
+making causal inference, with simple regression-based techniques unable
+to distinguish a true causal association from reverse causation or
+confounding\textsuperscript{3}.
+
+An established statistical design that can resolve this problem is
+randomisation. Given the hypothesis that an exposure is causally related
+to an outcome, randomisation can be employed to assess the causal nature
+of the association by randomly splitting the sample into two groups,
+subjecting one group to the exposure and leaving the other as a control.
+The association between the exposure and the outcome in this setting
+provides a robust estimate of the causal relationship. This is the
+theoretical basis behind randomised control trials, but in practice
+randomisation is often impossible to implement in an experimental
+context due to cost, scale or inability to manipulate the exposure. The
+principle, however, can be employed in extant observational data through
+the use of genetic variants associated with the exposure (instruments),
+where the inheritance of an allele serves as a random lifetime
+allocation of differential exposure levels\textsuperscript{3}. Two
+statistical approaches to exploiting the properties of genetic
+instruments are widely used: mediation and Mendelian randomisation (MR).
+
+Mediation is a commonly used statistical method for making causal
+inference in observational data, and it exists in many forms, from
+regression based systems\textsuperscript{4} to structural equation
+modeling (SEM)\textsuperscript{5}. In its simplest form mediation
+analysis seeks to explain the mechanism that underlies an observed
+relationship between an independent and a dependent variable by
+introducing a third, `mediator' variable. If the association between the
+independent and dependent variables disappears when conditioning on the
+mediator variable then this provides evidence that the independent
+variable influences the dependent variable through the mediator.
+
+This simple form, however, is prone to being unreliable. Alternative
+models of reverse causation or confounding are often indistinguishable
+from the hypothesised model of the independent variable causing the
+dependent variable via the mediator\textsuperscript{6}. Introducing
+pseudo-randomisation by using a genetic instrument as the independent
+variable can be used to improve on these issues. Supposing an exposure
+has a causal effect on an outcome, then an `instrument' that causes the
+exposure (a single nucleotide polymorophism (SNP), for example) should
+also influence the outcome. Therefore the influence of the instrument on
+the outcome conditional on the exposure should be zero. This forms the
+basis of a number of methods such as genetical
+genomics\textsuperscript{7}, the regression-based causal inference test
+(CIT)\textsuperscript{8,9}, an SEM implementation in the NEO
+software\textsuperscript{5}, and various other methods including
+Bayesian approaches\textsuperscript{10}. They have been employed by a
+number of recent publications that make causal inferences, often in
+large scale `omics datasets\textsuperscript{10--15}.
+
+MR employs the same data under but uses a different approach in how the
+instrumental variable is employed. Here an instrumental variable
+(typically a SNP) that has a robost, direct association with the
+exposure is used as a surrogate for the exposure in an association test
+with the outcome, avoiding the problems of confounding or reverse
+causation by testing for association between the outcome and the SNP.
+The causal association between the exposure and the outcome can be
+estimated by scaling the association between the SNP and the outcome by
+the association between the SNP and the exposure.
+
+By utilising genetic instruments in different ways, mediation and MR
+models have properties that confer some advantages and some
+disadvantages. The main advantage of mediation over MR is that the true
+underlying biology of the genetic instrument doesn't necessarily need to
+be known as long as it associates with the putative exposure. In the CIT
+framework (described fully in the Methods) for example, the test
+statistic is different if you test for the exposure causing the outcome
+or the outcome causing the exposure, allowing the researcher to infer
+the direction of causality between two variables by performing the test
+in both directions and choosing the model with the strongest evidence.
+
+Such is not the case for MR, where in order to infer the direction of
+causality the instrument must be known to influence the exposure
+directly, associating with the outcome only through the exposure. This
+requirement can be problematic because if there exists a putative
+association between two variables, with the instrumental SNP being
+robustly associated with each, it can be difficult to determine which of
+the two variables is subject to the direct effect of the SNP (i.e., for
+which of the two variables is the SNP a valid instrument?). By
+definition, we expect that if the association is causal then a SNP for
+the exposure will be associated with the outcome, so if the researcher
+erroneously uses the SNP as an instrument for the outcome then they are
+likely to see an apparently robust causal association of outcome on
+exposure. Such a situation can arise in many scenarios. Genome wide
+association studies (GWASs) that identify genetic associations for
+complex traits are, by design, hypothesis free and agnostic of genomic
+function, and it often takes years of follow up studies to understand
+the biological nature of a putative GWAS hit\textsuperscript{16}. For
+example, the same FTO variant that is associated with body mass index
+(BMI) is also associated with C-reactive protein (CRP). It is known that
+this is because BMI causes CRP\textsuperscript{17}, but supposing we
+only had \emph{a priori} knowledge that FTO associated with CRP, and we
+used it to instrument CRP in an MR analysis of CRP against BMI, we would
+mistakenly infer that CRP causes BMI. Another situation is where the
+causal directions between 'omic measures need to be determined, for
+example if a DNA methylation probe is associated with expression of an
+adjacent gene, then is a cis-acting SNP an instrument for the DNA
+methylation level, or the gene expression level\textsuperscript{18}?
+
+MR does however have some important advantages over mediation based
+approaches. First, MR is used as a tool to not just deduce the existance
+of causality, but to also estimate the causal effect size. Second, using
+mediation requires that the exposure, outcome and instrument variables
+are all measured in the same data, whereas recent extensions to MR
+circumvent this requirement, allowing causal inference to be drawn when
+exposure variables and outcome variables are measured in different
+samples\textsuperscript{19}. This has the crucial advantage of improving
+statistical power by allowing analysis in much larger sample sizes.
+Third, and though this is often underappreciated in fields outside
+classical epidemiology it is of crucial importance, MR is substantially
+more robust to there being measurement error in the exposure and outcome
+variables. Indeed instrumental variable (IV) analysis was in part
+initially introduced as a correction for measurement
+error\textsuperscript{20}, whereas it has been noted that both basic
+mediation\textsuperscript{21--24} and mediation methods that use
+instrumental variables\textsuperscript{25,26} are prone to be unreliable
+in its presence. We argue that this is a serious limitation.
+
+Measurement (or observational) error is the difference between the
+measured value of a quantity and its true value. Such variability can
+arise through a whole plethora of mechanisms, which are often unique to
+the study design and difficult to avoid\textsuperscript{27,28}. Array
+technology is now commonly used to obtain high throughput phenotyping at
+low cost, but they come with the problem of having imperfect resolution,
+for instance methylation levels as measured by the Illumina450k chip are
+prone to have some amount of noise around the true
+value\textsuperscript{29,30}. Relatedly, if the measurement of
 biological interest is the methylation level in a T cell, then
@@
 levels from whole blood samples because the measured value will be an
-assay of many cell types\textsuperscript{19}.
+assay of many cell types\textsuperscript{31}.
 
 Measurement error will of course arise in more low-tech data too, for
-example when measuring body mass index (BMI) one is typically interested
-in using this as a proxy for obesity, but it is clear that the
-correlation between BMI and underlying adiposity is not
-perfect\textsuperscript{20}. A similar problem of biological
-misspecification is unavoidable in disease diagnosis. Measurement error
-can also be introduced after the data has been collected, for example
-the transformation of non-normal data for the purpose of statistical
-analysis will lead to a new variable that will typically have both bias
-and imprecision compared to the original variable. The sources of
-measurement error are not limited to this list\textsuperscript{16}, and
-its impact has been explored in the context of mediation analysis in the
-epidemiological literature extensively\textsuperscript{9,12}.
-
-An alternative statistical method that gained prominence as a solution
-to the problem of measurement error is Mendelian randomisation
-(MR)\textsuperscript{21,22}. Here an instrumental variable (typically a
-SNP) that has a robost, causal association with the exposure is used to
-stand for the exposure in an association test with the outcome. If there
-is no measurement error in the SNP then the causal association between
-the exposure and the outcome can be estimated by scaling the association
-between the SNP and the outcome by the association between the SNP and
-the exposure. Indeed this approach was in part pioneered (in the form of
-instrumental variable (IV) analysis) as a correction for measurement
-error\textsuperscript{23}, but it also has the useful property that it
-guards against unmeasured confounders driving the association between
-exposure and outcome; and if the biological effect of the SNP is
-understood then it can also aid against reverse causality driving the
-association.
-
-Here, however, arises a potential issue: often the biological effect of
-a SNP is not known. If there exists a putative association between two
-variables, with a SNP being robustly associated with each, it can be
-difficult to determine which of the two variables is subject to the
-direct effect of the SNP (i.e., for which of the two variables is the
-SNP a valid instrument?). By definition, we expect that if the
-association is causal then a SNP for the exposure will be associated
-with the outcome, so if the researcher erroneously uses the SNP as an
-instrument for the outcome then they are likely to see an apparently
-robust causal association of outcome on exposure. Such a situation can
-arise in many scenarios. For example genome wide association studies
-(GWASs) that identify genetic associations for complex traits are, by
-design, hypothesis free and agnostic of genomic function, and it often
-takes years of follow up studies to understand the biological nature of
-a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
-- the same FTO variant that is famously associated with BMI is also
-associated with C-reactive protein (CRP). It is known that this is
-because BMI causes CRP\textsuperscript{25}, but supposing we only had
-\emph{a priori} knowledge that FTO associated with CRP, and we used it
-to instrument CRP in an MR analysis of CRP against BMI, we would infer
-that CRP causes BMI. Another situation is where the causal direction
-between 'omic measures need to be determined, for example if a DNA
-methylation probe is associated with expression of an adjacent gene,
-then is a cis-acting SNP an instrument for the DNA methylation level, or
-the gene expression level\textsuperscript{8}?
-
-Mediation based methods, for example CIT and other related approaches
-typically attempt to use the data to resolve this issue using the
-following logic: if the SNP-exposure association is stronger than the
-SNP-outcome association then the causal effect must be in the direction
-of exposure to outcome. The same logic can indeed be applied to MR, but
-often when applied the sensitivity of these approaches to measurement
-error is not considered.
-
-When being objective one must assume that measurement error is
-ubiquitous, and any measured variable is only an imperfect proxy of the
-biological quantity that the researcher intended to obtain. Here we show
-using theory and simulations how measurement error can lead to
-unreliable causal inference in the mediation-based CIT method, and we
-propose an extension to MR that, although does not eliminate the problem
-of measurement error, provides a framework in which to ascertain the
-causal direction when the biology of the instrument is not fully
-understood.
+example when measuring BMI one is typically interested in using this as
+a proxy for adiposity, but it is clear that the correlation between BMI
+and underlying adiposity is not perfect\textsuperscript{32}. A similar
+problem of biological misspecification is unavoidable in disease
+diagnosis, and measuring behaviour such as smoking or diet is
+notoriously difficult to do accurately. Measurement error can also be
+introduced after the data has been collected, for example the
+transformation of non-normal data for the purpose of statistical
+analysis will lead to a new variable that will typically incur both
+changes in scale and imprecision (noise) compared to the original
+variable. The sources of measurement error are not limited to this
+list\textsuperscript{28}, and its impact has been explored in the
+context of mediation analysis in the epidemiologyical literature
+extensively\textsuperscript{6,21,24}.
+
+Given the near-ubiquitous presence of measurement error in biological
+data it is vital to understand its impact on the tools we use for causal
+inference. Here we show using theory and simulations how
+non-differential measurement error and measurement imprecision can lead
+to unreliable causal inference in the mediation-based CIT method. We
+also present an extension to MR that allows researchers to ascertain the
+causal direction even when the biology of the instrument is not fully
+understood. This improves the utility of MR in cases where mediation
+based methods might have otherwise been used preferentially, and we
+apply this new method to infer the direction of causation between DNA
+methylation levels and gene expression levels.
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{26} and code is made
+the R programming language\textsuperscript{33} and code is made
 available at \href{}{github url to go here}.
@@
 
-The CIT method\textsuperscript{1} is implemented in the R package
-\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
-obtain an omnibus p-value of causality between an exposure \(x\) and an
-outcome \(y\) using a mediator \(g\) by making inferences based on a
-number of conditions that we examine in the results. To infer the
-direction of causality using the CIT method, an omnibus p-value
-generated by CIT, \(p_{CIT}\), was estimated for the correct model of
-\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
-some significance threshold \(\alpha\), causal direction was inferred
-based on the following scenarios:
+The CIT method\textsuperscript{8} is implemented in the R package
+\emph{R/cit}\textsuperscript{9}. The methodology of the CIT is as
+follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
+the exposure \(x\) is causally related to an outcome \(y\). Thus,
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+\end{aligned}
+\]
+
+The following tests are then performed:
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
+  associates with the exposure}
+\item
+  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
+  associates with the outcome}
+\item
+  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
+  associates with the outcome}
+\item
+  \(H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0\);
+  \emph{the SNP is independent of the outcome when the outcome is
+  adjusted for the exposure}
+\end{enumerate}
+
+where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
+residual of \(y\) after adjusting for the \(x\), where \(x\) is assumed
+to mediate the association between the SNP and the outcome. If all four
+tests reject the null hypothesis then it is inferred that \(x\) is
+causally related to \(y\). The CIT measures the strength of causality by
+generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
+(least extreme) p-value of the four tests, the intuition being that
+causal inference is only as strong as the weakest link in the chain of
+tests.
+
+In these analyses the \emph{cit.cp} function was used to obtain an
+omnibus p-value. To infer the direction of causality using the CIT
+method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
+for the correct model of \(x\) causing \(y\), and for incorrect model of
+\(y\) causing \(x\). For some significance threshold \(\alpha\), the
+existence of causality and its direction was inferred based on the
+following scenarios:
+
+\begin{itemize}
 \itemsep1pt\parskip0pt\parsep0pt
@@
   If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
-  then the an alternative causal model is accepted
+  then an alternative causal model is accepted
 \item
@@
   then no inference is made
-\end{enumerate}
+\end{itemize}
 
 For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value.
+\(\alpha = 0.05\) value, though we should stress that for real analyses
+it is not good practice to use p-values for making causal inference, nor
+is it reliable to depend on arbitrary significance thresholds.
 
@@
 
-Supposing there are two correlated variables, \(x\) and \(y\), with a
-variant \(g\) associated with both. Assuming no horizontal pleiotropy
-(i.e. \(g\) only has an effect on one of the variables through the other
-variable) it is desirable to know which of the variables it has a direct
-influence on. This can be achieved by testing for a difference in the
-correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
-for correlated correlations within a population\textsuperscript{28}. It
-is calculated as
+Two stage least squares (2SLS) is a commonly used technique for
+performing MR when the exposure, outcome and instrument data are all
+available in the same sample. Assuming the following model
+
+\[
+y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
+\]
+
+where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
+relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
+is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
+R package \(R/systemfit\)\textsuperscript{34}.
+
+The direction of the causal association was inferred as follows.
+Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
+of the variables through the other variable) it is desirable to know
+which of the variables, \(x\) or \(y\), it has a direct influence on.
+This can be achieved by assessing which of the two variables has the
+biggest absolute correlation with \(g\) (Appendix 2). This test can be
+formalised by testing for a difference in the correlations \(\rho_{gx}\)
+and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
+within a population\textsuperscript{35}. It is calculated as
 
@@
 
-and a p-value is generated from the \(Z\) value to indicate confidence
-of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
-for a causal association is then obtained by performing two sample MR in
-the direction that is predicted to be the correct model based on the Z
-score test. For the purposes of compiling simulation results, the test
-is deemed to have obtained a robust association if the Z score test has
-a p-value \(< 0.05\) and the two sample MR analysis has a p-value
-\(< 0.05\), though in practice we do not advocate using an arbitrary
-\(p\) value threshold for denoting significance in making causal
-inference.
+and a p-value is generated from the \(Z\) value to indicate the
+probability of obtaining the observed difference in correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis that both
+correlations are identical. The existence of causality and its direction
+is inferred based on the following scenarios:
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
+  a causal association for the correct model is accepted
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
+  a causal association for the incorrect model is accepted
+\item
+  Otherwise no inference is made
+\end{itemize}
+
+For the purposes of compiling simulation results, we use an arbitrary
+\(\alpha = 0.05\) value.
 
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{29} setting, where the Steiger test of two
-independent correlations is applied instead of the case detailed here
-where correlations share one variable in common. An advantage of using
-the Steiger test in the two sample context is that it can compare
-correlations in independent samples where sample sizes are different.
+two-sample MR\textsuperscript{19} setting. Two-sample MR refers to the
+case where the SNP-exposure association and SNP-outcome association are
+calculated in different samples (or from publicly available summary
+statistics). Here the Steiger test of two independent correlations can
+be applied where.
+
+\[
+Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
+\]
+
+An advantage of using the Steiger test in the two sample context is that
+it can compare correlations in independent samples where sample sizes
+are different. Steiger test statistics were calculated using the
+\emph{r.test} function in the R package
+\emph{R/psych}\textsuperscript{36}.
 
@@
 
-All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
-generated such that
+where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
+\(\beta_{mx}\) are parameters that represent non-differential
+measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
+and \(\beta_{my}\) are parameters for non-differential measurement error
+in the outcome \(y\).
+
+All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
+Normally distribted values of \(\epsilon_*\) were generated such that
 
@@
 
-giving a total of 432 combinations of parameters. Each of these sets of
-variables was performed 100 times, and the CIT and MR methods were
-applied to each in order to evaluate the causal association of the
-simualted variables. Similar patterns of results were obtained for
-different values of \(cor(g, x)\).
+giving a total of 432 combinations of parameters. Simulations using each
+of these sets of variables were performed 100 times, and the CIT and MR
+methods were applied to each in order to evaluate the causal association
+of the simulated variables. Similar patterns of results were obtained
+for different values of \(cor(g, x)\).
+
+\subsection{Two sample MR}\label{two-sample-mr}
+
+Two sample MR\textsuperscript{19} was performed using the summary
+statistics presented in\textsuperscript{18} for each of 458 associations
+between DNA methylation and gene expression levels. Because only summary
+statistics were available (effect, standard error, effect allele, sample
+size, p-values) for the instrumental SNP on the methylation and gene
+expression levels, the Steiger test of two independent correlations was
+used to infer the direction of causality for each of the associations.
+The Wald ratio test was then used to estimate the causal effect size for
+the estimated direction for each association.
 
@@
 \subsubsection{The influence of measurement error on
-CIT}\label{the-influence-of-measurement-error-on-cit}
+mediation}\label{the-influence-of-measurement-error-on-mediation}
 
 Measurement error of an exposure can be modeled as some transformation
-of the true value that leads to the observed value, \(x_O=f(x)\). For
-example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
-where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
-of \(x\), and \(\epsilon_m\) represents the imprecision in the
-measurement of \(x\). Here the true value of the exposure is partially
-explained by the genetic instrument, \(g\), such that
+of the true value that leads to the observed value, \(x_O = f(x)\). For
+example, we can define
+\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
+\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
+measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
+in the measurement of \(x\). Here the true value of the exposure is
+partially explained by the genetic instrument, \(g\), such that
 
@@
 where \(\beta_g\) is the effect of the SNP on the exposure, and
-\(\epsilon_g\) is the residual value of \(x\); and the outcome is
-partially explained by the exposure
+\(\epsilon_g\) is the normally distributed residual value of \(x\); and
+the outcome is partially explained by the exposure
 
@@
 four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
-associated with \(y\); 3) \(g\) is associated with \(x|y\); and 4) \(g\)
-is independent of \(y|x\). The 4th condition is necessary for causal
-inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\), where
-\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement error
-is introduced we can show using basic covariance properties (Appendix 1)
-that
+associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
+is independent of \(y|x\). The 4th condition employs mediation for
+causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement
+error is introduced it can be shown using basic covariance properties
+(Appendix 1) that
 
@@
 \[
-D = \frac{\beta^2_m var(x)} {\beta^2_m var(x) + var(\epsilon_m)}
-\]
-
-Thus an observational study will find \(cov(g, y - \hat{y}) = 0\) when
-the true model is causal only when \(D = 1\). Therefore, if there is any
-measurement error that incurs imprecision (i.e.
-\(var(\epsilon_m) \neq 0\)) then there will remain an association
-between \(g\) and \(y|x\), which is in violation of the the 4th
-condition of the CIT. Note that measurement bias alone is insufficient
-to lead to a violation of the test statistic assumptions.
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+\]
+
+Thus an observational study will find \(cov(g, y_O - \hat{y_O}) = 0\)
+when the true model is causal only when \(D = 1\). Therefore, if there
+is any measurement error that incurs imprecision (i.e.
+\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
+between \(g\) and \(y_O | x\), which is in violation of the the 4th
+condition of the CIT. Note that linear transformation of \(x\) or \(y\)
+without any incurred imprecision is insufficient to lead to a violation
+of the test statistic assumptions.
 
@@
 direction. Also of concern here is that increasing sample size does not
-solve the issue, indeed it only strengthens the evidence of the
-incorrect inference.
-
-\subsubsection{Inferring direction of causality using
-MR}\label{inferring-direction-of-causality-using-mr}
+solve the issue, indeed it only strengthens the apparent evidence for
+the incorrect inference.
+
+\subsubsection{Using MR to infer the existence of
+causality}\label{using-mr-to-infer-the-existence-of-causality}
 
 When selecting an instrument \(g\) that has a direct influence on \(x\)
-and \(x\) is causally related to \(y\), one can reason that the
+where \(x\) is causally related to \(y\), one can reason that the
 influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
@@
 \[
-\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
-\]
-
-which reduces to
-
-\[
-\lim_{n \to \infty} \hat{\beta}_{MR} = \frac{\beta_x cov(x + \epsilon_x)}{cov(x + \epsilon_x)} = \beta_x
-\]
-
-thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
-unbiased by measurement error in \(x\).
-
-However, if we do not know whether the SNP \(g\) has a direct influence
-on \(x\) or \(y\) then further efforts are required. If \(x\) causes
-\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
-Steiger test will obtain the correct result because the value
-\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0 when the causal
-association between \(x\) and \(y\) is \(\rho_{x, y} < 1\). But it can
-be shown that in the presence of measurement error,
-\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), thus it is
-important to note that under certain conditions when measurement error
-is present the Steiger test could make erroneous inference about causal
+\begin{aligned}
+\beta_{MR} = \frac{\beta_{y_O}}{\beta_g} & = \frac{cov(y_O, g)}{cov(x_O, g)} \\
+                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
+                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
+\end{aligned}
+\]
+
+Thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
+scaled by the measurement error in \(x\) and \(y\), but is unrelated to
+measurement imprecision. Measurement error and imprecision are likely to
+reduce the power of MR inferring a causal association (increased false
+negative rate), but importantly the evidence for causal association,
+based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
+inflated when the null hypothesis is true.
+
+We performed simulations to compare the performance of MR against CIT in
+detecting a causal association between simulated variables under
+different levels of imprecision simulated in the exposure. Figure 3
+shows the true positive rates between the CIT and MR for detecting a
+causal association. We observe that the CIT has lower power in all
+cases, with performance declining as measurement imprecision increases
+in the exposure. The performance of MR in detecting an association is
+unrelated to measurement error in. Measurement error in the outcome does
+not induce a substantive difference in performance between CIT and MR.
+
+\subsubsection{Using MR to infer the direction of
+causality}\label{using-mr-to-infer-the-direction-of-causality}
+
+If we do not know whether the SNP \(g\) has a direct influence on \(x\)
+or \(y\) then further efforts are required. If \(x\) causes \(y\) and
+\(g\) is a valid instrument for \(x\) then we expect that \(\rho_{g,x}\)
+to be greater than \(\rho_{g,y}\) because the association of \(g\) with
+\(y\) is the proportion of the true causal association between \(x\) and
+\(y\) scaled by the association between \(g\) and \(x\). We can use the
+Steiger test of correlated correlations to formally test the difference
+in magnitude of \(\rho_{g,x}\) and \(\rho_{g,y}\), where the variable
+with the largest correlation on \(g\) is deemed to be the variable which
+is imposing the causal effect.
+
+This is a simple method to orient the direction of causality in an MR
+analysis when the underlying biology of the SNP is not fully understood.
+But it can be shown that in the presence of measurement imprecision,
+\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), where \(d\)
+is a metric of the degree to which the Steiger test is liable to provide
+the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
+likely to be correct about the causal direction). Thus it is important
+to note that under certain conditions when measurement imprecision is
+present the Steiger test could make erroneous inference about the causal
 direction because the value of \(d\) is not restricted to being greater
-than 0. However, using the MR approach there are two potential
-advantages over the CIT. First, we can conduct a formal test for the
-direction of causality. Second the correct direction of causality can be
-inferred when measurement error in the exposure variable is present, as
-long as it is lower than the product of the measurement error in the
-outcome and the causal correlation between the exposure and the outcome.
-Figure 2 shows that in most cases, especially when the causal effect
-between \(x\) and \(y\) is not very large, the condition of \(d > 0\) is
-satisfied.
+than 0. However Figure 2 shows that when there is no measurement error
+in \(y\), the Steiger test is unlikely to infer the wrong direction of
+causality even if there is measurement error in \(x\). It also shows
+that in most cases where \(y\) is measured with error, especially when
+the causal effect between \(x\) and \(y\) is not very large, the
+condition of \(d > 0\) is satisfied.
 
 We performed simulations to explore the performance of this approach in
-comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
-MR analysis is liable to infer the wrong direction of causality, and
-that this erroneous result is more likely to occur with increasing
-sample size. However, in a large number of cases the MR analysis infers
-the wrong direction of causality at a much lower rate than the CIT. When
-\(d > 0\) is satisfied we observe that in most cases the MR method has
-greater power to obtain evidence for causality than CIT, and always
-obtains the correct direction of causality.
+comparison to CIT in terms of the rate at which a strong causal
+relationship is obtained for the correct direction of causality, and the
+rate at which a strong causal relationship is obtained where the
+reported direction of causality is incorrect. Figure 4 shows that when
+\(d < 0\) the MR analysis is indeed liable to infer the wrong direction
+of causality, and that this erroneous result is more likely to occur
+with increasing sample size. However, the CIT is in general much more
+fallable to reporting a robust causal association for the wrong
+direction of causality. When \(d > 0\) is satisfied we observe that in
+most cases the MR method has greater power to obtain evidence for
+causality than CIT, and always obtains the correct direction of
+causality.
+
+\subsubsection{MR analysis of gene expression and DNA
+methylation}\label{mr-analysis-of-gene-expression-and-dna-methylation}
+
+We used the Steiger test to test if DNA methylation levels are more
+likely to causally influence gene expression levels or vice versa. To do
+this we obtained a list of 458 gene expression - DNA methylation
+associations as reported in Shakhbazov et al\textsuperscript{18}. These
+were filtered to be located on the same chromosome, have robost
+correlations after correcting for multiple testing, and to share a SNP
+that had a robust cis-acting effect on both the DNA methylation probe
+and the gene expression probe. Using the reported summary data we used
+the Steiger test of two independent correlations to identify which
+direction of causality was most likely for each association. We found
+that the causal direction commonly goes in both directions (Table 1) but
+methylation levels were more likely to influence expression levels than
+vice versa (probability of expression causing methylation = 0.0024038
+(95\% CI 0.3446222 - 0.4667222) amongst 262 associations with
+\(p_{Steiger} < 0.05\)).
+
+We performed two sample MR\textsuperscript{19} for each association in
+the direction of causality that we estimated to be the correct direction
+from the Steiger test. We observed that the sign of the MR estimate was
+always in the same direction as the Pearson correlation coefficient
+reported by Shakhbazov et al\textsuperscript{18}, and that there was a
+moderate correlation between the magnitude of the causal correlation and
+the observational Pearson correlation (Table 1).
+
+We also observed that for associations where methylation caused gene
+expression the causal effect was more likely to be negative than for the
+associations where gene expression caused methylation (OR = 0.505822
+(95\% CI 0.2813174 - 0.89394)), suggesting that reducing gene expression
+levels at a controlling CpG typically leads to increased gene expression
+levels, consistent with expectation\textsuperscript{2}.
 
@@
 inferences using a statistical framework on observational data, and
-unfortunately there are no solutions that work perfectly in all
+using genetic instruments as surrogates for randomisation is a useful
+approach. MR and the CIT employ genetic instruments in different ways,
+and unfortunately neither offer solutions that work perfectly in all
 scenarios. In the face of measurement error it is evident that causal
-inference drawn through mediation analysis will be difficult to
-interpret, and this is a serious problem considering that it is often
-impossible to estimate the extent of measurement error for most
-experimental designs.
-
-In the epidemiological literature this is relatively well understood and
-our initial analysis simply confirms that for widely used methods such
-as CIT they are indeed liable to the same issues as standard mediation
-based analysis. Specifically, we show that as measurement error in the
-exposure variable increases, CIT is likely to infer the wrong direction
-of causality. We also demonstrate that, though seemingly unintuitive,
-increasing sample size does not resolve the issue, rather it increases
-the significance of the incorrect inference.
+inference drawn through mediation based methods such as the CIT analysis
+typically fare worse than MR and will be difficult to interpret, and
+this is a serious problem considering that it is often impossible to
+estimate the extent of measurement error present in observational data.
+
+In the epidemiological literature issues of measurement error in
+mediation is relatively well explored\textsuperscript{6}. Our analysis
+extends this to related methods such as CIT that are widely used in
+predominantly 'omic data. These methods are indeed liable to the same
+issues as standard mediation based analysis, and specifically we show
+that as measurement error in the exposure variable increases, CIT is
+likely have reduced statistical power, and liable to infer the wrong
+direction of causality. We also demonstrate that, though seemingly
+unintuitive, increasing sample size does not resolve the issue, rather
+it leads to more extreme p-values for the model that predicts the wrong
+direction of causality.
 
 Under many circumstances a practical solution to this problem is to use
-Mendelian randomisation instead of mediation based methods. Mendelian
-randomisation is robust in the face of measurement error and, given that
-the researcher has knowledge about the biology of the instrument being
-used in the analysis, can offer a direct solution to the issues that CIT
-faces. This assumption is often reasonable, for example SNPs are
-commonly used as instruments when they are found in genes with known
-biological relevance to the trait of interest. But on many occasions
-this is not the case, and it may be tempting to resort to using methods
-based on mediation in order to be able to both ascertain if there is a
-causal association and to infer the direction of causality. Here we have
-described a simple extension to MR which can be used as an alternative
-to mediation based methods. We show that this method is still liable to
-measurement error, but because it has different properties to the CIT it
-offers two main advantages. First, it uses a formal statistical
-framework to test for the robostness of the assumed direction of
-causality. Second, after testing in a comprehensive range of scenarios
-the MR based approach is less likely to infer the wrong direction of
-causality compared to CIT, while substantially improving power over CIT
-in the cases where \(d > 0\).
+Mendelian randomisation instead of methods such as the CIT or similar
+that are based on mediation. Inferring the existence of causality using
+Mendelian randomisation is robust in the face of measurement error and,
+if the researcher has knowledge about the biology of the instrument
+being used in the analysis, can offer a direct solution to the issues
+that the CIT faces. This assumption is often reasonable, for example
+SNPs are commonly used as instruments when they are found in genes with
+known biological relevance for the trait of interest. But on many
+occasions this is not the case, and it may be tempting to resort to
+using methods based on mediation in order to be able to both ascertain
+if there is a causal association and to infer the direction of
+causality. Here we have described a simple extension to MR which can be
+used as an alternative to mediation based methods. We show that this
+method is still liable to measurement error, but because it has
+different properties to the CIT it offers two main advantages. First, it
+uses a formal statistical framework to test for the robostness of the
+assumed direction of causality. Second, after testing in a comprehensive
+range of scenarios the MR based approach is less likely to infer the
+wrong direction of causality compared to CIT, while substantially
+improving power over CIT in the cases where \(d > 0\).
 
 Mediation based network approaches are very well
-established\textsuperscript{14} and have a number of extensions that
+established\textsuperscript{26} and have a number of extensions that
 make them valuable tools, including for example network construction.
@@
 mediation they are liable to suffer from the same issues of measurement
-error. Recent advances in MR methodology, for example network
-MR\textsuperscript{30} and mediation through MR\textsuperscript{31} may
-offer more robust solutions alternatives for these more complicated
-problems.
-
-There are a number of limitations to the analysis that we performed. We
-focused on the simple case of a single instrument in a single sample
-setting, and assumed that pleiotropy (the influence of the instrument on
-the outcome through a mechanism other than the exposure) was not
-present. To give an example in which this might arise consider the case
-that C-reactive protein (CRP) is being tested for causal association
-with some outcome, and the FTO variant is used as an instrument. Though
-FTO is robustly associated with CRP, it is through the upstream
-influence of BMI\textsuperscript{25}. Thus, if BMI has a direct
-influence on the outcome, the conditions of MR will be violated and the
-Steiger test proposed here will be unreliable.
+error. Recent advances in MR methodology, for example applying MR to
+genetical genomics\textsuperscript{37}, multivariate
+MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
+through MR\textsuperscript{40} may offer more robust alternatives for
+these more complicated problems.
+
+In our simulations we focused on the simple case of a single instrument
+in a single sample setting, and assumed that pleiotropy (the influence
+of the instrument on the outcome through a mechanism other than the
+exposure) was not present. Recent method developments in
+MR\textsuperscript{41} have focused on accounting for the issues that
+horizontal pleiotropy can introduce, but how they perform in the
+presence of measurement error remains to be explored. An important
+advantage that MR confers over most mediation based analysis is that it
+can be performed in two samples, which can considerably improve power
+and expand the scope of analysis. However, whether there is a
+substantive difference in two sample MR versus one sample MR in how
+measurement error has an effect is not yet fully understood. We have
+also assumed no measurement error in the genetic instrument, which is
+not unreasonable given the strict QC protocols that ensure high quality
+genotype data is available to most studies. We have restricted the scope
+to only exploring non-differential measurement error and avoided the
+complications incurred if measurement error in the exposure and outcome
+is correlated, however our analysis goes beyond many previous
+explorations of measurement error by assessing the impacts of both
+imprecision (noise) and linear transformations of the true variable on
+causal inference.
+
+Our analysis of DNA methylation and gene expression summary data
+demonstates the simplicity of inferring causal existence, direction and
+magnitude in the two sample MR framework. We note that verifying these
+results with regards to the extent to which measurement error has
+influenced them empirically is a challenging task and beyond the scope
+of this work.
 
@@
 method used, inferring causal direction using an instrument of unknown
-biology is very susceptible to measurement error. With the presence of
-measurement error near ubiquitous in most cross sectional studies, and
-our ability to measure it limited, we argue that it should be given
-serious consideration in designing statistical analyses that attempt to
-make causal inference, and that any putitive results should be
-accompanied with appropriate sensitivity analysis that assesses their
-robustness under varying levels of measurement error.
+biology is highly sensitive to measurement error. With the presence of
+measurement error near ubiquitous in most observational data, and our
+ability to measure it limited, we argue that it needs to be central to
+any consideration of approaches which are used in attempt to strengthen
+causal inference, and any putative results should be accompanied with
+appropriate sensitivity analysis that assesses their robustness under
+varying levels of measurement error.
+
+\newpage
+
+\subsection{Tables}\label{tables}
+
+\begin{longtable}[c]{@{}lrrrr@{}}
+\toprule
+Causal direction & Count & Count (\$p\_\{Steiger\} \textless{} 0.05) &
+P(+ve effect) & \(cor(\rho_{MR}, \rho_{P})\)\tabularnewline
+\midrule
+\endhead
+Expression causes Methylation & 246 & 156 & 0.3902439 &
+0.5629453\tabularnewline
+Methylation causes Expression & 212 & 106 & 0.2877358 &
+0.4384660\tabularnewline
+\bottomrule
+\end{longtable}
+
+Table 1: Summary of MR analysis on gene expression and DNA methylation
+associations. `Count (\(p_{Steiger} < 0.05\))' denotes the number of
+associations that have a p-value \textless{} 0.05 from the Steiger test.
+P(+ve effect) denotes the proportion of associations that were positive.
+\(cor(\rho_{MR}, \rho_{P})\) is the correlation of the MR estimate
+obtained in this analysis and the Pearson correlation between reported
+in Shakhbazov et al 2016.
 
@@
 
-\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
-Figure 2: Plots depicting the function
-\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).
+\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}
+
+Figure 2: Plots depicting the parameter space in which the function
+\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
+negative the Steiger test is liable to infer the wrong direction of
+causality. \(d\) is plotted on the y axis, with measurement precision of
+\(x\) (\(cor(x,x_O)\)) on the x-axis. Higher measurement precision of
+\(y\) (\(cor(y,y_O)\)) is depicted by lighter coloured lines, and
+columns of boxes are for different true values of \(cor(x,y)\). Shaded
+regions show the parameter space where \(d\) is negative. The graph
+shows that for the majority of the parameter space of the function,
+\(d\) is positive, especially where causal relationships are relatively
+weak.
 
@@
 
+\includegraphics{manuscript_files/figure-latex/causality_exists_tpr-1.pdf}
+
+Figure 3: Outcomes were simulated to be causally influenced by exposures
+with varying degrees of measurement imprecision applied to the exposure
+variable (x axis). True positive rates (y axis) for MR and CIT were
+compared with for varying levels of measurement imprecision in the
+outcome variable (rows of boxes) and sample sizes (columns of boxes).
+
+\newpage
+
 \includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}
 
-Figure 3: Outcomes were simulated to be causally influenced by
+Figure 4: Outcomes were simulated to be causally influenced by
 exposures, with varying degrees of measurement error applied to both.
@@
 that has a direct effect on \(x\), \(x_O\) is the measured quantity of
-\(x\), where measurement error is incurred from bias in \(\alpha_m\) and
-\(\beta_m\) and imprecision from \(\epsilon_m\), and \(y_O\) is the
-measured quantity of \(y\), where measurement error is incurred from
-bias in \(\alpha_n\) and \(\beta_n\) and imprecision from
-\(\epsilon_n\), our objective is to estimate the expected magnitude of
-association between \(g\) and \(y\) after conditioning on \(x\). Under
-the CIT, this is expected to be \(cov(g, y_O - \hat{y}_O) = 0\) when
-\(x\) causes \(y\), where
+\(x\), where measurement error is incurred from linear transformation in
+\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
+\(\epsilon_{mx}\), and \(y_O\) is the measured quantity of \(y\), where
+measurement error is incurred from linear transformation in
+\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
+\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
+of association between \(g\) and \(y\) after conditioning on \(x\).
+Under the CIT, this is expected to be \(cov(g, y_O - \hat{y}_O) = 0\)
+when \(x\) causes \(y\), where
 \(\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O\) is the predicted
@@
 
-Steiger test is used to infer if the variant \(g\) has a direct
+The Steiger test is used to infer if the variant \(g\) has a direct
 influence on \(x\) or \(y\) when it is known that it associates with
@@
 
-where \(\hat{x} = \hat{alpha}_1 + \hat{\beta}_1 g\). Writing in scale
+where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
 free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
@@
 
-1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
-molecular relationships with a causal inference test. \emph{BMC
-genetics} \textbf{10,} 23 (2009).
-
-2.Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
-causal associations between gene expression and disease. \emph{Nature
-Genetics} \textbf{37,} 710--717 (2005).
-
-3.Gayan, J., Gonzalez-Perez, A. \& Ruiz, A. `Does replication groups
-scoring reduce false positive rate in SNP interaction discovery?
-Response'. \emph{BMC Genomics} \textbf{11,} 403 (2010).
-
-4.Tavtigian, S. V. \emph{et al.} Rare, Evolutionarily Unlikely Missense
-Substitutions in ATM Confer Increased Risk of Breast Cancer. \emph{The
-American Journal of Human Genetics} \textbf{85,} 427--446 (2009).
+1.Kettunen, J. \emph{et al.} Genome-wide association study identifies
+multiple loci influencing human serum metabolite levels. \emph{Nature
+genetics} \textbf{44,} 269--76 (2012).
+
+2.Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+\& development} \textbf{16,} 6--21 (2002).
+
+3.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
+potentials, and limitations. \emph{International journal of
+epidemiology} \textbf{33,} 30--42 (2004).
+
+4.Baron, R. M. \& Kenny, D. A. The moderator-mediator variable
+distinction in social psychological research: conceptual, strategic, and
+statistical considerations. \emph{Journal of personality and social
+psychology} \textbf{51,} 1173--82 (1986).
 
@@
 
-6.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
+6.Fewell, Z., {Davey Smith}, G. \& Sterne, J. A. C. The impact of
+residual and unmeasured confounding in epidemiologic studies: a
+simulation study. \emph{American journal of epidemiology} \textbf{166,}
+646--55 (2007).
+
+7.Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
+causal associations between gene expression and disease. \emph{Nature
+Genetics} \textbf{37,} 710--717 (2005).
+
+8.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
+molecular relationships with a causal inference test. \emph{BMC
+genetics} \textbf{10,} 23 (2009).
+
+9.Millstein, J. \emph{cit: Causal Inference Test. R package version
+1.9}. (2016). at
+\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
+
+10.Waszak, S. M. \emph{et al.} Variation and genetic control of
+chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
+(2015).
+
+11.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
 epigenetic marks that mediate genetic risk for epithelial ovarian
@@
 
-7.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+12.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
 methylation as an intermediary of genetic risk in rheumatoid arthritis.
@@
 
-8.Waszak, S. M. \emph{et al.} Variation and genetic control of chromatin
-architecture in humans. \emph{Cell} \textbf{162,} 1039--1050 (2015).
-
-9.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+13.Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
+diabetes susceptibility loci in monozygotic twins. \emph{Nature
+communications} \textbf{5,} 5719 (2014).
+
+14.Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
+with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
+translational medicine} \textbf{6,} 257ra139 (2014).
+
+15.Hong, X. \emph{et al.} Genome-wide association study identifies
+peanut allergy-specific loci and evidence of epigenetic mediation in US
+children. \emph{Nature communications} \textbf{6,} 6304 (2015).
+
+16.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+Adipocyte Browning in Humans. \emph{The New England journal of medicine}
+\textbf{373,} 895--907 (2015).
+
+17.Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+index: elucidating direction of causation through reciprocal Mendelian
+randomization. \emph{International journal of obesity (2005)}
+\textbf{35,} 300--8 (2011).
+
+18.Shakhbazov, K. \emph{et al.} Shared genetic control of expression and
+methylation in peripheral blood. \emph{BMC genomics} \textbf{17,} 278
+(2016).
+
+19.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+randomization studies: subsample and 2-sample instrumental variable
+estimators. \emph{American journal of epidemiology} \textbf{178,}
+1177--84 (2013).
+
+20.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
+Schooling from a New Sample of Twins. \emph{The American Economic
+Review} \textbf{84,} 1157--1173 (1994).
+
+21.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
 Vandenbroucke, J. P. Quantification of bias in direct effects estimates
@@
 
-10.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+22.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
 
-11.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+23.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 
-12.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+24.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
 mediator matters when estimating indirect effects. \emph{Journal of
@@
 
-13.Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+25.Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
 gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
 
-14.Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+26.Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
 I. in \emph{Uncertainty in biology: A computational modeling approach}
@@
 
-15.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
+27.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
 meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
@@
 
-16.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
+28.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
 measurement bias. \emph{American journal of epidemiology} \textbf{170,}
@@
 
-17.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
+29.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
 pathway analysis: two potential perils in cancer studies involving DNA
@@
 
-18.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
+30.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
 polymorphic CpGs in the Illumina Infinium HumanMethylation450
@@
 
-19.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
+31.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
 measures of cell mixture distribution. \emph{BMC bioinformatics}
@@
 
-20.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
+32.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
 obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
@@
 
-21.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
-potentials, and limitations. \emph{International journal of
-epidemiology} \textbf{33,} 30--42 (2004).
-
-22.{Davey Smith}, G. \& Hemani, G. Mendelian randomization: genetic
-anchors for causal inference in epidemiological studies. \emph{Human
-molecular genetics} \textbf{23,} R89-----R98 (2014).
-
-23.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
-Schooling from a New Sample of Twins. \emph{The American Economic
-Review} \textbf{84,} 1157--1173 (1994).
-
-24.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
-Adipocyte Browning in Humans. \emph{The New England journal of medicine}
-\textbf{373,} 895--907 (2015).
-
-25.Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
-index: elucidating direction of causation through reciprocal Mendelian
-randomization. \emph{International journal of obesity (2005)}
-\textbf{35,} 300--8 (2011).
-
-26.R Core Team. \emph{R: A Language and Environment for Statistical
+33.R Core Team. \emph{R: A Language and Environment for Statistical
 Computing}. (R Foundation for Statistical Computing, 2015). at
@@
 
-27.Millstein, J. \emph{cit: Causal Inference Test. R package version
-1.9}. (2016). at
-\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
-
-28.Steiger, J. H. Tests for comparing elements of a correlation matrix.
+34.Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. \emph{Journal of Statistical
+Software} \textbf{23,} 1--40 (2007).
+
+35.Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
 
-29.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
-randomization studies: subsample and 2-sample instrumental variable
-estimators. \emph{American journal of epidemiology} \textbf{178,}
-1177--84 (2013).
-
-30.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+36.Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
+and Personality Research}. (Northwestern University, 2015). at
+\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
+
+37.Relton, C. L. \& {Davey Smith}, G. Two-step epigenetic Mendelian
+randomization: a strategy for establishing the causal role of epigenetic
+processes in pathways to disease. \emph{International journal of
+epidemiology} \textbf{41,} 161--76 (2012).
+
+38.Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+G. Using multivariable Mendelian randomization to disentangle the causal
+effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).
+
+39.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 
-31.Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+40.Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 
+41.Bowden, J., {Davey Smith}, G. \& Burgess, S. Mendelian randomization
+with invalid instruments: effect estimation and bias detection through
+Egger regression. \emph{International Journal of Epidemiology}
+\textbf{In press,} (2015).
+
 \end{document}

### 2016-05-09 scripts/mr_directionality_analysis.R

@@ -6,6 +6,7 @@ suppressPackageStartupMessages(library(tidyr))
 suppressPackageStartupMessages(library(psych))
 suppressPackageStartupMessages(library(lattice))
 suppressPackageStartupMessages(library(latex2exp))
+suppressPackageStartupMessages(library(knitr))
 
 get_cor_from_pval <- function(p, n)
 {
@@ -96,7 +97,7 @@ pl$direction_p_value[is.infinite(pl$direction_p_value)] <- max(is.finite(pl$dire
 # Reduce
 psum1 <- pl %>%
 	group_by(n, p, r_ab, r_za, noisea, noiseb, test) %>%
-	summarise(
+	dplyr::summarise(
 		prop_correct_direction=sum(correct_direction, na.rm=TRUE)/n(),
 		prop_correct_direction_sig=sum(correct_direction & direction_p_value > -log10(0.05) & test_p_value > -log10(0.05), na.rm=TRUE)/n(),
 		cor_aap=mean(cor_aap, na.rm=TRUE),
@@ -173,8 +174,8 @@ labs(x=TeX("$cor(x, x_O)$"), y="d", colour=TeX("$cor(y, y_O)$"))
 ## ---- cit_mr_comparison_figure ----
 
 psum2 <- gather(psum1, eval, value, prop_sig_correct, prop_sig_incorrect, prop_nonsig, factor_key=TRUE) %>%
-	group_by(n, rhs_lhs_diff_bin, r_za, eval, test) %>%
-	summarise(value=mean(value, na.rm=TRUE), nsim=n())
+	dplyr::group_by(n, rhs_lhs_diff_bin, r_za, eval, test) %>%
+	dplyr::summarise(value=mean(value, na.rm=TRUE), nsim=n())
 
 temp2 <- do.call(rbind, strsplit(as.character(psum2$rhs_lhs_diff_bin), split=","))
 psum2$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp2[,1]))
@@ -502,5 +503,141 @@ geom_bar(stat="identity", position="dodge", aes(fill=as.factor(test))) +
 facet_grid(noisea ~ noiseb) +
[…]

### 2016-12-13 docs/manuscript.Rmd
upated

@@
 
+## Appendix 3
+
+Sensitivity analysis for the Steiger test involves evaluating the probability that the inferred causal direction is incorrect due to differences in measurement error between the hypothesised exposure, $x$, and hypothesised outcome, $y$, variables.
+The objective is to calculate what the true $\rho_{g, x}$ and $\rho_{g, y}$ values would be given empirical estimates of $\rho_{g, x_O}$ and $\rho_{g, y_O}$ across all possible values of measurement error.
+
+$$
+\rho_{g,x} = \frac{\rho_{x,x_O}}{\rho_{g,x_O}}
+$$
+
+and
+
+$$
+\rho_{g,y} = \frac{\rho_{y,y_O}}{\rho_{g,y_O}}
+$$
+
+If $\rho_{g,x} > \rho_{g,y}$ then we infer that the causal direction is from exposure to outcome.
+Assume that all measurement error values for $x$ and $y$ are equally probable, the surface of possible values of this inequality is $z = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}$, and the objective is to calculate $P(z > 0)$, which corresponds to the proportion of measurement error values that would lead to the Steiger test being incorrect.
+
+The surface area of $z$ across the domain $\rho_{g,y_O} \leq \rho_{y,y_O} \leq 1; \rho_{g,x_O} \leq \rho_{x,x_O} \leq 1$ is obtained from
+
+$$
+\begin{aligned}
+S_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{  \frac{\partial}{\partial \rho_{x,x_o}} \left (\frac{\rho_{g,x_O}}{\rho_{x,x_O}} \right )^2 +  \left (\frac{\rho_{g,y_O}}{\rho_{y,y_O}} \right )^2   +1 }\, dydx \\
+& = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+\end{aligned}
+$$
+
+and the proportion of the surface for which $z > 0$ is obtained from
+
+$$
+\begin{aligned}
+S_{z>0} = \int^1_{\rho_{g,x_o}} \int^{R}_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+\end{aligned}
+$$
+
+where $R$ is the function for which $0 = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}$, i.e.
+
+$$
+R = \frac{\rho_{g,y_o}\rho_{g,x}}{\rho_{g,x_o}}
+$$
+
+Finally, the probability that the inference of $x \rightarrow y$ obtained from the Steiger test is incorrect due to measurement error is obtained from
+
+$$
+P(M) = \frac{S_{z>0}}{S_z}
+$$
+
+$P(M)$ can range from 0 to 1, where a small value indicates that the Steiger test is relatively insensitive to measurement error. This value is estimated through simulation.
+
 ## References

### 2016-12-13 docs/manuscript.tex
upated

@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{hyperref}
+\hypersetup{unicode=true,
+            pdftitle={Determining the existance and direction of causality in the face of measurement error},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage{longtable,booktabs}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Determining the existance and direction of causality in the face of measurement error},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
   \postdate{\par}
-  \date{09 May 2016}
+  \date{13 December 2016}
 
 \begin{document}
-
 \maketitle
@@
 \def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
@@
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
@@
 methylation levels were more likely to influence expression levels than
-vice versa (probability of expression causing methylation = 0.0024038
-(95\% CI 0.3446222 - 0.4667222) amongst 262 associations with
+vice versa (probability of expression causing methylation = 0.0019317
+(95\% CI 0.3455756 - 0.4653544) amongst 272 associations with
 \(p_{Steiger} < 0.05\)).
@@
 expression the causal effect was more likely to be negative than for the
-associations where gene expression caused methylation (OR = 0.505822
-(95\% CI 0.2813174 - 0.89394)), suggesting that reducing gene expression
-levels at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation\textsuperscript{2}.
+associations where gene expression caused methylation (OR = 0.4615363
+(95\% CI 0.259998 - 0.8052745)), suggesting that reducing gene
+expression levels at a controlling CpG typically leads to increased gene
+expression levels, consistent with expectation\textsuperscript{2}.
 
@@
 
-\begin{longtable}[c]{@{}lrrrr@{}}
+\begin{longtable}[]{@{}lrrrr@{}}
 \toprule
-Causal direction & Count & Count (\$p\_\{Steiger\} \textless{} 0.05) &
-P(+ve effect) & \(cor(\rho_{MR}, \rho_{P})\)\tabularnewline
+Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
+\(cor(\rho_{MR}, \rho_{P})\)\tabularnewline
 \midrule
 \endhead
-Expression causes Methylation & 246 & 156 & 0.3902439 &
+Expression causes Methylation & 246 & 162 & 0.3902439 &
 0.5629453\tabularnewline
-Methylation causes Expression & 212 & 106 & 0.2877358 &
+Methylation causes Expression & 212 & 110 & 0.2877358 &
 0.4384660\tabularnewline
@@
 
+\subsection{Appendix 3}\label{appendix-3}
+
+Sensitivity analysis for the Steiger test involves evaluating the
+probability that the inferred causal direction is incorrect due to
+differences in measurement error between the hypothesised exposure,
+\(x\), and hypothesised outcome, \(y\), variables. The objective is to
+calculate what the true \(\rho_{g, x}\) and \(\rho_{g, y}\) values would
+be given empirical estimates of \(\rho_{g, x_O}\) and \(\rho_{g, y_O}\)
+across all possible values of measurement error.
+
+\[
+\rho_{g,x} = \frac{\rho_{x,x_O}}{\rho_{g,x_O}}
+\]
+
+and
+
+\[
+\rho_{g,y} = \frac{\rho_{y,y_O}}{\rho_{g,y_O}}
+\]
+
+If \(\rho_{g,x} > \rho_{g,y}\) then we infer that the causal direction
+is from exposure to outcome. Assume that all measurement error values
+for \(x\) and \(y\) are equally probable, the surface of possible values
+of this inequality is
+\(z = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
+and the objective is to calculate \(P(z > 0)\), which corresponds to the
+proportion of measurement error values that would lead to the Steiger
+test being incorrect.
+
+The surface area of \(z\) across the domain
+\(\rho_{g,y_O} \leq \rho_{y,y_O} \leq 1; \rho_{g,x_O} \leq \rho_{x,x_O} \leq 1\)
+is obtained from
+
+\[
+\begin{aligned}
+S_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{  \frac{\partial}{\partial \rho_{x,x_o}} \left (\frac{\rho_{g,x_O}}{\rho_{x,x_O}} \right )^2 +  \left (\frac{\rho_{g,y_O}}{\rho_{y,y_O}} \right )^2   +1 }\, dydx \\
+& = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+\end{aligned}
+\]
+
+and the proportion of the surface for which \(z > 0\) is obtained from
+
+\[
+\begin{aligned}
+S_{z>0} = \int^1_{\rho_{g,x_o}} \int^{R}_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+\end{aligned}
+\]
+
+where \(R\) is the function for which
+\(0 = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
+i.e.
+
+\[
+R = \frac{\rho_{g,y_o}\rho_{g,x}}{\rho_{g,x_o}}
+\]
+
+Finally, the probability that the inference of \(x \rightarrow y\)
+obtained from the Steiger test is incorrect due to measurement error is
+obtained from
+
+\[
+P(M) = \frac{S_{z>0}}{S_z}
+\]
+
+\(P(M)\) can range from 0 to 1, where a small value indicates that the
+Steiger test is relatively insensitive to measurement error. This value
+is estimated through simulation.
+
 \subsection*{References}\label{references}
@@
 
-1.Kettunen, J. \emph{et al.} Genome-wide association study identifies
+\hypertarget{refs}{}
+\hypertarget{ref-Kettunen2012}{}
+1. Kettunen, J. \emph{et al.} Genome-wide association study identifies
 multiple loci influencing human serum metabolite levels. \emph{Nature
@@
 
-2.Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+\hypertarget{ref-Bird2002}{}
+2. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
 \& development} \textbf{16,} 6--21 (2002).
 
-3.{Davey Smith}, G. \& Ebrahim, S. Mendelian randomization: prospects,
+\hypertarget{ref-DaveySmith2004}{}
+3. Davey Smith, G. \& Ebrahim, S. Mendelian randomization: prospects,
 potentials, and limitations. \emph{International journal of
@@
 
-4.Baron, R. M. \& Kenny, D. A. The moderator-mediator variable
+\hypertarget{ref-Baron1986}{}
+4. Baron, R. M. \& Kenny, D. A. The moderator-mediator variable
 distinction in social psychological research: conceptual, strategic, and
@@
 
-5.Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
+\hypertarget{ref-Aten2008}{}
+5. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
 markers to orient the edges in quantitative trait networks: the NEO
@@
 
-6.Fewell, Z., {Davey Smith}, G. \& Sterne, J. A. C. The impact of
+\hypertarget{ref-Fewell2007}{}
+6. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
 residual and unmeasured confounding in epidemiologic studies: a
@@
 
-7.Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
+\hypertarget{ref-Schadt2005}{}
+7. Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
 causal associations between gene expression and disease. \emph{Nature
@@
 
-8.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
+\hypertarget{ref-Millstein2009}{}
+8. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
 molecular relationships with a causal inference test. \emph{BMC
@@
 
-9.Millstein, J. \emph{cit: Causal Inference Test. R package version
+\hypertarget{ref-Millstein2016}{}
+9. Millstein, J. \emph{cit: Causal Inference Test. R package version
 1.9}. (2016). at
@@
 
-10.Waszak, S. M. \emph{et al.} Variation and genetic control of
+\hypertarget{ref-Waszak2015}{}
+10. Waszak, S. M. \emph{et al.} Variation and genetic control of
 chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
@@
 
-11.Koestler, D. C. \emph{et al.} Integrative genomic analysis identifies
-epigenetic marks that mediate genetic risk for epithelial ovarian
-cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
-
-12.Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+\hypertarget{ref-Koestler2014}{}
+11. Koestler, D. C. \emph{et al.} Integrative genomic analysis
+identifies epigenetic marks that mediate genetic risk for epithelial
+ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
+
+\hypertarget{ref-Liu2013}{}
+12. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
 methylation as an intermediary of genetic risk in rheumatoid arthritis.
@@
 
-13.Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
+\hypertarget{ref-Yuan2014}{}
+13. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
 diabetes susceptibility loci in monozygotic twins. \emph{Nature
@@
 
-14.Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
+\hypertarget{ref-Tang2014}{}
+14. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
 with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
@@
 
-15.Hong, X. \emph{et al.} Genome-wide association study identifies
+\hypertarget{ref-Hong2015}{}
+15. Hong, X. \emph{et al.} Genome-wide association study identifies
 peanut allergy-specific loci and evidence of epigenetic mediation in US
@@
 
-16.Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+\hypertarget{ref-Claussnitzer2015}{}
+16. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
 Adipocyte Browning in Humans. \emph{The New England journal of medicine}
@@
 
-17.Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+\hypertarget{ref-Timpson2011}{}
+17. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
 index: elucidating direction of causation through reciprocal Mendelian
@@
 
-18.Shakhbazov, K. \emph{et al.} Shared genetic control of expression and
-methylation in peripheral blood. \emph{BMC genomics} \textbf{17,} 278
-(2016).
-
-19.Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+\hypertarget{ref-Shakhbazov2016}{}
+18. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
+278 (2016).
+
+\hypertarget{ref-Pierce2013}{}
+19. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
 randomization studies: subsample and 2-sample instrumental variable
@@
 
-20.Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return to
-Schooling from a New Sample of Twins. \emph{The American Economic
+\hypertarget{ref-Ashenfelter1994}{}
+20. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+to Schooling from a New Sample of Twins. \emph{The American Economic
 Review} \textbf{84,} 1157--1173 (1994).
 
-21.Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+\hypertarget{ref-LeCessie2012}{}
+21. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
 Vandenbroucke, J. P. Quantification of bias in direct effects estimates
@@
 
-22.Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+\hypertarget{ref-Nagarajan2013}{}
+22. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
 
-23.Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+\hypertarget{ref-Shpitser2010}{}
+23. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 
-24.Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+\hypertarget{ref-Blakely2013}{}
+24. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
 mediator matters when estimating indirect effects. \emph{Journal of
@@
 
-25.Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+\hypertarget{ref-Wang2015}{}
+25. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
 gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
 
-26.Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+\hypertarget{ref-Lagani2015}{}
+26. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
 I. in \emph{Uncertainty in biology: A computational modeling approach}
 47 (Springer, 2015). at
-\textless{}\url{https://books.google.com/books?id=8SLUCgAAQBAJ\{/\&\}pgis=1}\textgreater{}
-
-27.Houle, D., P{é}labon, C., Wagner, G. \& Hansen, T. Measurement and
+\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+
+\hypertarget{ref-Houle2011}{}
+27. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
 meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
@@
 
-28.Hern{á}n, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
+\hypertarget{ref-Hernan2009}{}
+28. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
 measurement bias. \emph{American journal of epidemiology} \textbf{170,}
@@
 
-29.Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
+\hypertarget{ref-Harper2013}{}
+29. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
 pathway analysis: two potential perils in cancer studies involving DNA
@@
 
-30.Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
+\hypertarget{ref-Chen2013a}{}
+30. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
 polymorphic CpGs in the Illumina Infinium HumanMethylation450
@@
 
-31.Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
+\hypertarget{ref-Houseman2012}{}
+31. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
 measures of cell mixture distribution. \emph{BMC bioinformatics}
@@
 
-32.Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
+\hypertarget{ref-Ahima2013}{}
+32. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
 obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
@@
 
-33.R Core Team. \emph{R: A Language and Environment for Statistical
+\hypertarget{ref-RCoreTeam2015}{}
+33. R Core Team. \emph{R: A Language and Environment for Statistical
 Computing}. (R Foundation for Statistical Computing, 2015). at
@@
 
-34.Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+\hypertarget{ref-Henningsen2007}{}
+34. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. \emph{Journal of Statistical
@@
 
-35.Steiger, J. H. Tests for comparing elements of a correlation matrix.
+\hypertarget{ref-Steiger1980}{}
+35. Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
 
-36.Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
+\hypertarget{ref-Revelle2015}{}
+36. Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
 and Personality Research}. (Northwestern University, 2015). at
@@
 
-37.Relton, C. L. \& {Davey Smith}, G. Two-step epigenetic Mendelian
+\hypertarget{ref-Relton2012}{}
+37. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 
-38.Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+\hypertarget{ref-Burgess2014a}{}
+38. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 
-39.Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+\hypertarget{ref-Burgess2015}{}
+39. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 
-40.Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+\hypertarget{ref-Varbo2015}{}
+40. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 
-41.Bowden, J., {Davey Smith}, G. \& Burgess, S. Mendelian randomization
+\hypertarget{ref-Bowden2015}{}
+41. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
 with invalid instruments: effect estimation and bias detection through

### 2016-12-13 scripts/mr_directionality_analysis.R

@@ -8,13 +8,6 @@ suppressPackageStartupMessages(library(lattice))
 suppressPackageStartupMessages(library(latex2exp))
 suppressPackageStartupMessages(library(knitr))
 
-get_cor_from_pval <- function(p, n)
-{
-	Fval <- qf(p, 1, n-1, low=FALSE)
-	R2 <- Fval / (n - 2 + Fval)
-	return(R2)
-}
-
 
 ## ---- read_data ----
 
@@ -509,6 +502,31 @@ labs(y="True positive rate", x = expression(cor(X, X[O])), fill="Test")
 
 ## ---- shakhbazov ----
 
+
+get_r_from_pn <- function(p, n)
+{
+	Fval <- qf(p, 1, n-1, low=FALSE)
+	# print(Fval)
+	if(!is.finite(Fval))
+	{
+		get_p_from_rn <- function(r, n)
+		{
+			fval <- r * (n-2) / (1 - r)
+			pval <- pf(fval, 1, n-1, low=FALSE)
+			return(pval)
+		}
+		optim.get_p_from_rn <- function(x, sample_size, pvalue)
+		{
+			abs(-log10(get_p_from_rn(x, sample_size)) - -log10(pvalue))
+		}
+		R2 <- optim(0.1, optim.get_p_from_rn, sample_size=n, pvalue=p)$par
+		# print(R2)
+		return(R2)
+	}
+	R2 <- Fval / (n - 2 + Fval)
+	return(R2)
+}
+
 mr_steiger <- function(p_exp, p_out, n_exp, n_out) 
 {
 	require(psych)
@@ -594,7 +612,7 @@ shakhtab <- dplyr::group_by(shakhbazov, dir) %>%
 		pos = sum(mr_eff > 0) / n(),
 		corr = cor(pearson^2, mr_r^2)
 	) %>% as.data.frame()
-names(shakhtab) <- c("Causal direction", "Count", "Count ($p_{Steiger} < 0.05$)", "P(+ve effect)", "$cor(\\rho_{MR}, \\rho_{P})$")
+names(shakhtab) <- c("Causal direction", "Count", "$p_{Steiger} < 0.05$", "P(+ve effect)", "$cor(\\rho_{MR}, \\rho_{P})$")
 
 
 shakhtest3 <- binom.test(sum(shakhbazov$dir == "Methylation causes Expression" & shakhbazov$dir_p < 0.05), sum(shakhbazov$dir_p < 0.05), 0.5)

### 2016-12-13 scripts/theory.R

@@ -145,3 +145,225 @@ summary(lm(a ~ b + u))
 
 
 
+# two variables, instrumented...
+
+# if x causes y then use steiger 
+
+we have r_iv^2 and steiger test statistic
+
+steiger requires rgx, rgy, ngx, ngy
+
+
+n <- 30000
+g <- rbinom(n, 2, 0.5)
+x <- rnorm(n) + g
+y <- 2 * x + rnorm(n)
+
+yo <- y + rnorm(n) * 5
+xo <- x + rnorm(n)
+
+bgx <- cov(g,x) / var(g)
+bgy <- cov(g,y) / var(g)
+(biv <- bgy / bgx)
+
+bgxo <- cov(g,xo) / var(g)
+bgyo <- cov(g,yo) / var(g)
+(bivo <- bgyo / bgxo)
+
+cor(g,x)^2
+cor(g,xo)^2
+cor(x,xo)^2 * cor(g,x)^2
+
+
+get_r_from_pn <- function(p, n)
+{
+	Fval <- qf(p, 1, n-1, low=FALSE)
+	R2 <- Fval / (n - 2 + Fval)
+	return(R2)
+}
+
+
+mr_steiger <- function(p_exp, p_out, n_exp, n_out) 
+{
+	require(psych)
+	index <- any(is.na(p_exp)) | any(is.na(p_out)) | any(is.na(n_exp)) | any(is.na(n_out))
+	p_exp <- p_exp[!index]
+	p_out <- p_out[!index]
+	n_exp <- n_exp[!index]
+	n_out <- n_out[!index]
+	r_exp <- get_r_from_pn(p_exp, n_exp)
+	r_out <- get_r_from_pn(p_out, n_out)
+	rtest <- psych::r.test(n = mean(n_exp), n2 = mean(n_out), r12 = r_exp, r34 = r_out)
+	l <- list(r2_exp = r_exp^2, r2_out = r_out^2, correct_causal_direction = r_exp > r_out, steiger_test = rtest$p)
+	return(l)
+}
+
+get_p_from_rn(0.46,890)
+
+
+optim()
+
+
+get_p_from_rn <- function(r, n)
+{
+	fval <- r * (n-2) / (1 - r)
+	pval <- pf(fval, 1, n-1, low=FALSE)
+	return(pval)
+}
+
+
[…]

### 2016-12-14 docs/manuscript.Rmd
updated

--- 2016-05-09 docs/manuscript.Rmd
@@
 ---
-title: "Determining the existance and direction of causality in the face of measurement error"
+title: "Determining the existence and direction of causality in the face of measurement error"
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 	"d_relationship_figure", 
-	caption="Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative.
+	caption="Plots depicting the parameter space in which the function $d = cor(x, x_o) - cor(x,y)cor(y, y_o)$ is negative.
 When $d$ is negative the Steiger test is liable to infer the wrong direction of causality.
-$d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis.
-Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$.
+$d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_o)$) on the x-axis.
+Higher measurement precision of $y$ ($cor(y,y_o)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$.
 Shaded regions show the parameter space where $d$ is negative.
@@
 CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
-The value of $d = \\rho_{x, x_O} - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality.
+The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality.
 Rows of graphs represent the sample size used in the simulations.
@@
 
-Simulations were conducting by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_O$, the outcome $y$, the measured values of the outcome $y_O$ and the instrument $g$.
+Simulations were conducting by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$.
 In all models $x$ causes $y$ and $g$ is an instrument for $x$.
@@
 x & = \alpha_g + \beta_g g + \epsilon_g \\
-x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
 y & = \alpha_x + \beta_x x + \epsilon_x \\
-y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
 \end{aligned}
@@
 
-Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O = f(x)$.
+Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$.
 For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$, and $\epsilon_{mx}$ represents the imprecision in the measurement of $x$.
@@
 In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
-The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$.
+The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
 When measurement error is introduced it can be shown using basic covariance properties (Appendix 1) that
@@
 \begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_O) - cov(g, \hat{y}_O)  \\
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
                     & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
@@
 
-Thus an observational study will find $cov(g, y_O - \hat{y_O}) = 0$ when the true model is causal only when $D = 1$.
+Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
 Therefore, if there is any measurement error that incurs imprecision (i.e.
-$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_O | x$, which is in violation of the the 4th condition of the CIT.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
 Note that linear transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions.
@@
 We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
 However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
@@
 \begin{aligned}
-\beta_{MR} = \frac{\beta_{y_O}}{\beta_g} & = \frac{cov(y_O, g)}{cov(x_O, g)} \\
+\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
                                          & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
@@
 
-Thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
+Thus we obtain an estimate of the effect of $x_o$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
 Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
@@
 This is a simple method to orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.
-But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
+But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
 Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0.
@@
 x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
 y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
 \end{aligned}
@@
 
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
+where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_o$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_o$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
 Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y_O - \hat{y}_O) = 0$ when $x$ causes $y$, where $\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O$ is the predicted value of $y_O$ using the measured value of $x_O$.
+Under the CIT, this is expected to be $cov(g, y_o - \hat{y}_o) = 0$ when $x$ causes $y$, where $\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o$ is the predicted value of $y_o$ using the measured value of $x_o$.
 
-We can split $cov(g, y_O - \hat{y}_O)$ into two parts, $cov(g, y_O)$ and $cov(g, \hat{y}_O)$.
+We can split $cov(g, y_o - \hat{y}_o)$ into two parts, $cov(g, y_o)$ and $cov(g, \hat{y}_o)$.
 
@@
 \begin{aligned}
-cov(g, y_O) & = cov(g, \beta_{my} y) \\
+cov(g, y_o) & = cov(g, \beta_{my} y) \\
             & = cov(g, \beta_{my} \beta_x x) \\
@@
 \begin{aligned}
-cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
-                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_O} \beta_{mx} \beta_g var(g)
+cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
+                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
+                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
+                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
 \end{aligned}
@@
 \begin{aligned}
-\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
+\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
                   & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
@@
 \begin{aligned}
-cov(g, \hat{y}_O) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
+cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                   & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
@@
 \begin{aligned}
-cov(g, y_O - \hat{y}_O) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
+cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
 \end{aligned}
@@
 
-thus $cov(g, y_O - \hat{y}_O) = 0$ if the measurement imprecision in $x_O$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_O - \hat{y}_O) = 0$ will not hold.
+thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.
 
@@
 Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
-In the presence of measurement error in $x$ and $y$, however, Steiger test will instead be assessing the inequality $\rho_{g, x_O} > \rho_{g, y_O}$, which can be simplified:
+In the presence of measurement error in $x$ and $y$, however, Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
 
@@
 \begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
+\rho_{g, x_o} & > \rho_{g, y_o} \\
+\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
+\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
+\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
 \end{aligned}
@@
 
+## Appendix 3
+
+The Steiger test is susceptible to inferring the wrong direction of causality under some measurement error parameters.
+For an observed association between the instrument $g$ and the measured exposure $x_o$, $\rho_{g,x_o}$, the association between $g$ and the true exposure (without measurement error) $x$, $\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}$ where the bounds $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$ are the theoretical limits of the correlation between true and measured $x$.
+The same logic applies to the association between the instrument and the outcome variable.
+
+For the values $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, the possible true $\rho_{g,x}$ and $\rho_{g,y}$ enumerated for all possible values of measurement error can be visualised as so:
+
+```{r, fig.width=6, fig.height=6}
+
+sensitivity_example <- steiger_sensitivity(0.3, 0.05)
+sensitivity_example$pl
+
+```
+
+Here, where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality.
+To evaluate the sensitivity, $S$ of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,
+
+$$
+S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
+$$
+
+$V_{z}$, the total volume of the function, can be obtained analytically by solving the double integral:
+
+$$
+\begin{aligned}
+V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
+& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
+\end{aligned}
+$$
+
+$V_{z>0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
+The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$.
+Hence,
+
+$$
+\begin{aligned}
+V_{z>0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
+& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
+\end{aligned}
+$$
+
+$S$ can take values from 0 to 0.5, where low values suggest low sensitivity of the Steiger test to measurement error.
+For the values used to plot the function above, $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, we obtain $S = `r round(sensitivity_example$sensitivity,3)`$.
+Hence, the majority of the parameter space of measurement error will not change the inferred causal direction of $x \rightarrow y$ for these values.
+
 ## References

### 2016-12-14 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Determining the existance and direction of causality in the face of measurement error},
+            pdftitle={Determining the existence and direction of causality in the face of measurement error},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Determining the existance and direction of causality in the face of
+  \title{Determining the existence and direction of causality in the face of
 measurement error}
@@
   \postdate{\par}
-  \date{13 December 2016}
+  \date{14 December 2016}
 
@@
 Simulations were conducting by creating variables of sample size \(n\)
-for the exposure \(x\), the measured values of the exposure \(x_O\), the
-outcome \(y\), the measured values of the outcome \(y_O\) and the
+for the exposure \(x\), the measured values of the exposure \(x_o\), the
+outcome \(y\), the measured values of the outcome \(y_o\) and the
 instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
@@
 Measurement error of an exposure can be modeled as some transformation
-of the true value that leads to the observed value, \(x_O = f(x)\). For
+of the true value that leads to the observed value, \(x_o = f(x)\). For
 example, we can define
@@
 causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
 error is introduced it can be shown using basic covariance properties
@@
 
-Thus an observational study will find \(cov(g, y_O - \hat{y_O}) = 0\)
+Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
 when the true model is causal only when \(D = 1\). Therefore, if there
@@
 \(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
-between \(g\) and \(y_O | x\), which is in violation of the the 4th
+between \(g\) and \(y_o | x\), which is in violation of the the 4th
 condition of the CIT. Note that linear transformation of \(x\) or \(y\)
@@
 the CIT method. Figure 1 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
+in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
 reliable in identifying the correct causal direction. However, as
@@
 
-Thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
+Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
 scaled by the measurement error in \(x\) and \(y\), but is unrelated to
@@
 But it can be shown that in the presence of measurement imprecision,
-\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), where \(d\)
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
 is a metric of the degree to which the Steiger test is liable to provide
@@
 
-\begin{longtable}[]{@{}lrrrr@{}}
+\begin{longtable}[]{@{}lrrrrr@{}}
 \toprule
 Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
-\(cor(\rho_{MR}, \rho_{P})\)\tabularnewline
+\(cor(\rho_{MR}, \rho_{P})\) & Mean sensitivity\tabularnewline
 \midrule
 \endhead
-Expression causes Methylation & 246 & 162 & 0.3902439 &
-0.5629453\tabularnewline
-Methylation causes Expression & 212 & 110 & 0.2877358 &
-0.4384660\tabularnewline
+Expression causes Methylation & 246 & 162 & 0.3902439 & 0.5629453 &
+0.2031556\tabularnewline
+Methylation causes Expression & 212 & 110 & 0.2877358 & 0.4384660 &
+0.2365816\tabularnewline
 \bottomrule
@@
 Figure 2: Plots depicting the parameter space in which the function
-\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
+\(d = cor(x, x_o) - cor(x,y)cor(y, y_o)\) is negative. When \(d\) is
 negative the Steiger test is liable to infer the wrong direction of
 causality. \(d\) is plotted on the y axis, with measurement precision of
-\(x\) (\(cor(x,x_O)\)) on the x-axis. Higher measurement precision of
-\(y\) (\(cor(y,y_O)\)) is depicted by lighter coloured lines, and
+\(x\) (\(cor(x,x_o)\)) on the x-axis. Higher measurement precision of
+\(y\) (\(cor(y,y_o)\)) is depicted by lighter coloured lines, and
 columns of boxes are for different true values of \(cor(x,y)\). Shaded
@@
 exposure and outcome, and to infer the direction of causality. The value
-of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
+of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\)
 is negative we expect the Steiger test to be more likely to be wrong
@@
 where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), \(x_O\) is the measured quantity of
+that has a direct effect on \(x\), \(x_o\) is the measured quantity of
 \(x\), where measurement error is incurred from linear transformation in
 \(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
-\(\epsilon_{mx}\), and \(y_O\) is the measured quantity of \(y\), where
+\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
 measurement error is incurred from linear transformation in
@@
 of association between \(g\) and \(y\) after conditioning on \(x\).
-Under the CIT, this is expected to be \(cov(g, y_O - \hat{y}_O) = 0\)
+Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
 when \(x\) causes \(y\), where
-\(\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O\) is the predicted
-value of \(y_O\) using the measured value of \(x_O\).
-
-We can split \(cov(g, y_O - \hat{y}_O)\) into two parts, \(cov(g, y_O)\)
-and \(cov(g, \hat{y}_O)\).
+\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
+value of \(y_o\) using the measured value of \(x_o\).
+
+We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
+and \(cov(g, \hat{y}_o)\).
 
@@
 
-thus \(cov(g, y_O - \hat{y}_O) = 0\) if the measurement imprecision in
-\(x_O\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
-then the condition \(cov(g, y_O - \hat{y}_O) = 0\) will not hold.
+thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
+\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
+then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.
 
@@
 however, Steiger test will instead be assessing the inequality
-\(\rho_{g, x_O} > \rho_{g, y_O}\), which can be simplified:
+\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
 
@@
 
-Sensitivity analysis for the Steiger test involves evaluating the
-probability that the inferred causal direction is incorrect due to
-differences in measurement error between the hypothesised exposure,
-\(x\), and hypothesised outcome, \(y\), variables. The objective is to
-calculate what the true \(\rho_{g, x}\) and \(\rho_{g, y}\) values would
-be given empirical estimates of \(\rho_{g, x_O}\) and \(\rho_{g, y_O}\)
-across all possible values of measurement error.
-
-\[
-\rho_{g,x} = \frac{\rho_{x,x_O}}{\rho_{g,x_O}}
-\]
-
-and
-
-\[
-\rho_{g,y} = \frac{\rho_{y,y_O}}{\rho_{g,y_O}}
-\]
-
-If \(\rho_{g,x} > \rho_{g,y}\) then we infer that the causal direction
-is from exposure to outcome. Assume that all measurement error values
-for \(x\) and \(y\) are equally probable, the surface of possible values
-of this inequality is
-\(z = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
-and the objective is to calculate \(P(z > 0)\), which corresponds to the
-proportion of measurement error values that would lead to the Steiger
-test being incorrect.
-
-The surface area of \(z\) across the domain
-\(\rho_{g,y_O} \leq \rho_{y,y_O} \leq 1; \rho_{g,x_O} \leq \rho_{x,x_O} \leq 1\)
-is obtained from
+The Steiger test is susceptible to inferring the wrong direction of
+causality under some measurement error parameters. For an observed
+association between the instrument \(g\) and the measured exposure
+\(x_o\), \(\rho_{g,x_o}\), the association between \(g\) and the true
+exposure (without measurement error) \(x\),
+\(\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}\) where the bounds
+\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\) are the theoretical limits of
+the correlation between true and measured \(x\). The same logic applies
+to the association between the instrument and the outcome variable.
+
+For the values \(\rho_{g,x_o} = 0.3\) and \(\rho_{g,y_o} = 0.05\), the
+possible true \(\rho_{g,x}\) and \(\rho_{g,y}\) enumerated for all
+possible values of measurement error can be visualised as so:
+
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}
+
+Here, where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error
+range in \(x_o\) and \(y_o\) will not change the qualitative inference
+of the direction of causality. To evaluate the sensitivity, \(S\) of the
+Steiger test to measurement error, the objective is to calculate the
+proportion of the total absolute volume of this function that is above
+the \(z = 0\) plane,
+
+\[
+S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
+\]
+
+\(V_{z}\), the total volume of the function, can be obtained
+analytically by solving the double integral:
 
@@
 \begin{aligned}
-S_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{  \frac{\partial}{\partial \rho_{x,x_o}} \left (\frac{\rho_{g,x_O}}{\rho_{x,x_O}} \right )^2 +  \left (\frac{\rho_{g,y_O}}{\rho_{y,y_O}} \right )^2   +1 }\, dydx \\
-& = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
+& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
 \end{aligned}
@@
 
-and the proportion of the surface for which \(z > 0\) is obtained from
+\(V_{z>0}\), the proportion of the volume that lies above the \(z=0\)
+plane, can also be obtained analytically. The region of this volume is
+bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) where
+\(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
+\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,
 
@@
 \begin{aligned}
-S_{z>0} = \int^1_{\rho_{g,x_o}} \int^{R}_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
+V_{z>0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
+& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
 \end{aligned}
@@
 
-where \(R\) is the function for which
-\(0 = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
-i.e.
-
-\[
-R = \frac{\rho_{g,y_o}\rho_{g,x}}{\rho_{g,x_o}}
-\]
-
-Finally, the probability that the inference of \(x \rightarrow y\)
-obtained from the Steiger test is incorrect due to measurement error is
-obtained from
-
-\[
-P(M) = \frac{S_{z>0}}{S_z}
-\]
-
-\(P(M)\) can range from 0 to 1, where a small value indicates that the
-Steiger test is relatively insensitive to measurement error. This value
-is estimated through simulation.
+\(S\) can take values from 0 to 0.5, where low values suggest low
+sensitivity of the Steiger test to measurement error. For the values
+used to plot the function above, \(\rho_{g,x_o} = 0.3\) and
+\(\rho_{g,y_o} = 0.05\), we obtain \(S = 0.032\). Hence, the majority of
+the parameter space of measurement error will not change the inferred
+causal direction of \(x \rightarrow y\) for these values.
 

### 2016-12-14 scripts/mr_directionality_analysis.R

@@ -502,7 +502,6 @@ labs(y="True positive rate", x = expression(cor(X, X[O])), fill="Test")
 
 ## ---- shakhbazov ----
 
-
 get_r_from_pn <- function(p, n)
 {
 	Fval <- qf(p, 1, n-1, low=FALSE)
@@ -527,6 +526,49 @@ get_r_from_pn <- function(p, n)
 	return(R2)
 }
 
+steiger_sensitivity <- function(rgx_o, rgy_o)
+{
+	if(rgy_o > rgx_o)
+	{
+		a <- rgy_o
+		b <- rgx_o
+	} else {
+		a <- rgx_o
+		b <- rgy_o
+	}
+
+	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=70), ryy_o=seq(rgy_o,1,length.out=70), type=c("A","B"))
+	d$rgy <- rgy_o / d$ryy_o
+	d$rgx <- rgx_o / d$rxx_o
+	d$z <- d$rgy - d$rgx
+	d$z[d$type=="A"] <- 0
+	temp <- wireframe(
+		z ~ rxx_o * ryy_o, 
+		groups=type, 
+		data=d, 
+		scales=list(arrows=FALSE), 
+		col.groups = colorRampPalette(c("red", "blue"))(2), 
+		drape=FALSE, 
+		xlab=expression(rho[xx[o]]), 
+		ylab=expression(rho[yy[o]]),
+		zlab=expression(rho[gy]-rho[gx])
+	)
+
+
+	vz <- a * log(a) - b * log(b) + a*b*(log(b)-log(a))
+	vz0 <- -2*b - b * log(a) - a*b*log(a) + 2*a*b
+
+	sensitivity <- vz0 / (2 * vz0 + abs(vz))
+
+	return(list(
+		vz = vz,
+		vz0 = vz0,
+		sensitivity = sensitivity,
+		pl = temp
+	))
+}
+
+
 mr_steiger <- function(p_exp, p_out, n_exp, n_out) 
 {
 	require(psych)
@@ -537,8 +579,20 @@ mr_steiger <- function(p_exp, p_out, n_exp, n_out)
 	n_out <- n_out[!index]
 	r_exp <- get_r_from_pn(p_exp, n_exp)
 	r_out <- get_r_from_pn(p_out, n_out)
+
+	sensitivity <- steiger_sensitivity(r_exp, r_out)
+
[…]

### 2016-12-14 scripts/theory.R

@@ -257,7 +257,6 @@ c = xy
 library(lattice)
 
 
-
 rgx_o <- 0.7
 rgy_o <- 0.2
 
@@ -267,7 +266,7 @@ d$rgx <- rgx_o / d$rxx_o
 d$z <- d$rgy - d$rgx
 d$z[d$type=="A"] <- 0
 d$col <- d$z > 0
-wireframe(z ~ ryy_o * rxx_o, groups=type, data=d, scales=list(arrows=FALSE), col.groups = colorRampPalette(c("red", "blue"))(2), drape=FALSE, pretty=TRUE)
+temp <- wireframe(z ~ ryy_o * rxx_o, groups=type, data=d, scales=list(arrows=FALSE), col.groups = colorRampPalette(c("red", "blue"))(2), drape=FALSE, pretty=TRUE)
 
 0.6*0.8
 
@@ -326,44 +325,102 @@ plot(act ~ bin)
 
 
 
-a=0.3
-b=0.1
-x=0.5
-
-a/x^2 + x^2/(2*a) - x^2*b^2/(6*a) - (a*b/x^2 + b*x^2/(2*a) - x^2*b^2/(6*a*b^3))
-a/x^2 * (1-b) + x^2/(2*a) * (1 - b^2/3 - b + 1/(3*b))
-
-a/x^2 - a*b/x^2
-a/x^2 * (1-b)
-
-x^2/(2*a) - x^2*b^2/(6*a) - b*x^2/(2*a) - x^2*b^2 / (6*a*b^3)
-x^2/(2*a) * (1 - b^2/3 - 1/(3*b))
-
-
-
-
-optim.get_p_from_rn <- function(x, sample_size, pvalue)
+get_r_from_pn <- function(p, n)
 {
-	abs(-log10(get_p_from_rn(x, sample_size)) - -log10(pvalue))
+	Fval <- qf(p, 1, n-1, low=FALSE)
+	# print(Fval)
+	if(!is.finite(Fval))
+	{
+		get_p_from_rn <- function(r, n)
+		{
+			fval <- r * (n-2) / (1 - r)
+			pval <- pf(fval, 1, n-1, low=FALSE)
+			return(pval)
+		}
+		optim.get_p_from_rn <- function(x, sample_size, pvalue)
+		{
+			abs(-log10(get_p_from_rn(x, sample_size)) - -log10(pvalue))
+		}
+		R2 <- optim(0.1, optim.get_p_from_rn, sample_size=n, pvalue=p)$par
[…]

### 2016-12-15 scripts/mr_directionality_analysis.R
updated

@@ -505,12 +505,11 @@ labs(y="True positive rate", x = expression(cor(X, X[O])), fill="Test")
 get_r_from_pn <- function(p, n)
 {
 	Fval <- qf(p, 1, n-1, low=FALSE)
-	# print(Fval)
 	if(!is.finite(Fval))
 	{
-		get_p_from_rn <- function(r, n)
+		get_p_from_r2n <- function(r2, n)
 		{
-			fval <- r * (n-2) / (1 - r)
+			fval <- r2 * (n-2) / (1 - r2)
 			pval <- pf(fval, 1, n-1, low=FALSE)
 			return(pval)
 		}
@@ -519,11 +518,10 @@ get_r_from_pn <- function(p, n)
 			abs(-log10(get_p_from_rn(x, sample_size)) - -log10(pvalue))
 		}
 		R2 <- optim(0.1, optim.get_p_from_rn, sample_size=n, pvalue=p)$par
-		# print(R2)
 		return(R2)
 	}
 	R2 <- Fval / (n - 2 + Fval)
-	return(R2)
+	return(sqrt(R2))
 }
 
 steiger_sensitivity <- function(rgx_o, rgy_o)
@@ -537,7 +535,7 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 		b <- rgy_o
 	}
 
-	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=70), ryy_o=seq(rgy_o,1,length.out=70), type=c("A","B"))
+	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=50), ryy_o=seq(rgy_o,1,length.out=50), type=c("A","B"))
 	d$rgy <- rgy_o / d$ryy_o
 	d$rgx <- rgx_o / d$rxx_o
 	d$z <- d$rgy - d$rgx
@@ -554,7 +552,6 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 		zlab=expression(rho[gy]-rho[gx])
 	)
 
-
 	vz <- a * log(a) - b * log(b) + a*b*(log(b)-log(a))
 	vz0 <- -2*b - b * log(a) - a*b*log(a) + 2*a*b
 
@@ -569,7 +566,7 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 }
 
 
[…]

### 2016-12-15 scripts/mr_directionality_analysis.R
updated

@@ -497,9 +497,6 @@ facet_grid(noisea ~ noiseb) +
 labs(y="True positive rate", x = expression(cor(X, X[O])), fill="Test")
 
 
-
-
-
 ## ---- shakhbazov ----
 
 get_r_from_pn <- function(p, n)
@@ -515,7 +512,7 @@ get_r_from_pn <- function(p, n)
 		}
 		optim.get_p_from_rn <- function(x, sample_size, pvalue)
 		{
-			abs(-log10(get_p_from_rn(x, sample_size)) - -log10(pvalue))
+			abs(-log10(get_p_from_r2n(x, sample_size)) - -log10(pvalue))
 		}
 		R2 <- optim(0.1, optim.get_p_from_rn, sample_size=n, pvalue=p)$par
 		return(R2)
@@ -612,16 +609,16 @@ mr_wald_ratio <- function(b_exp, b_out, se_exp, se_out, parameters)
 	return(list(b = b, se = se, pval = pval, nsnp = 1))
 }
 
-qtl <- read.csv("../data/12864_2016_2498_MOESM16_ESM.csv")
-qtl <- qtl[order(qtl$PVALUE_expr), ]
-qtl <- subset(qtl, !duplicated(paste0(meth_ind, exp_ind)))
+qtl_orig <- read.csv("../data/12864_2016_2498_MOESM16_ESM.csv")
+qtl_orig <- qtl_orig[order(qtl_orig$PVALUE_expr), ]
+qtl_orig <- subset(qtl_orig, !duplicated(paste0(meth_ind, exp_ind)))
 cors <- read.csv("../data/12864_2016_2498_MOESM8_ESM.csv")
 
 me <- expand.grid(rxx_o=c(0.5,0.75,1), ryy_o=c(0.5,0.75,1))
 qtl <- group_by(me, rxx_o, ryy_o) %>%
 	do({
 		x <- .
-		y <- qtl
+		y <- qtl_orig
 		y$rxx_o <- x$rxx_o[1]
 		y$ryy_o <- x$ryy_o[1]
 		return(y)
@@ -634,8 +631,6 @@ qtl$dir_p <- NA
 qtl$mr_eff <- NA
 qtl$mr_se <- NA
 qtl$mr_p <- NA
-
-
 for(i in 1:nrow(qtl))
 {
 	l <- mr_steiger(
[…]

### 2016-12-19 scripts/mr_directionality_multiple.R
uppdated

@@ -1,0 +1,181 @@
+library(TwoSampleMR)
+library(ggplot2)
+library(tidyr)
+
+sim_y_from_x <- function(x, r)
+{
+	x <- scale(x)
+	stopifnot(ncol(x) == length(r))
+	stopifnot(sum(r^2) <= 1)
+	e <- scale(rnorm(nrow(x))) * sqrt(1-sum(r^2))
+	g <- x %*% r
+	y <- g + e
+	return(y)
+}
+
+make_vector_effects <- function(x, r2, pleiotropy_direction=NULL, pleiotropy_cor=NULL)
+{
+	if(!is.null(pleiotropy_cor))
+	{
+		stopifnot(pleiotropy_cor <= 1 & pleiotropy_cor >= -1)
+		nsnp <- length(x)
+		eff <- sim_y_from_x(x, pleiotropy_cor)
+	} else {
+		nsnp <- x
+		eff <- rnorm(nsnp)
+	}
+	eff2 <- sign(eff) * eff^2
+	eff2 <- eff2 / sum(abs(eff2)) * r2
+	eff <- sign(eff2) * sqrt(abs(eff2))
+	if(!is.null(pleiotropy_direction))
+	{
+		eff <- eff + pleiotropy_direction * sign(x)
+	}
+	return(eff)
+}
+
+make_proxy <- function(x, noise, bias)
+{
+	y <- x*bias + rnorm(length(x), 0, sd=noise)
+	return(y)
+}
+
+
+make_system <- function(nid, nsnp, cor_A_B, cor_Z_A, cor_Z_B, pleiotropy_direction, cor_pleiotropy, noisea, noiseb)
+{
+
+	g <- matrix(rbinom(nid * nsnp, 2, 0.5), nid, nsnp)
+
+	a <- make_vector_effects(nsnp, cor_Z_A^2)
+	b <- make_vector_effects(a, cor_Z_B, pleiotropy_direction, cor_pleiotropy)
+	cor(a,b)
+
+	A <- sim_y_from_x(g, a)
+	B <- sim_y_from_x(cbind(A, g), c(cor_A_B, b))
+
+	Ap <- make_proxy(A, noisea, 1)
+	Bp <- make_proxy(B, noiseb, 1)
+
+	return(data.frame(
+		Z = g,
+		A = A,
+		B = B,
+		Ap = Ap,
+		Bp = Bp
+	))
+}
+
+
+get_regression <- function(x, y)
+{
+	n <- length(x)
+	bhat <- cov(y, x) / var(x)
[…]

### 2016-12-19 scripts/mr_directionality_multiple.R
updated

@@ -94,36 +94,23 @@ get_regressions <- function(X, y)
 	return(mat)
 }
 
-get_effects_from_system <- function(dat)
+get_effects_from_system <- function(dat, reverse=FALSE)
 {
 	inst_names <- grep("Z.", names(dat))
-	return(list(
+	l <- list(
 		A = get_regressions(dat[,inst_names], dat$A),
 		B = get_regressions(dat[,inst_names], dat$B),
 		Ap = get_regressions(dat[,inst_names], dat$Ap),
 		Bp = get_regressions(dat[,inst_names], dat$Bp)
-	))
+	)
+	if(reverse)
+	{
+		names(l) <- gsub("A", "C", names(l))
+		names(l) <- gsub("B", "A", names(l))
+		names(l) <- gsub("C", "B", names(l))
+	}
+	return(l)
 }
-
-dat <- make_system(2000, 100, sqrt(0.2), sqrt(0.1), sqrt(0.03), sqrt(0.5), sqrt(0.5), sqrt(0.5))
-out <- get_effects_from_system(dat)
-
-plot(out$B$b ~ out$A$b)
-
-
-dat <- make_system(20000, 40, sqrt(0.4), sqrt(0.2), sqrt(0.001), -0.05, NULL, sqrt(0.5), sqrt(0.5))
-out <- get_effects_from_system(dat)
-
-temp1 <- mr_ivw(out$A$b, out$B$b, out$A$se, out$B$se)
-temp2 <- mr_egger_regression(out$A$b, out$B$b, out$A$se, out$B$se)
-temp3 <- mr_weighted_median(out$A$b, out$B$b, out$A$se, out$B$se)
-
-plot(temp2$dat$b_out ~ temp2$dat$b_exp)
-abline(a=0,b=temp1$b)
-abline(a=temp2$b_i, b=temp2$b, col="red")
-temp1$b^2
-temp2$b^2
-
 
 make_adjusted_plot <- function(out)
 {	
@@ -159,23 +146,40 @@ make_adjusted_plot <- function(out)
 
 	s <- data.frame(intercept = c(0, temp2$b_i, 0),
 		slope = c(temp1$b, temp2$b, temp3$b),
+		pval = c(temp1$pval, temp2$pval, temp3$pval),
 		method=c("IVW", "MR Egger", "Weighted median")
 	)
 
 
-	ggplot(d, aes(x=b_exp, y=value)) +
[…]

### 2016-12-19 scripts/mr_directionality_multiple.R
updated

@@ -78,19 +78,19 @@ get_regression <- function(x, y)
 	Fval <- (SSF) / (SSR/(n-2))
 	pval <- pf(Fval, 1, n-2, lowe=F)
 	se <- sqrt(sum(residuals^2) / ((n - 2) * sum((x-mean(x))^2)))
-	return(c(ahat, bhat, se, pval))
+	return(c(ahat, bhat, se, pval, n))
 }
 
 get_regressions <- function(X, y)
 {
 	nsnp <- ncol(X)
-	mat <- matrix(0, nsnp, 4)
+	mat <- matrix(0, nsnp, 5)
 	for(i in 1:nsnp)
 	{
 		mat[i,] <- get_regression(X[,i], y)
 	}
 	mat <- as.data.frame(mat)
-	names(mat) <- c("intercept", "b", "se", "pval")
+	names(mat) <- c("intercept", "b", "se", "pval", "n")
 	return(mat)
 }
 
@@ -161,25 +161,76 @@ make_adjusted_plot <- function(out)
 }
 
 
+get_predicted_effects <- function(out)
+{
+	temp1 <- mr_ivw(out$A$b, out$B$b, out$A$se, out$B$se)
+	temp2 <- mr_egger_regression(out$A$b, out$B$b, out$A$se, out$B$se)
+	temp3 <- mr_weighted_median(out$A$b, out$B$b, out$A$se, out$B$se, default_parameters())
+
+	d <- temp2$dat
+
+	pred_ivw <- d$b_exp * temp1$b
+	pred_egg <- d$b_exp * temp2$b + temp2$b_i
+	pred_wme <- d$b_exp * temp3$b
+
+	r_a <- get_r_from_pn(out$A$pval, out$A$n)
+	r <- get_r_from_pn(out$B$pval, out$B$n)
+	coef <- abs(r / out$B$b)
+	r_ivw <- pred_ivw * coef
+	r_egg <- pred_egg * coef
+	r_wme <- pred_wme * coef
+
+	return(data.frame(r_a^2, r^2, r_ivw^2, r_egg^2, r_wme^2))
+}
+
+
+get_r_from_pn <- function(p, n)
+{
+	get_p_from_r2n <- function(r2, n)
+	{
+		fval <- r2 * (n-2) / (1 - r2)
+		pval <- pf(fval, 1, n-1, low=FALSE)
+		return(pval)
+	}
+	optim.get_p_from_rn <- function(x, sample_size, pvalue)
+	{
[…]

### 2016-12-21 scripts/mr_directionality_analysis.R
up;dated|

@@ -675,8 +675,10 @@ shakhbazov$dir <- as.character(shakhbazov$dir)
 shakhbazov$dir[shakhbazov$dir=="TRUE"] <- "Methylation causes Expression"
 shakhbazov$dir[shakhbazov$dir=="FALSE"] <- "Expression causes Methylation"
 
-shakhtest1 <- fisher.test(table(sign(shakhbazov$mr_eff), shakhbazov$dir))
-shakhtest2 <- fisher.test(table(sign(shakhbazov$mr_eff[shakhbazov$dir_p < 0.05]), shakhbazov$dir[shakhbazov$dir_p < 0.05]))
+real_index <- shakhbazov$rxx_o == 1 & shakhbazov$ryy_o == 1
+table(real_index)
+shakhtest1 <- fisher.test(table(sign(shakhbazov$mr_eff[real_index]), shakhbazov$dir[real_index]))
+shakhtest2 <- fisher.test(table(sign(shakhbazov$mr_eff[real_index & shakhbazov$dir_p < 0.05]), shakhbazov$dir[real_index & shakhbazov$dir_p < 0.05]))
 
 shakhsummary <- dplyr::group_by(shakhbazov, rxx_o, ryy_o, dir) %>%
 	dplyr::summarise(
@@ -688,8 +690,18 @@ shakhsummary <- dplyr::group_by(shakhbazov, rxx_o, ryy_o, dir) %>%
 	) %>% as.data.frame()
 
 
-shakhtest3 <- binom.test(sum(shakhbazov$dir == "Methylation causes Expression" & shakhbazov$dir_p < 0.05), sum(shakhbazov$dir_p < 0.05), 0.5)
-shakhtest4 <- binom.test(sum(shakhbazov$dir == "Methylation causes Expression"), nrow(shakhbazov), 0.5)
+shakhtest3 <- binom.test(sum(shakhbazov$dir == "Methylation causes Expression" & shakhbazov$dir_p < 0.05 & real_index), sum(shakhbazov$dir_p < 0.05 & real_index), 0.5)
+shakhtest4 <- binom.test(sum(shakhbazov$dir == "Methylation causes Expression" & real_index), sum(real_index), 0.5)
+library(dplyr)
+a <- subset(shakhbazov, dir_p < 0.05 & real_index) %>%
+	group_by(dir) %>%
[…]

### 2016-12-23 scripts/mr_directionality_analysis.R
upda

@@ -7,6 +7,131 @@ suppressPackageStartupMessages(library(psych))
 suppressPackageStartupMessages(library(lattice))
 suppressPackageStartupMessages(library(latex2exp))
 suppressPackageStartupMessages(library(knitr))
+suppressPackageStartupMessages(library(gridExtra))
+
+
+get_p_from_r2n <- function(r2, n)
+{
+	fval <- r2 * (n-2) / (1 - r2)
+	pval <- pf(fval, 1, n-1, low=FALSE)
+	return(pval)
+}
+
+get_r_from_pn <- function(p, n)
+{
+	optim.get_p_from_rn <- function(x, sample_size, pvalue)
+	{
+		abs(-log10(get_p_from_r2n(x, sample_size)) - -log10(pvalue))
+	}
+
+	if(length(p) > 1 & length(n) == 1)
+	{
+		message("Assuming n the same for all p values")
+		n <- rep(n, length(p))
+	}
+
+	Fval <- qf(p, 1, n-1, low=FALSE)
+	R2 <- Fval / (n - 2 + Fval)
+	index <- !is.finite(Fval)
+	if(any(index))
+	{
+		index <- which(index)
+		for(i in 1:length(index))
+		{
+			R2[index[i]] <- optim(0.001, optim.get_p_from_rn, sample_size=n[index[i]], pvalue=p[index[i]])$par
+		}
+	}
+	return(sqrt(R2))
+}
+
+
+steiger_sensitivity <- function(rgx_o, rgy_o)
+{
+	if(rgy_o > rgx_o)
+	{
+		a <- rgy_o
+		b <- rgx_o
+	} else {
+		a <- rgx_o
+		b <- rgy_o
+	}
+
+	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=50), ryy_o=seq(rgy_o,1,length.out=50), type=c("A","B"))
+	d$rgy <- d$ryy_o / rgy_o
+	d$rgx <- d$rxx_o / rgx_o
+	d$z <- d$rgy - d$rgx
+	d$z[d$type=="A"] <- 0
+	temp <- wireframe(
+		z ~ rxx_o * ryy_o, 
+		groups=type, 
+		data=d, 
+		scales=list(arrows=FALSE), 
[…]

### 2016-12-28 docs/manuscript.Rmd
up;dated|
updated

@@
 ---
-title: "Determining the existence and direction of causality in the face of measurement error"
+title: "Inferring the existence and direction of causal associations in the face of measurement error"
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
   pdf_document:
+    fig_caption: no
     keep_tex: true
@@
 
+```{r load_up }
+
+```
+
+```{r read_data }
+
+```
+
+```{r shakhbazov }
+
+```
+
+```{r shakhbazov_tests }
+
+```
+
+```{r steiger_sensitivity }
+
+```
+
 ```{r tablelist, cache=FALSE}
@@
 fig_nums(
+	"dag_figure",
+	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown.
+If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables.
+a) Methylation causes gene expression.
+The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels.
+The right figure shows the directed acyclic graph that represents this model.
+Faded symbols represent the measured values whereas solid symbols represent the true values.
+b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
+	display=FALSE
+)
+
+fig_nums(
 	"cit_measurement_error_figure", 
@@
 fig_nums(
-	"d_relationship_figure", 
-	caption="Plots depicting the parameter space in which the function $d = cor(x, x_o) - cor(x,y)cor(y, y_o)$ is negative.
-When $d$ is negative the Steiger test is liable to infer the wrong direction of causality.
-$d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_o)$) on the x-axis.
-Higher measurement precision of $y$ ($cor(y,y_o)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$.
-Shaded regions show the parameter space where $d$ is negative.
-The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
+	"steiger_sensitivity_plot", 
+	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface.
+When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction.
+Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y.
+A measure of sensitivity, therefore, is the volume of the positive space bound by the blue surface as a proportion of the total space bound by the blue surface.
+Sensitivity of 0 denotes no measurement error values that could lead to inferrence of incorrect direction, whereas the maximum value of 0.5 denotes measurement error equal chance of either direction across all measurement error ranges.
+In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the sensitivity is 0.18.
+b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger sensitivity values (y-axis) for different values of $\\rho_{g,x}^2$.
+Inferring the direction of causality is more susceptible to measurement error when there are stronger causal associations between X and Y.",
 	display=FALSE
@@
 	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
 	display=FALSE
@@
 For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.",
+	display=FALSE
+)
+
+fig_nums(
+	"shakhplot",
+	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them.
+The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels.
+The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes).
+If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor."),
 	display=FALSE
@@
 
-```{r load_up }
-
-```
-```{r read_data }
-
-```
-
-```{r shakhbazov }
-
-```
-
 ### Abstract
@@
 A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure.
-We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
+We show that in the face of measurement error this can lead to erroneous inference of both existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
 Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible.
-We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
-Using this method we demonstrate that for correlations between DNA methylation and gene expression levels, DNA methylation is likely to be the causal factor.
-We show that using MR to infer causal directions is straightforward, but because measurement error is ubiquitous in observational data, mediation-based causal inference in particular should be treated with caution.
+We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and introduce a simple extension to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
+This extension is less susceptible to measurement error than mediation-based analyses.
+Using these methods we demonstrate empirically that the inferred direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms, demanding caution when making such claims about causality.
+
+The last decade has seem a rapid acceleration in technological advances that enable us to assess the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits.
+This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
+For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.
 
@@
 
-The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits.
-This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
-For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.
-Despite the ubiquity of the question, commonly used observational study data are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004].
+Widely available observational study data are notoriously susceptible to pitfalls in making causal inference, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004; @Phillips1991; @DaveySmith2002].
+In response to this, genetic instrumentation has emerged as a valuable technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies, it is now a prominent tool that is applied in several different guises, and has been applied many forms of high throughput 'omic variables.
+However, an often neglected pitfall to these approaches is the influence of non-differential measurement error on the reliability of these methods.
 
-An established statistical design that can resolve this problem is randomisation.
-Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control.
+Non-differential measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value.
+Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
+Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
+Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
+
+Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
+Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
+The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
+
+Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.
+An established study design that can provide information about causality is randomisation.
+Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control.
 The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship.
-This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
+This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
 The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004].
-Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation and Mendelian randomisation (MR).
+Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).
 
-Mediation is a commonly used statistical method for making causal inference in observational data, and it exists in many forms, from regression based systems  [@Baron1986] to structural equation modeling (SEM) [@Aten2008].
-In its simplest form mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable.
-If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.
-
-This simple form, however, is prone to being unreliable.
-Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator [@Fewell2007].
-Introducing pseudo-randomisation by using a genetic instrument as the independent variable can be used to improve on these issues.
-Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome.
-Therefore the influence of the instrument on the outcome conditional on the exposure should be zero.
+Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome.
+If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the implied influence of the SNP on the outcome will disappear when conditioning on the exposure.
+Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome.
 This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015].
@@
 
-MR employs the same data under but uses a different approach in how the instrumental variable is employed.
-Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome, avoiding the problems of confounding or reverse causation by testing for association between the outcome and the SNP.
-The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
+MR can operate on the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner.
+Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure.
+The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
 
-By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages.
+By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages.
 The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
@@
 Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.
-This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?).
+This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?
+`r cfig("dag_figure")`).
 By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
 Such a situation can arise in many scenarios.
-Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
+Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
 For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP).
 It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI.
-Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016]?
+Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?
 
@@
 
-Measurement (or observational) error is the difference between the measured value of a quantity and its true value.
-Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
-Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
-Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
-
-Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
-A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
-Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
-The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
-
-Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference.
 Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
@@
 
-We used the Steiger test to test if DNA methylation levels are more likely to causally influence gene expression levels or vice versa.
-To do this we obtained a list of `r nrow(qtl)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
+We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels.
+To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
 These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
@@
 
-## Tables
+## Figures
 
-```{r shakhtab }
+![](../images/dag-01.png)
 
-```
-
-`r tab_nums("shakhtab")`
+`r fig_nums("dag_figure")`
 
 \newpage
-
-## Figures
 
@@
 
-```{r d_relationship_figure, fig.width=10, fig.height=6 }
+```{r steiger_sensitivity_plot, fig.width=12, fig.height=6 }
 
@@
 
-`r fig_nums("d_relationship_figure")`
+`r fig_nums("steiger_sensitivity_plot")`
 
@@
 `r fig_nums("cit_mr_comparison_figure")`
+
+\newpage
+
+```{r shakhplot}
+
+```
+
+`r fig_nums("shakhplot")`
 
@@
 Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
-In the presence of measurement error in $x$ and $y$, however, Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
 
-$$
-\begin{aligned}
-\rho_{g, x_o} & > \rho_{g, y_o} \\
-\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
-\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
-\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
-\end{aligned}
-$$
+In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$.
+In order to assess how sensitive the Steiger test is to measurement error, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.
 
-\newpage
+For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$.
+For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.
 
-## Appendix 3
-
-The Steiger test is susceptible to inferring the wrong direction of causality under some measurement error parameters.
-For an observed association between the instrument $g$ and the measured exposure $x_o$, $\rho_{g,x_o}$, the association between $g$ and the true exposure (without measurement error) $x$, $\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}$ where the bounds $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$ are the theoretical limits of the correlation between true and measured $x$.
-The same logic applies to the association between the instrument and the outcome variable.
-
-For the values $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, the possible true $\rho_{g,x}$ and $\rho_{g,y}$ enumerated for all possible values of measurement error can be visualised as so:
-
-```{r, fig.width=6, fig.height=6}
-
-sensitivity_example <- steiger_sensitivity(0.3, 0.05)
-sensitivity_example$pl
-
-```
-
-Here, where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality.
-To evaluate the sensitivity, $S$ of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,
+Where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality.
+To evaluate the sensitivity, $S$, of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,
 
@@
 
-$V_{z}$, the total volume of the function, can be obtained analytically by solving the double integral:
+$V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:
 
@@
 
-$S$ can take values from 0 to 0.5, where low values suggest low sensitivity of the Steiger test to measurement error.
-For the values used to plot the function above, $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, we obtain $S = `r round(sensitivity_example$sensitivity,3)`$.
-Hence, the majority of the parameter space of measurement error will not change the inferred causal direction of $x \rightarrow y$ for these values.
+The sensitivity, $S$, can take values from 0 to 0.5, where low values suggest there are
+
+\newpage
 

### 2016-12-28 docs/manuscript.tex
updated

@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
   \postdate{\par}
-  \date{14 December 2016}
+  \date{28 December 2016}
 
@@
 Two sample MR\textsuperscript{19} was performed using the summary
-statistics presented in\textsuperscript{18} for each of 458 associations
-between DNA methylation and gene expression levels. Because only summary
-statistics were available (effect, standard error, effect allele, sample
-size, p-values) for the instrumental SNP on the methylation and gene
-expression levels, the Steiger test of two independent correlations was
-used to infer the direction of causality for each of the associations.
-The Wald ratio test was then used to estimate the causal effect size for
-the estimated direction for each association.
+statistics presented in\textsuperscript{18} for each of 4122
+associations between DNA methylation and gene expression levels. Because
+only summary statistics were available (effect, standard error, effect
+allele, sample size, p-values) for the instrumental SNP on the
+methylation and gene expression levels, the Steiger test of two
+independent correlations was used to infer the direction of causality
+for each of the associations. The Wald ratio test was then used to
+estimate the causal effect size for the estimated direction for each
+association.
 
@@
 We performed simulations to verify that this problem does arise using
-the CIT method. Figure 1 shows that when there is no measurement error
+the CIT method. Figure 2 shows that when there is no measurement error
 in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
@@
 detecting a causal association between simulated variables under
-different levels of imprecision simulated in the exposure. Figure 3
+different levels of imprecision simulated in the exposure. Figure 4
 shows the true positive rates between the CIT and MR for detecting a
@@
 direction because the value of \(d\) is not restricted to being greater
-than 0. However Figure 2 shows that when there is no measurement error
+than 0. However Figure 7 shows that when there is no measurement error
 in \(y\), the Steiger test is unlikely to infer the wrong direction of
@@
 rate at which a strong causal relationship is obtained where the
-reported direction of causality is incorrect. Figure 4 shows that when
+reported direction of causality is incorrect. Figure 5 shows that when
 \(d < 0\) the MR analysis is indeed liable to infer the wrong direction
@@
 
-We used the Steiger test to test if DNA methylation levels are more
-likely to causally influence gene expression levels or vice versa. To do
-this we obtained a list of 458 gene expression - DNA methylation
-associations as reported in Shakhbazov et al\textsuperscript{18}. These
-were filtered to be located on the same chromosome, have robost
-correlations after correcting for multiple testing, and to share a SNP
-that had a robust cis-acting effect on both the DNA methylation probe
-and the gene expression probe. Using the reported summary data we used
-the Steiger test of two independent correlations to identify which
-direction of causality was most likely for each association. We found
-that the causal direction commonly goes in both directions (Table 1) but
-methylation levels were more likely to influence expression levels than
-vice versa (probability of expression causing methylation = 0.0019317
-(95\% CI 0.3455756 - 0.4653544) amongst 272 associations with
+We used the Steiger test to infer the direction of causality between DNA
+methylation levels and gene expression levels. To do this we obtained a
+list of 458 gene expression - DNA methylation associations as reported
+in Shakhbazov et al\textsuperscript{18}. These were filtered to be
+located on the same chromosome, have robost correlations after
+correcting for multiple testing, and to share a SNP that had a robust
+cis-acting effect on both the DNA methylation probe and the gene
+expression probe. Using the reported summary data we used the Steiger
+test of two independent correlations to identify which direction of
+causality was most likely for each association. We found that the causal
+direction commonly goes in both directions (Table 1) but methylation
+levels were more likely to influence expression levels than vice versa
+(probability of expression causing methylation = 0.0163873 (95\% CI
+0.3738727 - 0.4874269) amongst NA associations with
 \(p_{Steiger} < 0.05\)).
@@
 expression the causal effect was more likely to be negative than for the
-associations where gene expression caused methylation (OR = 0.4615363
-(95\% CI 0.259998 - 0.8052745)), suggesting that reducing gene
+associations where gene expression caused methylation (OR = 0.5326663
+(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
 expression levels at a controlling CpG typically leads to increased gene
@@
 
-\subsection{Tables}\label{tables}
-
-\begin{longtable}[]{@{}lrrrrr@{}}
-\toprule
-Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
-\(cor(\rho_{MR}, \rho_{P})\) & Mean sensitivity\tabularnewline
-\midrule
-\endhead
-Expression causes Methylation & 246 & 162 & 0.3902439 & 0.5629453 &
-0.2031556\tabularnewline
-Methylation causes Expression & 212 & 110 & 0.2877358 & 0.4384660 &
-0.2365816\tabularnewline
-\bottomrule
-\end{longtable}
-
-Table 1: Summary of MR analysis on gene expression and DNA methylation
-associations. `Count (\(p_{Steiger} < 0.05\))' denotes the number of
-associations that have a p-value \textless{} 0.05 from the Steiger test.
-P(+ve effect) denotes the proportion of associations that were positive.
-\(cor(\rho_{MR}, \rho_{P})\) is the correlation of the MR estimate
-obtained in this analysis and the Pearson correlation between reported
-in Shakhbazov et al 2016.
+\subsection{Figures}\label{figures}
+
+\includegraphics{../images/dag-01.png}
+
+Figure 1: Gene expression levels (blue blocks) and DNA methylation
+levels (green triangles) may be correlated but the causal direction is
+unknown. If a SNP (yellow circle) is associated with both DNA
+methylation and gene expression levels then it can be used as an
+instrument, but there are two competing models for these variables. a)
+Methylation causes gene expression. The left figure shows that the SNP
+influences methylation levels that in turn influence gene expression
+levels. The right figure shows the directed acyclic graph that
+represents this model. Faded symbols represent the measured values
+whereas solid symbols represent the true values. b) The same as in A,
+except the causal direction is from gene expression to DNA methylation.
 
@@
 
-\subsection{Figures}\label{figures}
-
 \includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}
 
-Figure 1: The CIT was performed on simulated variables where the
+Figure 2: The CIT was performed on simulated variables where the
 exposure influenced the outcome and the exposure was instrumented by a
@@
 
-\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}
-
-Figure 2: Plots depicting the parameter space in which the function
-\(d = cor(x, x_o) - cor(x,y)cor(y, y_o)\) is negative. When \(d\) is
-negative the Steiger test is liable to infer the wrong direction of
-causality. \(d\) is plotted on the y axis, with measurement precision of
-\(x\) (\(cor(x,x_o)\)) on the x-axis. Higher measurement precision of
-\(y\) (\(cor(y,y_o)\)) is depicted by lighter coloured lines, and
-columns of boxes are for different true values of \(cor(x,y)\). Shaded
-regions show the parameter space where \(d\) is negative. The graph
-shows that for the majority of the parameter space of the function,
-\(d\) is positive, especially where causal relationships are relatively
-weak.
+\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
+
+Figure 3: a) We can predict the values the Steiger test would take
+(z-axis) for different potential values of measurement error (x and y
+axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
+as denoted by range of values where the blue surface is above the red
+plane, those values of measurement error lead to our observed Steiger
+test inferring the wrong causal direction. Where the blue surface lies
+below the red plane, these measurement error values support the inferred
+causal direction of X to Y. A measure of sensitivity, therefore, is the
+volume of the positive space bound by the blue surface as a proportion
+of the total space bound by the blue surface. Sensitivity of 0 denotes
+no measurement error values that could lead to inferrence of incorrect
+direction, whereas the maximum value of 0.5 denotes measurement error
+equal chance of either direction across all measurement error ranges. In
+this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the
+sensitivity is 0.18. b) The influence of \(\rho_{x,y}\) (x-axis) on the
+Steiger sensitivity values (y-axis) for different values of
+\(\rho_{g,x}^2\). Inferring the direction of causality is more
+susceptible to measurement error when there are stronger causal
+associations between X and Y.
 
@@
 
-Figure 3: Outcomes were simulated to be causally influenced by exposures
+Figure 4: Outcomes were simulated to be causally influenced by exposures
 with varying degrees of measurement imprecision applied to the exposure
 variable (x axis). True positive rates (y axis) for MR and CIT were
-compared with for varying levels of measurement imprecision in the
-outcome variable (rows of boxes) and sample sizes (columns of boxes).
+compared for varying levels of measurement imprecision in the outcome
+variable (rows of boxes) and sample sizes (columns of boxes).
 
@@
 
-Figure 4: Outcomes were simulated to be causally influenced by
+Figure 5: Outcomes were simulated to be causally influenced by
 exposures, with varying degrees of measurement error applied to both.
@@
 evidence for causality.
+
+\newpage
+
+\includegraphics{manuscript_files/figure-latex/shakhplot-1.pdf}
+
+Figure 6: Using 458 putative associations between DNA methylation and
+gene expression we used the Steiger test to infer the direction of
+causality between them. The rightmost bar shows the proportion of
+associations for each of the two possible causal directions (colour key)
+assuming no measurement error in either gene expression or DNA
+methylation levels. The proportions change when we assume different
+levels of measurement error in gene expression levels (x-axis) or DNA
+methylation levels (columns of boxes). If there is systematically higher
+measurement error in one platform it will appear to be less likely to be
+the causal factor.
 
@@
 causal association between \(x\) and \(y\) (which is likely to be less
-than 1). In the presence of measurement error in \(x\) and \(y\),
-however, Steiger test will instead be assessing the inequality
-\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
-
-\[
-\begin{aligned}
-\rho_{g, x_o} & > \rho_{g, y_o} \\
-\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
-\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
-\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
-\end{aligned}
-\]
-
-\newpage
-
-\subsection{Appendix 3}\label{appendix-3}
-
-The Steiger test is susceptible to inferring the wrong direction of
-causality under some measurement error parameters. For an observed
-association between the instrument \(g\) and the measured exposure
-\(x_o\), \(\rho_{g,x_o}\), the association between \(g\) and the true
-exposure (without measurement error) \(x\),
-\(\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}\) where the bounds
-\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\) are the theoretical limits of
-the correlation between true and measured \(x\). The same logic applies
-to the association between the instrument and the outcome variable.
-
-For the values \(\rho_{g,x_o} = 0.3\) and \(\rho_{g,y_o} = 0.05\), the
-possible true \(\rho_{g,x}\) and \(\rho_{g,y}\) enumerated for all
-possible values of measurement error can be visualised as so:
-
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}
-
-Here, where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error
-range in \(x_o\) and \(y_o\) will not change the qualitative inference
-of the direction of causality. To evaluate the sensitivity, \(S\) of the
+than 1).
+
+In the presence of measurement error in \(x\) and \(y\), however, the
+Steiger test will instead be assessing the inequality
+\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how sensitive the
+Steiger test is to measurement error, we can evaluate the range of
+potential values of measurement error in \(x\) and \(y\) over which the
+empirical result from the Steiger test would return the wrong value.
+
+For different values of \(\rho_{x,x_o}\),
+\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
+\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
+\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
+and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).
+
+Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
+\(x_o\) and \(y_o\) will not change the qualitative inference of the
+direction of causality. To evaluate the sensitivity, \(S\), of the
 Steiger test to measurement error, the objective is to calculate the
@@
 
-\(V_{z}\), the total volume of the function, can be obtained
-analytically by solving the double integral:
+\(V_{z}\), the total volume of the function (Figure 3), can be obtained
+analytically by solving:
 
@@
 
-\(S\) can take values from 0 to 0.5, where low values suggest low
-sensitivity of the Steiger test to measurement error. For the values
-used to plot the function above, \(\rho_{g,x_o} = 0.3\) and
-\(\rho_{g,y_o} = 0.05\), we obtain \(S = 0.032\). Hence, the majority of
-the parameter space of measurement error will not change the inferred
-causal direction of \(x \rightarrow y\) for these values.
+The sensitivity, \(S\), can take values from 0 to 0.5, where low values
+suggest there are
+
+\newpage
 

### 2016-12-28 scripts/mr_directionality_analysis.R

@@ -8,6 +8,7 @@ suppressPackageStartupMessages(library(lattice))
 suppressPackageStartupMessages(library(latex2exp))
 suppressPackageStartupMessages(library(knitr))
 suppressPackageStartupMessages(library(gridExtra))
+suppressPackageStartupMessages(library(grid))
 
 
 get_p_from_r2n <- function(r2, n)
@@ -45,7 +46,7 @@ get_r_from_pn <- function(p, n)
 }
 
 
-steiger_sensitivity <- function(rgx_o, rgy_o)
+steiger_sensitivity <- function(rgx_o, rgy_o, ...)
 {
 	if(rgy_o > rgx_o)
 	{
@@ -57,8 +58,8 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 	}
 
 	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=50), ryy_o=seq(rgy_o,1,length.out=50), type=c("A","B"))
-	d$rgy <- d$ryy_o / rgy_o
-	d$rgx <- d$rxx_o / rgx_o
+	d$rgy <- rgy_o / d$ryy_o
+	d$rgx <- rgx_o / d$rxx_o
 	d$z <- d$rgy - d$rgx
 	d$z[d$type=="A"] <- 0
 	temp <- wireframe(
@@ -68,9 +69,11 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 		scales=list(arrows=FALSE), 
 		col.groups = colorRampPalette(c("red", "blue"))(2), 
 		drape=FALSE, 
-		xlab=expression(rho[xx[o]]), 
-		ylab=expression(rho[yy[o]]),
-		zlab=expression(rho[gy]-rho[gx])
+		ylab=expression(rho[xx[o]]), 
+		xlab=expression(rho[yy[o]]),
+		zlab=expression(rho[gy]-rho[gx]),
+		par.settings = list(axis.line=list(col="transparent")),
+		...
 	)
 
 	vz <- a * log(a) - b * log(b) + a*b*(log(b)-log(a))
@@ -87,7 +90,7 @@ steiger_sensitivity <- function(rgx_o, rgy_o)
 }
 
 
-mr_steiger <- function(p_exp, p_out, n_exp, n_out, r_xxo = 1, r_yyo=1) 
+mr_steiger <- function(p_exp, p_out, n_exp, n_out, r_xxo = 1, r_yyo=1, ...) 
 {
[…]

### 2016-12-29 docs/manuscript.Rmd
updated

@@
 
-The last decade has seem a rapid acceleration in technological advances that enable us to assess the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits.
-This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data.
-For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.
-
 ## Introduction
 
-Widely available observational study data are notoriously susceptible to pitfalls in making causal inference, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004; @Phillips1991; @DaveySmith2002].
-In response to this, genetic instrumentation has emerged as a valuable technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies, it is now a prominent tool that is applied in several different guises, and has been applied many forms of high throughput 'omic variables.
-However, an often neglected pitfall to these approaches is the influence of non-differential measurement error on the reliability of these methods.
+Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004].
+In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015].
+However, potential pitfalls still persist and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
 
@@
 Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
-The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
+The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
 
@@
 Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome.
-If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the implied influence of the SNP on the outcome will disappear when conditioning on the exposure.
+If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure.
 Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome.
-This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015].
-They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].
+This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], a structural equation modelling (SEM) implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015].
+They have been employed by a number of recent publications that make causal inferences in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].
 
@@
 
-By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages.
+By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference.
 The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
@@
 Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
-This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes.
-Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure and outcome variables.
-Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
+This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated.
+Third, MR is substantially more robust to there being measurement error in the exposure and outcome variables.
+Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
 We argue that this is a serious limitation.
@@
 Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
-We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood.
-This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
+We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and to evaluate the sensitivity of the result to measurement error.
+This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
+Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
 

### 2016-12-29 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Determining the existence and direction of causality in the face of measurement error},
+            pdftitle={Inferring the existence and direction of causal associations in the face of measurement error},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Determining the existence and direction of causality in the face of
-measurement error}
+  \title{Inferring the existence and direction of causal associations in the face
+of measurement error}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \postdate{\par}
-  \date{28 December 2016}
+  \date{29 December 2016}
 
@@
 after adjusting for the exposure. We show that in the face of
-measurement error this can lead to erroneous results, and that
-increasing sample size, rather than reducing bias, can often have the
-effect of increasing certainty in the wrong answer. Commonly used tools
-that are predicated on this methodology such as the causal inference
-test (CIT) are shown to be susceptible. We demonstrate that Mendelian
-randomization (MR) is a method for causal inference that is robust to
-measurement error, and that a simple extension to MR can provide a
-method to estimate the direction of causality between correlated
-variables even when the biology of the instrument is unknown. Using this
-method we demonstrate that for correlations between DNA methylation and
-gene expression levels, DNA methylation is likely to be the causal
-factor. We show that using MR to infer causal directions is
-straightforward, but because measurement error is ubiquitous in
-observational data, mediation-based causal inference in particular
-should be treated with caution.
-
-\subsection{Introduction}\label{introduction}
+measurement error this can lead to erroneous inference of both existence
+of causality and the causal direction, and that increasing sample size,
+rather than reducing bias, can often have the effect of increasing
+certainty in the wrong answer. Commonly used tools that are predicated
+on this methodology such as the causal inference test (CIT) are shown to
+be susceptible. We demonstrate that Mendelian randomization (MR) is a
+method for causal inference that is robust to measurement error, and
+introduce a simple extension to estimate the direction of causality
+between correlated variables even when the biology of the instrument is
+unknown. This extension is less susceptible to measurement error than
+mediation-based analyses. Using these methods we demonstrate empirically
+that the inferred direction of causality between correleted DNA
+methylation and gene expression levels is highly susceptible to
+differences in measurement error between the two assaying platforms,
+demanding caution when making such claims about causality.
 
 The last decade has seem a rapid acceleration in technological advances
-that enable us to measure the human phenome in ever increasing detail,
+that enable us to assess the human phenome in ever increasing detail,
 furnishing biologists with the tantalising opportunity to map the
@@
 understanding the causal nature of any putative observational
-associations. Despite the ubiquity of the question, commonly used
-observational study data are notoriously susceptible to pitfalls in
-making causal inference, with simple regression-based techniques unable
-to distinguish a true causal association from reverse causation or
-confounding\textsuperscript{3}.
-
-An established statistical design that can resolve this problem is
-randomisation. Given the hypothesis that an exposure is causally related
-to an outcome, randomisation can be employed to assess the causal nature
-of the association by randomly splitting the sample into two groups,
-subjecting one group to the exposure and leaving the other as a control.
-The association between the exposure and the outcome in this setting
-provides a robust estimate of the causal relationship. This is the
-theoretical basis behind randomised control trials, but in practice
-randomisation is often impossible to implement in an experimental
-context due to cost, scale or inability to manipulate the exposure. The
-principle, however, can be employed in extant observational data through
-the use of genetic variants associated with the exposure (instruments),
-where the inheritance of an allele serves as a random lifetime
-allocation of differential exposure levels\textsuperscript{3}. Two
-statistical approaches to exploiting the properties of genetic
-instruments are widely used: mediation and Mendelian randomisation (MR).
-
-Mediation is a commonly used statistical method for making causal
-inference in observational data, and it exists in many forms, from
-regression based systems\textsuperscript{4} to structural equation
-modeling (SEM)\textsuperscript{5}. In its simplest form mediation
-analysis seeks to explain the mechanism that underlies an observed
-relationship between an independent and a dependent variable by
-introducing a third, `mediator' variable. If the association between the
-independent and dependent variables disappears when conditioning on the
-mediator variable then this provides evidence that the independent
-variable influences the dependent variable through the mediator.
-
-This simple form, however, is prone to being unreliable. Alternative
-models of reverse causation or confounding are often indistinguishable
-from the hypothesised model of the independent variable causing the
-dependent variable via the mediator\textsuperscript{6}. Introducing
-pseudo-randomisation by using a genetic instrument as the independent
-variable can be used to improve on these issues. Supposing an exposure
-has a causal effect on an outcome, then an `instrument' that causes the
-exposure (a single nucleotide polymorophism (SNP), for example) should
-also influence the outcome. Therefore the influence of the instrument on
-the outcome conditional on the exposure should be zero. This forms the
-basis of a number of methods such as genetical
-genomics\textsuperscript{7}, the regression-based causal inference test
-(CIT)\textsuperscript{8,9}, an SEM implementation in the NEO
-software\textsuperscript{5}, and various other methods including
-Bayesian approaches\textsuperscript{10}. They have been employed by a
-number of recent publications that make causal inferences, often in
-large scale `omics datasets\textsuperscript{10--15}.
-
-MR employs the same data under but uses a different approach in how the
-instrumental variable is employed. Here an instrumental variable
-(typically a SNP) that has a robost, direct association with the
-exposure is used as a surrogate for the exposure in an association test
-with the outcome, avoiding the problems of confounding or reverse
-causation by testing for association between the outcome and the SNP.
-The causal association between the exposure and the outcome can be
-estimated by scaling the association between the SNP and the outcome by
-the association between the SNP and the exposure.
-
-By utilising genetic instruments in different ways, mediation and MR
-models have properties that confer some advantages and some
-disadvantages. The main advantage of mediation over MR is that the true
-underlying biology of the genetic instrument doesn't necessarily need to
-be known as long as it associates with the putative exposure. In the CIT
-framework (described fully in the Methods) for example, the test
-statistic is different if you test for the exposure causing the outcome
-or the outcome causing the exposure, allowing the researcher to infer
-the direction of causality between two variables by performing the test
-in both directions and choosing the model with the strongest evidence.
-
-Such is not the case for MR, where in order to infer the direction of
-causality the instrument must be known to influence the exposure
-directly, associating with the outcome only through the exposure. This
-requirement can be problematic because if there exists a putative
-association between two variables, with the instrumental SNP being
-robustly associated with each, it can be difficult to determine which of
-the two variables is subject to the direct effect of the SNP (i.e., for
-which of the two variables is the SNP a valid instrument?). By
-definition, we expect that if the association is causal then a SNP for
-the exposure will be associated with the outcome, so if the researcher
-erroneously uses the SNP as an instrument for the outcome then they are
-likely to see an apparently robust causal association of outcome on
-exposure. Such a situation can arise in many scenarios. Genome wide
-association studies (GWASs) that identify genetic associations for
-complex traits are, by design, hypothesis free and agnostic of genomic
-function, and it often takes years of follow up studies to understand
-the biological nature of a putative GWAS hit\textsuperscript{16}. For
-example, the same FTO variant that is associated with body mass index
-(BMI) is also associated with C-reactive protein (CRP). It is known that
-this is because BMI causes CRP\textsuperscript{17}, but supposing we
-only had \emph{a priori} knowledge that FTO associated with CRP, and we
-used it to instrument CRP in an MR analysis of CRP against BMI, we would
-mistakenly infer that CRP causes BMI. Another situation is where the
-causal directions between 'omic measures need to be determined, for
-example if a DNA methylation probe is associated with expression of an
-adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level\textsuperscript{18}?
-
-MR does however have some important advantages over mediation based
-approaches. First, MR is used as a tool to not just deduce the existance
-of causality, but to also estimate the causal effect size. Second, using
-mediation requires that the exposure, outcome and instrument variables
-are all measured in the same data, whereas recent extensions to MR
-circumvent this requirement, allowing causal inference to be drawn when
-exposure variables and outcome variables are measured in different
-samples\textsuperscript{19}. This has the crucial advantage of improving
-statistical power by allowing analysis in much larger sample sizes.
-Third, and though this is often underappreciated in fields outside
-classical epidemiology it is of crucial importance, MR is substantially
-more robust to there being measurement error in the exposure and outcome
-variables. Indeed instrumental variable (IV) analysis was in part
-initially introduced as a correction for measurement
-error\textsuperscript{20}, whereas it has been noted that both basic
-mediation\textsuperscript{21--24} and mediation methods that use
-instrumental variables\textsuperscript{25,26} are prone to be unreliable
-in its presence. We argue that this is a serious limitation.
-
-Measurement (or observational) error is the difference between the
+associations.
+
+\subsection{Introduction}\label{introduction}
+
+Observational study data are now widely available but notoriously
+susceptible to pitfalls in making causal inference, with basic
+regression-based techniques unable to distinguish a true causal
+association from reverse causation or confounding\textsuperscript{3--5}.
+In response to this, genetic instrumentation has emerged as a valuable
+technique for improving the reliability of causal inference in
+observational data, and with the coincident rise in genome-wide
+association studies it is now a prominent tool that is applied in
+several different guises, and has been applied many forms of high
+throughput 'omic variables. However, an often neglected pitfall to these
+approaches is the influence of non-differential measurement error on the
+reliability of causal inference.
+
+Non-differential measurement error (or observational error, henceforth
+referred to simply as measurement error) is the difference between the
 measured value of a quantity and its true value. Such variability can
 arise through a whole plethora of mechanisms, which are often unique to
-the study design and difficult to avoid\textsuperscript{27,28}. Array
+the study design and difficult to avoid\textsuperscript{6,7}. Array
 technology is now commonly used to obtain high throughput phenotyping at
@@
 prone to have some amount of noise around the true
-value\textsuperscript{29,30}. Relatedly, if the measurement of
-biological interest is the methylation level in a T cell, then
-measurement error of this value can be introduced by using methylation
-levels from whole blood samples because the measured value will be an
-assay of many cell types\textsuperscript{31}.
+value\textsuperscript{8,9}. Relatedly, if the measurement of biological
+interest is the methylation level in a T cell, then measurement error of
+this value can be introduced by using methylation levels from whole
+blood samples because the measured value will be an assay of many cell
+types\textsuperscript{10}.
 
@@
 a proxy for adiposity, but it is clear that the correlation between BMI
-and underlying adiposity is not perfect\textsuperscript{32}. A similar
+and underlying adiposity is not perfect\textsuperscript{11}. A similar
 problem of biological misspecification is unavoidable in disease
@@
 variable. The sources of measurement error are not limited to this
-list\textsuperscript{28}, and its impact has been explored in the
-context of mediation analysis in the epidemiologyical literature
-extensively\textsuperscript{6,21,24}.
-
-Given the near-ubiquitous presence of measurement error in biological
-data it is vital to understand its impact on the tools we use for causal
-inference. Here we show using theory and simulations how
-non-differential measurement error and measurement imprecision can lead
-to unreliable causal inference in the mediation-based CIT method. We
-also present an extension to MR that allows researchers to ascertain the
-causal direction even when the biology of the instrument is not fully
-understood. This improves the utility of MR in cases where mediation
-based methods might have otherwise been used preferentially, and we
-apply this new method to infer the direction of causation between DNA
-methylation levels and gene expression levels.
+list\textsuperscript{7}, and its impact has been explored in the
+epidemiological literature extensively\textsuperscript{12--14}.
+
+Given the near-ubiquitous presence of measurement error in phenomic data
+it is vital to understand its impact on the tools we use for causal
+inference. An established study design that can provide information
+about causality is randomisation. Given the hypothesis that trait A
+(henceforth referred to as the exposure) is causally related to trait B
+(henceforth referred to as the outcome), randomisation can be employed
+to assess the causal nature of the association by randomly splitting the
+sample into two groups, subjecting one group to the exposure and leaving
+the other as a control. The association between the exposure and the
+outcome in this setting provides a robust estimate of the causal
+relationship. This provides the theoretical basis behind randomised
+control trials, but in practice randomisation is often impossible to
+implement in an experimental context due to cost, scale or inability to
+manipulate the exposure. The principle, however, can be employed in
+extant observational data through the use of genetic variants associated
+with the exposure (instruments), where the inheritance of an allele
+serves as a random lifetime allocation of differential exposure
+levels\textsuperscript{3}. Two statistical approaches to exploiting the
+properties of genetic instruments are widely used: mediation-based
+approaches and Mendelian randomisation (MR).
+
+Mediation-based approaches employ genetic instruments (typically single
+nucleotide polymorphisms, SNPs) to orient the causal direction between
+the exposure and the outcome. If a SNP is associated with an exposure,
+and the exposure is associated with some outcome, then it logically
+follows that the indirect influence of the SNP on the outcome will
+disappear when conditioning on the exposure. Here, the exposure
+`mediates' the association between the SNP and the outcome, providing
+information about the causal influence of the exposure on the outcome.
+This forms the basis of a number of methods such as genetical
+genomics\textsuperscript{15}, the regression-based causal inference test
+(CIT)\textsuperscript{16,17}, a structural equation modelling (SEM)
+implementation in the NEO software\textsuperscript{18}, and various
+other methods including Bayesian approaches\textsuperscript{19}. They
+have been employed by a number of recent publications that make causal
+inferences in large scale `omics datasets\textsuperscript{19--24}.
+
+MR can operate on the same data - phenotypic measures of the exposure
+and the outcome variables and a genetic instrument for the exposure -
+but the genetic instrument is employed in a subtly different manner.
+Here the SNP, which is assumed to have a direct association with the
+exposure, is used as a surrogate for the exposure. The causal effect of
+the exposure and the outcome can be estimated by scaling the association
+between the SNP and the outcome by the association between the SNP and
+the exposure.
+
+By utilising genetic instruments in different ways, mediation-based
+analysis and MR models have properties that confer some advantages and
+some disadvantages for reliable causal inference. The main advantage of
+mediation over MR is that the true underlying biology of the genetic
+instrument doesn't necessarily need to be known as long as it associates
+with the putative exposure. In the CIT framework (described fully in the
+Methods) for example, the test statistic is different if you test for
+the exposure causing the outcome or the outcome causing the exposure,
+allowing the researcher to infer the direction of causality between two
+variables by performing the test in both directions and choosing the
+model with the strongest evidence.
+
+Such is not the case for MR, where in order to infer the direction of
+causality the instrument must be known to influence the exposure
+directly, associating with the outcome only through the exposure. This
+requirement can be problematic because if there exists a putative
+association between two variables, with the instrumental SNP being
+robustly associated with each, it can be difficult to determine which of
+the two variables is subject to the direct effect of the SNP (i.e., for
+which of the two variables is the SNP a valid instrument? Figure 1). By
+definition, we expect that if the association is causal then a SNP for
+the exposure will be associated with the outcome, so if the researcher
+erroneously uses the SNP as an instrument for the outcome then they are
+likely to see an apparently robust causal association of outcome on
+exposure. Such a situation can arise in many scenarios. Genome-wide
+association studies (GWASs) that identify genetic associations for
+complex traits are, by design, hypothesis free and agnostic of genomic
+function, and it often takes years of follow up studies to understand
+the biological nature of a putative GWAS hit\textsuperscript{25}. For
+example, the same FTO variant that is associated with body mass index
+(BMI) is also associated with C-reactive protein (CRP). It is known that
+this is because BMI causes CRP\textsuperscript{26}, but supposing we
+only had \emph{a priori} knowledge that FTO associated with CRP, and we
+used it to instrument CRP in an MR analysis of CRP against BMI, we would
+mistakenly infer that CRP causes BMI. Another situation is where the
+causal directions between 'omic measures need to be determined, for
+example if a DNA methylation probe is associated with expression of an
+adjacent gene, then is a cis-acting SNP an instrument for the DNA
+methylation level, or the gene expression level\textsuperscript{27}
+(Figure 1)?
+
+MR does however have some important advantages over mediation based
+approaches. First, MR is used as a tool to not just deduce the existance
+of causality, but to also estimate the causal effect size. Second, using
+mediation requires that the exposure, outcome and instrument variables
+are all measured in the same data, whereas recent extensions to MR
+circumvent this requirement, allowing causal inference to be drawn when
+exposure variables and outcome variables are measured in different
+samples\textsuperscript{28}. This has the crucial advantage of improving
+statistical power by allowing analysis in much larger sample sizes, and
+dramatically expands the breadth of possible phenotypic relationships
+that can be evaluated. Third, MR is substantially more robust to there
+being measurement error in the exposure and outcome variables. Indeed
+instrumental variable (IV) analysis was in part initially introduced as
+a correction for measurement error\textsuperscript{29}, whereas it has
+been noted that both classic mediation-based
+analyses\textsuperscript{13,14,30,31} and mediation-based methods that
+use instrumental variables\textsuperscript{32,33} are prone to be
+unreliable in its presence. We argue that this is a serious limitation.
+
+Here we show using theory and simulations how non-differential
+measurement error and measurement imprecision can lead to unreliable
+causal inference in the mediation-based CIT method. We then present an
+extension to MR that allows researchers to ascertain the causal
+direction of an association even when the biology of the instruments are
+not fully understood, and to evaluate the sensitivity of the result to
+measurement error. This improves the utility of MR in cases where
+mediation based methods might have otherwise been used preferentially.
+Finally, we apply this new method to infer the direction of causation
+between DNA methylation levels and gene expression levels.
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{33} and code is made
+the R programming language\textsuperscript{34} and code is made
 available at \href{}{github url to go here}.
@@
 
-The CIT method\textsuperscript{8} is implemented in the R package
-\emph{R/cit}\textsuperscript{9}. The methodology of the CIT is as
+The CIT method\textsuperscript{16} is implemented in the R package
+\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
 follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
@@
 is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
-R package \(R/systemfit\)\textsuperscript{34}.
+R package \(R/systemfit\)\textsuperscript{35}.
 
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population\textsuperscript{35}. It is calculated as
+within a population\textsuperscript{36}. It is calculated as
 
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{19} setting. Two-sample MR refers to the
+two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
 case where the SNP-exposure association and SNP-outcome association are
@@
 \emph{r.test} function in the R package
-\emph{R/psych}\textsuperscript{36}.
+\emph{R/psych}\textsuperscript{37}.
 
@@
 
-Two sample MR\textsuperscript{19} was performed using the summary
-statistics presented in\textsuperscript{18} for each of 4122
+Two sample MR\textsuperscript{28} was performed using the summary
+statistics presented in\textsuperscript{27} for each of 4122
 associations between DNA methylation and gene expression levels. Because
@@
 list of 458 gene expression - DNA methylation associations as reported
-in Shakhbazov et al\textsuperscript{18}. These were filtered to be
+in Shakhbazov et al\textsuperscript{27}. These were filtered to be
 located on the same chromosome, have robost correlations after
@@
 
-We performed two sample MR\textsuperscript{19} for each association in
+We performed two sample MR\textsuperscript{28} for each association in
 the direction of causality that we estimated to be the correct direction
@@
 always in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{18}, and that there was a
+reported by Shakhbazov et al\textsuperscript{27}, and that there was a
 moderate correlation between the magnitude of the causal correlation and
@@
 In the epidemiological literature issues of measurement error in
-mediation is relatively well explored\textsuperscript{6}. Our analysis
+mediation is relatively well explored\textsuperscript{12}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 Mediation based network approaches are very well
-established\textsuperscript{26} and have a number of extensions that
+established\textsuperscript{33} and have a number of extensions that
 make them valuable tools, including for example network construction.
@@
 error. Recent advances in MR methodology, for example applying MR to
-genetical genomics\textsuperscript{37}, multivariate
-MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
-through MR\textsuperscript{40} may offer more robust alternatives for
+genetical genomics\textsuperscript{38}, multivariate
+MR\textsuperscript{39}, network MR\textsuperscript{40} and mediation
+through MR\textsuperscript{41} may offer more robust alternatives for
 these more complicated problems.
@@
 exposure) was not present. Recent method developments in
-MR\textsuperscript{41} have focused on accounting for the issues that
+MR\textsuperscript{42} have focused on accounting for the issues that
 horizontal pleiotropy can introduce, but how they perform in the
@@
 
-\hypertarget{ref-Baron1986}{}
-4. Baron, R. M. \& Kenny, D. A. The moderator-mediator variable
-distinction in social psychological research: conceptual, strategic, and
-statistical considerations. \emph{Journal of personality and social
-psychology} \textbf{51,} 1173--82 (1986).
-
-\hypertarget{ref-Aten2008}{}
-5. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
-markers to orient the edges in quantitative trait networks: the NEO
-software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
-
-\hypertarget{ref-Fewell2007}{}
-6. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
-residual and unmeasured confounding in epidemiologic studies: a
-simulation study. \emph{American journal of epidemiology} \textbf{166,}
-646--55 (2007).
-
-\hypertarget{ref-Schadt2005}{}
-7. Schadt, E. E. \emph{et al.} An integrative genomics approach to infer
-causal associations between gene expression and disease. \emph{Nature
-Genetics} \textbf{37,} 710--717 (2005).
-
-\hypertarget{ref-Millstein2009}{}
-8. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
-molecular relationships with a causal inference test. \emph{BMC
-genetics} \textbf{10,} 23 (2009).
-
-\hypertarget{ref-Millstein2016}{}
-9. Millstein, J. \emph{cit: Causal Inference Test. R package version
-1.9}. (2016). at
-\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
-
-\hypertarget{ref-Waszak2015}{}
-10. Waszak, S. M. \emph{et al.} Variation and genetic control of
-chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
-(2015).
-
-\hypertarget{ref-Koestler2014}{}
-11. Koestler, D. C. \emph{et al.} Integrative genomic analysis
-identifies epigenetic marks that mediate genetic risk for epithelial
-ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
-
-\hypertarget{ref-Liu2013}{}
-12. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
-methylation as an intermediary of genetic risk in rheumatoid arthritis.
-\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).
-
-\hypertarget{ref-Yuan2014}{}
-13. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
-diabetes susceptibility loci in monozygotic twins. \emph{Nature
-communications} \textbf{5,} 5719 (2014).
-
-\hypertarget{ref-Tang2014}{}
-14. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
-with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
-translational medicine} \textbf{6,} 257ra139 (2014).
-
-\hypertarget{ref-Hong2015}{}
-15. Hong, X. \emph{et al.} Genome-wide association study identifies
-peanut allergy-specific loci and evidence of epigenetic mediation in US
-children. \emph{Nature communications} \textbf{6,} 6304 (2015).
-
-\hypertarget{ref-Claussnitzer2015}{}
-16. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
-Adipocyte Browning in Humans. \emph{The New England journal of medicine}
-\textbf{373,} 895--907 (2015).
-
-\hypertarget{ref-Timpson2011}{}
-17. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
-index: elucidating direction of causation through reciprocal Mendelian
-randomization. \emph{International journal of obesity (2005)}
-\textbf{35,} 300--8 (2011).
-
-\hypertarget{ref-Shakhbazov2016}{}
-18. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
-and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
-278 (2016).
-
-\hypertarget{ref-Pierce2013}{}
-19. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
-randomization studies: subsample and 2-sample instrumental variable
-estimators. \emph{American journal of epidemiology} \textbf{178,}
-1177--84 (2013).
-
-\hypertarget{ref-Ashenfelter1994}{}
-20. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
-to Schooling from a New Sample of Twins. \emph{The American Economic
-Review} \textbf{84,} 1157--1173 (1994).
-
-\hypertarget{ref-LeCessie2012}{}
-21. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
-Vandenbroucke, J. P. Quantification of bias in direct effects estimates
-due to different types of measurement error in the mediator.
-\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
-
-\hypertarget{ref-Nagarajan2013}{}
-22. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
-inference. \emph{PloS one} \textbf{8,} e80735 (2013).
-
-\hypertarget{ref-Shpitser2010}{}
-23. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
-covariate adjustment for estimating causal effects. \emph{Proceedings of
-the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
-(UAI-10)} 527--536 (2010).
-
-\hypertarget{ref-Blakely2013}{}
-24. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
-mediator matters when estimating indirect effects. \emph{Journal of
-epidemiology and community health} \textbf{67,} 458--66 (2013).
-
-\hypertarget{ref-Wang2015}{}
-25. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
-gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
-
-\hypertarget{ref-Lagani2015}{}
-26. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
-I. in \emph{Uncertainty in biology: A computational modeling approach}
-47 (Springer, 2015). at
-\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+\hypertarget{ref-Phillips1991}{}
+4. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
+effects? relative risk estimation when correlated exposures are measured
+imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
+1223--1231 (1991).
+
+\hypertarget{ref-DaveySmith2002}{}
+5. Davey Smith, G. \& Ebrahim, S. Data dredging, bias, or confounding.
+\emph{BMJ} \textbf{325,} (2002).
 
 \hypertarget{ref-Houle2011}{}
-27. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
+6. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
 meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
@@
 \hypertarget{ref-Hernan2009}{}
-28. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
+7. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
 measurement bias. \emph{American journal of epidemiology} \textbf{170,}
@@
 \hypertarget{ref-Harper2013}{}
-29. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
+8. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
 pathway analysis: two potential perils in cancer studies involving DNA
@@
 \hypertarget{ref-Chen2013a}{}
-30. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
+9. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
 polymorphic CpGs in the Illumina Infinium HumanMethylation450
@@
 \hypertarget{ref-Houseman2012}{}
-31. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
+10. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
 measures of cell mixture distribution. \emph{BMC bioinformatics}
@@
 \hypertarget{ref-Ahima2013}{}
-32. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
+11. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
 obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
@@
 
+\hypertarget{ref-Fewell2007}{}
+12. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
+residual and unmeasured confounding in epidemiologic studies: a
+simulation study. \emph{American journal of epidemiology} \textbf{166,}
+646--55 (2007).
+
+\hypertarget{ref-LeCessie2012}{}
+13. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+Vandenbroucke, J. P. Quantification of bias in direct effects estimates
+due to different types of measurement error in the mediator.
+\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
+
+\hypertarget{ref-Blakely2013}{}
+14. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+mediator matters when estimating indirect effects. \emph{Journal of
+epidemiology and community health} \textbf{67,} 458--66 (2013).
+
+\hypertarget{ref-Schadt2005}{}
+15. Schadt, E. E. \emph{et al.} An integrative genomics approach to
+infer causal associations between gene expression and disease.
+\emph{Nature Genetics} \textbf{37,} 710--717 (2005).
+
+\hypertarget{ref-Millstein2009}{}
+16. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
+molecular relationships with a causal inference test. \emph{BMC
+genetics} \textbf{10,} 23 (2009).
+
+\hypertarget{ref-Millstein2016}{}
+17. Millstein, J. cit: Causal Inference Test. R package version 1.9.
+(2016). at
+\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
+
+\hypertarget{ref-Aten2008}{}
+18. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using
+genetic markers to orient the edges in quantitative trait networks: the
+NEO software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
+
+\hypertarget{ref-Waszak2015}{}
+19. Waszak, S. M. \emph{et al.} Variation and genetic control of
+chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
+(2015).
+
+\hypertarget{ref-Koestler2014}{}
+20. Koestler, D. C. \emph{et al.} Integrative genomic analysis
+identifies epigenetic marks that mediate genetic risk for epithelial
+ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
+
+\hypertarget{ref-Liu2013}{}
+21. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+methylation as an intermediary of genetic risk in rheumatoid arthritis.
+\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).
+
+\hypertarget{ref-Yuan2014}{}
+22. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
+diabetes susceptibility loci in monozygotic twins. \emph{Nature
+communications} \textbf{5,} 5719 (2014).
+
+\hypertarget{ref-Tang2014}{}
+23. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
+with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
+translational medicine} \textbf{6,} 257ra139 (2014).
+
+\hypertarget{ref-Hong2015}{}
+24. Hong, X. \emph{et al.} Genome-wide association study identifies
+peanut allergy-specific loci and evidence of epigenetic mediation in US
+children. \emph{Nature communications} \textbf{6,} 6304 (2015).
+
+\hypertarget{ref-Claussnitzer2015}{}
+25. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+Adipocyte Browning in Humans. \emph{The New England journal of medicine}
+\textbf{373,} 895--907 (2015).
+
+\hypertarget{ref-Timpson2011}{}
+26. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+index: elucidating direction of causation through reciprocal Mendelian
+randomization. \emph{International journal of obesity (2005)}
+\textbf{35,} 300--8 (2011).
+
+\hypertarget{ref-Shakhbazov2016}{}
+27. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
+278 (2016).
+
+\hypertarget{ref-Pierce2013}{}
+28. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+randomization studies: subsample and 2-sample instrumental variable
+estimators. \emph{American journal of epidemiology} \textbf{178,}
+1177--84 (2013).
+
+\hypertarget{ref-Ashenfelter1994}{}
+29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+to Schooling from a New Sample of Twins. \emph{The American Economic
+Review} \textbf{84,} 1157--1173 (1994).
+
+\hypertarget{ref-Nagarajan2013}{}
+30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+inference. \emph{PloS one} \textbf{8,} e80735 (2013).
+
+\hypertarget{ref-Shpitser2010}{}
+31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+covariate adjustment for estimating causal effects. \emph{Proceedings of
+the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
+(UAI-10)} 527--536 (2010).
+
+\hypertarget{ref-Wang2015}{}
+32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
+
+\hypertarget{ref-Lagani2015}{}
+33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+I. in \emph{Uncertainty in biology: A computational modeling approach}
+47 (Springer, 2015). at
+\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+
 \hypertarget{ref-RCoreTeam2015}{}
-33. R Core Team. \emph{R: A Language and Environment for Statistical
-Computing}. (R Foundation for Statistical Computing, 2015). at
+34. R Core Team. R: A Language and Environment for Statistical
+Computing. (2015). at
 \textless{}\url{https://www.r-project.org/}\textgreater{}
@@
 \hypertarget{ref-Henningsen2007}{}
-34. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+35. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. \emph{Journal of Statistical
@@
 \hypertarget{ref-Steiger1980}{}
-35. Steiger, J. H. Tests for comparing elements of a correlation matrix.
+36. Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
@@
 \hypertarget{ref-Revelle2015}{}
-36. Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
-and Personality Research}. (Northwestern University, 2015). at
+37. Revelle, W. psych: Procedures for Psychological, Psychometric, and
+Personality Research. (2015). at
 \textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
@@
 \hypertarget{ref-Relton2012}{}
-37. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+38. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-38. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+39. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Burgess2015}{}
-39. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+40. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 \hypertarget{ref-Varbo2015}{}
-40. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+41. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Bowden2015}{}
-41. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
+42. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
 with invalid instruments: effect estimation and bias detection through

### 2016-12-31 docs/manuscript.Rmd
updated

@@
 Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y.
-A measure of sensitivity, therefore, is the volume of the positive space bound by the blue surface as a proportion of the total space bound by the blue surface.
-Sensitivity of 0 denotes no measurement error values that could lead to inferrence of incorrect direction, whereas the maximum value of 0.5 denotes measurement error equal chance of either direction across all measurement error ranges.
-In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the sensitivity is 0.18.
-b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger sensitivity values (y-axis) for different values of $\\rho_{g,x}^2$.
-Inferring the direction of causality is more susceptible to measurement error when there are stronger causal associations between X and Y.",
+A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$.
+In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$.
+b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger reliability values (y-axis) for different values of $\\rho_{g,x}^2$.
+Inferring the direction of causality is more sensitive to measurement error when there are stronger causal associations between X and Y.",
 	display=FALSE
@@
 
-Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
-We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and to evaluate the sensitivity of the result to measurement error.
+Using theory and simulations we show how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
+We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result to measurement error.
 This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
@@
 
+### Steiger sensitivity analysis
+
+The Steiger test for testing if $x \rightarrow y$ is based on evaluating $\\rho_{gx} > \\rho_{gy}$.
+However, $\\rho_{gx}$ and $\\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely.
+If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\\rho_{g,x_o} > \\rho_{g,y_o}$.
+To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we calculate the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
+We also calculate the measurement error values which are not compatible.
+The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction.
+A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
+Full details are provided in Appendix 2.
+
 ### Simulations
@@
 Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0.
-However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $y$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $x$.
-It also shows that in most cases where $y$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
+However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
+It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.
 
@@
 In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$.
-In order to assess how sensitive the Steiger test is to measurement error, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.
+In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.
 
@@
 Where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality.
-To evaluate the sensitivity, $S$, of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,
+To evaluate the reliability, $R$, of the Steiger test with regards to measurement error, the objective is to compare the weighted proportion of the parameter space that agrees with the inferred direction against the weighted proportion which does not:
 
 $$
-S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
+R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
 $$
 
+If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values.
+If $R > 1$ then $R$ times as much of the weighted parameter space favours the inferred direction of causality.
 $V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:
@@
 
-$V_{z>0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
+$V_{z \ge 0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
 The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$.
@@
 \begin{aligned}
-V_{z>0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
+V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
 & = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
@@
 
-The sensitivity, $S$, can take values from 0 to 0.5, where low values suggest there are
+Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.
 

### 2016-12-31 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{29 December 2016}
+  \date{31 December 2016}
 
@@
 
-The last decade has seem a rapid acceleration in technological advances
-that enable us to assess the human phenome in ever increasing detail,
-furnishing biologists with the tantalising opportunity to map the
-molecular basis of complex traits. This does, however, require
-overcoming the statistical challenge of correctly defining the causal
-architectures that underlie these high dimensional data. For example,
-identifying metabolic perturbations that lead to
-disease\textsuperscript{1}, or understanding the influence of epigenomic
-influences on gene expression\textsuperscript{2} all depend on
-understanding the causal nature of any putative observational
-associations.
-
 \subsection{Introduction}\label{introduction}
 
-Observational study data are now widely available but notoriously
-susceptible to pitfalls in making causal inference, with basic
-regression-based techniques unable to distinguish a true causal
-association from reverse causation or confounding\textsuperscript{3--5}.
-In response to this, genetic instrumentation has emerged as a valuable
-technique for improving the reliability of causal inference in
-observational data, and with the coincident rise in genome-wide
-association studies it is now a prominent tool that is applied in
-several different guises, and has been applied many forms of high
-throughput 'omic variables. However, an often neglected pitfall to these
-approaches is the influence of non-differential measurement error on the
-reliability of causal inference.
+Observational measures of the human phenome are growing ever more
+abundant, but using these data to make causal inference is notoriously
+susceptible to many pitfalls, with basic regression-based techniques
+unable to distinguish a true causal association from reverse causation
+or confounding\textsuperscript{1--3}. In response to this, genetic
+instrumentation has emerged as a technique for improving the reliability
+of causal inference in observational data, and with the coincident rise
+in genome-wide association studies it is now a prominent tool that is
+applied in several different guises\textsuperscript{3--6}. However,
+potential pitfalls still persist and one that is often neglected is the
+influence of non-differential measurement error on the reliability of
+causal inference.
 
@@
 arise through a whole plethora of mechanisms, which are often unique to
-the study design and difficult to avoid\textsuperscript{6,7}. Array
+the study design and difficult to avoid\textsuperscript{7,8}. Array
 technology is now commonly used to obtain high throughput phenotyping at
@@
 prone to have some amount of noise around the true
-value\textsuperscript{8,9}. Relatedly, if the measurement of biological
+value\textsuperscript{9,10}. Relatedly, if the measurement of biological
 interest is the methylation level in a T cell, then measurement error of
@@
 blood samples because the measured value will be an assay of many cell
-types\textsuperscript{10}.
+types\textsuperscript{11}.
 
@@
 a proxy for adiposity, but it is clear that the correlation between BMI
-and underlying adiposity is not perfect\textsuperscript{11}. A similar
+and underlying adiposity is not perfect\textsuperscript{12}. A similar
 problem of biological misspecification is unavoidable in disease
@@
 variable. The sources of measurement error are not limited to this
-list\textsuperscript{7}, and its impact has been explored in the
-epidemiological literature extensively\textsuperscript{12--14}.
+list\textsuperscript{8}, and its impact has been explored in the
+epidemiological literature extensively\textsuperscript{13--15}.
 
@@
 This forms the basis of a number of methods such as genetical
-genomics\textsuperscript{15}, the regression-based causal inference test
-(CIT)\textsuperscript{16,17}, a structural equation modelling (SEM)
-implementation in the NEO software\textsuperscript{18}, and various
-other methods including Bayesian approaches\textsuperscript{19}. They
-have been employed by a number of recent publications that make causal
-inferences in large scale `omics datasets\textsuperscript{19--24}.
+genomics\textsuperscript{16}, the regression-based causal inference test
+(CIT)\textsuperscript{4,17}, a structural equation modelling (SEM)
+implementation in the NEO software\textsuperscript{5}, and various other
+methods including Bayesian approaches\textsuperscript{6}. They have been
+employed by a number of recent publications that make causal inferences
+in large scale `omics datasets\textsuperscript{6,18--22}.
 
@@
 function, and it often takes years of follow up studies to understand
-the biological nature of a putative GWAS hit\textsuperscript{25}. For
+the biological nature of a putative GWAS hit\textsuperscript{23}. For
 example, the same FTO variant that is associated with body mass index
 (BMI) is also associated with C-reactive protein (CRP). It is known that
-this is because BMI causes CRP\textsuperscript{26}, but supposing we
+this is because BMI causes CRP\textsuperscript{24}, but supposing we
 only had \emph{a priori} knowledge that FTO associated with CRP, and we
@@
 adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level\textsuperscript{27}
+methylation level, or the gene expression level\textsuperscript{25}
 (Figure 1)?
@@
 exposure variables and outcome variables are measured in different
-samples\textsuperscript{28}. This has the crucial advantage of improving
+samples\textsuperscript{26}. This has the crucial advantage of improving
 statistical power by allowing analysis in much larger sample sizes, and
@@
 instrumental variable (IV) analysis was in part initially introduced as
-a correction for measurement error\textsuperscript{29}, whereas it has
+a correction for measurement error\textsuperscript{27}, whereas it has
 been noted that both classic mediation-based
-analyses\textsuperscript{13,14,30,31} and mediation-based methods that
-use instrumental variables\textsuperscript{32,33} are prone to be
+analyses\textsuperscript{14,15,28,29} and mediation-based methods that
+use instrumental variables\textsuperscript{30,31} are prone to be
 unreliable in its presence. We argue that this is a serious limitation.
 
-Here we show using theory and simulations how non-differential
-measurement error and measurement imprecision can lead to unreliable
-causal inference in the mediation-based CIT method. We then present an
+Using theory and simulations we show how non-differential measurement
+error and measurement imprecision can lead to unreliable causal
+inference in the mediation-based CIT method. We then present an
 extension to MR that allows researchers to ascertain the causal
 direction of an association even when the biology of the instruments are
-not fully understood, and to evaluate the sensitivity of the result to
-measurement error. This improves the utility of MR in cases where
-mediation based methods might have otherwise been used preferentially.
-Finally, we apply this new method to infer the direction of causation
-between DNA methylation levels and gene expression levels.
+not fully understood, and also a method to evaluate the sensitivity of
+the result to measurement error. This improves the utility of MR in
+cases where mediation based methods might have otherwise been used
+preferentially. Finally, we apply this new method to infer the direction
+of causation between DNA methylation levels and gene expression levels.
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{34} and code is made
+the R programming language\textsuperscript{32} and code is made
 available at \href{}{github url to go here}.
@@
 
-The CIT method\textsuperscript{16} is implemented in the R package
+The CIT method\textsuperscript{4} is implemented in the R package
 \emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
@@
 is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
-R package \(R/systemfit\)\textsuperscript{35}.
+R package \(R/systemfit\)\textsuperscript{33}.
 
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population\textsuperscript{36}. It is calculated as
+within a population\textsuperscript{34}. It is calculated as
 
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
+two-sample MR\textsuperscript{26} setting. Two-sample MR refers to the
 case where the SNP-exposure association and SNP-outcome association are
@@
 \emph{r.test} function in the R package
-\emph{R/psych}\textsuperscript{37}.
+\emph{R/psych}\textsuperscript{35}.
+
+\subsubsection{Steiger sensitivity
+analysis}\label{steiger-sensitivity-analysis}
+
+The Steiger test for testing if \(x \rightarrow y\) is based on
+evaluating \(\\rho_{gx} > \\rho_{gy}\). However, \(\\rho_{gx}\) and
+\(\\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
+imprecisely. If, for example, \(x\) has lower measurement precision than
+\(y\) then we might empirically obtain
+\(\\rho_{g,x_o} > \\rho_{g,y_o}\). To evaluate how reliable, \(R\), the
+Steiger test is to potential measurement imprecision in \(x\) and \(y\)
+we calculate the possible measurement error values which are compatible
+with the direction of causality obtained empirically from the Steiger
+test. We also calculate the measurement error values which are not
+compatible. The ratio of these quantities is an indication of how much
+of the measurement error space supports the inferred causal direction. A
+ratio \(R=1\) indicates that the Steiger test is highly sensitive to
+measurement error, with equal weight of the measurement error parameter
+space supporting both directions of causality. Full details are provided
+in Appendix 2.
 
@@
 
-Two sample MR\textsuperscript{28} was performed using the summary
-statistics presented in\textsuperscript{27} for each of 4122
+Two sample MR\textsuperscript{26} was performed using the summary
+statistics presented in\textsuperscript{25} for each of 4122
 associations between DNA methylation and gene expression levels. Because
@@
 than 0. However Figure 7 shows that when there is no measurement error
-in \(y\), the Steiger test is unlikely to infer the wrong direction of
-causality even if there is measurement error in \(x\). It also shows
-that in most cases where \(y\) is measured with error, especially when
+in \(x\), the Steiger test is unlikely to infer the wrong direction of
+causality even if there is measurement error in \(y\). It also shows
+that in most cases where \(x\) is measured with error, especially when
 the causal effect between \(x\) and \(y\) is not very large, the
-condition of \(d > 0\) is satisfied.
+sensitivity of the Steiger test to measurement error is relatively low.
 
@@
 list of 458 gene expression - DNA methylation associations as reported
-in Shakhbazov et al\textsuperscript{27}. These were filtered to be
+in Shakhbazov et al\textsuperscript{25}. These were filtered to be
 located on the same chromosome, have robost correlations after
@@
 
-We performed two sample MR\textsuperscript{28} for each association in
+We performed two sample MR\textsuperscript{26} for each association in
 the direction of causality that we estimated to be the correct direction
@@
 always in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{27}, and that there was a
+reported by Shakhbazov et al\textsuperscript{25}, and that there was a
 moderate correlation between the magnitude of the causal correlation and
@@
 expression levels at a controlling CpG typically leads to increased gene
-expression levels, consistent with expectation\textsuperscript{2}.
+expression levels, consistent with expectation\textsuperscript{36}.
 
@@
 In the epidemiological literature issues of measurement error in
-mediation is relatively well explored\textsuperscript{12}. Our analysis
+mediation is relatively well explored\textsuperscript{13}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 Mediation based network approaches are very well
-established\textsuperscript{33} and have a number of extensions that
+established\textsuperscript{31} and have a number of extensions that
 make them valuable tools, including for example network construction.
@@
 error. Recent advances in MR methodology, for example applying MR to
-genetical genomics\textsuperscript{38}, multivariate
-MR\textsuperscript{39}, network MR\textsuperscript{40} and mediation
-through MR\textsuperscript{41} may offer more robust alternatives for
+genetical genomics\textsuperscript{37}, multivariate
+MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
+through MR\textsuperscript{40} may offer more robust alternatives for
 these more complicated problems.
@@
 exposure) was not present. Recent method developments in
-MR\textsuperscript{42} have focused on accounting for the issues that
+MR\textsuperscript{41} have focused on accounting for the issues that
 horizontal pleiotropy can introduce, but how they perform in the
@@
 
+\begin{verbatim}
+## $r2_exp
+## [1] 0.1
+## 
+## $r2_out
+## [1] 0.03162278
+## 
+## $r2_exp_adj
+## [1] 0.1
+## 
+## $r2_out_adj
+## [1] 0.03162278
+## 
+## $correct_causal_direction
+## [1] TRUE
+## 
+## $steiger_test
+## [1] 0
+## 
+## $correct_causal_direction_adj
+## [1] TRUE
+## 
+## $steiger_test_adj
+## [1] 0
+## 
+## $vz
+## [1] -0.08934366
+## 
+## $vz0
+## [1] 0.0262863
+## 
+## $vz1
+## [1] 0.11563
+## 
+## $sensitivity
+## [1] 0.185224
+## 
+## $sensitivity_ratio
+## [1] 4.398868
+## 
+## $sensitivity_plot
+\end{verbatim}
+
 \includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
+\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-2.pdf}
+\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-3.pdf}
 
@@
 below the red plane, these measurement error values support the inferred
-causal direction of X to Y. A measure of sensitivity, therefore, is the
-volume of the positive space bound by the blue surface as a proportion
-of the total space bound by the blue surface. Sensitivity of 0 denotes
-no measurement error values that could lead to inferrence of incorrect
-direction, whereas the maximum value of 0.5 denotes measurement error
-equal chance of either direction across all measurement error ranges. In
-this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the
-sensitivity is 0.18. b) The influence of \(\rho_{x,y}\) (x-axis) on the
-Steiger sensitivity values (y-axis) for different values of
-\(\rho_{g,x}^2\). Inferring the direction of causality is more
-susceptible to measurement error when there are stronger causal
-associations between X and Y.
+causal direction of X to Y. A measure of reliability, therefore, is the
+ratio of the negative and positive volumes of the total space bound by
+the blue and red surfaces, \(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\).
+In this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\),
+the \(R = 4.40\), which means that 4.40 times as much of the possible
+measurement error values are in support of the \(x \rightarrow y\)
+direction of causality than \(y \rightarrow x\). b) The influence of
+\(\rho_{x,y}\) (x-axis) on the Steiger reliability values (y-axis) for
+different values of \(\rho_{g,x}^2\). Inferring the direction of
+causality is more sensitive to measurement error when there are stronger
+causal associations between X and Y.
 
@@
 Steiger test will instead be assessing the inequality
-\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how sensitive the
-Steiger test is to measurement error, we can evaluate the range of
-potential values of measurement error in \(x\) and \(y\) over which the
-empirical result from the Steiger test would return the wrong value.
+\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how reliable the
+Steiger test is in the presence of measurement imprecision, we can
+evaluate the range of potential values of measurement error in \(x\) and
+\(y\) over which the empirical result from the Steiger test would return
+the wrong value.
 
@@
 \(x_o\) and \(y_o\) will not change the qualitative inference of the
-direction of causality. To evaluate the sensitivity, \(S\), of the
-Steiger test to measurement error, the objective is to calculate the
-proportion of the total absolute volume of this function that is above
-the \(z = 0\) plane,
-
-\[
-S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
-\]
-
-\(V_{z}\), the total volume of the function (Figure 3), can be obtained
-analytically by solving:
+direction of causality. To evaluate the reliability, \(R\), of the
+Steiger test to measurement error, the objective is to compare the
+weighted proportion of the parameter space that agrees with the inferred
+direction against the weighted proportion which does not:
+
+\[
+R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
+\]
+
+If \(R=1\) then the direction of causality is equally probable across
+the range of possible measurement error values. If \(R > 1\) then \(R\)
+times as much of the weighted parameter space favours the inferred
+direction of causality. \(V_{z}\), the total volume of the function
+(Figure 3), can be obtained analytically by solving:
 
@@
 
-\(V_{z>0}\), the proportion of the volume that lies above the \(z=0\)
-plane, can also be obtained analytically. The region of this volume is
-bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) where
-\(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
+\(V_{z \ge 0}\), the proportion of the volume that lies above the
+\(z=0\) plane, can also be obtained analytically. The region of this
+volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
+where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
 \(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,
@@
 
-The sensitivity, \(S\), can take values from 0 to 0.5, where low values
-suggest there are
+Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Kettunen2012}{}
-1. Kettunen, J. \emph{et al.} Genome-wide association study identifies
-multiple loci influencing human serum metabolite levels. \emph{Nature
-genetics} \textbf{44,} 269--76 (2012).
-
-\hypertarget{ref-Bird2002}{}
-2. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
-\& development} \textbf{16,} 6--21 (2002).
+\hypertarget{ref-Phillips1991}{}
+1. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
+effects? relative risk estimation when correlated exposures are measured
+imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
+1223--1231 (1991).
+
+\hypertarget{ref-DaveySmith2002}{}
+2. Davey Smith, G. \& Ebrahim, S. Data dredging, bias, or confounding.
+\emph{BMJ} \textbf{325,} (2002).
 
@@
 
-\hypertarget{ref-Phillips1991}{}
-4. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
-effects? relative risk estimation when correlated exposures are measured
-imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
-1223--1231 (1991).
-
-\hypertarget{ref-DaveySmith2002}{}
-5. Davey Smith, G. \& Ebrahim, S. Data dredging, bias, or confounding.
-\emph{BMJ} \textbf{325,} (2002).
+\hypertarget{ref-Millstein2009}{}
+4. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
+molecular relationships with a causal inference test. \emph{BMC
+genetics} \textbf{10,} 23 (2009).
+
+\hypertarget{ref-Aten2008}{}
+5. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
+markers to orient the edges in quantitative trait networks: the NEO
+software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
+
+\hypertarget{ref-Waszak2015}{}
+6. Waszak, S. M. \emph{et al.} Variation and genetic control of
+chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
+(2015).
 
 \hypertarget{ref-Houle2011}{}
-6. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
+7. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
 meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
@@
 \hypertarget{ref-Hernan2009}{}
-7. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
+8. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
 measurement bias. \emph{American journal of epidemiology} \textbf{170,}
@@
 \hypertarget{ref-Harper2013}{}
-8. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
+9. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
 pathway analysis: two potential perils in cancer studies involving DNA
@@
 \hypertarget{ref-Chen2013a}{}
-9. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
+10. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
 polymorphic CpGs in the Illumina Infinium HumanMethylation450
@@
 \hypertarget{ref-Houseman2012}{}
-10. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
+11. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
 measures of cell mixture distribution. \emph{BMC bioinformatics}
@@
 \hypertarget{ref-Ahima2013}{}
-11. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
+12. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
 obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
@@
 \hypertarget{ref-Fewell2007}{}
-12. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
+13. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
 residual and unmeasured confounding in epidemiologic studies: a
@@
 \hypertarget{ref-LeCessie2012}{}
-13. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+14. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
 Vandenbroucke, J. P. Quantification of bias in direct effects estimates
@@
 \hypertarget{ref-Blakely2013}{}
-14. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+15. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
 mediator matters when estimating indirect effects. \emph{Journal of
@@
 \hypertarget{ref-Schadt2005}{}
-15. Schadt, E. E. \emph{et al.} An integrative genomics approach to
+16. Schadt, E. E. \emph{et al.} An integrative genomics approach to
 infer causal associations between gene expression and disease.
 \emph{Nature Genetics} \textbf{37,} 710--717 (2005).
-
-\hypertarget{ref-Millstein2009}{}
-16. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
-molecular relationships with a causal inference test. \emph{BMC
-genetics} \textbf{10,} 23 (2009).
 
@@
 
-\hypertarget{ref-Aten2008}{}
-18. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using
-genetic markers to orient the edges in quantitative trait networks: the
-NEO software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
-
-\hypertarget{ref-Waszak2015}{}
-19. Waszak, S. M. \emph{et al.} Variation and genetic control of
-chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
-(2015).
-
 \hypertarget{ref-Koestler2014}{}
-20. Koestler, D. C. \emph{et al.} Integrative genomic analysis
+18. Koestler, D. C. \emph{et al.} Integrative genomic analysis
 identifies epigenetic marks that mediate genetic risk for epithelial
@@
 \hypertarget{ref-Liu2013}{}
-21. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+19. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
 methylation as an intermediary of genetic risk in rheumatoid arthritis.
@@
 \hypertarget{ref-Yuan2014}{}
-22. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
+20. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
 diabetes susceptibility loci in monozygotic twins. \emph{Nature
@@
 \hypertarget{ref-Tang2014}{}
-23. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
+21. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
 with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
@@
 \hypertarget{ref-Hong2015}{}
-24. Hong, X. \emph{et al.} Genome-wide association study identifies
+22. Hong, X. \emph{et al.} Genome-wide association study identifies
 peanut allergy-specific loci and evidence of epigenetic mediation in US
@@
 \hypertarget{ref-Claussnitzer2015}{}
-25. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+23. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
 Adipocyte Browning in Humans. \emph{The New England journal of medicine}
@@
 \hypertarget{ref-Timpson2011}{}
-26. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+24. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
 index: elucidating direction of causation through reciprocal Mendelian
@@
 \hypertarget{ref-Shakhbazov2016}{}
-27. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+25. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
 and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
@@
 \hypertarget{ref-Pierce2013}{}
-28. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+26. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
 randomization studies: subsample and 2-sample instrumental variable
@@
 \hypertarget{ref-Ashenfelter1994}{}
-29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+27. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
 to Schooling from a New Sample of Twins. \emph{The American Economic
@@
 \hypertarget{ref-Nagarajan2013}{}
-30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+28. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
@@
 \hypertarget{ref-Shpitser2010}{}
-31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+29. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 \hypertarget{ref-Wang2015}{}
-32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+30. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
 gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
@@
 \hypertarget{ref-Lagani2015}{}
-33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+31. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
 I. in \emph{Uncertainty in biology: A computational modeling approach}
@@
 \hypertarget{ref-RCoreTeam2015}{}
-34. R Core Team. R: A Language and Environment for Statistical
+32. R Core Team. R: A Language and Environment for Statistical
 Computing. (2015). at
@@
 \hypertarget{ref-Henningsen2007}{}
-35. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+33. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. \emph{Journal of Statistical
@@
 \hypertarget{ref-Steiger1980}{}
-36. Steiger, J. H. Tests for comparing elements of a correlation matrix.
+34. Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
@@
 \hypertarget{ref-Revelle2015}{}
-37. Revelle, W. psych: Procedures for Psychological, Psychometric, and
+35. Revelle, W. psych: Procedures for Psychological, Psychometric, and
 Personality Research. (2015). at
@@
 
+\hypertarget{ref-Bird2002}{}
+36. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+\& development} \textbf{16,} 6--21 (2002).
+
 \hypertarget{ref-Relton2012}{}
-38. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+37. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-39. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+38. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Burgess2015}{}
-40. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+39. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 \hypertarget{ref-Varbo2015}{}
-41. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+40. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Bowden2015}{}
-42. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
+41. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
 with invalid instruments: effect estimation and bias detection through

### 2016-12-31 scripts/mr_directionality_analysis.R

@@ -79,12 +79,17 @@ steiger_sensitivity <- function(rgx_o, rgy_o, ...)
 	vz <- a * log(a) - b * log(b) + a*b*(log(b)-log(a))
 	vz0 <- -2*b - b * log(a) - a*b*log(a) + 2*a*b
 
+	vz1 <- abs(vz - vz0)
+
 	sensitivity <- vz0 / (2 * vz0 + abs(vz))
+	sensitivity_ratio <- vz1 / vz0
 
 	return(list(
 		vz = vz,
 		vz0 = vz0,
+		vz1 = vz1,
 		sensitivity = sensitivity,
+		sensitivity_ratio = sensitivity_ratio,
 		pl = temp
 	))
 }
@@ -119,7 +124,9 @@ mr_steiger <- function(p_exp, p_out, n_exp, n_out, r_xxo = 1, r_yyo=1, ...)
 		steiger_test_adj = rtest_adj$p,
 		vz = sensitivity$vz,
 		vz0 = sensitivity$vz0,
+		vz1 = sensitivity$vz1,
 		sensitivity = sensitivity$sensitivity,
+		sensitivity_ratio = sensitivity$sensitivity_ratio,
 		sensitivity_plot = sensitivity$pl
 	)
 	return(l)
@@ -821,11 +828,13 @@ labs(x=TeX("$|\\rho_{P}|$"), y=TeX("$|\\rho_{MR}|$"))
 ## ---- steiger_sensitivity ----
 
 sensitivity_parameters <- expand.grid(
-	r_xy = c(seq(-1,-0.1,0.1), -0.001, 0.001, seq(0.1,1,0.1)),
+	r_xy = seq(-1,1,0.05),
 	r_gx = sqrt(c(0.01,0.05,0.1)),
 	n=1000,
 	sensitivity = NA
 )
+
+sensitivity_parameters$r_gy <- sensitivity_parameters$r_gx^2 * sensitivity_parameters$r_xy^2
 
 sensitivity_parameters$r_gy <- sensitivity_parameters$r_xy * sensitivity_parameters$r_gx
 sensitivity_parameters$p_gy <- get_p_from_r2n(sensitivity_parameters$r_gy^2, sensitivity_parameters$n)
@@ -833,7 +842,7 @@ sensitivity_parameters$p_gx <- get_p_from_r2n(sensitivity_parameters$r_gx^2, sen
 
 for(i in 1:nrow(sensitivity_parameters))
 {
[…]

### 2016-12-31 docs/manuscript.Rmd
updated

@@
 Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association.
-We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).
+We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`).
+We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
+`r cfig("shakhplot")` shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
 
@@
 
-In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$.
+In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
+
+$$
+\begin{aligned}
+\rho_{g, x_O} & > \rho_{g, y_O} \\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
+\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
+\end{aligned}
+$$
+
 In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.

### 2017-01-02 docs/manuscript.Rmd
updated

@@
 	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them.
-The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels.
+a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels.
 The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes).
-If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor."),
+If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor.
+b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis).
+c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
 	display=FALSE
@@
 
-### MR analysis of gene expression and DNA methylation
+### MR analysis of gene expression and DNA methylation levels
 
@@
 Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association.
-We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`).
+We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a).
 We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
-`r cfig("shakhplot")` shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
+`r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
 
 We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test.
-We observed that the sign of the MR estimate was always in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], and that there was a moderate correlation between the magnitude of the causal correlation and the observational Pearson correlation (`r ctab("shakhtab")`).
+We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], but with a few exceptions (`r cfig("shakhplot")`b).
+There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index]))`).
+Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.
 
-We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`)), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
+We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
 
@@
 
-Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and using genetic instruments as surrogates for randomisation is a useful approach.
-MR and the CIT employ genetic instruments in different ways, and unfortunately neither offer solutions that work perfectly in all scenarios.
-In the face of measurement error it is evident that causal inference drawn through mediation based methods such as the CIT analysis typically fare worse than MR and will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error present in observational data.
-
-In the epidemiological literature issues of measurement error in mediation is relatively well explored [@Fewell2007].
+Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data.
+In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Fewell2007].
 Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data.
-These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely have reduced statistical power, and liable to infer the wrong direction of causality.
-We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
+These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
+We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
 
@@
 
-Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
+We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels.
+We demonstrate that inferring the causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
+For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association.
+Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.
+
+Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
 But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
@@
 In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
-Recent method developments in MR [@Bowden2015] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
+Recent method developments in MR [@Bowden2015; @Bowden2016] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
 An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.

### 2017-01-02 scripts/mr_directionality_analysis.R

@@ -682,9 +682,11 @@ for(i in 1:nrow(qtl))
 	l$sensitivity_plot
 	qtl$r_exp[i] <- l$r2_out_adj
 	qtl$r_meth[i] <- l$r2_exp_adj
+	qtl$steiger[i] <- l$steiger_test
 	qtl$dir[i] <- l$correct_causal_direction_adj
 	qtl$dir_p[i] <- l$steiger_test_adj
 	qtl$sensitivity[i] <- l$sensitivity
+	qtl$reliability[i] <- l$sensitivity_ratio
 
 	if(qtl$AL1_meth[i] != qtl$AL1_expr[i])
 	{
@@ -713,6 +715,8 @@ shakhbazov$mr_r <- sign(shakhbazov$mr_eff) * sqrt(abs(qnorm(shakhbazov$mr_p, low
 shakhbazov$dir <- as.character(shakhbazov$dir)
 shakhbazov$dir[shakhbazov$dir=="TRUE"] <- "Methylation causes Expression"
 shakhbazov$dir[shakhbazov$dir=="FALSE"] <- "Expression causes Methylation"
+shakhbazov$r_diff <- abs(shakhbazov$r_exp - shakhbazov$r_meth)
+
 real_index <- shakhbazov$rxx_o == 1 & shakhbazov$ryy_o == 1
 
 
@@ -722,7 +726,8 @@ shakhsummary <- dplyr::group_by(shakhbazov, rxx_o, ryy_o, dir) %>%
 		nsig = sum(dir_p < 0.05, na.rm=TRUE),
 		pos = sum(mr_eff > 0) / n(),
 		corr = cor(pearson^2, mr_r^2),
-		sens = mean(sensitivity, na.rm=TRUE)
+		sens = mean(sensitivity, na.rm=TRUE),
+		reliability = mean(reliability, na.rm=TRUE)
 	) %>% as.data.frame()
 
 
@@ -791,12 +796,35 @@ temp$ryy_o_lab <- paste0("cor(M,Mo) = ", temp$ryy_o)
 # facet_grid(rxx_o_lab ~ ryy_o_lab) 
 
 
-ggplot(subset(temp, key=="n"), aes(x=as.factor(rxx_o), y=value)) +
+p1 <- ggplot(subset(temp, key=="n"), aes(x=as.factor(rxx_o), y=value)) +
 geom_bar(stat="Identity", aes(fill=dir), position="stack") +
[…]

### 2017-01-03 docs/manuscript.Rmd
updated

@@
     keep_tex: true
+  word_document: default
 bibliography: manuscript.bib
@@
 
-With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables.
-A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure.
-We show that in the face of measurement error this can lead to erroneous inference of both existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
-Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible.
-We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and introduce a simple extension to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
+A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses.
+Here we show that when measurement error is present, commonly-used tools predicated on this approach are susceptible to making erroneous inference of both the existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
+We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
 This extension is less susceptible to measurement error than mediation-based analyses.
-Using these methods we demonstrate empirically that the inferred direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms, demanding caution when making such claims about causality.
+We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms.
+Other 'omic platforms are likely to be similarly susceptible, necessitating the use of sensitivity analyses to evaluate reliability.
 
@@
 
-Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+Measurement error will of course arise in more basic data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
 Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
-The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].
+The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].
 
@@
 This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
-The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004].
+The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014].
 Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).
@@
 By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference.
-The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
+An important advantage of mediation-based analysis is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
 In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.
+Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.
 
-Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.
-This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?
+Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e.
+for which of the two variables is the SNP a valid instrument?
 `r cfig("dag_figure")`).
@@
 It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI.
-Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?
+If multiple instruments are available for an hypothesised exposure then this problem can be ameliorated [@DaveySmithHemani2014], however this is not always the case, especially in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known.
+For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?
 
-MR does however have some important advantages over mediation based approaches.
+MR does however have some important advantages over the mediation-based approaches.
 First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size.
-Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
-This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated.
-Third, MR is substantially more robust to there being measurement error in the exposure and outcome variables.
-Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
-We argue that this is a serious limitation.
+Second, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
+This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
+Third, MR is substantially more robust to there being measurement error in the exposure variable.
+Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
 
@@
 Included in the MR analysis is a simple extension that is used to make inference of the causal direction when it is unknown which of the correlated variables the instrument is directly acting upon.
-All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]().
+All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
 
@@
 
-where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
+where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
 If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$.
@@
 In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
-To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$.
+To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for the incorrect model of $y$ causing $x$.
 For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:
@@
 
-For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to use p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds.
+For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to rely on p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds [@Sterne2001].
 
@@
 
-The direction of the causal association was inferred as follows.
-Assuming no horizontal pleiotropy (i.e.
-$g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables, $x$ or $y$, it has a direct influence on.
-This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2).
-This test can be formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
+In what follows we assume no horizontal pleiotropy, such that $g$ only has an effect on one of the variables through the other variable.
+To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$.
+This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
 It is calculated as
@@
 Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting.
-Two-sample MR refers to the case where the SNP-exposure association and SNP-outcome association are calculated in different samples (or from publicly available summary statistics).
+Two-sample MR refers to the case where the SNP-exposure association and SNP-outcome association are calculated in different samples (e.g.
+from publicly available summary statistics).
 Here the Steiger test of two independent correlations can be applied where.
@@
 
-Simulations were conducting by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$.
+Simulations were conducted by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$.
 In all models $x$ causes $y$ and $g$ is an instrument for $x$.
@@
 Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels.
+To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
+These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
 Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations.
@@
 
-We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels.
-To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
-These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
-Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association.
-We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a).
+We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a).
 We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
@@
 
-We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels.
-We demonstrate that inferring the causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
+We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data.
+The causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
 For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association.
@@
 
-Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
-But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
-Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a], network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.
-
-In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
-Recent method developments in MR [@Bowden2015; @Bowden2016] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
+In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available.
+However, MR strategies exist for other scenarios, for example if there are multiple instruments for the hypothesised exposure then the problems described here may be circumvented [@DaveySmithHemani2014].
+Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014].
+In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
+Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
 An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.
@@
 
-Our analysis of DNA methylation and gene expression summary data demonstates the simplicity of inferring causal existence, direction and magnitude in the two sample MR framework.
-We note that verifying these results with regards to the extent to which measurement error has influenced them empirically is a challenging task and beyond the scope of this work.
+Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
+But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
+Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015] may offer more robust alternatives for these more complicated problems.
 

### 2017-01-03 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{31 December 2016}
+  \date{03 January 2017}
 
@@
 
-With our ability to characterize the human phenome ever improving it is
-becoming increasingly common to use statistical tests to infer the
-causal relationships between correlated variables. A simple and commonly
-used method for inferring if an exposure is on the causal pathway to an
-outcome is through using genetic instruments in mediation analysis,
-where the effect of an instrument on the outcome is tested before and
-after adjusting for the exposure. We show that in the face of
-measurement error this can lead to erroneous inference of both existence
-of causality and the causal direction, and that increasing sample size,
-rather than reducing bias, can often have the effect of increasing
-certainty in the wrong answer. Commonly used tools that are predicated
-on this methodology such as the causal inference test (CIT) are shown to
-be susceptible. We demonstrate that Mendelian randomization (MR) is a
-method for causal inference that is robust to measurement error, and
-introduce a simple extension to estimate the direction of causality
-between correlated variables even when the biology of the instrument is
-unknown. This extension is less susceptible to measurement error than
-mediation-based analyses. Using these methods we demonstrate empirically
-that the inferred direction of causality between correleted DNA
-methylation and gene expression levels is highly susceptible to
-differences in measurement error between the two assaying platforms,
-demanding caution when making such claims about causality.
+A common approach for inferring causal relationships between two traits
+is to use genetic instruments in mediation-based analyses. Here we show
+that when measurement error is present, commonly-used tools predicated
+on this approach are susceptible to making erroneous inference of both
+the existence of causality and the causal direction, and that increasing
+sample size, rather than reducing bias, can often have the effect of
+increasing certainty in the wrong answer. We demonstrate that Mendelian
+randomization (MR) is a technique for causal inference that is robust to
+measurement error, and here we extend the method with new sensitivity
+analyses to estimate the direction of causality between correlated
+variables even when the biology of the instrument is unknown. This
+extension is less susceptible to measurement error than mediation-based
+analyses. We use it to demonstrate empirically that inferring the
+direction of causality between correleted DNA methylation and gene
+expression levels is highly susceptible to differences in measurement
+error between the two assaying platforms. Other 'omic platforms are
+likely to be similarly susceptible, necessitating the use of sensitivity
+analyses to evaluate reliability.
 
@@
 
-Measurement error will of course arise in more low-tech data too, for
+Measurement error will of course arise in more basic data too, for
 example when measuring BMI one is typically interested in using this as
@@
 list\textsuperscript{8}, and its impact has been explored in the
-epidemiological literature extensively\textsuperscript{13--15}.
+epidemiological literature extensively\textsuperscript{13,14}.
 
@@
 serves as a random lifetime allocation of differential exposure
-levels\textsuperscript{3}. Two statistical approaches to exploiting the
-properties of genetic instruments are widely used: mediation-based
+levels\textsuperscript{15,16}. Two statistical approaches to exploiting
+the properties of genetic instruments are widely used: mediation-based
 approaches and Mendelian randomisation (MR).
@@
 This forms the basis of a number of methods such as genetical
-genomics\textsuperscript{16}, the regression-based causal inference test
-(CIT)\textsuperscript{4,17}, a structural equation modelling (SEM)
+genomics\textsuperscript{17}, the regression-based causal inference test
+(CIT)\textsuperscript{4,18}, a structural equation modelling (SEM)
 implementation in the NEO software\textsuperscript{5}, and various other
@@
 employed by a number of recent publications that make causal inferences
-in large scale `omics datasets\textsuperscript{6,18--22}.
+in large scale `omics datasets\textsuperscript{6,19--23}.
 
@@
 analysis and MR models have properties that confer some advantages and
-some disadvantages for reliable causal inference. The main advantage of
-mediation over MR is that the true underlying biology of the genetic
-instrument doesn't necessarily need to be known as long as it associates
-with the putative exposure. In the CIT framework (described fully in the
-Methods) for example, the test statistic is different if you test for
-the exposure causing the outcome or the outcome causing the exposure,
-allowing the researcher to infer the direction of causality between two
-variables by performing the test in both directions and choosing the
-model with the strongest evidence.
-
-Such is not the case for MR, where in order to infer the direction of
-causality the instrument must be known to influence the exposure
-directly, associating with the outcome only through the exposure. This
-requirement can be problematic because if there exists a putative
-association between two variables, with the instrumental SNP being
-robustly associated with each, it can be difficult to determine which of
-the two variables is subject to the direct effect of the SNP (i.e., for
-which of the two variables is the SNP a valid instrument? Figure 1). By
-definition, we expect that if the association is causal then a SNP for
-the exposure will be associated with the outcome, so if the researcher
-erroneously uses the SNP as an instrument for the outcome then they are
-likely to see an apparently robust causal association of outcome on
-exposure. Such a situation can arise in many scenarios. Genome-wide
-association studies (GWASs) that identify genetic associations for
-complex traits are, by design, hypothesis free and agnostic of genomic
-function, and it often takes years of follow up studies to understand
-the biological nature of a putative GWAS hit\textsuperscript{23}. For
-example, the same FTO variant that is associated with body mass index
-(BMI) is also associated with C-reactive protein (CRP). It is known that
-this is because BMI causes CRP\textsuperscript{24}, but supposing we
-only had \emph{a priori} knowledge that FTO associated with CRP, and we
-used it to instrument CRP in an MR analysis of CRP against BMI, we would
-mistakenly infer that CRP causes BMI. Another situation is where the
-causal directions between 'omic measures need to be determined, for
-example if a DNA methylation probe is associated with expression of an
-adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level\textsuperscript{25}
-(Figure 1)?
-
-MR does however have some important advantages over mediation based
+some disadvantages for reliable causal inference. An important advantage
+of mediation-based analysis is that the true underlying biology of the
+genetic instrument doesn't necessarily need to be known as long as it
+associates with the putative exposure. In the CIT framework (described
+fully in the Methods) for example, the test statistic is different if
+you test for the exposure causing the outcome or the outcome causing the
+exposure, allowing the researcher to infer the direction of causality
+between two variables by performing the test in both directions and
+choosing the model with the strongest evidence. Such is not the case for
+MR, where in order to infer the direction of causality the instrument
+must be known to influence the exposure directly, associating with the
+outcome only through the exposure.
+
+Assuming biological knowledge of genetic associations can be problematic
+because if there exists a putative association between two variables,
+with the instrumental SNP being robustly associated with each, it can be
+difficult to determine which of the two variables is subject to the
+direct effect of the SNP (i.e.~for which of the two variables is the SNP
+a valid instrument? Figure 1). By definition, we expect that if the
+association is causal then a SNP for the exposure will be associated
+with the outcome, so if the researcher erroneously uses the SNP as an
+instrument for the outcome then they are likely to see an apparently
+robust causal association of outcome on exposure. Such a situation can
+arise in many scenarios. Genome-wide association studies (GWASs) that
+identify genetic associations for complex traits are, by design,
+hypothesis free and agnostic of genomic function, and it often takes
+years of follow up studies to understand the biological nature of a
+putative GWAS hit\textsuperscript{24}. For example, the same FTO variant
+that is associated with body mass index (BMI) is also associated with
+C-reactive protein (CRP). It is known that this is because BMI causes
+CRP\textsuperscript{25}, but supposing we only had \emph{a priori}
+knowledge that FTO associated with CRP, and we used it to instrument CRP
+in an MR analysis of CRP against BMI, we would mistakenly infer that CRP
+causes BMI. If multiple instruments are available for an hypothesised
+exposure then this problem can be ameliorated\textsuperscript{16},
+however this is not always the case, especially in the case of
+determining causal directions between 'omic measures where typically
+only one cis-acting SNP is known. For example if a DNA methylation probe
+is associated with expression of an adjacent gene, then is a cis-acting
+SNP an instrument for the DNA methylation level, or the gene expression
+level\textsuperscript{26} (Figure 1)?
+
+MR does however have some important advantages over the mediation-based
 approaches. First, MR is used as a tool to not just deduce the existance
-of causality, but to also estimate the causal effect size. Second, using
-mediation requires that the exposure, outcome and instrument variables
-are all measured in the same data, whereas recent extensions to MR
-circumvent this requirement, allowing causal inference to be drawn when
-exposure variables and outcome variables are measured in different
-samples\textsuperscript{26}. This has the crucial advantage of improving
-statistical power by allowing analysis in much larger sample sizes, and
-dramatically expands the breadth of possible phenotypic relationships
-that can be evaluated. Third, MR is substantially more robust to there
-being measurement error in the exposure and outcome variables. Indeed
-instrumental variable (IV) analysis was in part initially introduced as
-a correction for measurement error\textsuperscript{27}, whereas it has
-been noted that both classic mediation-based
-analyses\textsuperscript{14,15,28,29} and mediation-based methods that
-use instrumental variables\textsuperscript{30,31} are prone to be
-unreliable in its presence. We argue that this is a serious limitation.
+of causality, but to also estimate the causal effect size. Second, the
+mediation-based approaches require that the exposure, outcome and
+instrumental variables are all measured in the same data, whereas recent
+extensions to MR circumvent this requirement, allowing causal inference
+to be drawn when exposure variables and outcome variables are measured
+in different samples\textsuperscript{27}. This has the crucial advantage
+of improving statistical power by allowing analysis in much larger
+sample sizes, and dramatically expands the breadth of possible
+phenotypic relationships that can be evaluated\textsuperscript{28}.
+Third, MR is substantially more robust to there being measurement error
+in the exposure variable. Indeed instrumental variable (IV) analysis was
+in part initially introduced as a correction for measurement error in
+the exposure\textsuperscript{29}, whereas it has been noted that both
+classic mediation-based analyses\textsuperscript{13,14,30,31} and
+mediation-based methods that use instrumental
+variables\textsuperscript{32,33} are prone to be unreliable in its
+presence.
 
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{32} and code is made
-available at \href{}{github url to go here}.
+the R programming language\textsuperscript{34} and code is made
+available at \href{}{github url to go here} and implemented in the
+MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{28}.
 
@@
 The CIT method\textsuperscript{4} is implemented in the R package
-\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
+\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
 follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
@@
 where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
-residual of \(y\) after adjusting for the \(x\), where \(x\) is assumed
-to mediate the association between the SNP and the outcome. If all four
+residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
+mediate the association between the SNP and the outcome. If all four
 tests reject the null hypothesis then it is inferred that \(x\) is
@@
 method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
-for the correct model of \(x\) causing \(y\), and for incorrect model of
-\(y\) causing \(x\). For some significance threshold \(\alpha\), the
-existence of causality and its direction was inferred based on the
-following scenarios:
+for the correct model of \(x\) causing \(y\), and for the incorrect
+model of \(y\) causing \(x\). For some significance threshold
+\(\alpha\), the existence of causality and its direction was inferred
+based on the following scenarios:
 
@@
 \(\alpha = 0.05\) value, though we should stress that for real analyses
-it is not good practice to use p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds.
+it is not good practice to rely on p-values for making causal inference,
+nor is it reliable to depend on arbitrary significance
+thresholds\textsuperscript{35}.
 
@@
 is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
-R package \(R/systemfit\)\textsuperscript{33}.
-
-The direction of the causal association was inferred as follows.
-Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
-of the variables through the other variable) it is desirable to know
-which of the variables, \(x\) or \(y\), it has a direct influence on.
-This can be achieved by assessing which of the two variables has the
-biggest absolute correlation with \(g\) (Appendix 2). This test can be
+R package \(R/systemfit\)\textsuperscript{36}.
+
+In what follows we assume no horizontal pleiotropy, such that \(g\) only
+has an effect on one of the variables through the other variable. To
+infer the direction of causality it is desirable to know which of the
+variables, \(x\) or \(y\), is being directly influenced by the
+instrument \(g\). This can be achieved by assessing which of the two
+variables has the biggest absolute correlation with \(g\) (Appendix 2),
 formalised by testing for a difference in the correlations \(\rho_{gx}\)
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population\textsuperscript{34}. It is calculated as
+within a population\textsuperscript{37}. It is calculated as
 
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{26} setting. Two-sample MR refers to the
+two-sample MR\textsuperscript{27} setting. Two-sample MR refers to the
 case where the SNP-exposure association and SNP-outcome association are
-calculated in different samples (or from publicly available summary
+calculated in different samples (e.g.~from publicly available summary
 statistics). Here the Steiger test of two independent correlations can
@@
 \emph{r.test} function in the R package
-\emph{R/psych}\textsuperscript{35}.
+\emph{R/psych}\textsuperscript{38}.
 
@@
 
-Simulations were conducting by creating variables of sample size \(n\)
+Simulations were conducted by creating variables of sample size \(n\)
 for the exposure \(x\), the measured values of the exposure \(x_o\), the
@@
 
-Two sample MR\textsuperscript{26} was performed using the summary
-statistics presented in\textsuperscript{25} for each of 4122
-associations between DNA methylation and gene expression levels. Because
-only summary statistics were available (effect, standard error, effect
-allele, sample size, p-values) for the instrumental SNP on the
-methylation and gene expression levels, the Steiger test of two
-independent correlations was used to infer the direction of causality
-for each of the associations. The Wald ratio test was then used to
-estimate the causal effect size for the estimated direction for each
-association.
+Two sample MR\textsuperscript{27} was performed using the summary
+statistics presented in\textsuperscript{26} for each of 4122
+associations between DNA methylation and gene expression levels. To do
+this we obtained a list of 458 gene expression - DNA methylation
+associations as reported in Shakhbazov et al\textsuperscript{26}. These
+were filtered to be located on the same chromosome, have robost
+correlations after correcting for multiple testing, and to share a SNP
+that had a robust cis-acting effect on both the DNA methylation probe
+and the gene expression probe. Because only summary statistics were
+available (effect, standard error, effect allele, sample size, p-values)
+for the instrumental SNP on the methylation and gene expression levels,
+the Steiger test of two independent correlations was used to infer the
+direction of causality for each of the associations. The Wald ratio test
+was then used to estimate the causal effect size for the estimated
+direction for each association.
 
@@
 
-\subsubsection{MR analysis of gene expression and DNA
-methylation}\label{mr-analysis-of-gene-expression-and-dna-methylation}
+\subsubsection{MR analysis of gene expression and DNA methylation
+levels}\label{mr-analysis-of-gene-expression-and-dna-methylation-levels}
 
 We used the Steiger test to infer the direction of causality between DNA
-methylation levels and gene expression levels. To do this we obtained a
-list of 458 gene expression - DNA methylation associations as reported
-in Shakhbazov et al\textsuperscript{25}. These were filtered to be
-located on the same chromosome, have robost correlations after
-correcting for multiple testing, and to share a SNP that had a robust
-cis-acting effect on both the DNA methylation probe and the gene
-expression probe. Using the reported summary data we used the Steiger
-test of two independent correlations to identify which direction of
-causality was most likely for each association. We found that the causal
-direction commonly goes in both directions (Table 1) but methylation
-levels were more likely to influence expression levels than vice versa
-(probability of expression causing methylation = 0.0163873 (95\% CI
-0.3738727 - 0.4874269) amongst NA associations with
-\(p_{Steiger} < 0.05\)).
-
-We performed two sample MR\textsuperscript{26} for each association in
+methylation levels and gene expression levels and found that the causal
+direction commonly goes in both directions (Figure 6a). We then went on
+to predict the causal directions of the associations for varying levels
+of possible measurement error. Figure 6a shows that the qualitative
+result of which of DNA methylation or gene expression is more commonly
+the causal factor is very strongly determined by measurement error.
+
+We performed two sample MR\textsuperscript{27} for each association in
 the direction of causality that we estimated to be the correct direction
 from the Steiger test. We observed that the sign of the MR estimate was
-always in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{25}, and that there was a
-moderate correlation between the magnitude of the causal correlation and
-the observational Pearson correlation (Table 1).
+generally in the same direction as the Pearson correlation coefficient
+reported by Shakhbazov et al\textsuperscript{26}, but with a few
+exceptions (Figure 6b). There was a moderate correlation between the
+absolute magnitudes of the causal correlation and the observational
+Pearson correlation (r = 0.4423627). Together these inferences suggest
+that even for relatively low level associations it is important to use
+causal inference methods over observational associations to infer causal
+effect sizes.
 
@@
 associations where gene expression caused methylation (OR = 0.5326663
-(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
-expression levels at a controlling CpG typically leads to increased gene
-expression levels, consistent with expectation\textsuperscript{36}.
+(95\% CI 0.3172203 - 0.8851868), Figure 6c), suggesting that reducing
+gene expression levels at a controlling CpG typically leads to increased
+gene expression levels, consistent with expectation\textsuperscript{39}.
 
@@
 Researchers are often confronted with the problem of making causal
-inferences using a statistical framework on observational data, and
-using genetic instruments as surrogates for randomisation is a useful
-approach. MR and the CIT employ genetic instruments in different ways,
-and unfortunately neither offer solutions that work perfectly in all
-scenarios. In the face of measurement error it is evident that causal
-inference drawn through mediation based methods such as the CIT analysis
-typically fare worse than MR and will be difficult to interpret, and
-this is a serious problem considering that it is often impossible to
-estimate the extent of measurement error present in observational data.
-
-In the epidemiological literature issues of measurement error in
-mediation is relatively well explored\textsuperscript{13}. Our analysis
+inferences using a statistical framework on observational data. In the
+epidemiological literature issues of measurement error in mediation
+analysis is relatively well explored\textsuperscript{40}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 that as measurement error in the exposure variable increases, CIT is
-likely have reduced statistical power, and liable to infer the wrong
-direction of causality. We also demonstrate that, though seemingly
-unintuitive, increasing sample size does not resolve the issue, rather
-it leads to more extreme p-values for the model that predicts the wrong
-direction of causality.
+likely to have reduced statistical power, and liable to infer the wrong
+direction of causality. We also demonstrate that, though unintuitive,
+increasing sample size does not resolve the issue, rather it leads to
+more extreme p-values for the model that predicts the wrong direction of
+causality.
 
@@
 
-Mediation based network approaches are very well
-established\textsuperscript{31} and have a number of extensions that
-make them valuable tools, including for example network construction.
-But because they are predicated on the basic underlying principles of
-mediation they are liable to suffer from the same issues of measurement
-error. Recent advances in MR methodology, for example applying MR to
-genetical genomics\textsuperscript{37}, multivariate
-MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
-through MR\textsuperscript{40} may offer more robust alternatives for
-these more complicated problems.
+We demonstrate this new method by evaluating the causal relationships of
+458 known associations between DNA methylation and gene expression
+levels using summary level data. The causal direction is heavily
+influenced by how much measurement error is present in the different
+assaying platforms. For example, if DNA methylation measures typically
+have higher measuremet error than gene expression measures then our
+analysis suggests that DNA methylation levels would be more often the
+causal factor in the association. Indeed, previous studies which have
+evaluated measurement error in these platforms do support this
+position\textsuperscript{41,42}, though making strong conclusions for
+this analysis is difficult because measurement error is likely to be
+study specific.
 
 In our simulations we focused on the simple case of a single instrument
-in a single sample setting, and assumed that pleiotropy (the influence
-of the instrument on the outcome through a mechanism other than the
-exposure) was not present. Recent method developments in
-MR\textsuperscript{41} have focused on accounting for the issues that
+in a single sample setting with a view to making a fair comparison
+between MR and the various mediation-based methods available. However,
+MR strategies exist for other scenarios, for example if there are
+multiple instruments for the hypothesised exposure then the problems
+described here may be circumvented\textsuperscript{16}. Likewise, if
+there is at least one instrument for each trait then bi-directional MR
+can offer solutions to inferring the causal
+direction\textsuperscript{43}. In this work we assumed that pleiotropy
+(the influence of the instrument on the outcome through a mechanism
+other than the exposure) was not present. Recent method developments in
+MR\textsuperscript{44,45} have focused on accounting for the issues that
 horizontal pleiotropy can introduce, but how they perform in the
@@
 
-Our analysis of DNA methylation and gene expression summary data
-demonstates the simplicity of inferring causal existence, direction and
-magnitude in the two sample MR framework. We note that verifying these
-results with regards to the extent to which measurement error has
-influenced them empirically is a challenging task and beyond the scope
-of this work.
+Mediation based network approaches, that go beyond analyses of two
+variables, are very well established\textsuperscript{33} and have a
+number of extensions that make them valuable tools, including for
+example network construction. But because they are predicated on the
+basic underlying principles of mediation they are liable to suffer from
+the same issues of measurement error. Recent advances in MR methodology,
+for example applying MR to genetical genomics\textsuperscript{46},
+multivariate MR\textsuperscript{47} and mediation through
+MR\textsuperscript{48,49} may offer more robust alternatives for these
+more complicated problems.
 
@@
 
-\begin{verbatim}
-## $r2_exp
-## [1] 0.1
-## 
-## $r2_out
-## [1] 0.03162278
-## 
-## $r2_exp_adj
-## [1] 0.1
-## 
-## $r2_out_adj
-## [1] 0.03162278
-## 
-## $correct_causal_direction
-## [1] TRUE
-## 
-## $steiger_test
-## [1] 0
-## 
-## $correct_causal_direction_adj
-## [1] TRUE
-## 
-## $steiger_test_adj
-## [1] 0
-## 
-## $vz
-## [1] -0.08934366
-## 
-## $vz0
-## [1] 0.0262863
-## 
-## $vz1
-## [1] 0.11563
-## 
-## $sensitivity
-## [1] 0.185224
-## 
-## $sensitivity_ratio
-## [1] 4.398868
-## 
-## $sensitivity_plot
-\end{verbatim}
-
 \includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
-\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-2.pdf}
-\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-3.pdf}
 
@@
 gene expression we used the Steiger test to infer the direction of
-causality between them. The rightmost bar shows the proportion of
+causality between them. a) The rightmost bar shows the proportion of
 associations for each of the two possible causal directions (colour key)
@@
 measurement error in one platform it will appear to be less likely to be
-the causal factor.
+the causal factor. b) The relationship between the Pearson correlation
+between DNA methylation and gene expression levels (x-axis) and the
+causal estimate (scaled to be in standard deviation units, y-axis). c)
+Distribution of estimated causal effect sizes, stratified into
+associations inferred to be due to DNA methylation causing expression
+(blue) and expression causing DNA methylation (red).
 
@@
 Steiger test will instead be assessing the inequality
-\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how reliable the
-Steiger test is in the presence of measurement imprecision, we can
-evaluate the range of potential values of measurement error in \(x\) and
-\(y\) over which the empirical result from the Steiger test would return
-the wrong value.
+\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
+
+\[
+\begin{aligned}
+\rho_{g, x_O} & > \rho_{g, y_O} \\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
+\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
+\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
+\end{aligned}
+\]
+
+In order to assess how reliable the Steiger test is in the presence of
+measurement imprecision, we can evaluate the range of potential values
+of measurement error in \(x\) and \(y\) over which the empirical result
+from the Steiger test would return the wrong value.
 
@@
 direction of causality. To evaluate the reliability, \(R\), of the
-Steiger test to measurement error, the objective is to compare the
-weighted proportion of the parameter space that agrees with the inferred
-direction against the weighted proportion which does not:
+Steiger test with regards to measurement error, the objective is to
+compare the weighted proportion of the parameter space that agrees with
+the inferred direction against the weighted proportion which does not:
 
@@
 
+\hypertarget{ref-LeCessie2012}{}
+13. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
+Vandenbroucke, J. P. Quantification of bias in direct effects estimates
+due to different types of measurement error in the mediator.
+\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
+
+\hypertarget{ref-Blakely2013}{}
+14. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
+mediator matters when estimating indirect effects. \emph{Journal of
+epidemiology and community health} \textbf{67,} 458--66 (2013).
+
+\hypertarget{ref-DaveySmith2003}{}
+15. Davey Smith, G. \& Ebrahim, S. 'Mendelian randomization': can
+genetic epidemiology contribute to understanding environmental
+determinants of disease? \emph{International Journal of Epidemiology}
+\textbf{32,} 1--22 (2003).
+
+\hypertarget{ref-DaveySmithHemani2014}{}
+16. Davey Smith, G. \& Hemani, G. Mendelian randomization: genetic
+anchors for causal inference in epidemiological studies. \emph{Human
+molecular genetics} \textbf{23,} R89-----R98 (2014).
+
+\hypertarget{ref-Schadt2005}{}
+17. Schadt, E. E. \emph{et al.} An integrative genomics approach to
+infer causal associations between gene expression and disease.
+\emph{Nature Genetics} \textbf{37,} 710--717 (2005).
+
+\hypertarget{ref-Millstein2016}{}
+18. Millstein, J. cit: Causal Inference Test. R package version 1.9.
+(2016). at
+\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
+
+\hypertarget{ref-Koestler2014}{}
+19. Koestler, D. C. \emph{et al.} Integrative genomic analysis
+identifies epigenetic marks that mediate genetic risk for epithelial
+ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
+
+\hypertarget{ref-Liu2013}{}
+20. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
+methylation as an intermediary of genetic risk in rheumatoid arthritis.
+\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).
+
+\hypertarget{ref-Yuan2014}{}
+21. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
+diabetes susceptibility loci in monozygotic twins. \emph{Nature
+communications} \textbf{5,} 5719 (2014).
+
+\hypertarget{ref-Tang2014}{}
+22. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
+with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
+translational medicine} \textbf{6,} 257ra139 (2014).
+
+\hypertarget{ref-Hong2015}{}
+23. Hong, X. \emph{et al.} Genome-wide association study identifies
+peanut allergy-specific loci and evidence of epigenetic mediation in US
+children. \emph{Nature communications} \textbf{6,} 6304 (2015).
+
+\hypertarget{ref-Claussnitzer2015}{}
+24. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+Adipocyte Browning in Humans. \emph{The New England journal of medicine}
+\textbf{373,} 895--907 (2015).
+
+\hypertarget{ref-Timpson2011}{}
+25. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
+index: elucidating direction of causation through reciprocal Mendelian
+randomization. \emph{International journal of obesity (2005)}
+\textbf{35,} 300--8 (2011).
+
+\hypertarget{ref-Shakhbazov2016}{}
+26. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
+278 (2016).
+
+\hypertarget{ref-Pierce2013}{}
+27. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+randomization studies: subsample and 2-sample instrumental variable
+estimators. \emph{American journal of epidemiology} \textbf{178,}
+1177--84 (2013).
+
+\hypertarget{ref-Hemani2016}{}
+28. Hemani, G. \emph{et al.} MR-Base: a platform for systematic causal
+inference across the phenome using billions of genetic associations.
+\emph{BioRxiv} \textbf{10.1101/07,} (2016).
+
+\hypertarget{ref-Ashenfelter1994}{}
+29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+to Schooling from a New Sample of Twins. \emph{The American Economic
+Review} \textbf{84,} 1157--1173 (1994).
+
+\hypertarget{ref-Nagarajan2013}{}
+30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+inference. \emph{PloS one} \textbf{8,} e80735 (2013).
+
+\hypertarget{ref-Shpitser2010}{}
+31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+covariate adjustment for estimating causal effects. \emph{Proceedings of
+the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
+(UAI-10)} 527--536 (2010).
+
+\hypertarget{ref-Wang2015}{}
+32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
+
+\hypertarget{ref-Lagani2015}{}
+33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+I. in \emph{Uncertainty in biology: A computational modeling approach}
+47 (Springer, 2015). at
+\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+
+\hypertarget{ref-RCoreTeam2015}{}
+34. R Core Team. R: A Language and Environment for Statistical
+Computing. (2015). at
+\textless{}\url{https://www.r-project.org/}\textgreater{}
+
+\hypertarget{ref-Sterne2001}{}
+35. Sterne, J. A. C., Cox, D. R. \& Smith, G. D. Sifting the
+evidence---what's wrong with significance tests?Another comment on the
+role of statistical methods. \emph{BMJ} \textbf{322,} (2001).
+
+\hypertarget{ref-Henningsen2007}{}
+36. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. \emph{Journal of Statistical
+Software} \textbf{23,} 1--40 (2007).
+
+\hypertarget{ref-Steiger1980}{}
+37. Steiger, J. H. Tests for comparing elements of a correlation matrix.
+\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
+
+\hypertarget{ref-Revelle2015}{}
+38. Revelle, W. psych: Procedures for Psychological, Psychometric, and
+Personality Research. (2015). at
+\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
+
+\hypertarget{ref-Bird2002}{}
+39. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+\& development} \textbf{16,} 6--21 (2002).
+
 \hypertarget{ref-Fewell2007}{}
-13. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
+40. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
 residual and unmeasured confounding in epidemiologic studies: a
@@
 
-\hypertarget{ref-LeCessie2012}{}
-14. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
-Vandenbroucke, J. P. Quantification of bias in direct effects estimates
-due to different types of measurement error in the mediator.
-\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
-
-\hypertarget{ref-Blakely2013}{}
-15. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
-mediator matters when estimating indirect effects. \emph{Journal of
-epidemiology and community health} \textbf{67,} 458--66 (2013).
-
-\hypertarget{ref-Schadt2005}{}
-16. Schadt, E. E. \emph{et al.} An integrative genomics approach to
-infer causal associations between gene expression and disease.
-\emph{Nature Genetics} \textbf{37,} 710--717 (2005).
-
-\hypertarget{ref-Millstein2016}{}
-17. Millstein, J. cit: Causal Inference Test. R package version 1.9.
-(2016). at
-\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
-
-\hypertarget{ref-Koestler2014}{}
-18. Koestler, D. C. \emph{et al.} Integrative genomic analysis
-identifies epigenetic marks that mediate genetic risk for epithelial
-ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
-
-\hypertarget{ref-Liu2013}{}
-19. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
-methylation as an intermediary of genetic risk in rheumatoid arthritis.
-\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).
-
-\hypertarget{ref-Yuan2014}{}
-20. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
-diabetes susceptibility loci in monozygotic twins. \emph{Nature
-communications} \textbf{5,} 5719 (2014).
-
-\hypertarget{ref-Tang2014}{}
-21. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
-with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
-translational medicine} \textbf{6,} 257ra139 (2014).
-
-\hypertarget{ref-Hong2015}{}
-22. Hong, X. \emph{et al.} Genome-wide association study identifies
-peanut allergy-specific loci and evidence of epigenetic mediation in US
-children. \emph{Nature communications} \textbf{6,} 6304 (2015).
-
-\hypertarget{ref-Claussnitzer2015}{}
-23. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
-Adipocyte Browning in Humans. \emph{The New England journal of medicine}
-\textbf{373,} 895--907 (2015).
-
-\hypertarget{ref-Timpson2011}{}
-24. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
-index: elucidating direction of causation through reciprocal Mendelian
-randomization. \emph{International journal of obesity (2005)}
-\textbf{35,} 300--8 (2011).
-
-\hypertarget{ref-Shakhbazov2016}{}
-25. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
-and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
-278 (2016).
-
-\hypertarget{ref-Pierce2013}{}
-26. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
-randomization studies: subsample and 2-sample instrumental variable
-estimators. \emph{American journal of epidemiology} \textbf{178,}
-1177--84 (2013).
-
-\hypertarget{ref-Ashenfelter1994}{}
-27. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
-to Schooling from a New Sample of Twins. \emph{The American Economic
-Review} \textbf{84,} 1157--1173 (1994).
-
-\hypertarget{ref-Nagarajan2013}{}
-28. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
-inference. \emph{PloS one} \textbf{8,} e80735 (2013).
-
-\hypertarget{ref-Shpitser2010}{}
-29. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
-covariate adjustment for estimating causal effects. \emph{Proceedings of
-the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
-(UAI-10)} 527--536 (2010).
-
-\hypertarget{ref-Wang2015}{}
-30. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
-gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
-
-\hypertarget{ref-Lagani2015}{}
-31. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
-I. in \emph{Uncertainty in biology: A computational modeling approach}
-47 (Springer, 2015). at
-\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
-
-\hypertarget{ref-RCoreTeam2015}{}
-32. R Core Team. R: A Language and Environment for Statistical
-Computing. (2015). at
-\textless{}\url{https://www.r-project.org/}\textgreater{}
-
-\hypertarget{ref-Henningsen2007}{}
-33. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. \emph{Journal of Statistical
-Software} \textbf{23,} 1--40 (2007).
-
-\hypertarget{ref-Steiger1980}{}
-34. Steiger, J. H. Tests for comparing elements of a correlation matrix.
-\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
-
-\hypertarget{ref-Revelle2015}{}
-35. Revelle, W. psych: Procedures for Psychological, Psychometric, and
-Personality Research. (2015). at
-\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
-
-\hypertarget{ref-Bird2002}{}
-36. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
-\& development} \textbf{16,} 6--21 (2002).
+\hypertarget{ref-Bose2014}{}
+41. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
+methylation measurement using technical replicates: the Atherosclerosis
+Risk In Communities (ARIC) Study. \emph{BMC Bioinformatics} \textbf{15,}
+312 (2014).
+
+\hypertarget{ref-Bryant2011}{}
+42. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
+Biological Variability in a Microarray Experiment but Both Are
+Outweighed by Changes Induced by Stimulation. \emph{PLoS ONE}
+\textbf{6,} e19556 (2011).
+
+\hypertarget{ref-Richmond2014}{}
+43. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
+between Child Adiposity and Physical Activity Levels: A Mendelian
+Randomization Analysis. \emph{PLoS Medicine} \textbf{11,} e1001618
+(2014).
+
+\hypertarget{ref-Bowden2015}{}
+44. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
+with invalid instruments: effect estimation and bias detection through
+Egger regression. \emph{International Journal of Epidemiology}
+\textbf{In press,} (2015).
+
+\hypertarget{ref-Bowden2016b}{}
+45. Bowden, J., Davey Smith, G., Haycock, P. C. \& Burgess, S.
+Consistent Estimation in Mendelian Randomization with Some Invalid
+Instruments Using a Weighted Median Estimator. \emph{Genetic
+Epidemiology} \textbf{40,} 304--314 (2016).
 
 \hypertarget{ref-Relton2012}{}
-37. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+46. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-38. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+47. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 
+\hypertarget{ref-Varbo2015}{}
+48. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+cholesterol, and blood pressure as mediators from obesity to ischemic
+heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).
+
 \hypertarget{ref-Burgess2015}{}
-39. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+49. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 
-\hypertarget{ref-Varbo2015}{}
-40. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
-cholesterol, and blood pressure as mediators from obesity to ischemic
-heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).
-
-\hypertarget{ref-Bowden2015}{}
-41. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
-with invalid instruments: effect estimation and bias detection through
-Egger regression. \emph{International Journal of Epidemiology}
-\textbf{In press,} (2015).
-
 \end{document}

### 2017-01-03 scripts/mr_directionality_analysis.R

@@ -2,6 +2,7 @@
 
 suppressPackageStartupMessages(library(dplyr))
 suppressPackageStartupMessages(library(ggplot2))
+suppressPackageStartupMessages(library(ggExtra))
 suppressPackageStartupMessages(library(tidyr))
 suppressPackageStartupMessages(library(psych))
 suppressPackageStartupMessages(library(lattice))
@@ -881,8 +882,6 @@ example_sensitivity <- mr_steiger(
 	screen=list(z = 55, x = -60, y = 3), bty="n"
 )
 
-example_sensitivity
-
 p1 <- example_sensitivity$sensitivity_plot
 
 p2 <- ggplot(sensitivity_parameters, aes(y=sensitivity, x=r_xy, group=factor(r_gx^2))) +
@@ -890,7 +889,7 @@ p2 <- ggplot(sensitivity_parameters, aes(y=sensitivity, x=r_xy, group=factor(r_g
 	geom_line(aes(colour=factor(r_gx^2))) +
 	labs(x=expression(rho[xy]), y=expression(V[z<0]/V[z>=0]), colour=expression(rho[gx]^2)) +
 	scale_colour_brewer(type="qual")
-p2 + scale_y_log10()
+
 grid.arrange(
 	textGrob("a)", x=unit(0.1, "npc")), textGrob("b)", x=unit(0.1, "npc")),
 	p1, p2, 

### 2017-01-25 docs/manuscript.Rmd
updates

@@
 A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses.
-Here we show that when measurement error is present, commonly-used tools predicated on this approach are susceptible to making erroneous inference of both the existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
-We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
-This extension is less susceptible to measurement error than mediation-based analyses.
+Here we show that when measurement error is present, widely used tools predicated on this approach, such as the causal inference test (CIT) are susceptible to making erroneous inference of both the existence of causality and the causal direction.
+Increasing sample size for these tests, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
+We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is relatively robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
+This extension is less susceptible to measurement error than mediation-based analyses and only requires GWAS summary data.
 We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms.
@@
 In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015].
-However, potential pitfalls still persist and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
+However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
 
-Non-differential measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value.
+Measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value.
+This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement.
 Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
-Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a].
+Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a].
 Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].
@@
 The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].
+Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.
 
-Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.
 An established study design that can provide information about causality is randomisation.
@@
 
-MR can operate on the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner.
+MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner.
 Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure.

### 2017-01-25 scripts/mr_directionality_analysis.R

@@ -191,19 +191,20 @@ get_cit_direction <- function(p1, p2, thresh)
 parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.05)
 parameters$cit_correct_direction <- parameters$cit_res == 1
 
-# Get the p-value for the CIT direction
-# ab < 0.05 & ba > 0.05 = correct
-# ab > 0.05 & ba < 0.05 = incorrect
-# ab < 0.05 & ba < 0.05 = no call
-# ab > 0.05 & ba > 0.05 = incorrect
 
 
 parameters$cit_p <- parameters$cit_AB
 parameters$cit_p[parameters$cit_res == 2] <- parameters$cit_BA[parameters$cit_res == 2]
 parameters$cit_p[parameters$cit_res == 3] <- 0.5
-parameters$cit_p[parameters$cit_res == 4] <- 0.01
+parameters$cit_p[parameters$cit_res == 4] <- 0.5
 
 # Do we classify 4 as being significant wrong causal direction, or no causal inference?
+
+# Get the p-value for the CIT direction
+# ab < 0.05 & ba > 0.05 = correct
+# ab > 0.05 & ba < 0.05 = incorrect
+# ab < 0.05 & ba < 0.05 = no call
+# ab > 0.05 & ba > 0.05 = no call
 
 
 parameters$cit_p <- -log10(parameters$cit_p)

### 2017-01-26 scripts/mr_directionality_with_confounding.R
added extra models for confounding

@@ -1,0 +1,203 @@
+library(reshape2)
+library(ggplot2)
+library(plyr)
+library(cit)
+
+#' 
+#' Calculate fitted values directly
+#'
+#' @param y Response variable
+#' @param  x Predictor
+#' @export
+#' @return Array
+getFittedVals <- function(y, x)
+{
+	n <- length(x)
+	bhat <- cov(y, x) / var(x)
+	ahat <- mean(y) - bhat * mean(x)
+	fitted <- ahat + x * bhat
+	return(fitted)
+}
+
+#' Get residual values from OLS
+#'
+#'
+#' @param y Response variable
+#' @param x Predictor
+#' @export
+#' @return Array
+getResiduals <- function(y, x)
+{
+	fitted <- getFittedVals(y, x)
+	return(y - fitted)
+}
+
+# 		zA_adj <- zA
+# 		zB_adj <- getResiduals(zB, zA)
+# 	}
+# 	return(list(zA = zA_adj, zB = zB_adj))
+# }
+
+
+
+#' Get pval directly
+#' @param y Response variable
+#' @param x Predictor
+#' @export
+#' @return p value
+getPval <- function(y, x)
+{
+	n <- length(x)
+	fitted <- getFittedVals(y, x)
+	residuals <- y - fitted
+	SSR <- sum((residuals - mean(residuals))^2)
+	SSF <- sum((fitted - mean(fitted))^2)
+	Fval <- (SSF) / (SSR/(n-2))
+	pval <- pf(Fval, 1, n-2, lowe=F)
+	return(pval)
+}
+
+
+makeProxy <- function(x, noise, bias)
+{
+	y <- x*bias + rnorm(length(x), 0, sd=noise)
+	return(y)
+}
+
+testProxy <- function(x, Noise=seq(0.1, 1, by=0.1), Bias=seq(0.1, 1, by=0.1))
+{
+	require(lattice)
+	dat <- expand.grid(noise=Noise, bias=Bias, rsq = NA)
+	for(i in 1:nrow(dat))
+	{
+		dat$rsq[i] <- cor(x, makeProxy(x, dat$noise[i], dat$bias[i]))^2
+	}
+	print(wireframe(rsq ~ noise*bias, dat, drape=TRUE))
+	return(dat)	
+}
+
+
+make_geno <- function(n, p)
+{
+	stopifnot(all(p >= 0 & p <= 1))
[…]

### 2017-01-26 scripts/mr_directionality_with_confounding.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+#PBS -N me
+#PBS -o job_reports/me-output
+#PBS -e job_reports/me-error
+#PBS -l walltime=12:00:00
+#PBS -t 37-100
+#PBS -l nodes=1:ppn=1
+#PBS -S /bin/bash
+
+set -e
+
+echo "Running on ${HOSTNAME}"
+
+if [ -n "${1}" ]; then
+  echo "${1}"
+  PBS_ARRAYID=${1}
+fi
+
+i=${PBS_ARRAYID}
+splits=1944
+
+Rscript \
+	~/repo/cit_measurement_error/scripts/mr_directionality_with_confounding.R \
+	${i} \
+	${splits}

### 2017-01-26 scripts/mr_directionality_with_confounding_aggregate.R

@@ -1,0 +1,21 @@
+library(dplyr)
+
+splits <- 100
+savefile <- "~/repo/cit_measurement_error/results/mr_directionality_20170126.RData"
+
+l <- list()
+for(i in 1:splits)
+{
+	cat(i, "\n")
+	filename <- paste0("~/repo/cit_measurement_error/scratch/resultswc_", i, ".RData")
+	if(file.exists(filename))
+	{
+		load(filename)
+		l[[i]] <- parameters
+	} else {
+		message("Missing ", filename)
+	}
+}
+
+parameters <- rbind_all(l)
+save(parameters, file = savefile)

### 2017-01-26 docs/manuscript.Rmd

@@
 ### Abstract
+
+Genetic associations have the desirable property of ensuring the causal direction is from genotype to trait.
+Mediation based approaches such as the causal inference test (CIT) use this property to infer the causal direction between two traits, while Mendelian randomization uses the property to estimate the causal effect size if knowledge of the
+
+ by evaluating if the genetic association on the consequential trait is abbrogated by adjusting for the causal trait.
+Mendelian randomization, by contrast assumes a priori knowledge of which of the two traits the genetic variant is acting on primarily, and then creates a proxy to estimate its causal effect on
 

### 2017-01-27 scripts/mr_directionality_analysis.R
updated graphs

@@ -10,6 +10,9 @@ suppressPackageStartupMessages(library(latex2exp))
 suppressPackageStartupMessages(library(knitr))
 suppressPackageStartupMessages(library(gridExtra))
 suppressPackageStartupMessages(library(grid))
+suppressPackageStartupMessages(library(network))
+suppressPackageStartupMessages(library(sna))
+suppressPackageStartupMessages(library(ggnetwork))
 
 
 get_p_from_r2n <- function(r2, n)
@@ -147,9 +150,16 @@ mr_wald_ratio <- function(b_exp, b_out, se_exp, se_out, parameters)
 
 ## ---- read_data ----
 
+load("~/repo/cit_measurement_error/results/mr_directionality_20170126.RData")
+parametersc <- parameters
+parametersc$model <- "Non-causal"
 load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
+parameters$model <- "Causal"
+parameters <- rbind(parameters, parametersc)
 
 parameters <- subset(parameters, r_ab != 1)
+index_causal <- parameters$model == "Causal"
+index_ncausal <- parameters$model == "Non-causal"
 
 parameters$p_az <- -log10(parameters$p_az)
 parameters$p_bz <- -log10(parameters$p_bz)
@@ -166,7 +176,8 @@ parameters$rhs_lhs_diff <- parameters$lhs - parameters$rhs
 a <- r.test(n=parameters$n, r12=parameters$cor_zap, r13=parameters$cor_zbp, r23=parameters$cor_abp)
 parameters$cortest_t <- a$t
 parameters$cortest_p <- -log10(a$p)
-parameters$cortest_correct_direction <- sign(parameters$cortest_t) == 1
+parameters$cortest_correct_direction[index_causal] <- sign(parameters$cortest_t[index_causal]) == 1
+parameters$cortest_correct_direction[index_ncausal] <- FALSE
 parameters$cortest_correct_direction[is.na(parameters$cortest_correct_direction)] <- FALSE
 
 
@@ -189,21 +200,28 @@ get_cit_direction <- function(p1, p2, thresh)
 }
 
 parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.05)
[…]

### 2017-01-31 docs/manuscript.Rmd
updated

@@
 	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown.
-If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables.
+If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables.
+The causal inference test attempts to delineate between them.
 a) Methylation causes gene expression.

### 2017-01-31 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{03 January 2017}
+  \date{27 January 2017}
 
@@
 
+Genetic associations have the desirable property of ensuring the causal
+direction is from genotype to trait. Mediation based approaches such as
+the causal inference test (CIT) use this property to infer the causal
+direction between two traits, while Mendelian randomization uses the
+property to estimate the causal effect size if knowledge of the
+
+by evaluating if the genetic association on the consequential trait is
+abbrogated by adjusting for the causal trait. Mendelian randomization,
+by contrast assumes a priori knowledge of which of the two traits the
+genetic variant is acting on primarily, and then creates a proxy to
+estimate its causal effect on
+
 A common approach for inferring causal relationships between two traits
 is to use genetic instruments in mediation-based analyses. Here we show
-that when measurement error is present, commonly-used tools predicated
-on this approach are susceptible to making erroneous inference of both
-the existence of causality and the causal direction, and that increasing
-sample size, rather than reducing bias, can often have the effect of
-increasing certainty in the wrong answer. We demonstrate that Mendelian
-randomization (MR) is a technique for causal inference that is robust to
-measurement error, and here we extend the method with new sensitivity
-analyses to estimate the direction of causality between correlated
-variables even when the biology of the instrument is unknown. This
-extension is less susceptible to measurement error than mediation-based
-analyses. We use it to demonstrate empirically that inferring the
-direction of causality between correleted DNA methylation and gene
-expression levels is highly susceptible to differences in measurement
-error between the two assaying platforms. Other 'omic platforms are
-likely to be similarly susceptible, necessitating the use of sensitivity
-analyses to evaluate reliability.
+that when measurement error is present, widely used tools predicated on
+this approach, such as the causal inference test (CIT) are susceptible
+to making erroneous inference of both the existence of causality and the
+causal direction. Increasing sample size for these tests, rather than
+reducing bias, can often have the effect of increasing certainty in the
+wrong answer. We demonstrate that Mendelian randomization (MR) is a
+technique for causal inference that is relatively robust to measurement
+error, and here we extend the method with new sensitivity analyses to
+estimate the direction of causality between correlated variables even
+when the biology of the instrument is unknown. This extension is less
+susceptible to measurement error than mediation-based analyses and only
+requires GWAS summary data. We use it to demonstrate empirically that
+inferring the direction of causality between correleted DNA methylation
+and gene expression levels is highly susceptible to differences in
+measurement error between the two assaying platforms. Other 'omic
+platforms are likely to be similarly susceptible, necessitating the use
+of sensitivity analyses to evaluate reliability.
 
@@
 applied in several different guises\textsuperscript{3--6}. However,
-potential pitfalls still persist and one that is often neglected is the
+potential pitfalls remain and one that is often neglected is the
 influence of non-differential measurement error on the reliability of
@@
 
-Non-differential measurement error (or observational error, henceforth
-referred to simply as measurement error) is the difference between the
-measured value of a quantity and its true value. Such variability can
-arise through a whole plethora of mechanisms, which are often unique to
-the study design and difficult to avoid\textsuperscript{7,8}. Array
-technology is now commonly used to obtain high throughput phenotyping at
-low cost, but they come with the problem of having imperfect resolution,
-for instance methylation levels as measured by the Illumina450k chip are
-prone to have some amount of noise around the true
-value\textsuperscript{9,10}. Relatedly, if the measurement of biological
-interest is the methylation level in a T cell, then measurement error of
-this value can be introduced by using methylation levels from whole
-blood samples because the measured value will be an assay of many cell
-types\textsuperscript{11}.
+Measurement error (or observational error, henceforth referred to simply
+as measurement error) is the difference between the measured value of a
+quantity and its true value. This study focuses specifically on
+non-differential measurement error where all strata of a measured
+variable have the same error rate, which can manifest as changes in
+scale or levels of imprecision (noise) of the measurement. Such
+variability can arise through a whole plethora of mechanisms, which are
+often unique to the study design and difficult to
+avoid\textsuperscript{7,8}. Array technology is now commonly used to
+obtain high throughput phenotyping at low cost, but comes with the
+problem of having imperfect resolution, for instance methylation levels
+as measured by the Illumina450k chip are prone to have some amount of
+noise around the true value due to imperfect
+sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
+biological interest is the methylation level in a T cell, then
+measurement error of this value can be introduced by using methylation
+levels from whole blood samples because the measured value will be an
+assay of many cell types\textsuperscript{11}.
 
@@
 list\textsuperscript{8}, and its impact has been explored in the
-epidemiological literature extensively\textsuperscript{13,14}.
-
-Given the near-ubiquitous presence of measurement error in phenomic data
-it is vital to understand its impact on the tools we use for causal
-inference. An established study design that can provide information
-about causality is randomisation. Given the hypothesis that trait A
-(henceforth referred to as the exposure) is causally related to trait B
-(henceforth referred to as the outcome), randomisation can be employed
-to assess the causal nature of the association by randomly splitting the
-sample into two groups, subjecting one group to the exposure and leaving
-the other as a control. The association between the exposure and the
-outcome in this setting provides a robust estimate of the causal
-relationship. This provides the theoretical basis behind randomised
-control trials, but in practice randomisation is often impossible to
-implement in an experimental context due to cost, scale or inability to
-manipulate the exposure. The principle, however, can be employed in
-extant observational data through the use of genetic variants associated
-with the exposure (instruments), where the inheritance of an allele
-serves as a random lifetime allocation of differential exposure
+epidemiological literature extensively\textsuperscript{13,14}. Given the
+near-ubiquitous presence of measurement error in phenomic data it is
+vital to understand its impact on the tools we use for causal inference.
+
+An established study design that can provide information about causality
+is randomisation. Given the hypothesis that trait A (henceforth referred
+to as the exposure) is causally related to trait B (henceforth referred
+to as the outcome), randomisation can be employed to assess the causal
+nature of the association by randomly splitting the sample into two
+groups, subjecting one group to the exposure and leaving the other as a
+control. The association between the exposure and the outcome in this
+setting provides a robust estimate of the causal relationship. This
+provides the theoretical basis behind randomised control trials, but in
+practice randomisation is often impossible to implement in an
+experimental context due to cost, scale or inability to manipulate the
+exposure. The principle, however, can be employed in extant
+observational data through the use of genetic variants associated with
+the exposure (instruments), where the inheritance of an allele serves as
+a random lifetime allocation of differential exposure
 levels\textsuperscript{15,16}. Two statistical approaches to exploiting
@@
 
-MR can operate on the same data - phenotypic measures of the exposure
+MR can be applied to the same data - phenotypic measures of the exposure
 and the outcome variables and a genetic instrument for the exposure -

### 2017-01-31 scripts/cit_tests.R

@@ -1,0 +1,253 @@
+## ---- cit_tests ----
+
+suppressWarnings(suppressPackageStartupMessages(library(network)))
+suppressWarnings(suppressPackageStartupMessages(library(sna)))
+suppressWarnings(suppressPackageStartupMessages(library(ggnetwork)))
+suppressWarnings(suppressPackageStartupMessages(library(gridExtra)))
+suppressWarnings(suppressPackageStartupMessages(library(dplyr)))
+suppressWarnings(suppressPackageStartupMessages(library(tidyr)))
+
+nid <- 10000
+l <- m <- list()
+
+# 1
+
+mat <- matrix(c(
+	0,0,0,
+	1,0,0,
+	0,1,0
+), 3, 3)
+
+n <- network(mat, directed=TRUE)
+n %v% "what" <- c("G", "A", "B")
+l[[1]] <- ggplot(ggnetwork(n, layout="circle", arrow.gap=0.05), aes(x = x, y = y, xend = xend, yend = yend)) +
+  # geom_nodes(size = 11, aes(color=what)) +
+  geom_edges(arrow = arrow(type = "closed")) +
+  geom_nodelabel(aes(label = what), fontface = "bold") +
+  theme_blank() +
+  theme(legend.position="none") +
+  labs(title="Model 1")
+
+g <- rnorm(nid)
+a <- g + rnorm(nid)
+b <- a + rnorm(nid)
+m[[1]] <- rbind(
+	cit.cp(L=g, G=a, T=b),
+	cit.cp(L=g, G=b, T=a)
+)
+m[[1]] <- as.data.frame(m[[1]])
+m[[1]]$model <- 1
+m[[1]]$direction <- c("Testing A to B", "Testing B to A")
+
+
+# 2
+
+mat <- matrix(c(
+	0,0,0,
+	0,0,1,
+	1,0,0
+), 3, 3)
+
+n <- network(mat, directed=TRUE)
+n %v% "what" <- c("G", "A", "B")
+l[[2]] <- ggplot(ggnetwork(n, layout="circle", arrow.gap=0.05), aes(x = x, y = y, xend = xend, yend = yend)) +
+  # geom_nodes(size = 11, aes(color=what)) +
+  geom_edges(arrow = arrow(type = "closed")) +
[…]

### 2017-01-31 scripts/mr_directionality_analysis.R

@@ -275,7 +275,7 @@ levels(dat$eval) <- c("Correct causal model", "Incorrect causal model")
 
 ggplot(subset(dat, n %in% c(100, 1000, 10000)), aes(x=sqrt(rsq),y=-log10(pval))) +
 geom_point(aes(colour=eval)) +
-labs(y=expression(-log[10]*p), x=expression(cor(x, x[O])), colour="") +
+labs(y=expression(-log[10]*p), x=expression(rho[xx[o]]), colour="") +
 facet_grid(n ~ ., scale="free_y")
 
 
@@ -393,7 +393,7 @@ p4 <- ggplot(ggnetwork(n, layout="circle", arrow.gap=0.05), aes(x = x, y = y, xe
 
 cowplot::plot_grid(
 	gridExtra::arrangeGrob(
-		grobs=list(p3, p1, legend, p4, p2), 
+		grobs=list(p4, p1, legend, p3, p2), 
 		ncol=3, 
 		nrow=2, 
 		heights=c(4,4),

### 2017-01-31 scripts/mr_directionality_with_confounding.R

@@ -189,7 +189,15 @@ for(i in 1:nrow(parameters))
 	message(i, " of ", nrow(parameters))
 	dat <- with(parameters[i,], make_system(n, p, r_ab, r_za, noisea, noiseb))
 	parameters$cit_AB[i] <- cit.cp(dat$Z, dat$Ap, dat$Bp)[1]
+	parameters$cit_AB1[i] <- cit.cp(dat$Z, dat$Ap, dat$Bp)[2]
+	parameters$cit_AB2[i] <- cit.cp(dat$Z, dat$Ap, dat$Bp)[3]
+	parameters$cit_AB3[i] <- cit.cp(dat$Z, dat$Ap, dat$Bp)[4]
+	parameters$cit_AB4[i] <- cit.cp(dat$Z, dat$Ap, dat$Bp)[5]
 	parameters$cit_BA[i] <- cit.cp(dat$Z, dat$Bp, dat$Ap)[1]
+	parameters$cit_BA1[i] <- cit.cp(dat$Z, dat$Bp, dat$Ap)[2]
+	parameters$cit_BA2[i] <- cit.cp(dat$Z, dat$Bp, dat$Ap)[3]
+	parameters$cit_BA3[i] <- cit.cp(dat$Z, dat$Bp, dat$Ap)[4]
+	parameters$cit_BA4[i] <- cit.cp(dat$Z, dat$Bp, dat$Ap)[5]
 	parameters$cor_aap[i] <- cor(dat$A, dat$Ap)
 	parameters$cor_bbp[i] <- cor(dat$B, dat$Bp)
 	parameters$cor_abp[i] <- cor(dat$Ap, dat$Bp)

### 2017-01-31 scripts/mr_directionality_with_confounding.sh
updated:

@@ -4,7 +4,7 @@
 #PBS -o job_reports/me-output
 #PBS -e job_reports/me-error
 #PBS -l walltime=12:00:00
-#PBS -t 37-100
+#PBS -t 17-100
 #PBS -l nodes=1:ppn=1
 #PBS -S /bin/bash
 

### 2017-02-01 docs/manuscript.Rmd
updated

--- 2017-01-25 docs/manuscript.Rmd
@@
 	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown.
-If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables.
+If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables.
+The causal inference test (CIT) attempts to delineate between them.
 a) Methylation causes gene expression.
@@
 Faded symbols represent the measured values whereas solid symbols represent the true values.
-b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
+b) The same as in A, except the causal direction is from gene expression to DNA methylation.
+c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.
+",
 	display=FALSE
@@
 	"cit_mr_comparison_figure", 
-	caption="Outcomes were simulated to be causally influenced by exposures, with varying degrees of measurement error applied to both.
+	caption="a) Outcomes were simulated to be causally influenced by exposures as shown in the graph, with varying degrees of measurement error applied to both.
 CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
@@
 Rows of graphs represent the sample size used in the simulations.
-For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.",
+For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.
+b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$.
+MR is unable to identify this model, so any significant associations are deemed to be incorrect.
+Outcome 3 denotes evidence for the correct model for the CIT method.",
 	display=FALSE
@@
 
-A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses.
-Here we show that when measurement error is present, widely used tools predicated on this approach, such as the causal inference test (CIT) are susceptible to making erroneous inference of both the existence of causality and the causal direction.
-Increasing sample size for these tests, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer.
-We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is relatively robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown.
-This extension is less susceptible to measurement error than mediation-based analyses and only requires GWAS summary data.
-We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms.
-Other 'omic platforms are likely to be similarly susceptible, necessitating the use of sensitivity analyses to evaluate reliability.
+Inference of the causal relationship between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others.
+We show here, however, that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
+Here we develop an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages.
+First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; and third, it can be performed using only summary level data from genome-wide association studies.
+We apply the method to infer the causal direction between DNA methylation and gene expression levels.
+Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but that such high level conclusions are susceptible to differential measurement error between the platforms.
+We emphasise that the use of several different methods is important to triangulate reliable conclusions.
 
@@
 MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner.
-Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure.
-The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
+Here the SNP is used as a surrogate for the exposure.
+Assuming the SNP associates with the outcome only through the exposure, the causal effect of the exposure on the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
+Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016].
 
@@
 In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.
-Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.
+The CIT can also distinguish between several putative models of association between the traits and the SNP [@Millstein2009; @Millstein2016].
+Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.
 
-Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e.
+Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e.
 for which of the two variables is the SNP a valid instrument?
 `r cfig("dag_figure")`).
-By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
+By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
 Such a situation can arise in many scenarios.
 Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
-For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP).
-It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI.
-If multiple instruments are available for an hypothesised exposure then this problem can be ameliorated [@DaveySmithHemani2014], however this is not always the case, especially in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known.
+If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014].
+But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known.
 For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?
@@
 MR does however have some important advantages over the mediation-based approaches.
-First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size.
-Second, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
+The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
-Third, MR is substantially more robust to there being measurement error in the exposure variable.
+Additionally, when MR assumptions are satisfied the method is substantially more robust to there being measurement error in the exposure variable.
 Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
 
-Using theory and simulations we show how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method.
-We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result to measurement error.
+Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method.
+We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error.
 This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
@@
 In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
-To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for the incorrect model of $y$ causing $x$.
+To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$, and for the direction model of $y$ causing $x$.
 For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:
 
-- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
-- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
-- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then an alternative causal model is accepted
-- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then the $x \rightarrow y$ model is accepted
+- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the $y \rightarrow x$ model is accepted
+- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no inference is made
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).
 
-For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to rely on p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds [@Sterne2001].
+For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we stress that for real analyses it is not good practice to rely on p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds [@Sterne2001].
 
@@
 
-In what follows we assume no horizontal pleiotropy, such that $g$ only has an effect on one of the variables through the other variable.
+The method that we will now describe is able to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$.
+Unlike the CIT framework, this approach cannot infer if the true model is $x \leftarrow g \rightarrow y$.
+We also assume all genetic effects are additive.
+
 To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$.
@@
 
-The Steiger test for testing if $x \rightarrow y$ is based on evaluating $\\rho_{gx} > \\rho_{gy}$.
-However, $\\rho_{gx}$ and $\\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely.
-If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\\rho_{g,x_o} > \\rho_{g,y_o}$.
-To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we calculate the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
-We also calculate the measurement error values which are not compatible.
+The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$.
+However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely.
+If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$.
+As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$).
+It is possible to simulate what the inferred causal direction would be for all values within these bounds.
+To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
+We also do the same for the measurement error values which are not compatible.
 The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction.
@@
 Simulations were conducted by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$.
-In all models $x$ causes $y$ and $g$ is an instrument for $x$.
-Each variable was simulated such that:
+One of two models are simulated, the "causal model" where $x$ causes $y$ and $g$ is an instrument for $x$; or the "non-causal model" where $g$ influences a confounder $u$ which in turn causes both $x$ and $y$.
+Here $x$ and $y$ are correlated but not causally related.
+Each variable in the causal model was simulated such that:
 
@@
 where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.
+Similarly in the non-causal model:
+
+$$
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+u & = \alpha_g + \beta_g g + \epsilon_g \\
+x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+$$
 
@@
 cor(g, x)^2 & = 0.1 \\
+cor(g, u)^2 & = 0.1 \\
 cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
@@
 Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables.
-Similar patterns of results were obtained for different values of $cor(g, x)$.
+Similar patterns of results were obtained for different values of $cor(g, x)$ and $cor(g, u)$.
 
@@
 Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$.
-For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$, and $\epsilon_{mx}$ represents the imprecision in the measurement of $x$.
+For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
 Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that
@@
 The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error is introduced it can be shown using basic covariance properties (Appendix 1) that
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
 
@@
 $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
-Note that linear transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions.
+Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
 
@@
 We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
-The performance of MR in detecting an association is unrelated to measurement error in.
-Measurement error in the outcome does not induce a substantive difference in performance between CIT and MR.
+When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
+Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
 
@@
 We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect.
-`r cfig("cit_mr_comparison_figure")` shows that when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
-However, the CIT is in general much more fallable to reporting a robust causal association for the wrong direction of causality.
+Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
+
+`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
+However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality.
 When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate this confounding model as being causal.
 
@@
 
-\newpage
-
 ## Appendix 2

### 2017-02-01 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{27 January 2017}
+  \date{01 February 2017}
 
@@
 
-Genetic associations have the desirable property of ensuring the causal
-direction is from genotype to trait. Mediation based approaches such as
-the causal inference test (CIT) use this property to infer the causal
-direction between two traits, while Mendelian randomization uses the
-property to estimate the causal effect size if knowledge of the
-
-by evaluating if the genetic association on the consequential trait is
-abbrogated by adjusting for the causal trait. Mendelian randomization,
-by contrast assumes a priori knowledge of which of the two traits the
-genetic variant is acting on primarily, and then creates a proxy to
-estimate its causal effect on
-
-A common approach for inferring causal relationships between two traits
-is to use genetic instruments in mediation-based analyses. Here we show
-that when measurement error is present, widely used tools predicated on
-this approach, such as the causal inference test (CIT) are susceptible
-to making erroneous inference of both the existence of causality and the
-causal direction. Increasing sample size for these tests, rather than
-reducing bias, can often have the effect of increasing certainty in the
-wrong answer. We demonstrate that Mendelian randomization (MR) is a
-technique for causal inference that is relatively robust to measurement
-error, and here we extend the method with new sensitivity analyses to
-estimate the direction of causality between correlated variables even
-when the biology of the instrument is unknown. This extension is less
-susceptible to measurement error than mediation-based analyses and only
-requires GWAS summary data. We use it to demonstrate empirically that
-inferring the direction of causality between correleted DNA methylation
-and gene expression levels is highly susceptible to differences in
-measurement error between the two assaying platforms. Other 'omic
-platforms are likely to be similarly susceptible, necessitating the use
-of sensitivity analyses to evaluate reliability.
+Inference of the causal relationship between two traits can be achieved
+by combining genetic associations with a mediation-based approach, as is
+done in the causal inference test (CIT) and others. We show here,
+however, that measurement error in the phenotypes can lead to
+mediation-based approaches inferring the wrong causal direction, and
+that increasing sample sizes has the adverse effect of increasing
+confidence in the wrong answer. Here we develop an extension to
+Mendelian randomisation, a method that uses genetic associations in an
+instrumentation framework, that enables inference of the causal
+direction with some advantages. First, it is less susceptible to bias in
+the presence of measurement error; second, it is more statistically
+efficient; and third, it can be performed using only summary level data
+from genome-wide association studies. We apply the method to infer the
+causal direction between DNA methylation and gene expression levels. Our
+results demonstrate that, in general, DNA methylation is more likely to
+be the causal factor, but that such high level conclusions are
+susceptible to differential measurement error between the platforms. We
+emphasise that the use of several different methods is important to
+triangulate reliable conclusions.
 
@@
 but the genetic instrument is employed in a subtly different manner.
-Here the SNP, which is assumed to have a direct association with the
-exposure, is used as a surrogate for the exposure. The causal effect of
-the exposure and the outcome can be estimated by scaling the association
-between the SNP and the outcome by the association between the SNP and
-the exposure.
+Here the SNP is used as a surrogate for the exposure. Assuming the SNP
+associates with the outcome only through the exposure, the causal effect
+of the exposure on the outcome can be estimated by scaling the
+association between the SNP and the outcome by the association between
+the SNP and the exposure. Though difficult to test empirically, this
+assumption can be relaxed in various ways when multiple instruments are
+available for a putative exposure\textsuperscript{24,25} and a number of
+sensitivity tests are now available to improve
+reliability\textsuperscript{26}.
 
@@
 between two variables by performing the test in both directions and
-choosing the model with the strongest evidence. Such is not the case for
-MR, where in order to infer the direction of causality the instrument
-must be known to influence the exposure directly, associating with the
-outcome only through the exposure.
+choosing the model with the strongest evidence. The CIT can also
+distinguish between several putative models of association between the
+traits and the SNP\textsuperscript{4,18}. Such is not the case for MR,
+where in order to infer the direction of causality between two traits
+the instrument must be known to influence the exposure primarily,
+associating with the outcome only through the exposure.
 
@@
 difficult to determine which of the two variables is subject to the
-direct effect of the SNP (i.e.~for which of the two variables is the SNP
-a valid instrument? Figure 1). By definition, we expect that if the
+primary effect of the SNP (i.e.~for which of the two variables is the
+SNP a valid instrument? Figure 1). By definition, we expect that if the
 association is causal then a SNP for the exposure will be associated
-with the outcome, so if the researcher erroneously uses the SNP as an
-instrument for the outcome then they are likely to see an apparently
-robust causal association of outcome on exposure. Such a situation can
-arise in many scenarios. Genome-wide association studies (GWASs) that
-identify genetic associations for complex traits are, by design,
-hypothesis free and agnostic of genomic function, and it often takes
-years of follow up studies to understand the biological nature of a
-putative GWAS hit\textsuperscript{24}. For example, the same FTO variant
-that is associated with body mass index (BMI) is also associated with
-C-reactive protein (CRP). It is known that this is because BMI causes
-CRP\textsuperscript{25}, but supposing we only had \emph{a priori}
-knowledge that FTO associated with CRP, and we used it to instrument CRP
-in an MR analysis of CRP against BMI, we would mistakenly infer that CRP
-causes BMI. If multiple instruments are available for an hypothesised
-exposure then this problem can be ameliorated\textsuperscript{16},
-however this is not always the case, especially in the case of
-determining causal directions between 'omic measures where typically
-only one cis-acting SNP is known. For example if a DNA methylation probe
-is associated with expression of an adjacent gene, then is a cis-acting
-SNP an instrument for the DNA methylation level, or the gene expression
-level\textsuperscript{26} (Figure 1)?
+with the outcome, such that if the researcher erroneously uses the SNP
+as an instrument for the outcome then they are likely to see an
+apparently robust causal association of outcome on exposure. Such a
+situation can arise in many scenarios. Genome-wide association studies
+(GWASs) that identify genetic associations for complex traits are, by
+design, hypothesis free and agnostic of genomic function, and it often
+takes years of follow up studies to understand the biological nature of
+a putative GWAS hit\textsuperscript{27}. If multiple instruments are
+available for an hypothesised exposure, which is increasingly typical
+for complex traits that are analysed in large GWAS consortia, then
+techniques can be applied to mitigate these issues\textsuperscript{16}.
+But these techniques cannot always be applied in the case of determining
+causal directions between 'omic measures where typically only one
+cis-acting SNP is known. For example if a DNA methylation probe is
+associated with expression of an adjacent gene, then is a cis-acting SNP
+an instrument for the DNA methylation level, or the gene expression
+level\textsuperscript{28} (Figure 1)?
 
 MR does however have some important advantages over the mediation-based
-approaches. First, MR is used as a tool to not just deduce the existance
-of causality, but to also estimate the causal effect size. Second, the
-mediation-based approaches require that the exposure, outcome and
-instrumental variables are all measured in the same data, whereas recent
-extensions to MR circumvent this requirement, allowing causal inference
-to be drawn when exposure variables and outcome variables are measured
-in different samples\textsuperscript{27}. This has the crucial advantage
-of improving statistical power by allowing analysis in much larger
-sample sizes, and dramatically expands the breadth of possible
-phenotypic relationships that can be evaluated\textsuperscript{28}.
-Third, MR is substantially more robust to there being measurement error
-in the exposure variable. Indeed instrumental variable (IV) analysis was
-in part initially introduced as a correction for measurement error in
-the exposure\textsuperscript{29}, whereas it has been noted that both
-classic mediation-based analyses\textsuperscript{13,14,30,31} and
-mediation-based methods that use instrumental
-variables\textsuperscript{32,33} are prone to be unreliable in its
-presence.
+approaches. The mediation-based approaches require that the exposure,
+outcome and instrumental variables are all measured in the same data,
+whereas recent extensions to MR circumvent this requirement, allowing
+causal inference to be drawn when exposure variables and outcome
+variables are measured in different samples\textsuperscript{29}. This
+has the crucial advantage of improving statistical power by allowing
+analysis in much larger sample sizes, and dramatically expands the
+breadth of possible phenotypic relationships that can be
+evaluated\textsuperscript{26}. Additionally, when MR assumptions are
+satisfied the method is substantially more robust to there being
+measurement error in the exposure variable. Indeed instrumental variable
+(IV) analysis was in part initially introduced as a correction for
+measurement error in the exposure\textsuperscript{30}, whereas it has
+been noted that both classic mediation-based
+analyses\textsuperscript{13,14,31,32} and mediation-based methods that
+use instrumental variables\textsuperscript{33,34} are prone to be
+unreliable in its presence.
 
 Using theory and simulations we show how non-differential measurement
-error and measurement imprecision can lead to unreliable causal
-inference in the mediation-based CIT method. We then present an
-extension to MR that allows researchers to ascertain the causal
-direction of an association even when the biology of the instruments are
-not fully understood, and also a method to evaluate the sensitivity of
-the result to measurement error. This improves the utility of MR in
-cases where mediation based methods might have otherwise been used
+error in phenotypes can lead to unreliable causal inference in the
+mediation-based CIT method. We then present an extension to MR that
+allows researchers to ascertain the causal direction of an association
+even when the biology of the instruments are not fully understood, and
+also a method to evaluate the sensitivity of the result of this
+extension to measurement error. This improves the utility of MR in cases
+where mediation based methods might have otherwise been used
 preferentially. Finally, we apply this new method to infer the direction
@@
 the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{34} and code is made
+the R programming language\textsuperscript{35} and code is made
 available at \href{}{github url to go here} and implemented in the
-MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{28}.
+MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.
 
@@
 method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
-for the correct model of \(x\) causing \(y\), and for the incorrect
-model of \(y\) causing \(x\). For some significance threshold
-\(\alpha\), the existence of causality and its direction was inferred
-based on the following scenarios:
+for the direction of \(x\) causing \(y\), and for the direction model of
+\(y\) causing \(x\). For some significance threshold \(\alpha\), the
+existence of causality and its direction was inferred based on the
+following scenarios:
 
@@
 \item
-  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
-  then the correct model is accepted
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then the \(x \rightarrow y\)
+  model is accepted
 \item
-  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
-  then the incorrect model is accepted
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then the \(y \rightarrow x\)
+  model is accepted
 \item
-  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
-  then an alternative causal model is accepted
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then no inference is made
 \item
-  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
-  then no inference is made
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
+  accepted (\(x \leftarrow g \rightarrow y\)).
 \end{itemize}
@@
 For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value, though we should stress that for real analyses
-it is not good practice to rely on p-values for making causal inference,
-nor is it reliable to depend on arbitrary significance
-thresholds\textsuperscript{35}.
+\(\alpha = 0.05\) value, though we stress that for real analyses it is
+not good practice to rely on p-values for making causal inference, nor
+is it reliable to depend on arbitrary significance
+thresholds\textsuperscript{36}.
 
@@
 is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
-R package \(R/systemfit\)\textsuperscript{36}.
-
-In what follows we assume no horizontal pleiotropy, such that \(g\) only
-has an effect on one of the variables through the other variable. To
-infer the direction of causality it is desirable to know which of the
+R package \(R/systemfit\)\textsuperscript{37}.
+
+The method that we will now describe is able to distinguish between two
+models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
+framework, this approach cannot infer if the true model is
+\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
+additive.
+
+To infer the direction of causality it is desirable to know which of the
 variables, \(x\) or \(y\), is being directly influenced by the
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population\textsuperscript{37}. It is calculated as
+within a population\textsuperscript{38}. It is calculated as
 
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{27} setting. Two-sample MR refers to the
+two-sample MR\textsuperscript{29} setting. Two-sample MR refers to the
 case where the SNP-exposure association and SNP-outcome association are
@@
 \emph{r.test} function in the R package
-\emph{R/psych}\textsuperscript{38}.
+\emph{R/psych}\textsuperscript{39}.
 
@@
 
-The Steiger test for testing if \(x \rightarrow y\) is based on
-evaluating \(\\rho_{gx} > \\rho_{gy}\). However, \(\\rho_{gx}\) and
-\(\\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
+The Steiger test for inferring if \(x \rightarrow y\) is based on
+evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
+\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
 imprecisely. If, for example, \(x\) has lower measurement precision than
-\(y\) then we might empirically obtain
-\(\\rho_{g,x_o} > \\rho_{g,y_o}\). To evaluate how reliable, \(R\), the
-Steiger test is to potential measurement imprecision in \(x\) and \(y\)
-we calculate the possible measurement error values which are compatible
-with the direction of causality obtained empirically from the Steiger
-test. We also calculate the measurement error values which are not
-compatible. The ratio of these quantities is an indication of how much
-of the measurement error space supports the inferred causal direction. A
-ratio \(R=1\) indicates that the Steiger test is highly sensitive to
-measurement error, with equal weight of the measurement error parameter
-space supporting both directions of causality. Full details are provided
-in Appendix 2.
+\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
+As we show in the appendix, the maximum measurement error of \(x_o\) is
+\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
+possible to simulate what the inferred causal direction would be for all
+values within these bounds. To evaluate how reliable, \(R\), the Steiger
+test is to potential measurement imprecision in \(x\) and \(y\) we
+integrate over the simulated \(R\) values for the possible measurement
+error values which are compatible with the direction of causality
+obtained empirically from the Steiger test. We also do the same for the
+measurement error values which are not compatible. The ratio of these
+quantities is an indication of how much of the measurement error space
+supports the inferred causal direction. A ratio \(R=1\) indicates that
+the Steiger test is highly sensitive to measurement error, with equal
+weight of the measurement error parameter space supporting both
+directions of causality. Full details are provided in Appendix 2.
 
@@
 outcome \(y\), the measured values of the outcome \(y_o\) and the
-instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
-instrument for \(x\). Each variable was simulated such that:
+instrument \(g\). One of two models are simulated, the ``causal model''
+where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
+``non-causal model'' where \(g\) influences a confounder \(u\) which in
+turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
+but not causally related. Each variable in the causal model was
+simulated such that:
 
@@
 and \(\beta_{my}\) are parameters for non-differential measurement error
-in the outcome \(y\).
+in the outcome \(y\). Similarly in the non-causal model:
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+u & = \alpha_g + \beta_g g + \epsilon_g \\
+x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+\]
 
@@
 of the simulated variables. Similar patterns of results were obtained
-for different values of \(cor(g, x)\).
+for different values of \(cor(g, x)\) and \(cor(g, u)\).
 
@@
 
-Two sample MR\textsuperscript{27} was performed using the summary
-statistics presented in\textsuperscript{26} for each of 4122
+Two sample MR\textsuperscript{29} was performed using the summary
+statistics presented in\textsuperscript{28} for each of 4122
 associations between DNA methylation and gene expression levels. To do
 this we obtained a list of 458 gene expression - DNA methylation
-associations as reported in Shakhbazov et al\textsuperscript{26}. These
+associations as reported in Shakhbazov et al\textsuperscript{28}. These
 were filtered to be located on the same chromosome, have robost
@@
 \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
-measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
-in the measurement of \(x\). Here the true value of the exposure is
-partially explained by the genetic instrument, \(g\), such that
+measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
+represents the imprecision (or noise) in the measurement of \(x\). Here
+the true value of the exposure is partially explained by the genetic
+instrument, \(g\), such that
 
@@
 where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
-error is introduced it can be shown using basic covariance properties
-(Appendix 1) that
+error in scale and imprecision is introduced, such that \(y_o\) is the
+measured value of \(y\), it can be shown using basic covariance
+properties (Appendix 1) that
 
@@
 between \(g\) and \(y_o | x\), which is in violation of the the 4th
-condition of the CIT. Note that linear transformation of \(x\) or \(y\)
+condition of the CIT. Note that scale transformation of \(x\) or \(y\)
 without any incurred imprecision is insufficient to lead to a violation
-of the test statistic assumptions.
+of the test statistic assumptions, and henceforth mention of measurement
+error will relate to imprecision unless otherwise stated.
 
@@
 cases, with performance declining as measurement imprecision increases
-in the exposure. The performance of MR in detecting an association is
-unrelated to measurement error in. Measurement error in the outcome does
-not induce a substantive difference in performance between CIT and MR.
+in the exposure. When MR assumptions are satisfied, notably that it is
+known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
+the performance of MR in detecting an association is unrelated to
+measurement error in the exposure. Measurement error in the outcome does
+reduce power, but does not induce a substantive difference in
+performance between CIT and MR.
 
@@
 rate at which a strong causal relationship is obtained where the
-reported direction of causality is incorrect. Figure 5 shows that when
-\(d < 0\) the MR analysis is indeed liable to infer the wrong direction
-of causality, and that this erroneous result is more likely to occur
-with increasing sample size. However, the CIT is in general much more
-fallable to reporting a robust causal association for the wrong
-direction of causality. When \(d > 0\) is satisfied we observe that in
-most cases the MR method has greater power to obtain evidence for
-causality than CIT, and always obtains the correct direction of
-causality.
+reported direction of causality is incorrect. Simulations were performed
+for two models, one for a ``causal model'' where there was a causal
+relationship between \(x\) and \(y\); and one for a ``non-causal model''
+where \(x\) and \(y\) were not causally related, but had a confounded
+association induced by the SNP \(g\) influencing a confounder variable
+\(u\).
+
+Figure 5a shows that, for the ``causal model'', when \(d < 0\) the MR
+analysis is indeed liable to infer the wrong direction of causality, and
+that this erroneous result is more likely to occur with increasing
+sample size. However, the CIT is in general more fallable to reporting a
+robust causal association for the wrong direction of causality. When
+\(d > 0\) is satisfied we observe that in most cases the MR method has
+greater power to obtain evidence for causality than CIT, and always
+obtains the correct direction of causality. The CIT, unlike the Steiger
+test for MR, is able to distinguish the ``non-causal model'' from the
+``causal model'' (Methods, Figure 5b), but it is evident that
+measurement error will often lead the CIT to evaluate this confounding
+model as being causal.
 
@@
 
-We performed two sample MR\textsuperscript{27} for each association in
+We performed two sample MR\textsuperscript{29} for each association in
 the direction of causality that we estimated to be the correct direction
@@
 generally in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{26}, but with a few
+reported by Shakhbazov et al\textsuperscript{28}, but with a few
 exceptions (Figure 6b). There was a moderate correlation between the
 absolute magnitudes of the causal correlation and the observational
-Pearson correlation (r = 0.4423627). Together these inferences suggest
+Pearson correlation (r = 0.4538942). Together these inferences suggest
 that even for relatively low level associations it is important to use
@@
 expression the causal effect was more likely to be negative than for the
-associations where gene expression caused methylation (OR = 0.5326663
-(95\% CI 0.3172203 - 0.8851868), Figure 6c), suggesting that reducing
+associations where gene expression caused methylation (OR = 0.6103961
+(95\% CI 0.3622428 - 1.029752), Figure 6c), suggesting that reducing
 gene expression levels at a controlling CpG typically leads to increased
-gene expression levels, consistent with expectation\textsuperscript{39}.
+gene expression levels, consistent with expectation\textsuperscript{40}.
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored\textsuperscript{40}. Our analysis
+analysis is relatively well explored\textsuperscript{41}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 evaluated measurement error in these platforms do support this
-position\textsuperscript{41,42}, though making strong conclusions for
+position\textsuperscript{42,43}, though making strong conclusions for
 this analysis is difficult because measurement error is likely to be
@@
 can offer solutions to inferring the causal
-direction\textsuperscript{43}. In this work we assumed that pleiotropy
+direction\textsuperscript{44}. In this work we assumed that pleiotropy
 (the influence of the instrument on the outcome through a mechanism
 other than the exposure) was not present. Recent method developments in
-MR\textsuperscript{44,45} have focused on accounting for the issues that
+MR\textsuperscript{24,25} have focused on accounting for the issues that
 horizontal pleiotropy can introduce, but how they perform in the
@@
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established\textsuperscript{33} and have a
+variables, are very well established\textsuperscript{34} and have a
 number of extensions that make them valuable tools, including for
@@
 the same issues of measurement error. Recent advances in MR methodology,
-for example applying MR to genetical genomics\textsuperscript{46},
-multivariate MR\textsuperscript{47} and mediation through
-MR\textsuperscript{48,49} may offer more robust alternatives for these
+for example applying MR to genetical genomics\textsuperscript{45},
+multivariate MR\textsuperscript{46} and mediation through
+MR\textsuperscript{47,48} may offer more robust alternatives for these
 more complicated problems.
@@
 methylation and gene expression levels then it can be used as an
-instrument, but there are two competing models for these variables. a)
+instrument, but there are three competing models for these variables.
+The causal inference test (CIT) attempts to delineate between them. a)
 Methylation causes gene expression. The left figure shows that the SNP
@@
 except the causal direction is from gene expression to DNA methylation.
+c) A model of confounding, where gene expression and DNA methylation are
+not causally related, but the SNP influences them each through separate
+pathways or a confounder.
 
@@
 
-Figure 5: Outcomes were simulated to be causally influenced by
-exposures, with varying degrees of measurement error applied to both.
-CIT and MR were used to infer evidence for causality between the
-exposure and outcome, and to infer the direction of causality. The value
-of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\)
-is negative we expect the Steiger test to be more likely to be wrong
-about the direction of causality. Rows of graphs represent the sample
-size used in the simulations. For the CIT method, outcome 1 denoted
-evidence for causality with correct model, outcomes 2 or 3 denoted
-evidence for causality with incorrect model, and outcome 4 denoted no
-evidence for causality.
+Figure 5: a) Outcomes were simulated to be causally influenced by
+exposures as shown in the graph, with varying degrees of measurement
+error applied to both. CIT and MR were used to infer evidence for
+causality between the exposure and outcome, and to infer the direction
+of causality. The value of
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\) is
+negative we expect the Steiger test to be more likely to be wrong about
+the direction of causality. Rows of graphs represent the sample size
+used in the simulations. For the CIT method, outcome 1 denoted evidence
+for causality with correct model, outcomes 2 or 3 denoted evidence for
+causality with incorrect model, and outcome 4 denoted no evidence for
+causality. b) As in (a) except the simulated model was non-causal, and a
+genetic confounder induces an association between \(x\) and \(y\). MR is
+unable to identify this model, so any significant associations are
+deemed to be incorrect. Outcome 3 denotes evidence for the correct model
+for the CIT method.
 
@@
 
-\newpage
-
 \subsection{Appendix 2}\label{appendix-2}
@@
 
+\hypertarget{ref-Bowden2015}{}
+24. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
+with invalid instruments: effect estimation and bias detection through
+Egger regression. \emph{International Journal of Epidemiology}
+\textbf{In press,} (2015).
+
+\hypertarget{ref-Bowden2016b}{}
+25. Bowden, J., Davey Smith, G., Haycock, P. C. \& Burgess, S.
+Consistent Estimation in Mendelian Randomization with Some Invalid
+Instruments Using a Weighted Median Estimator. \emph{Genetic
+Epidemiology} \textbf{40,} 304--314 (2016).
+
+\hypertarget{ref-Hemani2016}{}
+26. Hemani, G. \emph{et al.} MR-Base: a platform for systematic causal
+inference across the phenome using billions of genetic associations.
+\emph{BioRxiv} \textbf{10.1101/07,} (2016).
+
 \hypertarget{ref-Claussnitzer2015}{}
-24. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
+27. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
 Adipocyte Browning in Humans. \emph{The New England journal of medicine}
@@
 
-\hypertarget{ref-Timpson2011}{}
-25. Timpson, N. J. \emph{et al.} C-reactive protein levels and body mass
-index: elucidating direction of causation through reciprocal Mendelian
-randomization. \emph{International journal of obesity (2005)}
-\textbf{35,} 300--8 (2011).
-
 \hypertarget{ref-Shakhbazov2016}{}
-26. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+28. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
 and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
@@
 \hypertarget{ref-Pierce2013}{}
-27. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+29. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
 randomization studies: subsample and 2-sample instrumental variable
@@
 
-\hypertarget{ref-Hemani2016}{}
-28. Hemani, G. \emph{et al.} MR-Base: a platform for systematic causal
-inference across the phenome using billions of genetic associations.
-\emph{BioRxiv} \textbf{10.1101/07,} (2016).
-
 \hypertarget{ref-Ashenfelter1994}{}
-29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+30. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
 to Schooling from a New Sample of Twins. \emph{The American Economic
@@
 \hypertarget{ref-Nagarajan2013}{}
-30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+31. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
@@
 \hypertarget{ref-Shpitser2010}{}
-31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+32. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 \hypertarget{ref-Wang2015}{}
-32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+33. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
 gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
@@
 \hypertarget{ref-Lagani2015}{}
-33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+34. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
 I. in \emph{Uncertainty in biology: A computational modeling approach}
@@
 \hypertarget{ref-RCoreTeam2015}{}
-34. R Core Team. R: A Language and Environment for Statistical
+35. R Core Team. R: A Language and Environment for Statistical
 Computing. (2015). at
@@
 \hypertarget{ref-Sterne2001}{}
-35. Sterne, J. A. C., Cox, D. R. \& Smith, G. D. Sifting the
+36. Sterne, J. A. C., Cox, D. R. \& Smith, G. D. Sifting the
 evidence---what's wrong with significance tests?Another comment on the
@@
 \hypertarget{ref-Henningsen2007}{}
-36. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
+37. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. \emph{Journal of Statistical
@@
 \hypertarget{ref-Steiger1980}{}
-37. Steiger, J. H. Tests for comparing elements of a correlation matrix.
+38. Steiger, J. H. Tests for comparing elements of a correlation matrix.
 \emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
@@
 \hypertarget{ref-Revelle2015}{}
-38. Revelle, W. psych: Procedures for Psychological, Psychometric, and
+39. Revelle, W. psych: Procedures for Psychological, Psychometric, and
 Personality Research. (2015). at
@@
 \hypertarget{ref-Bird2002}{}
-39. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+40. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
 \& development} \textbf{16,} 6--21 (2002).
@@
 \hypertarget{ref-Fewell2007}{}
-40. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
+41. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
 residual and unmeasured confounding in epidemiologic studies: a
@@
 \hypertarget{ref-Bose2014}{}
-41. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
+42. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
 methylation measurement using technical replicates: the Atherosclerosis
@@
 \hypertarget{ref-Bryant2011}{}
-42. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
+43. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
 Biological Variability in a Microarray Experiment but Both Are
@@
 \hypertarget{ref-Richmond2014}{}
-43. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
+44. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
 between Child Adiposity and Physical Activity Levels: A Mendelian
@@
 
-\hypertarget{ref-Bowden2015}{}
-44. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
-with invalid instruments: effect estimation and bias detection through
-Egger regression. \emph{International Journal of Epidemiology}
-\textbf{In press,} (2015).
-
-\hypertarget{ref-Bowden2016b}{}
-45. Bowden, J., Davey Smith, G., Haycock, P. C. \& Burgess, S.
-Consistent Estimation in Mendelian Randomization with Some Invalid
-Instruments Using a Weighted Median Estimator. \emph{Genetic
-Epidemiology} \textbf{40,} 304--314 (2016).
-
 \hypertarget{ref-Relton2012}{}
-46. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+45. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-47. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+46. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Varbo2015}{}
-48. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+47. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Burgess2015}{}
-49. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+48. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental

### 2017-02-01 scripts/cit_tests.R

@@ -6,6 +6,7 @@ suppressWarnings(suppressPackageStartupMessages(library(ggnetwork)))
 suppressWarnings(suppressPackageStartupMessages(library(gridExtra)))
 suppressWarnings(suppressPackageStartupMessages(library(dplyr)))
 suppressWarnings(suppressPackageStartupMessages(library(tidyr)))
+suppressWarnings(suppressPackageStartupMessages(library(cit)))
 
 nid <- 10000
 l <- m <- list()
@@ -90,8 +91,9 @@ l[[3]] <- ggplot(ggnetwork(n, layout="circle", arrow.gap=0.05), aes(x = x, y = y
   labs(title="Model 3")
 
 g <- rnorm(nid)
-a <- g + rnorm(nid)
-b <- g + rnorm(nid)
+u <- g + rnorm(nid)
+a <- u + rnorm(nid)
+b <- u + rnorm(nid)
 m[[3]] <- rbind(
 	cit.cp(L=g, G=a, T=b),
 	cit.cp(L=g, G=b, T=a)

### 2017-02-01 scripts/mr_directionality_analysis.R

@@ -155,6 +155,8 @@ parametersc <- parameters
 parametersc$model <- "Non-causal"
 load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
 parameters$model <- "Causal"
+parameters$cit_AB1 <- parameters$cit_AB2 <- parameters$cit_AB3 <- parameters$cit_AB4 <- NA
+parameters$cit_BA1 <- parameters$cit_BA2 <- parameters$cit_BA3 <- parameters$cit_BA4 <- NA
 parameters <- rbind(parameters, parametersc)
 
 parameters <- subset(parameters, r_ab != 1)
@@ -205,7 +207,7 @@ parameters$cit_correct_direction[index_ncausal] <- parameters$cit_res[index_ncau
 
 parameters$cit_p[index_causal] <- parameters$cit_AB[index_causal]
 parameters$cit_p[parameters$cit_res == 2 & index_causal] <- parameters$cit_BA[parameters$cit_res == 2 & index_causal]
-parameters$cit_p[parameters$cit_res == 3 & index_causal] <- 0.01
+parameters$cit_p[parameters$cit_res == 3 & index_causal] <- 0.0001
 parameters$cit_p[parameters$cit_res == 4 & index_causal] <- 0.5
 
 
@@ -359,36 +361,36 @@ geom_bar(stat="identity", aes(fill=eval)) +
 facet_grid(n ~ rhs_lhs_diff_bin) +
 scale_fill_brewer(type="qual") +
 labs(x="Method", y="Proportion of simulations", fill="") +
-theme(legend.position="none")
+theme(legend.position="none", strip.text.x=element_text(angle=90), axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
 
 p2 <- ggplot(subset(psum2, round(r_za,2)==0.01 & model == "Non-causal"), aes(x=test, y=value)) +
 geom_bar(stat="identity", aes(fill=eval)) +
 facet_grid(n ~ rhs_lhs_diff_bin) +
 scale_fill_brewer(type="qual") +
 labs(x="Method", y="Proportion of simulations", fill="") +
-theme(legend.position="none")
+theme(legend.position="none", strip.text.x=element_text(angle=90), axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
 
[…]

### 2017-02-01 docs/manuscript.Rmd
updated

@@
 b) The same as in A, except the causal direction is from gene expression to DNA methylation.
-c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.
-",
+c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
 	display=FALSE
@@
 
-Inference of the causal relationship between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others.
-We show here, however, that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
-Here we develop an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages.
-First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; and third, it can be performed using only summary level data from genome-wide association studies.
+Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others.
+However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
+Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages.
+First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated.
 We apply the method to infer the causal direction between DNA methylation and gene expression levels.
 Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but that such high level conclusions are susceptible to differential measurement error between the platforms.
-We emphasise that the use of several different methods is important to triangulate reliable conclusions.
+We emphasise that the use of several different methods is important to triangulate reliable conclusions about causality.
 

### 2017-02-01 docs/manuscript.Rmd
updated

@@
 	"dag_figure",
-	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown.
-If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables.
+	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown.
+If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables.
 The causal inference test (CIT) attempts to delineate between them.
@@
 Rows of plots represent the sample sizes used for the simulations.
-As measurement error increases (decreasing values on x-axis) the test statistic for the incorrect model gets stronger and the test statistic for the correct model gets weaker.",
+As measurement imprecision increases (decreasing values on x-axis) the test statistic for the incorrect model gets stronger and the test statistic for the correct model gets weaker.",
 	display=FALSE
@@
 The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes).
-If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor.
+If there is systematically higher measurement error in one platform than the other it will appear to be less likely to be the causal factor.
 b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis).
@@
 However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
-Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages.
+Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages.
 First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated.
 We apply the method to infer the causal direction between DNA methylation and gene expression levels.
-Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but that such high level conclusions are susceptible to differential measurement error between the platforms.
-We emphasise that the use of several different methods is important to triangulate reliable conclusions about causality.
+Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms.
+We emphasise that, where possible, implementing MR alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.
 
@@
 
-Measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value.
+Measurement error is the difference between the measured value of a quantity and its true value.
 This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement.
@@
 
-Measurement error will of course arise in more basic data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+Measurement error will of course arise in other types of data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
@@
 By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference.
-An important advantage of mediation-based analysis is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
+An important advantage of mediation-based analysis is that in order to infer the causal direction the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
 In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.
-The CIT can also distinguish between several putative models of association between the traits and the SNP [@Millstein2009; @Millstein2016].
+The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`).
 Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.
 
-Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e.
+Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e.
 for which of the two variables is the SNP a valid instrument?
@@
 But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known.
-For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?
+For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level (`r cfig("dag_figure")`)?
 
@@
 Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
+Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].
 
@@
 A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
-Full details are provided in Appendix 2.
+Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.
 
@@
 To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
-These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
+These were filtered to be located on the same chromosome, have robust correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe.
 Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations.
@@
 
-### MR analysis of gene expression and DNA methylation levels
+### The causal relationship between gene expression and DNA methylation levels
 
-We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a).
+We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
 We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
@@
 We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test.
-We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], but with a few exceptions (`r cfig("shakhplot")`b).
-There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index]))`).
+We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b).
+There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`).
 Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.
 
-We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
+We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
 
@@
 Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data.
-In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Fewell2007].
+In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Cole2014].
 Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data.
@@
 This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest.
-But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
-Here we have described a simple extension to MR which can be used as an alternative to mediation based methods.
-We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages.
-First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality.
+But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation in have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
+Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods.
+We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages.
+First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality.
 Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.
@@
 But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
-Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015] may offer more robust alternatives for these more complicated problems.
+Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015; @Richmond2016] may offer more robust alternatives for these more complicated problems.
 

### 2017-02-01 docs/manuscript.tex
updated

@@
 
-Inference of the causal relationship between two traits can be achieved
-by combining genetic associations with a mediation-based approach, as is
-done in the causal inference test (CIT) and others. We show here,
-however, that measurement error in the phenotypes can lead to
-mediation-based approaches inferring the wrong causal direction, and
-that increasing sample sizes has the adverse effect of increasing
-confidence in the wrong answer. Here we develop an extension to
-Mendelian randomisation, a method that uses genetic associations in an
-instrumentation framework, that enables inference of the causal
-direction with some advantages. First, it is less susceptible to bias in
-the presence of measurement error; second, it is more statistically
-efficient; and third, it can be performed using only summary level data
-from genome-wide association studies. We apply the method to infer the
-causal direction between DNA methylation and gene expression levels. Our
-results demonstrate that, in general, DNA methylation is more likely to
-be the causal factor, but that such high level conclusions are
-susceptible to differential measurement error between the platforms. We
-emphasise that the use of several different methods is important to
-triangulate reliable conclusions.
+Inference of the causal structure that induces correlations between two
+traits can be achieved by combining genetic associations with a
+mediation-based approach, as is done in the causal inference test (CIT)
+and others. However, we show that measurement error in the phenotypes
+can lead to mediation-based approaches inferring the wrong causal
+direction, and that increasing sample sizes has the adverse effect of
+increasing confidence in the wrong answer. Here we introduce an
+extension to Mendelian randomisation, a method that uses genetic
+associations in an instrumentation framework, that enables inference of
+the causal direction between traits, with some advantages. First, it is
+less susceptible to bias in the presence of measurement error; second,
+it is more statistically efficient; third, it can be performed using
+only summary level data from genome-wide association studies; and
+fourth, its sensitivity to measurement error can be evaluated. We apply
+the method to infer the causal direction between DNA methylation and
+gene expression levels. Our results demonstrate that, in general, DNA
+methylation is more likely to be the causal factor, but this result is
+highly susceptible to bias induced by systematic differences in
+measurement error between the platforms. We emphasise that, where
+possible, implementing MR alongside other approaches such as CIT is
+important to triangulate reliable conclusions about causality.
 
@@
 
-Measurement error (or observational error, henceforth referred to simply
-as measurement error) is the difference between the measured value of a
+Measurement error is the difference between the measured value of a
 quantity and its true value. This study focuses specifically on
@@
 
-Measurement error will of course arise in more basic data too, for
+Measurement error will of course arise in other types of data too, for
 example when measuring BMI one is typically interested in using this as
@@
 some disadvantages for reliable causal inference. An important advantage
-of mediation-based analysis is that the true underlying biology of the
-genetic instrument doesn't necessarily need to be known as long as it
-associates with the putative exposure. In the CIT framework (described
-fully in the Methods) for example, the test statistic is different if
-you test for the exposure causing the outcome or the outcome causing the
-exposure, allowing the researcher to infer the direction of causality
-between two variables by performing the test in both directions and
-choosing the model with the strongest evidence. The CIT can also
-distinguish between several putative models of association between the
-traits and the SNP\textsuperscript{4,18}. Such is not the case for MR,
-where in order to infer the direction of causality between two traits
-the instrument must be known to influence the exposure primarily,
+of mediation-based analysis is that in order to infer the causal
+direction the true underlying biology of the genetic instrument doesn't
+necessarily need to be known as long as it associates with the putative
+exposure. In the CIT framework (described fully in the Methods) for
+example, the test statistic is different if you test for the exposure
+causing the outcome or the outcome causing the exposure, allowing the
+researcher to infer the direction of causality between two variables by
+performing the test in both directions and choosing the model with the
+strongest evidence. The CIT also has the valuable property of being able
+to distinguish between several putative causal graphs that link the
+traits with the SNP (Figure 1). Such is not the case for MR, where in
+order to infer the direction of causality between two traits the
+instrument must be known to influence the exposure primarily,
 associating with the outcome only through the exposure.
@@
 because if there exists a putative association between two variables,
-with the instrumental SNP being robustly associated with each, it can be
-difficult to determine which of the two variables is subject to the
-primary effect of the SNP (i.e.~for which of the two variables is the
-SNP a valid instrument? Figure 1). By definition, we expect that if the
-association is causal then a SNP for the exposure will be associated
-with the outcome, such that if the researcher erroneously uses the SNP
-as an instrument for the outcome then they are likely to see an
-apparently robust causal association of outcome on exposure. Such a
-situation can arise in many scenarios. Genome-wide association studies
-(GWASs) that identify genetic associations for complex traits are, by
-design, hypothesis free and agnostic of genomic function, and it often
-takes years of follow up studies to understand the biological nature of
-a putative GWAS hit\textsuperscript{27}. If multiple instruments are
+with the SNP being robustly associated with each, it can be difficult to
+determine which of the two variables is subject to the primary effect of
+the SNP (i.e.~for which of the two variables is the SNP a valid
+instrument? Figure 1). By definition, we expect that if the association
+is causal then a SNP for the exposure will be associated with the
+outcome, such that if the researcher erroneously uses the SNP as an
+instrument for the outcome then they are likely to see an apparently
+robust causal association of outcome on exposure. Such a situation can
+arise in many scenarios. Genome-wide association studies (GWASs) that
+identify genetic associations for complex traits are, by design,
+hypothesis free and agnostic of genomic function, and it often takes
+years of follow up studies to understand the biological nature of a
+putative GWAS hit\textsuperscript{27}. If multiple instruments are
 available for an hypothesised exposure, which is increasingly typical
@@
 an instrument for the DNA methylation level, or the gene expression
-level\textsuperscript{28} (Figure 1)?
+level (Figure 1)?
 
@@
 causal inference to be drawn when exposure variables and outcome
-variables are measured in different samples\textsuperscript{29}. This
+variables are measured in different samples\textsuperscript{28}. This
 has the crucial advantage of improving statistical power by allowing
@@
 (IV) analysis was in part initially introduced as a correction for
-measurement error in the exposure\textsuperscript{30}, whereas it has
+measurement error in the exposure\textsuperscript{29}, whereas it has
 been noted that both classic mediation-based
-analyses\textsuperscript{13,14,31,32} and mediation-based methods that
-use instrumental variables\textsuperscript{33,34} are prone to be
+analyses\textsuperscript{13,14,30,31} and mediation-based methods that
+use instrumental variables\textsuperscript{32,33} are prone to be
 unreliable in its presence.
@@
 of causation between DNA methylation levels and gene expression levels.
+Our analyses highlight that because these different causal inference
+techniques have varying strengths and weaknesses, triangulation of
+evidence from as many sources as possible should be practiced in causal
+inference\textsuperscript{34}.
 
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{29} setting. Two-sample MR refers to the
+two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
 case where the SNP-exposure association and SNP-outcome association are
@@
 weight of the measurement error parameter space supporting both
-directions of causality. Full details are provided in Appendix 2.
+directions of causality. Full details are provided in Appendix 2 and
+Figure 3a.
 
@@
 
-Two sample MR\textsuperscript{29} was performed using the summary
-statistics presented in\textsuperscript{28} for each of 4122
+Two sample MR\textsuperscript{28} was performed using the summary
+statistics presented in\textsuperscript{40} for each of 4122
 associations between DNA methylation and gene expression levels. To do
 this we obtained a list of 458 gene expression - DNA methylation
-associations as reported in Shakhbazov et al\textsuperscript{28}. These
-were filtered to be located on the same chromosome, have robost
+associations as reported in Shakhbazov et al\textsuperscript{40}. These
+were filtered to be located on the same chromosome, have robust
 correlations after correcting for multiple testing, and to share a SNP
@@
 
-\subsubsection{MR analysis of gene expression and DNA methylation
-levels}\label{mr-analysis-of-gene-expression-and-dna-methylation-levels}
+\subsubsection{The causal relationship between gene expression and DNA
+methylation
+levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}
 
@@
 methylation levels and gene expression levels and found that the causal
-direction commonly goes in both directions (Figure 6a). We then went on
-to predict the causal directions of the associations for varying levels
-of possible measurement error. Figure 6a shows that the qualitative
-result of which of DNA methylation or gene expression is more commonly
-the causal factor is very strongly determined by measurement error.
-
-We performed two sample MR\textsuperscript{29} for each association in
+direction commonly goes in both directions (Figure 6a), but assuming no
+or equal measurement error DNA methylation levels were the predominant
+causal factor (\(p = 1.3\times 10^{-5}\)). We then went on to predict
+the causal directions of the associations for varying levels of possible
+measurement error. Figure 6a shows that the qualitative result of which
+of DNA methylation or gene expression is more commonly the causal factor
+is very strongly determined by measurement error.
+
+We performed two sample MR\textsuperscript{28} for each association in
 the direction of causality that we estimated to be the correct direction
@@
 generally in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{28}, but with a few
-exceptions (Figure 6b). There was a moderate correlation between the
-absolute magnitudes of the causal correlation and the observational
-Pearson correlation (r = 0.4538942). Together these inferences suggest
-that even for relatively low level associations it is important to use
-causal inference methods over observational associations to infer causal
-effect sizes.
+reported by Shakhbazov et al\textsuperscript{40} (Figure 6b). There was
+a moderate correlation between the absolute magnitudes of the causal
+correlation and the observational Pearson correlation (r = 0.45).
+Together these inferences suggest that even for relatively low level
+associations it is important to use causal inference methods over
+observational associations to infer causal effect sizes.
 
@@
 expression the causal effect was more likely to be negative than for the
-associations where gene expression caused methylation (OR = 0.6103961
-(95\% CI 0.3622428 - 1.029752), Figure 6c), suggesting that reducing
-gene expression levels at a controlling CpG typically leads to increased
-gene expression levels, consistent with expectation\textsuperscript{40}.
+associations where gene expression caused methylation (OR = 0.61 (95\%
+CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
+levels at a controlling CpG typically leads to increased gene expression
+levels, consistent with expectation\textsuperscript{41}.
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored\textsuperscript{41}. Our analysis
+analysis is relatively well explored\textsuperscript{42}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 known biological relevance for the trait of interest. But on many
-occasions this is not the case, and it may be tempting to resort to
-using methods based on mediation in order to be able to both ascertain
-if there is a causal association and to infer the direction of
-causality. Here we have described a simple extension to MR which can be
-used as an alternative to mediation based methods. We show that this
-method is still liable to measurement error, but because it has
-different properties to the CIT it offers two main advantages. First, it
-uses a formal statistical framework to test for the robostness of the
-assumed direction of causality. Second, after testing in a comprehensive
-range of scenarios the MR based approach is less likely to infer the
-wrong direction of causality compared to CIT, while substantially
-improving power over CIT in the cases where \(d > 0\).
+occasions, especially in the realm of 'omic data, this is not the case,
+and methods based on mediation in have been valuable in order to be able
+to both ascertain if there is a causal association and to infer the
+direction of causality. Here we have described a simple extension to MR
+which can be used as an alternative to or in conjunction with mediation
+based methods. We show that this method is still liable to measurement
+error, but because it has different properties to the CIT it offers
+several main advantages. First, it uses a formal statistical framework
+to test for the robustness of the assumed direction of causality.
+Second, after testing in a comprehensive range of scenarios the MR based
+approach is less likely to infer the wrong direction of causality
+compared to CIT, while substantially improving power over CIT in the
+cases where \(d > 0\).
 
@@
 evaluated measurement error in these platforms do support this
-position\textsuperscript{42,43}, though making strong conclusions for
+position\textsuperscript{43,44}, though making strong conclusions for
 this analysis is difficult because measurement error is likely to be
@@
 can offer solutions to inferring the causal
-direction\textsuperscript{44}. In this work we assumed that pleiotropy
+direction\textsuperscript{45}. In this work we assumed that pleiotropy
 (the influence of the instrument on the outcome through a mechanism
@@
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established\textsuperscript{34} and have a
+variables, are very well established\textsuperscript{33} and have a
 number of extensions that make them valuable tools, including for
@@
 the same issues of measurement error. Recent advances in MR methodology,
-for example applying MR to genetical genomics\textsuperscript{45},
-multivariate MR\textsuperscript{46} and mediation through
-MR\textsuperscript{47,48} may offer more robust alternatives for these
+for example applying MR to genetical genomics\textsuperscript{46},
+multivariate MR\textsuperscript{47} and mediation through
+MR\textsuperscript{48--50} may offer more robust alternatives for these
 more complicated problems.
@@
 Figure 1: Gene expression levels (blue blocks) and DNA methylation
-levels (green triangles) may be correlated but the causal direction is
+levels (green triangles) may be correlated but the causal structure is
 unknown. If a SNP (yellow circle) is associated with both DNA
 methylation and gene expression levels then it can be used as an
-instrument, but there are three competing models for these variables.
-The causal inference test (CIT) attempts to delineate between them. a)
-Methylation causes gene expression. The left figure shows that the SNP
-influences methylation levels that in turn influence gene expression
-levels. The right figure shows the directed acyclic graph that
-represents this model. Faded symbols represent the measured values
+instrument, but there are three basic competing models for these
+variables. The causal inference test (CIT) attempts to delineate between
+them. a) Methylation causes gene expression. The left figure shows that
+the SNP influences methylation levels that in turn influence gene
+expression levels. The right figure shows the directed acyclic graph
+that represents this model. Faded symbols represent the measured values
 whereas solid symbols represent the true values. b) The same as in A,
@@
 the exposure (false model) is in green. Rows of plots represent the
-sample sizes used for the simulations. As measurement error increases
-(decreasing values on x-axis) the test statistic for the incorrect model
-gets stronger and the test statistic for the correct model gets weaker.
+sample sizes used for the simulations. As measurement imprecision
+increases (decreasing values on x-axis) the test statistic for the
+incorrect model gets stronger and the test statistic for the correct
+model gets weaker.
 
@@
 methylation levels (columns of boxes). If there is systematically higher
-measurement error in one platform it will appear to be less likely to be
-the causal factor. b) The relationship between the Pearson correlation
-between DNA methylation and gene expression levels (x-axis) and the
-causal estimate (scaled to be in standard deviation units, y-axis). c)
-Distribution of estimated causal effect sizes, stratified into
-associations inferred to be due to DNA methylation causing expression
-(blue) and expression causing DNA methylation (red).
+measurement error in one platform than the other it will appear to be
+less likely to be the causal factor. b) The relationship between the
+Pearson correlation between DNA methylation and gene expression levels
+(x-axis) and the causal estimate (scaled to be in standard deviation
+units, y-axis). c) Distribution of estimated causal effect sizes,
+stratified into associations inferred to be due to DNA methylation
+causing expression (blue) and expression causing DNA methylation (red).
 
@@
 
-\hypertarget{ref-Shakhbazov2016}{}
-28. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
-and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
-278 (2016).
-
 \hypertarget{ref-Pierce2013}{}
-29. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
+28. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
 randomization studies: subsample and 2-sample instrumental variable
@@
 \hypertarget{ref-Ashenfelter1994}{}
-30. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
 to Schooling from a New Sample of Twins. \emph{The American Economic
@@
 \hypertarget{ref-Nagarajan2013}{}
-31. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
+30. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
 inference. \emph{PloS one} \textbf{8,} e80735 (2013).
@@
 \hypertarget{ref-Shpitser2010}{}
-32. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
+31. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
 covariate adjustment for estimating causal effects. \emph{Proceedings of
@@
 \hypertarget{ref-Wang2015}{}
-33. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
+32. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
 gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
@@
 \hypertarget{ref-Lagani2015}{}
-34. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
+33. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
 I. in \emph{Uncertainty in biology: A computational modeling approach}
@@
 \textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+
+\hypertarget{ref-Lawlor2017}{}
+34. Lawlor, D. A., Tilling, K. \& Davey Smith, G. Triangulation in
+aetiological epidemiology. \emph{International Journal of Epidemiology}
+\textbf{19,} dyw314 (2017).
 
@@
 
+\hypertarget{ref-Shakhbazov2016}{}
+40. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
+and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
+278 (2016).
+
 \hypertarget{ref-Bird2002}{}
-40. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+41. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
 \& development} \textbf{16,} 6--21 (2002).
 
-\hypertarget{ref-Fewell2007}{}
-41. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
-residual and unmeasured confounding in epidemiologic studies: a
-simulation study. \emph{American journal of epidemiology} \textbf{166,}
-646--55 (2007).
+\hypertarget{ref-Cole2014}{}
+42. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
+Potentially Serious and Misleading Consequences Due to Uncorrected
+Measurement Error. \emph{Psychological Methods} \textbf{19,} 300--315
+(2014).
 
 \hypertarget{ref-Bose2014}{}
-42. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
+43. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
 methylation measurement using technical replicates: the Atherosclerosis
@@
 \hypertarget{ref-Bryant2011}{}
-43. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
+44. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
 Biological Variability in a Microarray Experiment but Both Are
@@
 \hypertarget{ref-Richmond2014}{}
-44. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
+45. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
 between Child Adiposity and Physical Activity Levels: A Mendelian
@@
 \hypertarget{ref-Relton2012}{}
-45. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+46. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-46. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+47. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Varbo2015}{}
-47. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+48. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Burgess2015}{}
-48. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+49. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 
+\hypertarget{ref-Richmond2016}{}
+50. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
+C. L. Challenges and novel approaches for investigating molecular
+mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
+(2016).
+
 \end{document}

### 2017-02-01 scripts/mr_directionality_analysis.R

@@ -10,6 +10,7 @@ suppressPackageStartupMessages(library(latex2exp))
 suppressPackageStartupMessages(library(knitr))
 suppressPackageStartupMessages(library(gridExtra))
 suppressPackageStartupMessages(library(grid))
+suppressPackageStartupMessages(library(cowplot))
 suppressPackageStartupMessages(library(network))
 suppressPackageStartupMessages(library(sna))
 suppressPackageStartupMessages(library(ggnetwork))
@@ -271,6 +272,7 @@ psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=seq(-0.2, 1.1, by=0.1))
 
 ## ---- cit_measurement_error_figure ----
 
+
 load("~/repo/cit_measurement_error/results/20141114.RData")
 dat <- gather(dat, eval, pval, GT, TG, factor_key=TRUE)
 levels(dat$eval) <- c("Correct causal model", "Incorrect causal model")
@@ -334,6 +336,7 @@ grid.arrange(p1, p2, ncol=1)
 
 
 ## ---- cit_mr_comparison_figure ----
+
 
 psum2 <- gather(psum1, eval, value, prop_sig_correct, prop_sig_incorrect, prop_nonsig, factor_key=TRUE) %>%
 	dplyr::group_by(n, rhs_lhs_diff_bin, r_za, eval, test, model) %>%
@@ -609,6 +612,7 @@ facet_grid(noisea ~ noiseb)
 
 
 ## ---- causality_exists_tpr ----
+
 
 pl$causality_exists <- TRUE
 pl$causality_exists[pl$test == "MR"] <- pl$p_bz[pl$test == "MR"] > -log10(0.05)
@@ -944,6 +948,7 @@ for(i in 1:nrow(sensitivity_parameters))
 
 
 ## ---- steiger_sensitivity_plot ----
+
 
 example_sensitivity <- mr_steiger(
 	get_p_from_r2n(sqrt(0.01), 10001),

### 2017-02-09 docs/manuscript.Rmd
updated

@@
 Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms.
-We emphasise that, where possible, implementing MR alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.
+We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.
 
@@
 
-Measurement error will of course arise in other types of data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+Measurement error will of course arise in other types of data too.
+For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
@@
 By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference.
-An important advantage of mediation-based analysis is that in order to infer the causal direction the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure.
 In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.
@@
 
-MR does however have some important advantages over the mediation-based approaches.
+MR has some important advantages over the mediation-based approaches.
 The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
-Additionally, when MR assumptions are satisfied the method is substantially more robust to there being measurement error in the exposure variable.
+Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable.
 Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
@@
 Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].
+
+## Model
 

### 2017-02-15 docs/manuscript.Rmd
updated

@@
 
+We can model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that
+
+$$
+y = \alpha_x + \beta_x x + \epsilon_x
+$$
+
+In addition, the exposure is influenced by a SNP $g$ such that
+
+$$
+x = \alpha_g + \beta_g g + \epsilon_g
+$$
+
+Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that
+
+$$
+cov(g, y - \hat{y}) = 0
+$$
+
+where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$
+
+MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
+
+$$
+\begin{aligned}
+\hat{x} & = \hat{\beta}_g g \\
+y & = \beta_{MR}\hat{x} + \epsilon_{MR}
+\end{aligned}
+$$
+
+where $\beta_{MR} \neq 0$.
+
+Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$.
+For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
+Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that
+
+$$
+x = \alpha_g + \beta_g g + \epsilon_g
+$$
+
+where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure
+
+$$
+y = \alpha_x + \beta_x x + \epsilon_x
+$$
+
+where $\beta_x$ is the true effect of the exposure on the outcome.
+In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
+The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
+
+$$
+\begin{aligned}
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
+\end{aligned}
+$$
+
+where
+
+$$
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+$$
+
+Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
+Therefore, if there is any measurement error that incurs imprecision (i.e.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
+Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
+
+We performed simulations to verify that this problem does arise using the CIT method.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
+Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
+
+When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
+Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
+
+$$
+\begin{aligned}
+\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
+                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
+                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
+\end{aligned}
+$$
+
+Thus we obtain an estimate of the effect of $x_o$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
+Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
+
+We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
+We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
+When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
+Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
+
 ## Methods
@@
 3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
-4. $H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*
+4. $H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*
 
@@
 ## Results
-
-### The influence of measurement error on mediation
-
-Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$.
-For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
-Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that
-
-$$
-x = \alpha_g + \beta_g g + \epsilon_g
-$$
-
-where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure
-
-$$
-y = \alpha_x + \beta_x x + \epsilon_x
-$$
-
-where $\beta_x$ is the true effect of the exposure on the outcome.
-In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
-The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
-                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
-\end{aligned}
-$$
-
-where
-
-$$
-D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
-$$
-
-Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
-Therefore, if there is any measurement error that incurs imprecision (i.e.
-$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
-Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
-
-We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
-However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
-Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
-
-### Using MR to infer the existence of causality
-
-When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
-Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
-
-$$
-\begin{aligned}
-\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
-                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
-                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
-\end{aligned}
-$$
-
-Thus we obtain an estimate of the effect of $x_o$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
-Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
-
-We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
-`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
-We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
-When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
-Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
 

### 2017-02-15 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{01 February 2017}
+  \date{15 February 2017}
 
@@
 measurement error between the platforms. We emphasise that, where
-possible, implementing MR alongside other approaches such as CIT is
-important to triangulate reliable conclusions about causality.
+possible, implementing MR and appropriate sensitivity analyses alongside
+other approaches such as CIT is important to triangulate reliable
+conclusions about causality.
 
@@
 
-Measurement error will of course arise in other types of data too, for
+Measurement error will of course arise in other types of data too. For
 example when measuring BMI one is typically interested in using this as
@@
 analysis and MR models have properties that confer some advantages and
-some disadvantages for reliable causal inference. An important advantage
-of mediation-based analysis is that in order to infer the causal
-direction the true underlying biology of the genetic instrument doesn't
-necessarily need to be known as long as it associates with the putative
-exposure. In the CIT framework (described fully in the Methods) for
-example, the test statistic is different if you test for the exposure
-causing the outcome or the outcome causing the exposure, allowing the
-researcher to infer the direction of causality between two variables by
-performing the test in both directions and choosing the model with the
-strongest evidence. The CIT also has the valuable property of being able
-to distinguish between several putative causal graphs that link the
-traits with the SNP (Figure 1). Such is not the case for MR, where in
-order to infer the direction of causality between two traits the
-instrument must be known to influence the exposure primarily,
-associating with the outcome only through the exposure.
+some disadvantages for reliable causal inference. In the CIT framework
+(described fully in the Methods) for example, the test statistic is
+different if you test for the exposure causing the outcome or the
+outcome causing the exposure, allowing the researcher to infer the
+direction of causality between two variables by performing the test in
+both directions and choosing the model with the strongest evidence. The
+CIT also has the valuable property of being able to distinguish between
+several putative causal graphs that link the traits with the SNP (Figure
+1). Such is not the case for MR, where in order to infer the direction
+of causality between two traits the instrument must be known to
+influence the exposure primarily, associating with the outcome only
+through the exposure.
 
@@
 
-MR does however have some important advantages over the mediation-based
-approaches. The mediation-based approaches require that the exposure,
-outcome and instrumental variables are all measured in the same data,
-whereas recent extensions to MR circumvent this requirement, allowing
-causal inference to be drawn when exposure variables and outcome
-variables are measured in different samples\textsuperscript{28}. This
-has the crucial advantage of improving statistical power by allowing
-analysis in much larger sample sizes, and dramatically expands the
-breadth of possible phenotypic relationships that can be
-evaluated\textsuperscript{26}. Additionally, when MR assumptions are
-satisfied the method is substantially more robust to there being
-measurement error in the exposure variable. Indeed instrumental variable
-(IV) analysis was in part initially introduced as a correction for
-measurement error in the exposure\textsuperscript{29}, whereas it has
-been noted that both classic mediation-based
+MR has some important advantages over the mediation-based approaches.
+The mediation-based approaches require that the exposure, outcome and
+instrumental variables are all measured in the same data, whereas recent
+extensions to MR circumvent this requirement, allowing causal inference
+to be drawn when exposure variables and outcome variables are measured
+in different samples\textsuperscript{28}. This has the crucial advantage
+of improving statistical power by allowing analysis in much larger
+sample sizes, and dramatically expands the breadth of possible
+phenotypic relationships that can be evaluated\textsuperscript{26}.
+Additionally, when MR assumptions are satisfied the method is robust to
+there being measurement error in the exposure variable. Indeed
+instrumental variable (IV) analysis was in part initially introduced as
+a correction for measurement error in the exposure\textsuperscript{29},
+whereas it has been noted that both classic mediation-based
 analyses\textsuperscript{13,14,30,31} and mediation-based methods that
@@
 
+\subsection{Model}\label{model}
+
+We can model a system whereby some exposure \(x\) has a causal influence
+\(\beta_x\) on an outcome \(y\) such that
+
+\[
+y = \alpha_x + \beta_x x + \epsilon_x
+\]
+
+In addition, the exposure is influenced by a SNP \(g\) such that
+
+\[
+x = \alpha_g + \beta_g g + \epsilon_g
+\]
+
+Mediation-based analyses that test whether \(x\) causally relates to
+\(y\) impinge on evaluating if the influence of \(g\) on \(y\) can be
+accounted for by conditioning on \(x\), such that
+
+\[
+cov(g, y - \hat{y}) = 0
+\]
+
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\)
+
+MR analysis estimates the causal influence of \(x\) on \(y\) by using
+the instrument as a proxy for \(x\), such that
+
+\[
+\begin{aligned}
+\hat{x} & = \hat{\beta}_g g \\
+y & = \beta_{MR}\hat{x} + \epsilon_{MR}
+\end{aligned}
+\]
+
+where \(\beta_{MR} \neq 0\).
+
+Measurement error of an exposure can be modeled as some transformation
+of the true value that leads to the observed value, \(x_o = f(x)\). For
+example, we can define
+\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
+\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
+measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
+represents the imprecision (or noise) in the measurement of \(x\). Here
+the true value of the exposure is partially explained by the genetic
+instrument, \(g\), such that
+
+\[
+x = \alpha_g + \beta_g g + \epsilon_g
+\]
+
+where \(\beta_g\) is the effect of the SNP on the exposure, and
+\(\epsilon_g\) is the normally distributed residual value of \(x\); and
+the outcome is partially explained by the exposure
+
+\[
+y = \alpha_x + \beta_x x + \epsilon_x
+\]
+
+where \(\beta_x\) is the true effect of the exposure on the outcome. In
+the causal inference test (CIT), an omnibus p-value is generated from
+four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
+associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
+is independent of \(y|x\). The 4th condition employs mediation for
+causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
+error in scale and imprecision is introduced, such that \(y_o\) is the
+measured value of \(y\), it can be shown using basic covariance
+properties (Appendix 1) that
+
+\[
+\begin{aligned}
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
+\end{aligned}
+\]
+
+where
+
+\[
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+\]
+
+Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
+when the true model is causal only when \(D = 1\). Therefore, if there
+is any measurement error that incurs imprecision (i.e.
+\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
+between \(g\) and \(y_o | x\), which is in violation of the the 4th
+condition of the CIT. Note that scale transformation of \(x\) or \(y\)
+without any incurred imprecision is insufficient to lead to a violation
+of the test statistic assumptions, and henceforth mention of measurement
+error will relate to imprecision unless otherwise stated.
+
+We performed simulations to verify that this problem does arise using
+the CIT method. Figure 2 shows that when there is no measurement error
+in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
+reliable in identifying the correct causal direction. However, as
+measurement error increases in the exposure variable, eventually the CIT
+is more likely to infer a robust causal association in the wrong
+direction. Also of concern here is that increasing sample size does not
+solve the issue, indeed it only strengthens the apparent evidence for
+the incorrect inference.
+
+When selecting an instrument \(g\) that has a direct influence on \(x\)
+where \(x\) is causally related to \(y\), one can reason that the
+influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
+\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
+effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as
+
+\[
+\begin{aligned}
+\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
+                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
+                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
+\end{aligned}
+\]
+
+Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
+scaled by the measurement error in \(x\) and \(y\), but is unrelated to
+measurement imprecision. Measurement error and imprecision are likely to
+reduce the power of MR inferring a causal association (increased false
+negative rate), but importantly the evidence for causal association,
+based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
+inflated when the null hypothesis is true.
+
+We performed simulations to compare the performance of MR against CIT in
+detecting a causal association between simulated variables under
+different levels of imprecision simulated in the exposure. Figure 4
+shows the true positive rates between the CIT and MR for detecting a
+causal association. We observe that the CIT has lower power in all
+cases, with performance declining as measurement imprecision increases
+in the exposure. When MR assumptions are satisfied, notably that it is
+known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
+the performance of MR in detecting an association is unrelated to
+measurement error in the exposure. Measurement error in the outcome does
+reduce power, but does not induce a substantive difference in
+performance between CIT and MR.
+
 \subsection{Methods}\label{methods}
@@
 \item
-  \(H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0\);
+  \(H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0\);
   \emph{the SNP is independent of the outcome when the outcome is
@@
 \subsection{Results}\label{results}
-
-\subsubsection{The influence of measurement error on
-mediation}\label{the-influence-of-measurement-error-on-mediation}
-
-Measurement error of an exposure can be modeled as some transformation
-of the true value that leads to the observed value, \(x_o = f(x)\). For
-example, we can define
-\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
-\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
-measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
-represents the imprecision (or noise) in the measurement of \(x\). Here
-the true value of the exposure is partially explained by the genetic
-instrument, \(g\), such that
-
-\[
-x = \alpha_g + \beta_g g + \epsilon_g
-\]
-
-where \(\beta_g\) is the effect of the SNP on the exposure, and
-\(\epsilon_g\) is the normally distributed residual value of \(x\); and
-the outcome is partially explained by the exposure
-
-\[
-y = \alpha_x + \beta_x x + \epsilon_x
-\]
-
-where \(\beta_x\) is the true effect of the exposure on the outcome. In
-the causal inference test (CIT), an omnibus p-value is generated from
-four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
-associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
-is independent of \(y|x\). The 4th condition employs mediation for
-causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
-error in scale and imprecision is introduced, such that \(y_o\) is the
-measured value of \(y\), it can be shown using basic covariance
-properties (Appendix 1) that
-
-\[
-\begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
-                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
-\end{aligned}
-\]
-
-where
-
-\[
-D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
-\]
-
-Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
-when the true model is causal only when \(D = 1\). Therefore, if there
-is any measurement error that incurs imprecision (i.e.
-\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
-between \(g\) and \(y_o | x\), which is in violation of the the 4th
-condition of the CIT. Note that scale transformation of \(x\) or \(y\)
-without any incurred imprecision is insufficient to lead to a violation
-of the test statistic assumptions, and henceforth mention of measurement
-error will relate to imprecision unless otherwise stated.
-
-We performed simulations to verify that this problem does arise using
-the CIT method. Figure 2 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
-reliable in identifying the correct causal direction. However, as
-measurement error increases in the exposure variable, eventually the CIT
-is more likely to infer a robust causal association in the wrong
-direction. Also of concern here is that increasing sample size does not
-solve the issue, indeed it only strengthens the apparent evidence for
-the incorrect inference.
-
-\subsubsection{Using MR to infer the existence of
-causality}\label{using-mr-to-infer-the-existence-of-causality}
-
-When selecting an instrument \(g\) that has a direct influence on \(x\)
-where \(x\) is causally related to \(y\), one can reason that the
-influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
-\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
-effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as
-
-\[
-\begin{aligned}
-\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
-                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
-                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
-\end{aligned}
-\]
-
-Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
-scaled by the measurement error in \(x\) and \(y\), but is unrelated to
-measurement imprecision. Measurement error and imprecision are likely to
-reduce the power of MR inferring a causal association (increased false
-negative rate), but importantly the evidence for causal association,
-based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
-inflated when the null hypothesis is true.
-
-We performed simulations to compare the performance of MR against CIT in
-detecting a causal association between simulated variables under
-different levels of imprecision simulated in the exposure. Figure 4
-shows the true positive rates between the CIT and MR for detecting a
-causal association. We observe that the CIT has lower power in all
-cases, with performance declining as measurement imprecision increases
-in the exposure. When MR assumptions are satisfied, notably that it is
-known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
-the performance of MR in detecting an association is unrelated to
-measurement error in the exposure. Measurement error in the outcome does
-reduce power, but does not induce a substantive difference in
-performance between CIT and MR.
 

### 2017-02-15 docs/manuscript.Rmd
updated

@@
 
-where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$
-
-MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
+where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
 
@@
 
-where $\beta_{MR} \neq 0$.
+where $\beta_{MR} \neq 0$ denotes the existence of causality, and the $\beta_{MR}$ is an estimate of the causal effect.
 
-Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$.
-For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
-Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that
+Measurement error of an exposure can be modeled as a transformation of the true value that leads to the observed value, $x_o = f(x)$.
+For example, following Pierce and VanderWeele [-@Pierce2012] we can define
 
 $$
-x = \alpha_g + \beta_g g + \epsilon_g
+f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}
 $$
 
-where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure
-
-$$
-y = \alpha_x + \beta_x x + \epsilon_x
-$$
-
-where $\beta_x$ is the true effect of the exposure on the outcome.
-In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
-The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
-                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
-\end{aligned}
-$$
-
-where
-
-$$
-D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
-$$
-
-Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
-Therefore, if there is any measurement error that incurs imprecision (i.e.
-$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
-Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
-
-We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
-However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
-Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
-
-When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$.
-Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as
-
-$$
-\begin{aligned}
-\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
-                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
-                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
-\end{aligned}
-$$
-
-Thus we obtain an estimate of the effect of $x_o$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision.
-Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.
-
-We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
-`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
-We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
-When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
-Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
+where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
+The same model of measurement error can be applied to the outcome variable $y$.
+In this study we assume there is no measurement error in the SNP.
 
@@
 ## Results
+
+### Mediation-based causal inference under measurement error
+
+In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
+The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
+
+$$
+\begin{aligned}
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
+\end{aligned}
+$$
+
+where
+
+$$
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+$$
+
+Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
+Therefore, if there is any measurement error that incurs imprecision (i.e.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
+Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
+
+We performed simulations to verify that this problem does arise using the CIT method.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
+Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
+
+We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
+We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
+When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
+Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
 

### 2017-02-18 docs/manuscript.Rmd
updated

@@
 
-The analysis proceeds by simulating variables that have known causal effects and valid instruments, and then applying measurement error in different configurations.
-We then assess the performance of the CIT test and MR analysis for a) inference of a causal relationship and b) inferring the correct direction of causality.
-Included in the MR analysis is a simple extension that is used to make inference of the causal direction when it is unknown which of the correlated variables the instrument is directly acting upon.
-All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
-
 ### CIT test
@@
 
-### Steiger sensitivity analysis
+### Causal direction sensitivity analysis
 
@@
 It is possible to simulate what the inferred causal direction would be for all values within these bounds.
-To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
+To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
 We also do the same for the measurement error values which are not compatible.
@@
 
+All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
+
 ## Results
@@
 
-### Using MR to infer the direction of causality
+### Using MR Steiger to infer the direction of causality
 
-If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required.
-If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that $\rho_{g,x}$ to be greater than $\rho_{g,y}$ because the association of $g$ with $y$ is the proportion of the true causal association between $x$ and $y$ scaled by the association between $g$ and $x$.
-We can use the Steiger test of correlated correlations to formally test the difference in magnitude of $\rho_{g,x}$ and $\rho_{g,y}$, where the variable with the largest correlation on $g$ is deemed to be the variable which is imposing the causal effect.
+If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
+Here we introduce the MR Steiger approach to also orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.
 
-This is a simple method to orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.
-But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
 Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0.
@@
 
-The Steiger test is used to infer if the variant $g$ has a direct influence on $x$ or $y$ when it is known that it associates with both, but the direction of causality between $x$ and $y$ is unknown.
+Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude.
+The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.
+
 Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:
@@
 
-In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
+In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
 
@@
 
-In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.
+In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.
 

### 2017-02-18 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{15 February 2017}
+  \date{18 February 2017}
 
@@
 
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\)
-
-MR analysis estimates the causal influence of \(x\) on \(y\) by using
-the instrument as a proxy for \(x\), such that
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). MR analysis
+estimates the causal influence of \(x\) on \(y\) by using the instrument
+as a proxy for \(x\), such that
 
@@
 
-where \(\beta_{MR} \neq 0\).
-
-Measurement error of an exposure can be modeled as some transformation
-of the true value that leads to the observed value, \(x_o = f(x)\). For
-example, we can define
-\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
-\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
+where \(\beta_{MR} \neq 0\) denotes the existence of causality, and the
+\(\beta_{MR}\) is an estimate of the causal effect.
+
+Measurement error of an exposure can be modeled as a transformation of
+the true value that leads to the observed value, \(x_o = f(x)\). For
+example, following Pierce and VanderWeele\textsuperscript{35} we can
+define
+
+\[
+f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}
+\]
+
+where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
 measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
-represents the imprecision (or noise) in the measurement of \(x\). Here
-the true value of the exposure is partially explained by the genetic
-instrument, \(g\), such that
-
-\[
-x = \alpha_g + \beta_g g + \epsilon_g
-\]
-
-where \(\beta_g\) is the effect of the SNP on the exposure, and
-\(\epsilon_g\) is the normally distributed residual value of \(x\); and
-the outcome is partially explained by the exposure
-
-\[
-y = \alpha_x + \beta_x x + \epsilon_x
-\]
-
-where \(\beta_x\) is the true effect of the exposure on the outcome. In
-the causal inference test (CIT), an omnibus p-value is generated from
-four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
-associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
-is independent of \(y|x\). The 4th condition employs mediation for
-causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
-error in scale and imprecision is introduced, such that \(y_o\) is the
-measured value of \(y\), it can be shown using basic covariance
-properties (Appendix 1) that
-
-\[
-\begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
-                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
-\end{aligned}
-\]
-
-where
-
-\[
-D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
-\]
-
-Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
-when the true model is causal only when \(D = 1\). Therefore, if there
-is any measurement error that incurs imprecision (i.e.
-\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
-between \(g\) and \(y_o | x\), which is in violation of the the 4th
-condition of the CIT. Note that scale transformation of \(x\) or \(y\)
-without any incurred imprecision is insufficient to lead to a violation
-of the test statistic assumptions, and henceforth mention of measurement
-error will relate to imprecision unless otherwise stated.
-
-We performed simulations to verify that this problem does arise using
-the CIT method. Figure 2 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
-reliable in identifying the correct causal direction. However, as
-measurement error increases in the exposure variable, eventually the CIT
-is more likely to infer a robust causal association in the wrong
-direction. Also of concern here is that increasing sample size does not
-solve the issue, indeed it only strengthens the apparent evidence for
-the incorrect inference.
-
-When selecting an instrument \(g\) that has a direct influence on \(x\)
-where \(x\) is causally related to \(y\), one can reason that the
-influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
-\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
-effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as
-
-\[
-\begin{aligned}
-\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
-                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
-                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
-\end{aligned}
-\]
-
-Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
-scaled by the measurement error in \(x\) and \(y\), but is unrelated to
-measurement imprecision. Measurement error and imprecision are likely to
-reduce the power of MR inferring a causal association (increased false
-negative rate), but importantly the evidence for causal association,
-based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
-inflated when the null hypothesis is true.
-
-We performed simulations to compare the performance of MR against CIT in
-detecting a causal association between simulated variables under
-different levels of imprecision simulated in the exposure. Figure 4
-shows the true positive rates between the CIT and MR for detecting a
-causal association. We observe that the CIT has lower power in all
-cases, with performance declining as measurement imprecision increases
-in the exposure. When MR assumptions are satisfied, notably that it is
-known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
-the performance of MR in detecting an association is unrelated to
-measurement error in the exposure. Measurement error in the outcome does
-reduce power, but does not induce a substantive difference in
-performance between CIT and MR.
+represents the imprecision (or noise) in the measurement of \(x\). The
+same model of measurement error can be applied to the outcome variable
+\(y\). In this study we assume there is no measurement error in the SNP.
 
 \subsection{Methods}\label{methods}
-
-The analysis proceeds by simulating variables that have known causal
-effects and valid instruments, and then applying measurement error in
-different configurations. We then assess the performance of the CIT test
-and MR analysis for a) inference of a causal relationship and b)
-inferring the correct direction of causality. Included in the MR
-analysis is a simple extension that is used to make inference of the
-causal direction when it is unknown which of the correlated variables
-the instrument is directly acting upon. All analysis was performed using
-the R programming language\textsuperscript{35} and code is made
-available at \href{}{github url to go here} and implemented in the
-MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.
 
@@
 
-\subsubsection{Steiger sensitivity
-analysis}\label{steiger-sensitivity-analysis}
+\subsubsection{Causal direction sensitivity
+analysis}\label{causal-direction-sensitivity-analysis}
 
@@
 possible to simulate what the inferred causal direction would be for all
-values within these bounds. To evaluate how reliable, \(R\), the Steiger
-test is to potential measurement imprecision in \(x\) and \(y\) we
-integrate over the simulated \(R\) values for the possible measurement
-error values which are compatible with the direction of causality
-obtained empirically from the Steiger test. We also do the same for the
-measurement error values which are not compatible. The ratio of these
-quantities is an indication of how much of the measurement error space
-supports the inferred causal direction. A ratio \(R=1\) indicates that
-the Steiger test is highly sensitive to measurement error, with equal
-weight of the measurement error parameter space supporting both
-directions of causality. Full details are provided in Appendix 2 and
-Figure 3a.
+values within these bounds. To evaluate how reliable, \(R\), the
+inference of the causal direction is to potential measurement
+imprecision in \(x\) and \(y\) we integrate over the simulated \(R\)
+values for the possible measurement error values which are compatible
+with the direction of causality obtained empirically from the Steiger
+test. We also do the same for the measurement error values which are not
+compatible. The ratio of these quantities is an indication of how much
+of the measurement error space supports the inferred causal direction. A
+ratio \(R=1\) indicates that the Steiger test is highly sensitive to
+measurement error, with equal weight of the measurement error parameter
+space supporting both directions of causality. Full details are provided
+in Appendix 2 and Figure 3a.
 
@@
 
+All analysis was performed using the R programming
+language\textsuperscript{41} and code is made available at
+\href{}{github url to go here} and implemented in the MR-Base
+(\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.
+
 \subsection{Results}\label{results}
 
-\subsubsection{Using MR to infer the direction of
-causality}\label{using-mr-to-infer-the-direction-of-causality}
-
-If we do not know whether the SNP \(g\) has a direct influence on \(x\)
-or \(y\) then further efforts are required. If \(x\) causes \(y\) and
-\(g\) is a valid instrument for \(x\) then we expect that \(\rho_{g,x}\)
-to be greater than \(\rho_{g,y}\) because the association of \(g\) with
-\(y\) is the proportion of the true causal association between \(x\) and
-\(y\) scaled by the association between \(g\) and \(x\). We can use the
-Steiger test of correlated correlations to formally test the difference
-in magnitude of \(\rho_{g,x}\) and \(\rho_{g,y}\), where the variable
-with the largest correlation on \(g\) is deemed to be the variable which
-is imposing the causal effect.
-
-This is a simple method to orient the direction of causality in an MR
-analysis when the underlying biology of the SNP is not fully understood.
-But it can be shown that in the presence of measurement imprecision,
+\subsubsection{Mediation-based causal inference under measurement
+error}\label{mediation-based-causal-inference-under-measurement-error}
+
+In the causal inference test (CIT), an omnibus p-value is generated from
+four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
+associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
+is independent of \(y|x\). The 4th condition employs mediation for
+causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
+where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
+error in scale and imprecision is introduced, such that \(y_o\) is the
+measured value of \(y\), it can be shown using basic covariance
+properties (Appendix 1) that
+
+\[
+\begin{aligned}
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
+\end{aligned}
+\]
+
+where
+
+\[
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+\]
+
+Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
+when the true model is causal only when \(D = 1\). Therefore, if there
+is any measurement error that incurs imprecision (i.e.
+\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
+between \(g\) and \(y_o | x\), which is in violation of the the 4th
+condition of the CIT. Note that scale transformation of \(x\) or \(y\)
+without any incurred imprecision is insufficient to lead to a violation
+of the test statistic assumptions, and henceforth mention of measurement
+error will relate to imprecision unless otherwise stated.
+
+We performed simulations to verify that this problem does arise using
+the CIT method. Figure 2 shows that when there is no measurement error
+in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
+reliable in identifying the correct causal direction. However, as
+measurement error increases in the exposure variable, eventually the CIT
+is more likely to infer a robust causal association in the wrong
+direction. Also of concern here is that increasing sample size does not
+solve the issue, indeed it only strengthens the apparent evidence for
+the incorrect inference.
+
+We performed simulations to compare the performance of MR against CIT in
+detecting a causal association between simulated variables under
+different levels of imprecision simulated in the exposure. Figure 4
+shows the true positive rates between the CIT and MR for detecting a
+causal association. We observe that the CIT has lower power in all
+cases, with performance declining as measurement imprecision increases
+in the exposure. When MR assumptions are satisfied, notably that it is
+known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
+the performance of MR in detecting an association is unrelated to
+measurement error in the exposure. Measurement error in the outcome does
+reduce power, but does not induce a substantive difference in
+performance between CIT and MR.
+
+\subsubsection{Using MR Steiger to infer the direction of
+causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}
+
+If we do not know whether the SNP \(g\) has a primary influence on \(x\)
+or \(y\) then CIT can attempt to infer the causal direction. Here we
+introduce the MR Steiger approach to also orient the direction of
+causality in an MR analysis when the underlying biology of the SNP is
+not fully understood.
+
+We show that in the presence of measurement imprecision,
 \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
@@
 levels at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation\textsuperscript{41}.
+levels, consistent with expectation\textsuperscript{42}.
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored\textsuperscript{42}. Our analysis
+analysis is relatively well explored\textsuperscript{43}. Our analysis
 extends this to related methods such as CIT that are widely used in
@@
 evaluated measurement error in these platforms do support this
-position\textsuperscript{43,44}, though making strong conclusions for
+position\textsuperscript{44,45}, though making strong conclusions for
 this analysis is difficult because measurement error is likely to be
@@
 can offer solutions to inferring the causal
-direction\textsuperscript{45}. In this work we assumed that pleiotropy
+direction\textsuperscript{46}. In this work we assumed that pleiotropy
 (the influence of the instrument on the outcome through a mechanism
@@
 the same issues of measurement error. Recent advances in MR methodology,
-for example applying MR to genetical genomics\textsuperscript{46},
-multivariate MR\textsuperscript{47} and mediation through
-MR\textsuperscript{48--50} may offer more robust alternatives for these
+for example applying MR to genetical genomics\textsuperscript{47},
+multivariate MR\textsuperscript{48} and mediation through
+MR\textsuperscript{49--51} may offer more robust alternatives for these
 more complicated problems.
@@
 
-The Steiger test is used to infer if the variant \(g\) has a direct
-influence on \(x\) or \(y\) when it is known that it associates with
-both, but the direction of causality between \(x\) and \(y\) is unknown.
+Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
+causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
+and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
+hypothesis test that provides a p-value for observing the difference in
+these correlations under the null hypothesis that they are equal.
+
 Assuming the causal direction is \(x \to y\), two stage MR is formulated
@@
 In the presence of measurement error in \(x\) and \(y\), however, the
-Steiger test will instead be assessing the inequality
-\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
+empirical inference of the causal direction will instead be based on
+evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
 
@@
 
-In order to assess how reliable the Steiger test is in the presence of
-measurement imprecision, we can evaluate the range of potential values
-of measurement error in \(x\) and \(y\) over which the empirical result
-from the Steiger test would return the wrong value.
+In order to assess how reliable the inference of the causal direction is
+in the presence of measurement imprecision, we can evaluate the range of
+potential values of measurement error in \(x\) and \(y\) over which the
+empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
+return the wrong causal direction.
 
@@
 \hypertarget{ref-Millstein2016}{}
-18. Millstein, J. cit: Causal Inference Test. R package version 1.9.
-(2016). at
-\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}
+18. Millstein, J., Chen, G. K. \& Breton, C. V. Cit : Hypothesis Testing
+Software for Mediation Analysis in Genomic Applications.
+\emph{Bioinformatics} \textbf{AOP,} 1--10 (2016).
 
@@
 
-\hypertarget{ref-RCoreTeam2015}{}
-35. R Core Team. R: A Language and Environment for Statistical
-Computing. (2015). at
-\textless{}\url{https://www.r-project.org/}\textgreater{}
+\hypertarget{ref-Pierce2012}{}
+35. Pierce, B. L. \& VanderWeele, T. J. The effect of non-differential
+measurement error on bias, precision and power in Mendelian
+randomization studies. \emph{International Journal of Epidemiology}
+\textbf{41,} 1383--1393 (2012).
 
@@
 \hypertarget{ref-Revelle2015}{}
-39. Revelle, W. psych: Procedures for Psychological, Psychometric, and
-Personality Research. (2015). at
+39. Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
+and Personality Research}. (Northwestern University, 2015). at
 \textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
@@
 
+\hypertarget{ref-RCoreTeam2015}{}
+41. R Core Team. \emph{R: A Language and Environment for Statistical
+Computing}. (R Foundation for Statistical Computing, 2015). at
+\textless{}\url{https://www.r-project.org/}\textgreater{}
+
 \hypertarget{ref-Bird2002}{}
-41. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
+42. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
 \& development} \textbf{16,} 6--21 (2002).
@@
 \hypertarget{ref-Cole2014}{}
-42. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
+43. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
 Potentially Serious and Misleading Consequences Due to Uncorrected
@@
 \hypertarget{ref-Bose2014}{}
-43. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
+44. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
 methylation measurement using technical replicates: the Atherosclerosis
@@
 \hypertarget{ref-Bryant2011}{}
-44. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
+45. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
 Biological Variability in a Microarray Experiment but Both Are
@@
 \hypertarget{ref-Richmond2014}{}
-45. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
+46. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
 between Child Adiposity and Physical Activity Levels: A Mendelian
@@
 \hypertarget{ref-Relton2012}{}
-46. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+47. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-47. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+48. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Varbo2015}{}
-48. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+49. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Burgess2015}{}
-49. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+50. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 \hypertarget{ref-Richmond2016}{}
-50. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
+51. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
 C. L. Challenges and novel approaches for investigating molecular

### 2017-02-18 docs/manuscript.Rmd
updated

@@
 fig_nums(
+	"causality_exists_tpr",
+	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
+True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+	display=FALSE
+)
+
+fig_nums(
 	"steiger_sensitivity_plot", 
@@
 In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$.
-b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger reliability values (y-axis) for different values of $\\rho_{g,x}^2$.
-Inferring the direction of causality is more sensitive to measurement error when there are stronger causal associations between X and Y.",
-	display=FALSE
-)
-
-fig_nums(
-	"causality_exists_tpr",
-	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative.
+When $d$ is negative the Steiger test is liable to infer the wrong direction of causality.
+Shaded regions show the parameter space where $d$ is negative.
+The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
+	",
 	display=FALSE
@@
 An established study design that can provide information about causality is randomisation.
-Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control.
+Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and treating the other as a control.
 The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship.
@@
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
-Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable.
+Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012].
 Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
@@
 Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method.
-We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error.
-This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
+We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a matric to evaluate the sensitivity of the result of this extension to measurement error.
+Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
 Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
@@
 
-where $\beta_{MR} \neq 0$ denotes the existence of causality, and the $\beta_{MR}$ is an estimate of the causal effect.
+where $\beta_{MR} \neq 0$ denotes the existence of causality, and $\beta_{MR}$ is an estimate of the causal effect.
 
@@
 The same model of measurement error can be applied to the outcome variable $y$.
-In this study we assume there is no measurement error in the SNP.
+In this study we assume there is no measurement error in the SNP and measurement error in the exposure and the outcome are uncorrelated.
 
@@
 
-The method that we will now describe is able to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$.
+The method that we will now describe is designed to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$.
 Unlike the CIT framework, this approach cannot infer if the true model is $x \leftarrow g \rightarrow y$.
@@
 A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
-Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.
+Full details are provided in Appendix 2.
 
@@
 
-In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$.
-The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
+In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
 When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
@@
 If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
-Here we introduce the MR Steiger approach to also orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.
+Here we introduce the MR Steiger approach to similarly orient the direction of causality but in an MR analysis when the underlying biology of the SNP is not fully understood.
 
-We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
-Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0.
-However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
+For a particular association, it is of interest to identify the range of possible measurement error values agree and disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2).
+This metric can be used to evaluate the reliability of MR Steiger.
+
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
+`r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
 It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.
 
-We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect.
+### Comparison of CIT and MR Steiger for obtaining the correct direction of causality
+
+We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect.
 Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
@@
 When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
-The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate this confounding model as being causal.
+The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate erroneously this confounding model as being causal.
 
@@
 
-We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
+We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
 We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
@@
 
+```{r causality_exists_tpr, fig.width=10, fig.height=10 }
+
+```
+
+`r fig_nums("causality_exists_tpr")`
+
+\newpage
+
 ```{r steiger_sensitivity_plot, fig.width=12, fig.height=6 }
@@
 `r fig_nums("steiger_sensitivity_plot")`
-
-\newpage
-
-```{r causality_exists_tpr, fig.width=10, fig.height=10 }
-
-```
-
-`r fig_nums("causality_exists_tpr")`
 

### 2017-02-18 docs/manuscript.tex
updated

@@
 nature of the association by randomly splitting the sample into two
-groups, subjecting one group to the exposure and leaving the other as a
+groups, subjecting one group to the exposure and treating the other as a
 control. The association between the exposure and the outcome in this
@@
 Additionally, when MR assumptions are satisfied the method is robust to
-there being measurement error in the exposure variable. Indeed
-instrumental variable (IV) analysis was in part initially introduced as
-a correction for measurement error in the exposure\textsuperscript{29},
-whereas it has been noted that both classic mediation-based
-analyses\textsuperscript{13,14,30,31} and mediation-based methods that
-use instrumental variables\textsuperscript{32,33} are prone to be
-unreliable in its presence.
+there being measurement error in the exposure
+variable\textsuperscript{29}. Indeed instrumental variable (IV) analysis
+was in part initially introduced as a correction for measurement error
+in the exposure\textsuperscript{30}, whereas it has been noted that both
+classic mediation-based analyses\textsuperscript{13,14,31,32} and
+mediation-based methods that use instrumental
+variables\textsuperscript{33,34} are prone to be unreliable in its
+presence.
 
@@
 even when the biology of the instruments are not fully understood, and
-also a method to evaluate the sensitivity of the result of this
-extension to measurement error. This improves the utility of MR in cases
-where mediation based methods might have otherwise been used
-preferentially. Finally, we apply this new method to infer the direction
-of causation between DNA methylation levels and gene expression levels.
-Our analyses highlight that because these different causal inference
-techniques have varying strengths and weaknesses, triangulation of
-evidence from as many sources as possible should be practiced in causal
-inference\textsuperscript{34}.
+also a matric to evaluate the sensitivity of the result of this
+extension to measurement error. Together these extensions improve the
+utility of MR in cases where mediation based methods might have
+otherwise been used preferentially. Finally, we apply this new method to
+infer the direction of causation between DNA methylation levels and gene
+expression levels. Our analyses highlight that because these different
+causal inference techniques have varying strengths and weaknesses,
+triangulation of evidence from as many sources as possible should be
+practiced in causal inference\textsuperscript{35}.
 
@@
 
-where \(\beta_{MR} \neq 0\) denotes the existence of causality, and the
+where \(\beta_{MR} \neq 0\) denotes the existence of causality, and
 \(\beta_{MR}\) is an estimate of the causal effect.
@@
 the true value that leads to the observed value, \(x_o = f(x)\). For
-example, following Pierce and VanderWeele\textsuperscript{35} we can
+example, following Pierce and VanderWeele\textsuperscript{29} we can
 define
@@
 same model of measurement error can be applied to the outcome variable
-\(y\). In this study we assume there is no measurement error in the SNP.
+\(y\). In this study we assume there is no measurement error in the SNP
+and measurement error in the exposure and the outcome are uncorrelated.
 
@@
 
-The method that we will now describe is able to distinguish between two
-models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
+The method that we will now describe is designed to distinguish between
+two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
 framework, this approach cannot infer if the true model is
@@
 space supporting both directions of causality. Full details are provided
-in Appendix 2 and Figure 3a.
+in Appendix 2.
 
@@
 
-In the causal inference test (CIT), an omnibus p-value is generated from
-four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
-associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
-is independent of \(y|x\). The 4th condition employs mediation for
-causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
-error in scale and imprecision is introduced, such that \(y_o\) is the
+In the causal inference test (CIT), the 4th condition (see Methods)
+employs mediation for causal inference, and can be expressed as
+\(cov(g, y - \hat{y}) = 0\), where
+\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement error
+in scale and imprecision is introduced, such that \(y_o\) is the
 measured value of \(y\), it can be shown using basic covariance
@@
 detecting a causal association between simulated variables under
-different levels of imprecision simulated in the exposure. Figure 4
+different levels of imprecision simulated in the exposure. Figure 3
 shows the true positive rates between the CIT and MR for detecting a
@@
 or \(y\) then CIT can attempt to infer the causal direction. Here we
-introduce the MR Steiger approach to also orient the direction of
-causality in an MR analysis when the underlying biology of the SNP is
-not fully understood.
+introduce the MR Steiger approach to similarly orient the direction of
+causality but in an MR analysis when the underlying biology of the SNP
+is not fully understood.
+
+For a particular association, it is of interest to identify the range of
+possible measurement error values agree and disagree with the
+empirically inferred causal direction (Figure 4a, Appendix 2). This
+metric can be used to evaluate the reliability of MR Steiger.
 
 We show that in the presence of measurement imprecision,
-\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
-is a metric of the degree to which the Steiger test is liable to provide
-the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
-likely to be correct about the causal direction). Thus it is important
-to note that under certain conditions when measurement imprecision is
-present the Steiger test could make erroneous inference about the causal
-direction because the value of \(d\) is not restricted to being greater
-than 0. However Figure 7 shows that when there is no measurement error
-in \(x\), the Steiger test is unlikely to infer the wrong direction of
-causality even if there is measurement error in \(y\). It also shows
-that in most cases where \(x\) is measured with error, especially when
-the causal effect between \(x\) and \(y\) is not very large, the
-sensitivity of the Steiger test to measurement error is relatively low.
-
-We performed simulations to explore the performance of this approach in
-comparison to CIT in terms of the rate at which a strong causal
-relationship is obtained for the correct direction of causality, and the
-rate at which a strong causal relationship is obtained where the
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
+the range of parameters around which the Steiger test is liable to
+provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
+the Steiger test is likely to be correct about the causal direction).
+Figure 4b shows that when there is no measurement error in \(x\), the
+Steiger test is unlikely to infer the wrong direction of causality even
+if there is measurement error in \(y\). It also shows that in most cases
+where \(x\) is measured with error, especially when the causal effect
+between \(x\) and \(y\) is not very large, the sensitivity of the
+Steiger test to measurement error is relatively low.
+
+\subsubsection{Comparison of CIT and MR Steiger for obtaining the
+correct direction of
+causality}\label{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}
+
+We used simulations to explore the performance of the MR Steiger
+approach in comparison to CIT in terms of the rate at which a strong
+causal relationship is obtained for the correct direction of causality,
+and the rate at which a strong causal relationship is obtained where the
 reported direction of causality is incorrect. Simulations were performed
@@
 ``causal model'' (Methods, Figure 5b), but it is evident that
-measurement error will often lead the CIT to evaluate this confounding
-model as being causal.
+measurement error will often lead the CIT to evaluate erroneously this
+confounding model as being causal.
 
@@
 direction commonly goes in both directions (Figure 6a), but assuming no
-or equal measurement error DNA methylation levels were the predominant
+or equal measurement error, DNA methylation levels were the predominant
 causal factor (\(p = 1.3\times 10^{-5}\)). We then went on to predict
@@
 
+\includegraphics{manuscript_files/figure-latex/causality_exists_tpr-1.pdf}
+
+Figure 3: Outcomes were simulated to be causally influenced by exposures
+with varying degrees of measurement imprecision applied to the exposure
+variable (x axis). True positive rates (y axis) for MR and CIT were
+compared for varying levels of measurement imprecision in the outcome
+variable (rows of boxes) and sample sizes (columns of boxes).
+
+\newpage
+
 \includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
 
-Figure 3: a) We can predict the values the Steiger test would take
+Figure 4: a) We can predict the values the Steiger test would take
 (z-axis) for different potential values of measurement error (x and y
@@
 measurement error values are in support of the \(x \rightarrow y\)
-direction of causality than \(y \rightarrow x\). b) The influence of
-\(\rho_{x,y}\) (x-axis) on the Steiger reliability values (y-axis) for
-different values of \(\rho_{g,x}^2\). Inferring the direction of
-causality is more sensitive to measurement error when there are stronger
-causal associations between X and Y.
-
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/causality_exists_tpr-1.pdf}
-
-Figure 4: Outcomes were simulated to be causally influenced by exposures
-with varying degrees of measurement imprecision applied to the exposure
-variable (x axis). True positive rates (y axis) for MR and CIT were
-compared for varying levels of measurement imprecision in the outcome
-variable (rows of boxes) and sample sizes (columns of boxes).
+direction of causality than \(y \rightarrow x\). b) Plots depicting the
+parameter space in which the function
+\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
+negative the Steiger test is liable to infer the wrong direction of
+causality. Shaded regions show the parameter space where \(d\) is
+negative. The graph shows that for the majority of the parameter space
+of the function, \(d\) is positive, especially where causal
+relationships are relatively weak.
 
@@
 direction of causality. \(V_{z}\), the total volume of the function
-(Figure 3), can be obtained analytically by solving:
+(Figure 4), can be obtained analytically by solving:
 
@@
 
+\hypertarget{ref-Pierce2012}{}
+29. Pierce, B. L. \& VanderWeele, T. J. The effect of non-differential
+measurement error on bias, precision and power in Mendelian
+randomization studies. \emph{International Journal of Epidemiology}
+\textbf{41,} 1383--1393 (2012).
+
 \hypertarget{ref-Ashenfelter1994}{}
-29. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
+30. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
 to Schooling from a New Sample of Twins. \emph{The American Economic
@@
 \hypertarget{ref-Lawlor2017}{}
-34. Lawlor, D. A., Tilling, K. \& Davey Smith, G. Triangulation in
+35. Lawlor, D. A., Tilling, K. \& Davey Smith, G. Triangulation in
 aetiological epidemiology. \emph{International Journal of Epidemiology}
 \textbf{19,} dyw314 (2017).
-
-\hypertarget{ref-Pierce2012}{}
-35. Pierce, B. L. \& VanderWeele, T. J. The effect of non-differential
-measurement error on bias, precision and power in Mendelian
-randomization studies. \emph{International Journal of Epidemiology}
-\textbf{41,} 1383--1393 (2012).
 

### 2017-02-18 scripts/mr_directionality_analysis.R

@@ -962,12 +962,62 @@ p1 <- example_sensitivity$sensitivity_plot
 p2 <- ggplot(sensitivity_parameters, aes(y=sensitivity, x=r_xy, group=factor(r_gx^2))) +
 	geom_point(aes(colour=factor(r_gx^2))) +
 	geom_line(aes(colour=factor(r_gx^2))) +
-	labs(x=expression(rho[xy]), y=expression(V[z<0]/V[z>=0]), colour=expression(rho[gx]^2)) +
-	scale_colour_brewer(type="qual")
+	labs(x=expression(rho[xy]), y=expression(V[z>=0]/-V[z<0]), colour=expression(rho[gx]^2)) +
+	scale_colour_brewer(type="qual") +
+	scale_y_log10()
+
+ineq <- expand.grid(
+	cora = seq(0, 1, by=0.02),
+	corb = seq(0, 1, by=0.2),
+	ab = c(0.1, 0.5, 0.9)
+)
+
+ineq$lhs <- ineq$cora
+ineq$rhs <- ineq$corb * ineq$ab
+ineq$d <- ineq$lhs - ineq$rhs
+ineq$ablab <- paste0("cor(x,y) = ", ineq$ab)
+
+
+area <- group_by(ineq, ablab) %>%
+	do({
+
+		temp1 <- subset(., corb == max(corb) & d <= 0)
+
+		y1 <- min(temp1$d)
+		x1 <- min(temp1$cora)
+
+		y2 <- min(abs(temp1$d), 0)
+		x2 <- max(temp1$cora)
+
+		temp2 <- subset(., corb == min(corb))
+		y3 <- min(abs(temp2$d), 0)
+		x3 <- min(temp2$cora)
+
+		data.frame(cora=c(x1, x2, x3), d=c(y1, y2, y3), f=1)
+	})
+
+
+
+p3 <- ggplot(ineq, aes(x=cora, y=d)) +
+geom_polygon(data=area, fill="grey", aes(x=cora, y=d)) +
+geom_line(aes(colour=corb, group=corb)) +
+geom_hline(yintercept=0, linetype=2) +
+facet_wrap(~ ablab) +
+labs(x=TeX("$cor(x, x_o)$"), y="d", colour=TeX("$cor(y, y_o)$")) +
+theme(axis.text.x=element_text(angle=45, vjust=0.5))
+
+
+suppressPackageStartupMessages(library(grid))
[…]

### 2017-02-19 docs/manuscript.Rmd
updated

@@
 
-We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
-We then went on to predict the causal directions of the associations for varying levels of possible measurement error.
+We used the Steiger test to infer the direction of causality between `r nrow(qtl_orig)` putative associations between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
+The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (95% CI `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`).
+We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms.
 `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
 
-We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test.
+We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test.
 We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b).
@@
 This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest.
-But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation in have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
+But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
 Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods.
@@
 Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014].
+We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].
+
 In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
-Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored.
+Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored.
 An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.
@@
 
-The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly sensitive to measurement error.
+The overarching result from our simulations is that, regardless of the method used, inferring the causal direction using an instrument of unknown biology is highly sensitive to measurement error.
 With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

### 2017-02-19 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{18 February 2017}
+  \date{19 February 2017}
 
@@
 
-We used the Steiger test to infer the direction of causality between DNA
-methylation levels and gene expression levels and found that the causal
-direction commonly goes in both directions (Figure 6a), but assuming no
-or equal measurement error, DNA methylation levels were the predominant
-causal factor (\(p = 1.3\times 10^{-5}\)). We then went on to predict
-the causal directions of the associations for varying levels of possible
-measurement error. Figure 6a shows that the qualitative result of which
-of DNA methylation or gene expression is more commonly the causal factor
-is very strongly determined by measurement error.
+We used the Steiger test to infer the direction of causality between 458
+putative associations between DNA methylation levels and gene expression
+levels and found that the causal direction commonly goes in both
+directions (Figure 6a), but assuming no or equal measurement error, DNA
+methylation levels were the predominant causal factor
+(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
+was 3.92 (95\% CI 1.08 - 37.11). We then went on to predict the causal
+directions of the associations for varying levels of systematic
+measurement error for the different platforms. Figure 6a shows that the
+qualitative result of which of DNA methylation or gene expression is
+more commonly the causal factor is very strongly determined by
+measurement error.
 
 We performed two sample MR\textsuperscript{28} for each association in
-the direction of causality that we estimated to be the correct direction
-from the Steiger test. We observed that the sign of the MR estimate was
-generally in the same direction as the Pearson correlation coefficient
-reported by Shakhbazov et al\textsuperscript{40} (Figure 6b). There was
-a moderate correlation between the absolute magnitudes of the causal
-correlation and the observational Pearson correlation (r = 0.45).
-Together these inferences suggest that even for relatively low level
-associations it is important to use causal inference methods over
-observational associations to infer causal effect sizes.
+the direction of causality inferred by the Stieger test. We observed
+that the sign of the MR estimate was generally in the same direction as
+the Pearson correlation coefficient reported by Shakhbazov et
+al\textsuperscript{40} (Figure 6b). There was a moderate correlation
+between the absolute magnitudes of the causal correlation and the
+observational Pearson correlation (r = 0.45). Together these inferences
+suggest that even for relatively low level associations it is important
+to use causal inference methods over observational associations to infer
+causal effect sizes.
 
@@
 occasions, especially in the realm of 'omic data, this is not the case,
-and methods based on mediation in have been valuable in order to be able
-to both ascertain if there is a causal association and to infer the
+and methods based on mediation have been valuable in order to be able to
+both ascertain if there is a causal association and to infer the
 direction of causality. Here we have described a simple extension to MR
@@
 can offer solutions to inferring the causal
-direction\textsuperscript{46}. In this work we assumed that pleiotropy
-(the influence of the instrument on the outcome through a mechanism
-other than the exposure) was not present. Recent method developments in
-MR\textsuperscript{24,25} have focused on accounting for the issues that
-horizontal pleiotropy can introduce, but how they perform in the
-presence of measurement error remains to be explored. An important
-advantage that MR confers over most mediation based analysis is that it
-can be performed in two samples, which can considerably improve power
-and expand the scope of analysis. However, whether there is a
-substantive difference in two sample MR versus one sample MR in how
+direction\textsuperscript{46}. We restricted the simulations to
+evaluating the causal inference between quantitative traits, but it is
+possible that the analysis could be extended to binary traits by using
+the genetic variance explained on the liability scale, taking into
+account the population prevalence\textsuperscript{47}.
+
+In this work we assumed that pleiotropy (the influence of the instrument
+on the outcome through a mechanism other than the exposure) was not
+present. Recent method developments in MR\textsuperscript{24,25} have
+focused on accounting for the issues that horizontal pleiotropy can
+introduce when multiple instruments are available, but how they perform
+in the presence of measurement error remains to be explored. An
+important advantage that MR confers over most mediation based analysis
+is that it can be performed in two samples, which can considerably
+improve power and expand the scope of analysis. However, whether there
+is a substantive difference in two sample MR versus one sample MR in how
 measurement error has an effect is not yet fully understood. We have
@@
 the same issues of measurement error. Recent advances in MR methodology,
-for example applying MR to genetical genomics\textsuperscript{47},
-multivariate MR\textsuperscript{48} and mediation through
-MR\textsuperscript{49--51} may offer more robust alternatives for these
+for example applying MR to genetical genomics\textsuperscript{48},
+multivariate MR\textsuperscript{49} and mediation through
+MR\textsuperscript{50--52} may offer more robust alternatives for these
 more complicated problems.
@@
 The overarching result from our simulations is that, regardless of the
-method used, inferring causal direction using an instrument of unknown
-biology is highly sensitive to measurement error. With the presence of
-measurement error near ubiquitous in most observational data, and our
-ability to measure it limited, we argue that it needs to be central to
-any consideration of approaches which are used in attempt to strengthen
-causal inference, and any putative results should be accompanied with
-appropriate sensitivity analysis that assesses their robustness under
-varying levels of measurement error.
+method used, inferring the causal direction using an instrument of
+unknown biology is highly sensitive to measurement error. With the
+presence of measurement error near ubiquitous in most observational
+data, and our ability to measure it limited, we argue that it needs to
+be central to any consideration of approaches which are used in attempt
+to strengthen causal inference, and any putative results should be
+accompanied with appropriate sensitivity analysis that assesses their
+robustness under varying levels of measurement error.
 
@@
 
+\hypertarget{ref-Lee2013c}{}
+47. Lee, S. H. \& Wray, N. R. Novel genetic analysis for case-control
+genome-wide association studies: quantification of power and genomic
+prediction accuracy. \emph{PLoS One} \textbf{8,} e71494 (2013).
+
 \hypertarget{ref-Relton2012}{}
-47. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+48. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Burgess2014a}{}
-48. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
+49. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
 G. Using multivariable Mendelian randomization to disentangle the causal
@@
 \hypertarget{ref-Varbo2015}{}
-49. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+50. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
@@
 \hypertarget{ref-Burgess2015}{}
-50. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
+51. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
 Network Mendelian randomization: using genetic variants as instrumental
@@
 \hypertarget{ref-Richmond2016}{}
-51. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
+52. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
 C. L. Challenges and novel approaches for investigating molecular

### 2017-02-19 scripts/mr_directionality_analysis.R

@@ -842,6 +842,21 @@ shakhtest4 <- binom.test(
 	0.5
 )
 
+temp <- subset(shakhbazov, rxx_o==1 & ryy_o==1)
+#table(temp$steiger < 0.05 & temp$mr_p < 0.05)
+
+#hist(temp$reliability, breaks=50)
+
+#exp(mean(log(temp$reliability)))
+
+shakh_median_reliability <- median(temp$reliability)
+
+
+shakh_r_loci <- quantile(temp$reliability, c(0.025))
+shakh_r_upci <- quantile(temp$reliability, c(0.975))
+
+
+## ---- shakhtemp ----
 
 
 ## ---- shakhplot ----

### 2017-02-19 docs/manuscript.Rmd
updated

@@
 If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables.
-The causal inference test (CIT) attempts to delineate between them.
+The causal inference test (CIT) attempts to distinguish between them.
 a) Methylation causes gene expression.
@@
 	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface.
-When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction.
+When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction.
 Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y.
@@
 	"cit_mr_comparison_figure", 
-	caption="a) Outcomes were simulated to be causally influenced by exposures as shown in the graph, with varying degrees of measurement error applied to both.
+	caption="a) Outcome $y$ was simulated to be caused by exposure $x$ as shown in the graph, with varying degrees of measurement error applied to both.
 CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
@@
 
-In addition, the exposure is influenced by a SNP $g$ such that
+In addition, the exposure is influenced by a SNP $g$ with an effect of $\beta_g$ such that
 
@@
 
+The $\alpha_*$ terms represent intercepts, and the $\epsilon_*$ denote random error.
 Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that
@@
 The methodology of the CIT is as follows.
-Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ is causally related to an outcome $y$.
-Thus,
-
-$$
-\begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_g + \beta_g g + \epsilon_g \\
-y & = \alpha_x + \beta_x x + \epsilon_x \\
-\end{aligned}
-$$
-
+Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ causes an outcome $y$, as described above.
 The following tests are then performed:
@@
 where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
-If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$.
+If all four tests reject the null hypothesis then it is inferred that $x$ causes $y$.
 The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.
@@
 In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
-To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$, and for the direction model of $y$ causing $x$.
+To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$ (Model 2).
 For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:
 
-- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then the $x \rightarrow y$ model is accepted
-- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the $y \rightarrow x$ model is accepted
-- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no inference is made
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
+- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
+- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
 - If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).
@@
 
-Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. Assuming the following model
-
-$$
-y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
-$$
-
-where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected.
+Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample.
 A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].
@@
 - If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
-- Otherwise no inference is made
+- Otherwise no evidence for a causal relationship
 
@@
 However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely.
-If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$.
-As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$).
+If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} < \rho_{g,y_o}$ because $\rho_{g,x_o}$ could be underestimated more than $\rho_{g,y_o}$.
+
+As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
+The maximum measurement error of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$.
+The minimum is 0, denoting perfectly measured trait values (the same logic applies to $y_o$).
 It is possible to simulate what the inferred causal direction would be for all values within these bounds.
-To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test.
-We also do the same for the measurement error values which are not compatible.
-The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction.
-A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
+
+To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement error in $x$ and $y$ we need to predict the values of $\rho_{gy} - \rho_{gx}$ for those values of measurement error.
+We integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values.
+We find the ratio of the volume that agrees with the inferred direction of causality over the volume that disagrees with the inferred direction of causality.
+A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
+In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction.
 Full details are provided in Appendix 2.
@@
 
-Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels.
+Two sample MR [@Pierce2013] was performed using the summary statistics for genetic influences on gene expression and DNA methylation.
 To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
@@
 
-Where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality.
-To evaluate the reliability, $R$, of the Steiger test with regards to measurement error, the objective is to compare the weighted proportion of the parameter space that agrees with the inferred direction against the weighted proportion which does not:
+Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$.
+If $z < 0$ then we infer that $x \rightarrow y$.
+There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$.
+To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:
 
@@
 If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values.
-If $R > 1$ then $R$ times as much of the weighted parameter space favours the inferred direction of causality.
+If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality.
 $V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:

### 2017-02-19 docs/manuscript.tex
updated

@@
 
-In addition, the exposure is influenced by a SNP \(g\) such that
+In addition, the exposure is influenced by a SNP \(g\) with an effect of
+\(\beta_g\) such that
 
@@
 
-Mediation-based analyses that test whether \(x\) causally relates to
-\(y\) impinge on evaluating if the influence of \(g\) on \(y\) can be
-accounted for by conditioning on \(x\), such that
+The \(\alpha_*\) terms represent intercepts, and the \(\epsilon_*\)
+denote random error. Mediation-based analyses that test whether \(x\)
+causally relates to \(y\) impinge on evaluating if the influence of
+\(g\) on \(y\) can be accounted for by conditioning on \(x\), such that
 
@@
 follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
-the exposure \(x\) is causally related to an outcome \(y\). Thus,
-
-\[
-\begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_g + \beta_g g + \epsilon_g \\
-y & = \alpha_x + \beta_x x + \epsilon_x \\
-\end{aligned}
-\]
-
-The following tests are then performed:
+the exposure \(x\) causes an outcome \(y\), as described above. The
+following tests are then performed:
 
@@
 mediate the association between the SNP and the outcome. If all four
-tests reject the null hypothesis then it is inferred that \(x\) is
-causally related to \(y\). The CIT measures the strength of causality by
-generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
-(least extreme) p-value of the four tests, the intuition being that
-causal inference is only as strong as the weakest link in the chain of
-tests.
+tests reject the null hypothesis then it is inferred that \(x\) causes
+\(y\). The CIT measures the strength of causality by generating an
+omnibus p-value, \(p_{CIT}\), which is simply the largest (least
+extreme) p-value of the four tests, the intuition being that causal
+inference is only as strong as the weakest link in the chain of tests.
 
@@
 method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
-for the direction of \(x\) causing \(y\), and for the direction model of
-\(y\) causing \(x\). For some significance threshold \(\alpha\), the
-existence of causality and its direction was inferred based on the
-following scenarios:
+for the direction of \(x\) causing \(y\) (Model 1), and for the
+direction of \(y\) causing \(x\) (Model 2). For some significance
+threshold \(\alpha\), the existence of causality and its direction was
+inferred based on the following scenarios:
 
@@
   If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then the \(x \rightarrow y\)
-  model is accepted
+  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
 \item
   If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then the \(y \rightarrow x\)
-  model is accepted
+  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
 \item
   If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then no inference is made
+  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
+  relationship
 \item
@@
 performing MR when the exposure, outcome and instrument data are all
-available in the same sample. Assuming the following model
-
-\[
-y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
-\]
-
-where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
-relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
-is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
-R package \(R/systemfit\)\textsuperscript{37}.
+available in the same sample. A p-value for this test, \(p_{MR}\), was
+obtained using the R package \(R/systemfit\)\textsuperscript{37}.
 
@@
 \item
-  Otherwise no inference is made
+  Otherwise no evidence for a causal relationship
 \end{itemize}
@@
 imprecisely. If, for example, \(x\) has lower measurement precision than
-\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
-As we show in the appendix, the maximum measurement error of \(x_o\) is
-\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
-possible to simulate what the inferred causal direction would be for all
-values within these bounds. To evaluate how reliable, \(R\), the
-inference of the causal direction is to potential measurement
-imprecision in \(x\) and \(y\) we integrate over the simulated \(R\)
-values for the possible measurement error values which are compatible
-with the direction of causality obtained empirically from the Steiger
-test. We also do the same for the measurement error values which are not
-compatible. The ratio of these quantities is an indication of how much
-of the measurement error space supports the inferred causal direction. A
-ratio \(R=1\) indicates that the Steiger test is highly sensitive to
-measurement error, with equal weight of the measurement error parameter
-space supporting both directions of causality. Full details are provided
-in Appendix 2.
+\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
+because \(\rho_{g,x_o}\) could be underestimated more than
+\(\rho_{g,y_o}\).
+
+As we show in Appendix 2 it is possible to infer the bounds of
+measurement error on \(x_o\) or \(y_o\) given known genetic
+associations. The maximum measurement error of \(x_o\) is
+\(\rho_{g,x_o}\), because it is known that at least that much of the
+variance has been explained in \(x_o\) by \(g\). The minimum is 0,
+denoting perfectly measured trait values (the same logic applies to
+\(y_o\)). It is possible to simulate what the inferred causal direction
+would be for all values within these bounds.
+
+To evaluate how reliable, \(R\), the inference of the causal direction
+is to potential measurement error in \(x\) and \(y\) we need to predict
+the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
+error. We integrate over the entire range of \(\rho_{gy} - \rho_{gx}\)
+values for possible measurement error values. We find the ratio of the
+volume that agrees with the inferred direction of causality over the
+volume that disagrees with the inferred direction of causality. A ratio
+\(R=1\) indicates that the inferred causal direction is highly sensitive
+to measurement error, with equal weight of the measurement error
+parameter space supporting both directions of causality. In general, the
+\(R\) value denotes that the inferred direction of causality is \(R\)
+times more likely to be the empirical result than the opposite
+direction. Full details are provided in Appendix 2.
 
@@
 Two sample MR\textsuperscript{28} was performed using the summary
-statistics presented in\textsuperscript{40} for each of 4122
-associations between DNA methylation and gene expression levels. To do
-this we obtained a list of 458 gene expression - DNA methylation
-associations as reported in Shakhbazov et al\textsuperscript{40}. These
-were filtered to be located on the same chromosome, have robust
-correlations after correcting for multiple testing, and to share a SNP
-that had a robust cis-acting effect on both the DNA methylation probe
-and the gene expression probe. Because only summary statistics were
-available (effect, standard error, effect allele, sample size, p-values)
-for the instrumental SNP on the methylation and gene expression levels,
-the Steiger test of two independent correlations was used to infer the
-direction of causality for each of the associations. The Wald ratio test
-was then used to estimate the causal effect size for the estimated
-direction for each association.
+statistics for genetic influences on gene expression and DNA
+methylation. To do this we obtained a list of 458 gene expression - DNA
+methylation associations as reported in Shakhbazov et
+al\textsuperscript{40}. These were filtered to be located on the same
+chromosome, have robust correlations after correcting for multiple
+testing, and to share a SNP that had a robust cis-acting effect on both
+the DNA methylation probe and the gene expression probe. Because only
+summary statistics were available (effect, standard error, effect
+allele, sample size, p-values) for the instrumental SNP on the
+methylation and gene expression levels, the Steiger test of two
+independent correlations was used to infer the direction of causality
+for each of the associations. The Wald ratio test was then used to
+estimate the causal effect size for the estimated direction for each
+association.
 
@@
 instrument, but there are three basic competing models for these
-variables. The causal inference test (CIT) attempts to delineate between
-them. a) Methylation causes gene expression. The left figure shows that
-the SNP influences methylation levels that in turn influence gene
-expression levels. The right figure shows the directed acyclic graph
-that represents this model. Faded symbols represent the measured values
-whereas solid symbols represent the true values. b) The same as in A,
-except the causal direction is from gene expression to DNA methylation.
-c) A model of confounding, where gene expression and DNA methylation are
-not causally related, but the SNP influences them each through separate
-pathways or a confounder.
+variables. The causal inference test (CIT) attempts to distinguish
+between them. a) Methylation causes gene expression. The left figure
+shows that the SNP influences methylation levels that in turn influence
+gene expression levels. The right figure shows the directed acyclic
+graph that represents this model. Faded symbols represent the measured
+values whereas solid symbols represent the true values. b) The same as
+in A, except the causal direction is from gene expression to DNA
+methylation. c) A model of confounding, where gene expression and DNA
+methylation are not causally related, but the SNP influences them each
+through separate pathways or a confounder.
 
@@
 axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
-as denoted by range of values where the blue surface is above the red
-plane, those values of measurement error lead to our observed Steiger
-test inferring the wrong causal direction. Where the blue surface lies
-below the red plane, these measurement error values support the inferred
-causal direction of X to Y. A measure of reliability, therefore, is the
-ratio of the negative and positive volumes of the total space bound by
-the blue and red surfaces, \(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\).
-In this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\),
-the \(R = 4.40\), which means that 4.40 times as much of the possible
-measurement error values are in support of the \(x \rightarrow y\)
-direction of causality than \(y \rightarrow x\). b) Plots depicting the
-parameter space in which the function
-\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
-negative the Steiger test is liable to infer the wrong direction of
-causality. Shaded regions show the parameter space where \(d\) is
-negative. The graph shows that for the majority of the parameter space
-of the function, \(d\) is positive, especially where causal
-relationships are relatively weak.
+as denoted by the range of values where the blue surface is above the
+red plane, those values of measurement error lead to our observed
+Steiger test inferring the wrong causal direction. Where the blue
+surface lies below the red plane, these measurement error values support
+the inferred causal direction of X to Y. A measure of reliability,
+therefore, is the ratio of the negative and positive volumes of the
+total space bound by the blue and red surfaces,
+\(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\). In this case, where
+\(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the \(R = 4.40\),
+which means that 4.40 times as much of the possible measurement error
+values are in support of the \(x \rightarrow y\) direction of causality
+than \(y \rightarrow x\). b) Plots depicting the parameter space in
+which the function \(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is
+negative. When \(d\) is negative the Steiger test is liable to infer the
+wrong direction of causality. Shaded regions show the parameter space
+where \(d\) is negative. The graph shows that for the majority of the
+parameter space of the function, \(d\) is positive, especially where
+causal relationships are relatively weak.
 
@@
 
-Figure 5: a) Outcomes were simulated to be causally influenced by
-exposures as shown in the graph, with varying degrees of measurement
-error applied to both. CIT and MR were used to infer evidence for
-causality between the exposure and outcome, and to infer the direction
-of causality. The value of
-\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\) is
-negative we expect the Steiger test to be more likely to be wrong about
-the direction of causality. Rows of graphs represent the sample size
-used in the simulations. For the CIT method, outcome 1 denoted evidence
-for causality with correct model, outcomes 2 or 3 denoted evidence for
-causality with incorrect model, and outcome 4 denoted no evidence for
-causality. b) As in (a) except the simulated model was non-causal, and a
-genetic confounder induces an association between \(x\) and \(y\). MR is
-unable to identify this model, so any significant associations are
-deemed to be incorrect. Outcome 3 denotes evidence for the correct model
-for the CIT method.
+Figure 5: a) Outcome \(y\) was simulated to be caused by exposure \(x\)
+as shown in the graph, with varying degrees of measurement error applied
+to both. CIT and MR were used to infer evidence for causality between
+the exposure and outcome, and to infer the direction of causality. The
+value of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when
+\(d\) is negative we expect the Steiger test to be more likely to be
+wrong about the direction of causality. Rows of graphs represent the
+sample size used in the simulations. For the CIT method, outcome 1
+denoted evidence for causality with correct model, outcomes 2 or 3
+denoted evidence for causality with incorrect model, and outcome 4
+denoted no evidence for causality. b) As in (a) except the simulated
+model was non-causal, and a genetic confounder induces an association
+between \(x\) and \(y\). MR is unable to identify this model, so any
+significant associations are deemed to be incorrect. Outcome 3 denotes
+evidence for the correct model for the CIT method.
 
@@
 
-Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
-\(x_o\) and \(y_o\) will not change the qualitative inference of the
-direction of causality. To evaluate the reliability, \(R\), of the
-Steiger test with regards to measurement error, the objective is to
-compare the weighted proportion of the parameter space that agrees with
-the inferred direction against the weighted proportion which does not:
+Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
+explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
+we infer that \(x \rightarrow y\). There will be some values of
+\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
+\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
+causal direction with regards to measurement error, the objective is to
+compare the proportion of the parameter space that agrees with the
+inferred direction against the proportion which does not:
 
@@
 the range of possible measurement error values. If \(R > 1\) then \(R\)
-times as much of the weighted parameter space favours the inferred
-direction of causality. \(V_{z}\), the total volume of the function
-(Figure 4), can be obtained analytically by solving:
+times as much of the parameter space favours the inferred direction of
+causality. \(V_{z}\), the total volume of the function (Figure 4), can
+be obtained analytically by solving:
 

### 2017-02-25 docs/manuscript.Rmd
updated

@@
 bibliography: manuscript.bib
-csl: nature.csl
+csl: vancouver.csl
 ---
@@
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
-Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
+Measurement error can also be introduced after the data have been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable.
 The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].
@@
 The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship.
-This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
+This provides the theoretical basis behind randomised control trials, but in practice randomisation is often difficult or impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure.
 The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014].
@@
 The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`).
-Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.
+Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must have its most proximal link with the exposure, associating with the outcome only through the exposure.
 
@@
 Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method.
-We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a matric to evaluate the sensitivity of the result of this extension to measurement error.
+We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error.
 Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
-Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
+Finally, we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels.
 Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].
@@
 In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
-To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$ (Model 2).
-For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:
+To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT for each of two tests - $p_{CIT, x \rightarrow y}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$, $p_{CIT, y \rightarrow x}$ (Model 2).
+The results from each of these methods can then be used in combination to infer the existance and direction of causality.
+For some significance threshold $\alpha$ there are four possible outcomes from these two tests, and their interpretations are as follows:
 
@@
 
-and a p-value is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.
-The existence of causality and its direction is inferred based on the following scenarios:
+and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.
+
+The existence of causality and its direction is inferred based on combining information from the MR analysis and the Steiger test.
+The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}).
+For the purposes of compiling simulation results, these can be combined using an arbitrary $\alpha = 0.05$ value:
 
@@
 - Otherwise no evidence for a causal relationship
-
-For the purposes of compiling simulation results, we use an arbitrary $\alpha = 0.05$ value.
 
@@
 
-We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
+We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing methylation levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
 
@@
 We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages.
-First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality.
+First, it uses a formal statistical framework to test for the reliability of the assumed direction of causality.
 Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.
@@
 In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available.
-However, MR strategies exist for other scenarios, for example if there are multiple instruments for the hypothesised exposure then the problems described here may be circumvented [@DaveySmithHemani2014].
-Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014].
+We note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent.
+Multiple instruments can indeed help to circumvent some of the problems described here in other ways also, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
+Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014].
 We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].
@@
 We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies.
-We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
+We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated.
+We have also not addressed other issues pertaining to instrumental variables, such as exposure missspecification, where the instrument is used for a particular variable due to that variable's non-causal correlation with the true exposure.
+However our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
 
@@
 ## References
+
+\raggedright

### 2017-02-25 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{19 February 2017}
+  \date{24 February 2017}
 
@@
 unable to distinguish a true causal association from reverse causation
-or confounding\textsuperscript{1--3}. In response to this, genetic
-instrumentation has emerged as a technique for improving the reliability
-of causal inference in observational data, and with the coincident rise
-in genome-wide association studies it is now a prominent tool that is
-applied in several different guises\textsuperscript{3--6}. However,
-potential pitfalls remain and one that is often neglected is the
-influence of non-differential measurement error on the reliability of
-causal inference.
+or confounding (1--3). In response to this, genetic instrumentation has
+emerged as a technique for improving the reliability of causal inference
+in observational data, and with the coincident rise in genome-wide
+association studies it is now a prominent tool that is applied in
+several different guises (3--6). However, potential pitfalls remain and
+one that is often neglected is the influence of non-differential
+measurement error on the reliability of causal inference.
 
@@
 variability can arise through a whole plethora of mechanisms, which are
-often unique to the study design and difficult to
-avoid\textsuperscript{7,8}. Array technology is now commonly used to
-obtain high throughput phenotyping at low cost, but comes with the
-problem of having imperfect resolution, for instance methylation levels
-as measured by the Illumina450k chip are prone to have some amount of
-noise around the true value due to imperfect
-sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
+often unique to the study design and difficult to avoid (7,8). Array
+technology is now commonly used to obtain high throughput phenotyping at
+low cost, but comes with the problem of having imperfect resolution, for
+instance methylation levels as measured by the Illumina450k chip are
+prone to have some amount of noise around the true value due to
+imperfect sensitivity (9,10). Relatedly, if the measurement of
 biological interest is the methylation level in a T cell, then
@@
 levels from whole blood samples because the measured value will be an
-assay of many cell types\textsuperscript{11}.
+assay of many cell types (11).
 
@@
 a proxy for adiposity, but it is clear that the correlation between BMI
-and underlying adiposity is not perfect\textsuperscript{12}. A similar
-problem of biological misspecification is unavoidable in disease
-diagnosis, and measuring behaviour such as smoking or diet is
-notoriously difficult to do accurately. Measurement error can also be
-introduced after the data has been collected, for example the
-transformation of non-normal data for the purpose of statistical
-analysis will lead to a new variable that will typically incur both
-changes in scale and imprecision (noise) compared to the original
-variable. The sources of measurement error are not limited to this
-list\textsuperscript{8}, and its impact has been explored in the
-epidemiological literature extensively\textsuperscript{13,14}. Given the
+and underlying adiposity is not perfect (12). A similar problem of
+biological misspecification is unavoidable in disease diagnosis, and
+measuring behaviour such as smoking or diet is notoriously difficult to
+do accurately. Measurement error can also be introduced after the data
+have been collected, for example the transformation of non-normal data
+for the purpose of statistical analysis will lead to a new variable that
+will typically incur both changes in scale and imprecision (noise)
+compared to the original variable. The sources of measurement error are
+not limited to this list (8), and its impact has been explored in the
+epidemiological literature extensively (13,14). Given the
 near-ubiquitous presence of measurement error in phenomic data it is
@@
 provides the theoretical basis behind randomised control trials, but in
-practice randomisation is often impossible to implement in an
-experimental context due to cost, scale or inability to manipulate the
-exposure. The principle, however, can be employed in extant
+practice randomisation is often difficult or impossible to implement in
+an experimental context due to cost, scale or inability to manipulate
+the exposure. The principle, however, can be employed in extant
 observational data through the use of genetic variants associated with
 the exposure (instruments), where the inheritance of an allele serves as
-a random lifetime allocation of differential exposure
-levels\textsuperscript{15,16}. Two statistical approaches to exploiting
-the properties of genetic instruments are widely used: mediation-based
-approaches and Mendelian randomisation (MR).
+a random lifetime allocation of differential exposure levels (15,16).
+Two statistical approaches to exploiting the properties of genetic
+instruments are widely used: mediation-based approaches and Mendelian
+randomisation (MR).
 
@@
 information about the causal influence of the exposure on the outcome.
-This forms the basis of a number of methods such as genetical
-genomics\textsuperscript{17}, the regression-based causal inference test
-(CIT)\textsuperscript{4,18}, a structural equation modelling (SEM)
-implementation in the NEO software\textsuperscript{5}, and various other
-methods including Bayesian approaches\textsuperscript{6}. They have been
-employed by a number of recent publications that make causal inferences
-in large scale `omics datasets\textsuperscript{6,19--23}.
+This forms the basis of a number of methods such as genetical genomics
+(17), the regression-based causal inference test (CIT) (4,18), a
+structural equation modelling (SEM) implementation in the NEO software
+(5), and various other methods including Bayesian approaches (6). They
+have been employed by a number of recent publications that make causal
+inferences in large scale `omics datasets (6,19--23).
 
@@
 assumption can be relaxed in various ways when multiple instruments are
-available for a putative exposure\textsuperscript{24,25} and a number of
-sensitivity tests are now available to improve
-reliability\textsuperscript{26}.
+available for a putative exposure (24,25) and a number of sensitivity
+tests are now available to improve reliability (26).
 
@@
 1). Such is not the case for MR, where in order to infer the direction
-of causality between two traits the instrument must be known to
-influence the exposure primarily, associating with the outcome only
+of causality between two traits the instrument must have its most
+proximal link with the exposure, associating with the outcome only
 through the exposure.
@@
 years of follow up studies to understand the biological nature of a
-putative GWAS hit\textsuperscript{27}. If multiple instruments are
-available for an hypothesised exposure, which is increasingly typical
-for complex traits that are analysed in large GWAS consortia, then
-techniques can be applied to mitigate these issues\textsuperscript{16}.
-But these techniques cannot always be applied in the case of determining
-causal directions between 'omic measures where typically only one
-cis-acting SNP is known. For example if a DNA methylation probe is
-associated with expression of an adjacent gene, then is a cis-acting SNP
-an instrument for the DNA methylation level, or the gene expression
-level (Figure 1)?
+putative GWAS hit (27). If multiple instruments are available for an
+hypothesised exposure, which is increasingly typical for complex traits
+that are analysed in large GWAS consortia, then techniques can be
+applied to mitigate these issues (16). But these techniques cannot
+always be applied in the case of determining causal directions between
+'omic measures where typically only one cis-acting SNP is known. For
+example if a DNA methylation probe is associated with expression of an
+adjacent gene, then is a cis-acting SNP an instrument for the DNA
+methylation level, or the gene expression level (Figure 1)?
 
@@
 to be drawn when exposure variables and outcome variables are measured
-in different samples\textsuperscript{28}. This has the crucial advantage
-of improving statistical power by allowing analysis in much larger
-sample sizes, and dramatically expands the breadth of possible
-phenotypic relationships that can be evaluated\textsuperscript{26}.
-Additionally, when MR assumptions are satisfied the method is robust to
-there being measurement error in the exposure
-variable\textsuperscript{29}. Indeed instrumental variable (IV) analysis
-was in part initially introduced as a correction for measurement error
-in the exposure\textsuperscript{30}, whereas it has been noted that both
-classic mediation-based analyses\textsuperscript{13,14,31,32} and
-mediation-based methods that use instrumental
-variables\textsuperscript{33,34} are prone to be unreliable in its
+in different samples (28). This has the crucial advantage of improving
+statistical power by allowing analysis in much larger sample sizes, and
+dramatically expands the breadth of possible phenotypic relationships
+that can be evaluated (26). Additionally, when MR assumptions are
+satisfied the method is robust to there being measurement error in the
+exposure variable (29). Indeed instrumental variable (IV) analysis was
+in part initially introduced as a correction for measurement error in
+the exposure (30), whereas it has been noted that both classic
+mediation-based analyses (13,14,31,32) and mediation-based methods that
+use instrumental variables (33,34) are prone to be unreliable in its
 presence.
@@
 even when the biology of the instruments are not fully understood, and
-also a matric to evaluate the sensitivity of the result of this
+also a metric to evaluate the sensitivity of the result of this
 extension to measurement error. Together these extensions improve the
 utility of MR in cases where mediation based methods might have
-otherwise been used preferentially. Finally, we apply this new method to
+otherwise been used preferentially. Finally, we apply this method to
 infer the direction of causation between DNA methylation levels and gene
@@
 triangulation of evidence from as many sources as possible should be
-practiced in causal inference\textsuperscript{35}.
+practiced in causal inference (35).
 
@@
 the true value that leads to the observed value, \(x_o = f(x)\). For
-example, following Pierce and VanderWeele\textsuperscript{29} we can
-define
+example, following Pierce and VanderWeele (29) we can define
 
@@
 
-The CIT method\textsuperscript{4} is implemented in the R package
-\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
-follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
-the exposure \(x\) causes an outcome \(y\), as described above. The
-following tests are then performed:
+The CIT method (4) is implemented in the R package \emph{R/cit} (18).
+The methodology of the CIT is as follows. Assume an exposure \(x\) is
+instrumented by a SNP \(g\), and the exposure \(x\) causes an outcome
+\(y\), as described above. The following tests are then performed:
 
@@
 omnibus p-value. To infer the direction of causality using the CIT
-method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
-for the direction of \(x\) causing \(y\) (Model 1), and for the
-direction of \(y\) causing \(x\) (Model 2). For some significance
-threshold \(\alpha\), the existence of causality and its direction was
-inferred based on the following scenarios:
+method, an omnibus p-value generated by CIT for each of two tests -
+\(p_{CIT, x \rightarrow y}\), was estimated for the direction of \(x\)
+causing \(y\) (Model 1), and for the direction of \(y\) causing \(x\),
+\(p_{CIT, y \rightarrow x}\) (Model 2). The results from each of these
+methods can then be used in combination to infer the existance and
+direction of causality. For some significance threshold \(\alpha\) there
+are four possible outcomes from these two tests, and their
+interpretations are as follows:
 
@@
 not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance
-thresholds\textsuperscript{36}.
+is it reliable to depend on arbitrary significance thresholds (36).
 
@@
 available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the R package \(R/systemfit\)\textsuperscript{37}.
+obtained using the R package \(R/systemfit\) (37).
 
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population\textsuperscript{38}. It is calculated as
+within a population (38). It is calculated as
 
@@
 
-and a p-value is generated from the \(Z\) value to indicate the
-probability of obtaining the observed difference in correlations
-\(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis that both
-correlations are identical. The existence of causality and its direction
-is inferred based on the following scenarios:
+and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
+indicate the probability of obtaining the observed difference in
+correlations \(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis
+that both correlations are identical.
+
+The existence of causality and its direction is inferred based on
+combining information from the MR analysis and the Steiger test. The MR
+analysis indicates whether there is a potential causal relationship
+(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
+of the causal relationship and the confidence of the direction
+(\$p\_\{Steiger\}). For the purposes of compiling simulation results,
+these can be combined using an arbitrary \(\alpha = 0.05\) value:
 
@@
 
-For the purposes of compiling simulation results, we use an arbitrary
-\(\alpha = 0.05\) value.
-
 Note that the same correlation test approach can be applied to a
-two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
-case where the SNP-exposure association and SNP-outcome association are
-calculated in different samples (e.g.~from publicly available summary
-statistics). Here the Steiger test of two independent correlations can
-be applied where.
+two-sample MR (28) setting. Two-sample MR refers to the case where the
+SNP-exposure association and SNP-outcome association are calculated in
+different samples (e.g.~from publicly available summary statistics).
+Here the Steiger test of two independent correlations can be applied
+where.
 
@@
 are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package
-\emph{R/psych}\textsuperscript{39}.
+\emph{r.test} function in the R package \emph{R/psych} (39).
 
@@
 
-Two sample MR\textsuperscript{28} was performed using the summary
-statistics for genetic influences on gene expression and DNA
-methylation. To do this we obtained a list of 458 gene expression - DNA
-methylation associations as reported in Shakhbazov et
-al\textsuperscript{40}. These were filtered to be located on the same
-chromosome, have robust correlations after correcting for multiple
-testing, and to share a SNP that had a robust cis-acting effect on both
-the DNA methylation probe and the gene expression probe. Because only
-summary statistics were available (effect, standard error, effect
+Two sample MR (28) was performed using the summary statistics for
+genetic influences on gene expression and DNA methylation. To do this we
+obtained a list of 458 gene expression - DNA methylation associations as
+reported in Shakhbazov et al (40). These were filtered to be located on
+the same chromosome, have robust correlations after correcting for
+multiple testing, and to share a SNP that had a robust cis-acting effect
+on both the DNA methylation probe and the gene expression probe. Because
+only summary statistics were available (effect, standard error, effect
 allele, sample size, p-values) for the instrumental SNP on the
@@
 
-All analysis was performed using the R programming
-language\textsuperscript{41} and code is made available at
-\href{}{github url to go here} and implemented in the MR-Base
-(\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.
+All analysis was performed using the R programming language (41) and
+code is made available at \href{}{github url to go here} and implemented
+in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).
 
@@
 
-We performed two sample MR\textsuperscript{28} for each association in
-the direction of causality inferred by the Stieger test. We observed
-that the sign of the MR estimate was generally in the same direction as
-the Pearson correlation coefficient reported by Shakhbazov et
-al\textsuperscript{40} (Figure 6b). There was a moderate correlation
-between the absolute magnitudes of the causal correlation and the
-observational Pearson correlation (r = 0.45). Together these inferences
-suggest that even for relatively low level associations it is important
-to use causal inference methods over observational associations to infer
-causal effect sizes.
+We performed two sample MR (28) for each association in the direction of
+causality inferred by the Stieger test. We observed that the sign of the
+MR estimate was generally in the same direction as the Pearson
+correlation coefficient reported by Shakhbazov et al (40) (Figure 6b).
+There was a moderate correlation between the absolute magnitudes of the
+causal correlation and the observational Pearson correlation (r = 0.45).
+Together these inferences suggest that even for relatively low level
+associations it is important to use causal inference methods over
+observational associations to infer causal effect sizes.
 
@@
 associations where gene expression caused methylation (OR = 0.61 (95\%
-CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
-levels at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation\textsuperscript{42}.
+CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
+at a controlling CpG typically leads to increased gene expression
+levels, consistent with expectation (42).
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored\textsuperscript{43}. Our analysis
-extends this to related methods such as CIT that are widely used in
-predominantly 'omic data. These methods are indeed liable to the same
-issues as standard mediation based analysis, and specifically we show
-that as measurement error in the exposure variable increases, CIT is
-likely to have reduced statistical power, and liable to infer the wrong
-direction of causality. We also demonstrate that, though unintuitive,
-increasing sample size does not resolve the issue, rather it leads to
-more extreme p-values for the model that predicts the wrong direction of
-causality.
+analysis is relatively well explored (43). Our analysis extends this to
+related methods such as CIT that are widely used in predominantly 'omic
+data. These methods are indeed liable to the same issues as standard
+mediation based analysis, and specifically we show that as measurement
+error in the exposure variable increases, CIT is likely to have reduced
+statistical power, and liable to infer the wrong direction of causality.
+We also demonstrate that, though unintuitive, increasing sample size
+does not resolve the issue, rather it leads to more extreme p-values for
+the model that predicts the wrong direction of causality.
 
@@
 several main advantages. First, it uses a formal statistical framework
-to test for the robustness of the assumed direction of causality.
+to test for the reliability of the assumed direction of causality.
 Second, after testing in a comprehensive range of scenarios the MR based
@@
 causal factor in the association. Indeed, previous studies which have
-evaluated measurement error in these platforms do support this
-position\textsuperscript{44,45}, though making strong conclusions for
-this analysis is difficult because measurement error is likely to be
-study specific.
+evaluated measurement error in these platforms do support this position
+(44,45), though making strong conclusions for this analysis is difficult
+because measurement error is likely to be study specific.
 
@@
 in a single sample setting with a view to making a fair comparison
-between MR and the various mediation-based methods available. However,
-MR strategies exist for other scenarios, for example if there are
-multiple instruments for the hypothesised exposure then the problems
-described here may be circumvented\textsuperscript{16}. Likewise, if
-there is at least one instrument for each trait then bi-directional MR
-can offer solutions to inferring the causal
-direction\textsuperscript{46}. We restricted the simulations to
-evaluating the causal inference between quantitative traits, but it is
-possible that the analysis could be extended to binary traits by using
-the genetic variance explained on the liability scale, taking into
-account the population prevalence\textsuperscript{47}.
+between MR and the various mediation-based methods available. We note
+that it is straightforward to extend the MR Steiger approach to multiple
+instruments, requiring only that the total variance explained by all
+instruments be calculated under the assumption that they are
+independent. Multiple instruments can indeed help to circumvent some of
+the problems described here in other ways also, for example by
+evaluating the proportionality of the SNP-exposure and SNP-outcome
+effects (16). Additionally, if there is at least one instrument for each
+trait then bi-directional MR can offer solutions to inferring the causal
+direction (46). We restricted the simulations to evaluating the causal
+inference between quantitative traits, but it is possible that the
+analysis could be extended to binary traits by using the genetic
+variance explained on the liability scale, taking into account the
+population prevalence (47).
 
@@
 on the outcome through a mechanism other than the exposure) was not
-present. Recent method developments in MR\textsuperscript{24,25} have
-focused on accounting for the issues that horizontal pleiotropy can
-introduce when multiple instruments are available, but how they perform
-in the presence of measurement error remains to be explored. An
-important advantage that MR confers over most mediation based analysis
-is that it can be performed in two samples, which can considerably
-improve power and expand the scope of analysis. However, whether there
-is a substantive difference in two sample MR versus one sample MR in how
-measurement error has an effect is not yet fully understood. We have
-also assumed no measurement error in the genetic instrument, which is
-not unreasonable given the strict QC protocols that ensure high quality
-genotype data is available to most studies. We have restricted the scope
-to only exploring non-differential measurement error and avoided the
+present. Recent method developments in MR (24,25) have focused on
+accounting for the issues that horizontal pleiotropy can introduce when
+multiple instruments are available, but how they perform in the presence
+of measurement error remains to be explored. An important advantage that
+MR confers over most mediation based analysis is that it can be
+performed in two samples, which can considerably improve power and
+expand the scope of analysis. However, whether there is a substantive
+difference in two sample MR versus one sample MR in how measurement
+error has an effect is not yet fully understood. We have also assumed no
+measurement error in the genetic instrument, which is not unreasonable
+given the strict QC protocols that ensure high quality genotype data is
+available to most studies. We have restricted the scope to only
+exploring non-differential measurement error and avoided the
 complications incurred if measurement error in the exposure and outcome
-is correlated, however our analysis goes beyond many previous
-explorations of measurement error by assessing the impacts of both
-imprecision (noise) and linear transformations of the true variable on
-causal inference.
+is correlated. We have also not addressed other issues pertaining to
+instrumental variables, such as exposure missspecification, where the
+instrument is used for a particular variable due to that variable's
+non-causal correlation with the true exposure. However our analysis goes
+beyond many previous explorations of measurement error by assessing the
+impacts of both imprecision (noise) and linear transformations of the
+true variable on causal inference.
 
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established\textsuperscript{34} and have a
-number of extensions that make them valuable tools, including for
-example network construction. But because they are predicated on the
-basic underlying principles of mediation they are liable to suffer from
-the same issues of measurement error. Recent advances in MR methodology,
-for example applying MR to genetical genomics\textsuperscript{48},
-multivariate MR\textsuperscript{49} and mediation through
-MR\textsuperscript{50--52} may offer more robust alternatives for these
-more complicated problems.
+variables, are very well established (34) and have a number of
+extensions that make them valuable tools, including for example network
+construction. But because they are predicated on the basic underlying
+principles of mediation they are liable to suffer from the same issues
+of measurement error. Recent advances in MR methodology, for example
+applying MR to genetical genomics (48), multivariate MR (49) and
+mediation through MR (50--52) may offer more robust alternatives for
+these more complicated problems.
 
@@
 robustness under varying levels of measurement error.
+
+\begin{itemize}
+\tightlist
+\item
+  other types of measurement error, e.g.~differential; exposure
+  missspecification;
+\item
+  using multiple instruments
+\end{itemize}
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
+
+\raggedright
 
@@
 \hypertarget{ref-Phillips1991}{}
-1. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
+1. Phillips AN, Davey Smith G. How independent are ``independent''
 effects? relative risk estimation when correlated exposures are measured
-imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
-1223--1231 (1991).
+imprecisely. Journal of Clinical Epidemiology. 1991;44(11):1223--31.
 
 \hypertarget{ref-DaveySmith2002}{}
-2. Davey Smith, G. \& Ebrahim, S. Data dredging, bias, or confounding.
-\emph{BMJ} \textbf{325,} (2002).
+2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
+2002;325(7378):1437--8.
 
 \hypertarget{ref-DaveySmith2004}{}
-3. Davey Smith, G. \& Ebrahim, S. Mendelian randomization: prospects,
-potentials, and limitations. \emph{International journal of
-epidemiology} \textbf{33,} 30--42 (2004).
+3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
+potentials, and limitations. International journal of epidemiology
+{[}Internet{]}. 2004 Feb;33(1):30--42. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/15075143}
 
 \hypertarget{ref-Millstein2009}{}
-4. Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
-molecular relationships with a causal inference test. \emph{BMC
-genetics} \textbf{10,} 23 (2009).
+4. Millstein J, Zhang B, Zhu J, Schadt EE. Disentangling molecular
+relationships with a causal inference test. BMC genetics {[}Internet{]}.
+2009 Jan;10:23. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3224661\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3224661\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Aten2008}{}
-5. Aten, J. E., Fuller, T. F., Lusis, A. J. \& Horvath, S. Using genetic
-markers to orient the edges in quantitative trait networks: the NEO
-software. \emph{BMC systems biology} \textbf{2,} 34 (2008).
+5. Aten JE, Fuller TF, Lusis AJ, Horvath S. Using genetic markers to
+orient the edges in quantitative trait networks: the NEO software. BMC
+systems biology {[}Internet{]}. 2008 Jan;2:34. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2387136\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2387136\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Waszak2015}{}
-6. Waszak, S. M. \emph{et al.} Variation and genetic control of
-chromatin architecture in humans. \emph{Cell} \textbf{162,} 1039--1050
-(2015).
+6. Waszak SM, Delaneau O, Gschwind AR, Kilpinen H, Raghav SK, Witwicki
+RM, et al. Variation and genetic control of chromatin architecture in
+humans. Cell {[}Internet{]}. 2015;162(5):1039--50. Available from:
+\url{http://dx.doi.org/10.1016/j.cell.2015.08.001}
 
 \hypertarget{ref-Houle2011}{}
-7. Houle, D., Pélabon, C., Wagner, G. \& Hansen, T. Measurement and
-meaning in biology. \emph{The Quarterly Review of Biology} \textbf{86,}
-3--34 (2011).
+7. Houle D, Pélabon C, Wagner G, Hansen T. Measurement and meaning in
+biology. The Quarterly Review of Biology {[}Internet{]}.
+2011;86(1):3--34. Available from:
+\url{http://www.jstor.org/stable/10.1086/658408}
 
 \hypertarget{ref-Hernan2009}{}
-8. Hernán, M. a \& Cole, S. R. Invited Commentary: Causal diagrams and
-measurement bias. \emph{American journal of epidemiology} \textbf{170,}
-959--62; discussion 963--4 (2009).
+8. Hernán M a, Cole SR. Invited Commentary: Causal diagrams and
+measurement bias. American journal of epidemiology {[}Internet{]}. 2009
+Oct;170(8):959--62; discussion 963--4. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2765368\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2765368\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Harper2013}{}
-9. Harper, K. N., Peters, B. a \& Gamble, M. V. Batch effects and
-pathway analysis: two potential perils in cancer studies involving DNA
-methylation array analysis. \emph{Cancer epidemiology, biomarkers \&
-prevention : a publication of the American Association for Cancer
-Research, cosponsored by the American Society of Preventive Oncology}
-\textbf{22,} 1052--60 (2013).
+9. Harper KN, Peters B a, Gamble MV. Batch effects and pathway analysis:
+two potential perils in cancer studies involving DNA methylation array
+analysis. Cancer epidemiology, biomarkers \& prevention : a publication
+of the American Association for Cancer Research, cosponsored by the
+American Society of Preventive Oncology {[}Internet{]}. 2013
+Jun;22(6):1052--60. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3687782\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3687782\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Chen2013a}{}
-10. Chen, Y.-a. \emph{et al.} Discovery of cross-reactive probes and
-polymorphic CpGs in the Illumina Infinium HumanMethylation450
-microarray. \emph{Epigenetics : official journal of the DNA Methylation
-Society} \textbf{8,} 203--9 (2013).
+10. Chen Y-a, Lemire M, Choufani S, Butcher DT, Grafodatskaya D, Zanke
+BW, et al. Discovery of cross-reactive probes and polymorphic CpGs in
+the Illumina Infinium HumanMethylation450 microarray. Epigenetics :
+official journal of the DNA Methylation Society {[}Internet{]}. 2013
+Feb;8(2):203--9. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3592906\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3592906\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Houseman2012}{}
-11. Houseman, E. A. \emph{et al.} DNA methylation arrays as surrogate
-measures of cell mixture distribution. \emph{BMC bioinformatics}
-\textbf{13,} 86 (2012).
+11. Houseman EA, Accomando WP, Koestler DC, Christensen BC, Marsit CJ,
+Nelson HH, et al. DNA methylation arrays as surrogate measures of cell
+mixture distribution. BMC bioinformatics {[}Internet{]}. 2012 Jan;13:86.
+Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3532182\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3532182\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Ahima2013}{}
-12. Ahima, R. S. \& Lazar, M. A. Physiology. The health risk of
-obesity--better metrics imperative. \emph{Science (New York, N.Y.)}
-\textbf{341,} 856--8 (2013).
+12. Ahima RS, Lazar MA. Physiology. The health risk of obesity--better
+metrics imperative. Science (New York, NY) {[}Internet{]}. 2013
+Aug;341(6148):856--8. Available from:
+\url{http://science.sciencemag.org/content/341/6148/856.abstract}
 
 \hypertarget{ref-LeCessie2012}{}
-13. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
-Vandenbroucke, J. P. Quantification of bias in direct effects estimates
-due to different types of measurement error in the mediator.
-\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).
+13. Cessie S le, Debeij J, Rosendaal FR, Cannegieter SC, Vandenbroucke
+JP. Quantification of bias in direct effects estimates due to different
+types of measurement error in the mediator. Epidemiology (Cambridge,
+Mass) {[}Internet{]}. 2012 Jul;23(4):551--60. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/22526092}
 
 \hypertarget{ref-Blakely2013}{}
-14. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
-mediator matters when estimating indirect effects. \emph{Journal of
-epidemiology and community health} \textbf{67,} 458--66 (2013).
+14. Blakely T, McKenzie S, Carter K. Misclassification of the mediator
+matters when estimating indirect effects. Journal of epidemiology and
+community health {[}Internet{]}. 2013 May;67(5):458--66. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/23386673}
 
 \hypertarget{ref-DaveySmith2003}{}
-15. Davey Smith, G. \& Ebrahim, S. 'Mendelian randomization': can
-genetic epidemiology contribute to understanding environmental
-determinants of disease? \emph{International Journal of Epidemiology}
-\textbf{32,} 1--22 (2003).
+15. Davey Smith G, Ebrahim S. 'Mendelian randomization': can genetic
+epidemiology contribute to understanding environmental determinants of
+disease? International Journal of Epidemiology {[}Internet{]}. 2003
+Feb;32(1):1--22. Available from:
+\url{http://www.ije.oxfordjournals.org/cgi/doi/10.1093/ije/dyg070}
 
 \hypertarget{ref-DaveySmithHemani2014}{}
-16. Davey Smith, G. \& Hemani, G. Mendelian randomization: genetic
-anchors for causal inference in epidemiological studies. \emph{Human
-molecular genetics} \textbf{23,} R89-----R98 (2014).
+16. Davey Smith G, Hemani G. Mendelian randomization: genetic anchors
+for causal inference in epidemiological studies. Human molecular
+genetics. 2014 Jul;23(R1):R89-----R98.
 
 \hypertarget{ref-Schadt2005}{}
-17. Schadt, E. E. \emph{et al.} An integrative genomics approach to
-infer causal associations between gene expression and disease.
-\emph{Nature Genetics} \textbf{37,} 710--717 (2005).
+17. Schadt EE, Lamb J, Yang X, Zhu J, Edwards S, GuhaThakurta D, et al.
+An integrative genomics approach to infer causal associations between
+gene expression and disease. Nature Genetics {[}Internet{]}. 2005
+Jul;37(7):710--7. Available from:
+\href{http://dx.doi.org/10.1038/ng1589\%20http://www.nature.com/ng/journal/v37/n7/full/ng1589.html\%20http://www.nature.com/ng/journal/v37/n7/pdf/ng1589.pdf}{http://dx.doi.org/10.1038/ng1589 http://www.nature.com/ng/journal/v37/n7/full/ng1589.html http://www.nature.com/ng/journal/v37/n7/pdf/ng1589.pdf}
 
 \hypertarget{ref-Millstein2016}{}
-18. Millstein, J., Chen, G. K. \& Breton, C. V. Cit : Hypothesis Testing
-Software for Mediation Analysis in Genomic Applications.
-\emph{Bioinformatics} \textbf{AOP,} 1--10 (2016).
+18. Millstein J. cit: Causal Inference Test. R package version 1.9
+{[}Internet{]}. 2016. Available from:
+\url{http://cran.r-project.org/package=cit}
 
 \hypertarget{ref-Koestler2014}{}
-19. Koestler, D. C. \emph{et al.} Integrative genomic analysis
-identifies epigenetic marks that mediate genetic risk for epithelial
-ovarian cancer. \emph{BMC medical genomics} \textbf{7,} 8 (2014).
+19. Koestler DC, Chalise P, Cicek MS, Cunningham JM, Armasu S, Larson
+MC, et al. Integrative genomic analysis identifies epigenetic marks that
+mediate genetic risk for epithelial ovarian cancer. BMC medical genomics
+{[}Internet{]}. 2014 Jan;7(1):8. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3916313\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3916313\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Liu2013}{}
-20. Liu, Y. \emph{et al.} Epigenome-wide association data implicate DNA
-methylation as an intermediary of genetic risk in rheumatoid arthritis.
-\emph{Nature biotechnology} \textbf{31,} 142--7 (2013).
+20. Liu Y, Aryee MJ, Padyukov L, Fallin MD, Hesselberg E, Runarsson A,
+et al. Epigenome-wide association data implicate DNA methylation as an
+intermediary of genetic risk in rheumatoid arthritis. Nature
+biotechnology {[}Internet{]}. 2013 Feb;31(2):142--7. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3598632\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3598632\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Yuan2014}{}
-21. Yuan, W. \emph{et al.} An integrated epigenomic analysis for type 2
-diabetes susceptibility loci in monozygotic twins. \emph{Nature
-communications} \textbf{5,} 5719 (2014).
+21. Yuan W, Xia Y, Bell CG, Yet I, Ferreira T, Ward KJ, et al. An
+integrated epigenomic analysis for type 2 diabetes susceptibility loci
+in monozygotic twins. Nature communications {[}Internet{]}. 2014
+Jan;5:5719. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4284644\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4284644\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Tang2014}{}
-22. Tang, Y. \emph{et al.} Genotype-based treatment of type 2 diabetes
-with an \(\alpha\)2A-adrenergic receptor antagonist. \emph{Science
-translational medicine} \textbf{6,} 257ra139 (2014).
+22. Tang Y, Axelsson AS, Spégel P, Andersson LE, Mulder H, Groop LC, et
+al. Genotype-based treatment of type 2 diabetes with an
+\(\alpha\)2A-adrenergic receptor antagonist. Science translational
+medicine {[}Internet{]}. 2014 Oct;6(257):257ra139. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/25298321}
 
 \hypertarget{ref-Hong2015}{}
-23. Hong, X. \emph{et al.} Genome-wide association study identifies
-peanut allergy-specific loci and evidence of epigenetic mediation in US
-children. \emph{Nature communications} \textbf{6,} 6304 (2015).
+23. Hong X, Hao K, Ladd-Acosta C, Hansen KD, Tsai H-J, Liu X, et al.
+Genome-wide association study identifies peanut allergy-specific loci
+and evidence of epigenetic mediation in US children. Nature
+communications {[}Internet{]}. 2015 Jan;6:6304. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4340086\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4340086\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Bowden2015}{}
-24. Bowden, J., Davey Smith, G. \& Burgess, S. Mendelian randomization
-with invalid instruments: effect estimation and bias detection through
-Egger regression. \emph{International Journal of Epidemiology}
-\textbf{In press,} (2015).
+24. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
+invalid instruments: effect estimation and bias detection through Egger
+regression. International Journal of Epidemiology. 2015;44(2):512--25.
 
 \hypertarget{ref-Bowden2016b}{}
-25. Bowden, J., Davey Smith, G., Haycock, P. C. \& Burgess, S.
-Consistent Estimation in Mendelian Randomization with Some Invalid
-Instruments Using a Weighted Median Estimator. \emph{Genetic
-Epidemiology} \textbf{40,} 304--314 (2016).
+25. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent
+Estimation in Mendelian Randomization with Some Invalid Instruments
+Using a Weighted Median Estimator. Genetic Epidemiology {[}Internet{]}.
+2016 May;40(4):304--14. Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27061298\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733\%20http://doi.wiley.com/10.1002/gepi.21965}{http://www.ncbi.nlm.nih.gov/pubmed/27061298 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733 http://doi.wiley.com/10.1002/gepi.21965}
 
 \hypertarget{ref-Hemani2016}{}
-26. Hemani, G. \emph{et al.} MR-Base: a platform for systematic causal
-inference across the phenome using billions of genetic associations.
-\emph{BioRxiv} \textbf{10.1101/07,} (2016).
+26. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+MR-Base: a platform for systematic causal inference across the phenome
+using billions of genetic associations. BioRxiv. 2016;10.1101/07.
 
 \hypertarget{ref-Claussnitzer2015}{}
-27. Claussnitzer, M. \emph{et al.} FTO Obesity Variant Circuitry and
-Adipocyte Browning in Humans. \emph{The New England journal of medicine}
-\textbf{373,} 895--907 (2015).
+27. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
+al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
+New England journal of medicine. 2015;373(10):895--907.
 
 \hypertarget{ref-Pierce2013}{}
-28. Pierce, B. L. \& Burgess, S. Efficient design for Mendelian
-randomization studies: subsample and 2-sample instrumental variable
-estimators. \emph{American journal of epidemiology} \textbf{178,}
-1177--84 (2013).
+28. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+studies: subsample and 2-sample instrumental variable estimators.
+American journal of epidemiology {[}Internet{]}. 2013
+Oct;178(7):1177--84. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Pierce2012}{}
-29. Pierce, B. L. \& VanderWeele, T. J. The effect of non-differential
+29. Pierce BL, VanderWeele TJ. The effect of non-differential
 measurement error on bias, precision and power in Mendelian
-randomization studies. \emph{International Journal of Epidemiology}
-\textbf{41,} 1383--1393 (2012).
+randomization studies. International Journal of Epidemiology
+{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
+\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}
 
 \hypertarget{ref-Ashenfelter1994}{}
-30. Ashenfelter, O. \& Krueger, A. B. Estimates of the Economic Return
-to Schooling from a New Sample of Twins. \emph{The American Economic
-Review} \textbf{84,} 1157--1173 (1994).
+30. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
+Schooling from a New Sample of Twins. The American Economic Review.
+1994;84(5):1157--73.
 
 \hypertarget{ref-Nagarajan2013}{}
-31. Nagarajan, R. \& Scutari, M. Impact of noise on molecular network
-inference. \emph{PloS one} \textbf{8,} e80735 (2013).
+31. Nagarajan R, Scutari M. Impact of noise on molecular network
+inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
+from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Shpitser2010}{}
-32. Shpitser, I., VanderWeele, T. \& Robins, J. On the validity of
-covariate adjustment for estimating causal effects. \emph{Proceedings of
-the Twenty Sixth Conference on Uncertainty in Artificial Intelligence
-(UAI-10)} 527--536 (2010).
+32. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
+adjustment for estimating causal effects. Proceedings of the Twenty
+Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
+2010;527--36.
 
 \hypertarget{ref-Wang2015}{}
-33. Wang, L. \& Michoel, T. Detection of regulator genes and eQTLs in
-gene networks. \emph{arXiv} \textbf{arXiv:1512,} (2015).
+33. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
+networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
+\url{http://arxiv.org/abs/1512.05574}
 
 \hypertarget{ref-Lagani2015}{}
-34. Lagani, V., Triantafillou, S., Ball, G., Tegner, J. \& Tsamardinos,
-I. in \emph{Uncertainty in biology: A computational modeling approach}
-47 (Springer, 2015). at
-\textless{}\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}\textgreater{}
+34. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
+Probabilistic Computational Causal Discovery for Systems Biology. In:
+Uncertainty in biology: A computational modeling approach
+{[}Internet{]}. Springer; 2015. p. 47. Available from:
+\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}
 
 \hypertarget{ref-Lawlor2017}{}
-35. Lawlor, D. A., Tilling, K. \& Davey Smith, G. Triangulation in
-aetiological epidemiology. \emph{International Journal of Epidemiology}
-\textbf{19,} dyw314 (2017).
+35. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
+epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
+Jan;19(R1):dyw314. Available from:
+\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyw314}
 
 \hypertarget{ref-Sterne2001}{}
-36. Sterne, J. A. C., Cox, D. R. \& Smith, G. D. Sifting the
-evidence---what's wrong with significance tests?Another comment on the
-role of statistical methods. \emph{BMJ} \textbf{322,} (2001).
+36. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+significance tests? BMJ. 2001;322(7280):226--31.
 
 \hypertarget{ref-Henningsen2007}{}
-37. Henningsen, A. \& Hamann, J. D. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. \emph{Journal of Statistical
-Software} \textbf{23,} 1--40 (2007).
+37. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. Journal of Statistical Software
+{[}Internet{]}. 2007 Dec;23(4):1--40. Available from:
+\url{https://www.jstatsoft.org/index.php/jss/article/view/v023i04/v23i04.pdf}
 
 \hypertarget{ref-Steiger1980}{}
-38. Steiger, J. H. Tests for comparing elements of a correlation matrix.
-\emph{Psychological Bulletin} \textbf{87,} 245--251 (1980).
+38. Steiger JH. Tests for comparing elements of a correlation matrix.
+Psychological Bulletin. 1980;87(2):245--51.
 
 \hypertarget{ref-Revelle2015}{}
-39. Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
-and Personality Research}. (Northwestern University, 2015). at
-\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}
+39. Revelle W. psych: Procedures for Psychological, Psychometric, and
+Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
+University; 2015. Available from:
+\url{http://cran.r-project.org/package=psych}
 
 \hypertarget{ref-Shakhbazov2016}{}
-40. Shakhbazov, K. \emph{et al.} Shared genetic control of expression
-and methylation in peripheral blood. \emph{BMC genomics} \textbf{17,}
-278 (2016).
+40. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+PM, et al. Shared genetic control of expression and methylation in
+peripheral blood. BMC genomics {[}Internet{]}. 2016 Jan;17(1):278.
+Available from:
+\url{http://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-016-2498-4}
 
 \hypertarget{ref-RCoreTeam2015}{}
-41. R Core Team. \emph{R: A Language and Environment for Statistical
-Computing}. (R Foundation for Statistical Computing, 2015). at
-\textless{}\url{https://www.r-project.org/}\textgreater{}
+41. R Core Team. R: A Language and Environment for Statistical Computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2015. Available from: \url{https://www.r-project.org/}
 
 \hypertarget{ref-Bird2002}{}
-42. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
-\& development} \textbf{16,} 6--21 (2002).
+42. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
+\url{http://genesdev.cshlp.org/content/16/1/6.long}
 
 \hypertarget{ref-Cole2014}{}
-43. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
-Potentially Serious and Misleading Consequences Due to Uncorrected
-Measurement Error. \emph{Psychological Methods} \textbf{19,} 300--315
-(2014).
+43. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+Serious and Misleading Consequences Due to Uncorrected Measurement
+Error. Psychological Methods. 2014;19(2):300--15.
 
 \hypertarget{ref-Bose2014}{}
-44. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
-methylation measurement using technical replicates: the Atherosclerosis
-Risk In Communities (ARIC) Study. \emph{BMC Bioinformatics} \textbf{15,}
-312 (2014).
+44. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+Evaluation of microarray-based DNA methylation measurement using
+technical replicates: the Atherosclerosis Risk In Communities (ARIC)
+Study. BMC Bioinformatics {[}Internet{]}. 2014;15(1):312. Available
+from: \url{http://www.biomedcentral.com/1471-2105/15/312}
 
 \hypertarget{ref-Bryant2011}{}
-45. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
-Biological Variability in a Microarray Experiment but Both Are
-Outweighed by Changes Induced by Stimulation. \emph{PLoS ONE}
-\textbf{6,} e19556 (2011).
+45. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+et al. Technical Variability Is Greater than Biological Variability in a
+Microarray Experiment but Both Are Outweighed by Changes Induced by
+Stimulation. Khodursky AB, editor. PLoS ONE {[}Internet{]}. 2011
+May;6(5):e19556. Available from:
+\url{http://dx.plos.org/10.1371/journal.pone.0019556}
 
 \hypertarget{ref-Richmond2014}{}
-46. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
-between Child Adiposity and Physical Activity Levels: A Mendelian
-Randomization Analysis. \emph{PLoS Medicine} \textbf{11,} e1001618
-(2014).
+46. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
+NJ. Assessing Causality in the Association between Child Adiposity and
+Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
+editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
+from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}
 
 \hypertarget{ref-Lee2013c}{}
-47. Lee, S. H. \& Wray, N. R. Novel genetic analysis for case-control
-genome-wide association studies: quantification of power and genomic
-prediction accuracy. \emph{PLoS One} \textbf{8,} e71494 (2013).
+47. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+association studies: quantification of power and genomic prediction
+accuracy. PLoS One. 2013;8(8):e71494.
 
 \hypertarget{ref-Relton2012}{}
-48. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
+48. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
-processes in pathways to disease. \emph{International journal of
-epidemiology} \textbf{41,} 161--76 (2012).
+processes in pathways to disease. International journal of epidemiology
+{[}Internet{]}. 2012 Feb;41(1):161--76. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Burgess2014a}{}
-49. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
-G. Using multivariable Mendelian randomization to disentangle the causal
-effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).
+49. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+multivariable Mendelian randomization to disentangle the causal effects
+of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
+Available from:
+\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
 
 \hypertarget{ref-Varbo2015}{}
-50. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
+50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
 cholesterol, and blood pressure as mediators from obesity to ischemic
-heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).
+heart disease. Circulation research {[}Internet{]}. 2015
+Feb;116(4):665--73. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/25411050}
 
 \hypertarget{ref-Burgess2015}{}
-51. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
-Network Mendelian randomization: using genetic variants as instrumental
-variables to investigate mediation in causal pathways.
-\emph{International journal of epidemiology} \textbf{44,} 484--95
-(2015).
+51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+randomization: using genetic variants as instrumental variables to
+investigate mediation in causal pathways. International journal of
+epidemiology {[}Internet{]}. 2015 Apr;44(2):484--95. Available from:
+\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
 
 \hypertarget{ref-Richmond2016}{}
-52. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
-C. L. Challenges and novel approaches for investigating molecular
-mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
-(2016).
+52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+Challenges and novel approaches for investigating molecular mediation.
+Human molecular genetics {[}Internet{]}. 2016 Oct;25(R2):R149--56.
+Available from:
+\href{http://www.ncbi.nlm.nih.gov/pubmed/27439390\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}{http://www.ncbi.nlm.nih.gov/pubmed/27439390 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}
 

### 2017-03-03 docs/manuscript.Rmd
updated

@@
 ---
-title: "Inferring the existence and direction of causal associations in the face of measurement error"
+title: "Orienting the causal relationship between imprecisely measured traits using genetic instruments"
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome.
-If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure.
+If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that in this simple three-variable scenario the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure.
 Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome.
@@
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
+It is important to note that if there are any confounders influencing both the exposure and outcome, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome will not work.
+This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abbrogate this association one must also adjust for all (hidden or otherwise) confounders.
+MR does not suffer from this problem because it does not test for association through adjustment.
 Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012].
@@
 where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
+The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]).
+We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders are present due to inducing collider effects.
+
 If all four tests reject the null hypothesis then it is inferred that $x$ causes $y$.
@@
 We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].
+However, our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
 
@@
 We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated.
-We have also not addressed other issues pertaining to instrumental variables, such as exposure missspecification, where the instrument is used for a particular variable due to that variable's non-causal correlation with the true exposure.
-However our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
+We have also not addressed other issues pertaining to instrumental variables which are relevant to the question of instrument-exposure specification.
+One such problem is exposure misspecification, for example an instrument could associate with several closely related putative outcomes, with only one of them actually having a causal effect on the outcome.
+This problem has shown to be the case for SNPs influencing different lipid fractions, for example [@Do2013; @Burgess2014a].
 

### 2017-03-03 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Inferring the existence and direction of causal associations in the face of measurement error},
+            pdftitle={Orienting the causal relationship between imprecisely measured traits using genetic instruments},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Inferring the existence and direction of causal associations in the face
-of measurement error}
+  \title{Orienting the causal relationship between imprecisely measured traits
+using genetic instruments}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \postdate{\par}
-  \date{24 February 2017}
+  \date{03 March 2017}
 
@@
 and the exposure is associated with some outcome, then it logically
-follows that the indirect influence of the SNP on the outcome will
-disappear when conditioning on the exposure. Here, the exposure
-`mediates' the association between the SNP and the outcome, providing
-information about the causal influence of the exposure on the outcome.
-This forms the basis of a number of methods such as genetical genomics
-(17), the regression-based causal inference test (CIT) (4,18), a
-structural equation modelling (SEM) implementation in the NEO software
-(5), and various other methods including Bayesian approaches (6). They
-have been employed by a number of recent publications that make causal
-inferences in large scale `omics datasets (6,19--23).
+follows that in this simple three-variable scenario the indirect
+influence of the SNP on the outcome will disappear when conditioning on
+the exposure. Here, the exposure `mediates' the association between the
+SNP and the outcome, providing information about the causal influence of
+the exposure on the outcome. This forms the basis of a number of methods
+such as genetical genomics (17), the regression-based causal inference
+test (CIT) (4,18), a structural equation modelling (SEM) implementation
+in the NEO software (5), and various other methods including Bayesian
+approaches (6). They have been employed by a number of recent
+publications that make causal inferences in large scale `omics datasets
+(6,19--23).
 
@@
 dramatically expands the breadth of possible phenotypic relationships
-that can be evaluated (26). Additionally, when MR assumptions are
-satisfied the method is robust to there being measurement error in the
-exposure variable (29). Indeed instrumental variable (IV) analysis was
-in part initially introduced as a correction for measurement error in
-the exposure (30), whereas it has been noted that both classic
-mediation-based analyses (13,14,31,32) and mediation-based methods that
-use instrumental variables (33,34) are prone to be unreliable in its
-presence.
+that can be evaluated (26). It is important to note that if there are
+any confounders influencing both the exposure and outcome, the
+mediation-based approach of adjusting the outcome for the exposure to
+nullify the association between the SNP and the outcome will not work.
+This is because adjusting the outcome by the exposure induces a collider
+effect between the SNP and outcome (29), and the in order to fully
+abbrogate this association one must also adjust for all (hidden or
+otherwise) confounders. MR does not suffer from this problem because it
+does not test for association through adjustment. Additionally, when MR
+assumptions are satisfied the method is robust to there being
+measurement error in the exposure variable (30). Indeed instrumental
+variable (IV) analysis was in part initially introduced as a correction
+for measurement error in the exposure (31), whereas it has been noted
+that both classic mediation-based analyses (13,14,32,33) and
+mediation-based methods that use instrumental variables (34,35) are
+prone to be unreliable in its presence.
 
@@
 triangulation of evidence from as many sources as possible should be
-practiced in causal inference (35).
+practiced in causal inference (36).
 
@@
 the true value that leads to the observed value, \(x_o = f(x)\). For
-example, following Pierce and VanderWeele (29) we can define
+example, following Pierce and VanderWeele (30) we can define
 
@@
 residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
-mediate the association between the SNP and the outcome. If all four
-tests reject the null hypothesis then it is inferred that \(x\) causes
-\(y\). The CIT measures the strength of causality by generating an
-omnibus p-value, \(p_{CIT}\), which is simply the largest (least
-extreme) p-value of the four tests, the intuition being that causal
-inference is only as strong as the weakest link in the chain of tests.
+mediate the association between the SNP and the outcome. The 4th
+condition is formulated as an equivalence testing problem that is
+estimated using simulations, comparing the estimate against from the
+data against empirically obtained estimates for simulated variables
+where the independence model is true (full details are given in (4)). We
+note here that this approach is liable to fail, even when there is a
+true causal relationship, when confounders are present due to inducing
+collider effects.
+
+If all four tests reject the null hypothesis then it is inferred that
+\(x\) causes \(y\). The CIT measures the strength of causality by
+generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
+(least extreme) p-value of the four tests, the intuition being that
+causal inference is only as strong as the weakest link in the chain of
+tests.
 
@@
 not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds (36).
+is it reliable to depend on arbitrary significance thresholds (37).
 
@@
 available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the R package \(R/systemfit\) (37).
+obtained using the R package \(R/systemfit\) (38).
 
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population (38). It is calculated as
+within a population (39). It is calculated as
 
@@
 are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} (39).
+\emph{r.test} function in the R package \emph{R/psych} (40).
 
@@
 obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al (40). These were filtered to be located on
+reported in Shakhbazov et al (41). These were filtered to be located on
 the same chromosome, have robust correlations after correcting for
@@
 
-All analysis was performed using the R programming language (41) and
+All analysis was performed using the R programming language (42) and
 code is made available at \href{}{github url to go here} and implemented
@@
 MR estimate was generally in the same direction as the Pearson
-correlation coefficient reported by Shakhbazov et al (40) (Figure 6b).
+correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
 There was a moderate correlation between the absolute magnitudes of the
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation (42).
+levels, consistent with expectation (43).
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored (43). Our analysis extends this to
+analysis is relatively well explored (44). Our analysis extends this to
 related methods such as CIT that are widely used in predominantly 'omic
@@
 evaluated measurement error in these platforms do support this position
-(44,45), though making strong conclusions for this analysis is difficult
+(45,46), though making strong conclusions for this analysis is difficult
 because measurement error is likely to be study specific.
@@
 trait then bi-directional MR can offer solutions to inferring the causal
-direction (46). We restricted the simulations to evaluating the causal
+direction (47). We restricted the simulations to evaluating the causal
 inference between quantitative traits, but it is possible that the
@@
 variance explained on the liability scale, taking into account the
-population prevalence (47).
+population prevalence (48). However, our analysis goes beyond many
+previous explorations of measurement error by assessing the impacts of
+both imprecision (noise) and linear transformations of the true variable
+on causal inference.
 
@@
 is correlated. We have also not addressed other issues pertaining to
-instrumental variables, such as exposure missspecification, where the
-instrument is used for a particular variable due to that variable's
-non-causal correlation with the true exposure. However our analysis goes
-beyond many previous explorations of measurement error by assessing the
-impacts of both imprecision (noise) and linear transformations of the
-true variable on causal inference.
+instrumental variables which are relevant to the question of
+instrument-exposure specification. One such problem is exposure
+misspecification, for example an instrument could associate with several
+closely related putative outcomes, with only one of them actually having
+a causal effect on the outcome. This problem has shown to be the case
+for SNPs influencing different lipid fractions, for example (49,50).
 
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established (34) and have a number of
+variables, are very well established (35) and have a number of
 extensions that make them valuable tools, including for example network
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics (48), multivariate MR (49) and
-mediation through MR (50--52) may offer more robust alternatives for
+applying MR to genetical genomics (51), multivariate MR (50) and
+mediation through MR (52--54) may offer more robust alternatives for
 these more complicated problems.
@@
 robustness under varying levels of measurement error.
-
-\begin{itemize}
-\tightlist
-\item
-  other types of measurement error, e.g.~differential; exposure
-  missspecification;
-\item
-  using multiple instruments
-\end{itemize}
 
@@
 
+\hypertarget{ref-Hernan2004}{}
+29. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
+selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
+Sep;15(5):615--25. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}
+
 \hypertarget{ref-Pierce2012}{}
-29. Pierce BL, VanderWeele TJ. The effect of non-differential
+30. Pierce BL, VanderWeele TJ. The effect of non-differential
 measurement error on bias, precision and power in Mendelian
@@
 \hypertarget{ref-Ashenfelter1994}{}
-30. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
+31. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
 Schooling from a New Sample of Twins. The American Economic Review.
@@
 \hypertarget{ref-Nagarajan2013}{}
-31. Nagarajan R, Scutari M. Impact of noise on molecular network
+32. Nagarajan R, Scutari M. Impact of noise on molecular network
 inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
@@
 \hypertarget{ref-Shpitser2010}{}
-32. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
+33. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
 adjustment for estimating causal effects. Proceedings of the Twenty
@@
 \hypertarget{ref-Wang2015}{}
-33. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
+34. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
 networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
@@
 \hypertarget{ref-Lagani2015}{}
-34. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
+35. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
 Probabilistic Computational Causal Discovery for Systems Biology. In:
@@
 \hypertarget{ref-Lawlor2017}{}
-35. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
+36. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
 epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
@@
 \hypertarget{ref-Sterne2001}{}
-36. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+37. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
 significance tests? BMJ. 2001;322(7280):226--31.
@@
 \hypertarget{ref-Henningsen2007}{}
-37. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+38. Henningsen A, Hamann JD. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. Journal of Statistical Software
@@
 \hypertarget{ref-Steiger1980}{}
-38. Steiger JH. Tests for comparing elements of a correlation matrix.
+39. Steiger JH. Tests for comparing elements of a correlation matrix.
 Psychological Bulletin. 1980;87(2):245--51.
@@
 \hypertarget{ref-Revelle2015}{}
-39. Revelle W. psych: Procedures for Psychological, Psychometric, and
+40. Revelle W. psych: Procedures for Psychological, Psychometric, and
 Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
@@
 \hypertarget{ref-Shakhbazov2016}{}
-40. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+41. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
 PM, et al. Shared genetic control of expression and methylation in
@@
 \hypertarget{ref-RCoreTeam2015}{}
-41. R Core Team. R: A Language and Environment for Statistical Computing
+42. R Core Team. R: A Language and Environment for Statistical Computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-Bird2002}{}
-42. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+43. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
 development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
@@
 \hypertarget{ref-Cole2014}{}
-43. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+44. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
 Serious and Misleading Consequences Due to Uncorrected Measurement
@@
 \hypertarget{ref-Bose2014}{}
-44. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+45. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
 Evaluation of microarray-based DNA methylation measurement using
@@
 \hypertarget{ref-Bryant2011}{}
-45. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+46. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
 et al. Technical Variability Is Greater than Biological Variability in a
@@
 \hypertarget{ref-Richmond2014}{}
-46. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
+47. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
 NJ. Assessing Causality in the Association between Child Adiposity and
@@
 \hypertarget{ref-Lee2013c}{}
-47. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+48. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 
+\hypertarget{ref-Do2013}{}
+49. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+Common variants associated with plasma triglycerides and risk for
+coronary artery disease. Nature Genetics {[}Internet{]}. 2013
+Oct;45(11):1345--52. Available from:
+\url{http://www.nature.com/doifinder/10.1038/ng.2795}
+
+\hypertarget{ref-Burgess2014a}{}
+50. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+multivariable Mendelian randomization to disentangle the causal effects
+of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
+Available from:
+\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
+
 \hypertarget{ref-Relton2012}{}
-48. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+51. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 
-\hypertarget{ref-Burgess2014a}{}
-49. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
-multivariable Mendelian randomization to disentangle the causal effects
-of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
-Available from:
-\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
-
 \hypertarget{ref-Varbo2015}{}
-50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+52. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 \hypertarget{ref-Burgess2015}{}
-51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+53. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 \hypertarget{ref-Richmond2016}{}
-52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+54. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.

### 2017-03-03 scripts/cit_confounder.r

@@ -1,0 +1,93 @@
+library(cit)
+library(tidyverse)
+
+fastAssoc <- function(y, x)
+{
+	index <- is.finite(y) & is.finite(x)
+	n <- sum(index)
+	y <- y[index]
+	x <- x[index]
+	vx <- var(x)
+	bhat <- cov(y, x) / vx
+	ahat <- mean(y) - bhat * mean(x)
+	# fitted <- ahat + x * bhat
+	# residuals <- y - fitted
+	# SSR <- sum((residuals - mean(residuals))^2)
+	# SSF <- sum((fitted - mean(fitted))^2)
+
+	rsq <- (bhat * vx)^2 / (vx * var(y))
+	fval <- rsq * (n-2) / (1-rsq)
+	tval <- sqrt(fval)
+	se <- abs(bhat / tval)
+
+	# Fval <- (SSF) / (SSR/(n-2))
+	# pval <- pf(Fval, 1, n-2, lowe=F)
+	p <- pf(fval, 1, n-2, lowe=F)
+	return(list(
+		ahat=ahat, bhat=bhat, se=se, fval=fval, pval=p
+	))
+}
+
+
+# n <- 10000
+# u <- rnorm(n)
+# g <- rbinom(n, 2, 0.5)
+# x <- g + u + rnorm(n)
+# y <- x + u + rnorm(n)
+
+# i <- 18
+# u <- rnorm(param$n[i])
+# g <- rbinom(param$n[i], 2, 0.5)
+# x <- scale(g) * sqrt(param$vg[i]) + u * sqrt(param$vu[i]) + rnorm(param$n[i], sd = sqrt(1 - param$vg[i] - param$vu[i]))
+# y <- x * sqrt(param$vx[i]) + u * sqrt(param$vu[i]) + rnorm(param$n[i], sd = sqrt(1 - param$vx[i] - param$vu[i]))
+
+# cor(g,x)^2
+# cor(x,y)^2
+# cor(x,u)^2
+# cor(u,y)^2
+
+
+# cit.cp(G = x, L = g, T = y)
+
+
+# n <- 10000
+# u <- rnorm(n)
+# g <- rbinom(n, 2, 0.5)
+# x <- g + u + rnorm(n)
+# y <- x + u*0 + rnorm(n)
+
+
+# cit.cp(G = x, L = g, T = y)
+
+param <- expand.grid(
+	n=c(1000,3000,6000),
[…]

### 2017-03-03 scripts/cit_confounder.r
updated

@@ -79,15 +79,20 @@ for(i in 1:nrow(param))
 
 save(param, file="../results/cit_collider.rdata")
 
+###
+
+library(tidyverse)
+load("../results/cit_collider.rdata")
 
 res <- gather(param, key=test, value=pval, cit, mr)
 res$test <- as.factor(res$test)
 levels(res$test) <- c("CIT", "y ~ g")
 
-ggplot(subset(res, n==6000), aes(x=vu, y=-log10(pval))) +
+ggplot(subset(res), aes(x=vu, y=-log10(pval))) +
 geom_point(aes(colour=test)) +
 geom_smooth(aes(colour=test)) +
 scale_colour_brewer(type="qual", palette=2) +
+facet_grid(n ~ .) +
 labs(x="Confounder effect", colour="Test")
 
-
+ggsave("../images/cit_confounder.pdf")

### 2017-03-05 docs/manuscript.Rmd
updated

@@
 Multiple instruments can indeed help to circumvent some of the problems described here in other ways also, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
-Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014].
+Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017].
 We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].

### 2017-03-05 docs/manuscript.tex
added ref

@@
   \postdate{\par}
-  \date{03 March 2017}
+  \date{05 March 2017}
 
@@
 trait then bi-directional MR can offer solutions to inferring the causal
-direction (47). We restricted the simulations to evaluating the causal
-inference between quantitative traits, but it is possible that the
-analysis could be extended to binary traits by using the genetic
+direction (16,47,48). We restricted the simulations to evaluating the
+causal inference between quantitative traits, but it is possible that
+the analysis could be extended to binary traits by using the genetic
 variance explained on the liability scale, taking into account the
-population prevalence (48). However, our analysis goes beyond many
+population prevalence (49). However, our analysis goes beyond many
 previous explorations of measurement error by assessing the impacts of
@@
 a causal effect on the outcome. This problem has shown to be the case
-for SNPs influencing different lipid fractions, for example (49,50).
+for SNPs influencing different lipid fractions, for example (50,51).
 
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics (51), multivariate MR (50) and
-mediation through MR (52--54) may offer more robust alternatives for
+applying MR to genetical genomics (52), multivariate MR (51) and
+mediation through MR (53--55) may offer more robust alternatives for
 these more complicated problems.
@@
 \hypertarget{ref-Millstein2016}{}
-18. Millstein J. cit: Causal Inference Test. R package version 1.9
-{[}Internet{]}. 2016. Available from:
-\url{http://cran.r-project.org/package=cit}
+18. Millstein J, Chen GK, Breton CV. Cit : Hypothesis Testing Software
+for Mediation Analysis in Genomic Applications. Bioinformatics.
+2016;AOP(10.1093/bioinformatics/btw135):1--10.
 
@@
 
+\hypertarget{ref-Mancuso2017}{}
+48. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+Integrating Gene Expression with Summary Association Statistics to
+Identify Genes Associated with 30 Complex Traits. The American Journal
+of Human Genetics {[}Internet{]}. 2017 Mar;100(3):473--87. Available
+from:
+\url{http://linkinghub.elsevier.com/retrieve/pii/S0002929717300320}
+
 \hypertarget{ref-Lee2013c}{}
-48. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+49. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Do2013}{}
-49. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+50. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
 Common variants associated with plasma triglycerides and risk for
@@
 \hypertarget{ref-Burgess2014a}{}
-50. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+51. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Relton2012}{}
-51. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+52. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Varbo2015}{}
-52. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+53. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 \hypertarget{ref-Burgess2015}{}
-53. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+54. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 \hypertarget{ref-Richmond2016}{}
-54. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+55. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.

### 2017-03-11 docs/manuscript.Rmd
updated

@@
 
-read_chunk("~/repo/cit_measurement_error/scripts/mr_directionality_analysis.R")
+read_chunk("../scripts/mr_directionality_analysis.R")
 opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE)
@@
 Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004].
-In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015].
+In response to this, the use of genetic associations to instrument traits has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015].
 However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
@@
 Measurement error is the difference between the measured value of a quantity and its true value.
-This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement.
-Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009].
+This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or measurement imprecision (noise).
+Such variability can arise through a whole plethora of mechanisms, which are often specific to the study design and difficult to avoid [@Houle2011; @Hernan2009].
 Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a].
@@
 Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome.
-If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that in this simple three-variable scenario the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure.
-Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome.
+If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that in this simple three-variable scenario the estimated direct influence of the SNP on the outcome will be zero when conditioning on the exposure.
+Here, the exposure completely mediates the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome.
 This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], a structural equation modelling (SEM) implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015].
@@
 By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure.
-Such a situation can arise in many scenarios.
 Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015].
@@
 MR has some important advantages over the mediation-based approaches.
-The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
+First, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013].
 This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016].
-It is important to note that if there are any confounders influencing both the exposure and outcome, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome will not work.
-This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abbrogate this association one must also adjust for all (hidden or otherwise) confounders.
+Second, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome is affected by unmeasured confounding of the exposure and outcome.
+This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders.
 MR does not suffer from this problem because it does not test for association through adjustment.
-Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012].
+Third, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012].
 Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.
@@
 
-We can model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that
+We model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that
 
@@
 
-The $\alpha_*$ terms represent intercepts, and the $\epsilon_*$ denote random error.
-Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that
+The $\alpha_*$ terms represent intercepts, and henceforth can be ignored.
+The $\epsilon_*$ terms denote random error, assumed independently and normally distributed with mean zero.
+Mediation-based analyses that test whether $x$ causally relates to $y$ rely on evaluating whether the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that
 
@@
 
-where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
+where $\hat{y} = \hat{\beta}_x x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
 
@@
 
-Measurement error of an exposure can be modeled as a transformation of the true value that leads to the observed value, $x_o = f(x)$.
+Measurement error of an exposure can be modeled as a transformation of the true value ($x$) that leads to the observed value, $x_o = f(x)$.
 For example, following Pierce and VanderWeele [-@Pierce2012] we can define
@@
 The same model of measurement error can be applied to the outcome variable $y$.
-In this study we assume there is no measurement error in the SNP and measurement error in the exposure and the outcome are uncorrelated.
+In this study we assume there is no measurement error in the SNP, and that measurement error in the exposure and the outcome are uncorrelated.
 
@@
 
-The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
+First we describe how the CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
 The methodology of the CIT is as follows.
@@
 The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]).
-We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders are present due to inducing collider effects.
+We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders of the exposure and outcome are present, as these will induce collider bias.
 
@@
 
-In these analyses the *cit.cp* function was used to obtain an omnibus p-value.
+Now we describe how we used the CIT method in our simulations.
+The *cit.cp* function was used to obtain an omnibus p-value.
 To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT for each of two tests - $p_{CIT, x \rightarrow y}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$, $p_{CIT, y \rightarrow x}$ (Model 2).
@@
 Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample.
-A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].
+A p-value for this test, $p_{MR}$, was obtained using the $systemfit$ function in the R package $R/systemfit$ [@Henningsen2007].
+Note that the value of $p_{MR}$ is identical when using the same genetic variant to instrument the influence of the exposure $x$ on the outcome $y$, or erroneously, instrumenting the outcome $y$ on the exposure $x$.
 
@@
 
-and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.
+and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining a difference between correlations $\rho_{gx}$ and $\rho_{gy}$ at least as large as the one observed, under the null hypothesis that both correlations are identical.
 
 The existence of causality and its direction is inferred based on combining information from the MR analysis and the Steiger test.
-The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}).
+The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}$).
 For the purposes of compiling simulation results, these can be combined using an arbitrary $\alpha = 0.05$ value:
 
-- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
-- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
-- Otherwise no evidence for a causal relationship
+- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted, $x \rightarrow y$
+- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted, $y \rightarrow x$
+- Otherwise if $p_{Steiger} > \alpha$ or $p_{MR} > \alpha$, neither model is accepted
 
@@
 The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$.
-However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely.
+However, $\rho_{gx}$ (or $\rho_{gy}$) are underestimated if $x$ (or $y$) are measured imprecisely.
 If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} < \rho_{g,y_o}$ because $\rho_{g,x_o}$ could be underestimated more than $\rho_{g,y_o}$.
@@
 As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
-The maximum measurement error of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$.
+The maximum measurement imprecision of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$.
 The minimum is 0, denoting perfectly measured trait values (the same logic applies to $y_o$).
@@
 
-## Two sample MR
+## Applied example using two sample MR
 
@@
 
-All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
+All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [https://github.com/explodecomputer/causal-directions](https://github.com/explodecomputer/causal-directions) and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
 
@@
 Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
-Therefore, if there is any measurement error that incurs imprecision (i.e.
-$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT.
+Therefore, if there is any measurement error that incurs imprecision in $x$ (i.e.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x_o$, which is in violation of the the 4th condition of the CIT.
 Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
@@
 
-We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+We also performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
 `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
@@
 
-We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect.
+We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
 Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
 
-`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size.
+`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", the MR analysis is indeed liable to infer the wrong direction of causality when $d < 0$, and that this erroneous result is more likely to occur with increasing sample size.
 However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality.
-When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
-The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate erroneously this confounding model as being causal.
+When $d > 0$ we find that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.
 
@@
 
-We used the Steiger test to infer the direction of causality between `r nrow(qtl_orig)` putative associations between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
-The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (95% CI `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`).
+We used the Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations.
+We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
+The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (5%-95% quantiles `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`).
 We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms.
-`r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
+`r cfig("shakhplot")`a shows that the conclusions about the direction of causality between DNA methylation and gene expression are very sensitive to measurement error.
 
@@
 There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`).
-Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.
+Together these inferences suggest that even in estimating associations between 'omic' variables, which are considered to be low level phenotypes, it is important to use causal inference methods over observational associations to infer causal effect sizes.
 
@@
 Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data.
-In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Cole2014].
+In the epidemiological literature issues of measurement error in mediation analysis are relatively well explored [@Cole2014].
 Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data.
-These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
+These methods are indeed susceptible to the same problem as standard mediation based analysis, and specifically we show that as measurement error in the (true) exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
 We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

### 2017-03-11 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{05 March 2017}
+  \date{11 March 2017}
 
@@
 unable to distinguish a true causal association from reverse causation
-or confounding (1--3). In response to this, genetic instrumentation has
-emerged as a technique for improving the reliability of causal inference
-in observational data, and with the coincident rise in genome-wide
-association studies it is now a prominent tool that is applied in
-several different guises (3--6). However, potential pitfalls remain and
-one that is often neglected is the influence of non-differential
-measurement error on the reliability of causal inference.
+or confounding (1--3). In response to this, the use of genetic
+associations to instrument traits has emerged as a technique for
+improving the reliability of causal inference in observational data, and
+with the coincident rise in genome-wide association studies it is now a
+prominent tool that is applied in several different guises (3--6).
+However, potential pitfalls remain and one that is often neglected is
+the influence of non-differential measurement error on the reliability
+of causal inference.
 
@@
 variable have the same error rate, which can manifest as changes in
-scale or levels of imprecision (noise) of the measurement. Such
-variability can arise through a whole plethora of mechanisms, which are
-often unique to the study design and difficult to avoid (7,8). Array
-technology is now commonly used to obtain high throughput phenotyping at
-low cost, but comes with the problem of having imperfect resolution, for
-instance methylation levels as measured by the Illumina450k chip are
-prone to have some amount of noise around the true value due to
-imperfect sensitivity (9,10). Relatedly, if the measurement of
-biological interest is the methylation level in a T cell, then
-measurement error of this value can be introduced by using methylation
-levels from whole blood samples because the measured value will be an
-assay of many cell types (11).
+scale or measurement imprecision (noise). Such variability can arise
+through a whole plethora of mechanisms, which are often specific to the
+study design and difficult to avoid (7,8). Array technology is now
+commonly used to obtain high throughput phenotyping at low cost, but
+comes with the problem of having imperfect resolution, for instance
+methylation levels as measured by the Illumina450k chip are prone to
+have some amount of noise around the true value due to imperfect
+sensitivity (9,10). Relatedly, if the measurement of biological interest
+is the methylation level in a T cell, then measurement error of this
+value can be introduced by using methylation levels from whole blood
+samples because the measured value will be an assay of many cell types
+(11).
 
@@
 and the exposure is associated with some outcome, then it logically
-follows that in this simple three-variable scenario the indirect
-influence of the SNP on the outcome will disappear when conditioning on
-the exposure. Here, the exposure `mediates' the association between the
-SNP and the outcome, providing information about the causal influence of
-the exposure on the outcome. This forms the basis of a number of methods
-such as genetical genomics (17), the regression-based causal inference
-test (CIT) (4,18), a structural equation modelling (SEM) implementation
-in the NEO software (5), and various other methods including Bayesian
-approaches (6). They have been employed by a number of recent
-publications that make causal inferences in large scale `omics datasets
-(6,19--23).
+follows that in this simple three-variable scenario the estimated direct
+influence of the SNP on the outcome will be zero when conditioning on
+the exposure. Here, the exposure completely mediates the association
+between the SNP and the outcome, providing information about the causal
+influence of the exposure on the outcome. This forms the basis of a
+number of methods such as genetical genomics (17), the regression-based
+causal inference test (CIT) (4,18), a structural equation modelling
+(SEM) implementation in the NEO software (5), and various other methods
+including Bayesian approaches (6). They have been employed by a number
+of recent publications that make causal inferences in large scale `omics
+datasets (6,19--23).
 
@@
 instrument for the outcome then they are likely to see an apparently
-robust causal association of outcome on exposure. Such a situation can
-arise in many scenarios. Genome-wide association studies (GWASs) that
-identify genetic associations for complex traits are, by design,
-hypothesis free and agnostic of genomic function, and it often takes
-years of follow up studies to understand the biological nature of a
-putative GWAS hit (27). If multiple instruments are available for an
-hypothesised exposure, which is increasingly typical for complex traits
-that are analysed in large GWAS consortia, then techniques can be
-applied to mitigate these issues (16). But these techniques cannot
-always be applied in the case of determining causal directions between
-'omic measures where typically only one cis-acting SNP is known. For
-example if a DNA methylation probe is associated with expression of an
-adjacent gene, then is a cis-acting SNP an instrument for the DNA
-methylation level, or the gene expression level (Figure 1)?
+robust causal association of outcome on exposure. Genome-wide
+association studies (GWASs) that identify genetic associations for
+complex traits are, by design, hypothesis free and agnostic of genomic
+function, and it often takes years of follow up studies to understand
+the biological nature of a putative GWAS hit (27). If multiple
+instruments are available for an hypothesised exposure, which is
+increasingly typical for complex traits that are analysed in large GWAS
+consortia, then techniques can be applied to mitigate these issues (16).
+But these techniques cannot always be applied in the case of determining
+causal directions between 'omic measures where typically only one
+cis-acting SNP is known. For example if a DNA methylation probe is
+associated with expression of an adjacent gene, then is a cis-acting SNP
+an instrument for the DNA methylation level, or the gene expression
+level (Figure 1)?
 
 MR has some important advantages over the mediation-based approaches.
-The mediation-based approaches require that the exposure, outcome and
-instrumental variables are all measured in the same data, whereas recent
-extensions to MR circumvent this requirement, allowing causal inference
-to be drawn when exposure variables and outcome variables are measured
-in different samples (28). This has the crucial advantage of improving
-statistical power by allowing analysis in much larger sample sizes, and
-dramatically expands the breadth of possible phenotypic relationships
-that can be evaluated (26). It is important to note that if there are
-any confounders influencing both the exposure and outcome, the
-mediation-based approach of adjusting the outcome for the exposure to
-nullify the association between the SNP and the outcome will not work.
-This is because adjusting the outcome by the exposure induces a collider
-effect between the SNP and outcome (29), and the in order to fully
-abbrogate this association one must also adjust for all (hidden or
-otherwise) confounders. MR does not suffer from this problem because it
-does not test for association through adjustment. Additionally, when MR
-assumptions are satisfied the method is robust to there being
-measurement error in the exposure variable (30). Indeed instrumental
-variable (IV) analysis was in part initially introduced as a correction
-for measurement error in the exposure (31), whereas it has been noted
-that both classic mediation-based analyses (13,14,32,33) and
-mediation-based methods that use instrumental variables (34,35) are
-prone to be unreliable in its presence.
+First, the mediation-based approaches require that the exposure, outcome
+and instrumental variables are all measured in the same data, whereas
+recent extensions to MR circumvent this requirement, allowing causal
+inference to be drawn when exposure variables and outcome variables are
+measured in different samples (28). This has the crucial advantage of
+improving statistical power by allowing analysis in much larger sample
+sizes, and dramatically expands the breadth of possible phenotypic
+relationships that can be evaluated (26). Second, the mediation-based
+approach of adjusting the outcome for the exposure to nullify the
+association between the SNP and the outcome is affected by unmeasured
+confounding of the exposure and outcome. This is because adjusting the
+outcome by the exposure induces a collider effect between the SNP and
+outcome (29), and the in order to fully abrogate this association one
+must also adjust for all (hidden or otherwise) confounders. MR does not
+suffer from this problem because it does not test for association
+through adjustment. Third, when MR assumptions are satisfied the method
+is robust to there being measurement error in the exposure variable
+(30). Indeed instrumental variable (IV) analysis was in part initially
+introduced as a correction for measurement error in the exposure (31),
+whereas it has been noted that both classic mediation-based analyses
+(13,14,32,33) and mediation-based methods that use instrumental
+variables (34,35) are prone to be unreliable in its presence.
 
@@
 
-We can model a system whereby some exposure \(x\) has a causal influence
+We model a system whereby some exposure \(x\) has a causal influence
 \(\beta_x\) on an outcome \(y\) such that
@@
 
-The \(\alpha_*\) terms represent intercepts, and the \(\epsilon_*\)
-denote random error. Mediation-based analyses that test whether \(x\)
-causally relates to \(y\) impinge on evaluating if the influence of
-\(g\) on \(y\) can be accounted for by conditioning on \(x\), such that
+The \(\alpha_*\) terms represent intercepts, and henceforth can be
+ignored. The \(\epsilon_*\) terms denote random error, assumed
+independently and normally distributed with mean zero. Mediation-based
+analyses that test whether \(x\) causally relates to \(y\) rely on
+evaluating whether the influence of \(g\) on \(y\) can be accounted for
+by conditioning on \(x\), such that
 
@@
 
-where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). MR analysis
-estimates the causal influence of \(x\) on \(y\) by using the instrument
-as a proxy for \(x\), such that
+where \(\hat{y} = \hat{\beta}_x x\). MR analysis estimates the causal
+influence of \(x\) on \(y\) by using the instrument as a proxy for
+\(x\), such that
 
@@
 Measurement error of an exposure can be modeled as a transformation of
-the true value that leads to the observed value, \(x_o = f(x)\). For
-example, following Pierce and VanderWeele (30) we can define
+the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
+For example, following Pierce and VanderWeele (30) we can define
 
@@
 same model of measurement error can be applied to the outcome variable
-\(y\). In this study we assume there is no measurement error in the SNP
-and measurement error in the exposure and the outcome are uncorrelated.
+\(y\). In this study we assume there is no measurement error in the SNP,
+and that measurement error in the exposure and the outcome are
+uncorrelated.
 
@@
 
-The CIT method (4) is implemented in the R package \emph{R/cit} (18).
-The methodology of the CIT is as follows. Assume an exposure \(x\) is
-instrumented by a SNP \(g\), and the exposure \(x\) causes an outcome
-\(y\), as described above. The following tests are then performed:
+First we describe how the CIT method (4) is implemented in the R package
+\emph{R/cit} (18). The methodology of the CIT is as follows. Assume an
+exposure \(x\) is instrumented by a SNP \(g\), and the exposure \(x\)
+causes an outcome \(y\), as described above. The following tests are
+then performed:
 
@@
 note here that this approach is liable to fail, even when there is a
-true causal relationship, when confounders are present due to inducing
-collider effects.
+true causal relationship, when confounders of the exposure and outcome
+are present, as these will induce collider bias.
 
@@
 
-In these analyses the \emph{cit.cp} function was used to obtain an
-omnibus p-value. To infer the direction of causality using the CIT
-method, an omnibus p-value generated by CIT for each of two tests -
-\(p_{CIT, x \rightarrow y}\), was estimated for the direction of \(x\)
-causing \(y\) (Model 1), and for the direction of \(y\) causing \(x\),
-\(p_{CIT, y \rightarrow x}\) (Model 2). The results from each of these
-methods can then be used in combination to infer the existance and
-direction of causality. For some significance threshold \(\alpha\) there
-are four possible outcomes from these two tests, and their
-interpretations are as follows:
+Now we describe how we used the CIT method in our simulations. The
+\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
+the direction of causality using the CIT method, an omnibus p-value
+generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
+was estimated for the direction of \(x\) causing \(y\) (Model 1), and
+for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
+(Model 2). The results from each of these methods can then be used in
+combination to infer the existance and direction of causality. For some
+significance threshold \(\alpha\) there are four possible outcomes from
+these two tests, and their interpretations are as follows:
 
@@
 available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the R package \(R/systemfit\) (38).
+obtained using the \(systemfit\) function in the R package
+\(R/systemfit\) (38). Note that the value of \(p_{MR}\) is identical
+when using the same genetic variant to instrument the influence of the
+exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
+outcome \(y\) on the exposure \(x\).
 
@@
 and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
-indicate the probability of obtaining the observed difference in
-correlations \(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis
-that both correlations are identical.
+indicate the probability of obtaining a difference between correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
+under the null hypothesis that both correlations are identical.
 
@@
 of the causal relationship and the confidence of the direction
-(\$p\_\{Steiger\}). For the purposes of compiling simulation results,
+(\(p_{Steiger}\)). For the purposes of compiling simulation results,
 these can be combined using an arbitrary \(\alpha = 0.05\) value:
@@
   If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
-  a causal association for the correct model is accepted
+  a causal association for the correct model is accepted,
+  \(x \rightarrow y\)
 \item
   If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
-  a causal association for the incorrect model is accepted
+  a causal association for the incorrect model is accepted,
+  \(y \rightarrow x\)
 \item
-  Otherwise no evidence for a causal relationship
+  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
+  model is accepted
 \end{itemize}
@@
 The Steiger test for inferring if \(x \rightarrow y\) is based on
-evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
-\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
+evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
+\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
 imprecisely. If, for example, \(x\) has lower measurement precision than
@@
 measurement error on \(x_o\) or \(y_o\) given known genetic
-associations. The maximum measurement error of \(x_o\) is
+associations. The maximum measurement imprecision of \(x_o\) is
 \(\rho_{g,x_o}\), because it is known that at least that much of the
@@
 
-\subsection{Two sample MR}\label{two-sample-mr}
+\subsection{Applied example using two sample
+MR}\label{applied-example-using-two-sample-mr}
 
@@
 All analysis was performed using the R programming language (42) and
-code is made available at \href{}{github url to go here} and implemented
-in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).
+code is made available at
+\url{https://github.com/explodecomputer/causal-directions} and
+implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).
 
@@
 when the true model is causal only when \(D = 1\). Therefore, if there
-is any measurement error that incurs imprecision (i.e.
+is any measurement error that incurs imprecision in \(x\) (i.e.
 \(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
-between \(g\) and \(y_o | x\), which is in violation of the the 4th
+between \(g\) and \(y_o | x_o\), which is in violation of the the 4th
 condition of the CIT. Note that scale transformation of \(x\) or \(y\)
@@
 
-We performed simulations to compare the performance of MR against CIT in
-detecting a causal association between simulated variables under
+We also performed simulations to compare the performance of MR against
+CIT in detecting a causal association between simulated variables under
 different levels of imprecision simulated in the exposure. Figure 3
@@
 We used simulations to explore the performance of the MR Steiger
-approach in comparison to CIT in terms of the rate at which a strong
-causal relationship is obtained for the correct direction of causality,
-and the rate at which a strong causal relationship is obtained where the
-reported direction of causality is incorrect. Simulations were performed
-for two models, one for a ``causal model'' where there was a causal
-relationship between \(x\) and \(y\); and one for a ``non-causal model''
-where \(x\) and \(y\) were not causally related, but had a confounded
-association induced by the SNP \(g\) influencing a confounder variable
-\(u\).
-
-Figure 5a shows that, for the ``causal model'', when \(d < 0\) the MR
-analysis is indeed liable to infer the wrong direction of causality, and
-that this erroneous result is more likely to occur with increasing
+approach in comparison to CIT in terms of the rate at which evidence of
+a causal relationship is obtained for the correct direction of
+causality, and the rate at which evidence of a causal relationship is
+obtained where the reported direction of causality is incorrect.
+Simulations were performed for two models, one for a ``causal model''
+where there was a causal relationship between \(x\) and \(y\); and one
+for a ``non-causal model'' where \(x\) and \(y\) were not causally
+related, but had a confounded association induced by the SNP \(g\)
+influencing a confounder variable \(u\).
+
+Figure 5a shows that, for the ``causal model'', the MR analysis is
+indeed liable to infer the wrong direction of causality when \(d < 0\),
+and that this erroneous result is more likely to occur with increasing
 sample size. However, the CIT is in general more fallable to reporting a
 robust causal association for the wrong direction of causality. When
-\(d > 0\) is satisfied we observe that in most cases the MR method has
-greater power to obtain evidence for causality than CIT, and always
-obtains the correct direction of causality. The CIT, unlike the Steiger
-test for MR, is able to distinguish the ``non-causal model'' from the
-``causal model'' (Methods, Figure 5b), but it is evident that
-measurement error will often lead the CIT to evaluate erroneously this
-confounding model as being causal.
+\(d > 0\) we find that in most cases the MR method has greater power to
+obtain evidence for causality than CIT, and always obtains the correct
+direction of causality. The CIT, unlike the Steiger test for MR, is able
+to distinguish the ``non-causal model'' from the ``causal model''
+(Methods, Figure 5b), but it is evident that measurement error will
+often lead the CIT to identify the causal model as true, when in fact
+the underlying model is this non-causal model.
 
@@
 
-We used the Steiger test to infer the direction of causality between 458
-putative associations between DNA methylation levels and gene expression
-levels and found that the causal direction commonly goes in both
+We used the Steiger test to infer the direction of causality between DNA
+methylation and gene expression levels between 458 putative
+associations. We found that the causal direction commonly goes in both
 directions (Figure 6a), but assuming no or equal measurement error, DNA
@@
 (\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
-was 3.92 (95\% CI 1.08 - 37.11). We then went on to predict the causal
-directions of the associations for varying levels of systematic
-measurement error for the different platforms. Figure 6a shows that the
-qualitative result of which of DNA methylation or gene expression is
-more commonly the causal factor is very strongly determined by
-measurement error.
+was 3.92 (5\%-95\% quantiles 1.08 - 37.11). We then went on to predict
+the causal directions of the associations for varying levels of
+systematic measurement error for the different platforms. Figure 6a
+shows that the conclusions about the direction of causality between DNA
+methylation and gene expression are very sensitive to measurement error.
 
@@
 causal correlation and the observational Pearson correlation (r = 0.45).
-Together these inferences suggest that even for relatively low level
-associations it is important to use causal inference methods over
+Together these inferences suggest that even in estimating associations
+between `omic' variables, which are considered to be low level
+phenotypes, it is important to use causal inference methods over
 observational associations to infer causal effect sizes.
@@
 epidemiological literature issues of measurement error in mediation
-analysis is relatively well explored (44). Our analysis extends this to
+analysis are relatively well explored (44). Our analysis extends this to
 related methods such as CIT that are widely used in predominantly 'omic
-data. These methods are indeed liable to the same issues as standard
-mediation based analysis, and specifically we show that as measurement
-error in the exposure variable increases, CIT is likely to have reduced
-statistical power, and liable to infer the wrong direction of causality.
-We also demonstrate that, though unintuitive, increasing sample size
-does not resolve the issue, rather it leads to more extreme p-values for
-the model that predicts the wrong direction of causality.
+data. These methods are indeed susceptible to the same problem as
+standard mediation based analysis, and specifically we show that as
+measurement error in the (true) exposure variable increases, CIT is
+likely to have reduced statistical power, and liable to infer the wrong
+direction of causality. We also demonstrate that, though unintuitive,
+increasing sample size does not resolve the issue, rather it leads to
+more extreme p-values for the model that predicts the wrong direction of
+causality.
 

### 2017-03-13 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{11 March 2017}
+  \date{13 March 2017}
 
@@
 
-where \(\hat{y} = \hat{\beta}_x x\). MR analysis estimates the causal
-influence of \(x\) on \(y\) by using the instrument as a proxy for
-\(x\), such that
+where \(\hat{y} = \hat{\beta}_x x\) and assuming no intercept
+\(y - \hat{y} = \epsilon_x\). MR analysis estimates the causal influence
+of \(x\) on \(y\) by using the instrument as a proxy for \(x\), such
+that
 
@@
 (45,46), though making strong conclusions for this analysis is difficult
-because measurement error is likely to be study specific.
+because measurement error is likely to be study specific. We also
+haven't accounted for the influence of winner's curse, which can inflate
+estimates of the variance explained by SNPs, with higher inflation
+expected amongst lower powered studies. Using p-values for genetic
+associations from replication studies will mitigate this problem.
 

### 2017-03-15 docs/manuscript.Rmd
updated

@@
 
-where $\hat{y} = \hat{\beta}_x x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
+where $\hat{y} = \hat{\beta}_x x$ and assuming no intercept $y - \hat{y} = \epsilon_x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that
 
@@
 We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data.
-The causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
+The inferred causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
 For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association.
 Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.
+We also haven't accounted for the influence of winner's curse, which can inflate estimates of the variance explained by SNPs, with higher inflation expected amongst lower powered studies.
+Using p-values for genetic associations from replication studies will mitigate this problem.
 
 In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available.
-We note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent.
-Multiple instruments can indeed help to circumvent some of the problems described here in other ways also, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
+However, if there is only a single instrument it is difficult to separate between the two competing models of $g$ instrumenting a trait which causes another trait, and $g$ having pleiotropic effects on both traits independently [@Zhu2016].
+Under certain conditions of measurement error the CIT test can distinguish these models.
+We also note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent.
+Multiple instruments can indeed help to distinguish between the causal and pleiotropic models, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
 Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017].
@@
 With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
+
+## Acknowledgements
+
+This work was funded by the UK Medical Research Council (MC_UU_12013/1 and MC_UU_12013/9).
 

### 2017-03-15 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{13 March 2017}
+  \date{15 March 2017}
 
@@
 458 known associations between DNA methylation and gene expression
-levels using summary level data. The causal direction is heavily
-influenced by how much measurement error is present in the different
-assaying platforms. For example, if DNA methylation measures typically
-have higher measuremet error than gene expression measures then our
-analysis suggests that DNA methylation levels would be more often the
-causal factor in the association. Indeed, previous studies which have
-evaluated measurement error in these platforms do support this position
-(45,46), though making strong conclusions for this analysis is difficult
-because measurement error is likely to be study specific. We also
-haven't accounted for the influence of winner's curse, which can inflate
-estimates of the variance explained by SNPs, with higher inflation
-expected amongst lower powered studies. Using p-values for genetic
-associations from replication studies will mitigate this problem.
+levels using summary level data. The inferred causal direction is
+heavily influenced by how much measurement error is present in the
+different assaying platforms. For example, if DNA methylation measures
+typically have higher measuremet error than gene expression measures
+then our analysis suggests that DNA methylation levels would be more
+often the causal factor in the association. Indeed, previous studies
+which have evaluated measurement error in these platforms do support
+this position (45,46), though making strong conclusions for this
+analysis is difficult because measurement error is likely to be study
+specific. We also haven't accounted for the influence of winner's curse,
+which can inflate estimates of the variance explained by SNPs, with
+higher inflation expected amongst lower powered studies. Using p-values
+for genetic associations from replication studies will mitigate this
+problem.
 
@@
 in a single sample setting with a view to making a fair comparison
-between MR and the various mediation-based methods available. We note
-that it is straightforward to extend the MR Steiger approach to multiple
+between MR and the various mediation-based methods available. However,
+if there is only a single instrument it is difficult to separate between
+the two competing models of \(g\) instrumenting a trait which causes
+another trait, and \(g\) having pleiotropic effects on both traits
+independently (47). Under certain conditions of measurement error the
+CIT test can distinguish these models. We also note that it is
+straightforward to extend the MR Steiger approach to multiple
 instruments, requiring only that the total variance explained by all
 instruments be calculated under the assumption that they are
-independent. Multiple instruments can indeed help to circumvent some of
-the problems described here in other ways also, for example by
-evaluating the proportionality of the SNP-exposure and SNP-outcome
-effects (16). Additionally, if there is at least one instrument for each
-trait then bi-directional MR can offer solutions to inferring the causal
-direction (16,47,48). We restricted the simulations to evaluating the
-causal inference between quantitative traits, but it is possible that
-the analysis could be extended to binary traits by using the genetic
+independent. Multiple instruments can indeed help to distinguish between
+the causal and pleiotropic models, for example by evaluating the
+proportionality of the SNP-exposure and SNP-outcome effects (16).
+Additionally, if there is at least one instrument for each trait then
+bi-directional MR can offer solutions to inferring the causal direction
+(16,48,49). We restricted the simulations to evaluating the causal
+inference between quantitative traits, but it is possible that the
+analysis could be extended to binary traits by using the genetic
 variance explained on the liability scale, taking into account the
-population prevalence (49). However, our analysis goes beyond many
+population prevalence (50). However, our analysis goes beyond many
 previous explorations of measurement error by assessing the impacts of
@@
 a causal effect on the outcome. This problem has shown to be the case
-for SNPs influencing different lipid fractions, for example (50,51).
+for SNPs influencing different lipid fractions, for example (51,52).
 
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics (52), multivariate MR (51) and
-mediation through MR (53--55) may offer more robust alternatives for
+applying MR to genetical genomics (53), multivariate MR (52) and
+mediation through MR (54--56) may offer more robust alternatives for
 these more complicated problems.
@@
 robustness under varying levels of measurement error.
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+This work was funded by the UK Medical Research Council (MC\_UU\_12013/1
+and MC\_UU\_12013/9).
 
@@
 \hypertarget{ref-Millstein2016}{}
-18. Millstein J, Chen GK, Breton CV. Cit : Hypothesis Testing Software
-for Mediation Analysis in Genomic Applications. Bioinformatics.
-2016;AOP(10.1093/bioinformatics/btw135):1--10.
+18. Millstein J. cit: Causal Inference Test. R package version 1.9
+{[}Internet{]}. 2016. Available from:
+\url{http://cran.r-project.org/package=cit}
 
@@
 
+\hypertarget{ref-Zhu2016}{}
+47. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+Integration of summary data from GWAS and eQTL studies predicts complex
+trait gene targets. Nature Genetics {[}Internet{]}. 2016
+Mar;48(5):481--7. Available from:
+\url{http://www.nature.com/doifinder/10.1038/ng.3538}
+
 \hypertarget{ref-Richmond2014}{}
-47. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
+48. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
 NJ. Assessing Causality in the Association between Child Adiposity and
@@
 \hypertarget{ref-Mancuso2017}{}
-48. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+49. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
 Integrating Gene Expression with Summary Association Statistics to
@@
 \hypertarget{ref-Lee2013c}{}
-49. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+50. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Do2013}{}
-50. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+51. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
 Common variants associated with plasma triglycerides and risk for
@@
 \hypertarget{ref-Burgess2014a}{}
-51. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+52. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Relton2012}{}
-52. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+53. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Varbo2015}{}
-53. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+54. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 \hypertarget{ref-Burgess2015}{}
-54. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+55. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 \hypertarget{ref-Richmond2016}{}
-55. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+56. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.

### 2017-03-21 docs/manuscript.Rmd
updated for submission

@@
 We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.
+
+### Author summary
+
+Understanding the causal relationships between pairs of traits is crucial for unravelling the causes of disease.
+To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases.
+Here we demonstrate that when there is uncertainty about the nature of the genetic association several existing methods that are used for causal inference can be liable to give the wrong direction of causality.
+We introduce a new method that is more flexible and robust to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression.
+Our results show that DNA methylation is more likely to be the causal factor, but that systematic differences in measurement error between the platforms could influence such conclusions.
 

### 2017-03-21 docs/manuscript.tex
updated for submission

@@
   \postdate{\par}
-  \date{15 March 2017}
+  \date{20 March 2017}
 
@@
 conclusions about causality.
+
+\subsubsection{Author summary}\label{author-summary}
+
+Understanding the causal relationships between pairs of traits is
+crucial for unravelling the causes of disease. To this end, results from
+genome-wide association studies are valuable because if a trait is known
+to be influenced by a genetic variant then this knowledge can be used to
+test the trait's causal influences on other traits and diseases. Here we
+demonstrate that when there is uncertainty about the nature of the
+genetic association several existing methods that are used for causal
+inference can be liable to give the wrong direction of causality. We
+introduce a new method that is more flexible and robust to problems such
+as measurement error, and apply it to evaluate the causal relationships
+between DNA methylation levels and gene expression. Our results show
+that DNA methylation is more likely to be the causal factor, but that
+systematic differences in measurement error between the platforms could
+influence such conclusions.
 

### 2017-07-14 docs/manuscript.Rmd
updated

@@
 
-In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present.
+Our new method attempts to infer causal directions under the assumption that horizontal pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) is not present.
 Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored.
@@
 
+## Appendix 3
+
+We have assumed no unmeasured confounding in these simulations.
+Unmeasured confounding will however have potentially large influences on mediation-based methods for inferring causal directions, and can also adversely influence the estimate of the causal direction for the Steiger test.
+
+### Unmeasured confounding in mediation
+
+Including an unmeasured confounder, $u$, after ignoring intercept terms the exposure $x$ and outcome $y$ variables can be modelled as
+
+$$
+\begin{aligned}
+
+y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
+x & = \beta_g g + \beta_{ux} u + \epsilon_g
+\end{aligned}
+$$
+
+The observational estimate of the causal effect of $x$ on $y$, $\hat{\beta}_x$ is obtained from
+
+$$
+\begin{aligned}
+\hat{\beta}_x & = cov(x, y) / var(x) \\
+& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x x + \beta_{uy} u + \epsilon_x) / var(x) \\
+& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x x + \beta_{uy} u + \epsilon_x) / var(x) \\
+& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x \beta_g g + \beta_x \beta_{ux} u + \beta_x \epsilon_g + \beta_{uy} u + \epsilon_x) / var(x) \\
+& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
+\end{aligned}
+$$
+
+From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are both non-zero.
+In terms of mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
+
+$$
+\hat{y}* = \hat{\beta}_x x
+$$
+
+and
+
+$$
+\begin{aligned}
+cov(g, y - \hat{y}*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
+& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
+& = (\beta_x - \hat{\beta}_x) var(g)
+\end{aligned}
+$$
+
+Should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and the $y|x$, which will introduce errors in inferring causal directions.
+
+### Unmeasured confounding in the Steiger test
+
+Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
+
+$$
+cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
+$$
+
+$$
+\begin{aligned}
+cor(g,y)^2 & = cov()
+\end{aligned}
+$$
+
+\end{aligned}
+$$
+
 ## References

### 2017-07-14 scripts/unmeasured_confounding.R

@@ -1,0 +1,202 @@
+n <- 100000
+u <- rnorm(n)
+g <- rnorm(n)
+bxy <- 1
+bux <- 1
+buy <- -5
+bgx <- 05
+x <- bgx * g + bux * u + rnorm(n)
+y <- bxy * x + buy * u + rnorm(n)
+
+
+cor(g,x)^2
+bgx^2 / var(x)
+
+
+cor(u,x)^2
+bux^2/var(x)
+
+# cor(x,y)^2
+
+cor(g,y)^2
+
+
+cor(g,x)^2 * cor(y,x)^2
+cor(g,x)^2 * bxy^2 / var(y)
+
+cov(g, y)
+
+cov(g, bxy * (bgx * g + bux * u) + buy * u)
+cov(g, bxy * bgx * g) / var(g)
+cov(g,y) / var(g)
+
+(cov(x,y)/var(x))^2 * bgx^2 * var(g) + (cov(x,y)/var(x))^2 * bux^2 * var(u) + buy^2 * var(u) + 1
+var(y)
+
+(bxy^2 * bgx^2 * var(g)^2) / ((cov(x,y)/var(x))^2 * bgx^2 * var(g) + (cov(x,y)/var(x))^2 * bux^2 * var(u) + buy^2 * var(u) + 1)
+
+cor(g,y)^2
+
+
+get_bxy <- function(bxy, bgx, bux, buy, vg, vu, vex, vey)
+
+get_calcs <- function(bxy, bgx, bux, buy, vg, vu, vex, vey)
+{
+	bxyo <- ((bgx^2*bxy*vg + bux^2*bxy*vu + bux*buy*vu + bxy) / (vg*bgx^2 + vu*bux^2 + vex))
+
+	rsqgx <- bgx^2 / (vg * bgx^2 + vu * bux^2 + vex)
+	rsqgy <- (bxy^2 * bgx^2 * vg^2) / (bxyo^2 * bgx^2 * vg + bxyo^2 * bux^2 * vu + buy^2 * vu + vey)
+
+	rsqux <- bux^2 / (vg*bgx^2 + vu*bux^2 + vex)
+	rsquy <- buy^2 / (bxyo^2 * bgx^2 * vg + bxyo^2 * bux^2 * vu + buy^2 * vu + vey)
+	return(data.frame(bxyo=bxyo, rsqgx=rsqgx, rsqgy=rsqgy, rsqux=rsqux, rsquy=rsquy))
+}
+
+
+get_calcs()
+
+
+
[…]

### 2017-07-16 docs/manuscript.Rmd
updated

@@
 ---
-title: "Orienting the causal relationship between imprecisely measured traits using genetic instruments"
+title: "Orienting the causal relationship between imprecisely measured traits using GWAS summary data"
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 
+sfig_nums(
+	"unmeasured",
+	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the Steiger test returning the wrong causal direction.
+Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$)."),
+	display=FALSE
+)
+
 ```
@@
 Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages.
-First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated.
+First, it can be performed using only summary level data from genome-wide association studies; second, it is less susceptible to bias in the presence of measurement error or unmeasured confounding.
 We apply the method to infer the causal direction between DNA methylation and gene expression levels.
-Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms.
+Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms, and by horizontal pleiotropy.
 We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.
@@
 
-Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting.
+Note that the same correlation test approach can be applied to a two-sample MR setting.
 Two-sample MR refers to the case where the SNP-exposure association and SNP-outcome association are calculated in different samples (e.g.
-from publicly available summary statistics).
+from publicly available summary statistics [@Pierce2013; @Hemani2016]).
 Here the Steiger test of two independent correlations can be applied where.
@@
 
-### Causal direction sensitivity analysis
+The Steiger test assumes that there is a causal relationship between the two variables, and that the SNP is a valid instrument for one of them.
+However it is liable to give incorrect causal directions under some other circumstances also.
+First, some levels of horizontal pleiotropy, where the SNP influences the outcome through some pathway other than the exposure, could induce problems because this is a means by which the instrument is invalid.
+Second, some differential values of measurement error between the exposure and the outcome could induce incorrect causal directions, and this is explored in Appendix 2.
+Third, some levels of unmeasured confounding between the exposure and the outcome could induce the wrong causal direction, and this is explored in Appendix 3.
+
+### Causal direction sensitivity analysis for measurement error
 
@@
 To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement error in $x$ and $y$ we need to predict the values of $\rho_{gy} - \rho_{gx}$ for those values of measurement error.
-We integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values.
+We offer two tools in which to do this.
+First, the user can provide values of measurement error for $x$ and $y$ and obtain a revised inference of the causal direction.
+Second, we integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values, assuming that any measurement error value is equally likely.
 We find the ratio of the volume that agrees with the inferred direction of causality over the volume that disagrees with the inferred direction of causality.
@@
 \begin{aligned}
-
 y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
@@
 \hat{\beta}_x & = cov(x, y) / var(x) \\
-& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x x + \beta_{uy} u + \epsilon_x) / var(x) \\
-& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x x + \beta_{uy} u + \epsilon_x) / var(x) \\
-& = cov(\beta_g g + \beta_{ux} u + \epsilon_g, \beta_x \beta_g g + \beta_x \beta_{ux} u + \beta_x \epsilon_g + \beta_{uy} u + \epsilon_x) / var(x) \\
 & = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
@@
 
-From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are both non-zero.
-In terms of mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
+From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are non-zero.
+Relating to mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
 
@@
 
-Should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and the $y|x$, which will introduce errors in inferring causal directions.
+should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and $y|x$, which will introduce errors in inferring causal directions.
 
@@
 Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
+After some algebra
 
@@
 
+and
+
 $$
-\begin{aligned}
-cor(g,y)^2 & = cov()
-\end{aligned}
+cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
 $$
 
-\end{aligned}
-$$
+`r csfig("unmeasured")` shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$.
+We note that in most cases where the observational absolute $R^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the Steiger test.
+
+![](../images/unmeasured_conf_steiger.png)
+
+`r sfig_nums("unmeasured")`
+
+\newpage
 

### 2017-07-16 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{20 March 2017}
+  \date{16 July 2017}
 
@@
 
-In this work we assumed that pleiotropy (the influence of the instrument
-on the outcome through a mechanism other than the exposure) was not
-present. Recent method developments in MR (24,25) have focused on
-accounting for the issues that horizontal pleiotropy can introduce when
-multiple instruments are available, but how they perform in the presence
-of measurement error remains to be explored. An important advantage that
-MR confers over most mediation based analysis is that it can be
-performed in two samples, which can considerably improve power and
-expand the scope of analysis. However, whether there is a substantive
-difference in two sample MR versus one sample MR in how measurement
-error has an effect is not yet fully understood. We have also assumed no
-measurement error in the genetic instrument, which is not unreasonable
-given the strict QC protocols that ensure high quality genotype data is
-available to most studies. We have restricted the scope to only
-exploring non-differential measurement error and avoided the
-complications incurred if measurement error in the exposure and outcome
-is correlated. We have also not addressed other issues pertaining to
-instrumental variables which are relevant to the question of
-instrument-exposure specification. One such problem is exposure
-misspecification, for example an instrument could associate with several
-closely related putative outcomes, with only one of them actually having
-a causal effect on the outcome. This problem has shown to be the case
-for SNPs influencing different lipid fractions, for example (51,52).
+Our new method attempts to infer causal directions under the assumption
+that horizontal pleiotropy (the influence of the instrument on the
+outcome through a mechanism other than the exposure) is not present.
+Recent method developments in MR (24,25) have focused on accounting for
+the issues that horizontal pleiotropy can introduce when multiple
+instruments are available, but how they perform in the presence of
+measurement error remains to be explored. An important advantage that MR
+confers over most mediation based analysis is that it can be performed
+in two samples, which can considerably improve power and expand the
+scope of analysis. However, whether there is a substantive difference in
+two sample MR versus one sample MR in how measurement error has an
+effect is not yet fully understood. We have also assumed no measurement
+error in the genetic instrument, which is not unreasonable given the
+strict QC protocols that ensure high quality genotype data is available
+to most studies. We have restricted the scope to only exploring
+non-differential measurement error and avoided the complications
+incurred if measurement error in the exposure and outcome is correlated.
+We have also not addressed other issues pertaining to instrumental
+variables which are relevant to the question of instrument-exposure
+specification. One such problem is exposure misspecification, for
+example an instrument could associate with several closely related
+putative outcomes, with only one of them actually having a causal effect
+on the outcome. This problem has shown to be the case for SNPs
+influencing different lipid fractions, for example (51,52).
 
@@
 
+\subsection{Appendix 3}\label{appendix-3}
+
+We have assumed no unmeasured confounding in these simulations.
+Unmeasured confounding will however have potentially large influences on
+mediation-based methods for inferring causal directions, and can also
+adversely influence the estimate of the causal direction for the Steiger
+test.
+
+\subsubsection{Unmeasured confounding in
+mediation}\label{unmeasured-confounding-in-mediation}
+
+Including an unmeasured confounder, \(u\), after ignoring intercept
+terms the exposure \(x\) and outcome \(y\) variables can be modelled as
+
+\[
+\begin{aligned}
+y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
+x & = \beta_g g + \beta_{ux} u + \epsilon_g
+\end{aligned}
+\]
+
+The observational estimate of the causal effect of \(x\) on \(y\),
+\(\hat{\beta}_x\) is obtained from
+
+\[
+\begin{aligned}
+\hat{\beta}_x & = cov(x, y) / var(x) \\
+& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
+\end{aligned}
+\]
+
+From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
+when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
+mediation, where we attempt to test if \(g\) associates with \(y\) after
+adjusting \(y\) for \(x\), such that
+
+\[
+\hat{y}* = \hat{\beta}_x x
+\]
+
+and
+
+\[
+\begin{aligned}
+cov(g, y - \hat{y}*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
+& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
+& = (\beta_x - \hat{\beta}_x) var(g)
+\end{aligned}
+\]
+
+should any amount of unmeasured confounding exist, therefore, there will
+remain an association between \(g\) and \(y|x\), which will introduce
+errors in inferring causal directions.
+
+\subsubsection{Unmeasured confounding in the Steiger
+test}\label{unmeasured-confounding-in-the-steiger-test}
+
+Similarly, we can investigate the extent to which unmeasured confounding
+will lead to the wrong causal direction between \(x\) and \(y\) using
+the Steiger test, evaluating the liability for the inequality
+\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
+
+\[
+cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
+\]
+
+and
+
+\[
+cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
+\]
+
+Supplementary Figure 1 shows the relationship between the magnitude of
+the correlations between \(x\), \(y\) and the confounder \(u\). We note
+that in most cases where the observational absolute \(R^2_{xy}\) is less
+than 0.2, unmeasured confounding will not incur the wrong causal
+direction in the Steiger test.
+
+\includegraphics{../images/unmeasured_conf_steiger.png}
+
+Supplementary Figure 1: Graph representing the unmeasured confounding
+parameters that will lead to the Steiger test returning the wrong causal
+direction. Columns of boxes represent different signed values of the
+observational variance explained between \(x\) and \(y\) (\(R^2_{xy}\)).
+
+\newpage
+
 \subsection{References}\label{references}

### 2017-07-16 scripts/mr_directionality_analysis.R

@@ -151,10 +151,10 @@ mr_wald_ratio <- function(b_exp, b_out, se_exp, se_out, parameters)
 
 ## ---- read_data ----
 
-load("~/repo/cit_measurement_error/results/mr_directionality_20170126.RData")
+load("~/repo/causal-directions/results/mr_directionality_20170126.RData")
 parametersc <- parameters
 parametersc$model <- "Non-causal"
-load("~/repo/cit_measurement_error/results/mr_directionality_20160211.RData")
+load("~/repo/causal-directions/results/mr_directionality_20160211.RData")
 parameters$model <- "Causal"
 parameters$cit_AB1 <- parameters$cit_AB2 <- parameters$cit_AB3 <- parameters$cit_AB4 <- NA
 parameters$cit_BA1 <- parameters$cit_BA2 <- parameters$cit_BA3 <- parameters$cit_BA4 <- NA
@@ -273,7 +273,7 @@ psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=seq(-0.2, 1.1, by=0.1))
 ## ---- cit_measurement_error_figure ----
 
 
-load("~/repo/cit_measurement_error/results/20141114.RData")
+load("~/repo/causal-directions/results/20141114.RData")
 dat <- gather(dat, eval, pval, GT, TG, factor_key=TRUE)
 levels(dat$eval) <- c("Correct causal model", "Incorrect causal model")
 
@@ -623,13 +623,13 @@ pl$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp2[,1]))
 pl$rhs_lhs_diff_bin_lab <- as.factor(paste0("d = ", pl$rhs_lhs_diff_bin_numeric))
 pl$rhs_lhs_diff_bin_lab <- factor(pl$rhs_lhs_diff_bin_lab, levels=levels(pl$rhs_lhs_diff_bin_lab)[order(as.numeric(as.character(levels(pl$rhs_lhs_diff_bin_numeric))))])
 
-temp <- dplyr::group_by(pl, test, r_ab, r_za, n, noisea, noiseb) %>%
+temp <- dplyr::group_by(pl, test, r_ab, r_za, n, noiseb) %>%
 	dplyr::summarise(prop_causality_exists = sum(causality_exists) / n())
[…]

### 2017-07-16 scripts/unmeasured_confounding.R

@@ -1,12 +1,16 @@
 n <- 100000
-u <- rnorm(n)
-g <- rnorm(n)
+vu <- 1
+vg <- 1
 bxy <- 1
 bux <- 1
 buy <- -5
 bgx <- 05
-x <- bgx * g + bux * u + rnorm(n)
-y <- bxy * x + buy * u + rnorm(n)
+vey <- 1
+vex <- 1
+u <- rnorm(n, sd=sqrt(vu))
+g <- rnorm(n, sd=sqrt(vg))
+x <- bgx * g + bux * u + rnorm(n, sqrt(vex))
+y <- bxy * x + buy * u + rnorm(n, sqrt(vey))
 
 
 cor(g,x)^2
@@ -36,6 +40,17 @@ var(y)
 (bxy^2 * bgx^2 * var(g)^2) / ((cov(x,y)/var(x))^2 * bgx^2 * var(g) + (cov(x,y)/var(x))^2 * bux^2 * var(u) + buy^2 * var(u) + 1)
 
 cor(g,y)^2
+cor(g,x)^2
+
+cor(x,y)^2
+
+lm(y ~ x)
+
+var(x)
+var(y)
+
+lm(y ~ x)$coefficients[2]^2 * var(x)/var(y)
+
 
 
 get_bxy <- function(bxy, bgx, bux, buy, vg, vu, vex, vey)
@@ -44,47 +59,49 @@ get_calcs <- function(bxy, bgx, bux, buy, vg, vu, vex, vey)
 {
 	bxyo <- ((bgx^2*bxy*vg + bux^2*bxy*vu + bux*buy*vu + bxy) / (vg*bgx^2 + vu*bux^2 + vex))
 
+	vx <- bgx^2 * vg + bux^2 * vu + vex
+	vy <- bxyo^2 * vx + buy^2 * vu + vey
+
+	rsqxyo <- bxyo^2 * vx / vy
+	rsqxyos <- rsqxyo * sign(bxyo)
+	rsqxy <- bxy^2 * vx / vy
+	rsqxys <- rsqxy * sign(bxy)
 	rsqgx <- bgx^2 / (vg * bgx^2 + vu * bux^2 + vex)
 	rsqgy <- (bxy^2 * bgx^2 * vg^2) / (bxyo^2 * bgx^2 * vg + bxyo^2 * bux^2 * vu + buy^2 * vu + vey)
 
 	rsqux <- bux^2 / (vg*bgx^2 + vu*bux^2 + vex)
[…]

### 2017-07-17 scripts/unmeasured_confounding.R
updated

@@ -100,7 +100,7 @@ ggplot(dat, aes(x=rsquxs, y=rsquys)) +
 geom_point(aes(colour=as.factor((rsqgx-rsqgy)>0))) +
 facet_grid(. ~ rsqxysog) +
 labs(x=expression(Signed~R[ux]^2), y=expression(Signed~R[uy]^2), colour="Correct Steiger\ntest result")
-ggsave("../images/unmeasured_conf_steiger.pdf", width=10, height=6)
+ggsave("../images/unmeasured_conf_steiger.pdf", width=12, height=8)
 
 
 ggplot(dat, aes(x=bux, y=buy)) +

### 2017-07-17 docs/manuscript.Rmd
updated

@@
 
+# fig_nums(
+# 	"causality_exists_tpr",
+# 	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
+True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+# 	display=FALSE
+# )
+
 fig_nums(
-	"causality_exists_tpr",
-	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+	"bdmr", 
+	caption="Using the same simulations as in Figure 2, but allowing an instrument for both the exposure and the outcome, bi-direcitonal MR was performed under different levels of measurement error in the exposure.
+The y-axis shows the probability for each of the three possible outcomes - obtaining the correct causal direction, obtaining the incorrect causal direction, or having insufficient power to detect a causal relationship.
+As measurement imprecision increases (decreasing values on x-axis) the power decreases but bi-directional MR is not liable to giving the wrong causal direction.",
 	display=FALSE
@@
 where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.
-Similarly in the non-causal model:
+The simulations in `r cfig("bdmr")` are similar except we also simulate a second SNP which has a direct influence on $y$.
+MR can therefore be performed using the first SNP for estimating the effect of $x$ on $y$, and using the second SNP for estimating the effect of $y$ on $x$.
+
+We have formulated the the non-causal model as:
 
@@
 
-All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distribted values of $\epsilon_*$ were generated such that
+Here the $u$ variable mediates the influence of $g$ on $x$ and $g$ on $y$.
+The effects of $u$ on $x$ and $y$ are identical, therefore this simply attenuates the variance explained by $g$ on $x$ and $y$.
+All $\alpha$ values were set to 0, and $\beta$ values set to 1.
+Normally distribted values of $\epsilon_*$ were generated such that
 
@@
 
-We also performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
-`r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association.
-We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure.
-When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure.
-Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.
+The bi-directional MR approach does offer a solution to this problem (`r cfig("bdmr")`)
 
@@
 
-```{r causality_exists_tpr, fig.width=10, fig.height=10 }
+<!-- ```{r causality_exists_tpr, fig.width=10, fig.height=10 }
 
@@
 \newpage
+
+ -->
 

### 2017-07-17 scripts/analysis.R

@@ -73,3 +73,15 @@ labs(y="Frequency", x=expression(cor(X, X[p])), colour="")
 ggsave("~/repo/MethylationIV/cit/images/proxy_bdmr.pdf")
 
 
+
+
+
+
+dat$steiger <- psych::r.test(n=dat$n, n2=dat$n, r12=sqrt(dat$rsqG), r34=sqrt(dat$rsqT))$p
+
+a <- r.test(n=dat$n, n2=dat$n, r12=sqrt(dat$rsqG), r34=sqrt(dat$rsqT))
+
+dat$steiger_d <- dat$rsqG > dat$rsqT
+
+table(dat$steiger < 0.05 & dat$steiger_d)
+table(dat$steiger < 0.05 & !dat$steiger_d)

### 2017-07-17 scripts/mr_directionality_analysis.R

@@ -202,7 +202,7 @@ get_cit_direction <- function(p1, p2, thresh)
 	return(res)
 }
 
-parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.05)
+parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.25)
 parameters$cit_correct_direction[index_causal] <- parameters$cit_res[index_causal] == 1
 parameters$cit_correct_direction[index_ncausal] <- parameters$cit_res[index_ncausal] == 3
 
@@ -268,6 +268,33 @@ psum1 <- pl %>%
 		prop_nonsig=sum(direction_p_value <= -log10(0.05) | test_p_value <= -log10(0.05), na.rm=T)/n()
 	)
 psum1$rhs_lhs_diff_bin <- cut(psum1$rhs_lhs_diff, breaks=seq(-0.2, 1.1, by=0.1))
+
+
+
+## ---- causality_exists_tpr ----
+
+
+pl$causality_exists <- TRUE
+pl$causality_exists[pl$test == "MR"] <- pl$p_bz[pl$test == "MR"] > -log10(0.05)
+pl$causality_exists[pl$test == "CIT"] <- pl$cit_res[pl$test == "CIT"] <= 3
+pl$rhs_lhs_diff_bin <- cut(pl$rhs_lhs_diff, breaks=10)
+temp2 <- do.call(rbind, strsplit(as.character(pl$rhs_lhs_diff_bin), split=","))
+pl$rhs_lhs_diff_bin_numeric <- as.factor(gsub("\\(", "", temp2[,1]))
+pl$rhs_lhs_diff_bin_lab <- as.factor(paste0("d = ", pl$rhs_lhs_diff_bin_numeric))
+pl$rhs_lhs_diff_bin_lab <- factor(pl$rhs_lhs_diff_bin_lab, levels=levels(pl$rhs_lhs_diff_bin_lab)[order(as.numeric(as.character(levels(pl$rhs_lhs_diff_bin_numeric))))])
+
+temp <- dplyr::group_by(pl, test, r_ab, r_za, n, noiseb) %>%
+	dplyr::summarise(prop_causality_exists = sum(causality_exists) / n())
+temp$r_za <- paste0("cor(x,g) = ", temp$r_za)
+temp$n <- paste0("n = ", temp$n)
[…]

### 2017-07-17 scripts/proxy.R

@@ -146,6 +146,8 @@ inferCausality <- function(dat, threshold = 0.01)
 }
 
 
+
+
 makeProxy <- function(x, noise, bias)
 {
 	y <- x*bias + rnorm(length(x), 0, sd=noise)
@@ -188,6 +190,9 @@ runSim <- function(N, NoiseT=seq(0.1, 1, by=0.1), BiasT=seq(0.1, 1, by=0.1), Noi
 			dat$bdmr_p[i] <- inferCausality(twoStageLS(G1, T1, L1, L2, FALSE))
 			dat$rsqG[i] <- cor(G1, G)^2
 			dat$rsqT[i] <- cor(T1, T)^2
+			dat$rL1G[i] <- cor(L1, G1)
+			dat$rL1T[i] <- cor(L1, T1)
+
 		}
 		dat$n <- n
 		d[[j]] <- dat
@@ -195,9 +200,32 @@ runSim <- function(N, NoiseT=seq(0.1, 1, by=0.1), BiasT=seq(0.1, 1, by=0.1), Noi
 	dat <- rbind.fill(d)
 	return(dat)
 }
-
 arguments <- commandArgs(T)
 n <- as.numeric(arguments[1])
 dat <- runSim(n)
 filename <- paste0("~/repo/cit_measurement_error/results/20160504_", n, ".RData")
 save(dat, file=filename)
+
+
+# n <- 100
+# dat1 <- runSim(n)
+
+# n <- 1000
+# dat2 <- runSim(n)
+
+# n <- 10000
+# dat3 <- runSim(n)
+
+
+# dat <- rbind(dat1, dat2, dat3)
+
+
+# dat$steiger <- psych::r.test(n=dat$n, n2=dat$n, r12=dat$rL1G, r34=dat$rL1T)$p
+# dat$steiger_d <- dat$rL1G^2 > dat$rL1T^2
+
+# ggplot(dat, aes(x=noiseG, y=-log10(steiger))) +
+# geom_point(aes(colour=steiger_d)) +
+# facet_grid(n ~ .)
+
+
+# save(dat, file="../re")

### 2017-07-18 docs/manuscript.Rmd
updated

@@
 	caption="Summary of MR analysis on gene expression and DNA methylation associations.
-'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the Steiger test.
+'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the MR Steiger test.
 P(+ve effect) denotes the proportion of associations that were positive.
@@
 The causal inference test (CIT) attempts to distinguish between them.
-a) Methylation causes gene expression.
-The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels.
+a) Gene expression causes methylation.
+The left figure shows that the SNP influences gene expression levels that in turn influence methylation levels.
 The right figure shows the directed acyclic graph that represents this model.
 Faded symbols represent the measured values whereas solid symbols represent the true values.
-b) The same as in A, except the causal direction is from gene expression to DNA methylation.
+b) The same as in A, except the causal direction is from DNA methylation to Gene expression.
 c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
@@
 
+fig_nums(
+	"causality_exists_tpr",
+	caption="Outcomes were simulated to be unrelated to the exposure (bottom plot) or causally influenced by the exposure (top plot) with varying degrees of measurement imprecision applied to the exposure variable (x axis).
+True positive rates (y axis) for MR and CIT were compared for varying sample sizes (columns of boxes).",
+	display=FALSE
+)
+
 # fig_nums(
-# 	"causality_exists_tpr",
-# 	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
+# 	"bdmr", 
+# 	caption="Using the same simulations as in Figure 2, but allowing an instrument for both the exposure and the outcome, bi-direcitonal MR was performed under different levels of measurement error in the exposure.
+The y-axis shows the probability for each of the three possible outcomes - obtaining the correct causal direction, obtaining the incorrect causal direction, or having insufficient power to detect a causal relationship.
+As measurement imprecision increases (decreasing values on x-axis) the power decreases but bi-directional MR is not liable to giving the wrong causal direction.",
 # 	display=FALSE
@@
 fig_nums(
-	"bdmr", 
-	caption="Using the same simulations as in Figure 2, but allowing an instrument for both the exposure and the outcome, bi-direcitonal MR was performed under different levels of measurement error in the exposure.
-The y-axis shows the probability for each of the three possible outcomes - obtaining the correct causal direction, obtaining the incorrect causal direction, or having insufficient power to detect a causal relationship.
-As measurement imprecision increases (decreasing values on x-axis) the power decreases but bi-directional MR is not liable to giving the wrong causal direction.",
-	display=FALSE
-)
-
-fig_nums(
 	"steiger_sensitivity_plot", 
-	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface.
-When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction.
+	caption="a) We can predict the values the MR Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface.
+When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed MR Steiger test inferring the wrong causal direction.
 Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y.
@@
 b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative.
-When $d$ is negative the Steiger test is liable to infer the wrong direction of causality.
+When $d$ is negative the MR Steiger test is liable to infer the wrong direction of causality.
 Shaded regions show the parameter space where $d$ is negative.
@@
 CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality.
-The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality.
+The columns of graphs denote intervals for he value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the MR Steiger test to be more likely to be wrong about the direction of causality.
 Rows of graphs represent the sample size used in the simulations.
@@
 b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$.
-MR is unable to identify this model, so any significant associations are deemed to be incorrect.
-Outcome 3 denotes evidence for the correct model for the CIT method.",
+Neither CIT nor MR are able to identify this model, so any significant associations in MR are deemed to be incorrect, while outcomes 1 or 2 for the CIT are deemed to be incorrect.",
 	display=FALSE
@@
 	"shakhplot",
-	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them.
+	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the MR Steiger test to infer the direction of causality between them.
 a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels.
@@
 	"unmeasured",
-	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the Steiger test returning the wrong causal direction.
+	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
 Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$)."),
@@
 
-Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others.
-However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
+Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT).
+However, we show that measurement error in the phenotypes can lead to the CIT inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
+This problem is likely to be general to other mediation-based approaches.
 Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages.
@@
 To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases.
-Here we demonstrate that when there is uncertainty about the nature of the genetic association several existing methods that are used for causal inference can be liable to give the wrong direction of causality.
+Here we demonstrate that when there is uncertainty about the nature of the genetic association existing methods that are used for causal inference can be liable to give the wrong direction of causality.
 We introduce a new method that is more flexible and robust to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression.
@@
 In response to this, the use of genetic associations to instrument traits has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015].
-However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
+However, shifting from observational associations to instrumentation does require more (often untestable) assumptions, and potential pitfalls remain.
+One that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.
 
@@
 Measurement error will of course arise in other types of data too.
-For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013].
+For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013], leading to the problem that phenotypes may be imprecisely defined.
 A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately.
@@
 Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016].
+Additionally, if valid genetic instruments are known for both traits of interest then MR can be performed in both directions (bi-directional MR), testing the influence of one trait on the other and vice versa, to infer the causal direction between the two phenotypes.
 
@@
 Second, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome is affected by unmeasured confounding of the exposure and outcome.
-This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders.
+This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders.
 MR does not suffer from this problem because it does not test for association through adjustment.
@@
 Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method.
+Though we only examine the CIT method in detail, we show that attempting to adjust for mediating variables to make causal inference is susceptible to problems, which can be generalised to other mediation-based methods.
 We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error.
-Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially.
-Finally, we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels.
+Finally, to demonstrate the potential impact of measurement error we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels.
 Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].
@@
 where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
+Measurement imprecision can represent imprecise measurement due to limits on sensitivity of measuring equipment, or due to phenotypes being imprecisely defined.
 The same model of measurement error can be applied to the outcome variable $y$.
-In this study we assume there is no measurement error in the SNP, and that measurement error in the exposure and the outcome are uncorrelated.
+
+In this study we assume there is no measurement error in the SNP.
+Common genetic variants are typically less susceptible to measurement error due to strict quality control procedures prior to genome wide association studies.
+Any non-differential measurement error that might be present (either because the SNP is poorly typed or because the SNP is not in complete linkage disequilibrium with the causal variant) will reduce power in MR but will not incur bias [@DaveySmith2004; @Pierce2012; @LeCessie2012].
+We also assume that measurement error in the exposure and the outcome are uncorrelated.
 
@@
 - If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
-- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is likely confounding and no call is made ($x \leftarrow u \rightarrow y$).
 
@@
 Second, we integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values, assuming that any measurement error value is equally likely.
-We find the ratio of the volume that agrees with the inferred direction of causality over the volume that disagrees with the inferred direction of causality.
-A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality.
+Across all possible values of measurement error in $x$ and $y$ we find the volume that agrees with the inferred direction of causality and the volume that disagrees with the inferred direction of causality, and take the ratio of these two values.
+A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, because equal weight of the measurement error parameter space supporting both directions of causality.
 In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction.
@@
 
-where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.
-The simulations in `r cfig("bdmr")` are similar except we also simulate a second SNP which has a direct influence on $y$.
-MR can therefore be performed using the first SNP for estimating the effect of $x$ on $y$, and using the second SNP for estimating the effect of $y$ on $x$.
-
+where non-differential measurement error is represented by a noise (measurement imprecision) term $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and measurement bias terms $\alpha_{m*}$ and $\beta_{m*}$ for the exposure variable $x$ and the outcome variable $y$.
+Note that following the first section of the Results we no longer include the bias terms for simplicity.
 We have formulated the the non-causal model as:
@@
 g & \sim Binom(2, 0.5) \\
-u & = \alpha_g + \beta_g g + \epsilon_g \\
-x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
+x & = \alpha_{gx} + \beta_{gx} g + \epsilon_{gx} \\
 x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
+y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
 y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
@@
 
-Here the $u$ variable mediates the influence of $g$ on $x$ and $g$ on $y$.
-The effects of $u$ on $x$ and $y$ are identical, therefore this simply attenuates the variance explained by $g$ on $x$ and $y$.
-All $\alpha$ values were set to 0, and $\beta$ values set to 1.
-Normally distribted values of $\epsilon_*$ were generated such that
+All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distributed values of $\epsilon_*$ were generated such that
 
@@
 cor(g, x)^2 & = 0.1 \\
-cor(g, u)^2 & = 0.1 \\
 cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
@@
 We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}1$) the CIT is reliable in identifying the correct causal direction.
 However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
 Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
-
-The bi-directional MR approach does offer a solution to this problem (`r cfig("bdmr")`)
 
@@
 If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
-Here we introduce the MR Steiger approach to similarly orient the direction of causality but in an MR analysis when the underlying biology of the SNP is not fully understood.
+Though bi-directional MR can be used to orient causal directions, this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood.
+We go on to explore the scenarios in which the method is likely to return the correct or incorrect causal directions.
 
-For a particular association, it is of interest to identify the range of possible measurement error values agree and disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2).
-This metric can be used to evaluate the reliability of MR Steiger.
+We performed simulations to compare the power and type 1 error rates of MR and CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+Comparing the performance of methods with different sets of assumptions can be difficult, but a basic comparison is shown in `r cfig("causality_exists_tpr")`.
+We observe that the CIT is more conservative under the null model of no association owing to the omnibus test statistic comprising several statistical tests.
+The FDR using a p-value threshold of 0.05 appears to be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
+Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases faster than that of the MR Steiger method.
 
-We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction).
-`r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
-It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.
+For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2).
+This metric can be used to evaluate the reliability of MR Steiger test.
+
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
+`r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
+It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the MR Steiger test to measurement error is relatively low.
+
+Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Appendix 3).
+Though the relationship between unmeasured confounding and causal orientation is complex (`r sfig_nums("unmeasured")`), we note that when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction.
 
@@
 
-We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
+We used simulations to explore the performance of the MR Steiger approach in comparison to CIT for different levels of measurement error.
+The performance was compared in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
 Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
@@
 However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality.
-When $d > 0$ we find that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
-The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.
+When $d > 0$ we find that in most cases the MR Steiger method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+The CIT, unlike the MR Steiger test, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.
 
@@
 
-We used the Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations.
+We used the MR Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations.
 We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
@@
 `r cfig("shakhplot")`a shows that the conclusions about the direction of causality between DNA methylation and gene expression are very sensitive to measurement error.
+We made a strong assumption that either methylation influenced gene expression or vice versa, but it is certainly possible that the SNP is solely or additionally influencing some other trait that confounds the association between gene expression and DNA methylation.
 
@@
 In the epidemiological literature issues of measurement error in mediation analysis are relatively well explored [@Cole2014].
-Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data.
+Our analysis extends this to related methods such as CIT that are used in predominantly 'omic data.
 These methods are indeed susceptible to the same problem as standard mediation based analysis, and specifically we show that as measurement error in the (true) exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
@@
 
-<!-- ```{r causality_exists_tpr, fig.width=10, fig.height=10 }
+```{r causality_exists_tpr, fig.width=10, fig.height=10 }
 
@@
 \newpage
-
- -->
 
@@
 
-### Unmeasured confounding in the Steiger test
+### Unmeasured confounding in the MR Steiger test
 
-Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
+Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the MR Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
 After some algebra
@@
 `r csfig("unmeasured")` shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$.
-We note that in most cases where the observational absolute $R^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the Steiger test.
+We note that in most cases where the observational absolute $R^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
 

### 2017-07-18 docs/manuscript.tex
updated

@@
 \hypersetup{unicode=true,
-            pdftitle={Orienting the causal relationship between imprecisely measured traits using genetic instruments},
+            pdftitle={Orienting the causal relationship between imprecisely measured traits using GWAS summary data},
             pdfborder={0 0 0},
@@
   \title{Orienting the causal relationship between imprecisely measured traits
-using genetic instruments}
+using GWAS summary data}
   \pretitle{\vspace{\droptitle}\centering\huge}
@@
   \postdate{\par}
-  \date{16 July 2017}
+  \date{18 July 2017}
 
@@
 traits can be achieved by combining genetic associations with a
-mediation-based approach, as is done in the causal inference test (CIT)
-and others. However, we show that measurement error in the phenotypes
-can lead to mediation-based approaches inferring the wrong causal
-direction, and that increasing sample sizes has the adverse effect of
-increasing confidence in the wrong answer. Here we introduce an
-extension to Mendelian randomisation, a method that uses genetic
-associations in an instrumentation framework, that enables inference of
-the causal direction between traits, with some advantages. First, it is
-less susceptible to bias in the presence of measurement error; second,
-it is more statistically efficient; third, it can be performed using
-only summary level data from genome-wide association studies; and
-fourth, its sensitivity to measurement error can be evaluated. We apply
-the method to infer the causal direction between DNA methylation and
-gene expression levels. Our results demonstrate that, in general, DNA
-methylation is more likely to be the causal factor, but this result is
-highly susceptible to bias induced by systematic differences in
-measurement error between the platforms. We emphasise that, where
-possible, implementing MR and appropriate sensitivity analyses alongside
-other approaches such as CIT is important to triangulate reliable
-conclusions about causality.
+mediation-based approach, as is done in the causal inference test (CIT).
+However, we show that measurement error in the phenotypes can lead to
+the CIT inferring the wrong causal direction, and that increasing sample
+sizes has the adverse effect of increasing confidence in the wrong
+answer. This problem is likely to be general to other mediation-based
+approaches. Here we introduce an extension to Mendelian randomisation, a
+method that uses genetic associations in an instrumentation framework,
+that enables inference of the causal direction between traits, with some
+advantages. First, it can be performed using only summary level data
+from genome-wide association studies; second, it is less susceptible to
+bias in the presence of measurement error or unmeasured confounding. We
+apply the method to infer the causal direction between DNA methylation
+and gene expression levels. Our results demonstrate that, in general,
+DNA methylation is more likely to be the causal factor, but this result
+is highly susceptible to bias induced by systematic differences in
+measurement error between the platforms, and by horizontal pleiotropy.
+We emphasise that, where possible, implementing MR and appropriate
+sensitivity analyses alongside other approaches such as CIT is important
+to triangulate reliable conclusions about causality.
 
@@
 demonstrate that when there is uncertainty about the nature of the
-genetic association several existing methods that are used for causal
-inference can be liable to give the wrong direction of causality. We
-introduce a new method that is more flexible and robust to problems such
-as measurement error, and apply it to evaluate the causal relationships
+genetic association existing methods that are used for causal inference
+can be liable to give the wrong direction of causality. We introduce a
+new method that is more flexible and robust to problems such as
+measurement error, and apply it to evaluate the causal relationships
 between DNA methylation levels and gene expression. Our results show
@@
 prominent tool that is applied in several different guises (3--6).
-However, potential pitfalls remain and one that is often neglected is
-the influence of non-differential measurement error on the reliability
-of causal inference.
+However, shifting from observational associations to instrumentation
+does require more (often untestable) assumptions, and potential pitfalls
+remain. One that is often neglected is the influence of non-differential
+measurement error on the reliability of causal inference.
 
@@
 a proxy for adiposity, but it is clear that the correlation between BMI
-and underlying adiposity is not perfect (12). A similar problem of
+and underlying adiposity is not perfect (12), leading to the problem
+that phenotypes may be imprecisely defined. A similar problem of
 biological misspecification is unavoidable in disease diagnosis, and
@@
 available for a putative exposure (24,25) and a number of sensitivity
-tests are now available to improve reliability (26).
+tests are now available to improve reliability (26). Additionally, if
+valid genetic instruments are known for both traits of interest then MR
+can be performed in both directions (bi-directional MR), testing the
+influence of one trait on the other and vice versa, to infer the causal
+direction between the two phenotypes.
 
@@
 outcome by the exposure induces a collider effect between the SNP and
-outcome (29), and the in order to fully abrogate this association one
-must also adjust for all (hidden or otherwise) confounders. MR does not
+outcome (29), and in order to fully abrogate this association one must
+also adjust for all (hidden or otherwise) confounders. MR does not
 suffer from this problem because it does not test for association
@@
 error in phenotypes can lead to unreliable causal inference in the
-mediation-based CIT method. We then present an extension to MR that
-allows researchers to ascertain the causal direction of an association
-even when the biology of the instruments are not fully understood, and
-also a metric to evaluate the sensitivity of the result of this
-extension to measurement error. Together these extensions improve the
-utility of MR in cases where mediation based methods might have
-otherwise been used preferentially. Finally, we apply this method to
-infer the direction of causation between DNA methylation levels and gene
-expression levels. Our analyses highlight that because these different
-causal inference techniques have varying strengths and weaknesses,
-triangulation of evidence from as many sources as possible should be
-practiced in causal inference (36).
+mediation-based CIT method. Though we only examine the CIT method in
+detail, we show that attempting to adjust for mediating variables to
+make causal inference is susceptible to problems, which can be
+generalised to other mediation-based methods. We then present an
+extension to MR that allows researchers to ascertain the causal
+direction of an association even when the biology of the instruments are
+not fully understood, and also a metric to evaluate the sensitivity of
+the result of this extension to measurement error. Finally, to
+demonstrate the potential impact of measurement error we apply this
+method to infer the direction of causation between DNA methylation
+levels and gene expression levels. Our analyses highlight that because
+these different causal inference techniques have varying strengths and
+weaknesses, triangulation of evidence from as many sources as possible
+should be practiced in causal inference (36).
 
@@
 measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
-represents the imprecision (or noise) in the measurement of \(x\). The
-same model of measurement error can be applied to the outcome variable
-\(y\). In this study we assume there is no measurement error in the SNP,
-and that measurement error in the exposure and the outcome are
-uncorrelated.
+represents the imprecision (or noise) in the measurement of \(x\).
+Measurement imprecision can represent imprecise measurement due to
+limits on sensitivity of measuring equipment, or due to phenotypes being
+imprecisely defined. The same model of measurement error can be applied
+to the outcome variable \(y\).
+
+In this study we assume there is no measurement error in the SNP. Common
+genetic variants are typically less susceptible to measurement error due
+to strict quality control procedures prior to genome wide association
+studies. Any non-differential measurement error that might be present
+(either because the SNP is poorly typed or because the SNP is not in
+complete linkage disequilibrium with the causal variant) will reduce
+power in MR but will not incur bias (3,13,30). We also assume that
+measurement error in the exposure and the outcome are uncorrelated.
 
@@
   If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
-  accepted (\(x \leftarrow g \rightarrow y\)).
+  \(p_{CIT, y \rightarrow x} < \alpha\) then there is likely confounding
+  and no call is made (\(x \leftarrow u \rightarrow y\)).
 \end{itemize}
@@
 Note that the same correlation test approach can be applied to a
-two-sample MR (28) setting. Two-sample MR refers to the case where the
+two-sample MR setting. Two-sample MR refers to the case where the
 SNP-exposure association and SNP-outcome association are calculated in
-different samples (e.g.~from publicly available summary statistics).
-Here the Steiger test of two independent correlations can be applied
-where.
+different samples (e.g.~from publicly available summary statistics
+(26,28)). Here the Steiger test of two independent correlations can be
+applied where.
 
@@
 
-\subsubsection{Causal direction sensitivity
-analysis}\label{causal-direction-sensitivity-analysis}
+The Steiger test assumes that there is a causal relationship between the
+two variables, and that the SNP is a valid instrument for one of them.
+However it is liable to give incorrect causal directions under some
+other circumstances also. First, some levels of horizontal pleiotropy,
+where the SNP influences the outcome through some pathway other than the
+exposure, could induce problems because this is a means by which the
+instrument is invalid. Second, some differential values of measurement
+error between the exposure and the outcome could induce incorrect causal
+directions, and this is explored in Appendix 2. Third, some levels of
+unmeasured confounding between the exposure and the outcome could induce
+the wrong causal direction, and this is explored in Appendix 3.
+
+\subsubsection{Causal direction sensitivity analysis for measurement
+error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
 
@@
 the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
-error. We integrate over the entire range of \(\rho_{gy} - \rho_{gx}\)
-values for possible measurement error values. We find the ratio of the
-volume that agrees with the inferred direction of causality over the
-volume that disagrees with the inferred direction of causality. A ratio
-\(R=1\) indicates that the inferred causal direction is highly sensitive
-to measurement error, with equal weight of the measurement error
+error. We offer two tools in which to do this. First, the user can
+provide values of measurement error for \(x\) and \(y\) and obtain a
+revised inference of the causal direction. Second, we integrate over the
+entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
+measurement error values, assuming that any measurement error value is
+equally likely. Across all possible values of measurement error in \(x\)
+and \(y\) we find the volume that agrees with the inferred direction of
+causality and the volume that disagrees with the inferred direction of
+causality, and take the ratio of these two values. A ratio \(R=1\)
+indicates that the inferred causal direction is highly sensitive to
+measurement error, because equal weight of the measurement error
 parameter space supporting both directions of causality. In general, the
@@
 
-where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
-\(\beta_{mx}\) are parameters that represent non-differential
-measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
-and \(\beta_{my}\) are parameters for non-differential measurement error
-in the outcome \(y\). Similarly in the non-causal model:
+where non-differential measurement error is represented by a noise
+(measurement imprecision) term
+\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
+\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
+the outcome variable \(y\). Note that following the first section of the
+Results we no longer include the bias terms for simplicity. We have
+formulated the the non-causal model as:
 
@@
 All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
-Normally distribted values of \(\epsilon_*\) were generated such that
+Normally distributed values of \(\epsilon_*\) were generated such that
 
@@
 the CIT method. Figure 2 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
-reliable in identifying the correct causal direction. However, as
-measurement error increases in the exposure variable, eventually the CIT
-is more likely to infer a robust causal association in the wrong
-direction. Also of concern here is that increasing sample size does not
-solve the issue, indeed it only strengthens the apparent evidence for
-the incorrect inference.
-
-We also performed simulations to compare the performance of MR against
-CIT in detecting a causal association between simulated variables under
-different levels of imprecision simulated in the exposure. Figure 3
-shows the true positive rates between the CIT and MR for detecting a
-causal association. We observe that the CIT has lower power in all
-cases, with performance declining as measurement imprecision increases
-in the exposure. When MR assumptions are satisfied, notably that it is
-known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
-the performance of MR in detecting an association is unrelated to
-measurement error in the exposure. Measurement error in the outcome does
-reduce power, but does not induce a substantive difference in
-performance between CIT and MR.
+in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}1\))
+the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases in the exposure variable,
+eventually the CIT is more likely to infer a robust causal association
+in the wrong direction. Also of concern here is that increasing sample
+size does not solve the issue, indeed it only strengthens the apparent
+evidence for the incorrect inference.
 
@@
 If we do not know whether the SNP \(g\) has a primary influence on \(x\)
-or \(y\) then CIT can attempt to infer the causal direction. Here we
-introduce the MR Steiger approach to similarly orient the direction of
-causality but in an MR analysis when the underlying biology of the SNP
-is not fully understood.
+or \(y\) then CIT can attempt to infer the causal direction. Though
+bi-directional MR can be used to orient causal directions, this requires
+knowledge of a valid instrument for each trait, and we were motivated to
+develop the MR Steiger method that could operate on summary data to
+orient the direction of causality using the same conditions as the CIT,
+where the underlying biology of a single SNP is not fully understood. We
+go on to explore the scenarios in which the method is likely to return
+the correct or incorrect causal directions.
+
+We performed simulations to compare the power and type 1 error rates of
+MR and CIT in detecting a causal association between simulated variables
+under different levels of imprecision simulated in the exposure.
+Comparing the performance of methods with different sets of assumptions
+can be difficult, but a basic comparison is shown in Figure 3. We
+observe that the CIT is more conservative under the null model of no
+association owing to the omnibus test statistic comprising several
+statistical tests. The FDR using a p-value threshold of 0.05 appears to
+be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
+Using the same p-value thresholds to declare significance in the
+non-null simulations, the general trend appears to be that the CIT power
+reduces as measurement error in the exposure increases faster than that
+of the MR Steiger method.
 
 For a particular association, it is of interest to identify the range of
-possible measurement error values agree and disagree with the
-empirically inferred causal direction (Figure 4a, Appendix 2). This
-metric can be used to evaluate the reliability of MR Steiger.
+possible measurement error values for which the method will give results
+that agree or disagree with the empirically inferred causal direction
+(Figure 4a, Appendix 2). This metric can be used to evaluate the
+reliability of MR Steiger test.
 
@@
 \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
-the range of parameters around which the Steiger test is liable to
+the range of parameters around which the MR Steiger test is liable to
 provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
-the Steiger test is likely to be correct about the causal direction).
-Figure 4b shows that when there is no measurement error in \(x\), the
+the MR Steiger test is likely to be correct about the causal direction).
+Figure 4b shows that when there is no measurement error in \(x\), the MR
 Steiger test is unlikely to infer the wrong direction of causality even
@@
 where \(x\) is measured with error, especially when the causal effect
-between \(x\) and \(y\) is not very large, the sensitivity of the
+between \(x\) and \(y\) is not very large, the sensitivity of the MR
 Steiger test to measurement error is relatively low.
+
+Unmeasured confounding between the exposure and outcome can also give
+rise to problems with the MR Steiger approach (Appendix 3). Though the
+relationship between unmeasured confounding and causal orientation is
+complex (Supplementary Figure 1: Graph representing the unmeasured
+confounding parameters that will lead to the MR Steiger test returning
+the wrong causal direction. Columns of boxes represent different signed
+values of the observational variance explained between \(x\) and \(y\)
+(\(R^2_{xy}\)).), we note that when the magnitude of the observational
+variance explained between the exposure and the outcome is below 0.2 the
+MR Steiger method is unlikely to return the incorrect causal direction.
 
@@
 We used simulations to explore the performance of the MR Steiger
-approach in comparison to CIT in terms of the rate at which evidence of
-a causal relationship is obtained for the correct direction of
-causality, and the rate at which evidence of a causal relationship is
-obtained where the reported direction of causality is incorrect.
-Simulations were performed for two models, one for a ``causal model''
-where there was a causal relationship between \(x\) and \(y\); and one
-for a ``non-causal model'' where \(x\) and \(y\) were not causally
-related, but had a confounded association induced by the SNP \(g\)
-influencing a confounder variable \(u\).
+approach in comparison to CIT for different levels of measurement error.
+The performance was compared in terms of the rate at which evidence of a
+causal relationship is obtained for the correct direction of causality,
+and the rate at which evidence of a causal relationship is obtained
+where the reported direction of causality is incorrect. Simulations were
+performed for two models, one for a ``causal model'' where there was a
+causal relationship between \(x\) and \(y\); and one for a ``non-causal
+model'' where \(x\) and \(y\) were not causally related, but had a
+confounded association induced by the SNP \(g\) influencing a confounder
+variable \(u\).
 
@@
 robust causal association for the wrong direction of causality. When
-\(d > 0\) we find that in most cases the MR method has greater power to
-obtain evidence for causality than CIT, and always obtains the correct
-direction of causality. The CIT, unlike the Steiger test for MR, is able
-to distinguish the ``non-causal model'' from the ``causal model''
+\(d > 0\) we find that in most cases the MR Steiger method has greater
+power to obtain evidence for causality than CIT, and always obtains the
+correct direction of causality. The CIT, unlike the MR Steiger test, is
+able to distinguish the ``non-causal model'' from the ``causal model''
 (Methods, Figure 5b), but it is evident that measurement error will
@@
 
-We used the Steiger test to infer the direction of causality between DNA
-methylation and gene expression levels between 458 putative
+We used the MR Steiger test to infer the direction of causality between
+DNA methylation and gene expression levels between 458 putative
 associations. We found that the causal direction commonly goes in both
@@
 methylation and gene expression are very sensitive to measurement error.
+We made a strong assumption that either methylation influenced gene
+expression or vice versa, but it is certainly possible that the SNP is
+solely or additionally influencing some other trait that confounds the
+association between gene expression and DNA methylation.
 
@@
 analysis are relatively well explored (44). Our analysis extends this to
-related methods such as CIT that are widely used in predominantly 'omic
-data. These methods are indeed susceptible to the same problem as
-standard mediation based analysis, and specifically we show that as
-measurement error in the (true) exposure variable increases, CIT is
-likely to have reduced statistical power, and liable to infer the wrong
-direction of causality. We also demonstrate that, though unintuitive,
-increasing sample size does not resolve the issue, rather it leads to
-more extreme p-values for the model that predicts the wrong direction of
-causality.
+related methods such as CIT that are used in predominantly 'omic data.
+These methods are indeed susceptible to the same problem as standard
+mediation based analysis, and specifically we show that as measurement
+error in the (true) exposure variable increases, CIT is likely to have
+reduced statistical power, and liable to infer the wrong direction of
+causality. We also demonstrate that, though unintuitive, increasing
+sample size does not resolve the issue, rather it leads to more extreme
+p-values for the model that predicts the wrong direction of causality.
 
@@
 variables. The causal inference test (CIT) attempts to distinguish
-between them. a) Methylation causes gene expression. The left figure
-shows that the SNP influences methylation levels that in turn influence
-gene expression levels. The right figure shows the directed acyclic
-graph that represents this model. Faded symbols represent the measured
-values whereas solid symbols represent the true values. b) The same as
-in A, except the causal direction is from gene expression to DNA
-methylation. c) A model of confounding, where gene expression and DNA
-methylation are not causally related, but the SNP influences them each
-through separate pathways or a confounder.
+between them. a) Gene expression causes methylation. The left figure
+shows that the SNP influences gene expression levels that in turn
+influence methylation levels. The right figure shows the directed
+acyclic graph that represents this model. Faded symbols represent the
+measured values whereas solid symbols represent the true values. b) The
+same as in A, except the causal direction is from DNA methylation to
+Gene expression. c) A model of confounding, where gene expression and
+DNA methylation are not causally related, but the SNP influences them
+each through separate pathways or a confounder.
 
@@
 
-Figure 3: Outcomes were simulated to be causally influenced by exposures
-with varying degrees of measurement imprecision applied to the exposure
+Figure 3: Outcomes were simulated to be unrelated to the exposure
+(bottom plot) or causally influenced by the exposure (top plot) with
+varying degrees of measurement imprecision applied to the exposure
 variable (x axis). True positive rates (y axis) for MR and CIT were
-compared for varying levels of measurement imprecision in the outcome
-variable (rows of boxes) and sample sizes (columns of boxes).
+compared for varying sample sizes (columns of boxes).
 
@@
 
-Figure 4: a) We can predict the values the Steiger test would take
+Figure 4: a) We can predict the values the MR Steiger test would take
 (z-axis) for different potential values of measurement error (x and y
@@
 as denoted by the range of values where the blue surface is above the
-red plane, those values of measurement error lead to our observed
+red plane, those values of measurement error lead to our observed MR
 Steiger test inferring the wrong causal direction. Where the blue
@@
 which the function \(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is
-negative. When \(d\) is negative the Steiger test is liable to infer the
-wrong direction of causality. Shaded regions show the parameter space
-where \(d\) is negative. The graph shows that for the majority of the
-parameter space of the function, \(d\) is positive, especially where
+negative. When \(d\) is negative the MR Steiger test is liable to infer
+the wrong direction of causality. Shaded regions show the parameter
+space where \(d\) is negative. The graph shows that for the majority of
+the parameter space of the function, \(d\) is positive, especially where
 causal relationships are relatively weak.
@@
 the exposure and outcome, and to infer the direction of causality. The
-value of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when
-\(d\) is negative we expect the Steiger test to be more likely to be
-wrong about the direction of causality. Rows of graphs represent the
-sample size used in the simulations. For the CIT method, outcome 1
-denoted evidence for causality with correct model, outcomes 2 or 3
-denoted evidence for causality with incorrect model, and outcome 4
-denoted no evidence for causality. b) As in (a) except the simulated
-model was non-causal, and a genetic confounder induces an association
-between \(x\) and \(y\). MR is unable to identify this model, so any
-significant associations are deemed to be incorrect. Outcome 3 denotes
-evidence for the correct model for the CIT method.
+columns of graphs denote intervals for he value of
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\) is
+negative we expect the MR Steiger test to be more likely to be wrong
+about the direction of causality. Rows of graphs represent the sample
+size used in the simulations. For the CIT method, outcome 1 denoted
+evidence for causality with correct model, outcomes 2 or 3 denoted
+evidence for causality with incorrect model, and outcome 4 denoted no
+evidence for causality. b) As in (a) except the simulated model was
+non-causal, and a genetic confounder induces an association between
+\(x\) and \(y\). Neither CIT nor MR are able to identify this model, so
+any significant associations in MR are deemed to be incorrect, while
+outcomes 1 or 2 for the CIT are deemed to be incorrect.
 
@@
 Figure 6: Using 458 putative associations between DNA methylation and
-gene expression we used the Steiger test to infer the direction of
+gene expression we used the MR Steiger test to infer the direction of
 causality between them. a) The rightmost bar shows the proportion of
@@
 
-\subsubsection{Unmeasured confounding in the Steiger
-test}\label{unmeasured-confounding-in-the-steiger-test}
+\subsubsection{Unmeasured confounding in the MR Steiger
+test}\label{unmeasured-confounding-in-the-mr-steiger-test}
 
@@
 will lead to the wrong causal direction between \(x\) and \(y\) using
-the Steiger test, evaluating the liability for the inequality
+the MR Steiger test, evaluating the liability for the inequality
 \(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
@@
 than 0.2, unmeasured confounding will not incur the wrong causal
-direction in the Steiger test.
+direction in the MR Steiger test.
 
@@
 Supplementary Figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the Steiger test returning the wrong causal
-direction. Columns of boxes represent different signed values of the
-observational variance explained between \(x\) and \(y\) (\(R^2_{xy}\)).
+parameters that will lead to the MR Steiger test returning the wrong
+causal direction. Columns of boxes represent different signed values of
+the observational variance explained between \(x\) and \(y\)
+(\(R^2_{xy}\)).
 

### 2017-07-18 scripts/mr_directionality_analysis.R

@@ -62,6 +62,9 @@ steiger_sensitivity <- function(rgx_o, rgy_o, ...)
 		b <- rgy_o
 	}
 
+	mycolors.trans = rgb(c(0,0), c(0,0), 
+               c(0,255),alpha = c(70,255), maxColorValue = 255) 
+
 	d <- expand.grid(rxx_o=seq(rgx_o,1,length.out=50), ryy_o=seq(rgy_o,1,length.out=50), type=c("A","B"))
 	d$rgy <- rgy_o / d$ryy_o
 	d$rgx <- rgx_o / d$rxx_o
@@ -72,7 +75,7 @@ steiger_sensitivity <- function(rgx_o, rgy_o, ...)
 		groups=type, 
 		data=d, 
 		scales=list(arrows=FALSE), 
-		col.groups = colorRampPalette(c("red", "blue"))(2), 
+		col.groups = mycolors.trans, 
 		drape=FALSE, 
 		ylab=expression(rho[xx[o]]), 
 		xlab=expression(rho[yy[o]]),
@@ -202,9 +205,9 @@ get_cit_direction <- function(p1, p2, thresh)
 	return(res)
 }
 
-parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.25)
+parameters$cit_res <- get_cit_direction(parameters$cit_AB, parameters$cit_BA, 0.05)
 parameters$cit_correct_direction[index_causal] <- parameters$cit_res[index_causal] == 1
-parameters$cit_correct_direction[index_ncausal] <- parameters$cit_res[index_ncausal] == 3
+parameters$cit_correct_direction[index_ncausal] <- parameters$cit_res[index_ncausal] == 99
 
 parameters$cit_p[index_causal] <- parameters$cit_AB[index_causal]
 parameters$cit_p[parameters$cit_res == 2 & index_causal] <- parameters$cit_BA[parameters$cit_res == 2 & index_causal]
@@ -213,7 +216,8 @@ parameters$cit_p[parameters$cit_res == 4 & index_causal] <- 0.5
 
 
 
-parameters$cit_p[index_ncausal] <- pmax(parameters$cit_AB[index_ncausal], parameters$cit_BA[index_ncausal])
[…]

### 2017-07-24 docs/manuscript.Rmd
updated

@@
 
-Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT).
+Inference about the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT).
 However, we show that measurement error in the phenotypes can lead to the CIT inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer.
@@
 To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases.
-Here we demonstrate that when there is uncertainty about the nature of the genetic association existing methods that are used for causal inference can be liable to give the wrong direction of causality.
-We introduce a new method that is more flexible and robust to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression.
-Our results show that DNA methylation is more likely to be the causal factor, but that systematic differences in measurement error between the platforms could influence such conclusions.
+Here we discuss scenarios where the nature of the genetic association with the causal trait can pose lead existing causal inference methods to give the wrong direction of causality.
+We introduce a new method that can be applied to summary level data and is potentially less susceptible to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression.
+While our results show that DNA methylation is more likely to be the causal factor, we point out that is it crucial to acknowledge that systematic differences in measurement error between the platforms could influence such conclusions.
 
@@
 Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method.
-Though we only examine the CIT method in detail, we show that attempting to adjust for mediating variables to make causal inference is susceptible to problems, which can be generalised to other mediation-based methods.
+Though we only examine the CIT method in detail, we believe that attempting to adjust for mediating variables to make causal inference is susceptible to problems, which can be generalised to other mediation-based methods.
 We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error.
@@
 where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$.
-Measurement imprecision can represent imprecise measurement due to limits on sensitivity of measuring equipment, or due to phenotypes being imprecisely defined.
+Measurement imprecision can represent imprecise measurement due to limits on sensitivity of measuring equipment, or arise because of phenotypes being imprecisely defined.
 The same model of measurement error can be applied to the outcome variable $y$.
@@
 The Steiger test assumes that there is a causal relationship between the two variables, and that the SNP is a valid instrument for one of them.
-However it is liable to give incorrect causal directions under some other circumstances also.
+However it is liable to give incorrect causal directions under some other circumstances.
 First, some levels of horizontal pleiotropy, where the SNP influences the outcome through some pathway other than the exposure, could induce problems because this is a means by which the instrument is invalid.
@@
 \begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_{gx} + \beta_{gx} g + \epsilon_{gx} \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
 y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
 \end{aligned}
@@
 Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Appendix 3).
-Though the relationship between unmeasured confounding and causal orientation is complex (`r sfig_nums("unmeasured")`), we note that when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction.
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (`r sfig_nums("unmeasured")`).
+Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
 
@@
 `r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", the MR analysis is indeed liable to infer the wrong direction of causality when $d < 0$, and that this erroneous result is more likely to occur with increasing sample size.
-However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality.
+However, the CIT is in general more fallible to reporting a robust causal association for the wrong direction of causality.
 When $d > 0$ we find that in most cases the MR Steiger method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
@@
 $$
-\hat{y}* = \hat{\beta}_x x
+\hat{y}^* = \hat{\beta}_x x
 $$
@@
 \begin{aligned}
-cov(g, y - \hat{y}*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
+cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
 & = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
@@
 
-`r csfig("unmeasured")` shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$.
-We note that in most cases where the observational absolute $R^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
+`r csfig("unmeasured")` shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
+The pattern of results were similar for different values of $\beta_{gx}$.
+We note that in most cases for the parameter values explored, where the observational $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
 

### 2017-07-24 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{18 July 2017}
+  \date{24 July 2017}
 
@@
 
-Inference of the causal structure that induces correlations between two
-traits can be achieved by combining genetic associations with a
+Inference about the causal structure that induces correlations between
+two traits can be achieved by combining genetic associations with a
 mediation-based approach, as is done in the causal inference test (CIT).
@@
 test the trait's causal influences on other traits and diseases. Here we
-demonstrate that when there is uncertainty about the nature of the
-genetic association existing methods that are used for causal inference
-can be liable to give the wrong direction of causality. We introduce a
-new method that is more flexible and robust to problems such as
-measurement error, and apply it to evaluate the causal relationships
-between DNA methylation levels and gene expression. Our results show
-that DNA methylation is more likely to be the causal factor, but that
-systematic differences in measurement error between the platforms could
-influence such conclusions.
+discuss scenarios where the nature of the genetic association with the
+causal trait can pose lead existing causal inference methods to give the
+wrong direction of causality. We introduce a new method that can be
+applied to summary level data and is potentially less susceptible to
+problems such as measurement error, and apply it to evaluate the causal
+relationships between DNA methylation levels and gene expression. While
+our results show that DNA methylation is more likely to be the causal
+factor, we point out that is it crucial to acknowledge that systematic
+differences in measurement error between the platforms could influence
+such conclusions.
 
@@
 mediation-based CIT method. Though we only examine the CIT method in
-detail, we show that attempting to adjust for mediating variables to
+detail, we believe that attempting to adjust for mediating variables to
 make causal inference is susceptible to problems, which can be
@@
 Measurement imprecision can represent imprecise measurement due to
-limits on sensitivity of measuring equipment, or due to phenotypes being
-imprecisely defined. The same model of measurement error can be applied
-to the outcome variable \(y\).
+limits on sensitivity of measuring equipment, or arise because of
+phenotypes being imprecisely defined. The same model of measurement
+error can be applied to the outcome variable \(y\).
 
@@
 However it is liable to give incorrect causal directions under some
-other circumstances also. First, some levels of horizontal pleiotropy,
-where the SNP influences the outcome through some pathway other than the
+other circumstances. First, some levels of horizontal pleiotropy, where
+the SNP influences the outcome through some pathway other than the
 exposure, could induce problems because this is a means by which the
@@
 Unmeasured confounding between the exposure and outcome can also give
-rise to problems with the MR Steiger approach (Appendix 3). Though the
+rise to problems with the MR Steiger approach (Appendix 3). The
 relationship between unmeasured confounding and causal orientation is
-complex (Supplementary Figure 1: Graph representing the unmeasured
-confounding parameters that will lead to the MR Steiger test returning
-the wrong causal direction. Columns of boxes represent different signed
-values of the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).), we note that when the magnitude of the observational
-variance explained between the exposure and the outcome is below 0.2 the
-MR Steiger method is unlikely to return the incorrect causal direction.
+complex across the parameter space of possible confounding values
+(Supplementary Figure 1: Graph representing the unmeasured confounding
+parameters that will lead to the MR Steiger test returning the wrong
+causal direction. Columns of boxes represent different signed values of
+the observational variance explained between \(x\) and \(y\)
+(\(R^2_{xy}\)).). Based on the range of parameter values that we
+explored, when the magnitude of the observational variance explained
+between the exposure and the outcome is below 0.2 the MR Steiger method
+is unlikely to return the incorrect causal direction due to unmeasured
+confounding.
 
@@
 and that this erroneous result is more likely to occur with increasing
-sample size. However, the CIT is in general more fallable to reporting a
+sample size. However, the CIT is in general more fallible to reporting a
 robust causal association for the wrong direction of causality. When
@@
 \[
-\hat{y}* = \hat{\beta}_x x
+\hat{y}^* = \hat{\beta}_x x
 \]
@@
 Supplementary Figure 1 shows the relationship between the magnitude of
-the correlations between \(x\), \(y\) and the confounder \(u\). We note
-that in most cases where the observational absolute \(R^2_{xy}\) is less
-than 0.2, unmeasured confounding will not incur the wrong causal
-direction in the MR Steiger test.
+the correlations between \(x\), \(y\) and the confounder \(u\) for a
+range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
+confounder effects. The pattern of results were similar for different
+values of \(\beta_{gx}\). We note that in most cases for the parameter
+values explored, where the observational \(\hat{R}^2_{xy}\) is less than
+0.2, unmeasured confounding will not incur the wrong causal direction in
+the MR Steiger test.
 

### 2017-07-27 scripts/mr_directionality_with_confounding.R
updated

@@ -127,10 +127,10 @@ make_phen <- function(effs, indep, vy=1, vx=rep(1, length(effs)))
 make_system <- function(n, p, r_ab, r_za, noisea, noiseb)
 {
 	Z <- make_geno(n, p)
-	U <- make_phen(r_za, Z)
-	eff <- sqrt(r_ab)
-	A <- make_phen(eff, U)
-	B <- make_phen(eff, U)
+	# U <- make_phen(r_za, Z)
+	eff <- sqrt(r_za)
+	A <- make_phen(eff, Z)
+	B <- make_phen(eff, Z)
 	Ap <- makeProxy(A, noisea, 1)
 	Bp <- makeProxy(B, noiseb, 1)
 	return(data.frame(Z, A, B, Ap, Bp))
@@ -167,7 +167,7 @@ parameters <- expand.grid(
 
 arguments <- commandArgs(T)
 jid <- "all"
-outdir <- "~/repo/cit_measurement_error/scratch/"
+outdir <- "~/repo/causal-directions/scratch/"
 if(length(arguments) > 0)
 {
 	jid <- as.numeric(arguments[1])

### 2017-07-27 scripts/mr_directionality_with_confounding.sh

@@ -21,6 +21,6 @@ i=${PBS_ARRAYID}
 splits=1944
 
 Rscript \
-	~/repo/cit_measurement_error/scripts/mr_directionality_with_confounding.R \
+	~/repo/causal-directions/scripts/mr_directionality_with_confounding.R \
 	${i} \
 	${splits}

### 2017-07-27 docs/manuscript.Rmd
submitted revision

@@
 To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases.
-Here we discuss scenarios where the nature of the genetic association with the causal trait can pose lead existing causal inference methods to give the wrong direction of causality.
+Here we discuss scenarios where the nature of the genetic association with the causal trait can lead existing causal inference methods to give the wrong direction of causality.
 We introduce a new method that can be applied to summary level data and is potentially less susceptible to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression.
@@
 Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016].
-Additionally, if valid genetic instruments are known for both traits of interest then MR can be performed in both directions (bi-directional MR), testing the influence of one trait on the other and vice versa, to infer the causal direction between the two phenotypes.
+Additionally, if valid genetic instruments are known for both traits of interest then MR can be performed in both directions (bi-directional MR), testing the influence of one trait on the other and vice versa, to infer the causal direction between the two phenotypes [@Timpson2011; @Richmond2014].
 
@@
 If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
-Though bi-directional MR can be used to orient causal directions, this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood.
+Though bi-directional MR can be used to orient causal directions [@Timpson2011], this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood.
 We go on to explore the scenarios in which the method is likely to return the correct or incorrect causal directions.
@@
 The FDR using a p-value threshold of 0.05 appears to be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
-Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases faster than that of the MR Steiger method.
+Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases more steeply than that of the MR Steiger method.
 
@@
 Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Appendix 3).
-The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (`r sfig_nums("unmeasured")`).
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (`r csfig("unmeasured")`).
 Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
@@
 The pattern of results were similar for different values of $\beta_{gx}$.
-We note that in most cases for the parameter values explored, where the observational $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
+We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
 

### 2017-07-27 docs/manuscript.tex
submitted revision

@@
   \postdate{\par}
-  \date{24 July 2017}
+  \date{27 July 2017}
 
@@
 discuss scenarios where the nature of the genetic association with the
-causal trait can pose lead existing causal inference methods to give the
+causal trait can lead existing causal inference methods to give the
 wrong direction of causality. We introduce a new method that can be
@@
 influence of one trait on the other and vice versa, to infer the causal
-direction between the two phenotypes.
+direction between the two phenotypes (27,28).
 
@@
 function, and it often takes years of follow up studies to understand
-the biological nature of a putative GWAS hit (27). If multiple
+the biological nature of a putative GWAS hit (29). If multiple
 instruments are available for an hypothesised exposure, which is
@@
 inference to be drawn when exposure variables and outcome variables are
-measured in different samples (28). This has the crucial advantage of
+measured in different samples (30). This has the crucial advantage of
 improving statistical power by allowing analysis in much larger sample
@@
 outcome by the exposure induces a collider effect between the SNP and
-outcome (29), and in order to fully abrogate this association one must
+outcome (31), and in order to fully abrogate this association one must
 also adjust for all (hidden or otherwise) confounders. MR does not
@@
 is robust to there being measurement error in the exposure variable
-(30). Indeed instrumental variable (IV) analysis was in part initially
-introduced as a correction for measurement error in the exposure (31),
+(32). Indeed instrumental variable (IV) analysis was in part initially
+introduced as a correction for measurement error in the exposure (33),
 whereas it has been noted that both classic mediation-based analyses
-(13,14,32,33) and mediation-based methods that use instrumental
-variables (34,35) are prone to be unreliable in its presence.
+(13,14,34,35) and mediation-based methods that use instrumental
+variables (36,37) are prone to be unreliable in its presence.
 
@@
 weaknesses, triangulation of evidence from as many sources as possible
-should be practiced in causal inference (36).
+should be practiced in causal inference (38).
 
@@
 the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
-For example, following Pierce and VanderWeele (30) we can define
+For example, following Pierce and VanderWeele (32) we can define
 
@@
 complete linkage disequilibrium with the causal variant) will reduce
-power in MR but will not incur bias (3,13,30). We also assume that
+power in MR but will not incur bias (3,13,32). We also assume that
 measurement error in the exposure and the outcome are uncorrelated.
@@
 not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds (37).
+is it reliable to depend on arbitrary significance thresholds (39).
 
@@
 obtained using the \(systemfit\) function in the R package
-\(R/systemfit\) (38). Note that the value of \(p_{MR}\) is identical
+\(R/systemfit\) (40). Note that the value of \(p_{MR}\) is identical
 when using the same genetic variant to instrument the influence of the
@@
 and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population (39). It is calculated as
+within a population (41). It is calculated as
 
@@
 different samples (e.g.~from publicly available summary statistics
-(26,28)). Here the Steiger test of two independent correlations can be
+(26,30)). Here the Steiger test of two independent correlations can be
 applied where.
@@
 are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} (40).
+\emph{r.test} function in the R package \emph{R/psych} (42).
 
@@
 
-Two sample MR (28) was performed using the summary statistics for
+Two sample MR (30) was performed using the summary statistics for
 genetic influences on gene expression and DNA methylation. To do this we
 obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al (41). These were filtered to be located on
+reported in Shakhbazov et al (43). These were filtered to be located on
 the same chromosome, have robust correlations after correcting for
@@
 
-All analysis was performed using the R programming language (42) and
+All analysis was performed using the R programming language (44) and
 code is made available at
@@
 or \(y\) then CIT can attempt to infer the causal direction. Though
-bi-directional MR can be used to orient causal directions, this requires
-knowledge of a valid instrument for each trait, and we were motivated to
-develop the MR Steiger method that could operate on summary data to
-orient the direction of causality using the same conditions as the CIT,
-where the underlying biology of a single SNP is not fully understood. We
-go on to explore the scenarios in which the method is likely to return
-the correct or incorrect causal directions.
+bi-directional MR can be used to orient causal directions (27), this
+requires knowledge of a valid instrument for each trait, and we were
+motivated to develop the MR Steiger method that could operate on summary
+data to orient the direction of causality using the same conditions as
+the CIT, where the underlying biology of a single SNP is not fully
+understood. We go on to explore the scenarios in which the method is
+likely to return the correct or incorrect causal directions.
 
@@
 non-null simulations, the general trend appears to be that the CIT power
-reduces as measurement error in the exposure increases faster than that
-of the MR Steiger method.
+reduces as measurement error in the exposure increases more steeply than
+that of the MR Steiger method.
 
@@
 complex across the parameter space of possible confounding values
-(Supplementary Figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the MR Steiger test returning the wrong
-causal direction. Columns of boxes represent different signed values of
-the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).). Based on the range of parameter values that we
+(Supplementary Figure 1). Based on the range of parameter values that we
 explored, when the magnitude of the observational variance explained
@@
 
-We performed two sample MR (28) for each association in the direction of
+We performed two sample MR (30) for each association in the direction of
 causality inferred by the Stieger test. We observed that the sign of the
 MR estimate was generally in the same direction as the Pearson
-correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
+correlation coefficient reported by Shakhbazov et al (43) (Figure 6b).
 There was a moderate correlation between the absolute magnitudes of the
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation (43).
+levels, consistent with expectation (45).
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis are relatively well explored (44). Our analysis extends this to
+analysis are relatively well explored (46). Our analysis extends this to
 related methods such as CIT that are used in predominantly 'omic data.
@@
 which have evaluated measurement error in these platforms do support
-this position (45,46), though making strong conclusions for this
+this position (47,48), though making strong conclusions for this
 analysis is difficult because measurement error is likely to be study
@@
 another trait, and \(g\) having pleiotropic effects on both traits
-independently (47). Under certain conditions of measurement error the
+independently (49). Under certain conditions of measurement error the
 CIT test can distinguish these models. We also note that it is
@@
 bi-directional MR can offer solutions to inferring the causal direction
-(16,48,49). We restricted the simulations to evaluating the causal
+(16,28,50). We restricted the simulations to evaluating the causal
 inference between quantitative traits, but it is possible that the
@@
 variance explained on the liability scale, taking into account the
-population prevalence (50). However, our analysis goes beyond many
+population prevalence (51). However, our analysis goes beyond many
 previous explorations of measurement error by assessing the impacts of
@@
 on the outcome. This problem has shown to be the case for SNPs
-influencing different lipid fractions, for example (51,52).
+influencing different lipid fractions, for example (52,53).
 
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established (35) and have a number of
+variables, are very well established (37) and have a number of
 extensions that make them valuable tools, including for example network
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics (53), multivariate MR (52) and
-mediation through MR (54--56) may offer more robust alternatives for
+applying MR to genetical genomics (54), multivariate MR (53) and
+mediation through MR (55--57) may offer more robust alternatives for
 these more complicated problems.
@@
 values of \(\beta_{gx}\). We note that in most cases for the parameter
-values explored, where the observational \(\hat{R}^2_{xy}\) is less than
-0.2, unmeasured confounding will not incur the wrong causal direction in
-the MR Steiger test.
+values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
+less than 0.2, unmeasured confounding will not incur the wrong causal
+direction in the MR Steiger test.
 
@@
 
+\hypertarget{ref-Timpson2011}{}
+27. Timpson NJ, Nordestgaard BG, Harbord RM, Zacho J, Frayling TM,
+Tybjærg-Hansen a, et al. C-reactive protein levels and body mass index:
+elucidating direction of causation through reciprocal Mendelian
+randomization. International journal of obesity (2005) {[}Internet{]}.
+2011 Feb;35(2):300--8. Available from:
+\url{http://www.ncbi.nlm.nih.gov/pubmed/20714329}
+
+\hypertarget{ref-Richmond2014}{}
+28. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
+NJ. Assessing Causality in the Association between Child Adiposity and
+Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
+editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
+from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}
+
 \hypertarget{ref-Claussnitzer2015}{}
-27. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
+29. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
 al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
@@
 \hypertarget{ref-Pierce2013}{}
-28. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+30. Pierce BL, Burgess S. Efficient design for Mendelian randomization
 studies: subsample and 2-sample instrumental variable estimators.
@@
 \hypertarget{ref-Hernan2004}{}
-29. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
+31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
 selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
@@
 \hypertarget{ref-Pierce2012}{}
-30. Pierce BL, VanderWeele TJ. The effect of non-differential
+32. Pierce BL, VanderWeele TJ. The effect of non-differential
 measurement error on bias, precision and power in Mendelian
@@
 \hypertarget{ref-Ashenfelter1994}{}
-31. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
+33. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
 Schooling from a New Sample of Twins. The American Economic Review.
@@
 \hypertarget{ref-Nagarajan2013}{}
-32. Nagarajan R, Scutari M. Impact of noise on molecular network
+34. Nagarajan R, Scutari M. Impact of noise on molecular network
 inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
@@
 \hypertarget{ref-Shpitser2010}{}
-33. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
+35. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
 adjustment for estimating causal effects. Proceedings of the Twenty
@@
 \hypertarget{ref-Wang2015}{}
-34. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
+36. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
 networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
@@
 \hypertarget{ref-Lagani2015}{}
-35. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
+37. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
 Probabilistic Computational Causal Discovery for Systems Biology. In:
@@
 \hypertarget{ref-Lawlor2017}{}
-36. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
+38. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
 epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
@@
 \hypertarget{ref-Sterne2001}{}
-37. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
 significance tests? BMJ. 2001;322(7280):226--31.
@@
 \hypertarget{ref-Henningsen2007}{}
-38. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
 Systems of Simultaneous Equations in R. Journal of Statistical Software
@@
 \hypertarget{ref-Steiger1980}{}
-39. Steiger JH. Tests for comparing elements of a correlation matrix.
+41. Steiger JH. Tests for comparing elements of a correlation matrix.
 Psychological Bulletin. 1980;87(2):245--51.
@@
 \hypertarget{ref-Revelle2015}{}
-40. Revelle W. psych: Procedures for Psychological, Psychometric, and
+42. Revelle W. psych: Procedures for Psychological, Psychometric, and
 Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
@@
 \hypertarget{ref-Shakhbazov2016}{}
-41. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+43. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
 PM, et al. Shared genetic control of expression and methylation in
@@
 \hypertarget{ref-RCoreTeam2015}{}
-42. R Core Team. R: A Language and Environment for Statistical Computing
+44. R Core Team. R: A Language and Environment for Statistical Computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-Bird2002}{}
-43. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
 development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
@@
 \hypertarget{ref-Cole2014}{}
-44. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+46. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
 Serious and Misleading Consequences Due to Uncorrected Measurement
@@
 \hypertarget{ref-Bose2014}{}
-45. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+47. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
 Evaluation of microarray-based DNA methylation measurement using
@@
 \hypertarget{ref-Bryant2011}{}
-46. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+48. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
 et al. Technical Variability Is Greater than Biological Variability in a
@@
 \hypertarget{ref-Zhu2016}{}
-47. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+49. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 
-\hypertarget{ref-Richmond2014}{}
-48. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
-NJ. Assessing Causality in the Association between Child Adiposity and
-Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
-editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
-from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}
-
 \hypertarget{ref-Mancuso2017}{}
-49. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+50. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
 Integrating Gene Expression with Summary Association Statistics to
@@
 \hypertarget{ref-Lee2013c}{}
-50. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+51. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Do2013}{}
-51. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+52. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
 Common variants associated with plasma triglycerides and risk for
@@
 \hypertarget{ref-Burgess2014a}{}
-52. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+53. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Relton2012}{}
-53. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+54. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Varbo2015}{}
-54. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+55. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 \hypertarget{ref-Burgess2015}{}
-55. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+56. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 \hypertarget{ref-Richmond2016}{}
-56. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+57. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.

### 2017-10-05 docs/manuscript.Rmd
split revision into supplement and main texts

@@
 c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
-	display=FALSE
-)
-
-sfig_nums(
-	"unmeasured",
-	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
-Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$)."),
 	display=FALSE
@@
 To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$.
-This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
+This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Supplementary text 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
 It is calculated as
@@
 First, some levels of horizontal pleiotropy, where the SNP influences the outcome through some pathway other than the exposure, could induce problems because this is a means by which the instrument is invalid.
-Second, some differential values of measurement error between the exposure and the outcome could induce incorrect causal directions, and this is explored in Appendix 2.
-Third, some levels of unmeasured confounding between the exposure and the outcome could induce the wrong causal direction, and this is explored in Appendix 3.
+Second, some differential values of measurement error between the exposure and the outcome could induce incorrect causal directions, and this is explored in Supplementary text 2.
+Third, some levels of unmeasured confounding between the exposure and the outcome could induce the wrong causal direction, and this is explored in Supplementary text 3.
 
@@
 
-As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
+As we show in Supplementary text 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
 The maximum measurement imprecision of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$.
@@
 In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction.
-Full details are provided in Appendix 2.
+Full details are provided in Supplementary text 2.
 
@@
 In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Supplementary text 1) that
 
@@
 
-For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2).
+For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Supplementary text 2).
 This metric can be used to evaluate the reliability of MR Steiger test.
 
-We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Supplementary text 2) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
 `r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
@@
 
-Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Appendix 3).
-The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (`r csfig("unmeasured")`).
+Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Supplementary text 3).
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 1).
 Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
@@
 
-## Acknowledgements
-
-This work was funded by the UK Medical Research Council (MC_UU_12013/1 and MC_UU_12013/9).
-
 \newpage
@@
 
-## Appendix 1
-
-We assume the following model
-
-$$
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-$$
-
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_o$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_o$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
-Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y_o - \hat{y}_o) = 0$ when $x$ causes $y$, where $\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o$ is the predicted value of $y_o$ using the measured value of $x_o$.
-
-We can split $cov(g, y_o - \hat{y}_o)$ into two parts, $cov(g, y_o)$ and $cov(g, \hat{y}_o)$.
-
-**Part 1**
-
-$$
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-**Part 2**
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-$$
-
-Simpifying further
-
-$$
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-$$
-
-which can be substituted back to give
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-Finally
-
-$$
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.
-
-## Appendix 2
-
-Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude.
-The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:
-
-$$
-x = \alpha_1 + \beta_1 g + e_1
-$$
-
-for the first stage and
-
-$$
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-$$
-
-where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$.
-Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
-
-In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
-
-$$
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-$$
-
-In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.
-
-For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$.
-For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.
-
-Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$.
-If $z < 0$ then we infer that $x \rightarrow y$.
-There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$.
-To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:
-
-$$
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-$$
-
-If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values.
-If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality.
-$V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:
-
-$$
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-$$
-
-$V_{z \ge 0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
-The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$.
-Hence,
-
-$$
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-$$
-
-Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.
-
 \newpage
 
-## Appendix 3
+## Supplementary information
 
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on mediation-based methods for inferring causal directions, and can also adversely influence the estimate of the causal direction for the Steiger test.
+**Supplementary text 1** The influence of measurement error in the exposure on mediation-based estimated
 
-### Unmeasured confounding in mediation
+**Supplementary text 2** Sensitivity analysis for measurement error on the MR Steiger test
 
-Including an unmeasured confounder, $u$, after ignoring intercept terms the exposure $x$ and outcome $y$ variables can be modelled as
+**Supplementary text 3** The influence of unmeasured confounding on the inference of causal directions
 
-$$
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-$$
-
-The observational estimate of the causal effect of $x$ on $y$, $\hat{\beta}_x$ is obtained from
-
-$$
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-$$
-
-From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are non-zero.
-Relating to mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
-
-$$
-\hat{y}^* = \hat{\beta}_x x
-$$
-
-and
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-$$
-
-should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and $y|x$, which will introduce errors in inferring causal directions.
-
-### Unmeasured confounding in the MR Steiger test
-
-Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the MR Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
-After some algebra
-
-$$
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-$$
-
-and
-
-$$
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-$$
-
-`r csfig("unmeasured")` shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
-The pattern of results were similar for different values of $\beta_{gx}$.
-We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
-
-![](../images/unmeasured_conf_steiger.png)
-
-`r sfig_nums("unmeasured")`
+**Supplementary figure 1** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
+Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).
 

### 2017-10-05 docs/manuscript.tex
split revision into supplement and main texts

@@
   \postdate{\par}
-  \date{27 July 2017}
+  \date{28 July 2017}
 
@@
 instrument \(g\). This can be achieved by assessing which of the two
-variables has the biggest absolute correlation with \(g\) (Appendix 2),
-formalised by testing for a difference in the correlations \(\rho_{gx}\)
-and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population (41). It is calculated as
+variables has the biggest absolute correlation with \(g\) (Supplementary
+text 2), formalised by testing for a difference in the correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) using Steiger's Z-test for correlated
+correlations within a population (41). It is calculated as
 
@@
 error between the exposure and the outcome could induce incorrect causal
-directions, and this is explored in Appendix 2. Third, some levels of
-unmeasured confounding between the exposure and the outcome could induce
-the wrong causal direction, and this is explored in Appendix 3.
+directions, and this is explored in Supplementary text 2. Third, some
+levels of unmeasured confounding between the exposure and the outcome
+could induce the wrong causal direction, and this is explored in
+Supplementary text 3.
 
@@
 
-As we show in Appendix 2 it is possible to infer the bounds of
+As we show in Supplementary text 2 it is possible to infer the bounds of
 measurement error on \(x_o\) or \(y_o\) given known genetic
@@
 times more likely to be the empirical result than the opposite
-direction. Full details are provided in Appendix 2.
+direction. Full details are provided in Supplementary text 2.
 
@@
 measured value of \(y\), it can be shown using basic covariance
-properties (Appendix 1) that
+properties (Supplementary text 1) that
 
@@
 that agree or disagree with the empirically inferred causal direction
-(Figure 4a, Appendix 2). This metric can be used to evaluate the
-reliability of MR Steiger test.
+(Figure 4a, Supplementary text 2). This metric can be used to evaluate
+the reliability of MR Steiger test.
 
 We show that in the presence of measurement imprecision,
-\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
-the range of parameters around which the MR Steiger test is liable to
-provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
-the MR Steiger test is likely to be correct about the causal direction).
-Figure 4b shows that when there is no measurement error in \(x\), the MR
-Steiger test is unlikely to infer the wrong direction of causality even
-if there is measurement error in \(y\). It also shows that in most cases
-where \(x\) is measured with error, especially when the causal effect
-between \(x\) and \(y\) is not very large, the sensitivity of the MR
-Steiger test to measurement error is relatively low.
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Supplementary text 2)
+determines the range of parameters around which the MR Steiger test is
+liable to provide the wrong direction of causality (\emph{i.e.} if
+\(d>0\) then the MR Steiger test is likely to be correct about the
+causal direction). Figure 4b shows that when there is no measurement
+error in \(x\), the MR Steiger test is unlikely to infer the wrong
+direction of causality even if there is measurement error in \(y\). It
+also shows that in most cases where \(x\) is measured with error,
+especially when the causal effect between \(x\) and \(y\) is not very
+large, the sensitivity of the MR Steiger test to measurement error is
+relatively low.
 
 Unmeasured confounding between the exposure and outcome can also give
-rise to problems with the MR Steiger approach (Appendix 3). The
-relationship between unmeasured confounding and causal orientation is
-complex across the parameter space of possible confounding values
-(Supplementary Figure 1). Based on the range of parameter values that we
+rise to problems with the MR Steiger approach (Supplementary text 3).
+The relationship between unmeasured confounding and causal orientation
+is complex across the parameter space of possible confounding values
+(Supplementary figure 1). Based on the range of parameter values that we
 explored, when the magnitude of the observational variance explained
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
-
-This work was funded by the UK Medical Research Council (MC\_UU\_12013/1
-and MC\_UU\_12013/9).
-
 \newpage
@@
 
-\subsection{Appendix 1}\label{appendix-1}
-
-We assume the following model
-
-\[
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-\]
-
-where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), \(x_o\) is the measured quantity of
-\(x\), where measurement error is incurred from linear transformation in
-\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
-\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
-measurement error is incurred from linear transformation in
-\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
-\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
-of association between \(g\) and \(y\) after conditioning on \(x\).
-Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
-when \(x\) causes \(y\), where
-\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
-value of \(y_o\) using the measured value of \(x_o\).
-
-We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
-and \(cov(g, \hat{y}_o)\).
-
-\textbf{Part 1}
-
-\[
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-\textbf{Part 2}
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-\]
-
-Simpifying further
-
-\[
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-\]
-
-which can be substituted back to give
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-Finally
-
-\[
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
-\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
-then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.
-
-\subsection{Appendix 2}\label{appendix-2}
-
-Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
-causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
-and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
-hypothesis test that provides a p-value for observing the difference in
-these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is \(x \to y\), two stage MR is formulated
-using the following regression models:
-
-\[
-x = \alpha_1 + \beta_1 g + e_1
-\]
-
-for the first stage and
-
-\[
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-\]
-
-where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
-free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
-the exposure variable \(x\), and it is expected that
-\(\rho_{g, x} > \rho_{g, y}\) because
-\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
-causal association between \(x\) and \(y\) (which is likely to be less
-than 1).
-
-In the presence of measurement error in \(x\) and \(y\), however, the
-empirical inference of the causal direction will instead be based on
-evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
-
-\[
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-\]
-
-In order to assess how reliable the inference of the causal direction is
-in the presence of measurement imprecision, we can evaluate the range of
-potential values of measurement error in \(x\) and \(y\) over which the
-empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
-return the wrong causal direction.
-
-For different values of \(\rho_{x,x_o}\),
-\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
-\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
-\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
-and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).
-
-Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
-explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
-we infer that \(x \rightarrow y\). There will be some values of
-\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
-\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
-causal direction with regards to measurement error, the objective is to
-compare the proportion of the parameter space that agrees with the
-inferred direction against the proportion which does not:
-
-\[
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-\]
-
-If \(R=1\) then the direction of causality is equally probable across
-the range of possible measurement error values. If \(R > 1\) then \(R\)
-times as much of the parameter space favours the inferred direction of
-causality. \(V_{z}\), the total volume of the function (Figure 4), can
-be obtained analytically by solving:
-
-\[
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-\]
-
-\(V_{z \ge 0}\), the proportion of the volume that lies above the
-\(z=0\) plane, can also be obtained analytically. The region of this
-volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
-where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
-\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,
-
-\[
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-\]
-
-Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).
-
 \newpage
 
-\subsection{Appendix 3}\label{appendix-3}
-
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on
-mediation-based methods for inferring causal directions, and can also
-adversely influence the estimate of the causal direction for the Steiger
-test.
-
-\subsubsection{Unmeasured confounding in
-mediation}\label{unmeasured-confounding-in-mediation}
-
-Including an unmeasured confounder, \(u\), after ignoring intercept
-terms the exposure \(x\) and outcome \(y\) variables can be modelled as
-
-\[
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-\]
-
-The observational estimate of the causal effect of \(x\) on \(y\),
-\(\hat{\beta}_x\) is obtained from
-
-\[
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-\]
-
-From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
-when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
-mediation, where we attempt to test if \(g\) associates with \(y\) after
-adjusting \(y\) for \(x\), such that
-
-\[
-\hat{y}^* = \hat{\beta}_x x
-\]
-
-and
-
-\[
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-\]
-
-should any amount of unmeasured confounding exist, therefore, there will
-remain an association between \(g\) and \(y|x\), which will introduce
-errors in inferring causal directions.
-
-\subsubsection{Unmeasured confounding in the MR Steiger
-test}\label{unmeasured-confounding-in-the-mr-steiger-test}
-
-Similarly, we can investigate the extent to which unmeasured confounding
-will lead to the wrong causal direction between \(x\) and \(y\) using
-the MR Steiger test, evaluating the liability for the inequality
-\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
-
-\[
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-\]
-
-and
-
-\[
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-\]
-
-Supplementary Figure 1 shows the relationship between the magnitude of
-the correlations between \(x\), \(y\) and the confounder \(u\) for a
-range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
-confounder effects. The pattern of results were similar for different
-values of \(\beta_{gx}\). We note that in most cases for the parameter
-values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
-less than 0.2, unmeasured confounding will not incur the wrong causal
-direction in the MR Steiger test.
-
-\includegraphics{../images/unmeasured_conf_steiger.png}
-
-Supplementary Figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the MR Steiger test returning the wrong
-causal direction. Columns of boxes represent different signed values of
-the observational variance explained between \(x\) and \(y\)
+\subsection{Supplementary information}\label{supplementary-information}
+
+\textbf{Supplementary text 1} The influence of measurement error in the
+exposure on mediation-based estimated
+
+\textbf{Supplementary text 2} Sensitivity analysis for measurement error
+on the MR Steiger test
+
+\textbf{Supplementary text 3} The influence of unmeasured confounding on
+the inference of causal directions
+
+\textbf{Supplementary figure 1} Graph representing the unmeasured
+confounding parameters that will lead to the MR Steiger test returning
+the wrong causal direction. Columns of boxes represent different signed
+values of the observational variance explained between \(x\) and \(y\)
 (\(R^2_{xy}\)).

### 2017-10-05 docs/supplementary.Rmd
split revision into supplement and main texts

---
title: "Supplementary materials: Orienting the causal relationship between imprecisely measured traits using GWAS summary data"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
---

## Supplementary text 1

We assume the following model

$$
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
$$

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_o$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_o$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$. Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$. Under the CIT, this is expected to be $cov(g, y_o - \hat{y}_o) = 0$ when $x$ causes $y$, where $\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o$ is the predicted value of $y_o$ using the measured value of $x_o$.

We can split $cov(g, y_o - \hat{y}_o)$ into two parts, $cov(g, y_o)$ and $cov(g, \hat{y}_o)$.

**Part 1**

$$
\begin{aligned}
cov(g, y_o) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$

**Part 2**

$$
\begin{aligned}
cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
\end{aligned}
$$

Simpifying further

$$
\begin{aligned}
\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
$$

which can be substituted back to give

$$
\begin{aligned}
cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$

Finally

$$
\begin{aligned}
cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$

thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.

\newpage

## Supplementary text 2

Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude. The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.

Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:

$$
x = \alpha_1 + \beta_1 g + e_1
$$

for the first stage and

$$
y = \alpha_2 + \beta_2 \hat{x} + e_2
$$

where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$. Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).

In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:

$$
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
$$

In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.

For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$. For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.

Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$. If $z < 0$ then we infer that $x \rightarrow y$. There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$. To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:

$$
R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
$$

If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values. If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality. $V_{z}$, the total volume of the function (Supplementary figure 4), can be obtained analytically by solving:

$$
\begin{aligned}
V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
\end{aligned}
$$

$V_{z \ge 0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically. The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$. Hence,

$$
\begin{aligned}
V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
$$

Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.

\newpage

## Supplementary text 3

We have assumed no unmeasured confounding in these simulations. Unmeasured confounding will however have potentially large influences on mediation-based methods for inferring causal directions, and can also adversely influence the estimate of the causal direction for the Steiger test.

### Unmeasured confounding in mediation

Including an unmeasured confounder, $u$, after ignoring intercept terms the exposure $x$ and outcome $y$ variables can be modelled as

$$
\begin{aligned}
y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
x & = \beta_g g + \beta_{ux} u + \epsilon_g
\end{aligned}
$$

The observational estimate of the causal effect of $x$ on $y$, $\hat{\beta}_x$ is obtained from

$$
\begin{aligned}
\hat{\beta}_x & = cov(x, y) / var(x) \\
& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
\end{aligned}
$$

From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are non-zero. Relating to mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that

$$
\hat{y}^* = \hat{\beta}_x x
$$

and

$$
\begin{aligned}
cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
& = (\beta_x - \hat{\beta}_x) var(g)
\end{aligned}
$$

should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and $y|x$, which will introduce errors in inferring causal directions.

### Unmeasured confounding in the MR Steiger test

Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the MR Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect. After some algebra

$$
cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
$$

and

$$
cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
$$

Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects. The pattern of results were similar for different values of $\beta_{gx}$. We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.

\newpage

![](../images/unmeasured_conf_steiger.png)

Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-05 docs/supplementary.tex
split revision into supplement and main texts

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Supplementary materials: Orienting the causal relationship between imprecisely measured traits using GWAS summary data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Supplementary materials: Orienting the causal relationship between
imprecisely measured traits using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 July 2017}

\begin{document}
\maketitle

\subsection{Supplementary text 1}\label{supplementary-text-1}

We assume the following model

\[
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), \(x_o\) is the measured quantity of
\(x\), where measurement error is incurred from linear transformation in
\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
measurement error is incurred from linear transformation in
\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
of association between \(g\) and \(y\) after conditioning on \(x\).
Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
when \(x\) causes \(y\), where
\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
value of \(y_o\) using the measured value of \(x_o\).

We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
and \(cov(g, \hat{y}_o)\).

\textbf{Part 1}

\[
\begin{aligned}
cov(g, y_o) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

\textbf{Part 2}

\[
\begin{aligned}
cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
\end{aligned}
\]

Simpifying further

\[
\begin{aligned}
\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
\]

which can be substituted back to give

\[
\begin{aligned}
cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

Finally

\[
\begin{aligned}
cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.

\newpage

\subsection{Supplementary text 2}\label{supplementary-text-2}

Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
hypothesis test that provides a p-value for observing the difference in
these correlations under the null hypothesis that they are equal.

Assuming the causal direction is \(x \to y\), two stage MR is formulated
using the following regression models:

\[
x = \alpha_1 + \beta_1 g + e_1
\]

for the first stage and

\[
y = \alpha_2 + \beta_2 \hat{x} + e_2
\]

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1).

In the presence of measurement error in \(x\) and \(y\), however, the
empirical inference of the causal direction will instead be based on
evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
\]

In order to assess how reliable the inference of the causal direction is
in the presence of measurement imprecision, we can evaluate the range of
potential values of measurement error in \(x\) and \(y\) over which the
empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
return the wrong causal direction.

For different values of \(\rho_{x,x_o}\),
\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).

Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
we infer that \(x \rightarrow y\). There will be some values of
\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
causal direction with regards to measurement error, the objective is to
compare the proportion of the parameter space that agrees with the
inferred direction against the proportion which does not:

\[
R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
\]

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the parameter space favours the inferred direction of
causality. \(V_{z}\), the total volume of the function (Supplementary
figure 4), can be obtained analytically by solving:

\[
\begin{aligned}
V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
\end{aligned}
\]

\(V_{z \ge 0}\), the proportion of the volume that lies above the
\(z=0\) plane, can also be obtained analytically. The region of this
volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,

\[
\begin{aligned}
V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
\]

Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).

\newpage

\subsection{Supplementary text 3}\label{supplementary-text-3}

We have assumed no unmeasured confounding in these simulations.
Unmeasured confounding will however have potentially large influences on
mediation-based methods for inferring causal directions, and can also
adversely influence the estimate of the causal direction for the Steiger
test.

\subsubsection{Unmeasured confounding in
mediation}\label{unmeasured-confounding-in-mediation}

Including an unmeasured confounder, \(u\), after ignoring intercept
terms the exposure \(x\) and outcome \(y\) variables can be modelled as

\[
\begin{aligned}
y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
x & = \beta_g g + \beta_{ux} u + \epsilon_g
\end{aligned}
\]

The observational estimate of the causal effect of \(x\) on \(y\),
\(\hat{\beta}_x\) is obtained from

\[
\begin{aligned}
\hat{\beta}_x & = cov(x, y) / var(x) \\
& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
\end{aligned}
\]

From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
mediation, where we attempt to test if \(g\) associates with \(y\) after
adjusting \(y\) for \(x\), such that

\[
\hat{y}^* = \hat{\beta}_x x
\]

and

\[
\begin{aligned}
cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
& = (\beta_x - \hat{\beta}_x) var(g)
\end{aligned}
\]

should any amount of unmeasured confounding exist, therefore, there will
remain an association between \(g\) and \(y|x\), which will introduce
errors in inferring causal directions.

\subsubsection{Unmeasured confounding in the MR Steiger
test}\label{unmeasured-confounding-in-the-mr-steiger-test}

Similarly, we can investigate the extent to which unmeasured confounding
will lead to the wrong causal direction between \(x\) and \(y\) using
the MR Steiger test, evaluating the liability for the inequality
\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra

\[
cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
\]

and

\[
cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
\]

Supplementary figure 1 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
less than 0.2, unmeasured confounding will not incur the wrong causal
direction in the MR Steiger test.

\newpage

\includegraphics{../images/unmeasured_conf_steiger.png}

Supplementary figure 1: Graph representing the unmeasured confounding
parameters that will lead to the MR Steiger test returning the wrong
causal direction. Columns of boxes represent different signed values of
the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).

\end{document}

### 2017-10-05 docs/manuscript.Rmd
updating for revision

@@
 	"causality_exists_tpr",
-	caption="Outcomes were simulated to be unrelated to the exposure (bottom plot) or causally influenced by the exposure (top plot) with varying degrees of measurement imprecision applied to the exposure variable (x axis).
-True positive rates (y axis) for MR and CIT were compared for varying sample sizes (columns of boxes).",
+	caption="Outcomes were simulated to be unrelated to the exposure (bottom plot, showing false positive rates on the y-axis) or causally influenced by the exposure (top plot, showing true positive rates on the y-axis) with varying degrees of measurement imprecision applied to the exposure variable (x axis).
+Results for MR and CIT were compared for varying sample sizes (columns of boxes).",
 	display=FALSE
@@
 	caption="a) We can predict the values the MR Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface.
-When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed MR Steiger test inferring the wrong causal direction.
-Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y.
-A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$.
+When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the black plane, those values of measurement error lead to our observed MR Steiger test inferring the wrong causal direction.
+Where the blue surface lies below the black plane, these measurement error values support the inferred causal direction of X to Y.
+A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and black surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$.
 In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$.
@@
 We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}1$) the CIT is reliable in identifying the correct causal direction.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}=1$) the CIT is reliable in identifying the correct causal direction.
 However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.

### 2017-10-05 scripts/cit_confounder.r
checking for influence of confounding on cit

@@ -96,3 +96,58 @@ facet_grid(n ~ .) +
 labs(x="Confounder effect", colour="Test")
 
 ggsave("../images/cit_confounder.pdf")
+
+
+
+#####
+
+library(cit)
+library(dplyr)
+library(ggplot2)
+
+nid <- 5000
+
+eff <- c(0.1, 0.25, 0.5, 1, 1.5, 2, 3, 4, 5)
+eff <- expand.grid(conf=c(eff, -eff, 0), geff=c(eff, -eff, 0))
+dim(eff)
+l <- list()
+for(i in 1:nrow(eff))
+{
+	message(i)
+	g <- rnorm(nid)
+	u <- rnorm(nid)
+	a <- g * eff$geff[i] + u * eff$conf[i] + rnorm(nid)
+	b <- g * eff$geff[i] + u * eff$conf[i] + rnorm(nid)
+	x <- as.data.frame(rbind(
+		cit.cp(L=g, G=a, T=b),
+		cit.cp(L=g, G=b, T=a)
+	))
+	x$sim <- i
+	x$dir <- c("AB", "BA")
+	l[[i]] <- x
+}
+
+
+
+l1 <- bind_rows(l)
+l1$conf <- rep(eff$conf, each=2)
+l1$geff <- rep(eff$geff, each=2)
+
+l2 <- group_by(l1, sim) %>%
+	summarise(nsig = sum(p_cit < 0.05), conf=conf[1], geff=geff[1])
+
+save(l1, l2, file="../results/cit_confounder.rdata")
+
+ggplot(l1, aes(x=conf, y=geff)) +
+geom_point(aes(colour=p_cit < 0.05)) +
+facet_grid(. ~ dir)
+
+
+l2$nsig <- as.factor(l2$nsig)
+levels(l2$nsig) <- c("None", "One direction", "Both directions")
+
+ggplot(l2, aes(x=conf, y=geff)) +
+geom_point(aes(colour=as.factor(nsig)), size=4) +
+scale_colour_brewer(type="qual") +
+labs(x="Confounder effect", y="Genetic effect", colour="CIT significance")
+

### 2017-10-10 docs/manuscript.Rmd
updated

@@
 
-1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
-2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
-3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
-4. $H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*
+1. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
+2. $H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0$; *the SNP associates with the exposure conditional on the outcome*
+3. $H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0$; *the exposure associates with the outcome conditional on the SNP*
+4. $H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0$; *the SNP is independent of the outcome conditional on the exposure*
 
-where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome.
-The 4th condition is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate against from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]).
+The term in the 4th test can be rewritten as a $cov(g, y|x) = cov(g, y - \hat{y})$ where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, and $x$ is assumed to mediate the association between the SNP and the outcome.
+The condition in the 4th test is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]).
 We note here that this approach is liable to fail, even when there is a true causal relationship, when confounders of the exposure and outcome are present, as these will induce collider bias.
@@
 - If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
-- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is likely confounding and no call is made ($x \leftarrow u \rightarrow y$).
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is potentially confounding (Supplementary figure 1) and no call is made.
 
@@
 Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Supplementary text 3).
-The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 1).
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 2).
 Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
@@
 
-**Supplementary figure 1** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
+**Supplementary figure 1** Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
+Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype.
+Each point represents a single simulation.
+Where power is high (when the absolute values of the $x$ and $y$ axes are large) the CIT returns a significant result ($p < 0.01$) when testing the causal effect of $x$ on $y$, and when testing the causal effect of $y$ on $x$.
+
+**Supplementary figure 2** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
 Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-10 docs/manuscript.tex
updated

@@
   \postdate{\par}
-  \date{28 July 2017}
+  \date{10 October 2017}
 
@@
 \item
-  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
-  associates with the exposure}
-\item
   \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
@@
 \item
-  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
-  associates with the outcome}
+  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
+  associates with the exposure conditional on the outcome}
 \item
-  \(H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0\);
-  \emph{the SNP is independent of the outcome when the outcome is
-  adjusted for the exposure}
+  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
+  associates with the outcome conditional on the SNP}
+\item
+  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
+  independent of the outcome conditional on the exposure}
 \end{enumerate}
 
-where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
-residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
-mediate the association between the SNP and the outcome. The 4th
-condition is formulated as an equivalence testing problem that is
-estimated using simulations, comparing the estimate against from the
-data against empirically obtained estimates for simulated variables
-where the independence model is true (full details are given in (4)). We
-note here that this approach is liable to fail, even when there is a
-true causal relationship, when confounders of the exposure and outcome
-are present, as these will induce collider bias.
+The term in the 4th test can be rewritten as a
+\(cov(g, y|x) = cov(g, y - \hat{y})\) where
+\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
+of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
+association between the SNP and the outcome. The condition in the 4th
+test is formulated as an equivalence testing problem that is estimated
+using simulations, comparing the estimate from the data against
+empirically obtained estimates for simulated variables where the
+independence model is true (full details are given in (4)). We note here
+that this approach is liable to fail, even when there is a true causal
+relationship, when confounders of the exposure and outcome are present,
+as these will induce collider bias.
 
@@
   If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then there is likely confounding
-  and no call is made (\(x \leftarrow u \rightarrow y\)).
+  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
+  confounding (Supplementary figure 1) and no call is made.
 \end{itemize}
@@
 the CIT method. Figure 2 shows that when there is no measurement error
-in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}1\))
+in the exposure or outcome variables (\(\rho_{x, x_o}=\rho_{y, y_o}=1\))
 the CIT is reliable in identifying the correct causal direction.
@@
 is complex across the parameter space of possible confounding values
-(Supplementary figure 1). Based on the range of parameter values that we
+(Supplementary figure 2). Based on the range of parameter values that we
 explored, when the magnitude of the observational variance explained
@@
 Figure 3: Outcomes were simulated to be unrelated to the exposure
-(bottom plot) or causally influenced by the exposure (top plot) with
-varying degrees of measurement imprecision applied to the exposure
-variable (x axis). True positive rates (y axis) for MR and CIT were
-compared for varying sample sizes (columns of boxes).
+(bottom plot, showing false positive rates on the y-axis) or causally
+influenced by the exposure (top plot, showing true positive rates on the
+y-axis) with varying degrees of measurement imprecision applied to the
+exposure variable (x axis). Results for MR and CIT were compared for
+varying sample sizes (columns of boxes).
 
@@
 as denoted by the range of values where the blue surface is above the
-red plane, those values of measurement error lead to our observed MR
+black plane, those values of measurement error lead to our observed MR
 Steiger test inferring the wrong causal direction. Where the blue
-surface lies below the red plane, these measurement error values support
-the inferred causal direction of X to Y. A measure of reliability,
-therefore, is the ratio of the negative and positive volumes of the
-total space bound by the blue and red surfaces,
+surface lies below the black plane, these measurement error values
+support the inferred causal direction of X to Y. A measure of
+reliability, therefore, is the ratio of the negative and positive
+volumes of the total space bound by the blue and black surfaces,
 \(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\). In this case, where
@@
 
-\textbf{Supplementary figure 1} Graph representing the unmeasured
+\textbf{Supplementary figure 1} Illustrative simulations (\(n = 5000\))
+showing the results from CIT analysis under a model of confounding.
+Here, the phenotypes \(x\) and \(y\) are not causally related, but there
+is a genetic effect and a confounder both influencing each phenotype.
+Each point represents a single simulation. Where power is high (when the
+absolute values of the \(x\) and \(y\) axes are large) the CIT returns a
+significant result (\(p < 0.01\)) when testing the causal effect of
+\(x\) on \(y\), and when testing the causal effect of \(y\) on \(x\).
+
+\textbf{Supplementary figure 2} Graph representing the unmeasured
 confounding parameters that will lead to the MR Steiger test returning

### 2017-10-10 docs/supplementary.Rmd
updated

@@
 If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality.
-$V_{z}$, the total volume of the function (Supplementary figure 4), can be obtained analytically by solving:
+$V_{z}$, the total volume of the function (Figure 4), can be obtained analytically by solving:
 
@@
 
-Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
+Supplementary figure 2 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
 The pattern of results were similar for different values of $\beta_{gx}$.
@@
 
+![](../images/cit_confounder2.pdf)
+
+Supplementary figure 1: Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
+Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype.
+Each point represents a single simulation.
+Where power is high (when the absolute values of the $x$ and $y$ axes are large) the CIT returns a significant result ($p < 0.01$) when testing the causal effect of $x$ on $y$, and when testing the causal effect of $y$ on $x$.
+
+\newpage
+
 ![](../images/unmeasured_conf_steiger.png)
 
-Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
+Supplementary figure 2: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
 Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-10 docs/supplementary.tex
updated

@@
 times as much of the parameter space favours the inferred direction of
-causality. \(V_{z}\), the total volume of the function (Supplementary
-figure 4), can be obtained analytically by solving:
+causality. \(V_{z}\), the total volume of the function (Figure 4), can
+be obtained analytically by solving:
 
@@
 
-Supplementary figure 1 shows the relationship between the magnitude of
+Supplementary figure 2 shows the relationship between the magnitude of
 the correlations between \(x\), \(y\) and the confounder \(u\) for a
@@
 
+\includegraphics{../images/cit_confounder.pdf}
+
+Supplementary figure 1: Illustrative simulations (\(n = 5000\)) showing
+the results from CIT analysis under a model of confounding. Here, the
+phenotypes \(x\) and \(y\) are not causally related, but there is a
+genetic effect and a confounder both influencing each phenotype. Each
+point represents a single simulation. Where power is high (when the
+absolute values of the \(x\) and \(y\) axes are large) the CIT returns a
+significant result (\(p < 0.01\)) when testing the causal effect of
+\(x\) on \(y\), and when testing the causal effect of \(y\) on \(x\).
+
+\newpage
+
 \includegraphics{../images/unmeasured_conf_steiger.png}
 
-Supplementary figure 1: Graph representing the unmeasured confounding
+Supplementary figure 2: Graph representing the unmeasured confounding
 parameters that will lead to the MR Steiger test returning the wrong

### 2017-10-10 scripts/cit_confounder.r

@@ -138,6 +138,8 @@ l2 <- group_by(l1, sim) %>%
 
 save(l1, l2, file="../results/cit_confounder.rdata")
 
+
+load("../results/cit_confounder.rdata")
 ggplot(l1, aes(x=conf, y=geff)) +
 geom_point(aes(colour=p_cit < 0.05)) +
 facet_grid(. ~ dir)
@@ -150,4 +152,4 @@ ggplot(l2, aes(x=conf, y=geff)) +
 geom_point(aes(colour=as.factor(nsig)), size=4) +
 scale_colour_brewer(type="qual") +
 labs(x="Confounder effect", y="Genetic effect", colour="CIT significance")
-
+ggsave("../images/cit_confounder2.pdf")

### 2017-10-24 docs/manuscript.Rmd
updated, preparing for proofs

@@
 bibliography: manuscript.bib
-csl: vancouver.csl
+csl: plos.csl
 ---
@@
 - If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
-- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is potentially confounding (Supplementary figure 1) and no call is made.
+- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is potentially confounding (S1 fig) and no call is made.
 
@@
 To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$.
-This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Supplementary text 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
+This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (S2 Text), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980].
 It is calculated as
@@
 First, some levels of horizontal pleiotropy, where the SNP influences the outcome through some pathway other than the exposure, could induce problems because this is a means by which the instrument is invalid.
-Second, some differential values of measurement error between the exposure and the outcome could induce incorrect causal directions, and this is explored in Supplementary text 2.
-Third, some levels of unmeasured confounding between the exposure and the outcome could induce the wrong causal direction, and this is explored in Supplementary text 3.
+Second, some differential values of measurement error between the exposure and the outcome could lead to incorrect inference of the causal direction (S2 Text).
+Third, some levels of unmeasured confounding between the exposure and the outcome could lead to inference of the wrong causal direction (S3 Text).
 
@@
 
-As we show in Supplementary text 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
+As we show in S2 Text it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations.
 The maximum measurement imprecision of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$.
@@
 Across all possible values of measurement error in $x$ and $y$ we find the volume that agrees with the inferred direction of causality and the volume that disagrees with the inferred direction of causality, and take the ratio of these two values.
-A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, because equal weight of the measurement error parameter space supporting both directions of causality.
-In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction.
-Full details are provided in Supplementary text 2.
+A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, because equal weight of the measurement error parameter space supports each direction of causality.
+In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction (S2 Text).
 
@@
 Note that following the first section of the Results we no longer include the bias terms for simplicity.
-We have formulated the the non-causal model as:
+We have formulated the non-causal model as:
 
@@
 Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables.
-Similar patterns of results were obtained for different values of $cor(g, x)$ and $cor(g, u)$.
+Similar patterns of results were obtained for different values of $cor(g, x)$.
 
@@
 
-Two sample MR [@Pierce2013] was performed using the summary statistics for genetic influences on gene expression and DNA methylation.
+Two sample MR [@Pierce2013] was performed using summary statistics for genetic influences on gene expression and DNA methylation.
 To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016].
@@
 In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Supplementary text 1) that
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (S1 Text) that
 
@@
 We observe that the CIT is more conservative under the null model of no association owing to the omnibus test statistic comprising several statistical tests.
-The FDR using a p-value threshold of 0.05 appears to be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
+The FDR using a p-value threshold of 0.05 appears to be close to zero, whereas for the MR Steiger method the FDR is around 0.05.
 Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases more steeply than that of the MR Steiger method.
 
-For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Supplementary text 2).
+For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, S2 Text).
 This metric can be used to evaluate the reliability of MR Steiger test.
 
-We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Supplementary text 2) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (S2 Text) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
 `r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
@@
 
-Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Supplementary text 3).
-The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (Supplementary figure 2).
+Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (S3 Text).
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (S2 Fig).
 Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
@@
 The performance was compared in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
-Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
+Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing $x$ and $y$ independently.
 
@@
 
-![](../images/dag-01.png)
+<!-- ![](../images/dag-01.png) -->
 
@@
 
-\newpage
+<!-- \newpage -->
 
+<!-- 
 ```{r cit_measurement_error_figure, fig.width=10, fig.height=10 }
@@
 ```	
-
+ -->
 `r fig_nums("cit_measurement_error_figure")`
 
-\newpage
+<!-- \newpage -->
 
-```{r causality_exists_tpr, fig.width=10, fig.height=10 }
+<!-- ```{r causality_exists_tpr, fig.width=10, fig.height=10 }
 
 ```
-
+ -->
 `r fig_nums("causality_exists_tpr")`
 
-\newpage
+<!-- \newpage -->
 
-```{r steiger_sensitivity_plot, fig.width=12, fig.height=6 }
+<!-- ```{r steiger_sensitivity_plot, fig.width=12, fig.height=6 }
 
 ```
-
+ -->
 `r fig_nums("steiger_sensitivity_plot")`
 
-\newpage
+<!-- \newpage -->
 
+<!-- 
 ```{r cit_mr_comparison_figure, fig.width=10, fig.height=10 }
@@
 ```
-
+ -->
 `r fig_nums("cit_mr_comparison_figure")`
 
-\newpage
+<!-- \newpage -->
 
-```{r shakhplot}
+<!-- ```{r shakhplot}
 
 ```
-
+ -->
 `r fig_nums("shakhplot")`
 
-\newpage
+<!-- \newpage -->
 
-\newpage
+<!-- \newpage -->
 
@@
 
-**Supplementary text 1** The influence of measurement error in the exposure on mediation-based estimated
+**S1 Text. The influence of measurement error in the exposure on mediation-based estimated**
 
-**Supplementary text 2** Sensitivity analysis for measurement error on the MR Steiger test
+**S2 Text. Sensitivity analysis for measurement error on the MR Steiger test**
 
-**Supplementary text 3** The influence of unmeasured confounding on the inference of causal directions
+**S3 Text. The influence of unmeasured confounding on the inference of causal directions**
 
-**Supplementary figure 1** Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
+**S1 Fig.
+Influence of confounding on CIT.** Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
 Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype.
@@
 
-**Supplementary figure 2** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
+**S2 Fig.
+Influence of confounding on MR Steiger.** Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
 Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-24 docs/manuscript.tex
updated, preparing for proofs

@@
   \postdate{\par}
-  \date{10 October 2017}
+  \date{24 October 2017}
 
@@
 unable to distinguish a true causal association from reverse causation
-or confounding (1--3). In response to this, the use of genetic
+or confounding {[}1--3{]}. In response to this, the use of genetic
 associations to instrument traits has emerged as a technique for
@@
 with the coincident rise in genome-wide association studies it is now a
-prominent tool that is applied in several different guises (3--6).
+prominent tool that is applied in several different guises {[}3--6{]}.
 However, shifting from observational associations to instrumentation
@@
 through a whole plethora of mechanisms, which are often specific to the
-study design and difficult to avoid (7,8). Array technology is now
+study design and difficult to avoid {[}7,8{]}. Array technology is now
 commonly used to obtain high throughput phenotyping at low cost, but
@@
 have some amount of noise around the true value due to imperfect
-sensitivity (9,10). Relatedly, if the measurement of biological interest
-is the methylation level in a T cell, then measurement error of this
-value can be introduced by using methylation levels from whole blood
-samples because the measured value will be an assay of many cell types
-(11).
+sensitivity {[}9,10{]}. Relatedly, if the measurement of biological
+interest is the methylation level in a T cell, then measurement error of
+this value can be introduced by using methylation levels from whole
+blood samples because the measured value will be an assay of many cell
+types {[}11{]}.
 
@@
 a proxy for adiposity, but it is clear that the correlation between BMI
-and underlying adiposity is not perfect (12), leading to the problem
+and underlying adiposity is not perfect {[}12{]}, leading to the problem
 that phenotypes may be imprecisely defined. A similar problem of
@@
 compared to the original variable. The sources of measurement error are
-not limited to this list (8), and its impact has been explored in the
-epidemiological literature extensively (13,14). Given the
+not limited to this list {[}8{]}, and its impact has been explored in
+the epidemiological literature extensively {[}13,14{]}. Given the
 near-ubiquitous presence of measurement error in phenomic data it is
@@
 the exposure (instruments), where the inheritance of an allele serves as
-a random lifetime allocation of differential exposure levels (15,16).
-Two statistical approaches to exploiting the properties of genetic
-instruments are widely used: mediation-based approaches and Mendelian
-randomisation (MR).
+a random lifetime allocation of differential exposure levels
+{[}15,16{]}. Two statistical approaches to exploiting the properties of
+genetic instruments are widely used: mediation-based approaches and
+Mendelian randomisation (MR).
 
@@
 influence of the exposure on the outcome. This forms the basis of a
-number of methods such as genetical genomics (17), the regression-based
-causal inference test (CIT) (4,18), a structural equation modelling
-(SEM) implementation in the NEO software (5), and various other methods
-including Bayesian approaches (6). They have been employed by a number
-of recent publications that make causal inferences in large scale `omics
-datasets (6,19--23).
+number of methods such as genetical genomics {[}17{]}, the
+regression-based causal inference test (CIT) {[}4,18{]}, a structural
+equation modelling (SEM) implementation in the NEO software {[}5{]}, and
+various other methods including Bayesian approaches {[}6{]}. They have
+been employed by a number of recent publications that make causal
+inferences in large scale `omics datasets {[}6,19--23{]}.
 
@@
 assumption can be relaxed in various ways when multiple instruments are
-available for a putative exposure (24,25) and a number of sensitivity
-tests are now available to improve reliability (26). Additionally, if
-valid genetic instruments are known for both traits of interest then MR
-can be performed in both directions (bi-directional MR), testing the
-influence of one trait on the other and vice versa, to infer the causal
-direction between the two phenotypes (27,28).
+available for a putative exposure {[}24,25{]} and a number of
+sensitivity tests are now available to improve reliability {[}26{]}.
+Additionally, if valid genetic instruments are known for both traits of
+interest then MR can be performed in both directions (bi-directional
+MR), testing the influence of one trait on the other and vice versa, to
+infer the causal direction between the two phenotypes {[}27,28{]}.
 
@@
 function, and it often takes years of follow up studies to understand
-the biological nature of a putative GWAS hit (29). If multiple
+the biological nature of a putative GWAS hit {[}29{]}. If multiple
 instruments are available for an hypothesised exposure, which is
 increasingly typical for complex traits that are analysed in large GWAS
-consortia, then techniques can be applied to mitigate these issues (16).
-But these techniques cannot always be applied in the case of determining
-causal directions between 'omic measures where typically only one
-cis-acting SNP is known. For example if a DNA methylation probe is
-associated with expression of an adjacent gene, then is a cis-acting SNP
-an instrument for the DNA methylation level, or the gene expression
+consortia, then techniques can be applied to mitigate these issues
+{[}16{]}. But these techniques cannot always be applied in the case of
+determining causal directions between 'omic measures where typically
+only one cis-acting SNP is known. For example if a DNA methylation probe
+is associated with expression of an adjacent gene, then is a cis-acting
+SNP an instrument for the DNA methylation level, or the gene expression
 level (Figure 1)?
@@
 inference to be drawn when exposure variables and outcome variables are
-measured in different samples (30). This has the crucial advantage of
-improving statistical power by allowing analysis in much larger sample
-sizes, and dramatically expands the breadth of possible phenotypic
-relationships that can be evaluated (26). Second, the mediation-based
-approach of adjusting the outcome for the exposure to nullify the
-association between the SNP and the outcome is affected by unmeasured
-confounding of the exposure and outcome. This is because adjusting the
-outcome by the exposure induces a collider effect between the SNP and
-outcome (31), and in order to fully abrogate this association one must
-also adjust for all (hidden or otherwise) confounders. MR does not
-suffer from this problem because it does not test for association
-through adjustment. Third, when MR assumptions are satisfied the method
-is robust to there being measurement error in the exposure variable
-(32). Indeed instrumental variable (IV) analysis was in part initially
-introduced as a correction for measurement error in the exposure (33),
-whereas it has been noted that both classic mediation-based analyses
-(13,14,34,35) and mediation-based methods that use instrumental
-variables (36,37) are prone to be unreliable in its presence.
+measured in different samples {[}30{]}. This has the crucial advantage
+of improving statistical power by allowing analysis in much larger
+sample sizes, and dramatically expands the breadth of possible
+phenotypic relationships that can be evaluated {[}26{]}. Second, the
+mediation-based approach of adjusting the outcome for the exposure to
+nullify the association between the SNP and the outcome is affected by
+unmeasured confounding of the exposure and outcome. This is because
+adjusting the outcome by the exposure induces a collider effect between
+the SNP and outcome {[}31{]}, and in order to fully abrogate this
+association one must also adjust for all (hidden or otherwise)
+confounders. MR does not suffer from this problem because it does not
+test for association through adjustment. Third, when MR assumptions are
+satisfied the method is robust to there being measurement error in the
+exposure variable {[}32{]}. Indeed instrumental variable (IV) analysis
+was in part initially introduced as a correction for measurement error
+in the exposure {[}33{]}, whereas it has been noted that both classic
+mediation-based analyses {[}13,14,34,35{]} and mediation-based methods
+that use instrumental variables {[}36,37{]} are prone to be unreliable
+in its presence.
 
@@
 weaknesses, triangulation of evidence from as many sources as possible
-should be practiced in causal inference (38).
+should be practiced in causal inference {[}38{]}.
 
@@
 the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
-For example, following Pierce and VanderWeele (32) we can define
+For example, following Pierce and VanderWeele {[}32{]} we can define
 
@@
 complete linkage disequilibrium with the causal variant) will reduce
-power in MR but will not incur bias (3,13,32). We also assume that
+power in MR but will not incur bias {[}3,13,32{]}. We also assume that
 measurement error in the exposure and the outcome are uncorrelated.
@@
 
-First we describe how the CIT method (4) is implemented in the R package
-\emph{R/cit} (18). The methodology of the CIT is as follows. Assume an
-exposure \(x\) is instrumented by a SNP \(g\), and the exposure \(x\)
-causes an outcome \(y\), as described above. The following tests are
-then performed:
+First we describe how the CIT method {[}4{]} is implemented in the R
+package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
+Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
+exposure \(x\) causes an outcome \(y\), as described above. The
+following tests are then performed:
 
@@
 empirically obtained estimates for simulated variables where the
-independence model is true (full details are given in (4)). We note here
-that this approach is liable to fail, even when there is a true causal
-relationship, when confounders of the exposure and outcome are present,
-as these will induce collider bias.
+independence model is true (full details are given in {[}4{]}). We note
+here that this approach is liable to fail, even when there is a true
+causal relationship, when confounders of the exposure and outcome are
+present, as these will induce collider bias.
 
@@
   \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
-  confounding (Supplementary figure 1) and no call is made.
+  confounding (S1 fig) and no call is made.
 \end{itemize}
@@
 not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds (39).
+is it reliable to depend on arbitrary significance thresholds {[}39{]}.
 
@@
 obtained using the \(systemfit\) function in the R package
-\(R/systemfit\) (40). Note that the value of \(p_{MR}\) is identical
+\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
 when using the same genetic variant to instrument the influence of the
@@
 instrument \(g\). This can be achieved by assessing which of the two
-variables has the biggest absolute correlation with \(g\) (Supplementary
-text 2), formalised by testing for a difference in the correlations
-\(\rho_{gx}\) and \(\rho_{gy}\) using Steiger's Z-test for correlated
-correlations within a population (41). It is calculated as
+variables has the biggest absolute correlation with \(g\) (S2 Text),
+formalised by testing for a difference in the correlations \(\rho_{gx}\)
+and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
+within a population {[}41{]}. It is calculated as
 
@@
 different samples (e.g.~from publicly available summary statistics
-(26,30)). Here the Steiger test of two independent correlations can be
-applied where.
+{[}26,30{]}). Here the Steiger test of two independent correlations can
+be applied where.
 
@@
 are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} (42).
+\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.
 
@@
 instrument is invalid. Second, some differential values of measurement
-error between the exposure and the outcome could induce incorrect causal
-directions, and this is explored in Supplementary text 2. Third, some
-levels of unmeasured confounding between the exposure and the outcome
-could induce the wrong causal direction, and this is explored in
-Supplementary text 3.
+error between the exposure and the outcome could lead to incorrect
+inference of the causal direction (S2 Text). Third, some levels of
+unmeasured confounding between the exposure and the outcome could lead
+to inference of the wrong causal direction (S3 Text).
 
@@
 
-As we show in Supplementary text 2 it is possible to infer the bounds of
-measurement error on \(x_o\) or \(y_o\) given known genetic
-associations. The maximum measurement imprecision of \(x_o\) is
-\(\rho_{g,x_o}\), because it is known that at least that much of the
-variance has been explained in \(x_o\) by \(g\). The minimum is 0,
-denoting perfectly measured trait values (the same logic applies to
-\(y_o\)). It is possible to simulate what the inferred causal direction
-would be for all values within these bounds.
+As we show in S2 Text it is possible to infer the bounds of measurement
+error on \(x_o\) or \(y_o\) given known genetic associations. The
+maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
+it is known that at least that much of the variance has been explained
+in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
+values (the same logic applies to \(y_o\)). It is possible to simulate
+what the inferred causal direction would be for all values within these
+bounds.
 
@@
 measurement error, because equal weight of the measurement error
-parameter space supporting both directions of causality. In general, the
+parameter space supports each direction of causality. In general, the
 \(R\) value denotes that the inferred direction of causality is \(R\)
-times more likely to be the empirical result than the opposite
-direction. Full details are provided in Supplementary text 2.
+times more likely to be the empirical result than the opposite direction
+(S2 Text).
 
@@
 Results we no longer include the bias terms for simplicity. We have
-formulated the the non-causal model as:
+formulated the non-causal model as:
 
@@
 of the simulated variables. Similar patterns of results were obtained
-for different values of \(cor(g, x)\) and \(cor(g, u)\).
+for different values of \(cor(g, x)\).
 
@@
 
-Two sample MR (30) was performed using the summary statistics for
+Two sample MR {[}30{]} was performed using summary statistics for
 genetic influences on gene expression and DNA methylation. To do this we
 obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al (43). These were filtered to be located on
-the same chromosome, have robust correlations after correcting for
+reported in Shakhbazov et al {[}43{]}. These were filtered to be located
+on the same chromosome, have robust correlations after correcting for
 multiple testing, and to share a SNP that had a robust cis-acting effect
@@
 
-All analysis was performed using the R programming language (44) and
+All analysis was performed using the R programming language {[}44{]} and
 code is made available at
 \url{https://github.com/explodecomputer/causal-directions} and
-implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).
+implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
+{[}26{]}.
 
@@
 measured value of \(y\), it can be shown using basic covariance
-properties (Supplementary text 1) that
+properties (S1 Text) that
 
@@
 or \(y\) then CIT can attempt to infer the causal direction. Though
-bi-directional MR can be used to orient causal directions (27), this
+bi-directional MR can be used to orient causal directions {[}27{]}, this
 requires knowledge of a valid instrument for each trait, and we were
@@
 statistical tests. The FDR using a p-value threshold of 0.05 appears to
-be close to 0, whereas for the MR Steiger method the FDR is around 0.05.
-Using the same p-value thresholds to declare significance in the
+be close to zero, whereas for the MR Steiger method the FDR is around
+0.05. Using the same p-value thresholds to declare significance in the
 non-null simulations, the general trend appears to be that the CIT power
@@
 that agree or disagree with the empirically inferred causal direction
-(Figure 4a, Supplementary text 2). This metric can be used to evaluate
-the reliability of MR Steiger test.
+(Figure 4a, S2 Text). This metric can be used to evaluate the
+reliability of MR Steiger test.
 
 We show that in the presence of measurement imprecision,
-\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Supplementary text 2)
-determines the range of parameters around which the MR Steiger test is
-liable to provide the wrong direction of causality (\emph{i.e.} if
-\(d>0\) then the MR Steiger test is likely to be correct about the
-causal direction). Figure 4b shows that when there is no measurement
-error in \(x\), the MR Steiger test is unlikely to infer the wrong
-direction of causality even if there is measurement error in \(y\). It
-also shows that in most cases where \(x\) is measured with error,
-especially when the causal effect between \(x\) and \(y\) is not very
-large, the sensitivity of the MR Steiger test to measurement error is
-relatively low.
+\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (S2 Text) determines the
+range of parameters around which the MR Steiger test is liable to
+provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
+the MR Steiger test is likely to be correct about the causal direction).
+Figure 4b shows that when there is no measurement error in \(x\), the MR
+Steiger test is unlikely to infer the wrong direction of causality even
+if there is measurement error in \(y\). It also shows that in most cases
+where \(x\) is measured with error, especially when the causal effect
+between \(x\) and \(y\) is not very large, the sensitivity of the MR
+Steiger test to measurement error is relatively low.
 
 Unmeasured confounding between the exposure and outcome can also give
-rise to problems with the MR Steiger approach (Supplementary text 3).
-The relationship between unmeasured confounding and causal orientation
-is complex across the parameter space of possible confounding values
-(Supplementary figure 2). Based on the range of parameter values that we
-explored, when the magnitude of the observational variance explained
-between the exposure and the outcome is below 0.2 the MR Steiger method
-is unlikely to return the incorrect causal direction due to unmeasured
-confounding.
+rise to problems with the MR Steiger approach (S3 Text). The
+relationship between unmeasured confounding and causal orientation is
+complex across the parameter space of possible confounding values (S2
+Fig). Based on the range of parameter values that we explored, when the
+magnitude of the observational variance explained between the exposure
+and the outcome is below 0.2 the MR Steiger method is unlikely to return
+the incorrect causal direction due to unmeasured confounding.
 
@@
 model'' where \(x\) and \(y\) were not causally related, but had a
-confounded association induced by the SNP \(g\) influencing a confounder
-variable \(u\).
+confounded association induced by the SNP \(g\) influencing \(x\) and
+\(y\) independently.
 
@@
 
-We performed two sample MR (30) for each association in the direction of
-causality inferred by the Stieger test. We observed that the sign of the
-MR estimate was generally in the same direction as the Pearson
-correlation coefficient reported by Shakhbazov et al (43) (Figure 6b).
-There was a moderate correlation between the absolute magnitudes of the
-causal correlation and the observational Pearson correlation (r = 0.45).
-Together these inferences suggest that even in estimating associations
-between `omic' variables, which are considered to be low level
-phenotypes, it is important to use causal inference methods over
-observational associations to infer causal effect sizes.
+We performed two sample MR {[}30{]} for each association in the
+direction of causality inferred by the Stieger test. We observed that
+the sign of the MR estimate was generally in the same direction as the
+Pearson correlation coefficient reported by Shakhbazov et al {[}43{]}
+(Figure 6b). There was a moderate correlation between the absolute
+magnitudes of the causal correlation and the observational Pearson
+correlation (r = 0.45). Together these inferences suggest that even in
+estimating associations between `omic' variables, which are considered
+to be low level phenotypes, it is important to use causal inference
+methods over observational associations to infer causal effect sizes.
 
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation (45).
+levels, consistent with expectation {[}45{]}.
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis are relatively well explored (46). Our analysis extends this to
-related methods such as CIT that are used in predominantly 'omic data.
-These methods are indeed susceptible to the same problem as standard
-mediation based analysis, and specifically we show that as measurement
-error in the (true) exposure variable increases, CIT is likely to have
-reduced statistical power, and liable to infer the wrong direction of
-causality. We also demonstrate that, though unintuitive, increasing
-sample size does not resolve the issue, rather it leads to more extreme
-p-values for the model that predicts the wrong direction of causality.
+analysis are relatively well explored {[}46{]}. Our analysis extends
+this to related methods such as CIT that are used in predominantly 'omic
+data. These methods are indeed susceptible to the same problem as
+standard mediation based analysis, and specifically we show that as
+measurement error in the (true) exposure variable increases, CIT is
+likely to have reduced statistical power, and liable to infer the wrong
+direction of causality. We also demonstrate that, though unintuitive,
+increasing sample size does not resolve the issue, rather it leads to
+more extreme p-values for the model that predicts the wrong direction of
+causality.
 
@@
 which have evaluated measurement error in these platforms do support
-this position (47,48), though making strong conclusions for this
+this position {[}47,48{]}, though making strong conclusions for this
 analysis is difficult because measurement error is likely to be study
@@
 another trait, and \(g\) having pleiotropic effects on both traits
-independently (49). Under certain conditions of measurement error the
-CIT test can distinguish these models. We also note that it is
+independently {[}49{]}. Under certain conditions of measurement error
+the CIT test can distinguish these models. We also note that it is
 straightforward to extend the MR Steiger approach to multiple
@@
 the causal and pleiotropic models, for example by evaluating the
-proportionality of the SNP-exposure and SNP-outcome effects (16).
+proportionality of the SNP-exposure and SNP-outcome effects {[}16{]}.
 Additionally, if there is at least one instrument for each trait then
 bi-directional MR can offer solutions to inferring the causal direction
-(16,28,50). We restricted the simulations to evaluating the causal
+{[}16,28,50{]}. We restricted the simulations to evaluating the causal
 inference between quantitative traits, but it is possible that the
@@
 variance explained on the liability scale, taking into account the
-population prevalence (51). However, our analysis goes beyond many
+population prevalence {[}51{]}. However, our analysis goes beyond many
 previous explorations of measurement error by assessing the impacts of
@@
 outcome through a mechanism other than the exposure) is not present.
-Recent method developments in MR (24,25) have focused on accounting for
-the issues that horizontal pleiotropy can introduce when multiple
+Recent method developments in MR {[}24,25{]} have focused on accounting
+for the issues that horizontal pleiotropy can introduce when multiple
 instruments are available, but how they perform in the presence of
@@
 on the outcome. This problem has shown to be the case for SNPs
-influencing different lipid fractions, for example (52,53).
+influencing different lipid fractions, for example {[}52,53{]}.
 
 Mediation based network approaches, that go beyond analyses of two
-variables, are very well established (37) and have a number of
+variables, are very well established {[}37{]} and have a number of
 extensions that make them valuable tools, including for example network
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics (54), multivariate MR (53) and
-mediation through MR (55--57) may offer more robust alternatives for
+applying MR to genetical genomics {[}54{]}, multivariate MR {[}53{]} and
+mediation through MR {[}55--57{]} may offer more robust alternatives for
 these more complicated problems.
@@
 \subsection{Figures}\label{figures}
-
-\includegraphics{../images/dag-01.png}
 
@@
 
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/cit_measurement_error_figure-1.pdf}
-
 Figure 2: The CIT was performed on simulated variables where the
@@
 
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/causality_exists_tpr-1.pdf}
-
 Figure 3: Outcomes were simulated to be unrelated to the exposure
@@
 varying sample sizes (columns of boxes).
-
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
 
@@
 
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}
-
 Figure 5: a) Outcome \(y\) was simulated to be caused by exposure \(x\)
@@
 
-\newpage
-
-\includegraphics{manuscript_files/figure-latex/shakhplot-1.pdf}
-
 Figure 6: Using 458 putative associations between DNA methylation and
@@
 
-\newpage
-
-\newpage
-
 \subsection{Supplementary information}\label{supplementary-information}
 
-\textbf{Supplementary text 1} The influence of measurement error in the
-exposure on mediation-based estimated
-
-\textbf{Supplementary text 2} Sensitivity analysis for measurement error
-on the MR Steiger test
-
-\textbf{Supplementary text 3} The influence of unmeasured confounding on
-the inference of causal directions
-
-\textbf{Supplementary figure 1} Illustrative simulations (\(n = 5000\))
-showing the results from CIT analysis under a model of confounding.
-Here, the phenotypes \(x\) and \(y\) are not causally related, but there
-is a genetic effect and a confounder both influencing each phenotype.
-Each point represents a single simulation. Where power is high (when the
-absolute values of the \(x\) and \(y\) axes are large) the CIT returns a
-significant result (\(p < 0.01\)) when testing the causal effect of
-\(x\) on \(y\), and when testing the causal effect of \(y\) on \(x\).
-
-\textbf{Supplementary figure 2} Graph representing the unmeasured
-confounding parameters that will lead to the MR Steiger test returning
-the wrong causal direction. Columns of boxes represent different signed
-values of the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).
+\textbf{S1 Text. The influence of measurement error in the exposure on
+mediation-based estimated}
+
+\textbf{S2 Text. Sensitivity analysis for measurement error on the MR
+Steiger test}
+
+\textbf{S3 Text. The influence of unmeasured confounding on the
+inference of causal directions}
+
+\textbf{S1 Fig. Influence of confounding on CIT.} Illustrative
+simulations (\(n = 5000\)) showing the results from CIT analysis under a
+model of confounding. Here, the phenotypes \(x\) and \(y\) are not
+causally related, but there is a genetic effect and a confounder both
+influencing each phenotype. Each point represents a single simulation.
+Where power is high (when the absolute values of the \(x\) and \(y\)
+axes are large) the CIT returns a significant result (\(p < 0.01\)) when
+testing the causal effect of \(x\) on \(y\), and when testing the causal
+effect of \(y\) on \(x\).
+
+\textbf{S2 Fig. Influence of confounding on MR Steiger.} Graph
+representing the unmeasured confounding parameters that will lead to the
+MR Steiger test returning the wrong causal direction. Columns of boxes
+represent different signed values of the observational variance
+explained between \(x\) and \(y\) (\(R^2_{xy}\)).
 
@@
 effects? relative risk estimation when correlated exposures are measured
-imprecisely. Journal of Clinical Epidemiology. 1991;44(11):1223--31.
+imprecisely. Journal of Clinical Epidemiology. Pergamon; 1991;44:
+1223--1231.
+doi:\href{https://doi.org/10.1016/0895-4356(91)90155-3}{10.1016/0895-4356(91)90155-3}
 
@@
 2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
-2002;325(7378):1437--8.
+2002;325: 1437--8.
 
@@
 3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
-potentials, and limitations. International journal of epidemiology
-{[}Internet{]}. 2004 Feb;33(1):30--42. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/15075143}
+potentials, and limitations. International journal of epidemiology.
+2004;33: 30--42.
+doi:\href{https://doi.org/10.1093/ije/dyh132}{10.1093/ije/dyh132}
 
@@
 4. Millstein J, Zhang B, Zhu J, Schadt EE. Disentangling molecular
-relationships with a causal inference test. BMC genetics {[}Internet{]}.
-2009 Jan;10:23. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3224661\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3224661\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+relationships with a causal inference test. BMC genetics. 2009;10: 23.
+doi:\href{https://doi.org/10.1186/1471-2156-10-23}{10.1186/1471-2156-10-23}
 
@@
 orient the edges in quantitative trait networks: the NEO software. BMC
-systems biology {[}Internet{]}. 2008 Jan;2:34. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2387136\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2387136\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+systems biology. 2008;2: 34.
+doi:\href{https://doi.org/10.1186/1752-0509-2-34}{10.1186/1752-0509-2-34}
 
@@
 RM, et al. Variation and genetic control of chromatin architecture in
-humans. Cell {[}Internet{]}. 2015;162(5):1039--50. Available from:
-\url{http://dx.doi.org/10.1016/j.cell.2015.08.001}
+humans. Cell. Elsevier Inc. 2015;162: 1039--1050.
+doi:\href{https://doi.org/10.1016/j.cell.2015.08.001}{10.1016/j.cell.2015.08.001}
 
@@
 7. Houle D, Pélabon C, Wagner G, Hansen T. Measurement and meaning in
-biology. The Quarterly Review of Biology {[}Internet{]}.
-2011;86(1):3--34. Available from:
+biology. The Quarterly Review of Biology. 2011;86: 3--34. Available:
 \url{http://www.jstor.org/stable/10.1086/658408}
@@
 8. Hernán M a, Cole SR. Invited Commentary: Causal diagrams and
-measurement bias. American journal of epidemiology {[}Internet{]}. 2009
-Oct;170(8):959--62; discussion 963--4. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2765368\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=2765368\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+measurement bias. American journal of epidemiology. 2009;170: 959--62;
+discussion 963--4.
+doi:\href{https://doi.org/10.1093/aje/kwp293}{10.1093/aje/kwp293}
 
@@
 of the American Association for Cancer Research, cosponsored by the
-American Society of Preventive Oncology {[}Internet{]}. 2013
-Jun;22(6):1052--60. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3687782\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3687782\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+American Society of Preventive Oncology. 2013;22: 1052--60.
+doi:\href{https://doi.org/10.1158/1055-9965.EPI-13-0114}{10.1158/1055-9965.EPI-13-0114}
 
@@
 the Illumina Infinium HumanMethylation450 microarray. Epigenetics :
-official journal of the DNA Methylation Society {[}Internet{]}. 2013
-Feb;8(2):203--9. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3592906\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3592906\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+official journal of the DNA Methylation Society. 2013;8: 203--9.
+doi:\href{https://doi.org/10.4161/epi.23470}{10.4161/epi.23470}
 
@@
 Nelson HH, et al. DNA methylation arrays as surrogate measures of cell
-mixture distribution. BMC bioinformatics {[}Internet{]}. 2012 Jan;13:86.
-Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3532182\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3532182\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+mixture distribution. BMC bioinformatics. 2012;13: 86.
+doi:\href{https://doi.org/10.1186/1471-2105-13-86}{10.1186/1471-2105-13-86}
 
@@
 12. Ahima RS, Lazar MA. Physiology. The health risk of obesity--better
-metrics imperative. Science (New York, NY) {[}Internet{]}. 2013
-Aug;341(6148):856--8. Available from:
-\url{http://science.sciencemag.org/content/341/6148/856.abstract}
+metrics imperative. Science (New York, NY). American Association for the
+Advancement of Science; 2013;341: 856--8.
+doi:\href{https://doi.org/10.1126/science.1241244}{10.1126/science.1241244}
 
@@
 types of measurement error in the mediator. Epidemiology (Cambridge,
-Mass) {[}Internet{]}. 2012 Jul;23(4):551--60. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/22526092}
+Mass). 2012;23: 551--60.
+doi:\href{https://doi.org/10.1097/EDE.0b013e318254f5de}{10.1097/EDE.0b013e318254f5de}
 
@@
 matters when estimating indirect effects. Journal of epidemiology and
-community health {[}Internet{]}. 2013 May;67(5):458--66. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/23386673}
+community health. 2013;67: 458--66.
+doi:\href{https://doi.org/10.1136/jech-2012-201813}{10.1136/jech-2012-201813}
 
@@
 epidemiology contribute to understanding environmental determinants of
-disease? International Journal of Epidemiology {[}Internet{]}. 2003
-Feb;32(1):1--22. Available from:
-\url{http://www.ije.oxfordjournals.org/cgi/doi/10.1093/ije/dyg070}
+disease? International Journal of Epidemiology. 2003;32: 1--22.
+doi:\href{https://doi.org/10.1093/ije/dyg070}{10.1093/ije/dyg070}
 
@@
 for causal inference in epidemiological studies. Human molecular
-genetics. 2014 Jul;23(R1):R89-----R98.
+genetics. Oxford Univ Press; 2014;23: R89-----R98.
+doi:\href{https://doi.org/10.1093/hmg/ddu328}{10.1093/hmg/ddu328}
 
@@
 An integrative genomics approach to infer causal associations between
-gene expression and disease. Nature Genetics {[}Internet{]}. 2005
-Jul;37(7):710--7. Available from:
-\href{http://dx.doi.org/10.1038/ng1589\%20http://www.nature.com/ng/journal/v37/n7/full/ng1589.html\%20http://www.nature.com/ng/journal/v37/n7/pdf/ng1589.pdf}{http://dx.doi.org/10.1038/ng1589 http://www.nature.com/ng/journal/v37/n7/full/ng1589.html http://www.nature.com/ng/journal/v37/n7/pdf/ng1589.pdf}
+gene expression and disease. Nature Genetics. 2005;37: 710--717.
+doi:\href{https://doi.org/10.1038/ng1589}{10.1038/ng1589}
 
@@
 18. Millstein J. cit: Causal Inference Test. R package version 1.9
-{[}Internet{]}. 2016. Available from:
+{[}Internet{]}. 2016. Available:
 \url{http://cran.r-project.org/package=cit}
@@
 MC, et al. Integrative genomic analysis identifies epigenetic marks that
-mediate genetic risk for epithelial ovarian cancer. BMC medical genomics
-{[}Internet{]}. 2014 Jan;7(1):8. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3916313\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3916313\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+mediate genetic risk for epithelial ovarian cancer. BMC medical
+genomics. BMC Medical Genomics; 2014;7: 8.
+doi:\href{https://doi.org/10.1186/1755-8794-7-8}{10.1186/1755-8794-7-8}
 
@@
 intermediary of genetic risk in rheumatoid arthritis. Nature
-biotechnology {[}Internet{]}. 2013 Feb;31(2):142--7. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3598632\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3598632\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+biotechnology. 2013;31: 142--7.
+doi:\href{https://doi.org/10.1038/nbt.2487}{10.1038/nbt.2487}
 
@@
 integrated epigenomic analysis for type 2 diabetes susceptibility loci
-in monozygotic twins. Nature communications {[}Internet{]}. 2014
-Jan;5:5719. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4284644\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4284644\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+in monozygotic twins. Nature communications. 2014;5: 5719.
+doi:\href{https://doi.org/10.1038/ncomms6719}{10.1038/ncomms6719}
 
@@
 \(\alpha\)2A-adrenergic receptor antagonist. Science translational
-medicine {[}Internet{]}. 2014 Oct;6(257):257ra139. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/25298321}
+medicine. 2014;6: 257ra139.
+doi:\href{https://doi.org/10.1126/scitranslmed.3009934}{10.1126/scitranslmed.3009934}
 
@@
 and evidence of epigenetic mediation in US children. Nature
-communications {[}Internet{]}. 2015 Jan;6:6304. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4340086\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4340086\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+communications. 2015;6: 6304.
+doi:\href{https://doi.org/10.1038/ncomms7304}{10.1038/ncomms7304}
 
@@
 invalid instruments: effect estimation and bias detection through Egger
-regression. International Journal of Epidemiology. 2015;44(2):512--25.
+regression. International Journal of Epidemiology. 2015;44: 512--25.
 
@@
 Estimation in Mendelian Randomization with Some Invalid Instruments
-Using a Weighted Median Estimator. Genetic Epidemiology {[}Internet{]}.
-2016 May;40(4):304--14. Available from:
-\href{http://www.ncbi.nlm.nih.gov/pubmed/27061298\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733\%20http://doi.wiley.com/10.1002/gepi.21965}{http://www.ncbi.nlm.nih.gov/pubmed/27061298 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC4849733 http://doi.wiley.com/10.1002/gepi.21965}
+Using a Weighted Median Estimator. Genetic Epidemiology. 2016;40:
+304--314.
+doi:\href{https://doi.org/10.1002/gepi.21965}{10.1002/gepi.21965}
 
@@
 elucidating direction of causation through reciprocal Mendelian
-randomization. International journal of obesity (2005) {[}Internet{]}.
-2011 Feb;35(2):300--8. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/20714329}
+randomization. International journal of obesity (2005). 2011;35: 300--8.
+doi:\href{https://doi.org/10.1038/ijo.2010.137}{10.1038/ijo.2010.137}
 
@@
 Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
-editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
-from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}
+editor. PLoS Medicine. 2014;11: e1001618.
+doi:\href{https://doi.org/10.1371/journal.pmed.1001618}{10.1371/journal.pmed.1001618}
 
@@
 al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
-New England journal of medicine. 2015;373(10):895--907.
+New England journal of medicine. 2015;373: 895--907.
+doi:\href{https://doi.org/10.1056/NEJMoa1502214}{10.1056/NEJMoa1502214}
 
@@
 studies: subsample and 2-sample instrumental variable estimators.
-American journal of epidemiology {[}Internet{]}. 2013
-Oct;178(7):1177--84. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3783091\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+American journal of epidemiology. 2013;178: 1177--84.
+doi:\href{https://doi.org/10.1093/aje/kwt084}{10.1093/aje/kwt084}
 
@@
 31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
-selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
-Sep;15(5):615--25. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}
+selection bias. Epidemiology (Cambridge, Mass). 2004;15: 615--25.
+Available: \url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}
 
@@
 measurement error on bias, precision and power in Mendelian
-randomization studies. International Journal of Epidemiology
-{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
-\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}
+randomization studies. International Journal of Epidemiology. Oxford
+University Press; 2012;41: 1383--1393.
+doi:\href{https://doi.org/10.1093/ije/dys141}{10.1093/ije/dys141}
 
@@
 Schooling from a New Sample of Twins. The American Economic Review.
-1994;84(5):1157--73.
+1994;84: 1157--1173.
+doi:\href{https://doi.org/10.1016/S0272-7757(98)00038-7}{10.1016/S0272-7757(98)00038-7}
 
@@
 34. Nagarajan R, Scutari M. Impact of noise on molecular network
-inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
-from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+inference. PloS one. 2013;8: e80735.
+doi:\href{https://doi.org/10.1371/journal.pone.0080735}{10.1371/journal.pone.0080735}
 
@@
 Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
-2010;527--36.
+2010; 527--536.
 
@@
 36. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
-networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
+networks. arXiv. 2015;arXiv:1512. Available:
 \url{http://arxiv.org/abs/1512.05574}
@@
 37. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
-Probabilistic Computational Causal Discovery for Systems Biology. In:
-Uncertainty in biology: A computational modeling approach
-{[}Internet{]}. Springer; 2015. p. 47. Available from:
+Probabilistic Computational Causal Discovery for Systems Biology.
+Uncertainty in biology: A computational modeling approach. Springer;
+2015. p. 47. Available:
 \href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}
@@
 38. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
-epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
-Jan;19(R1):dyw314. Available from:
-\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyw314}
+epidemiology. International Journal of Epidemiology. 2017;19: dyw314.
+doi:\href{https://doi.org/10.1093/ije/dyw314}{10.1093/ije/dyw314}
 
@@
 39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
-significance tests? BMJ. 2001;322(7280):226--31.
+significance tests? BMJ. 2001;322: 226--231.
 
@@
 40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. Journal of Statistical Software
-{[}Internet{]}. 2007 Dec;23(4):1--40. Available from:
-\url{https://www.jstatsoft.org/index.php/jss/article/view/v023i04/v23i04.pdf}
+Systems of Simultaneous Equations in R. Journal of Statistical Software.
+2007;23: 1--40.
+doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
 
@@
 41. Steiger JH. Tests for comparing elements of a correlation matrix.
-Psychological Bulletin. 1980;87(2):245--51.
+Psychological Bulletin. 1980;87: 245--251.
+doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
 
@@
 Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
-University; 2015. Available from:
+University; 2015. Available:
 \url{http://cran.r-project.org/package=psych}
@@
 PM, et al. Shared genetic control of expression and methylation in
-peripheral blood. BMC genomics {[}Internet{]}. 2016 Jan;17(1):278.
-Available from:
-\url{http://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-016-2498-4}
+peripheral blood. BMC genomics. BioMed Central; 2016;17: 278.
+doi:\href{https://doi.org/10.1186/s12864-016-2498-4}{10.1186/s12864-016-2498-4}
 
@@
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2015. Available from: \url{https://www.r-project.org/}
+2015. Available: \url{https://www.r-project.org/}
 
@@
 45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
-development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
-\url{http://genesdev.cshlp.org/content/16/1/6.long}
+development. 2002;16: 6--21.
+doi:\href{https://doi.org/10.1101/gad.947102}{10.1101/gad.947102}
 
@@
 Serious and Misleading Consequences Due to Uncorrected Measurement
-Error. Psychological Methods. 2014;19(2):300--15.
+Error. Psychological Methods. 2014;19: 300--315.
+doi:\href{https://doi.org/10.1037/a0033805}{10.1037/a0033805}
 
@@
 technical replicates: the Atherosclerosis Risk In Communities (ARIC)
-Study. BMC Bioinformatics {[}Internet{]}. 2014;15(1):312. Available
-from: \url{http://www.biomedcentral.com/1471-2105/15/312}
+Study. BMC Bioinformatics. 2014;15: 312.
+doi:\href{https://doi.org/10.1186/1471-2105-15-312}{10.1186/1471-2105-15-312}
 
@@
 Microarray Experiment but Both Are Outweighed by Changes Induced by
-Stimulation. Khodursky AB, editor. PLoS ONE {[}Internet{]}. 2011
-May;6(5):e19556. Available from:
-\url{http://dx.plos.org/10.1371/journal.pone.0019556}
+Stimulation. Khodursky AB, editor. PLoS ONE. Public Library of Science;
+2011;6: e19556.
+doi:\href{https://doi.org/10.1371/journal.pone.0019556}{10.1371/journal.pone.0019556}
 
@@
 Integration of summary data from GWAS and eQTL studies predicts complex
-trait gene targets. Nature Genetics {[}Internet{]}. 2016
-Mar;48(5):481--7. Available from:
-\url{http://www.nature.com/doifinder/10.1038/ng.3538}
+trait gene targets. Nature Genetics. Nature Research; 2016;48: 481--487.
+doi:\href{https://doi.org/10.1038/ng.3538}{10.1038/ng.3538}
 
@@
 Identify Genes Associated with 30 Complex Traits. The American Journal
-of Human Genetics {[}Internet{]}. 2017 Mar;100(3):473--87. Available
-from:
-\url{http://linkinghub.elsevier.com/retrieve/pii/S0002929717300320}
+of Human Genetics. 2017;100: 473--487.
+doi:\href{https://doi.org/10.1016/j.ajhg.2017.01.031}{10.1016/j.ajhg.2017.01.031}
 
@@
 association studies: quantification of power and genomic prediction
-accuracy. PLoS One. 2013;8(8):e71494.
+accuracy. PLoS One. 2013;8: e71494.
+doi:\href{https://doi.org/10.1371/journal.pone.0071494\%20\%5Bdoi\%5D/rPONE-D-13-18776\%20\%5Bpii\%5D}{10.1371/journal.pone.0071494 {[}doi{]}\textbackslash{}rPONE-D-13-18776 {[}pii{]}}
 
@@
 Common variants associated with plasma triglycerides and risk for
-coronary artery disease. Nature Genetics {[}Internet{]}. 2013
-Oct;45(11):1345--52. Available from:
-\url{http://www.nature.com/doifinder/10.1038/ng.2795}
+coronary artery disease. Nature Genetics. Nature Research; 2013;45:
+1345--1352. doi:\href{https://doi.org/10.1038/ng.2795}{10.1038/ng.2795}
 
@@
 multivariable Mendelian randomization to disentangle the causal effects
-of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
-Available from:
-\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
+of lipid fractions. PloS one. Public Library of Science; 2014;9:
+e108891.
+doi:\href{https://doi.org/10.1371/journal.pone.0108891}{10.1371/journal.pone.0108891}
 
@@
 randomization: a strategy for establishing the causal role of epigenetic
-processes in pathways to disease. International journal of epidemiology
-{[}Internet{]}. 2012 Feb;41(1):161--76. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+processes in pathways to disease. International journal of epidemiology.
+2012;41: 161--76.
+doi:\href{https://doi.org/10.1093/ije/dyr233}{10.1093/ije/dyr233}
 
@@
 cholesterol, and blood pressure as mediators from obesity to ischemic
-heart disease. Circulation research {[}Internet{]}. 2015
-Feb;116(4):665--73. Available from:
-\url{http://www.ncbi.nlm.nih.gov/pubmed/25411050}
+heart disease. Circulation research. 2015;116: 665--73.
+doi:\href{https://doi.org/10.1161/CIRCRESAHA.116.304846}{10.1161/CIRCRESAHA.116.304846}
 
@@
 investigate mediation in causal pathways. International journal of
-epidemiology {[}Internet{]}. 2015 Apr;44(2):484--95. Available from:
-\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}
+epidemiology. 2015;44: 484--95.
+doi:\href{https://doi.org/10.1093/ije/dyu176}{10.1093/ije/dyu176}
 
@@
 Challenges and novel approaches for investigating molecular mediation.
-Human molecular genetics {[}Internet{]}. 2016 Oct;25(R2):R149--56.
-Available from:
-\href{http://www.ncbi.nlm.nih.gov/pubmed/27439390\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}{http://www.ncbi.nlm.nih.gov/pubmed/27439390 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}
+Human molecular genetics. Oxford University Press; 2016;25: R149--R156.
+doi:\href{https://doi.org/10.1093/hmg/ddw197}{10.1093/hmg/ddw197}
 

### 2017-10-24 docs/s1text.Rmd
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.Rmd
@@
 title: "Supplementary materials: Orienting the causal relationship between imprecisely measured traits using GWAS summary data"
-date: "`r format(Sys.time(), '%d %B %Y')`"
 output:
@@
 
-## Supplementary text 1
+## S1 Text. The influence of measurement error in the exposure on mediation-based estimated
 
@@
 thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.
-
-\newpage
-
-## Supplementary text 2
-
-Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude.
-The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:
-
-$$
-x = \alpha_1 + \beta_1 g + e_1
-$$
-
-for the first stage and
-
-$$
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-$$
-
-where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$.
-Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
-
-In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
-
-$$
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-$$
-
-In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.
-
-For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$.
-For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.
-
-Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$.
-If $z < 0$ then we infer that $x \rightarrow y$.
-There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$.
-To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:
-
-$$
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-$$
-
-If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values.
-If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality.
-$V_{z}$, the total volume of the function (Supplementary figure 4), can be obtained analytically by solving:
-
-$$
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-$$
-
-$V_{z \ge 0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
-The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$.
-Hence,
-
-$$
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-$$
-
-Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.
-
-\newpage
-
-## Supplementary text 3
-
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on mediation-based methods for inferring causal directions, and can also adversely influence the estimate of the causal direction for the Steiger test.
-
-### Unmeasured confounding in mediation
-
-Including an unmeasured confounder, $u$, after ignoring intercept terms the exposure $x$ and outcome $y$ variables can be modelled as
-
-$$
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-$$
-
-The observational estimate of the causal effect of $x$ on $y$, $\hat{\beta}_x$ is obtained from
-
-$$
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-$$
-
-From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are non-zero.
-Relating to mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
-
-$$
-\hat{y}^* = \hat{\beta}_x x
-$$
-
-and
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-$$
-
-should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and $y|x$, which will introduce errors in inferring causal directions.
-
-### Unmeasured confounding in the MR Steiger test
-
-Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the MR Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
-After some algebra
-
-$$
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-$$
-
-and
-
-$$
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-$$
-
-Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
-The pattern of results were similar for different values of $\beta_{gx}$.
-We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
-
-\newpage
-
-![](../images/unmeasured_conf_steiger.png)
-
-Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
-Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-24 docs/s1text.tex
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.tex
@@
   \preauthor{}\postauthor{}
-  \predate{\centering\large\emph}
-  \postdate{\par}
-  \date{28 July 2017}
+  \date{}
+  \predate{}\postdate{}
 
@@
 
-\subsection{Supplementary text 1}\label{supplementary-text-1}
+\subsection{S1 Text. The influence of measurement error in the exposure
+on mediation-based
+estimated}\label{s1-text.-the-influence-of-measurement-error-in-the-exposure-on-mediation-based-estimated}
 
@@
 
-\newpage
-
-\subsection{Supplementary text 2}\label{supplementary-text-2}
-
-Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
-causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
-and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
-hypothesis test that provides a p-value for observing the difference in
-these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is \(x \to y\), two stage MR is formulated
-using the following regression models:
-
-\[
-x = \alpha_1 + \beta_1 g + e_1
-\]
-
-for the first stage and
-
-\[
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-\]
-
-where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
-free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
-the exposure variable \(x\), and it is expected that
-\(\rho_{g, x} > \rho_{g, y}\) because
-\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
-causal association between \(x\) and \(y\) (which is likely to be less
-than 1).
-
-In the presence of measurement error in \(x\) and \(y\), however, the
-empirical inference of the causal direction will instead be based on
-evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
-
-\[
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-\]
-
-In order to assess how reliable the inference of the causal direction is
-in the presence of measurement imprecision, we can evaluate the range of
-potential values of measurement error in \(x\) and \(y\) over which the
-empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
-return the wrong causal direction.
-
-For different values of \(\rho_{x,x_o}\),
-\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
-\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
-\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
-and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).
-
-Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
-explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
-we infer that \(x \rightarrow y\). There will be some values of
-\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
-\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
-causal direction with regards to measurement error, the objective is to
-compare the proportion of the parameter space that agrees with the
-inferred direction against the proportion which does not:
-
-\[
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-\]
-
-If \(R=1\) then the direction of causality is equally probable across
-the range of possible measurement error values. If \(R > 1\) then \(R\)
-times as much of the parameter space favours the inferred direction of
-causality. \(V_{z}\), the total volume of the function (Supplementary
-figure 4), can be obtained analytically by solving:
-
-\[
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-\]
-
-\(V_{z \ge 0}\), the proportion of the volume that lies above the
-\(z=0\) plane, can also be obtained analytically. The region of this
-volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
-where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
-\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,
-
-\[
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-\]
-
-Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).
-
-\newpage
-
-\subsection{Supplementary text 3}\label{supplementary-text-3}
-
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on
-mediation-based methods for inferring causal directions, and can also
-adversely influence the estimate of the causal direction for the Steiger
-test.
-
-\subsubsection{Unmeasured confounding in
-mediation}\label{unmeasured-confounding-in-mediation}
-
-Including an unmeasured confounder, \(u\), after ignoring intercept
-terms the exposure \(x\) and outcome \(y\) variables can be modelled as
-
-\[
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-\]
-
-The observational estimate of the causal effect of \(x\) on \(y\),
-\(\hat{\beta}_x\) is obtained from
-
-\[
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-\]
-
-From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
-when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
-mediation, where we attempt to test if \(g\) associates with \(y\) after
-adjusting \(y\) for \(x\), such that
-
-\[
-\hat{y}^* = \hat{\beta}_x x
-\]
-
-and
-
-\[
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-\]
-
-should any amount of unmeasured confounding exist, therefore, there will
-remain an association between \(g\) and \(y|x\), which will introduce
-errors in inferring causal directions.
-
-\subsubsection{Unmeasured confounding in the MR Steiger
-test}\label{unmeasured-confounding-in-the-mr-steiger-test}
-
-Similarly, we can investigate the extent to which unmeasured confounding
-will lead to the wrong causal direction between \(x\) and \(y\) using
-the MR Steiger test, evaluating the liability for the inequality
-\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
-
-\[
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-\]
-
-and
-
-\[
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-\]
-
-Supplementary figure 1 shows the relationship between the magnitude of
-the correlations between \(x\), \(y\) and the confounder \(u\) for a
-range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
-confounder effects. The pattern of results were similar for different
-values of \(\beta_{gx}\). We note that in most cases for the parameter
-values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
-less than 0.2, unmeasured confounding will not incur the wrong causal
-direction in the MR Steiger test.
-
-\newpage
-
-\includegraphics{../images/unmeasured_conf_steiger.png}
-
-Supplementary figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the MR Steiger test returning the wrong
-causal direction. Columns of boxes represent different signed values of
-the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).
-
 \end{document}

### 2017-10-24 docs/s2text.Rmd
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.Rmd
@@
 
-## Supplementary text 1
-
-We assume the following model
-
-$$
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-$$
-
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_o$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_o$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
-Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y_o - \hat{y}_o) = 0$ when $x$ causes $y$, where $\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o$ is the predicted value of $y_o$ using the measured value of $x_o$.
-
-We can split $cov(g, y_o - \hat{y}_o)$ into two parts, $cov(g, y_o)$ and $cov(g, \hat{y}_o)$.
-
-**Part 1**
-
-$$
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-**Part 2**
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-$$
-
-Simpifying further
-
-$$
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-$$
-
-which can be substituted back to give
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-Finally
-
-$$
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.
-
-\newpage
-
-## Supplementary text 2
+## S2 Text. Sensitivity analysis for measurement error on the MR Steiger test
 
@@
 Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.
-
-\newpage
-
-## Supplementary text 3
-
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on mediation-based methods for inferring causal directions, and can also adversely influence the estimate of the causal direction for the Steiger test.
-
-### Unmeasured confounding in mediation
-
-Including an unmeasured confounder, $u$, after ignoring intercept terms the exposure $x$ and outcome $y$ variables can be modelled as
-
-$$
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-$$
-
-The observational estimate of the causal effect of $x$ on $y$, $\hat{\beta}_x$ is obtained from
-
-$$
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-$$
-
-From this it is clear that $\beta_x$ and $\hat{\beta}_x$ will differ when both $\beta_{uy}$ and $\beta_{ux}$ are non-zero.
-Relating to mediation, where we attempt to test if $g$ associates with $y$ after adjusting $y$ for $x$, such that
-
-$$
-\hat{y}^* = \hat{\beta}_x x
-$$
-
-and
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-$$
-
-should any amount of unmeasured confounding exist, therefore, there will remain an association between $g$ and $y|x$, which will introduce errors in inferring causal directions.
-
-### Unmeasured confounding in the MR Steiger test
-
-Similarly, we can investigate the extent to which unmeasured confounding will lead to the wrong causal direction between $x$ and $y$ using the MR Steiger test, evaluating the liability for the inequality $cor(g,x)^2 > cor(g,y)^2$ being incorrect.
-After some algebra
-
-$$
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-$$
-
-and
-
-$$
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-$$
-
-Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
-The pattern of results were similar for different values of $\beta_{gx}$.
-We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
-
-\newpage
-
-![](../images/unmeasured_conf_steiger.png)
-
-Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
-Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-24 docs/s2text.tex
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.tex
@@
 
-\subsection{Supplementary text 1}\label{supplementary-text-1}
-
-We assume the following model
-
-\[
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-\]
-
-where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), \(x_o\) is the measured quantity of
-\(x\), where measurement error is incurred from linear transformation in
-\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
-\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
-measurement error is incurred from linear transformation in
-\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
-\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
-of association between \(g\) and \(y\) after conditioning on \(x\).
-Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
-when \(x\) causes \(y\), where
-\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
-value of \(y_o\) using the measured value of \(x_o\).
-
-We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
-and \(cov(g, \hat{y}_o)\).
-
-\textbf{Part 1}
-
-\[
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-\textbf{Part 2}
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-\]
-
-Simpifying further
-
-\[
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-\]
-
-which can be substituted back to give
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-Finally
-
-\[
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
-\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
-then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.
-
-\newpage
-
-\subsection{Supplementary text 2}\label{supplementary-text-2}
+\subsection{S2 Text. Sensitivity analysis for measurement error on the
+MR Steiger
+test}\label{s2-text.-sensitivity-analysis-for-measurement-error-on-the-mr-steiger-test}
 
@@
 
-\newpage
-
-\subsection{Supplementary text 3}\label{supplementary-text-3}
-
-We have assumed no unmeasured confounding in these simulations.
-Unmeasured confounding will however have potentially large influences on
-mediation-based methods for inferring causal directions, and can also
-adversely influence the estimate of the causal direction for the Steiger
-test.
-
-\subsubsection{Unmeasured confounding in
-mediation}\label{unmeasured-confounding-in-mediation}
-
-Including an unmeasured confounder, \(u\), after ignoring intercept
-terms the exposure \(x\) and outcome \(y\) variables can be modelled as
-
-\[
-\begin{aligned}
-y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
-x & = \beta_g g + \beta_{ux} u + \epsilon_g
-\end{aligned}
-\]
-
-The observational estimate of the causal effect of \(x\) on \(y\),
-\(\hat{\beta}_x\) is obtained from
-
-\[
-\begin{aligned}
-\hat{\beta}_x & = cov(x, y) / var(x) \\
-& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
-\end{aligned}
-\]
-
-From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
-when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
-mediation, where we attempt to test if \(g\) associates with \(y\) after
-adjusting \(y\) for \(x\), such that
-
-\[
-\hat{y}^* = \hat{\beta}_x x
-\]
-
-and
-
-\[
-\begin{aligned}
-cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
-& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
-& = (\beta_x - \hat{\beta}_x) var(g)
-\end{aligned}
-\]
-
-should any amount of unmeasured confounding exist, therefore, there will
-remain an association between \(g\) and \(y|x\), which will introduce
-errors in inferring causal directions.
-
-\subsubsection{Unmeasured confounding in the MR Steiger
-test}\label{unmeasured-confounding-in-the-mr-steiger-test}
-
-Similarly, we can investigate the extent to which unmeasured confounding
-will lead to the wrong causal direction between \(x\) and \(y\) using
-the MR Steiger test, evaluating the liability for the inequality
-\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra
-
-\[
-cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
-\]
-
-and
-
-\[
-cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
-\]
-
-Supplementary figure 1 shows the relationship between the magnitude of
-the correlations between \(x\), \(y\) and the confounder \(u\) for a
-range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
-confounder effects. The pattern of results were similar for different
-values of \(\beta_{gx}\). We note that in most cases for the parameter
-values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
-less than 0.2, unmeasured confounding will not incur the wrong causal
-direction in the MR Steiger test.
-
-\newpage
-
-\includegraphics{../images/unmeasured_conf_steiger.png}
-
-Supplementary figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the MR Steiger test returning the wrong
-causal direction. Columns of boxes represent different signed values of
-the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).
-
 \end{document}

### 2017-10-24 docs/s3text.Rmd
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.Rmd
@@
 
-## Supplementary text 1
-
-We assume the following model
-
-$$
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-$$
-
-where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_o$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_o$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$.
-Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$.
-Under the CIT, this is expected to be $cov(g, y_o - \hat{y}_o) = 0$ when $x$ causes $y$, where $\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o$ is the predicted value of $y_o$ using the measured value of $x_o$.
-
-We can split $cov(g, y_o - \hat{y}_o)$ into two parts, $cov(g, y_o)$ and $cov(g, \hat{y}_o)$.
-
-**Part 1**
-
-$$
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-**Part 2**
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-$$
-
-Simpifying further
-
-$$
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-$$
-
-which can be substituted back to give
-
-$$
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-Finally
-
-$$
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-$$
-
-thus $cov(g, y_o - \hat{y}_o) = 0$ if the measurement imprecision in $x_o$ is $var(\epsilon_m) = 0$. However if there is any imprecision then the condition $cov(g, y_o - \hat{y}_o) = 0$ will not hold.
-
-\newpage
-
-## Supplementary text 2
-
-Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude.
-The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:
-
-$$
-x = \alpha_1 + \beta_1 g + e_1
-$$
-
-for the first stage and
-
-$$
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-$$
-
-where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$.
-Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).
-
-In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:
-
-$$
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-$$
-
-In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.
-
-For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$.
-For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.
-
-Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$.
-If $z < 0$ then we infer that $x \rightarrow y$.
-There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$.
-To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:
-
-$$
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-$$
-
-If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values.
-If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality.
-$V_{z}$, the total volume of the function (Supplementary figure 4), can be obtained analytically by solving:
-
-$$
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-$$
-
-$V_{z \ge 0}$, the proportion of the volume that lies above the $z=0$ plane, can also be obtained analytically.
-The region of this volume is bound by the values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ where $0 = \rho_{g,y} - \rho_{g,x}$, which can be expanded to $\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}$.
-Hence,
-
-$$
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-$$
-
-Thus $V_{z < 0} = V_{z} - V_{z \geq 0}$.
-
-\newpage
-
-## Supplementary text 3
+## S3 Text. The influence of unmeasured confounding on the inference of causal directions
 
@@
 
-Supplementary figure 1 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
+S2 fig shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
 The pattern of results were similar for different values of $\beta_{gx}$.
 We note that in most cases for the parameter values explored, where the observational absolute $\hat{R}^2_{xy}$ is less than 0.2, unmeasured confounding will not incur the wrong causal direction in the MR Steiger test.
-
-\newpage
-
-![](../images/unmeasured_conf_steiger.png)
-
-Supplementary figure 1: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
-Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-24 docs/s3text.tex
updated, preparing for proofs

--- 2017-10-05 docs/supplementary.tex
@@
 
-\subsection{Supplementary text 1}\label{supplementary-text-1}
-
-We assume the following model
-
-\[
-\begin{aligned}
-x   & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y   & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
-\end{aligned}
-\]
-
-where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
-that has a direct effect on \(x\), \(x_o\) is the measured quantity of
-\(x\), where measurement error is incurred from linear transformation in
-\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
-\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
-measurement error is incurred from linear transformation in
-\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
-\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
-of association between \(g\) and \(y\) after conditioning on \(x\).
-Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
-when \(x\) causes \(y\), where
-\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
-value of \(y_o\) using the measured value of \(x_o\).
-
-We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
-and \(cov(g, \hat{y}_o)\).
-
-\textbf{Part 1}
-
-\[
-\begin{aligned}
-cov(g, y_o) & = cov(g, \beta_{my} y) \\
-            & = cov(g, \beta_{my} \beta_x x) \\
-            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
-            & = \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-\textbf{Part 2}
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
-                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
-                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
-\end{aligned}
-\]
-
-Simpifying further
-
-\[
-\begin{aligned}
-\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
-                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
-\end{aligned}
-\]
-
-which can be substituted back to give
-
-\[
-\begin{aligned}
-cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
-                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-Finally
-
-\[
-\begin{aligned}
-cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
-\end{aligned}
-\]
-
-thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
-\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
-then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.
-
-\newpage
-
-\subsection{Supplementary text 2}\label{supplementary-text-2}
-
-Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
-causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
-and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
-hypothesis test that provides a p-value for observing the difference in
-these correlations under the null hypothesis that they are equal.
-
-Assuming the causal direction is \(x \to y\), two stage MR is formulated
-using the following regression models:
-
-\[
-x = \alpha_1 + \beta_1 g + e_1
-\]
-
-for the first stage and
-
-\[
-y = \alpha_2 + \beta_2 \hat{x} + e_2
-\]
-
-where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
-free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
-the exposure variable \(x\), and it is expected that
-\(\rho_{g, x} > \rho_{g, y}\) because
-\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
-causal association between \(x\) and \(y\) (which is likely to be less
-than 1).
-
-In the presence of measurement error in \(x\) and \(y\), however, the
-empirical inference of the causal direction will instead be based on
-evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:
-
-\[
-\begin{aligned}
-\rho_{g, x_O} & > \rho_{g, y_O} \\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
-\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
-\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
-\end{aligned}
-\]
-
-In order to assess how reliable the inference of the causal direction is
-in the presence of measurement imprecision, we can evaluate the range of
-potential values of measurement error in \(x\) and \(y\) over which the
-empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
-return the wrong causal direction.
-
-For different values of \(\rho_{x,x_o}\),
-\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
-\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
-\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
-and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).
-
-Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
-explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
-we infer that \(x \rightarrow y\). There will be some values of
-\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
-\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
-causal direction with regards to measurement error, the objective is to
-compare the proportion of the parameter space that agrees with the
-inferred direction against the proportion which does not:
-
-\[
-R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
-\]
-
-If \(R=1\) then the direction of causality is equally probable across
-the range of possible measurement error values. If \(R > 1\) then \(R\)
-times as much of the parameter space favours the inferred direction of
-causality. \(V_{z}\), the total volume of the function (Supplementary
-figure 4), can be obtained analytically by solving:
-
-\[
-\begin{aligned}
-V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
-\end{aligned}
-\]
-
-\(V_{z \ge 0}\), the proportion of the volume that lies above the
-\(z=0\) plane, can also be obtained analytically. The region of this
-volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
-where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
-\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,
-
-\[
-\begin{aligned}
-V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
-& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
-\end{aligned}
-\]
-
-Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).
-
-\newpage
-
-\subsection{Supplementary text 3}\label{supplementary-text-3}
+\subsection{S3 Text. The influence of unmeasured confounding on the
+inference of causal
+directions}\label{s3-text.-the-influence-of-unmeasured-confounding-on-the-inference-of-causal-directions}
 
@@
 
-Supplementary figure 1 shows the relationship between the magnitude of
-the correlations between \(x\), \(y\) and the confounder \(u\) for a
-range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
-confounder effects. The pattern of results were similar for different
-values of \(\beta_{gx}\). We note that in most cases for the parameter
-values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
-less than 0.2, unmeasured confounding will not incur the wrong causal
+S2 fig shows the relationship between the magnitude of the correlations
+between \(x\), \(y\) and the confounder \(u\) for a range of
+\(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of confounder
+effects. The pattern of results were similar for different values of
+\(\beta_{gx}\). We note that in most cases for the parameter values
+explored, where the observational absolute \(\hat{R}^2_{xy}\) is less
+than 0.2, unmeasured confounding will not incur the wrong causal
 direction in the MR Steiger test.
 
-\newpage
-
-\includegraphics{../images/unmeasured_conf_steiger.png}
-
-Supplementary figure 1: Graph representing the unmeasured confounding
-parameters that will lead to the MR Steiger test returning the wrong
-causal direction. Columns of boxes represent different signed values of
-the observational variance explained between \(x\) and \(y\)
-(\(R^2_{xy}\)).
-
 \end{document}

### 2017-10-24 docs/supplementary.Rmd
updated, preparing for proofs

@@
 
-Supplementary figure 2 shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
+S2 fig shows the relationship between the magnitude of the correlations between $x$, $y$ and the confounder $u$ for a range of $\beta_{xy} = (-2,2)$, $\beta_{gx} = 0.1$ and a range of confounder effects.
 The pattern of results were similar for different values of $\beta_{gx}$.
@@
 
-Supplementary figure 1: Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
+S1 fig: Illustrative simulations ($n = 5000$) showing the results from CIT analysis under a model of confounding.
 Here, the phenotypes $x$ and $y$ are not causally related, but there is a genetic effect and a confounder both influencing each phenotype.
@@
 
-Supplementary figure 2: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
+S2 fig: Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction.
 Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$).

### 2017-10-24 scripts/mr_directionality_analysis.R

@@ -666,7 +666,7 @@ temp1$lab[temp1$r_ab==0] <- "Type 1 error"
 ggplot(subset(temp1, r_ab %in% c(0, 0.6) & r_za == "cor(x,g) = 0.01" & noiseb %in% c("cor(Y,Yo) = 0.4")), aes(x=noisea, y = prop_causality_exists)) +
 geom_bar(stat="identity", position="dodge", aes(fill=as.factor(test))) +
 facet_grid(lab ~ n) +
-labs(y="False positive rate", x = expression(epsilon[mx]), fill="Test") +
+labs(y="Discovery rate", x = expression(epsilon[mx]), fill="Test") +
 ylim(c(0,1))
 
 

### 2017-11-13 docs/manuscript.Rmd
reorganising manuscript for submission

@@
 
-## Model
+### Model
 
@@
 
+## Results
+
+### Mediation-based causal inference under measurement error
+
+In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
+When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (S1 Text) that
+
+$$
+\begin{aligned}
+cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
+                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
+\end{aligned}
+$$
+
+where
+
+$$
+D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
+$$
+
+Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
+Therefore, if there is any measurement error that incurs imprecision in $x$ (i.e.
+$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x_o$, which is in violation of the the 4th condition of the CIT.
+Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
+
+We performed simulations to verify that this problem does arise using the CIT method.
+`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}=1$) the CIT is reliable in identifying the correct causal direction.
+However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
+Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
+
+### Using MR Steiger to infer the direction of causality
+
+If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
+Though bi-directional MR can be used to orient causal directions [@Timpson2011], this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood.
+We go on to explore the scenarios in which the method is likely to return the correct or incorrect causal directions.
+
+We performed simulations to compare the power and type 1 error rates of MR and CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
+Comparing the performance of methods with different sets of assumptions can be difficult, but a basic comparison is shown in `r cfig("causality_exists_tpr")`.
+We observe that the CIT is more conservative under the null model of no association owing to the omnibus test statistic comprising several statistical tests.
+The FDR using a p-value threshold of 0.05 appears to be close to zero, whereas for the MR Steiger method the FDR is around 0.05.
+Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases more steeply than that of the MR Steiger method.
+
+For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, S2 Text).
+This metric can be used to evaluate the reliability of MR Steiger test.
+
+We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (S2 Text) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
+`r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
+It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the MR Steiger test to measurement error is relatively low.
+
+Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (S3 Text).
+The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (S2 Fig).
+Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
+
+### Comparison of CIT and MR Steiger for obtaining the correct direction of causality
+
+We used simulations to explore the performance of the MR Steiger approach in comparison to CIT for different levels of measurement error.
+The performance was compared in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
+Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing $x$ and $y$ independently.
+
+`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", the MR analysis is indeed liable to infer the wrong direction of causality when $d < 0$, and that this erroneous result is more likely to occur with increasing sample size.
+However, the CIT is in general more fallible to reporting a robust causal association for the wrong direction of causality.
+When $d > 0$ we find that in most cases the MR Steiger method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
+The CIT, unlike the MR Steiger test, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.
+
+### The causal relationship between gene expression and DNA methylation levels
+
+We used the MR Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations.
+We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
+The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (5%-95% quantiles `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`).
+We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms.
+`r cfig("shakhplot")`a shows that the conclusions about the direction of causality between DNA methylation and gene expression are very sensitive to measurement error.
+We made a strong assumption that either methylation influenced gene expression or vice versa, but it is certainly possible that the SNP is solely or additionally influencing some other trait that confounds the association between gene expression and DNA methylation.
+
+We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test.
+We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b).
+There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`).
+Together these inferences suggest that even in estimating associations between 'omic' variables, which are considered to be low level phenotypes, it is important to use causal inference methods over observational associations to infer causal effect sizes.
+
+We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing methylation levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
+
+## Discussion
+
+Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data.
+In the epidemiological literature issues of measurement error in mediation analysis are relatively well explored [@Cole2014].
+Our analysis extends this to related methods such as CIT that are used in predominantly 'omic data.
+These methods are indeed susceptible to the same problem as standard mediation based analysis, and specifically we show that as measurement error in the (true) exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
+We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
+
+Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation.
+Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces.
+This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest.
+But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
+Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods.
+We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages.
+First, it uses a formal statistical framework to test for the reliability of the assumed direction of causality.
+Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.
+
+We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data.
+The inferred causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
+For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association.
+Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.
+We also haven't accounted for the influence of winner's curse, which can inflate estimates of the variance explained by SNPs, with higher inflation expected amongst lower powered studies.
+Using p-values for genetic associations from replication studies will mitigate this problem.
+
+In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available.
+However, if there is only a single instrument it is difficult to separate between the two competing models of $g$ instrumenting a trait which causes another trait, and $g$ having pleiotropic effects on both traits independently [@Zhu2016].
+Under certain conditions of measurement error the CIT test can distinguish these models.
+We also note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent.
+Multiple instruments can indeed help to distinguish between the causal and pleiotropic models, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
+Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017].
+We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].
+However, our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
+
+Our new method attempts to infer causal directions under the assumption that horizontal pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) is not present.
+Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored.
+An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.
+However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood.
+We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies.
+We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated.
+We have also not addressed other issues pertaining to instrumental variables which are relevant to the question of instrument-exposure specification.
+One such problem is exposure misspecification, for example an instrument could associate with several closely related putative outcomes, with only one of them actually having a causal effect on the outcome.
+This problem has shown to be the case for SNPs influencing different lipid fractions, for example [@Do2013; @Burgess2014a].
+
+Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
+But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
+Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015; @Richmond2016] may offer more robust alternatives for these more complicated problems.
+
+The overarching result from our simulations is that, regardless of the method used, inferring the causal direction using an instrument of unknown biology is highly sensitive to measurement error.
+With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
+
 ## Methods
@@
 
-First we describe how the CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
-The methodology of the CIT is as follows.
+Here we describe how the CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016].
 Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ causes an outcome $y$, as described above.
@@
 
-The term in the 4th test can be rewritten as a $cov(g, y|x) = cov(g, y - \hat{y})$ where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, and $x$ is assumed to mediate the association between the SNP and the outcome.
+The term in the 4th test can be rewritten as $cov(g, y|x) = cov(g, y - \hat{y})$ where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, and $x$ is assumed to mediate the association between the SNP and the outcome.
 The condition in the 4th test is formulated as an equivalence testing problem that is estimated using simulations, comparing the estimate from the data against empirically obtained estimates for simulated variables where the independence model is true (full details are given in [@Millstein2009]).
@@
 
-## Applied example using two sample MR
+### Applied example using two sample MR
 
@@
 All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [https://github.com/explodecomputer/causal-directions](https://github.com/explodecomputer/causal-directions) and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].
-
-## Results
-
-### Mediation-based causal inference under measurement error
-
-In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$.
-When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (S1 Text) that
-
-$$
-\begin{aligned}
-cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
-                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
-\end{aligned}
-$$
-
-where
-
-$$
-D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
-$$
-
-Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$.
-Therefore, if there is any measurement error that incurs imprecision in $x$ (i.e.
-$var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x_o$, which is in violation of the the 4th condition of the CIT.
-Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.
-
-We performed simulations to verify that this problem does arise using the CIT method.
-`r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}=1$) the CIT is reliable in identifying the correct causal direction.
-However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction.
-Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.
-
-### Using MR Steiger to infer the direction of causality
-
-If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction.
-Though bi-directional MR can be used to orient causal directions [@Timpson2011], this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood.
-We go on to explore the scenarios in which the method is likely to return the correct or incorrect causal directions.
-
-We performed simulations to compare the power and type 1 error rates of MR and CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure.
-Comparing the performance of methods with different sets of assumptions can be difficult, but a basic comparison is shown in `r cfig("causality_exists_tpr")`.
-We observe that the CIT is more conservative under the null model of no association owing to the omnibus test statistic comprising several statistical tests.
-The FDR using a p-value threshold of 0.05 appears to be close to zero, whereas for the MR Steiger method the FDR is around 0.05.
-Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases more steeply than that of the MR Steiger method.
-
-For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, S2 Text).
-This metric can be used to evaluate the reliability of MR Steiger test.
-
-We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (S2 Text) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction).
-`r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$.
-It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the MR Steiger test to measurement error is relatively low.
-
-Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (S3 Text).
-The relationship between unmeasured confounding and causal orientation is complex across the parameter space of possible confounding values (S2 Fig).
-Based on the range of parameter values that we explored, when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction due to unmeasured confounding.
-
-### Comparison of CIT and MR Steiger for obtaining the correct direction of causality
-
-We used simulations to explore the performance of the MR Steiger approach in comparison to CIT for different levels of measurement error.
-The performance was compared in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect.
-Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing $x$ and $y$ independently.
-
-`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", the MR analysis is indeed liable to infer the wrong direction of causality when $d < 0$, and that this erroneous result is more likely to occur with increasing sample size.
-However, the CIT is in general more fallible to reporting a robust causal association for the wrong direction of causality.
-When $d > 0$ we find that in most cases the MR Steiger method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
-The CIT, unlike the MR Steiger test, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to identify the causal model as true, when in fact the underlying model is this non-causal model.
-
-### The causal relationship between gene expression and DNA methylation levels
-
-We used the MR Steiger test to infer the direction of causality between DNA methylation and gene expression levels between `r nrow(qtl_orig)` putative associations.
-We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$).
-The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (5%-95% quantiles `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`).
-We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms.
-`r cfig("shakhplot")`a shows that the conclusions about the direction of causality between DNA methylation and gene expression are very sensitive to measurement error.
-We made a strong assumption that either methylation influenced gene expression or vice versa, but it is certainly possible that the SNP is solely or additionally influencing some other trait that confounds the association between gene expression and DNA methylation.
-
-We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test.
-We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b).
-There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`).
-Together these inferences suggest that even in estimating associations between 'omic' variables, which are considered to be low level phenotypes, it is important to use causal inference methods over observational associations to infer causal effect sizes.
-
-We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing methylation levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].
-
-## Discussion
-
-Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data.
-In the epidemiological literature issues of measurement error in mediation analysis are relatively well explored [@Cole2014].
-Our analysis extends this to related methods such as CIT that are used in predominantly 'omic data.
-These methods are indeed susceptible to the same problem as standard mediation based analysis, and specifically we show that as measurement error in the (true) exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality.
-We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.
-
-Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation.
-Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces.
-This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest.
-But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality.
-Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods.
-We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages.
-First, it uses a formal statistical framework to test for the reliability of the assumed direction of causality.
-Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.
-
-We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data.
-The inferred causal direction is heavily influenced by how much measurement error is present in the different assaying platforms.
-For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association.
-Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.
-We also haven't accounted for the influence of winner's curse, which can inflate estimates of the variance explained by SNPs, with higher inflation expected amongst lower powered studies.
-Using p-values for genetic associations from replication studies will mitigate this problem.
-
-In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available.
-However, if there is only a single instrument it is difficult to separate between the two competing models of $g$ instrumenting a trait which causes another trait, and $g$ having pleiotropic effects on both traits independently [@Zhu2016].
-Under certain conditions of measurement error the CIT test can distinguish these models.
-We also note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent.
-Multiple instruments can indeed help to distinguish between the causal and pleiotropic models, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014].
-Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017].
-We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].
-However, our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.
-
-Our new method attempts to infer causal directions under the assumption that horizontal pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) is not present.
-Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored.
-An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis.
-However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood.
-We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies.
-We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated.
-We have also not addressed other issues pertaining to instrumental variables which are relevant to the question of instrument-exposure specification.
-One such problem is exposure misspecification, for example an instrument could associate with several closely related putative outcomes, with only one of them actually having a causal effect on the outcome.
-This problem has shown to be the case for SNPs influencing different lipid fractions, for example [@Do2013; @Burgess2014a].
-
-Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction.
-But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error.
-Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015; @Richmond2016] may offer more robust alternatives for these more complicated problems.
-
-The overarching result from our simulations is that, regardless of the method used, inferring the causal direction using an instrument of unknown biology is highly sensitive to measurement error.
-With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.
 

### 2017-11-13 docs/manuscript-plos.tex
updated

--- 2017-10-24 docs/manuscript.tex
@@
   \postdate{\par}
-  \date{24 October 2017}
+  \date{13 November 2017}
 
@@
 
-\subsubsection{Abstract}\label{abstract}
+\section{Abstract}\label{abstract}
 
@@
 
-\subsubsection{Author summary}\label{author-summary}
+\section{Author summary}\label{author-summary}
 
@@
 
-\subsection{Introduction}\label{introduction}
+\section{Introduction}\label{introduction}
 
@@
 
-\subsection{Methods}\label{methods}
-
-\subsubsection{CIT test}\label{cit-test}
-
-First we describe how the CIT method {[}4{]} is implemented in the R
-package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
-Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
-exposure \(x\) causes an outcome \(y\), as described above. The
-following tests are then performed:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
-  associates with the outcome}
-\item
-  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
-  associates with the exposure conditional on the outcome}
-\item
-  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
-  associates with the outcome conditional on the SNP}
-\item
-  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
-  independent of the outcome conditional on the exposure}
-\end{enumerate}
-
-The term in the 4th test can be rewritten as a
-\(cov(g, y|x) = cov(g, y - \hat{y})\) where
-\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
-of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
-association between the SNP and the outcome. The condition in the 4th
-test is formulated as an equivalence testing problem that is estimated
-using simulations, comparing the estimate from the data against
-empirically obtained estimates for simulated variables where the
-independence model is true (full details are given in {[}4{]}). We note
-here that this approach is liable to fail, even when there is a true
-causal relationship, when confounders of the exposure and outcome are
-present, as these will induce collider bias.
-
-If all four tests reject the null hypothesis then it is inferred that
-\(x\) causes \(y\). The CIT measures the strength of causality by
-generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
-(least extreme) p-value of the four tests, the intuition being that
-causal inference is only as strong as the weakest link in the chain of
-tests.
-
-Now we describe how we used the CIT method in our simulations. The
-\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
-the direction of causality using the CIT method, an omnibus p-value
-generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
-was estimated for the direction of \(x\) causing \(y\) (Model 1), and
-for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
-(Model 2). The results from each of these methods can then be used in
-combination to infer the existance and direction of causality. For some
-significance threshold \(\alpha\) there are four possible outcomes from
-these two tests, and their interpretations are as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
-  relationship
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
-  confounding (S1 fig) and no call is made.
-\end{itemize}
-
-For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value, though we stress that for real analyses it is
-not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds {[}39{]}.
-
-\subsubsection{MR causal test}\label{mr-causal-test}
-
-Two stage least squares (2SLS) is a commonly used technique for
-performing MR when the exposure, outcome and instrument data are all
-available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the \(systemfit\) function in the R package
-\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
-when using the same genetic variant to instrument the influence of the
-exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
-outcome \(y\) on the exposure \(x\).
-
-The method that we will now describe is designed to distinguish between
-two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
-framework, this approach cannot infer if the true model is
-\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
-additive.
-
-To infer the direction of causality it is desirable to know which of the
-variables, \(x\) or \(y\), is being directly influenced by the
-instrument \(g\). This can be achieved by assessing which of the two
-variables has the biggest absolute correlation with \(g\) (S2 Text),
-formalised by testing for a difference in the correlations \(\rho_{gx}\)
-and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population {[}41{]}. It is calculated as
-
-\[
-Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
-\]
-
-where Fisher's z-transformation is used to obtain
-\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
-
-\[
-h = \frac{1 - (frm^2)} {1 - rm^2}
-\]
-
-where
-
-\[
-f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
-\]
-
-and
-
-\[
-rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
-\]
-
-The \(Z\) value is interpreted such that
-
-\[
-Z \left\{
-\begin{array}{ll}
-> 0, & x \to y\\
-< 0, & y \to x\\
-= 0, & x \perp\!\!\!\perp y 
-\end{array} \right.
-\]
-
-and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
-indicate the probability of obtaining a difference between correlations
-\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
-under the null hypothesis that both correlations are identical.
-
-The existence of causality and its direction is inferred based on
-combining information from the MR analysis and the Steiger test. The MR
-analysis indicates whether there is a potential causal relationship
-(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
-of the causal relationship and the confidence of the direction
-(\(p_{Steiger}\)). For the purposes of compiling simulation results,
-these can be combined using an arbitrary \(\alpha = 0.05\) value:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
-  a causal association for the correct model is accepted,
-  \(x \rightarrow y\)
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
-  a causal association for the incorrect model is accepted,
-  \(y \rightarrow x\)
-\item
-  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
-  model is accepted
-\end{itemize}
-
-Note that the same correlation test approach can be applied to a
-two-sample MR setting. Two-sample MR refers to the case where the
-SNP-exposure association and SNP-outcome association are calculated in
-different samples (e.g.~from publicly available summary statistics
-{[}26,30{]}). Here the Steiger test of two independent correlations can
-be applied where.
-
-\[
-Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
-\]
-
-An advantage of using the Steiger test in the two sample context is that
-it can compare correlations in independent samples where sample sizes
-are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.
-
-The Steiger test assumes that there is a causal relationship between the
-two variables, and that the SNP is a valid instrument for one of them.
-However it is liable to give incorrect causal directions under some
-other circumstances. First, some levels of horizontal pleiotropy, where
-the SNP influences the outcome through some pathway other than the
-exposure, could induce problems because this is a means by which the
-instrument is invalid. Second, some differential values of measurement
-error between the exposure and the outcome could lead to incorrect
-inference of the causal direction (S2 Text). Third, some levels of
-unmeasured confounding between the exposure and the outcome could lead
-to inference of the wrong causal direction (S3 Text).
-
-\subsubsection{Causal direction sensitivity analysis for measurement
-error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
-
-The Steiger test for inferring if \(x \rightarrow y\) is based on
-evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
-\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
-imprecisely. If, for example, \(x\) has lower measurement precision than
-\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
-because \(\rho_{g,x_o}\) could be underestimated more than
-\(\rho_{g,y_o}\).
-
-As we show in S2 Text it is possible to infer the bounds of measurement
-error on \(x_o\) or \(y_o\) given known genetic associations. The
-maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
-it is known that at least that much of the variance has been explained
-in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
-values (the same logic applies to \(y_o\)). It is possible to simulate
-what the inferred causal direction would be for all values within these
-bounds.
-
-To evaluate how reliable, \(R\), the inference of the causal direction
-is to potential measurement error in \(x\) and \(y\) we need to predict
-the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
-error. We offer two tools in which to do this. First, the user can
-provide values of measurement error for \(x\) and \(y\) and obtain a
-revised inference of the causal direction. Second, we integrate over the
-entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
-measurement error values, assuming that any measurement error value is
-equally likely. Across all possible values of measurement error in \(x\)
-and \(y\) we find the volume that agrees with the inferred direction of
-causality and the volume that disagrees with the inferred direction of
-causality, and take the ratio of these two values. A ratio \(R=1\)
-indicates that the inferred causal direction is highly sensitive to
-measurement error, because equal weight of the measurement error
-parameter space supports each direction of causality. In general, the
-\(R\) value denotes that the inferred direction of causality is \(R\)
-times more likely to be the empirical result than the opposite direction
-(S2 Text).
-
-\subsubsection{Simulations}\label{simulations}
-
-Simulations were conducted by creating variables of sample size \(n\)
-for the exposure \(x\), the measured values of the exposure \(x_o\), the
-outcome \(y\), the measured values of the outcome \(y_o\) and the
-instrument \(g\). One of two models are simulated, the ``causal model''
-where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
-``non-causal model'' where \(g\) influences a confounder \(u\) which in
-turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
-but not causally related. Each variable in the causal model was
-simulated such that:
-
-\[
-\begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
-\end{aligned}
-\]
-
-where non-differential measurement error is represented by a noise
-(measurement imprecision) term
-\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
-\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
-the outcome variable \(y\). Note that following the first section of the
-Results we no longer include the bias terms for simplicity. We have
-formulated the non-causal model as:
-
-\[
-\begin{aligned}
-y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
-\end{aligned}
-\]
-
-All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
-Normally distributed values of \(\epsilon_*\) were generated such that
-
-\[
-\begin{aligned}
-cor(g, x)^2 & = 0.1 \\
-cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
-\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-n & = \{100, 1000, 10000\}
-\end{aligned}
-\]
-
-giving a total of 432 combinations of parameters. Simulations using each
-of these sets of variables were performed 100 times, and the CIT and MR
-methods were applied to each in order to evaluate the causal association
-of the simulated variables. Similar patterns of results were obtained
-for different values of \(cor(g, x)\).
-
-\subsection{Applied example using two sample
-MR}\label{applied-example-using-two-sample-mr}
-
-Two sample MR {[}30{]} was performed using summary statistics for
-genetic influences on gene expression and DNA methylation. To do this we
-obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al {[}43{]}. These were filtered to be located
-on the same chromosome, have robust correlations after correcting for
-multiple testing, and to share a SNP that had a robust cis-acting effect
-on both the DNA methylation probe and the gene expression probe. Because
-only summary statistics were available (effect, standard error, effect
-allele, sample size, p-values) for the instrumental SNP on the
-methylation and gene expression levels, the Steiger test of two
-independent correlations was used to infer the direction of causality
-for each of the associations. The Wald ratio test was then used to
-estimate the causal effect size for the estimated direction for each
-association.
-
-All analysis was performed using the R programming language {[}44{]} and
-code is made available at
-\url{https://github.com/explodecomputer/causal-directions} and
-implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
-{[}26{]}.
-
-\subsection{Results}\label{results}
-
-\subsubsection{Mediation-based causal inference under measurement
+\section{Results}\label{results}
+
+\subsection{Mediation-based causal inference under measurement
 error}\label{mediation-based-causal-inference-under-measurement-error}
@@
 
-\subsubsection{Using MR Steiger to infer the direction of
+\subsection{Using MR Steiger to infer the direction of
 causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}
@@
 
-\subsubsection{Comparison of CIT and MR Steiger for obtaining the
+\subsection{Comparison of CIT and MR Steiger for obtaining the
 correct direction of
@@
 
-\subsubsection{The causal relationship between gene expression and DNA
+\subsection{The causal relationship between gene expression and DNA
 methylation
@@
 the sign of the MR estimate was generally in the same direction as the
-Pearson correlation coefficient reported by Shakhbazov et al {[}43{]}
+Pearson correlation coefficient reported by Shakhbazov et al {[}39{]}
 (Figure 6b). There was a moderate correlation between the absolute
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation {[}45{]}.
-
-\subsection{Discussion}\label{discussion}
+levels, consistent with expectation {[}40{]}.
+
+\section{Discussion}\label{discussion}
 
@@
 epidemiological literature issues of measurement error in mediation
-analysis are relatively well explored {[}46{]}. Our analysis extends
+analysis are relatively well explored {[}41{]}. Our analysis extends
 this to related methods such as CIT that are used in predominantly 'omic
@@
 which have evaluated measurement error in these platforms do support
-this position {[}47,48{]}, though making strong conclusions for this
+this position {[}42,43{]}, though making strong conclusions for this
 analysis is difficult because measurement error is likely to be study
@@
 another trait, and \(g\) having pleiotropic effects on both traits
-independently {[}49{]}. Under certain conditions of measurement error
+independently {[}44{]}. Under certain conditions of measurement error
 the CIT test can distinguish these models. We also note that it is
@@
 bi-directional MR can offer solutions to inferring the causal direction
-{[}16,28,50{]}. We restricted the simulations to evaluating the causal
+{[}16,28,45{]}. We restricted the simulations to evaluating the causal
 inference between quantitative traits, but it is possible that the
@@
 variance explained on the liability scale, taking into account the
-population prevalence {[}51{]}. However, our analysis goes beyond many
+population prevalence {[}46{]}. However, our analysis goes beyond many
 previous explorations of measurement error by assessing the impacts of
@@
 on the outcome. This problem has shown to be the case for SNPs
-influencing different lipid fractions, for example {[}52,53{]}.
+influencing different lipid fractions, for example {[}47,48{]}.
 
@@
 of measurement error. Recent advances in MR methodology, for example
-applying MR to genetical genomics {[}54{]}, multivariate MR {[}53{]} and
-mediation through MR {[}55--57{]} may offer more robust alternatives for
+applying MR to genetical genomics {[}49{]}, multivariate MR {[}48{]} and
+mediation through MR {[}50--52{]} may offer more robust alternatives for
 these more complicated problems.
@@
 
+\section{Methods}\label{methods}
+
+\subsection{CIT test}\label{cit-test}
+
+Here we describe how the CIT method {[}4{]} is implemented in the R
+package \emph{R/cit} {[}18{]}. Assume an exposure \(x\) is instrumented
+by a SNP \(g\), and the exposure \(x\) causes an outcome \(y\), as
+described above. The following tests are then performed:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
+  associates with the outcome}
+\item
+  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
+  associates with the exposure conditional on the outcome}
+\item
+  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
+  associates with the outcome conditional on the SNP}
+\item
+  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
+  independent of the outcome conditional on the exposure}
+\end{enumerate}
+
+The term in the 4th test can be rewritten as
+\(cov(g, y|x) = cov(g, y - \hat{y})\) where
+\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
+of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
+association between the SNP and the outcome. The condition in the 4th
+test is formulated as an equivalence testing problem that is estimated
+using simulations, comparing the estimate from the data against
+empirically obtained estimates for simulated variables where the
+independence model is true (full details are given in {[}4{]}). We note
+here that this approach is liable to fail, even when there is a true
+causal relationship, when confounders of the exposure and outcome are
+present, as these will induce collider bias.
+
+If all four tests reject the null hypothesis then it is inferred that
+\(x\) causes \(y\). The CIT measures the strength of causality by
+generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
+(least extreme) p-value of the four tests, the intuition being that
+causal inference is only as strong as the weakest link in the chain of
+tests.
+
+Now we describe how we used the CIT method in our simulations. The
+\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
+the direction of causality using the CIT method, an omnibus p-value
+generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
+was estimated for the direction of \(x\) causing \(y\) (Model 1), and
+for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
+(Model 2). The results from each of these methods can then be used in
+combination to infer the existance and direction of causality. For some
+significance threshold \(\alpha\) there are four possible outcomes from
+these two tests, and their interpretations are as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
+  relationship
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
+  confounding (S1 fig) and no call is made.
+\end{itemize}
+
+For the purposes of compiling simulation results we use an arbitrary
+\(\alpha = 0.05\) value, though we stress that for real analyses it is
+not good practice to rely on p-values for making causal inference, nor
+is it reliable to depend on arbitrary significance thresholds {[}53{]}.
+
+\subsection{MR causal test}\label{mr-causal-test}
+
+Two stage least squares (2SLS) is a commonly used technique for
+performing MR when the exposure, outcome and instrument data are all
+available in the same sample. A p-value for this test, \(p_{MR}\), was
+obtained using the \(systemfit\) function in the R package
+\(R/systemfit\) {[}54{]}. Note that the value of \(p_{MR}\) is identical
+when using the same genetic variant to instrument the influence of the
+exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
+outcome \(y\) on the exposure \(x\).
+
+The method that we will now describe is designed to distinguish between
+two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
+framework, this approach cannot infer if the true model is
+\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
+additive.
+
+To infer the direction of causality it is desirable to know which of the
+variables, \(x\) or \(y\), is being directly influenced by the
+instrument \(g\). This can be achieved by assessing which of the two
+variables has the biggest absolute correlation with \(g\) (S2 Text),
+formalised by testing for a difference in the correlations \(\rho_{gx}\)
+and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
+within a population {[}55{]}. It is calculated as
+
+\[
+Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
+\]
+
+where Fisher's z-transformation is used to obtain
+\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
+
+\[
+h = \frac{1 - (frm^2)} {1 - rm^2}
+\]
+
+where
+
+\[
+f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
+\]
+
+and
+
+\[
+rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
+\]
+
+The \(Z\) value is interpreted such that
+
+\[
+Z \left\{
+\begin{array}{ll}
+> 0, & x \to y\\
+< 0, & y \to x\\
+= 0, & x \perp\!\!\!\perp y 
+\end{array} \right.
+\]
+
+and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
+indicate the probability of obtaining a difference between correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
+under the null hypothesis that both correlations are identical.
+
+The existence of causality and its direction is inferred based on
+combining information from the MR analysis and the Steiger test. The MR
+analysis indicates whether there is a potential causal relationship
+(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
+of the causal relationship and the confidence of the direction
+(\(p_{Steiger}\)). For the purposes of compiling simulation results,
+these can be combined using an arbitrary \(\alpha = 0.05\) value:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
+  a causal association for the correct model is accepted,
+  \(x \rightarrow y\)
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
+  a causal association for the incorrect model is accepted,
+  \(y \rightarrow x\)
+\item
+  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
+  model is accepted
+\end{itemize}
+
+Note that the same correlation test approach can be applied to a
+two-sample MR setting. Two-sample MR refers to the case where the
+SNP-exposure association and SNP-outcome association are calculated in
+different samples (e.g.~from publicly available summary statistics
+{[}26,30{]}). Here the Steiger test of two independent correlations can
+be applied where.
+
+\[
+Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
+\]
+
+An advantage of using the Steiger test in the two sample context is that
+it can compare correlations in independent samples where sample sizes
+are different. Steiger test statistics were calculated using the
+\emph{r.test} function in the R package \emph{R/psych} {[}56{]}.
+
+The Steiger test assumes that there is a causal relationship between the
+two variables, and that the SNP is a valid instrument for one of them.
+However it is liable to give incorrect causal directions under some
+other circumstances. First, some levels of horizontal pleiotropy, where
+the SNP influences the outcome through some pathway other than the
+exposure, could induce problems because this is a means by which the
+instrument is invalid. Second, some differential values of measurement
+error between the exposure and the outcome could lead to incorrect
+inference of the causal direction (S2 Text). Third, some levels of
+unmeasured confounding between the exposure and the outcome could lead
+to inference of the wrong causal direction (S3 Text).
+
+\subsection{Causal direction sensitivity analysis for measurement
+error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
+
+The Steiger test for inferring if \(x \rightarrow y\) is based on
+evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
+\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
+imprecisely. If, for example, \(x\) has lower measurement precision than
+\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
+because \(\rho_{g,x_o}\) could be underestimated more than
+\(\rho_{g,y_o}\).
+
+As we show in S2 Text it is possible to infer the bounds of measurement
+error on \(x_o\) or \(y_o\) given known genetic associations. The
+maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
+it is known that at least that much of the variance has been explained
+in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
+values (the same logic applies to \(y_o\)). It is possible to simulate
+what the inferred causal direction would be for all values within these
+bounds.
+
+To evaluate how reliable, \(R\), the inference of the causal direction
+is to potential measurement error in \(x\) and \(y\) we need to predict
+the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
+error. We offer two tools in which to do this. First, the user can
+provide values of measurement error for \(x\) and \(y\) and obtain a
+revised inference of the causal direction. Second, we integrate over the
+entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
+measurement error values, assuming that any measurement error value is
+equally likely. Across all possible values of measurement error in \(x\)
+and \(y\) we find the volume that agrees with the inferred direction of
+causality and the volume that disagrees with the inferred direction of
+causality, and take the ratio of these two values. A ratio \(R=1\)
+indicates that the inferred causal direction is highly sensitive to
+measurement error, because equal weight of the measurement error
+parameter space supports each direction of causality. In general, the
+\(R\) value denotes that the inferred direction of causality is \(R\)
+times more likely to be the empirical result than the opposite direction
+(S2 Text).
+
+\subsection{Simulations}\label{simulations}
+
+Simulations were conducted by creating variables of sample size \(n\)
+for the exposure \(x\), the measured values of the exposure \(x_o\), the
+outcome \(y\), the measured values of the outcome \(y_o\) and the
+instrument \(g\). One of two models are simulated, the ``causal model''
+where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
+``non-causal model'' where \(g\) influences a confounder \(u\) which in
+turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
+but not causally related. Each variable in the causal model was
+simulated such that:
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+\]
+
+where non-differential measurement error is represented by a noise
+(measurement imprecision) term
+\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
+\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
+the outcome variable \(y\). Note that following the first section of the
+Results we no longer include the bias terms for simplicity. We have
+formulated the non-causal model as:
+
+\[
+\begin{aligned}
+y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
+\end{aligned}
+\]
+
+All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
+Normally distributed values of \(\epsilon_*\) were generated such that
+
+\[
+\begin{aligned}
+cor(g, x)^2 & = 0.1 \\
+cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
+\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+n & = \{100, 1000, 10000\}
+\end{aligned}
+\]
+
+giving a total of 432 combinations of parameters. Simulations using each
+of these sets of variables were performed 100 times, and the CIT and MR
+methods were applied to each in order to evaluate the causal association
+of the simulated variables. Similar patterns of results were obtained
+for different values of \(cor(g, x)\).
+
+\subsection{Applied example using two sample
+MR}\label{applied-example-using-two-sample-mr}
+
+Two sample MR {[}30{]} was performed using summary statistics for
+genetic influences on gene expression and DNA methylation. To do this we
+obtained a list of 458 gene expression - DNA methylation associations as
+reported in Shakhbazov et al {[}39{]}. These were filtered to be located
+on the same chromosome, have robust correlations after correcting for
+multiple testing, and to share a SNP that had a robust cis-acting effect
+on both the DNA methylation probe and the gene expression probe. Because
+only summary statistics were available (effect, standard error, effect
+allele, sample size, p-values) for the instrumental SNP on the
+methylation and gene expression levels, the Steiger test of two
+independent correlations was used to infer the direction of causality
+for each of the associations. The Wald ratio test was then used to
+estimate the causal effect size for the estimated direction for each
+association.
+
+All analysis was performed using the R programming language {[}57{]} and
+code is made available at
+\url{https://github.com/explodecomputer/causal-directions} and
+implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
+{[}26{]}.
+
+\subsection{References}\label{references}
+
+\raggedright
+
+\hypertarget{refs}{}
+\hypertarget{ref-Phillips1991}{}
+1. Phillips AN, Davey Smith G. How independent are ``independent''
+effects? relative risk estimation when correlated exposures are measured
+imprecisely. Journal of Clinical Epidemiology. Pergamon; 1991;44:
+1223--1231.
+doi:\href{https://doi.org/10.1016/0895-4356(91)90155-3}{10.1016/0895-4356(91)90155-3}
+
+\hypertarget{ref-DaveySmith2002}{}
+2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
+2002;325: 1437--8.
+
+\hypertarget{ref-DaveySmith2004}{}
+3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
+potentials, and limitations. International journal of epidemiology.
+2004;33: 30--42.
+doi:\href{https://doi.org/10.1093/ije/dyh132}{10.1093/ije/dyh132}
+
+\hypertarget{ref-Millstein2009}{}
+4. Millstein J, Zhang B, Zhu J, Schadt EE. Disentangling molecular
+relationships with a causal inference test. BMC genetics. 2009;10: 23.
+doi:\href{https://doi.org/10.1186/1471-2156-10-23}{10.1186/1471-2156-10-23}
+
+\hypertarget{ref-Aten2008}{}
+5. Aten JE, Fuller TF, Lusis AJ, Horvath S. Using genetic markers to
+orient the edges in quantitative trait networks: the NEO software. BMC
+systems biology. 2008;2: 34.
+doi:\href{https://doi.org/10.1186/1752-0509-2-34}{10.1186/1752-0509-2-34}
+
+\hypertarget{ref-Waszak2015}{}
+6. Waszak SM, Delaneau O, Gschwind AR, Kilpinen H, Raghav SK, Witwicki
+RM, et al. Variation and genetic control of chromatin architecture in
+humans. Cell. Elsevier Inc. 2015;162: 1039--1050.
+doi:\href{https://doi.org/10.1016/j.cell.2015.08.001}{10.1016/j.cell.2015.08.001}
+
+\hypertarget{ref-Houle2011}{}
+7. Houle D, Pélabon C, Wagner G, Hansen T. Measurement and meaning in
+biology. The Quarterly Review of Biology. 2011;86: 3--34. Available:
+\url{http://www.jstor.org/stable/10.1086/658408}
+
+\hypertarget{ref-Hernan2009}{}
+8. Hernán M a, Cole SR. Invited Commentary: Causal diagrams and
+measurement bias. American journal of epidemiology. 2009;170: 959--62;
+discussion 963--4.
+doi:\href{https://doi.org/10.1093/aje/kwp293}{10.1093/aje/kwp293}
+
+\hypertarget{ref-Harper2013}{}
+9. Harper KN, Peters B a, Gamble MV. Batch effects and pathway analysis:
+two potential perils in cancer studies involving DNA methylation array
+analysis. Cancer epidemiology, biomarkers \& prevention : a publication
+of the American Association for Cancer Research, cosponsored by the
+American Society of Preventive Oncology. 2013;22: 1052--60.
+doi:\href{https://doi.org/10.1158/1055-9965.EPI-13-0114}{10.1158/1055-9965.EPI-13-0114}
+
+\hypertarget{ref-Chen2013a}{}
+10. Chen Y-a, Lemire M, Choufani S, Butcher DT, Grafodatskaya D, Zanke
+BW, et al. Discovery of cross-reactive probes and polymorphic CpGs in
+the Illumina Infinium HumanMethylation450 microarray. Epigenetics :
+official journal of the DNA Methylation Society. 2013;8: 203--9.
+doi:\href{https://doi.org/10.4161/epi.23470}{10.4161/epi.23470}
+
+\hypertarget{ref-Houseman2012}{}
+11. Houseman EA, Accomando WP, Koestler DC, Christensen BC, Marsit CJ,
+Nelson HH, et al. DNA methylation arrays as surrogate measures of cell
+mixture distribution. BMC bioinformatics. 2012;13: 86.
+doi:\href{https://doi.org/10.1186/1471-2105-13-86}{10.1186/1471-2105-13-86}
+
+\hypertarget{ref-Ahima2013}{}
+12. Ahima RS, Lazar MA. Physiology. The health risk of obesity--better
+metrics imperative. Science (New York, NY). American Association for the
+Advancement of Science; 2013;341: 856--8.
+doi:\href{https://doi.org/10.1126/science.1241244}{10.1126/science.1241244}
+
+\hypertarget{ref-LeCessie2012}{}
+13. Cessie S le, Debeij J, Rosendaal FR, Cannegieter SC, Vandenbroucke
+JP. Quantification of bias in direct effects estimates due to different
+types of measurement error in the mediator. Epidemiology (Cambridge,
+Mass). 2012;23: 551--60.
+doi:\href{https://doi.org/10.1097/EDE.0b013e318254f5de}{10.1097/EDE.0b013e318254f5de}
+
+\hypertarget{ref-Blakely2013}{}
+14. Blakely T, McKenzie S, Carter K. Misclassification of the mediator
+matters when estimating indirect effects. Journal of epidemiology and
+community health. 2013;67: 458--66.
+doi:\href{https://doi.org/10.1136/jech-2012-201813}{10.1136/jech-2012-201813}
+
+\hypertarget{ref-DaveySmith2003}{}
+15. Davey Smith G, Ebrahim S. 'Mendelian randomization': can genetic
+epidemiology contribute to understanding environmental determinants of
+disease? International Journal of Epidemiology. 2003;32: 1--22.
+doi:\href{https://doi.org/10.1093/ije/dyg070}{10.1093/ije/dyg070}
+
+\hypertarget{ref-DaveySmithHemani2014}{}
+16. Davey Smith G, Hemani G. Mendelian randomization: genetic anchors
+for causal inference in epidemiological studies. Human molecular
+genetics. Oxford Univ Press; 2014;23: R89-----R98.
+doi:\href{https://doi.org/10.1093/hmg/ddu328}{10.1093/hmg/ddu328}
+
+\hypertarget{ref-Schadt2005}{}
+17. Schadt EE, Lamb J, Yang X, Zhu J, Edwards S, GuhaThakurta D, et al.
+An integrative genomics approach to infer causal associations between
+gene expression and disease. Nature Genetics. 2005;37: 710--717.
+doi:\href{https://doi.org/10.1038/ng1589}{10.1038/ng1589}
+
+\hypertarget{ref-Millstein2016}{}
+18. Millstein J. cit: Causal Inference Test. R package version 1.9
+{[}Internet{]}. 2016. Available:
+\url{http://cran.r-project.org/package=cit}
+
+\hypertarget{ref-Koestler2014}{}
+19. Koestler DC, Chalise P, Cicek MS, Cunningham JM, Armasu S, Larson
+MC, et al. Integrative genomic analysis identifies epigenetic marks that
+mediate genetic risk for epithelial ovarian cancer. BMC medical
+genomics. BMC Medical Genomics; 2014;7: 8.
+doi:\href{https://doi.org/10.1186/1755-8794-7-8}{10.1186/1755-8794-7-8}
+
+\hypertarget{ref-Liu2013}{}
+20. Liu Y, Aryee MJ, Padyukov L, Fallin MD, Hesselberg E, Runarsson A,
+et al. Epigenome-wide association data implicate DNA methylation as an
+intermediary of genetic risk in rheumatoid arthritis. Nature
+biotechnology. 2013;31: 142--7.
+doi:\href{https://doi.org/10.1038/nbt.2487}{10.1038/nbt.2487}
+
+\hypertarget{ref-Yuan2014}{}
+21. Yuan W, Xia Y, Bell CG, Yet I, Ferreira T, Ward KJ, et al. An
+integrated epigenomic analysis for type 2 diabetes susceptibility loci
+in monozygotic twins. Nature communications. 2014;5: 5719.
+doi:\href{https://doi.org/10.1038/ncomms6719}{10.1038/ncomms6719}
+
+\hypertarget{ref-Tang2014}{}
+22. Tang Y, Axelsson AS, Spégel P, Andersson LE, Mulder H, Groop LC, et
+al. Genotype-based treatment of type 2 diabetes with an
+\(\alpha\)2A-adrenergic receptor antagonist. Science translational
+medicine. 2014;6: 257ra139.
+doi:\href{https://doi.org/10.1126/scitranslmed.3009934}{10.1126/scitranslmed.3009934}
+
+\hypertarget{ref-Hong2015}{}
+23. Hong X, Hao K, Ladd-Acosta C, Hansen KD, Tsai H-J, Liu X, et al.
+Genome-wide association study identifies peanut allergy-specific loci
+and evidence of epigenetic mediation in US children. Nature
+communications. 2015;6: 6304.
+doi:\href{https://doi.org/10.1038/ncomms7304}{10.1038/ncomms7304}
+
+\hypertarget{ref-Bowden2015}{}
+24. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
+invalid instruments: effect estimation and bias detection through Egger
+regression. International Journal of Epidemiology. 2015;44: 512--25.
+
+\hypertarget{ref-Bowden2016b}{}
+25. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent
+Estimation in Mendelian Randomization with Some Invalid Instruments
+Using a Weighted Median Estimator. Genetic Epidemiology. 2016;40:
+304--314.
+doi:\href{https://doi.org/10.1002/gepi.21965}{10.1002/gepi.21965}
+
+\hypertarget{ref-Hemani2016}{}
+26. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
+MR-Base: a platform for systematic causal inference across the phenome
+using billions of genetic associations. BioRxiv. 2016;10.1101/07.
+
+\hypertarget{ref-Timpson2011}{}
+27. Timpson NJ, Nordestgaard BG, Harbord RM, Zacho J, Frayling TM,
+Tybjærg-Hansen a, et al. C-reactive protein levels and body mass index:
+elucidating direction of causation through reciprocal Mendelian
+randomization. International journal of obesity (2005). 2011;35: 300--8.
+doi:\href{https://doi.org/10.1038/ijo.2010.137}{10.1038/ijo.2010.137}
+
+\hypertarget{ref-Richmond2014}{}
+28. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
+NJ. Assessing Causality in the Association between Child Adiposity and
+Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
+editor. PLoS Medicine. 2014;11: e1001618.
+doi:\href{https://doi.org/10.1371/journal.pmed.1001618}{10.1371/journal.pmed.1001618}
+
+\hypertarget{ref-Claussnitzer2015}{}
+29. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
+al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
+New England journal of medicine. 2015;373: 895--907.
+doi:\href{https://doi.org/10.1056/NEJMoa1502214}{10.1056/NEJMoa1502214}
+
+\hypertarget{ref-Pierce2013}{}
+30. Pierce BL, Burgess S. Efficient design for Mendelian randomization
+studies: subsample and 2-sample instrumental variable estimators.
+American journal of epidemiology. 2013;178: 1177--84.
+doi:\href{https://doi.org/10.1093/aje/kwt084}{10.1093/aje/kwt084}
+
+\hypertarget{ref-Hernan2004}{}
+31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
+selection bias. Epidemiology (Cambridge, Mass). 2004;15: 615--25.
+Available: \url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}
+
+\hypertarget{ref-Pierce2012}{}
+32. Pierce BL, VanderWeele TJ. The effect of non-differential
+measurement error on bias, precision and power in Mendelian
+randomization studies. International Journal of Epidemiology. Oxford
+University Press; 2012;41: 1383--1393.
+doi:\href{https://doi.org/10.1093/ije/dys141}{10.1093/ije/dys141}
+
+\hypertarget{ref-Ashenfelter1994}{}
+33. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
+Schooling from a New Sample of Twins. The American Economic Review.
+1994;84: 1157--1173.
+doi:\href{https://doi.org/10.1016/S0272-7757(98)00038-7}{10.1016/S0272-7757(98)00038-7}
+
+\hypertarget{ref-Nagarajan2013}{}
+34. Nagarajan R, Scutari M. Impact of noise on molecular network
+inference. PloS one. 2013;8: e80735.
+doi:\href{https://doi.org/10.1371/journal.pone.0080735}{10.1371/journal.pone.0080735}
+
+\hypertarget{ref-Shpitser2010}{}
+35. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
+adjustment for estimating causal effects. Proceedings of the Twenty
+Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
+2010; 527--536.
+
+\hypertarget{ref-Wang2015}{}
+36. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
+networks. arXiv. 2015;arXiv:1512. Available:
+\url{http://arxiv.org/abs/1512.05574}
+
+\hypertarget{ref-Lagani2015}{}
+37. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
+Probabilistic Computational Causal Discovery for Systems Biology.
+Uncertainty in biology: A computational modeling approach. Springer;
+2015. p. 47. Available:
+\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}
+
+\hypertarget{ref-Lawlor2017}{}
+38. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
+epidemiology. International Journal of Epidemiology. 2017;19: dyw314.
+doi:\href{https://doi.org/10.1093/ije/dyw314}{10.1093/ije/dyw314}
+
+\hypertarget{ref-Shakhbazov2016}{}
+39. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+PM, et al. Shared genetic control of expression and methylation in
+peripheral blood. BMC genomics. BioMed Central; 2016;17: 278.
+doi:\href{https://doi.org/10.1186/s12864-016-2498-4}{10.1186/s12864-016-2498-4}
+
+\hypertarget{ref-Bird2002}{}
+40. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+development. 2002;16: 6--21.
+doi:\href{https://doi.org/10.1101/gad.947102}{10.1101/gad.947102}
+
+\hypertarget{ref-Cole2014}{}
+41. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+Serious and Misleading Consequences Due to Uncorrected Measurement
+Error. Psychological Methods. 2014;19: 300--315.
+doi:\href{https://doi.org/10.1037/a0033805}{10.1037/a0033805}
+
+\hypertarget{ref-Bose2014}{}
+42. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+Evaluation of microarray-based DNA methylation measurement using
+technical replicates: the Atherosclerosis Risk In Communities (ARIC)
+Study. BMC Bioinformatics. 2014;15: 312.
+doi:\href{https://doi.org/10.1186/1471-2105-15-312}{10.1186/1471-2105-15-312}
+
+\hypertarget{ref-Bryant2011}{}
+43. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+et al. Technical Variability Is Greater than Biological Variability in a
+Microarray Experiment but Both Are Outweighed by Changes Induced by
+Stimulation. Khodursky AB, editor. PLoS ONE. Public Library of Science;
+2011;6: e19556.
+doi:\href{https://doi.org/10.1371/journal.pone.0019556}{10.1371/journal.pone.0019556}
+
+\hypertarget{ref-Zhu2016}{}
+44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+Integration of summary data from GWAS and eQTL studies predicts complex
+trait gene targets. Nature Genetics. Nature Research; 2016;48: 481--487.
+doi:\href{https://doi.org/10.1038/ng.3538}{10.1038/ng.3538}
+
+\hypertarget{ref-Mancuso2017}{}
+45. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+Integrating Gene Expression with Summary Association Statistics to
+Identify Genes Associated with 30 Complex Traits. The American Journal
+of Human Genetics. 2017;100: 473--487.
+doi:\href{https://doi.org/10.1016/j.ajhg.2017.01.031}{10.1016/j.ajhg.2017.01.031}
+
+\hypertarget{ref-Lee2013c}{}
+46. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+association studies: quantification of power and genomic prediction
+accuracy. PLoS One. 2013;8: e71494.
+doi:\href{https://doi.org/10.1371/journal.pone.0071494\%20\%5Bdoi\%5D/rPONE-D-13-18776\%20\%5Bpii\%5D}{10.1371/journal.pone.0071494 {[}doi{]}\textbackslash{}rPONE-D-13-18776 {[}pii{]}}
+
+\hypertarget{ref-Do2013}{}
+47. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+Common variants associated with plasma triglycerides and risk for
+coronary artery disease. Nature Genetics. Nature Research; 2013;45:
+1345--1352. doi:\href{https://doi.org/10.1038/ng.2795}{10.1038/ng.2795}
+
+\hypertarget{ref-Burgess2014a}{}
+48. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+multivariable Mendelian randomization to disentangle the causal effects
+of lipid fractions. PloS one. Public Library of Science; 2014;9:
+e108891.
+doi:\href{https://doi.org/10.1371/journal.pone.0108891}{10.1371/journal.pone.0108891}
+
+\hypertarget{ref-Relton2012}{}
+49. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+randomization: a strategy for establishing the causal role of epigenetic
+processes in pathways to disease. International journal of epidemiology.
+2012;41: 161--76.
+doi:\href{https://doi.org/10.1093/ije/dyr233}{10.1093/ije/dyr233}
+
+\hypertarget{ref-Varbo2015}{}
+50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
+cholesterol, and blood pressure as mediators from obesity to ischemic
+heart disease. Circulation research. 2015;116: 665--73.
+doi:\href{https://doi.org/10.1161/CIRCRESAHA.116.304846}{10.1161/CIRCRESAHA.116.304846}
+
+\hypertarget{ref-Burgess2015}{}
+51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+randomization: using genetic variants as instrumental variables to
+investigate mediation in causal pathways. International journal of
+epidemiology. 2015;44: 484--95.
+doi:\href{https://doi.org/10.1093/ije/dyu176}{10.1093/ije/dyu176}
+
+\hypertarget{ref-Richmond2016}{}
+52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+Challenges and novel approaches for investigating molecular mediation.
+Human molecular genetics. Oxford University Press; 2016;25: R149--R156.
+doi:\href{https://doi.org/10.1093/hmg/ddw197}{10.1093/hmg/ddw197}
+
+\hypertarget{ref-Sterne2001}{}
+53. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+significance tests? BMJ. 2001;322: 226--231.
+
+\hypertarget{ref-Henningsen2007}{}
+54. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. Journal of Statistical Software.
+2007;23: 1--40.
+doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
+
+\hypertarget{ref-Steiger1980}{}
+55. Steiger JH. Tests for comparing elements of a correlation matrix.
+Psychological Bulletin. 1980;87: 245--251.
+doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
+
+\hypertarget{ref-Revelle2015}{}
+56. Revelle W. psych: Procedures for Psychological, Psychometric, and
+Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
+University; 2015. Available:
+\url{http://cran.r-project.org/package=psych}
+
+\hypertarget{ref-RCoreTeam2015}{}
+57. R Core Team. R: A Language and Environment for Statistical Computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2015. Available: \url{https://www.r-project.org/}
+
 \newpage
 
-\subsection{Figures}\label{figures}
+\section{Figures}\label{figures}
 
@@
 
-\newpage
-
-\subsection{References}\label{references}
-
-\raggedright
-
-\hypertarget{refs}{}
-\hypertarget{ref-Phillips1991}{}
-1. Phillips AN, Davey Smith G. How independent are ``independent''
-effects? relative risk estimation when correlated exposures are measured
-imprecisely. Journal of Clinical Epidemiology. Pergamon; 1991;44:
-1223--1231.
-doi:\href{https://doi.org/10.1016/0895-4356(91)90155-3}{10.1016/0895-4356(91)90155-3}
-
-\hypertarget{ref-DaveySmith2002}{}
-2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
-2002;325: 1437--8.
-
-\hypertarget{ref-DaveySmith2004}{}
-3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
-potentials, and limitations. International journal of epidemiology.
-2004;33: 30--42.
-doi:\href{https://doi.org/10.1093/ije/dyh132}{10.1093/ije/dyh132}
-
-\hypertarget{ref-Millstein2009}{}
-4. Millstein J, Zhang B, Zhu J, Schadt EE. Disentangling molecular
-relationships with a causal inference test. BMC genetics. 2009;10: 23.
-doi:\href{https://doi.org/10.1186/1471-2156-10-23}{10.1186/1471-2156-10-23}
-
-\hypertarget{ref-Aten2008}{}
-5. Aten JE, Fuller TF, Lusis AJ, Horvath S. Using genetic markers to
-orient the edges in quantitative trait networks: the NEO software. BMC
-systems biology. 2008;2: 34.
-doi:\href{https://doi.org/10.1186/1752-0509-2-34}{10.1186/1752-0509-2-34}
-
-\hypertarget{ref-Waszak2015}{}
-6. Waszak SM, Delaneau O, Gschwind AR, Kilpinen H, Raghav SK, Witwicki
-RM, et al. Variation and genetic control of chromatin architecture in
-humans. Cell. Elsevier Inc. 2015;162: 1039--1050.
-doi:\href{https://doi.org/10.1016/j.cell.2015.08.001}{10.1016/j.cell.2015.08.001}
-
-\hypertarget{ref-Houle2011}{}
-7. Houle D, Pélabon C, Wagner G, Hansen T. Measurement and meaning in
-biology. The Quarterly Review of Biology. 2011;86: 3--34. Available:
-\url{http://www.jstor.org/stable/10.1086/658408}
-
-\hypertarget{ref-Hernan2009}{}
-8. Hernán M a, Cole SR. Invited Commentary: Causal diagrams and
-measurement bias. American journal of epidemiology. 2009;170: 959--62;
-discussion 963--4.
-doi:\href{https://doi.org/10.1093/aje/kwp293}{10.1093/aje/kwp293}
-
-\hypertarget{ref-Harper2013}{}
-9. Harper KN, Peters B a, Gamble MV. Batch effects and pathway analysis:
-two potential perils in cancer studies involving DNA methylation array
-analysis. Cancer epidemiology, biomarkers \& prevention : a publication
-of the American Association for Cancer Research, cosponsored by the
-American Society of Preventive Oncology. 2013;22: 1052--60.
-doi:\href{https://doi.org/10.1158/1055-9965.EPI-13-0114}{10.1158/1055-9965.EPI-13-0114}
-
-\hypertarget{ref-Chen2013a}{}
-10. Chen Y-a, Lemire M, Choufani S, Butcher DT, Grafodatskaya D, Zanke
-BW, et al. Discovery of cross-reactive probes and polymorphic CpGs in
-the Illumina Infinium HumanMethylation450 microarray. Epigenetics :
-official journal of the DNA Methylation Society. 2013;8: 203--9.
-doi:\href{https://doi.org/10.4161/epi.23470}{10.4161/epi.23470}
-
-\hypertarget{ref-Houseman2012}{}
-11. Houseman EA, Accomando WP, Koestler DC, Christensen BC, Marsit CJ,
-Nelson HH, et al. DNA methylation arrays as surrogate measures of cell
-mixture distribution. BMC bioinformatics. 2012;13: 86.
-doi:\href{https://doi.org/10.1186/1471-2105-13-86}{10.1186/1471-2105-13-86}
-
-\hypertarget{ref-Ahima2013}{}
-12. Ahima RS, Lazar MA. Physiology. The health risk of obesity--better
-metrics imperative. Science (New York, NY). American Association for the
-Advancement of Science; 2013;341: 856--8.
-doi:\href{https://doi.org/10.1126/science.1241244}{10.1126/science.1241244}
-
-\hypertarget{ref-LeCessie2012}{}
-13. Cessie S le, Debeij J, Rosendaal FR, Cannegieter SC, Vandenbroucke
-JP. Quantification of bias in direct effects estimates due to different
-types of measurement error in the mediator. Epidemiology (Cambridge,
-Mass). 2012;23: 551--60.
-doi:\href{https://doi.org/10.1097/EDE.0b013e318254f5de}{10.1097/EDE.0b013e318254f5de}
-
-\hypertarget{ref-Blakely2013}{}
-14. Blakely T, McKenzie S, Carter K. Misclassification of the mediator
-matters when estimating indirect effects. Journal of epidemiology and
-community health. 2013;67: 458--66.
-doi:\href{https://doi.org/10.1136/jech-2012-201813}{10.1136/jech-2012-201813}
-
-\hypertarget{ref-DaveySmith2003}{}
-15. Davey Smith G, Ebrahim S. 'Mendelian randomization': can genetic
-epidemiology contribute to understanding environmental determinants of
-disease? International Journal of Epidemiology. 2003;32: 1--22.
-doi:\href{https://doi.org/10.1093/ije/dyg070}{10.1093/ije/dyg070}
-
-\hypertarget{ref-DaveySmithHemani2014}{}
-16. Davey Smith G, Hemani G. Mendelian randomization: genetic anchors
-for causal inference in epidemiological studies. Human molecular
-genetics. Oxford Univ Press; 2014;23: R89-----R98.
-doi:\href{https://doi.org/10.1093/hmg/ddu328}{10.1093/hmg/ddu328}
-
-\hypertarget{ref-Schadt2005}{}
-17. Schadt EE, Lamb J, Yang X, Zhu J, Edwards S, GuhaThakurta D, et al.
-An integrative genomics approach to infer causal associations between
-gene expression and disease. Nature Genetics. 2005;37: 710--717.
-doi:\href{https://doi.org/10.1038/ng1589}{10.1038/ng1589}
-
-\hypertarget{ref-Millstein2016}{}
-18. Millstein J. cit: Causal Inference Test. R package version 1.9
-{[}Internet{]}. 2016. Available:
-\url{http://cran.r-project.org/package=cit}
-
-\hypertarget{ref-Koestler2014}{}
-19. Koestler DC, Chalise P, Cicek MS, Cunningham JM, Armasu S, Larson
-MC, et al. Integrative genomic analysis identifies epigenetic marks that
-mediate genetic risk for epithelial ovarian cancer. BMC medical
-genomics. BMC Medical Genomics; 2014;7: 8.
-doi:\href{https://doi.org/10.1186/1755-8794-7-8}{10.1186/1755-8794-7-8}
-
-\hypertarget{ref-Liu2013}{}
-20. Liu Y, Aryee MJ, Padyukov L, Fallin MD, Hesselberg E, Runarsson A,
-et al. Epigenome-wide association data implicate DNA methylation as an
-intermediary of genetic risk in rheumatoid arthritis. Nature
-biotechnology. 2013;31: 142--7.
-doi:\href{https://doi.org/10.1038/nbt.2487}{10.1038/nbt.2487}
-
-\hypertarget{ref-Yuan2014}{}
-21. Yuan W, Xia Y, Bell CG, Yet I, Ferreira T, Ward KJ, et al. An
-integrated epigenomic analysis for type 2 diabetes susceptibility loci
-in monozygotic twins. Nature communications. 2014;5: 5719.
-doi:\href{https://doi.org/10.1038/ncomms6719}{10.1038/ncomms6719}
-
-\hypertarget{ref-Tang2014}{}
-22. Tang Y, Axelsson AS, Spégel P, Andersson LE, Mulder H, Groop LC, et
-al. Genotype-based treatment of type 2 diabetes with an
-\(\alpha\)2A-adrenergic receptor antagonist. Science translational
-medicine. 2014;6: 257ra139.
-doi:\href{https://doi.org/10.1126/scitranslmed.3009934}{10.1126/scitranslmed.3009934}
-
-\hypertarget{ref-Hong2015}{}
-23. Hong X, Hao K, Ladd-Acosta C, Hansen KD, Tsai H-J, Liu X, et al.
-Genome-wide association study identifies peanut allergy-specific loci
-and evidence of epigenetic mediation in US children. Nature
-communications. 2015;6: 6304.
-doi:\href{https://doi.org/10.1038/ncomms7304}{10.1038/ncomms7304}
-
-\hypertarget{ref-Bowden2015}{}
-24. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
-invalid instruments: effect estimation and bias detection through Egger
-regression. International Journal of Epidemiology. 2015;44: 512--25.
-
-\hypertarget{ref-Bowden2016b}{}
-25. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent
-Estimation in Mendelian Randomization with Some Invalid Instruments
-Using a Weighted Median Estimator. Genetic Epidemiology. 2016;40:
-304--314.
-doi:\href{https://doi.org/10.1002/gepi.21965}{10.1002/gepi.21965}
-
-\hypertarget{ref-Hemani2016}{}
-26. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
-MR-Base: a platform for systematic causal inference across the phenome
-using billions of genetic associations. BioRxiv. 2016;10.1101/07.
-
-\hypertarget{ref-Timpson2011}{}
-27. Timpson NJ, Nordestgaard BG, Harbord RM, Zacho J, Frayling TM,
-Tybjærg-Hansen a, et al. C-reactive protein levels and body mass index:
-elucidating direction of causation through reciprocal Mendelian
-randomization. International journal of obesity (2005). 2011;35: 300--8.
-doi:\href{https://doi.org/10.1038/ijo.2010.137}{10.1038/ijo.2010.137}
-
-\hypertarget{ref-Richmond2014}{}
-28. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
-NJ. Assessing Causality in the Association between Child Adiposity and
-Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
-editor. PLoS Medicine. 2014;11: e1001618.
-doi:\href{https://doi.org/10.1371/journal.pmed.1001618}{10.1371/journal.pmed.1001618}
-
-\hypertarget{ref-Claussnitzer2015}{}
-29. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
-al. FTO Obesity Variant Circuitry and Adipocyte Browning in Humans. The
-New England journal of medicine. 2015;373: 895--907.
-doi:\href{https://doi.org/10.1056/NEJMoa1502214}{10.1056/NEJMoa1502214}
-
-\hypertarget{ref-Pierce2013}{}
-30. Pierce BL, Burgess S. Efficient design for Mendelian randomization
-studies: subsample and 2-sample instrumental variable estimators.
-American journal of epidemiology. 2013;178: 1177--84.
-doi:\href{https://doi.org/10.1093/aje/kwt084}{10.1093/aje/kwt084}
-
-\hypertarget{ref-Hernan2004}{}
-31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
-selection bias. Epidemiology (Cambridge, Mass). 2004;15: 615--25.
-Available: \url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}
-
-\hypertarget{ref-Pierce2012}{}
-32. Pierce BL, VanderWeele TJ. The effect of non-differential
-measurement error on bias, precision and power in Mendelian
-randomization studies. International Journal of Epidemiology. Oxford
-University Press; 2012;41: 1383--1393.
-doi:\href{https://doi.org/10.1093/ije/dys141}{10.1093/ije/dys141}
-
-\hypertarget{ref-Ashenfelter1994}{}
-33. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
-Schooling from a New Sample of Twins. The American Economic Review.
-1994;84: 1157--1173.
-doi:\href{https://doi.org/10.1016/S0272-7757(98)00038-7}{10.1016/S0272-7757(98)00038-7}
-
-\hypertarget{ref-Nagarajan2013}{}
-34. Nagarajan R, Scutari M. Impact of noise on molecular network
-inference. PloS one. 2013;8: e80735.
-doi:\href{https://doi.org/10.1371/journal.pone.0080735}{10.1371/journal.pone.0080735}
-
-\hypertarget{ref-Shpitser2010}{}
-35. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
-adjustment for estimating causal effects. Proceedings of the Twenty
-Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
-2010; 527--536.
-
-\hypertarget{ref-Wang2015}{}
-36. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
-networks. arXiv. 2015;arXiv:1512. Available:
-\url{http://arxiv.org/abs/1512.05574}
-
-\hypertarget{ref-Lagani2015}{}
-37. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
-Probabilistic Computational Causal Discovery for Systems Biology.
-Uncertainty in biology: A computational modeling approach. Springer;
-2015. p. 47. Available:
-\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}
-
-\hypertarget{ref-Lawlor2017}{}
-38. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
-epidemiology. International Journal of Epidemiology. 2017;19: dyw314.
-doi:\href{https://doi.org/10.1093/ije/dyw314}{10.1093/ije/dyw314}
-
-\hypertarget{ref-Sterne2001}{}
-39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
-significance tests? BMJ. 2001;322: 226--231.
-
-\hypertarget{ref-Henningsen2007}{}
-40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. Journal of Statistical Software.
-2007;23: 1--40.
-doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
-
-\hypertarget{ref-Steiger1980}{}
-41. Steiger JH. Tests for comparing elements of a correlation matrix.
-Psychological Bulletin. 1980;87: 245--251.
-doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
-
-\hypertarget{ref-Revelle2015}{}
-42. Revelle W. psych: Procedures for Psychological, Psychometric, and
-Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
-University; 2015. Available:
-\url{http://cran.r-project.org/package=psych}
-
-\hypertarget{ref-Shakhbazov2016}{}
-43. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
-PM, et al. Shared genetic control of expression and methylation in
-peripheral blood. BMC genomics. BioMed Central; 2016;17: 278.
-doi:\href{https://doi.org/10.1186/s12864-016-2498-4}{10.1186/s12864-016-2498-4}
-
-\hypertarget{ref-RCoreTeam2015}{}
-44. R Core Team. R: A Language and Environment for Statistical Computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2015. Available: \url{https://www.r-project.org/}
-
-\hypertarget{ref-Bird2002}{}
-45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
-development. 2002;16: 6--21.
-doi:\href{https://doi.org/10.1101/gad.947102}{10.1101/gad.947102}
-
-\hypertarget{ref-Cole2014}{}
-46. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
-Serious and Misleading Consequences Due to Uncorrected Measurement
-Error. Psychological Methods. 2014;19: 300--315.
-doi:\href{https://doi.org/10.1037/a0033805}{10.1037/a0033805}
-
-\hypertarget{ref-Bose2014}{}
-47. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
-Evaluation of microarray-based DNA methylation measurement using
-technical replicates: the Atherosclerosis Risk In Communities (ARIC)
-Study. BMC Bioinformatics. 2014;15: 312.
-doi:\href{https://doi.org/10.1186/1471-2105-15-312}{10.1186/1471-2105-15-312}
-
-\hypertarget{ref-Bryant2011}{}
-48. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
-et al. Technical Variability Is Greater than Biological Variability in a
-Microarray Experiment but Both Are Outweighed by Changes Induced by
-Stimulation. Khodursky AB, editor. PLoS ONE. Public Library of Science;
-2011;6: e19556.
-doi:\href{https://doi.org/10.1371/journal.pone.0019556}{10.1371/journal.pone.0019556}
-
-\hypertarget{ref-Zhu2016}{}
-49. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
-Integration of summary data from GWAS and eQTL studies predicts complex
-trait gene targets. Nature Genetics. Nature Research; 2016;48: 481--487.
-doi:\href{https://doi.org/10.1038/ng.3538}{10.1038/ng.3538}
-
-\hypertarget{ref-Mancuso2017}{}
-50. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
-Integrating Gene Expression with Summary Association Statistics to
-Identify Genes Associated with 30 Complex Traits. The American Journal
-of Human Genetics. 2017;100: 473--487.
-doi:\href{https://doi.org/10.1016/j.ajhg.2017.01.031}{10.1016/j.ajhg.2017.01.031}
-
-\hypertarget{ref-Lee2013c}{}
-51. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
-association studies: quantification of power and genomic prediction
-accuracy. PLoS One. 2013;8: e71494.
-doi:\href{https://doi.org/10.1371/journal.pone.0071494\%20\%5Bdoi\%5D/rPONE-D-13-18776\%20\%5Bpii\%5D}{10.1371/journal.pone.0071494 {[}doi{]}\textbackslash{}rPONE-D-13-18776 {[}pii{]}}
-
-\hypertarget{ref-Do2013}{}
-52. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
-Common variants associated with plasma triglycerides and risk for
-coronary artery disease. Nature Genetics. Nature Research; 2013;45:
-1345--1352. doi:\href{https://doi.org/10.1038/ng.2795}{10.1038/ng.2795}
-
-\hypertarget{ref-Burgess2014a}{}
-53. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
-multivariable Mendelian randomization to disentangle the causal effects
-of lipid fractions. PloS one. Public Library of Science; 2014;9:
-e108891.
-doi:\href{https://doi.org/10.1371/journal.pone.0108891}{10.1371/journal.pone.0108891}
-
-\hypertarget{ref-Relton2012}{}
-54. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
-randomization: a strategy for establishing the causal role of epigenetic
-processes in pathways to disease. International journal of epidemiology.
-2012;41: 161--76.
-doi:\href{https://doi.org/10.1093/ije/dyr233}{10.1093/ije/dyr233}
-
-\hypertarget{ref-Varbo2015}{}
-55. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
-Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
-cholesterol, and blood pressure as mediators from obesity to ischemic
-heart disease. Circulation research. 2015;116: 665--73.
-doi:\href{https://doi.org/10.1161/CIRCRESAHA.116.304846}{10.1161/CIRCRESAHA.116.304846}
-
-\hypertarget{ref-Burgess2015}{}
-56. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
-randomization: using genetic variants as instrumental variables to
-investigate mediation in causal pathways. International journal of
-epidemiology. 2015;44: 484--95.
-doi:\href{https://doi.org/10.1093/ije/dyu176}{10.1093/ije/dyu176}
-
-\hypertarget{ref-Richmond2016}{}
-57. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
-Challenges and novel approaches for investigating molecular mediation.
-Human molecular genetics. Oxford University Press; 2016;25: R149--R156.
-doi:\href{https://doi.org/10.1093/hmg/ddw197}{10.1093/hmg/ddw197}
-
 \end{document}

### 2017-11-20 docs/manuscript.tex
updated docs for proofs

@@
   \postdate{\par}
-  \date{24 October 2017}
+  \date{19 November 2017}
 
@@
 
-\subsubsection{Abstract}\label{abstract}
+\hypertarget{abstract}{%
+\subsubsection{Abstract}\label{abstract}}
 
@@
 
-\subsubsection{Author summary}\label{author-summary}
+\hypertarget{author-summary}{%
+\subsubsection{Author summary}\label{author-summary}}
 
@@
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
@@
 
-\subsection{Model}\label{model}
+\hypertarget{model}{%
+\subsubsection{Model}\label{model}}
 
@@
 
-\subsection{Methods}\label{methods}
-
-\subsubsection{CIT test}\label{cit-test}
-
-First we describe how the CIT method {[}4{]} is implemented in the R
-package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
-Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
-exposure \(x\) causes an outcome \(y\), as described above. The
-following tests are then performed:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
-  associates with the outcome}
-\item
-  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
-  associates with the exposure conditional on the outcome}
-\item
-  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
-  associates with the outcome conditional on the SNP}
-\item
-  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
-  independent of the outcome conditional on the exposure}
-\end{enumerate}
-
-The term in the 4th test can be rewritten as a
-\(cov(g, y|x) = cov(g, y - \hat{y})\) where
-\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
-of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
-association between the SNP and the outcome. The condition in the 4th
-test is formulated as an equivalence testing problem that is estimated
-using simulations, comparing the estimate from the data against
-empirically obtained estimates for simulated variables where the
-independence model is true (full details are given in {[}4{]}). We note
-here that this approach is liable to fail, even when there is a true
-causal relationship, when confounders of the exposure and outcome are
-present, as these will induce collider bias.
-
-If all four tests reject the null hypothesis then it is inferred that
-\(x\) causes \(y\). The CIT measures the strength of causality by
-generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
-(least extreme) p-value of the four tests, the intuition being that
-causal inference is only as strong as the weakest link in the chain of
-tests.
-
-Now we describe how we used the CIT method in our simulations. The
-\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
-the direction of causality using the CIT method, an omnibus p-value
-generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
-was estimated for the direction of \(x\) causing \(y\) (Model 1), and
-for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
-(Model 2). The results from each of these methods can then be used in
-combination to infer the existance and direction of causality. For some
-significance threshold \(\alpha\) there are four possible outcomes from
-these two tests, and their interpretations are as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
-  relationship
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
-  confounding (S1 fig) and no call is made.
-\end{itemize}
-
-For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value, though we stress that for real analyses it is
-not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds {[}39{]}.
-
-\subsubsection{MR causal test}\label{mr-causal-test}
-
-Two stage least squares (2SLS) is a commonly used technique for
-performing MR when the exposure, outcome and instrument data are all
-available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the \(systemfit\) function in the R package
-\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
-when using the same genetic variant to instrument the influence of the
-exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
-outcome \(y\) on the exposure \(x\).
-
-The method that we will now describe is designed to distinguish between
-two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
-framework, this approach cannot infer if the true model is
-\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
-additive.
-
-To infer the direction of causality it is desirable to know which of the
-variables, \(x\) or \(y\), is being directly influenced by the
-instrument \(g\). This can be achieved by assessing which of the two
-variables has the biggest absolute correlation with \(g\) (S2 Text),
-formalised by testing for a difference in the correlations \(\rho_{gx}\)
-and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population {[}41{]}. It is calculated as
-
-\[
-Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
-\]
-
-where Fisher's z-transformation is used to obtain
-\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
-
-\[
-h = \frac{1 - (frm^2)} {1 - rm^2}
-\]
-
-where
-
-\[
-f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
-\]
-
-and
-
-\[
-rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
-\]
-
-The \(Z\) value is interpreted such that
-
-\[
-Z \left\{
-\begin{array}{ll}
-> 0, & x \to y\\
-< 0, & y \to x\\
-= 0, & x \perp\!\!\!\perp y 
-\end{array} \right.
-\]
-
-and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
-indicate the probability of obtaining a difference between correlations
-\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
-under the null hypothesis that both correlations are identical.
-
-The existence of causality and its direction is inferred based on
-combining information from the MR analysis and the Steiger test. The MR
-analysis indicates whether there is a potential causal relationship
-(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
-of the causal relationship and the confidence of the direction
-(\(p_{Steiger}\)). For the purposes of compiling simulation results,
-these can be combined using an arbitrary \(\alpha = 0.05\) value:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
-  a causal association for the correct model is accepted,
-  \(x \rightarrow y\)
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
-  a causal association for the incorrect model is accepted,
-  \(y \rightarrow x\)
-\item
-  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
-  model is accepted
-\end{itemize}
-
-Note that the same correlation test approach can be applied to a
-two-sample MR setting. Two-sample MR refers to the case where the
-SNP-exposure association and SNP-outcome association are calculated in
-different samples (e.g.~from publicly available summary statistics
-{[}26,30{]}). Here the Steiger test of two independent correlations can
-be applied where.
-
-\[
-Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
-\]
-
-An advantage of using the Steiger test in the two sample context is that
-it can compare correlations in independent samples where sample sizes
-are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.
-
-The Steiger test assumes that there is a causal relationship between the
-two variables, and that the SNP is a valid instrument for one of them.
-However it is liable to give incorrect causal directions under some
-other circumstances. First, some levels of horizontal pleiotropy, where
-the SNP influences the outcome through some pathway other than the
-exposure, could induce problems because this is a means by which the
-instrument is invalid. Second, some differential values of measurement
-error between the exposure and the outcome could lead to incorrect
-inference of the causal direction (S2 Text). Third, some levels of
-unmeasured confounding between the exposure and the outcome could lead
-to inference of the wrong causal direction (S3 Text).
-
-\subsubsection{Causal direction sensitivity analysis for measurement
-error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
-
-The Steiger test for inferring if \(x \rightarrow y\) is based on
-evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
-\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
-imprecisely. If, for example, \(x\) has lower measurement precision than
-\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
-because \(\rho_{g,x_o}\) could be underestimated more than
-\(\rho_{g,y_o}\).
-
-As we show in S2 Text it is possible to infer the bounds of measurement
-error on \(x_o\) or \(y_o\) given known genetic associations. The
-maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
-it is known that at least that much of the variance has been explained
-in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
-values (the same logic applies to \(y_o\)). It is possible to simulate
-what the inferred causal direction would be for all values within these
-bounds.
-
-To evaluate how reliable, \(R\), the inference of the causal direction
-is to potential measurement error in \(x\) and \(y\) we need to predict
-the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
-error. We offer two tools in which to do this. First, the user can
-provide values of measurement error for \(x\) and \(y\) and obtain a
-revised inference of the causal direction. Second, we integrate over the
-entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
-measurement error values, assuming that any measurement error value is
-equally likely. Across all possible values of measurement error in \(x\)
-and \(y\) we find the volume that agrees with the inferred direction of
-causality and the volume that disagrees with the inferred direction of
-causality, and take the ratio of these two values. A ratio \(R=1\)
-indicates that the inferred causal direction is highly sensitive to
-measurement error, because equal weight of the measurement error
-parameter space supports each direction of causality. In general, the
-\(R\) value denotes that the inferred direction of causality is \(R\)
-times more likely to be the empirical result than the opposite direction
-(S2 Text).
-
-\subsubsection{Simulations}\label{simulations}
-
-Simulations were conducted by creating variables of sample size \(n\)
-for the exposure \(x\), the measured values of the exposure \(x_o\), the
-outcome \(y\), the measured values of the outcome \(y_o\) and the
-instrument \(g\). One of two models are simulated, the ``causal model''
-where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
-``non-causal model'' where \(g\) influences a confounder \(u\) which in
-turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
-but not causally related. Each variable in the causal model was
-simulated such that:
-
-\[
-\begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
-\end{aligned}
-\]
-
-where non-differential measurement error is represented by a noise
-(measurement imprecision) term
-\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
-\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
-the outcome variable \(y\). Note that following the first section of the
-Results we no longer include the bias terms for simplicity. We have
-formulated the non-causal model as:
-
-\[
-\begin{aligned}
-y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
-\end{aligned}
-\]
-
-All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
-Normally distributed values of \(\epsilon_*\) were generated such that
-
-\[
-\begin{aligned}
-cor(g, x)^2 & = 0.1 \\
-cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
-\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-n & = \{100, 1000, 10000\}
-\end{aligned}
-\]
-
-giving a total of 432 combinations of parameters. Simulations using each
-of these sets of variables were performed 100 times, and the CIT and MR
-methods were applied to each in order to evaluate the causal association
-of the simulated variables. Similar patterns of results were obtained
-for different values of \(cor(g, x)\).
-
-\subsection{Applied example using two sample
-MR}\label{applied-example-using-two-sample-mr}
-
-Two sample MR {[}30{]} was performed using summary statistics for
-genetic influences on gene expression and DNA methylation. To do this we
-obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al {[}43{]}. These were filtered to be located
-on the same chromosome, have robust correlations after correcting for
-multiple testing, and to share a SNP that had a robust cis-acting effect
-on both the DNA methylation probe and the gene expression probe. Because
-only summary statistics were available (effect, standard error, effect
-allele, sample size, p-values) for the instrumental SNP on the
-methylation and gene expression levels, the Steiger test of two
-independent correlations was used to infer the direction of causality
-for each of the associations. The Wald ratio test was then used to
-estimate the causal effect size for the estimated direction for each
-association.
-
-All analysis was performed using the R programming language {[}44{]} and
-code is made available at
-\url{https://github.com/explodecomputer/causal-directions} and
-implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
-{[}26{]}.
-
-\subsection{Results}\label{results}
-
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+\hypertarget{mediation-based-causal-inference-under-measurement-error}{%
 \subsubsection{Mediation-based causal inference under measurement
-error}\label{mediation-based-causal-inference-under-measurement-error}
+error}\label{mediation-based-causal-inference-under-measurement-error}}
 
@@
 
+\hypertarget{using-mr-steiger-to-infer-the-direction-of-causality}{%
 \subsubsection{Using MR Steiger to infer the direction of
-causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}
+causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}}
 
@@
 
+\hypertarget{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}{%
 \subsubsection{Comparison of CIT and MR Steiger for obtaining the
 correct direction of
-causality}\label{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}
+causality}\label{comparison-of-cit-and-mr-steiger-for-obtaining-the-correct-direction-of-causality}}
 
@@
 
+\hypertarget{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}{%
 \subsubsection{The causal relationship between gene expression and DNA
 methylation
-levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}
+levels}\label{the-causal-relationship-between-gene-expression-and-dna-methylation-levels}}
 
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation {[}45{]}.
-
-\subsection{Discussion}\label{discussion}
+levels, consistent with expectation {[}40{]}.
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+\hypertarget{cit-test}{%
+\subsubsection{CIT test}\label{cit-test}}
+
+Here we describe how the CIT method {[}4{]} is implemented in the R
+package \emph{R/cit} {[}18{]}. Assume an exposure \(x\) is instrumented
+by a SNP \(g\), and the exposure \(x\) causes an outcome \(y\), as
+described above. The following tests are then performed:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
+  associates with the outcome}
+\item
+  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
+  associates with the exposure conditional on the outcome}
+\item
+  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
+  associates with the outcome conditional on the SNP}
+\item
+  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
+  independent of the outcome conditional on the exposure}
+\end{enumerate}
+
+The term in the 4th test can be rewritten as
+\(cov(g, y|x) = cov(g, y - \hat{y})\) where
+\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
+of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
+association between the SNP and the outcome. The condition in the 4th
+test is formulated as an equivalence testing problem that is estimated
+using simulations, comparing the estimate from the data against
+empirically obtained estimates for simulated variables where the
+independence model is true (full details are given in {[}4{]}). We note
+here that this approach is liable to fail, even when there is a true
+causal relationship, when confounders of the exposure and outcome are
+present, as these will induce collider bias.
+
+If all four tests reject the null hypothesis then it is inferred that
+\(x\) causes \(y\). The CIT measures the strength of causality by
+generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
+(least extreme) p-value of the four tests, the intuition being that
+causal inference is only as strong as the weakest link in the chain of
+tests.
+
+Now we describe how we used the CIT method in our simulations. The
+\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
+the direction of causality using the CIT method, an omnibus p-value
+generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
+was estimated for the direction of \(x\) causing \(y\) (Model 1), and
+for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
+(Model 2). The results from each of these methods can then be used in
+combination to infer the existance and direction of causality. For some
+significance threshold \(\alpha\) there are four possible outcomes from
+these two tests, and their interpretations are as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
+  relationship
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
+  confounding (S1 fig) and no call is made.
+\end{itemize}
+
+For the purposes of compiling simulation results we use an arbitrary
+\(\alpha = 0.05\) value, though we stress that for real analyses it is
+not good practice to rely on p-values for making causal inference, nor
+is it reliable to depend on arbitrary significance thresholds {[}53{]}.
+
+\hypertarget{mr-causal-test}{%
+\subsubsection{MR causal test}\label{mr-causal-test}}
+
+Two stage least squares (2SLS) is a commonly used technique for
+performing MR when the exposure, outcome and instrument data are all
+available in the same sample. A p-value for this test, \(p_{MR}\), was
+obtained using the \(systemfit\) function in the R package
+\(R/systemfit\) {[}54{]}. Note that the value of \(p_{MR}\) is identical
+when using the same genetic variant to instrument the influence of the
+exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
+outcome \(y\) on the exposure \(x\).
+
+The method that we will now describe is designed to distinguish between
+two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
+framework, this approach cannot infer if the true model is
+\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
+additive.
+
+To infer the direction of causality it is desirable to know which of the
+variables, \(x\) or \(y\), is being directly influenced by the
+instrument \(g\). This can be achieved by assessing which of the two
+variables has the biggest absolute correlation with \(g\) (S2 Text),
+formalised by testing for a difference in the correlations \(\rho_{gx}\)
+and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
+within a population {[}55{]}. It is calculated as
+
+\[
+Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
+\]
+
+where Fisher's z-transformation is used to obtain
+\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
+
+\[
+h = \frac{1 - (frm^2)} {1 - rm^2}
+\]
+
+where
+
+\[
+f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
+\]
+
+and
+
+\[
+rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
+\]
+
+The \(Z\) value is interpreted such that
+
+\[
+Z \left\{
+\begin{array}{ll}
+> 0, & x \to y\\
+< 0, & y \to x\\
+= 0, & x \perp\!\!\!\perp y 
+\end{array} \right.
+\]
+
+and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
+indicate the probability of obtaining a difference between correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
+under the null hypothesis that both correlations are identical.
+
+The existence of causality and its direction is inferred based on
+combining information from the MR analysis and the Steiger test. The MR
+analysis indicates whether there is a potential causal relationship
+(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
+of the causal relationship and the confidence of the direction
+(\(p_{Steiger}\)). For the purposes of compiling simulation results,
+these can be combined using an arbitrary \(\alpha = 0.05\) value:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
+  a causal association for the correct model is accepted,
+  \(x \rightarrow y\)
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
+  a causal association for the incorrect model is accepted,
+  \(y \rightarrow x\)
+\item
+  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
+  model is accepted
+\end{itemize}
+
+Note that the same correlation test approach can be applied to a
+two-sample MR setting. Two-sample MR refers to the case where the
+SNP-exposure association and SNP-outcome association are calculated in
+different samples (e.g.~from publicly available summary statistics
+{[}26,30{]}). Here the Steiger test of two independent correlations can
+be applied where.
+
+\[
+Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
+\]
+
+An advantage of using the Steiger test in the two sample context is that
+it can compare correlations in independent samples where sample sizes
+are different. Steiger test statistics were calculated using the
+\emph{r.test} function in the R package \emph{R/psych} {[}56{]}.
+
+The Steiger test assumes that there is a causal relationship between the
+two variables, and that the SNP is a valid instrument for one of them.
+However it is liable to give incorrect causal directions under some
+other circumstances. First, some levels of horizontal pleiotropy, where
+the SNP influences the outcome through some pathway other than the
+exposure, could induce problems because this is a means by which the
+instrument is invalid. Second, some differential values of measurement
+error between the exposure and the outcome could lead to incorrect
+inference of the causal direction (S2 Text). Third, some levels of
+unmeasured confounding between the exposure and the outcome could lead
+to inference of the wrong causal direction (S3 Text).
+
+\hypertarget{causal-direction-sensitivity-analysis-for-measurement-error}{%
+\subsubsection{Causal direction sensitivity analysis for measurement
+error}\label{causal-direction-sensitivity-analysis-for-measurement-error}}
+
+The Steiger test for inferring if \(x \rightarrow y\) is based on
+evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
+\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
+imprecisely. If, for example, \(x\) has lower measurement precision than
+\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
+because \(\rho_{g,x_o}\) could be underestimated more than
+\(\rho_{g,y_o}\).
+
+As we show in S2 Text it is possible to infer the bounds of measurement
+error on \(x_o\) or \(y_o\) given known genetic associations. The
+maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
+it is known that at least that much of the variance has been explained
+in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
+values (the same logic applies to \(y_o\)). It is possible to simulate
+what the inferred causal direction would be for all values within these
+bounds.
+
+To evaluate how reliable, \(R\), the inference of the causal direction
+is to potential measurement error in \(x\) and \(y\) we need to predict
+the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
+error. We offer two tools in which to do this. First, the user can
+provide values of measurement error for \(x\) and \(y\) and obtain a
+revised inference of the causal direction. Second, we integrate over the
+entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
+measurement error values, assuming that any measurement error value is
+equally likely. Across all possible values of measurement error in \(x\)
+and \(y\) we find the volume that agrees with the inferred direction of
+causality and the volume that disagrees with the inferred direction of
+causality, and take the ratio of these two values. A ratio \(R=1\)
+indicates that the inferred causal direction is highly sensitive to
+measurement error, because equal weight of the measurement error
+parameter space supports each direction of causality. In general, the
+\(R\) value denotes that the inferred direction of causality is \(R\)
+times more likely to be the empirical result than the opposite direction
+(S2 Text).
+
+\hypertarget{simulations}{%
+\subsubsection{Simulations}\label{simulations}}
+
+Simulations were conducted by creating variables of sample size \(n\)
+for the exposure \(x\), the measured values of the exposure \(x_o\), the
+outcome \(y\), the measured values of the outcome \(y_o\) and the
+instrument \(g\). One of two models are simulated, the ``causal model''
+where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
+``non-causal model'' where \(g\) influences a confounder \(u\) which in
+turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
+but not causally related. Each variable in the causal model was
+simulated such that:
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+\]
+
+where non-differential measurement error is represented by a noise
+(measurement imprecision) term
+\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
+\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
+the outcome variable \(y\). Note that following the first section of the
+Results we no longer include the bias terms for simplicity. We have
+formulated the non-causal model as:
+
+\[
+\begin{aligned}
+y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
+\end{aligned}
+\]
+
+All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
+Normally distributed values of \(\epsilon_*\) were generated such that
+
+\[
+\begin{aligned}
+cor(g, x)^2 & = 0.1 \\
+cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
+\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+n & = \{100, 1000, 10000\}
+\end{aligned}
+\]
+
+giving a total of 432 combinations of parameters. Simulations using each
+of these sets of variables were performed 100 times, and the CIT and MR
+methods were applied to each in order to evaluate the causal association
+of the simulated variables. Similar patterns of results were obtained
+for different values of \(cor(g, x)\).
+
+\hypertarget{applied-example-using-two-sample-mr}{%
+\subsubsection{Applied example using two sample
+MR}\label{applied-example-using-two-sample-mr}}
+
+Two sample MR {[}30{]} was performed using summary statistics for
+genetic influences on gene expression and DNA methylation. To do this we
+obtained a list of 458 gene expression - DNA methylation associations as
+reported in Shakhbazov et al {[}39{]}. These were filtered to be located
+on the same chromosome, have robust correlations after correcting for
+multiple testing, and to share a SNP that had a robust cis-acting effect
+on both the DNA methylation probe and the gene expression probe. Because
+only summary statistics were available (effect, standard error, effect
+allele, sample size, p-values) for the instrumental SNP on the
+methylation and gene expression levels, the Steiger test of two
+independent correlations was used to infer the direction of causality
+for each of the associations. The Wald ratio test was then used to
+estimate the causal effect size for the estimated direction for each
+association.
+
+All analysis was performed using the R programming language {[}57{]} and
+code is made available at
+\url{https://github.com/explodecomputer/causal-directions} and
+implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
+{[}26{]}.
+
 \newpage
 
-\subsection{Figures}\label{figures}
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
 
@@
 
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
 Figure 6: Using 458 putative associations between DNA methylation and
@@
 
-\subsection{Supplementary information}\label{supplementary-information}
+\hypertarget{supplementary-information}{%
+\subsection{Supplementary information}\label{supplementary-information}}
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Phillips1991}{}
+\leavevmode\hypertarget{ref-Phillips1991}{}%
 1. Phillips AN, Davey Smith G. How independent are ``independent''
@@
 
-\hypertarget{ref-DaveySmith2002}{}
+\leavevmode\hypertarget{ref-DaveySmith2002}{}%
 2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
@@
 
-\hypertarget{ref-DaveySmith2004}{}
+\leavevmode\hypertarget{ref-DaveySmith2004}{}%
 3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
@@
 
-\hypertarget{ref-Millstein2009}{}
+\leavevmode\hypertarget{ref-Millstein2009}{}%
 4. Millstein J, Zhang B, Zhu J, Schadt EE. Disentangling molecular
@@
 
-\hypertarget{ref-Aten2008}{}
+\leavevmode\hypertarget{ref-Aten2008}{}%
 5. Aten JE, Fuller TF, Lusis AJ, Horvath S. Using genetic markers to
@@
 
-\hypertarget{ref-Waszak2015}{}
+\leavevmode\hypertarget{ref-Waszak2015}{}%
 6. Waszak SM, Delaneau O, Gschwind AR, Kilpinen H, Raghav SK, Witwicki
@@
 
-\hypertarget{ref-Houle2011}{}
+\leavevmode\hypertarget{ref-Houle2011}{}%
 7. Houle D, Pélabon C, Wagner G, Hansen T. Measurement and meaning in
@@
 
-\hypertarget{ref-Hernan2009}{}
+\leavevmode\hypertarget{ref-Hernan2009}{}%
 8. Hernán M a, Cole SR. Invited Commentary: Causal diagrams and
@@
 
-\hypertarget{ref-Harper2013}{}
+\leavevmode\hypertarget{ref-Harper2013}{}%
 9. Harper KN, Peters B a, Gamble MV. Batch effects and pathway analysis:
@@
 
-\hypertarget{ref-Chen2013a}{}
+\leavevmode\hypertarget{ref-Chen2013a}{}%
 10. Chen Y-a, Lemire M, Choufani S, Butcher DT, Grafodatskaya D, Zanke
@@
 
-\hypertarget{ref-Houseman2012}{}
+\leavevmode\hypertarget{ref-Houseman2012}{}%
 11. Houseman EA, Accomando WP, Koestler DC, Christensen BC, Marsit CJ,
@@
 
-\hypertarget{ref-Ahima2013}{}
+\leavevmode\hypertarget{ref-Ahima2013}{}%
 12. Ahima RS, Lazar MA. Physiology. The health risk of obesity--better
@@
 
-\hypertarget{ref-LeCessie2012}{}
+\leavevmode\hypertarget{ref-LeCessie2012}{}%
 13. Cessie S le, Debeij J, Rosendaal FR, Cannegieter SC, Vandenbroucke
@@
 
-\hypertarget{ref-Blakely2013}{}
+\leavevmode\hypertarget{ref-Blakely2013}{}%
 14. Blakely T, McKenzie S, Carter K. Misclassification of the mediator
@@
 
-\hypertarget{ref-DaveySmith2003}{}
+\leavevmode\hypertarget{ref-DaveySmith2003}{}%
 15. Davey Smith G, Ebrahim S. 'Mendelian randomization': can genetic
@@
 
-\hypertarget{ref-DaveySmithHemani2014}{}
+\leavevmode\hypertarget{ref-DaveySmithHemani2014}{}%
 16. Davey Smith G, Hemani G. Mendelian randomization: genetic anchors
@@
 
-\hypertarget{ref-Schadt2005}{}
+\leavevmode\hypertarget{ref-Schadt2005}{}%
 17. Schadt EE, Lamb J, Yang X, Zhu J, Edwards S, GuhaThakurta D, et al.
@@
 
-\hypertarget{ref-Millstein2016}{}
+\leavevmode\hypertarget{ref-Millstein2016}{}%
 18. Millstein J. cit: Causal Inference Test. R package version 1.9
@@
 
-\hypertarget{ref-Koestler2014}{}
+\leavevmode\hypertarget{ref-Koestler2014}{}%
 19. Koestler DC, Chalise P, Cicek MS, Cunningham JM, Armasu S, Larson
@@
 
-\hypertarget{ref-Liu2013}{}
+\leavevmode\hypertarget{ref-Liu2013}{}%
 20. Liu Y, Aryee MJ, Padyukov L, Fallin MD, Hesselberg E, Runarsson A,
@@
 
-\hypertarget{ref-Yuan2014}{}
+\leavevmode\hypertarget{ref-Yuan2014}{}%
 21. Yuan W, Xia Y, Bell CG, Yet I, Ferreira T, Ward KJ, et al. An
@@
 
-\hypertarget{ref-Tang2014}{}
+\leavevmode\hypertarget{ref-Tang2014}{}%
 22. Tang Y, Axelsson AS, Spégel P, Andersson LE, Mulder H, Groop LC, et
@@
 
-\hypertarget{ref-Hong2015}{}
+\leavevmode\hypertarget{ref-Hong2015}{}%
 23. Hong X, Hao K, Ladd-Acosta C, Hansen KD, Tsai H-J, Liu X, et al.
@@
 
-\hypertarget{ref-Bowden2015}{}
+\leavevmode\hypertarget{ref-Bowden2015}{}%
 24. Bowden J, Davey Smith G, Burgess S. Mendelian randomization with
@@
 
-\hypertarget{ref-Bowden2016b}{}
+\leavevmode\hypertarget{ref-Bowden2016b}{}%
 25. Bowden J, Davey Smith G, Haycock PC, Burgess S. Consistent
@@
 
-\hypertarget{ref-Hemani2016}{}
+\leavevmode\hypertarget{ref-Hemani2016}{}%
 26. Hemani G, Zheng J, Wade KH, Laurin C, Elsworth B, Burgess S, et al.
@@
 
-\hypertarget{ref-Timpson2011}{}
+\leavevmode\hypertarget{ref-Timpson2011}{}%
 27. Timpson NJ, Nordestgaard BG, Harbord RM, Zacho J, Frayling TM,
@@
 
-\hypertarget{ref-Richmond2014}{}
+\leavevmode\hypertarget{ref-Richmond2014}{}%
 28. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
@@
 
-\hypertarget{ref-Claussnitzer2015}{}
+\leavevmode\hypertarget{ref-Claussnitzer2015}{}%
 29. Claussnitzer M, Dankel SN, Kim K-H, Quon G, Meuleman W, Haugen C, et
@@
 
-\hypertarget{ref-Pierce2013}{}
+\leavevmode\hypertarget{ref-Pierce2013}{}%
 30. Pierce BL, Burgess S. Efficient design for Mendelian randomization
@@
 
-\hypertarget{ref-Hernan2004}{}
+\leavevmode\hypertarget{ref-Hernan2004}{}%
 31. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
@@
 
-\hypertarget{ref-Pierce2012}{}
+\leavevmode\hypertarget{ref-Pierce2012}{}%
 32. Pierce BL, VanderWeele TJ. The effect of non-differential
@@
 
-\hypertarget{ref-Ashenfelter1994}{}
+\leavevmode\hypertarget{ref-Ashenfelter1994}{}%
 33. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
@@
 
-\hypertarget{ref-Nagarajan2013}{}
+\leavevmode\hypertarget{ref-Nagarajan2013}{}%
 34. Nagarajan R, Scutari M. Impact of noise on molecular network
@@
 
-\hypertarget{ref-Shpitser2010}{}
+\leavevmode\hypertarget{ref-Shpitser2010}{}%
 35. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
@@
 
-\hypertarget{ref-Wang2015}{}
+\leavevmode\hypertarget{ref-Wang2015}{}%
 36. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
@@
 
-\hypertarget{ref-Lagani2015}{}
+\leavevmode\hypertarget{ref-Lagani2015}{}%
 37. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
@@
 
-\hypertarget{ref-Lawlor2017}{}
+\leavevmode\hypertarget{ref-Lawlor2017}{}%
 38. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
@@
 
-\hypertarget{ref-Sterne2001}{}
-39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
-significance tests? BMJ. 2001;322: 226--231.
-
-\hypertarget{ref-Henningsen2007}{}
-40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. Journal of Statistical Software.
-2007;23: 1--40.
-doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
-
-\hypertarget{ref-Steiger1980}{}
-41. Steiger JH. Tests for comparing elements of a correlation matrix.
-Psychological Bulletin. 1980;87: 245--251.
-doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
-
-\hypertarget{ref-Revelle2015}{}
-42. Revelle W. psych: Procedures for Psychological, Psychometric, and
-Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
-University; 2015. Available:
-\url{http://cran.r-project.org/package=psych}
-
-\hypertarget{ref-Shakhbazov2016}{}
-43. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+\leavevmode\hypertarget{ref-Shakhbazov2016}{}%
+39. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
 PM, et al. Shared genetic control of expression and methylation in
@@
 
-\hypertarget{ref-RCoreTeam2015}{}
-44. R Core Team. R: A Language and Environment for Statistical Computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2015. Available: \url{https://www.r-project.org/}
-
-\hypertarget{ref-Bird2002}{}
-45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+\leavevmode\hypertarget{ref-Bird2002}{}%
+40. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
 development. 2002;16: 6--21.
@@
 
-\hypertarget{ref-Cole2014}{}
-46. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+\leavevmode\hypertarget{ref-Cole2014}{}%
+41. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
 Serious and Misleading Consequences Due to Uncorrected Measurement
@@
 
-\hypertarget{ref-Bose2014}{}
-47. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+\leavevmode\hypertarget{ref-Bose2014}{}%
+42. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
 Evaluation of microarray-based DNA methylation measurement using
@@
 
-\hypertarget{ref-Bryant2011}{}
-48. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+\leavevmode\hypertarget{ref-Bryant2011}{}%
+43. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
 et al. Technical Variability Is Greater than Biological Variability in a
@@
 
-\hypertarget{ref-Zhu2016}{}
-49. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+\leavevmode\hypertarget{ref-Zhu2016}{}%
+44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 
-\hypertarget{ref-Mancuso2017}{}
-50. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+\leavevmode\hypertarget{ref-Mancuso2017}{}%
+45. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
 Integrating Gene Expression with Summary Association Statistics to
@@
 
-\hypertarget{ref-Lee2013c}{}
-51. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+\leavevmode\hypertarget{ref-Lee2013c}{}%
+46. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 
-\hypertarget{ref-Do2013}{}
-52. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+\leavevmode\hypertarget{ref-Do2013}{}%
+47. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
 Common variants associated with plasma triglycerides and risk for
@@
 
-\hypertarget{ref-Burgess2014a}{}
-53. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+\leavevmode\hypertarget{ref-Burgess2014a}{}%
+48. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 
-\hypertarget{ref-Relton2012}{}
-54. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+\leavevmode\hypertarget{ref-Relton2012}{}%
+49. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 
-\hypertarget{ref-Varbo2015}{}
-55. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+\leavevmode\hypertarget{ref-Varbo2015}{}%
+50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 
-\hypertarget{ref-Burgess2015}{}
-56. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+\leavevmode\hypertarget{ref-Burgess2015}{}%
+51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 
-\hypertarget{ref-Richmond2016}{}
-57. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+\leavevmode\hypertarget{ref-Richmond2016}{}%
+52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.
@@
 
+\leavevmode\hypertarget{ref-Sterne2001}{}%
+53. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+significance tests? BMJ. 2001;322: 226--231.
+
+\leavevmode\hypertarget{ref-Henningsen2007}{}%
+54. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. Journal of Statistical Software.
+2007;23: 1--40.
+doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
+
+\leavevmode\hypertarget{ref-Steiger1980}{}%
+55. Steiger JH. Tests for comparing elements of a correlation matrix.
+Psychological Bulletin. 1980;87: 245--251.
+doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
+
+\leavevmode\hypertarget{ref-Revelle2015}{}%
+56. Revelle W. psych: Procedures for Psychological, Psychometric, and
+Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
+University; 2015. Available:
+\url{http://cran.r-project.org/package=psych}
+
+\leavevmode\hypertarget{ref-RCoreTeam2015}{}%
+57. R Core Team. R: A Language and Environment for Statistical Computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2015. Available: \url{https://www.r-project.org/}
+
 \end{document}

### 2017-11-20 scripts/mr_directionality_analysis.R

@@ -775,7 +775,7 @@ qtl <- group_by(me, rxx_o, ryy_o) %>%
 		y <- qtl_orig
 		y$rxx_o <- x$rxx_o[1]
 		y$ryy_o <- x$ryy_o[1]
-		return(y)
+		y
 	})
 
 qtl$r_exp <- NA
@@ -931,7 +931,8 @@ p1 <- ggplot(subset(temp, key=="n"), aes(x=as.factor(rxx_o), y=value)) +
 geom_bar(stat="Identity", aes(fill=dir), position="stack") +
 geom_hline(yintercept = nrow(qtl_orig)/2, linetype="dotted") +
 facet_grid(. ~ ryy_o_lab) +
-labs(x="cor(E,Eo)", y="Count", fill=NULL, title="a)")
+labs(x="cor(E,Eo)", y="Count", fill=NULL, title="a)") +
+theme(strip.text=element_text(size=7), axis.text.x=element_text(size=7))
 
 
 p2 <- ggMarginal(
@@ -955,8 +956,9 @@ geom_density(aes(fill=dir), alpha=0.9) +
 labs(fill=NULL, x=TeX("$\\beta_{MR}$"), title="c)") +
 theme(legend.position="none")
 
+tiff("Fig6-2.tif", width=15.93, height=10.75, unit="cm", res=110)
 grid.arrange(p1, p3, p4, ncol=2, layout_matrix=rbind(c(1,1), c(2,3)))
-
+dev.off()
 
 ## ---- shakhtab ----
 

### 2017-11-20 docs/manuscript.tex

--- 2017-10-24 docs/manuscript.tex
@@
 
-\subsection{Model}\label{model}
+\subsubsection{Model}\label{model}
 
@@
 measurement error in the exposure and the outcome are uncorrelated.
-
-\subsection{Methods}\label{methods}
-
-\subsubsection{CIT test}\label{cit-test}
-
-First we describe how the CIT method {[}4{]} is implemented in the R
-package \emph{R/cit} {[}18{]}. The methodology of the CIT is as follows.
-Assume an exposure \(x\) is instrumented by a SNP \(g\), and the
-exposure \(x\) causes an outcome \(y\), as described above. The
-following tests are then performed:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
-  associates with the outcome}
-\item
-  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
-  associates with the exposure conditional on the outcome}
-\item
-  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
-  associates with the outcome conditional on the SNP}
-\item
-  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
-  independent of the outcome conditional on the exposure}
-\end{enumerate}
-
-The term in the 4th test can be rewritten as a
-\(cov(g, y|x) = cov(g, y - \hat{y})\) where
-\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
-of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
-association between the SNP and the outcome. The condition in the 4th
-test is formulated as an equivalence testing problem that is estimated
-using simulations, comparing the estimate from the data against
-empirically obtained estimates for simulated variables where the
-independence model is true (full details are given in {[}4{]}). We note
-here that this approach is liable to fail, even when there is a true
-causal relationship, when confounders of the exposure and outcome are
-present, as these will induce collider bias.
-
-If all four tests reject the null hypothesis then it is inferred that
-\(x\) causes \(y\). The CIT measures the strength of causality by
-generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
-(least extreme) p-value of the four tests, the intuition being that
-causal inference is only as strong as the weakest link in the chain of
-tests.
-
-Now we describe how we used the CIT method in our simulations. The
-\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
-the direction of causality using the CIT method, an omnibus p-value
-generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
-was estimated for the direction of \(x\) causing \(y\) (Model 1), and
-for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
-(Model 2). The results from each of these methods can then be used in
-combination to infer the existance and direction of causality. For some
-significance threshold \(\alpha\) there are four possible outcomes from
-these two tests, and their interpretations are as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
-\item
-  If \(p_{CIT, x \rightarrow y} > \alpha\) and
-  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
-  relationship
-\item
-  If \(p_{CIT, x \rightarrow y} < \alpha\) and
-  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
-  confounding (S1 fig) and no call is made.
-\end{itemize}
-
-For the purposes of compiling simulation results we use an arbitrary
-\(\alpha = 0.05\) value, though we stress that for real analyses it is
-not good practice to rely on p-values for making causal inference, nor
-is it reliable to depend on arbitrary significance thresholds {[}39{]}.
-
-\subsubsection{MR causal test}\label{mr-causal-test}
-
-Two stage least squares (2SLS) is a commonly used technique for
-performing MR when the exposure, outcome and instrument data are all
-available in the same sample. A p-value for this test, \(p_{MR}\), was
-obtained using the \(systemfit\) function in the R package
-\(R/systemfit\) {[}40{]}. Note that the value of \(p_{MR}\) is identical
-when using the same genetic variant to instrument the influence of the
-exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
-outcome \(y\) on the exposure \(x\).
-
-The method that we will now describe is designed to distinguish between
-two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
-framework, this approach cannot infer if the true model is
-\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
-additive.
-
-To infer the direction of causality it is desirable to know which of the
-variables, \(x\) or \(y\), is being directly influenced by the
-instrument \(g\). This can be achieved by assessing which of the two
-variables has the biggest absolute correlation with \(g\) (S2 Text),
-formalised by testing for a difference in the correlations \(\rho_{gx}\)
-and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
-within a population {[}41{]}. It is calculated as
-
-\[
-Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
-\]
-
-where Fisher's z-transformation is used to obtain
-\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
-
-\[
-h = \frac{1 - (frm^2)} {1 - rm^2}
-\]
-
-where
-
-\[
-f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
-\]
-
-and
-
-\[
-rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
-\]
-
-The \(Z\) value is interpreted such that
-
-\[
-Z \left\{
-\begin{array}{ll}
-> 0, & x \to y\\
-< 0, & y \to x\\
-= 0, & x \perp\!\!\!\perp y 
-\end{array} \right.
-\]
-
-and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
-indicate the probability of obtaining a difference between correlations
-\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
-under the null hypothesis that both correlations are identical.
-
-The existence of causality and its direction is inferred based on
-combining information from the MR analysis and the Steiger test. The MR
-analysis indicates whether there is a potential causal relationship
-(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
-of the causal relationship and the confidence of the direction
-(\(p_{Steiger}\)). For the purposes of compiling simulation results,
-these can be combined using an arbitrary \(\alpha = 0.05\) value:
-
-\begin{itemize}
-\tightlist
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
-  a causal association for the correct model is accepted,
-  \(x \rightarrow y\)
-\item
-  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
-  a causal association for the incorrect model is accepted,
-  \(y \rightarrow x\)
-\item
-  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
-  model is accepted
-\end{itemize}
-
-Note that the same correlation test approach can be applied to a
-two-sample MR setting. Two-sample MR refers to the case where the
-SNP-exposure association and SNP-outcome association are calculated in
-different samples (e.g.~from publicly available summary statistics
-{[}26,30{]}). Here the Steiger test of two independent correlations can
-be applied where.
-
-\[
-Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
-\]
-
-An advantage of using the Steiger test in the two sample context is that
-it can compare correlations in independent samples where sample sizes
-are different. Steiger test statistics were calculated using the
-\emph{r.test} function in the R package \emph{R/psych} {[}42{]}.
-
-The Steiger test assumes that there is a causal relationship between the
-two variables, and that the SNP is a valid instrument for one of them.
-However it is liable to give incorrect causal directions under some
-other circumstances. First, some levels of horizontal pleiotropy, where
-the SNP influences the outcome through some pathway other than the
-exposure, could induce problems because this is a means by which the
-instrument is invalid. Second, some differential values of measurement
-error between the exposure and the outcome could lead to incorrect
-inference of the causal direction (S2 Text). Third, some levels of
-unmeasured confounding between the exposure and the outcome could lead
-to inference of the wrong causal direction (S3 Text).
-
-\subsubsection{Causal direction sensitivity analysis for measurement
-error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
-
-The Steiger test for inferring if \(x \rightarrow y\) is based on
-evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
-\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
-imprecisely. If, for example, \(x\) has lower measurement precision than
-\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
-because \(\rho_{g,x_o}\) could be underestimated more than
-\(\rho_{g,y_o}\).
-
-As we show in S2 Text it is possible to infer the bounds of measurement
-error on \(x_o\) or \(y_o\) given known genetic associations. The
-maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
-it is known that at least that much of the variance has been explained
-in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
-values (the same logic applies to \(y_o\)). It is possible to simulate
-what the inferred causal direction would be for all values within these
-bounds.
-
-To evaluate how reliable, \(R\), the inference of the causal direction
-is to potential measurement error in \(x\) and \(y\) we need to predict
-the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
-error. We offer two tools in which to do this. First, the user can
-provide values of measurement error for \(x\) and \(y\) and obtain a
-revised inference of the causal direction. Second, we integrate over the
-entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
-measurement error values, assuming that any measurement error value is
-equally likely. Across all possible values of measurement error in \(x\)
-and \(y\) we find the volume that agrees with the inferred direction of
-causality and the volume that disagrees with the inferred direction of
-causality, and take the ratio of these two values. A ratio \(R=1\)
-indicates that the inferred causal direction is highly sensitive to
-measurement error, because equal weight of the measurement error
-parameter space supports each direction of causality. In general, the
-\(R\) value denotes that the inferred direction of causality is \(R\)
-times more likely to be the empirical result than the opposite direction
-(S2 Text).
-
-\subsubsection{Simulations}\label{simulations}
-
-Simulations were conducted by creating variables of sample size \(n\)
-for the exposure \(x\), the measured values of the exposure \(x_o\), the
-outcome \(y\), the measured values of the outcome \(y_o\) and the
-instrument \(g\). One of two models are simulated, the ``causal model''
-where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
-``non-causal model'' where \(g\) influences a confounder \(u\) which in
-turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
-but not causally related. Each variable in the causal model was
-simulated such that:
-
-\[
-\begin{aligned}
-g & \sim Binom(2, 0.5) \\
-x & = \alpha_g + \beta_g g + \epsilon_g \\
-x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
-y & = \alpha_x + \beta_x x + \epsilon_x \\
-y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
-\end{aligned}
-\]
-
-where non-differential measurement error is represented by a noise
-(measurement imprecision) term
-\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
-\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
-the outcome variable \(y\). Note that following the first section of the
-Results we no longer include the bias terms for simplicity. We have
-formulated the non-causal model as:
-
-\[
-\begin{aligned}
-y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
-\end{aligned}
-\]
-
-All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
-Normally distributed values of \(\epsilon_*\) were generated such that
-
-\[
-\begin{aligned}
-cor(g, x)^2 & = 0.1 \\
-cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
-\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
-n & = \{100, 1000, 10000\}
-\end{aligned}
-\]
-
-giving a total of 432 combinations of parameters. Simulations using each
-of these sets of variables were performed 100 times, and the CIT and MR
-methods were applied to each in order to evaluate the causal association
-of the simulated variables. Similar patterns of results were obtained
-for different values of \(cor(g, x)\).
-
-\subsection{Applied example using two sample
-MR}\label{applied-example-using-two-sample-mr}
-
-Two sample MR {[}30{]} was performed using summary statistics for
-genetic influences on gene expression and DNA methylation. To do this we
-obtained a list of 458 gene expression - DNA methylation associations as
-reported in Shakhbazov et al {[}43{]}. These were filtered to be located
-on the same chromosome, have robust correlations after correcting for
-multiple testing, and to share a SNP that had a robust cis-acting effect
-on both the DNA methylation probe and the gene expression probe. Because
-only summary statistics were available (effect, standard error, effect
-allele, sample size, p-values) for the instrumental SNP on the
-methylation and gene expression levels, the Steiger test of two
-independent correlations was used to infer the direction of causality
-for each of the associations. The Wald ratio test was then used to
-estimate the causal effect size for the estimated direction for each
-association.
-
-All analysis was performed using the R programming language {[}44{]} and
-code is made available at
-\url{https://github.com/explodecomputer/causal-directions} and
-implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
-{[}26{]}.
 
@@
 at a controlling CpG typically leads to increased gene expression
-levels, consistent with expectation {[}45{]}.
+levels, consistent with expectation {[}40{]}.
 
@@
 robustness under varying levels of measurement error.
+
+\subsection{Methods}\label{methods}
+
+\subsubsection{CIT test}\label{cit-test}
+
+Here we describe how the CIT method {[}4{]} is implemented in the R
+package \emph{R/cit} {[}18{]}. Assume an exposure \(x\) is instrumented
+by a SNP \(g\), and the exposure \(x\) causes an outcome \(y\), as
+described above. The following tests are then performed:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
+  associates with the outcome}
+\item
+  \(H_0: cov(g, x|y) = 0; H_1: cov(g, x|y) \neq 0\); \emph{the SNP
+  associates with the exposure conditional on the outcome}
+\item
+  \(H_0: cov(x, y|g) = 0; H_1: cov(x, y|g) \neq 0\); \emph{the exposure
+  associates with the outcome conditional on the SNP}
+\item
+  \(H_0: cov(g, y|x) \neq 0; H_1: cov(g, y|x) = 0\); \emph{the SNP is
+  independent of the outcome conditional on the exposure}
+\end{enumerate}
+
+The term in the 4th test can be rewritten as
+\(cov(g, y|x) = cov(g, y - \hat{y})\) where
+\(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the residual
+of \(y\) after adjusting for \(x\), and \(x\) is assumed to mediate the
+association between the SNP and the outcome. The condition in the 4th
+test is formulated as an equivalence testing problem that is estimated
+using simulations, comparing the estimate from the data against
+empirically obtained estimates for simulated variables where the
+independence model is true (full details are given in {[}4{]}). We note
+here that this approach is liable to fail, even when there is a true
+causal relationship, when confounders of the exposure and outcome are
+present, as these will induce collider bias.
+
+If all four tests reject the null hypothesis then it is inferred that
+\(x\) causes \(y\). The CIT measures the strength of causality by
+generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
+(least extreme) p-value of the four tests, the intuition being that
+causal inference is only as strong as the weakest link in the chain of
+tests.
+
+Now we describe how we used the CIT method in our simulations. The
+\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
+the direction of causality using the CIT method, an omnibus p-value
+generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
+was estimated for the direction of \(x\) causing \(y\) (Model 1), and
+for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
+(Model 2). The results from each of these methods can then be used in
+combination to infer the existance and direction of causality. For some
+significance threshold \(\alpha\) there are four possible outcomes from
+these two tests, and their interpretations are as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
+\item
+  If \(p_{CIT, x \rightarrow y} > \alpha\) and
+  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
+  relationship
+\item
+  If \(p_{CIT, x \rightarrow y} < \alpha\) and
+  \(p_{CIT, y \rightarrow x} < \alpha\) then there is potentially
+  confounding (S1 fig) and no call is made.
+\end{itemize}
+
+For the purposes of compiling simulation results we use an arbitrary
+\(\alpha = 0.05\) value, though we stress that for real analyses it is
+not good practice to rely on p-values for making causal inference, nor
+is it reliable to depend on arbitrary significance thresholds {[}53{]}.
+
+\subsubsection{MR causal test}\label{mr-causal-test}
+
+Two stage least squares (2SLS) is a commonly used technique for
+performing MR when the exposure, outcome and instrument data are all
+available in the same sample. A p-value for this test, \(p_{MR}\), was
+obtained using the \(systemfit\) function in the R package
+\(R/systemfit\) {[}54{]}. Note that the value of \(p_{MR}\) is identical
+when using the same genetic variant to instrument the influence of the
+exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
+outcome \(y\) on the exposure \(x\).
+
+The method that we will now describe is designed to distinguish between
+two models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
+framework, this approach cannot infer if the true model is
+\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
+additive.
+
+To infer the direction of causality it is desirable to know which of the
+variables, \(x\) or \(y\), is being directly influenced by the
+instrument \(g\). This can be achieved by assessing which of the two
+variables has the biggest absolute correlation with \(g\) (S2 Text),
+formalised by testing for a difference in the correlations \(\rho_{gx}\)
+and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
+within a population {[}55{]}. It is calculated as
+
+\[
+Z = (Z_{gx} - Z_{gy}) \frac{\sqrt{N-3}}{\sqrt{2(1-\rho_{xy})h}}
+\]
+
+where Fisher's z-transformation is used to obtain
+\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),
+
+\[
+h = \frac{1 - (frm^2)} {1 - rm^2}
+\]
+
+where
+
+\[
+f = \frac{1 - \rho_{xy}}{2(1 - rm^2)}
+\]
+
+and
+
+\[
+rm^2 = \frac{1}{2}(\rho_{gx}^2 + \rho_{gy}^2).
+\]
+
+The \(Z\) value is interpreted such that
+
+\[
+Z \left\{
+\begin{array}{ll}
+> 0, & x \to y\\
+< 0, & y \to x\\
+= 0, & x \perp\!\!\!\perp y 
+\end{array} \right.
+\]
+
+and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
+indicate the probability of obtaining a difference between correlations
+\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
+under the null hypothesis that both correlations are identical.
+
+The existence of causality and its direction is inferred based on
+combining information from the MR analysis and the Steiger test. The MR
+analysis indicates whether there is a potential causal relationship
+(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
+of the causal relationship and the confidence of the direction
+(\(p_{Steiger}\)). For the purposes of compiling simulation results,
+these can be combined using an arbitrary \(\alpha = 0.05\) value:
+
+\begin{itemize}
+\tightlist
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
+  a causal association for the correct model is accepted,
+  \(x \rightarrow y\)
+\item
+  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
+  a causal association for the incorrect model is accepted,
+  \(y \rightarrow x\)
+\item
+  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
+  model is accepted
+\end{itemize}
+
+Note that the same correlation test approach can be applied to a
+two-sample MR setting. Two-sample MR refers to the case where the
+SNP-exposure association and SNP-outcome association are calculated in
+different samples (e.g.~from publicly available summary statistics
+{[}26,30{]}). Here the Steiger test of two independent correlations can
+be applied where.
+
+\[
+Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
+\]
+
+An advantage of using the Steiger test in the two sample context is that
+it can compare correlations in independent samples where sample sizes
+are different. Steiger test statistics were calculated using the
+\emph{r.test} function in the R package \emph{R/psych} {[}56{]}.
+
+The Steiger test assumes that there is a causal relationship between the
+two variables, and that the SNP is a valid instrument for one of them.
+However it is liable to give incorrect causal directions under some
+other circumstances. First, some levels of horizontal pleiotropy, where
+the SNP influences the outcome through some pathway other than the
+exposure, could induce problems because this is a means by which the
+instrument is invalid. Second, some differential values of measurement
+error between the exposure and the outcome could lead to incorrect
+inference of the causal direction (S2 Text). Third, some levels of
+unmeasured confounding between the exposure and the outcome could lead
+to inference of the wrong causal direction (S3 Text).
+
+\subsubsection{Causal direction sensitivity analysis for measurement
+error}\label{causal-direction-sensitivity-analysis-for-measurement-error}
+
+The Steiger test for inferring if \(x \rightarrow y\) is based on
+evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
+\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
+imprecisely. If, for example, \(x\) has lower measurement precision than
+\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
+because \(\rho_{g,x_o}\) could be underestimated more than
+\(\rho_{g,y_o}\).
+
+As we show in S2 Text it is possible to infer the bounds of measurement
+error on \(x_o\) or \(y_o\) given known genetic associations. The
+maximum measurement imprecision of \(x_o\) is \(\rho_{g,x_o}\), because
+it is known that at least that much of the variance has been explained
+in \(x_o\) by \(g\). The minimum is 0, denoting perfectly measured trait
+values (the same logic applies to \(y_o\)). It is possible to simulate
+what the inferred causal direction would be for all values within these
+bounds.
+
+To evaluate how reliable, \(R\), the inference of the causal direction
+is to potential measurement error in \(x\) and \(y\) we need to predict
+the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
+error. We offer two tools in which to do this. First, the user can
+provide values of measurement error for \(x\) and \(y\) and obtain a
+revised inference of the causal direction. Second, we integrate over the
+entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
+measurement error values, assuming that any measurement error value is
+equally likely. Across all possible values of measurement error in \(x\)
+and \(y\) we find the volume that agrees with the inferred direction of
+causality and the volume that disagrees with the inferred direction of
+causality, and take the ratio of these two values. A ratio \(R=1\)
+indicates that the inferred causal direction is highly sensitive to
+measurement error, because equal weight of the measurement error
+parameter space supports each direction of causality. In general, the
+\(R\) value denotes that the inferred direction of causality is \(R\)
+times more likely to be the empirical result than the opposite direction
+(S2 Text).
+
+\subsubsection{Simulations}\label{simulations}
+
+Simulations were conducted by creating variables of sample size \(n\)
+for the exposure \(x\), the measured values of the exposure \(x_o\), the
+outcome \(y\), the measured values of the outcome \(y_o\) and the
+instrument \(g\). One of two models are simulated, the ``causal model''
+where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
+``non-causal model'' where \(g\) influences a confounder \(u\) which in
+turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
+but not causally related. Each variable in the causal model was
+simulated such that:
+
+\[
+\begin{aligned}
+g & \sim Binom(2, 0.5) \\
+x & = \alpha_g + \beta_g g + \epsilon_g \\
+x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
+y & = \alpha_x + \beta_x x + \epsilon_x \\
+y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
+\end{aligned}
+\]
+
+where non-differential measurement error is represented by a noise
+(measurement imprecision) term
+\(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), and measurement bias terms
+\(\alpha_{m*}\) and \(\beta_{m*}\) for the exposure variable \(x\) and
+the outcome variable \(y\). Note that following the first section of the
+Results we no longer include the bias terms for simplicity. We have
+formulated the non-causal model as:
+
+\[
+\begin{aligned}
+y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
+\end{aligned}
+\]
+
+All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
+Normally distributed values of \(\epsilon_*\) were generated such that
+
+\[
+\begin{aligned}
+cor(g, x)^2 & = 0.1 \\
+cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
+\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
+n & = \{100, 1000, 10000\}
+\end{aligned}
+\]
+
+giving a total of 432 combinations of parameters. Simulations using each
+of these sets of variables were performed 100 times, and the CIT and MR
+methods were applied to each in order to evaluate the causal association
+of the simulated variables. Similar patterns of results were obtained
+for different values of \(cor(g, x)\).
+
+\subsubsection{Applied example using two sample
+MR}\label{applied-example-using-two-sample-mr}
+
+Two sample MR {[}30{]} was performed using summary statistics for
+genetic influences on gene expression and DNA methylation. To do this we
+obtained a list of 458 gene expression - DNA methylation associations as
+reported in Shakhbazov et al {[}39{]}. These were filtered to be located
+on the same chromosome, have robust correlations after correcting for
+multiple testing, and to share a SNP that had a robust cis-acting effect
+on both the DNA methylation probe and the gene expression probe. Because
+only summary statistics were available (effect, standard error, effect
+allele, sample size, p-values) for the instrumental SNP on the
+methylation and gene expression levels, the Steiger test of two
+independent correlations was used to infer the direction of causality
+for each of the associations. The Wald ratio test was then used to
+estimate the causal effect size for the estimated direction for each
+association.
+
+All analysis was performed using the R programming language {[}57{]} and
+code is made available at
+\url{https://github.com/explodecomputer/causal-directions} and
+implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform
+{[}26{]}.
 
@@
 
-\hypertarget{ref-Sterne2001}{}
-39. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
-significance tests? BMJ. 2001;322: 226--231.
-
-\hypertarget{ref-Henningsen2007}{}
-40. Henningsen A, Hamann JD. systemfit : A Package for Estimating
-Systems of Simultaneous Equations in R. Journal of Statistical Software.
-2007;23: 1--40.
-doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
-
-\hypertarget{ref-Steiger1980}{}
-41. Steiger JH. Tests for comparing elements of a correlation matrix.
-Psychological Bulletin. 1980;87: 245--251.
-doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
-
-\hypertarget{ref-Revelle2015}{}
-42. Revelle W. psych: Procedures for Psychological, Psychometric, and
-Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
-University; 2015. Available:
-\url{http://cran.r-project.org/package=psych}
-
 \hypertarget{ref-Shakhbazov2016}{}
-43. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
+39. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
 PM, et al. Shared genetic control of expression and methylation in
@@
 
-\hypertarget{ref-RCoreTeam2015}{}
-44. R Core Team. R: A Language and Environment for Statistical Computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2015. Available: \url{https://www.r-project.org/}
-
 \hypertarget{ref-Bird2002}{}
-45. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
+40. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
 development. 2002;16: 6--21.
@@
 \hypertarget{ref-Cole2014}{}
-46. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
+41. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
 Serious and Misleading Consequences Due to Uncorrected Measurement
@@
 \hypertarget{ref-Bose2014}{}
-47. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
+42. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
 Evaluation of microarray-based DNA methylation measurement using
@@
 \hypertarget{ref-Bryant2011}{}
-48. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
+43. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
 et al. Technical Variability Is Greater than Biological Variability in a
@@
 \hypertarget{ref-Zhu2016}{}
-49. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
+44. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
 Integration of summary data from GWAS and eQTL studies predicts complex
@@
 \hypertarget{ref-Mancuso2017}{}
-50. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
+45. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
 Integrating Gene Expression with Summary Association Statistics to
@@
 \hypertarget{ref-Lee2013c}{}
-51. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
+46. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
 association studies: quantification of power and genomic prediction
@@
 \hypertarget{ref-Do2013}{}
-52. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
+47. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
 Common variants associated with plasma triglycerides and risk for
@@
 \hypertarget{ref-Burgess2014a}{}
-53. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
+48. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
 multivariable Mendelian randomization to disentangle the causal effects
@@
 \hypertarget{ref-Relton2012}{}
-54. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
+49. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
 randomization: a strategy for establishing the causal role of epigenetic
@@
 \hypertarget{ref-Varbo2015}{}
-55. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
+50. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
 Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
@@
 \hypertarget{ref-Burgess2015}{}
-56. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
+51. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
 randomization: using genetic variants as instrumental variables to
@@
 \hypertarget{ref-Richmond2016}{}
-57. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
+52. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
 Challenges and novel approaches for investigating molecular mediation.
@@
 
+\hypertarget{ref-Sterne2001}{}
+53. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
+significance tests? BMJ. 2001;322: 226--231.
+
+\hypertarget{ref-Henningsen2007}{}
+54. Henningsen A, Hamann JD. systemfit : A Package for Estimating
+Systems of Simultaneous Equations in R. Journal of Statistical Software.
+2007;23: 1--40.
+doi:\href{https://doi.org/10.18637/jss.v023.i04}{10.18637/jss.v023.i04}
+
+\hypertarget{ref-Steiger1980}{}
+55. Steiger JH. Tests for comparing elements of a correlation matrix.
+Psychological Bulletin. 1980;87: 245--251.
+doi:\href{https://doi.org/10.1037/0033-2909.87.2.245}{10.1037/0033-2909.87.2.245}
+
+\hypertarget{ref-Revelle2015}{}
+56. Revelle W. psych: Procedures for Psychological, Psychometric, and
+Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
+University; 2015. Available:
+\url{http://cran.r-project.org/package=psych}
+
+\hypertarget{ref-RCoreTeam2015}{}
+57. R Core Team. R: A Language and Environment for Statistical Computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2015. Available: \url{https://www.r-project.org/}
+
 \end{document}
